---
title: "Bulk RNA-seq differential expression with edgeR"
format:
  html:
    toc: true
    code-fold: true
execute:
  freeze: false
---

> Status: `draft`
>
> Template class: API / TUTORIAL REMINDER

## What it does

This is a compact reminder of the current edgeR quasi-likelihood workflow for
testing an explicit biological contrast from a filtered, normalized `DGEList`.
It is intentionally short because no complete reusable bulk differential-
expression implementation was found in the source repository.

::: {.callout-note}
## API reminder

No complete reusable bulk edgeR differential-expression workflow was found in
the source repository when this template was created.

This notebook is intentionally a concise reminder of the current
quasi-likelihood edgeR workflow. Check the current official edgeR User's Guide
before applying it to a new study, especially for design matrices, contrasts,
complex experiments, and filtering.
:::

## When to use it

Use this after `qc_normalization_edger.qmd` has produced a filtered `DGEList`
that still contains raw counts and edgeR normalization factors. The GLM uses
the count model; it does not accept z-scores, arbitrary normalized matrices, or
logCPM in place of counts.

## Parameters and input

```{r parameters}
input_dge <- "output/objects/dge_qc_normalized.qs2"
metadata_path <- "input/sample_metadata.tsv"
sample_id_column <- "sample_id"

# Edit these to match the actual design and the desired biological comparison.
design_formula <- stats::as.formula("~ 0 + group")
contrast_vector <- c(groupcase = 1, groupcontrol = -1)

fdr_cutoff <- 0.05
lfc_cutoff <- 1
output_results <- "output/tables/edgeR_qlf_results.tsv"
```

## Minimal current API

```{r edgeR-quasi-likelihood-workflow}
dge <- qs2::qs_read(input_dge)
metadata <- readr::read_tsv(metadata_path, show_col_types = FALSE)

stopifnot(sample_id_column %in% names(metadata))
metadata <- metadata[match(colnames(dge), metadata[[sample_id_column]]), , drop = FALSE]
stopifnot(identical(metadata[[sample_id_column]], colnames(dge)))
rownames(metadata) <- metadata[[sample_id_column]]

design <- stats::model.matrix(design_formula, data = metadata)
stopifnot(identical(rownames(design), colnames(dge)))
stopifnot(all(names(contrast_vector) %in% colnames(design)))

dge <- edgeR::estimateDisp(dge, design)
fit <- edgeR::glmQLFit(dge, design)
qlf <- edgeR::glmQLFTest(fit, contrast = contrast_vector)

result_tbl <- edgeR::topTags(
  qlf,
  n = nrow(dge),
  sort.by = "PValue"
)$table |>
  tibble::rownames_to_column("gene_id")

result_tbl <- result_tbl |>
  dplyr::mutate(
    reported = FDR <= fdr_cutoff & abs(logFC) >= lfc_cutoff
  )

dir.create(dirname(output_results), recursive = TRUE, showWarnings = FALSE)
readr::write_tsv(result_tbl, output_results)
```

## Main decisions

The design matrix is the scientific model, not boilerplate. Add batch,
pairing, blocking, or other covariates only when justified by the experimental
design and supported by the available replication. The contrast vector defines
the direction of `logFC`; inspect its names and signs before interpreting the
results.

Biological replication is required for dispersion estimation and inference.
Technical replicates should be handled according to the experimental design,
not counted as independent biological samples. `fdr_cutoff` and `lfc_cutoff`
are reporting choices in this reminder, not universal edgeR definitions.

## Outputs

The main output is a gene-level TSV containing the edgeR quasi-likelihood
results and an explicit reporting flag. The fitted objects are kept visible in
the notebook for inspection; persistence of a fitted model is optional and
should be added only when it is useful for a real analysis.

For GSEA, pass a complete ranked gene-level statistic table to
`gsea.qmd`. Do not silently reduce the ranking to only genes that passed a
significance cutoff.
