---
title: "Ranked gene-set enrichment analysis"
format:
  html:
    toc: true
    code-fold: true
execute:
  freeze: false
---

> Status: `draft`
>
> Template class: SOURCE-BACKED WORKFLOW

## Purpose

Run GO Biological Process and optional KEGG GSEA from a complete ranked
gene-level statistic table. This notebook consumes differential-expression
results; it does not rerun differential expression.

::: {.callout-note}
## Practical note

The ranking statistic is the key GSEA input. Filtering a ranked list to only
genes that passed an arbitrary significance cutoff discards information and
changes the question. I prefer to build the ranking from the full available
gene-level result table, then document the statistic and identifier conversion.
:::

## Inputs and parameters

The ranking table must contain one gene identifier column and one signed
numeric statistic column. This template maps `SYMBOL` or `ENSEMBL` identifiers
to Entrez IDs with `clusterProfiler::bitr()`; an `ENTREZID` input can be used
directly. The annotation database and organism must match the input species.

```{r setup}
library(clusterProfiler)
library(org.Hs.eg.db)
```

```{r parameters}
ranking_path <- "input/gene_ranking.tsv"
gene_id_column <- "gene_id"
statistic_column <- "rank_statistic"
input_id_type <- "SYMBOL"  # SYMBOL, ENSEMBL, or ENTREZID
annotation_db <- org.Hs.eg.db
organism_code <- "hsa"

run_go <- TRUE
run_kegg <- TRUE
ontology <- "BP"
min_gene_set_size <- 10
max_gene_set_size <- 500
pvalue_cutoff <- 0.05
p_adjust_method <- "BH"
simplify_go_terms <- TRUE
simplify_cutoff <- 0.7

output_table_dir <- "output/tables"
```

## Prepare the ranked list

```{r prepare-ranked-list}
stopifnot(input_id_type %in% c("SYMBOL", "ENSEMBL", "ENTREZID"))

ranking_tbl <- readr::read_tsv(ranking_path, show_col_types = FALSE)
stopifnot(all(c(gene_id_column, statistic_column) %in% names(ranking_tbl)))

ranking_tbl <- ranking_tbl |>
  dplyr::transmute(
    gene_id = as.character(.data[[gene_id_column]]),
    rank_statistic = as.numeric(.data[[statistic_column]])
  ) |>
  dplyr::filter(
    !is.na(gene_id),
    nzchar(gene_id),
    is.finite(rank_statistic)
  )

stopifnot(nrow(ranking_tbl) > 0L)

if (input_id_type == "ENTREZID") {
  mapped_tbl <- ranking_tbl |>
    dplyr::mutate(ENTREZID = gene_id)
} else {
  id_map <- clusterProfiler::bitr(
    unique(ranking_tbl$gene_id),
    fromType = input_id_type,
    toType = "ENTREZID",
    OrgDb = annotation_db
  )
  names(id_map)[names(id_map) == input_id_type] <- "gene_id"
  mapped_tbl <- ranking_tbl |>
    dplyr::inner_join(
      id_map |> dplyr::select(gene_id, ENTREZID),
      by = "gene_id"
    )
}

# Keep one mapped value per Entrez ID, retaining the strongest absolute statistic.
ranked_entrez <- mapped_tbl |>
  dplyr::filter(!is.na(ENTREZID), is.finite(rank_statistic)) |>
  dplyr::group_by(ENTREZID) |>
  dplyr::slice_max(abs(rank_statistic), n = 1, with_ties = FALSE) |>
  dplyr::ungroup() |>
  dplyr::arrange(dplyr::desc(rank_statistic))

gene_list <- ranked_entrez$rank_statistic
names(gene_list) <- ranked_entrez$ENTREZID
gene_list <- sort(gene_list, decreasing = TRUE)
stopifnot(!anyDuplicated(names(gene_list)))

dir.create(output_table_dir, recursive = TRUE, showWarnings = FALSE)
readr::write_tsv(ranked_entrez, file.path(output_table_dir, "gsea_ranked_gene_list.tsv"))
```

## GO and KEGG GSEA

```{r run-gsea}
if (run_go) {
  go_result <- clusterProfiler::gseGO(
    geneList = gene_list,
    OrgDb = annotation_db,
    keyType = "ENTREZID",
    ont = ontology,
    minGSSize = min_gene_set_size,
    maxGSSize = max_gene_set_size,
    pvalueCutoff = pvalue_cutoff,
    pAdjustMethod = p_adjust_method,
    verbose = FALSE
  )

  go_table <- as.data.frame(go_result)
  readr::write_tsv(go_table, file.path(output_table_dir, "gsea_go_results.tsv"))

  if (simplify_go_terms && nrow(go_table) > 0L) {
    go_simplified <- clusterProfiler::simplify(
      go_result,
      cutoff = simplify_cutoff,
      by = "p.adjust",
      select_fun = min
    )
    readr::write_tsv(
      as.data.frame(go_simplified),
      file.path(output_table_dir, "gsea_go_simplified_results.tsv")
    )
  }
}

if (run_kegg) {
  kegg_result <- clusterProfiler::gseKEGG(
    geneList = gene_list,
    organism = organism_code,
    keyType = "ncbi-geneid",
    minGSSize = min_gene_set_size,
    maxGSSize = max_gene_set_size,
    pvalueCutoff = pvalue_cutoff,
    pAdjustMethod = p_adjust_method,
    verbose = FALSE
  )
  readr::write_tsv(
    as.data.frame(kegg_result),
    file.path(output_table_dir, "gsea_kegg_results.tsv")
  )
}
```

## Outputs

The ranked list and enrichment result tables are exported as TSV. GO
simplification is a redundancy-reduction step for GO terms; it does not change
the original GSEA question. KEGG results are kept separate because KEGG does
not use the GO semantic hierarchy.

## Method notes

`logFC`, a signed test statistic, or a source-specific statistic such as
`sign(logFC) * sqrt(F)` can be appropriate rankings, but they are not
interchangeable. Select the statistic that matches the upstream model and
record that choice in the input table or analysis notes.

Missing identifiers are removed before enrichment and duplicate mapped IDs are
resolved by retaining the largest absolute ranking statistic. This is a
practical tie-breaking rule, not a biological claim. Change the annotation
database and KEGG organism code for another species, then consult the current
clusterProfiler documentation for advanced databases, methods, and parameters.
