---
title: "Sample-level gene-set activity with GSVA"
format:
  html:
    toc: true
    code-fold: true
execute:
  freeze: false
---

> Status: `draft`
>
> Template class: SOURCE-BACKED WORKFLOW

## Purpose

Transform a gene-by-sample expression matrix into a gene-set-by-sample
activity matrix with GSVA. The output is a sample-level activity score for
each supplied gene set; it is not an edgeR test or a ranked-list GSEA result.

::: {.callout-note}
## Source-backed workflow

The source scoring notebook uses the current parameter-object style
`GSVA::gsvaParam()` followed by `GSVA::gsva()`. The canonical template keeps
that method-specific route and makes gene-set coverage and input scale visible.
:::

## Inputs and parameters

The expression table must have genes in rows, samples in columns, and a first
column containing unique gene IDs. The gene-set table must contain `set_id`
and `gene_id`. Use a continuous normalized expression representation for the
default Gaussian kernel; do not pass arbitrary z-score or count transforms
without deciding whether the selected kernel is appropriate.

```{r parameters}
expression_path <- "input/normalized_expression.tsv"
gene_set_path <- "input/gene_sets.tsv"
expression_gene_column <- "gene_id"
gene_set_id_column <- "set_id"
gene_set_gene_column <- "gene_id"

kcdf <- "Gaussian"
minimum_gene_set_size <- 5L
maximum_gene_set_size <- 500L

output_score_table <- "output/tables/gsva_scores.tsv"
output_coverage_table <- "output/tables/gsva_gene_set_coverage.tsv"
```

## Analysis

```{r load-expression}
expression_tbl <- readr::read_tsv(expression_path, show_col_types = FALSE)
stopifnot(expression_gene_column %in% names(expression_tbl))

expr_mat <- expression_tbl |>
  tibble::column_to_rownames(expression_gene_column) |>
  as.matrix()
storage.mode(expr_mat) <- "numeric"

stopifnot(nrow(expr_mat) > 0L, ncol(expr_mat) > 0L)
stopifnot(!anyDuplicated(rownames(expr_mat)))
stopifnot(!anyDuplicated(colnames(expr_mat)))
stopifnot(all(is.finite(expr_mat)))
```

```{r gene-set-coverage}
gene_set_tbl <- readr::read_tsv(gene_set_path, show_col_types = FALSE)
stopifnot(all(c(gene_set_id_column, gene_set_gene_column) %in% names(gene_set_tbl)))

gene_set_tbl <- gene_set_tbl |>
  dplyr::transmute(
    set_id = as.character(.data[[gene_set_id_column]]),
    gene_id = as.character(.data[[gene_set_gene_column]])
  ) |>
  dplyr::filter(
    !is.na(set_id), nzchar(set_id),
    !is.na(gene_id), nzchar(gene_id)
  ) |>
  dplyr::distinct()

gene_sets <- split(gene_set_tbl$gene_id, gene_set_tbl$set_id)
gene_sets_found <- lapply(gene_sets, intersect, y = rownames(expr_mat))

coverage_table <- tibble::tibble(
  set_id = names(gene_sets),
  requested = lengths(gene_sets),
  found = lengths(gene_sets_found)
) |>
  dplyr::mutate(
    missing = requested - found,
    coverage = found / requested
  )

gene_sets_found <- gene_sets_found[
  lengths(gene_sets_found) >= minimum_gene_set_size
]
stopifnot(length(gene_sets_found) > 0L)
```

```{r run-gsva}
param <- GSVA::gsvaParam(
  exprData = expr_mat,
  geneSets = gene_sets_found,
  kcdf = kcdf,
  minSize = minimum_gene_set_size,
  maxSize = maximum_gene_set_size
)

gsva_result <- GSVA::gsva(param)
score_mat <- if (inherits(gsva_result, "SummarizedExperiment")) {
  SummarizedExperiment::assay(gsva_result, "es")
} else {
  as.matrix(gsva_result)
}

score_table <- as.data.frame(score_mat) |>
  tibble::rownames_to_column("gene_set")

dir.create(dirname(output_score_table), recursive = TRUE, showWarnings = FALSE)
dir.create(dirname(output_coverage_table), recursive = TRUE, showWarnings = FALSE)
readr::write_tsv(score_table, output_score_table)
readr::write_tsv(coverage_table, output_coverage_table)
```

## Outputs

`gsva_scores.tsv` contains one row per gene set and one column per sample.
`gsva_gene_set_coverage.tsv` records requested and measured genes before GSVA
filters gene sets by the declared minimum size.

## Method notes

The source uses Gaussian-kernel GSVA on a transformed expression matrix. The
canonical default keeps `kcdf = "Gaussian"` visible for continuous normalized
input. Count-scale input and other kernels are different methodological
choices; consult the current GSVA documentation before changing them.

GSVA and singscore answer related but different questions. GSVA estimates
sample-level gene-set activity from the expression distribution, whereas
singscore uses within-sample ranks. Do not merge their outputs into an unnamed
generic pathway score.

This notebook writes the score matrix and coverage diagnostics as TSV. A
separate qs2 object is not serialized because the score matrix is already the
useful analytical handoff.
