---
title: "Bulk RNA-seq QC and normalization with edgeR"
format:
  html:
    toc: true
    code-fold: true
execute:
  freeze: false
---

> Status: `draft`
>
> Template class: SOURCE-BACKED WORKFLOW

## Purpose

Inspect sample-level bulk RNA-seq quality, filter lowly expressed genes,
estimate library-composition normalization factors, and create a normalized
logCPM representation while retaining the filtered raw counts for edgeR
modeling.

::: {.callout-note}
## Source-backed workflow

This is the detailed QC baseline reconstructed from the source RNA-seq
workflow. It keeps library size, detected genes, expression distributions,
sample correlation, PCA, and clustering visible before any sample is removed.
:::

::: {.callout-warning}
## QC is not an automatic deletion rule

An unusual sample is a reason to investigate. It should not be removed because
of one plot or one threshold without considering library size, detected genes,
expression distributions, correlation, PCA, the experimental design, and the
source of the outlier signal.
:::

## Inputs

The input is a qs2-persisted edgeR `DGEList`, normally produced by
`tximport_edger.qmd`. Its counts must still be raw count-scale values. Metadata
may be attached in `dge$samples`; the first column is expected to remain the
edgeR library-size column and the row names must be sample IDs.

## Parameters

```{r parameters}
input_dge <- "output/objects/dge_imported.qs2"
output_dge <- "output/objects/dge_qc_normalized.qs2"
output_table_dir <- "output/tables"

min_cpm <- 1
min_samples <- 5
normalization_method <- "TMMwsp"
prior_count <- 1

# Optional column in dge$samples used only to color the PCA display.
pca_color_column <- NULL
```

## Analysis

```{r load-and-check-counts}
dge <- qs2::qs_read(input_dge)
stopifnot(inherits(dge, "DGEList"))
stopifnot(!is.null(dge$counts))
stopifnot(!is.null(rownames(dge$counts)))
stopifnot(!is.null(colnames(dge$counts)))
stopifnot(!anyDuplicated(rownames(dge$counts)))
stopifnot(!anyDuplicated(colnames(dge$counts)))
stopifnot(all(is.finite(dge$counts)))
stopifnot(all(dge$counts >= 0))

counts_before <- dge$counts
sample_ids <- colnames(counts_before)

qc_before <- tibble::tibble(
  sample_id = sample_ids,
  library_size = colSums(counts_before),
  detected_genes = colSums(counts_before > 0)
)
```

```{r filter-counts}
keep_gene <- rowSums(edgeR::cpm(dge) > min_cpm) >= min_samples

filter_accounting <- tibble::tibble(
  metric = c("genes_before", "genes_retained", "genes_removed"),
  value = c(
    nrow(dge),
    sum(keep_gene),
    sum(!keep_gene)
  )
)

# Keep the raw counts and library-size accounting consistent after filtering.
dge <- dge[keep_gene, , keep.lib.sizes = FALSE]
```

```{r normalize-and-logcpm}
dge <- edgeR::normLibSizes(dge, method = normalization_method)

# logCPM is for diagnostics, plots, correlations, and downstream scoring.
# The filtered raw counts remain in dge for edgeR likelihood modeling.
cpm_mat <- edgeR::cpm(dge)
logcpm_mat <- edgeR::cpm(
  dge,
  log = TRUE,
  prior.count = prior_count
)

qc_after <- tibble::tibble(
  sample_id = colnames(dge),
  library_size = colSums(dge$counts),
  detected_genes = colSums(dge$counts > 0),
  median_cpm = apply(cpm_mat, 2, stats::median),
  median_logCPM = apply(logcpm_mat, 2, stats::median),
  iqr_logCPM = apply(logcpm_mat, 2, stats::IQR),
  norm_factor = dge$samples$norm.factors
)

qc_sample_summary <- dplyr::left_join(
  qc_before,
  qc_after,
  by = "sample_id",
  suffix = c("_before", "_after")
)
```

## Diagnostics

### Library size and expression distributions

```{r distribution-diagnostics}
qc_sample_summary

boxplot(
  as.data.frame(logcpm_mat),
  las = 2,
  main = "TMMwsp-normalized logCPM by sample",
  ylab = "logCPM"
)

hist(
  qc_after$library_size,
  breaks = 20,
  main = "Library sizes after gene filtering",
  xlab = "Counts"
)
```

### Sample correlation and PCA

```{r correlation-pca}
stopifnot(nrow(logcpm_mat) >= 2L, ncol(logcpm_mat) >= 2L)

sample_correlation <- stats::cor(
  logcpm_mat,
  method = "spearman",
  use = "pairwise.complete.obs"
)

heatmap(
  sample_correlation,
  symm = TRUE,
  main = "Sample Spearman correlation from logCPM"
)

sample_correlation_for_review <- sample_correlation
diag(sample_correlation_for_review) <- NA_real_

pca_features <- apply(logcpm_mat, 1, stats::sd) > 0
stopifnot(any(pca_features))
pca <- stats::prcomp(
  t(logcpm_mat[pca_features, , drop = FALSE]),
  center = TRUE,
  scale. = TRUE
)

pca_coordinates <- as.data.frame(pca$x)
pca_coordinates$sample_id <- rownames(pca_coordinates)

pca_colors <- "grey40"
if (!is.null(pca_color_column)) {
  stopifnot(pca_color_column %in% names(dge$samples))
  pca_colors <- as.integer(as.factor(dge$samples[
    pca_coordinates$sample_id,
    pca_color_column
  ]))
}

plot(
  pca$x[, 1],
  pca$x[, 2],
  col = pca_colors,
  pch = 19,
  xlab = "PC1",
  ylab = "PC2",
  main = "PCA of normalized logCPM"
)
text(pca$x[, 1], pca$x[, 2], labels = rownames(pca$x), pos = 3, cex = 0.7)
```

### Clustering and outlier review

```{r clustering-diagnostic}
sample_hclust <- stats::hclust(
  stats::dist(t(logcpm_mat)),
  method = "complete"
)
plot(sample_hclust, main = "Sample clustering from normalized logCPM")

pca_distance <- sqrt(rowSums(
  sweep(pca$x[, seq_len(min(5, ncol(pca$x))), drop = FALSE], 2,
        colMeans(pca$x[, seq_len(min(5, ncol(pca$x))), drop = FALSE]), "-")^2
))

outlier_review <- tibble::tibble(
  sample_id = names(pca_distance),
  pca_distance = as.numeric(pca_distance),
  median_spearman = apply(sample_correlation_for_review, 1, stats::median, na.rm = TRUE)
)
outlier_review
```

## Outputs

```{r save-outputs}
dir.create(output_table_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(dirname(output_dge), recursive = TRUE, showWarnings = FALSE)

readr::write_tsv(filter_accounting, file.path(output_table_dir, "gene_filter_accounting.tsv"))
readr::write_tsv(qc_sample_summary, file.path(output_table_dir, "sample_qc_summary.tsv"))
readr::write_tsv(outlier_review, file.path(output_table_dir, "outlier_review.tsv"))
readr::write_tsv(
  as.data.frame(sample_correlation) |>
    tibble::rownames_to_column("sample_id"),
  file.path(output_table_dir, "sample_spearman_correlation.tsv")
)
readr::write_tsv(pca_coordinates, file.path(output_table_dir, "sample_pca_coordinates.tsv"))
readr::write_tsv(
  as.data.frame(cpm_mat) |>
    tibble::rownames_to_column("gene_id"),
  file.path(output_table_dir, "filtered_cpm.tsv")
)
readr::write_tsv(
  as.data.frame(logcpm_mat) |>
    tibble::rownames_to_column("gene_id"),
  file.path(output_table_dir, "filtered_logCPM.tsv")
)

qs2::qs_save(dge, output_dge)
```

The output DGEList contains filtered raw counts, edgeR library sizes,
normalization factors, and any non-conflicting sample metadata. The logCPM
table is a transformed expression representation; it is not a replacement for
raw-count likelihood modeling in edgeR.

## Method notes

The default `min_cpm = 1` in at least `min_samples = 5` samples is a visible
baseline, not a universal biological threshold. The source material contains
both a >1 CPM and a >5 CPM path, depending on the input branch. Thresholds
depend on library depth, sparsity, study design, and the downstream model.

`edgeR::filterByExpr()` is a useful design-aware alternative, but it is not
silently substituted here. If it is used, make the design and the change from
the CPM baseline explicit and check the current edgeR guidance.

`TMMwsp` corrects library-composition differences, especially in sparse data;
it is not batch correction. Batch and biological covariates belong in the
downstream design matrix.

I prefer reviewing several sample-level diagnostics together rather than
removing a sample because it appears unusual in one plot. This is a workflow
preference supported by the source QC practice, not an automatic outlier rule.

The source notebook also contains MAD-based low-tail thresholds and a
consensus outlier flag. Those are useful extensions to inspect for a specific
study, but they are not silently imposed here because the source has more than
one threshold path and sample exclusion still requires design-aware review.
