---
title: "Sample-level signature scoring with singscore"
format:
  html:
    toc: true
    code-fold: true
execute:
  freeze: false
---

> Status: `draft`
>
> Template class: SOURCE-BACKED WORKFLOW

## Purpose

Score a directional gene signature separately in each sample using within-
sample gene ranks. This is a sample-level score, not a gene-set enrichment
p-value and not differential expression.

::: {.callout-note}
## Source-backed workflow

The source Hallmark and sample-scoring notebooks use `rankGenes()` followed by
`simpleScore()`. The canonical version keeps the useful coverage checks and
up/down components while making the signature and expression input generic.
:::

## Inputs and parameters

Use an expression matrix with genes in rows and samples in columns. The matrix
should be a comparable expression representation such as logCPM, logTPM, or
another declared normalized scale. The signature TSV must contain `gene_id`
and `direction`, with directions `UP` and `DOWN`.

```{r parameters}
expression_path <- "input/normalized_expression.tsv"
signature_path <- "input/signature.tsv"
gene_id_column <- "gene_id"
direction_column <- "direction"
center_score <- TRUE
minimum_genes_per_direction <- 1L

output_score_table <- "output/tables/singscore_scores.tsv"
output_coverage_table <- "output/tables/singscore_gene_coverage.tsv"
```

## Analysis

```{r load-expression}
expression_tbl <- readr::read_tsv(expression_path, show_col_types = FALSE)
stopifnot(gene_id_column %in% names(expression_tbl))

expr_mat <- expression_tbl |>
  tibble::column_to_rownames(gene_id_column) |>
  as.matrix()
storage.mode(expr_mat) <- "numeric"

stopifnot(nrow(expr_mat) > 0L, ncol(expr_mat) > 0L)
stopifnot(!anyDuplicated(rownames(expr_mat)))
stopifnot(!anyDuplicated(colnames(expr_mat)))
stopifnot(all(is.finite(expr_mat)))
```

```{r signature-coverage}
signature_tbl <- readr::read_tsv(signature_path, show_col_types = FALSE)
stopifnot(all(c(gene_id_column, direction_column) %in% names(signature_tbl)))

signature_tbl <- signature_tbl |>
  dplyr::transmute(
    gene_id = as.character(.data[[gene_id_column]]),
    direction = toupper(as.character(.data[[direction_column]]))
  ) |>
  dplyr::filter(
    !is.na(gene_id),
    nzchar(gene_id),
    direction %in% c("UP", "DOWN")
  ) |>
  dplyr::distinct()

up_genes <- intersect(
  signature_tbl$gene_id[signature_tbl$direction == "UP"],
  rownames(expr_mat)
)
down_genes <- intersect(
  signature_tbl$gene_id[signature_tbl$direction == "DOWN"],
  rownames(expr_mat)
)

coverage_table <- tibble::tibble(
  direction = c("UP", "DOWN"),
  requested = c(
    sum(signature_tbl$direction == "UP"),
    sum(signature_tbl$direction == "DOWN")
  ),
  found = c(length(up_genes), length(down_genes))
) |>
  dplyr::mutate(
    missing = requested - found,
    coverage = found / requested
  )

stopifnot(
  length(up_genes) >= minimum_genes_per_direction,
  length(down_genes) >= minimum_genes_per_direction
)
```

```{r score-samples}
rank_data <- singscore::rankGenes(expr_mat)

score_table <- singscore::simpleScore(
  rankData = rank_data,
  upSet = up_genes,
  downSet = down_genes,
  centerScore = center_score
) |>
  as.data.frame() |>
  tibble::rownames_to_column("sample_id")

stopifnot(identical(score_table$sample_id, colnames(expr_mat)))

dir.create(dirname(output_score_table), recursive = TRUE, showWarnings = FALSE)
dir.create(dirname(output_coverage_table), recursive = TRUE, showWarnings = FALSE)
readr::write_tsv(score_table, output_score_table)
readr::write_tsv(coverage_table, output_coverage_table)
```

## Outputs

The score TSV retains `TotalScore`, `UpScore`, `DownScore`, and the associated
dispersion columns returned by singscore. Preserve these components before any
optional downstream standardization or group comparison.

## Method notes

Singscore ranks the measured genes within each sample. It is therefore
scientifically distinct from ranked-list GSEA, GSVA, single-cell UCell, and
regulatory-activity methods such as decoupler. The direction of the signature
is explicit: up genes and down genes contribute separate components to the
directional score.

Signature coverage is a diagnostic, not proof that a score is biologically
valid. Low coverage, many tied values, inconsistent gene identifiers, or
platform-specific missingness should be investigated before interpretation.
The source example uses Hallmark signatures, but Hallmark is not mandatory in
this reusable template.
