---
title: "tximport to edgeR object"
format:
  html:
    toc: true
    code-fold: true
execute:
  freeze: false
---

> Status: `draft`
>
> Template class: SOURCE-BACKED WORKFLOW

## Purpose

Convert a gene-level `tximport` result into an edgeR `DGEList` with explicit
gene IDs, sample alignment, optional annotation, and visible library-size
checks. Filtering, normalization, and downstream modeling belong to the next
template.

::: {.callout-note}
## Source-backed workflow

This template preserves the tximport-to-edgeR object construction used in the
source RNA-seq workflow. It does not try to discover Salmon directories,
support every quantifier, or replace the current tximport documentation.
:::

## Inputs

The supported input is one `tximport` R object persisted as qs2. It should
contain a gene-level `counts` matrix and the usual tximport sample columns.
Sample metadata is a TSV with one row per sample and a unique `sample_id`
column. A gene annotation TSV is optional and must contain a unique `gene_id`
column when supplied.

## Parameters

```{r parameters}
tximport_path <- "input/tximport_gene.qs2"
metadata_path <- "input/sample_metadata.tsv"
gene_annotation_path <- NULL

strip_gene_version <- TRUE
remove_all_zero_genes <- FALSE

output_dge <- "output/objects/dge_imported.qs2"
output_library_sizes <- "output/tables/imported_library_sizes.tsv"
```

## Analysis

```{r load-and-validate-inputs}
metadata <- readr::read_tsv(metadata_path, show_col_types = FALSE)
stopifnot("sample_id" %in% names(metadata))
stopifnot(!anyDuplicated(metadata$sample_id))

txi <- qs2::qs_read(tximport_path)
stopifnot(all(c("counts", "countsFromAbundance") %in% names(txi)))
stopifnot(is.matrix(txi$counts))
stopifnot(!is.null(rownames(txi$counts)))
stopifnot(!is.null(colnames(txi$counts)))

raw_gene_ids <- rownames(txi$counts)
gene_ids <- if (strip_gene_version) {
  sub("\\..*$", "", raw_gene_ids)
} else {
  raw_gene_ids
}

stopifnot(all(nzchar(gene_ids)))
if (anyDuplicated(gene_ids)) {
  stop(
    "Gene IDs become duplicated after identifier cleanup. ",
    "Resolve the annotation upstream rather than silently collapsing tximport rows."
  )
}

stopifnot(setequal(colnames(txi$counts), metadata$sample_id))
sample_order <- match(metadata$sample_id, colnames(txi$counts))

txi$counts <- txi$counts[, sample_order, drop = FALSE]
rownames(txi$counts) <- gene_ids

for (field in intersect(c("abundance", "length"), names(txi))) {
  if (is.matrix(txi[[field]]) && nrow(txi[[field]]) == length(raw_gene_ids)) {
    stopifnot(identical(rownames(txi[[field]]), raw_gene_ids))
    txi[[field]] <- txi[[field]][, sample_order, drop = FALSE]
    rownames(txi[[field]]) <- gene_ids
  }
}

stopifnot(identical(colnames(txi$counts), metadata$sample_id))
stopifnot(all(is.finite(txi$counts)))
stopifnot(all(txi$counts >= 0))
```

```{r construct-dgelist}
annotation <- NULL
if (!is.null(gene_annotation_path)) {
  annotation <- readr::read_tsv(gene_annotation_path, show_col_types = FALSE)
  stopifnot("gene_id" %in% names(annotation))
  stopifnot(!anyDuplicated(annotation$gene_id))
  annotation <- annotation[match(gene_ids, annotation$gene_id), , drop = FALSE]
  stopifnot(identical(annotation$gene_id, gene_ids))
}

dge <- edgeR::DGEListFromTximport(
  txi,
  genes = annotation,
  remove.zeros = remove_all_zero_genes
)

stopifnot(inherits(dge, "DGEList"))
stopifnot(identical(colnames(dge), metadata$sample_id))
stopifnot(!anyDuplicated(rownames(dge)))

# Keep sample-level metadata with the DGEList without overwriting edgeR's
# library-size and normalization-factor columns.
metadata_extra <- metadata[, setdiff(names(metadata), "sample_id"), drop = FALSE]
metadata_extra <- metadata_extra[, setdiff(names(metadata_extra), names(dge$samples)), drop = FALSE]
dge$samples <- cbind(dge$samples, metadata_extra)

library_size_tbl <- dge$samples |>
  tibble::rownames_to_column("sample_id") |>
  dplyr::select(sample_id, dplyr::everything())

dir.create(dirname(output_dge), recursive = TRUE, showWarnings = FALSE)
dir.create(dirname(output_library_sizes), recursive = TRUE, showWarnings = FALSE)
qs2::qs_save(dge, output_dge)
readr::write_tsv(library_size_tbl, output_library_sizes)
```

## Diagnostics

```{r object-summary}
cat("genes:", nrow(dge), "\n")
cat("samples:", ncol(dge), "\n")
print(dge$samples[, c("lib.size", "norm.factors"), drop = FALSE])
```

## Outputs

- `dge_imported.qs2`: the gene-level edgeR `DGEList` before QC filtering and
  normalization;
- `imported_library_sizes.tsv`: sample names, edgeR library sizes, normalized
  factors, and non-conflicting metadata columns.

## Method notes

`tximport` provides matrices whose rows are features and whose columns are
samples. The exact sample names must match the metadata; reordering is explicit
and the final object is checked again. Gene-version stripping is conservative:
if it creates duplicate gene IDs, the template stops rather than silently
changing the count model.

The source project also contains a raw count-matrix route that collapses
duplicate gene symbols by summing counts and removes all-zero rows. That is a
different input contract and belongs in the QC/normalization stage when it is
needed; this notebook intentionally supports one tximport object path.

`DGEListFromTximport()` creates the edgeR object. It does not perform the
scientific filtering, normalization, design specification, or differential
testing decisions documented in later templates.
