---
title: "Group Summary and Association Tests"
lang: en
format:
  html:
    theme:
      light: cosmo
      dark: darkly
    respect-user-color-scheme: true
    toc: true
    toc-title: "Contents"
    toc-depth: 4
    toc-location: right
    toc-expand: 2
    number-sections: true
    number-depth: 3
    smooth-scroll: true
    anchor-sections: true
    code-fold: true
    code-summary: "Show analysis code"
    code-copy: true
    code-overflow: wrap
    code-line-numbers: false
    fig-width: 16
    fig-height: 12
    fig-dpi: 300
    fig-format: png
    fig-align: center
    fig-responsive: true
    fig-cap-location: bottom
    tbl-cap-location: top
    df-print: paged
    crossrefs-hover: true
    footnotes-hover: true
    citations-hover: true
    embed-resources: true
    self-contained-math: true
    page-layout: article
    grid:
      sidebar-width: 250px
      body-width: 1050px
      margin-width: 300px
      gutter-width: 1.5rem
    link-external-newwindow: true
    link-external-icon: true
lightbox:
  match: auto
  effect: fade
  desc-position: bottom
  loop: false
execute:
  echo: true
  warning: false
  message: false
  error: false
knitr:
  opts_chunk:
    fig.align: center
    out.width: "100%"
    collapse: true
    comment: "#>"
crossref:
  fig-title: "Figure"
  tbl-title: "Table"
  sec-prefix: "Section"
  fig-prefix: "Figure"
  tbl-prefix: "Table"
  ref-hyperlink: true
---

## Purpose

Summarize object metadata and run simple association tests between reusable grouping variables.

::: {.callout-note}
## Archive status

This is a legacy notebook kept for reference. For current reusable workflows,
see the active templates in the repository.
:::

## Packages

Load the packages needed for this notebook.

```{r}
#| label: tpl-general-statistics-01-group-summary-and-association-tests-packages
#| echo: true
#| warning: false
#| message: false
# Load only the packages used in this template.
suppressPackageStartupMessages({
  library(qs2)
  library(readr)
  library(dplyr)
  library(ggplot2)
})
```

## Parameters

Set the parameters that control the analysis and output locations.

```{r}
#| label: tpl-general-statistics-01-group-summary-and-association-tests-user-config
#| echo: true
#| warning: false
#| message: false
# USER INPUT:
input_object <- "path/to/object_with_metadata.qs"
output_object <- "path/to/object_with_statistics.qs"
results_dir <- "results/group_statistics"
figures_dir <- "figures/group_statistics"
assay_name <- "RNA"
sample_column <- "sample_column"
group_column <- "group_column"
numeric_columns <- c("numeric_feature_1")
categorical_columns <- c("categorical_feature_1")
condition_1 <- "condition_1"
condition_2 <- "condition_2"
seed <- 1
n_threads <- 4
compress_level <- 6
```

## Inputs

Load the input object or tables used by the workflow.

```{r}
#| label: tpl-general-statistics-01-group-summary-and-association-tests-input-loading
#| echo: true
#| warning: false
#| message: false
# Create output folders and load configured inputs.
set.seed(seed)
dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figures_dir, recursive = TRUE, showWarnings = FALSE)
object <- qs2::qs_read(file = input_object, nthreads = n_threads)
metadata <- if (inherits(object, "Seurat")) object[[]] else as.data.frame(object)
```

## Validation

Check the required inputs before running the analysis.

```{r}
#| label: tpl-general-statistics-01-group-summary-and-association-tests-validation
#| echo: true
#| warning: false
#| message: false
# Stop early when required inputs or columns are missing.
stopifnot(all(c(sample_column, group_column) %in% colnames(metadata)))
numeric_columns <- intersect(numeric_columns, colnames(metadata))
categorical_columns <- intersect(categorical_columns, colnames(metadata))
```

## Input Summary

Record a compact summary of the inputs for later review.

```{r}
#| label: tbl-general-statistics-group-counts
#| tbl-cap: "Counts for the configured grouping variable."
#| echo: true
#| warning: false
#| message: false
# Write a small TSV summary of the loaded input.
group_counts <- metadata |> count(.data[[group_column]], name = "n")
readr::write_tsv(group_counts, file.path(results_dir, "group_counts.tsv"))
group_counts
```

## Analysis

Run the main analysis with the parameters defined above.

```{r}
#| label: tpl-general-statistics-01-group-summary-and-association-tests-main-analysis
#| echo: true
#| warning: false
#| message: false
# Run the reusable analysis step for this template.
numeric_tests <- lapply(numeric_columns, function(column) {
  x <- metadata |> filter(.data[[group_column]] %in% c(condition_1, condition_2))
  test <- wilcox.test(x[[column]] ~ x[[group_column]])
  tibble::tibble(feature = column, method = "wilcox", p_value = test$p.value)
}) |> bind_rows()
categorical_tests <- lapply(categorical_columns, function(column) {
  tbl <- table(metadata[[group_column]], metadata[[column]])
  test <- suppressWarnings(chisq.test(tbl))
  tibble::tibble(feature = column, method = "chisq", p_value = test$p.value)
}) |> bind_rows()
association_tests <- bind_rows(numeric_tests, categorical_tests) |> mutate(p_adj = p.adjust(p_value, method = "BH"))
```

## Diagnostics

Inspect the checks that help assess the analysis output.

```{r}
#| label: tpl-general-statistics-01-group-summary-and-association-tests-qc-checks
#| echo: true
#| warning: false
#| message: false
# Export checks that make failures easy to inspect.
missing_tbl <- tibble::tibble(column = c(numeric_columns, categorical_columns), missing = vapply(metadata[c(numeric_columns, categorical_columns)], function(x) sum(is.na(x)), integer(1)))
readr::write_tsv(missing_tbl, file.path(results_dir, "metadata_missingness.tsv"))
```

## Visualization

Save the main diagnostic or result figures to the output directory.

```{r}
#| label: tpl-general-statistics-01-group-summary-and-association-tests-visualization
#| echo: true
#| warning: false
#| message: false
# Save compact figures for the main outputs.
if (length(numeric_columns) > 0) {
  p <- ggplot(metadata, aes(x = .data[[group_column]], y = .data[[numeric_columns[1]]])) + geom_boxplot() + theme_bw()
  ggsave(file.path(figures_dir, "numeric_feature_by_group.pdf"), p, width = 6, height = 4)
}
```

## Object metadata

Store the derived metadata alongside the analysis object.

```{r}
#| label: tpl-general-statistics-01-group-summary-and-association-tests-object-result-storage
#| echo: true
#| warning: false
#| message: false
# Store reusable analysis metadata inside the object.
if (inherits(object, "Seurat")) {
  object@misc$group_statistics <- list(association_tests = association_tests, group_column = group_column)
} else {
  attr(object, "group_statistics") <- list(association_tests = association_tests, group_column = group_column)
}
```

## File outputs

Write the tabular outputs produced by this workflow.

```{r}
#| label: tpl-general-statistics-01-group-summary-and-association-tests-result-export
#| echo: true
#| warning: false
#| message: false
# Export the main reusable tables as TSV files.
readr::write_tsv(association_tests, file.path(results_dir, "association_tests.tsv"))
```

## Save

Save the resulting object in the format used by this notebook.

```{r}
#| label: tpl-general-statistics-01-group-summary-and-association-tests-final-save
#| echo: true
#| warning: false
#| message: false
# Save the updated object for the next template step.
qs2::qs_save(object, file = output_object, nthreads = n_threads, compress_level = compress_level)
```

## Output summary

Summarize the files and objects written by the workflow.

```{r}
#| label: tpl-general-statistics-01-group-summary-and-association-tests-output-summary
#| echo: true
#| warning: false
#| message: false
# Print a compact summary of generated outputs.
tibble::tibble(output_object = output_object, tests = nrow(association_tests))
```

## Reproducibility {.appendix}

This final chunk records the R session used when the notebook is rendered.

```{r}
#| label: session-information
#| echo: false
#| results: markup
sessionInfo()
```
