---
title: "TCGA-LUAD Multi-Omics Download"
subtitle: "Reproducible GDC acquisition and matched-matrix preparation for MOFA"
author: "Bianchi Bioinformatics"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    theme: lux
    df-print: paged
execute:
  echo: true
  warning: false
  message: false
  error: false
  freeze: auto
editor: source
params:
  run_downloads: false
  random_seed: 1234
  project_id: "TCGA-LUAD"
  tumor_code: "LUAD"
  sample_type: "TP"
  mirbase_version: "v22"
---

# Purpose

This notebook downloads and prepares TCGA-LUAD primary-tumor matrices for multi-omics modeling.
It was converted from `TCGA_download_compleate.R` and keeps the same biological workflow:

1. identify TCGA-LUAD primary tumor samples from miRNA-seq availability;
2. download and process miRNA-seq, RNA-seq, DNA methylation, RPPA proteomics, and somatic mutations;
3. collapse duplicated TCGA aliquots to 16-character sample barcodes;
4. retain the intersection of samples present in every modality;
5. write raw and matched matrices plus a sample summary.

The heavy GDC chunks are gated by `params$run_downloads`.
Render once with the default `false` value for documentation, then execute with `true` only when the environment, internet access, storage, and package versions are ready.

# Reproducibility Contract

The run is intended to be reproducible at the level of:

- explicit cohort: `r params$project_id`;
- explicit sample type: primary tumor, `r params$sample_type`;
- explicit miRBase target version: `r params$mirbase_version`;
- fixed random seed: `r params$random_seed`;
- all outputs written under project-local `input/tables/`;
- package versions, session metadata, parameters, and output manifest written to disk.

# Setup

```{r}
#| label: setup

set.seed(params$random_seed)

required_packages <- c(
  "TCGAbiolinks",
  "tidyverse",
  "SummarizedExperiment",
  "miRBaseConverter",
  "data.table",
  "vroom",
  "knitr"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]

if (length(missing_packages) > 0) {
  stop(
    "Install missing packages before running this notebook: ",
    paste(missing_packages, collapse = ", ")
  )
}

library(TCGAbiolinks)
library(tidyverse)
library(SummarizedExperiment)
library(miRBaseConverter)
library(data.table)
library(vroom)

if (requireNamespace("conflicted", quietly = TRUE)) {
  conflicted::conflicts_prefer(
    dplyr::filter,
    dplyr::select,
    dplyr::rename,
    dplyr::mutate,
    dplyr::summarise,
    dplyr::distinct,
    base::intersect,
    base::setdiff
  )
}

notebook_input <- knitr::current_input()
notebook_dir <- if (!is.null(notebook_input)) {
  dirname(normalizePath(notebook_input, mustWork = TRUE))
} else {
  getwd()
}

project_dir <- normalizePath(file.path(notebook_dir, ".."), mustWork = TRUE)
input_dir <- file.path(project_dir, "input")
table_dir <- file.path(input_dir, "tables")
metadata_dir <- file.path(input_dir, "metadata")

dir.create(table_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(metadata_dir, recursive = TRUE, showWarnings = FALSE)

run_started_at <- Sys.time()
```

```{r}
#| label: record-run-parameters

run_parameters <- tibble::tibble(
  parameter = names(params),
  value = vapply(params, as.character, character(1))
)

data.table::fwrite(
  run_parameters,
  file.path(metadata_dir, "TCGA_LUAD_download_parameters.csv")
)

knitr::kable(run_parameters)
```

```{r}
#| label: package-versions

package_versions <- tibble::tibble(
  package = required_packages,
  version = vapply(
    required_packages,
    function(pkg) as.character(utils::packageVersion(pkg)),
    character(1)
  )
)

data.table::fwrite(
  package_versions,
  file.path(metadata_dir, "TCGA_LUAD_download_package_versions.csv")
)

knitr::kable(package_versions)
```

# Helper Functions

```{r}
#| label: helper-functions

tcga_sample_barcode <- function(x) {
  substr(x, 1, 16)
}

collapse_duplicate_samples_mean <- function(df, feature_col) {
  df |>
    pivot_longer(
      cols = -all_of(feature_col),
      names_to = "sample",
      values_to = "value"
    ) |>
    mutate(sample = tcga_sample_barcode(sample)) |>
    group_by(.data[[feature_col]], sample) |>
    summarise(value = mean(value, na.rm = TRUE), .groups = "drop") |>
    pivot_wider(
      names_from = sample,
      values_from = value
    )
}

collapse_duplicate_samples_sum <- function(df, feature_col) {
  df |>
    pivot_longer(
      cols = -all_of(feature_col),
      names_to = "sample",
      values_to = "value"
    ) |>
    mutate(sample = tcga_sample_barcode(sample)) |>
    group_by(.data[[feature_col]], sample) |>
    summarise(value = sum(value, na.rm = TRUE), .groups = "drop") |>
    pivot_wider(
      names_from = sample,
      values_from = value,
      values_fill = 0
    )
}

collapse_duplicate_samples_max <- function(df, feature_col) {
  df |>
    pivot_longer(
      cols = -all_of(feature_col),
      names_to = "sample",
      values_to = "value"
    ) |>
    mutate(sample = tcga_sample_barcode(sample)) |>
    group_by(.data[[feature_col]], sample) |>
    summarise(value = max(value, na.rm = TRUE), .groups = "drop") |>
    pivot_wider(
      names_from = sample,
      values_from = value,
      values_fill = 0
    )
}

first_existing_column <- function(candidate_cols, observed_cols, label) {
  value <- intersect(candidate_cols, observed_cols)[1]

  if (is.na(value)) {
    stop(
      "Could not identify ", label, " column. Observed columns: ",
      paste(observed_cols, collapse = ", ")
    )
  }

  value
}

write_output_manifest <- function() {
  output_files <- list.files(
    table_dir,
    pattern = "\\.csv$",
    full.names = TRUE,
    recursive = FALSE
  )

  manifest <- tibble::tibble(
    file = output_files,
    basename = basename(output_files),
    size_bytes = file.info(output_files)$size,
    modified_time = as.character(file.info(output_files)$mtime)
  )

  data.table::fwrite(
    manifest,
    file.path(metadata_dir, "TCGA_LUAD_download_output_manifest.csv")
  )

  manifest
}
```

# 1. Primary Tumor Samples

```{r}
#| label: query-primary-tumor-samples
#| eval: !expr params$run_downloads

query_all <- GDCquery(
  project = params$project_id,
  data.category = "Transcriptome Profiling",
  data.type = "Isoform Expression Quantification",
  experimental.strategy = "miRNA-Seq"
)

samplesDown.luad <- getResults(query_all, cols = "cases")

tumor_samples <- TCGAquery_SampleTypes(
  barcode = samplesDown.luad,
  typesample = params$sample_type
)

tumor_samples <- unique(tcga_sample_barcode(tumor_samples))

stopifnot(length(tumor_samples) > 0)

tumor_sample_table <- tibble::tibble(sample = tumor_samples)

data.table::fwrite(
  tumor_sample_table,
  file.path(metadata_dir, "TCGA_LUAD_primary_tumor_samples.csv")
)

length(tumor_samples)
head(tumor_samples)
```

# 2. miRNA-Seq

```{r}
#| label: download-process-mirna
#| eval: !expr params$run_downloads

query.miRNA <- GDCquery(
  project = params$project_id,
  data.category = "Transcriptome Profiling",
  data.type = "Isoform Expression Quantification",
  experimental.strategy = "miRNA-Seq",
  barcode = tumor_samples
)

GDCdownload(query.miRNA)
miRNA_data <- GDCprepare(query.miRNA)

miRNA_mature <- miRNA_data |>
  filter(grepl("mature", miRNA_region)) |>
  mutate(MIMAT = sub("mature,(MIMAT\\d+)", "\\1", miRNA_region))

miRNA_counts <- miRNA_mature |>
  group_by(MIMAT, barcode) |>
  summarise(read_count = sum(read_count), .groups = "drop") |>
  mutate(read_count = as.numeric(read_count))

miRNA_counts <- miRNA_counts |>
  mutate(
    MatureName = miRNA_AccessionToName(
      MIMAT,
      targetVersion = params$mirbase_version
    )$TargetName
  )

miRNA_df <- miRNA_counts |>
  select(MatureName, barcode, read_count) |>
  group_by(MatureName, barcode) |>
  summarise(read_count = sum(read_count), .groups = "drop") |>
  pivot_wider(
    names_from = barcode,
    values_from = read_count,
    values_fill = 0
  ) |>
  rename(miRNA = MatureName)

miRNA_df <- collapse_duplicate_samples_sum(miRNA_df, "miRNA")

data.table::fwrite(
  miRNA_df,
  file.path(table_dir, "miRNA_TCGA_RAW_COUNTS.csv")
)

dim(miRNA_df)
```

# 3. RNA-Seq

```{r}
#| label: download-process-rnaseq
#| eval: !expr params$run_downloads

query.rna <- GDCquery(
  project = params$project_id,
  data.category = "Transcriptome Profiling",
  data.type = "Gene Expression Quantification",
  workflow.type = "STAR - Counts",
  barcode = tumor_samples
)

GDCdownload(query.rna)
rna_data <- GDCprepare(query.rna)

rna_matrix <- assay(rna_data) |>
  as.data.frame() |>
  rownames_to_column("ensembl_id") |>
  as_tibble()

rna_matrix <- rna_matrix |>
  mutate(ensembl_id = gsub("\\..*", "", ensembl_id))

rna_matrix <- collapse_duplicate_samples_sum(rna_matrix, "ensembl_id")

vroom::vroom_write(
  rna_matrix,
  file.path(table_dir, "RNAseq_TCGA_RAW_COUNTS.csv")
)

dim(rna_matrix)
```

# 4. DNA Methylation

```{r}
#| label: download-process-methylation
#| eval: !expr params$run_downloads

query.meth <- GDCquery(
  project = params$project_id,
  data.category = "DNA Methylation",
  data.type = "Methylation Beta Value",
  platform = "Illumina Human Methylation 450",
  barcode = tumor_samples
)

GDCdownload(query.meth)
meth_data <- GDCprepare(query.meth)

meth_beta <- assay(meth_data) |>
  as.data.frame() |>
  rownames_to_column("probe_id") |>
  as_tibble()

meth_beta <- collapse_duplicate_samples_mean(meth_beta, "probe_id")

meth_mvalue <- meth_beta |>
  mutate(
    across(
      -probe_id,
      ~ {
        beta <- as.numeric(.x)
        beta <- pmin(pmax(beta, 1e-6), 1 - 1e-6)
        log2(beta / (1 - beta))
      }
    )
  )

data.table::fwrite(
  meth_mvalue,
  file.path(table_dir, "Methylation_TCGA_M_VALUES.csv")
)

dim(meth_mvalue)
```

# 5. RPPA Proteomics

```{r}
#| label: download-process-rppa
#| eval: !expr params$run_downloads

query.rppa <- GDCquery(
  project = params$project_id,
  data.category = "Proteome Profiling",
  data.type = "Protein Expression Quantification",
  barcode = tumor_samples
)

GDCdownload(query.rppa)
rppa_data <- GDCprepare(query.rppa)

if (inherits(rppa_data, "SummarizedExperiment")) {
  rppa_matrix <- assay(rppa_data) |>
    as.data.frame() |>
    rownames_to_column("protein") |>
    as_tibble()
} else {
  rppa_matrix <- rppa_data |>
    as_tibble()

  feature_col <- first_existing_column(
    c("peptide_target", "protein_expression", "protein", "gene_name", "target"),
    colnames(rppa_matrix),
    "RPPA feature"
  )

  sample_col <- first_existing_column(
    c("barcode", "sample", "case_submitter_id", "submitter_id"),
    colnames(rppa_matrix),
    "RPPA sample"
  )

  value_col <- first_existing_column(
    c("protein_expression", "value", "expression", "normalized_value"),
    colnames(rppa_matrix),
    "RPPA value"
  )

  rppa_matrix <- rppa_matrix |>
    select(
      protein = all_of(feature_col),
      sample = all_of(sample_col),
      value = all_of(value_col)
    ) |>
    mutate(sample = tcga_sample_barcode(sample)) |>
    group_by(protein, sample) |>
    summarise(value = mean(as.numeric(value), na.rm = TRUE), .groups = "drop") |>
    pivot_wider(
      names_from = sample,
      values_from = value
    )
}

rppa_matrix <- collapse_duplicate_samples_mean(rppa_matrix, "protein")

data.table::fwrite(
  rppa_matrix,
  file.path(table_dir, "RPPA_TCGA_PROTEIN_EXPRESSION.csv")
)

dim(rppa_matrix)
```

# 6. Somatic Mutations

```{r}
#| label: download-process-mutations
#| eval: !expr params$run_downloads

maf <- GDCquery_Maf(
  tumor = params$tumor_code,
  pipelines = "mutect2"
)

mutation_matrix <- maf |>
  as_tibble() |>
  filter(!Variant_Classification %in% c(
    "Silent",
    "RNA",
    "Intron",
    "IGR",
    "3'UTR",
    "5'UTR"
  )) |>
  mutate(sample = tcga_sample_barcode(Tumor_Sample_Barcode)) |>
  filter(sample %in% tumor_samples) |>
  distinct(Hugo_Symbol, sample) |>
  mutate(mutated = 1L) |>
  pivot_wider(
    names_from = sample,
    values_from = mutated,
    values_fill = 0
  ) |>
  rename(gene = Hugo_Symbol)

mutation_matrix <- collapse_duplicate_samples_max(mutation_matrix, "gene")

data.table::fwrite(
  mutation_matrix,
  file.path(table_dir, "Mutation_TCGA_BINARY_MATRIX.csv")
)

dim(mutation_matrix)
```

# 7. Match Samples Across Modalities

```{r}
#| label: match-all-modalities
#| eval: !expr params$run_downloads

sample_lists <- list(
  miRNA = colnames(miRNA_df)[-1],
  RNA = colnames(rna_matrix)[-1],
  methylation = colnames(meth_mvalue)[-1],
  RPPA = colnames(rppa_matrix)[-1],
  mutation = colnames(mutation_matrix)[-1]
)

common_samples_all <- Reduce(intersect, sample_lists)

stopifnot(length(common_samples_all) > 0)

miRNA_matched <- miRNA_df |>
  select(miRNA, all_of(common_samples_all))

rna_matched <- rna_matrix |>
  select(ensembl_id, all_of(common_samples_all))

meth_matched <- meth_mvalue |>
  select(probe_id, all_of(common_samples_all))

rppa_matched <- rppa_matrix |>
  select(protein, all_of(common_samples_all))

mutation_matched <- mutation_matrix |>
  select(gene, all_of(common_samples_all))

length(common_samples_all)
head(common_samples_all)
```

# 8. Save Matched Matrices

```{r}
#| label: save-matched-matrices
#| eval: !expr params$run_downloads

data.table::fwrite(
  miRNA_matched,
  file.path(table_dir, "miRNA_TCGA_TUMOR_MATCHED.csv")
)

data.table::fwrite(
  rna_matched,
  file.path(table_dir, "RNAseq_TCGA_TUMOR_MATCHED.csv")
)

data.table::fwrite(
  meth_matched,
  file.path(table_dir, "Methylation_TCGA_TUMOR_MATCHED.csv")
)

data.table::fwrite(
  rppa_matched,
  file.path(table_dir, "RPPA_TCGA_TUMOR_MATCHED.csv")
)

data.table::fwrite(
  mutation_matched,
  file.path(table_dir, "Mutation_TCGA_TUMOR_MATCHED.csv")
)
```

# 9. Sample Summary

```{r}
#| label: sample-summary
#| eval: !expr params$run_downloads

sample_summary <- tibble(
  modality = names(sample_lists),
  n_samples = purrr::map_int(sample_lists, length)
) |>
  bind_rows(
    tibble(
      modality = "matched_all_modalities",
      n_samples = length(common_samples_all)
    )
  )

data.table::fwrite(
  sample_summary,
  file.path(table_dir, "TCGA_LUAD_modality_sample_summary.csv")
)

knitr::kable(sample_summary)
```

# 10. Output Manifest

```{r}
#| label: output-manifest
#| eval: !expr params$run_downloads

output_manifest <- write_output_manifest()

knitr::kable(output_manifest)
```

# 11. Session Information

```{r}
#| label: session-info

run_finished_at <- Sys.time()

run_metadata <- tibble::tibble(
  field = c(
    "started_at",
    "finished_at",
    "working_directory",
    "r_version",
    "platform"
  ),
  value = c(
    as.character(run_started_at),
    as.character(run_finished_at),
    getwd(),
    R.version.string,
    paste(Sys.info()[["sysname"]], Sys.info()[["release"]], Sys.info()[["machine"]])
  )
)

data.table::fwrite(
  run_metadata,
  file.path(metadata_dir, "TCGA_LUAD_download_run_metadata.csv")
)

session_info_text <- capture.output(sessionInfo())
writeLines(
  session_info_text,
  file.path(metadata_dir, "TCGA_LUAD_download_sessionInfo.txt")
)

knitr::kable(run_metadata)
```
