---
title: "title example"
format:
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    theme: lux
    fig-width: 10
    fig-height: 6
    fig-align: center
    df-print: paged
execute:
  echo: false
  warning: false
  message: false
editor: source
---

# Setup

```{r}
#| label: setup
#| echo: false
#| warning: false
#| message: false
#| include: false

library(data.table)
library(ggplot2)
library(pheatmap)
library(MOFA2)
library(GSVA)
library(singscore)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(future)
library(tidyverse)
library(fst)
library(vroom)
library(SecAct)
library(singscore)


# Keep choices explicit because these packages share common
# function names used later in the notebook.
conflicted::conflicts_prefer(
  dplyr::filter,
  dplyr::select,
  dplyr::first,
  janitor::clean_names,
  base::intersect,
  dplyr::count
)

N_WORKERS <- min(5, future::availableCores())
if (N_WORKERS > 1) {
  plan(multisession, workers = N_WORKERS)
} else {
  plan(sequential)
}
cat("Number of workers:", nbrOfWorkers(), "\n")

RANDOM_SEED <- 1234

dir.create("output/tables", recursive = TRUE, showWarnings = FALSE)
dir.create("output/figures", recursive = TRUE, showWarnings = FALSE)
dir.create("output/objects", recursive = TRUE, showWarnings = FALSE)

compute_cpm <- function(count_mat) {
  dge <- edgeR::DGEList(counts = count_mat)
  keep <- rowSums(edgeR::cpm(dge) > 1) >= 5
  dge <- dge[keep, , keep.lib.sizes = FALSE]
  dge <- edgeR::normLibSizes(dge, method = "TMMwsp")
  edgeR::cpm(dge,  prior.count = 1)
}

compute_logcpm <- function(count_mat) {
  dge <- edgeR::DGEList(counts = count_mat)
  keep <- rowSums(edgeR::cpm(dge) > 1) >= 5
  dge <- dge[keep, , keep.lib.sizes = FALSE]
  dge <- edgeR::normLibSizes(dge, method = "TMMwsp")
  edgeR::cpm(dge, log = TRUE, prior.count = 1)
}

library(tidyverse)
library(gt)
library(gtExtras)
library(scales)

plot_deconv_dunn_table <- function(
  dunn_tbl,
  deconv_col = NULL,
  deconv_label = "Cell type",
  title = "Dunn post-hoc comparisons across C1 consensus groups",
  subtitle = "Only BH-adjusted significant pairwise comparisons are shown",
  output_html = "Dunn_significant_comparisons.html",
  save_html = TRUE,
  table_id = "deconv_dunn_table"
) {
  
  if (is.null(deconv_col)) {
    if ("CIBERSORTx" %in% colnames(dunn_tbl)) {
      deconv_col <- "CIBERSORTx"
    } else if ("consensusTME" %in% colnames(dunn_tbl)) {
      deconv_col <- "consensusTME"
    } else {
      stop("Could not find a deconvolution column. Expected either 'CIBERSORTx' or 'consensusTME'.")
    }
  }
  
  dunn_tbl_clean <- dunn_tbl %>%
    mutate(
      deconv_feature = .data[[deconv_col]],
      comparison = paste(group1, "vs", group2),
      direction = case_when(
        statistic > 0 ~ paste0(group1, " higher"),
        statistic < 0 ~ paste0(group2, " higher"),
        TRUE ~ "No direction"
      ),
      p = as.numeric(p),
      p.adj = as.numeric(p.adj),
      statistic = round(statistic, 2),
      p_display = scales::pvalue(p, accuracy = 0.001, add_p = FALSE),
      p_adj_display = scales::pvalue(p.adj, accuracy = 0.001, add_p = FALSE),
      p.adj.signif = factor(
        p.adj.signif,
        levels = c("*", "**", "***", "****")
      )
    ) %>%
    arrange(.data$deconv_feature, .data$p.adj) %>%
    select(
      deconv_feature,
      comparison,
      direction,
      n1,
      n2,
      statistic,
      p,
      p.adj,
      p_display,
      p_adj_display,
      p.adj.signif
    )
  
  gt_tbl <- dunn_tbl_clean %>%
    gt(groupname_col = "deconv_feature", id = table_id) %>%
    tab_header(
      title = md(paste0("**", title, "**")),
      subtitle = subtitle
    ) %>%
    cols_label(
      comparison = "Comparison",
      direction = "Direction",
      n1 = "n1",
      n2 = "n2",
      statistic = "Dunn Z",
      p_display = "p",
      p_adj_display = "BH-adjusted p",
      p.adj.signif = "Significance"
    ) %>%
    tab_spanner(
      label = "Groups",
      columns = c(comparison, direction, n1, n2)
    ) %>%
    tab_spanner(
      label = "Statistics",
      columns = c(statistic, p_display, p_adj_display, p.adj.signif)
    ) %>%
    data_color(
      columns = statistic,
      palette = "RdBu",
      reverse = TRUE
    ) %>%
    data_color(
      columns = p.adj,
      colors = scales::col_numeric(
        palette = c("#2166AC", "#F7F7F7"),
        domain = c(0, 0.05)
      )
    ) %>%
    cols_hide(columns = c(p, p.adj)) %>%
    gtExtras::gt_theme_538(quiet = TRUE) %>%
    tab_style(
      style = cell_text(weight = "bold"),
      locations = cells_column_labels(everything())
    ) %>%
    tab_style(
      style = cell_text(weight = "bold"),
      locations = cells_row_groups()
    ) %>%
    tab_style(
      style = cell_fill(color = "#F3F4F6"),
      locations = cells_row_groups()
    ) %>%
    tab_style(
      style = cell_text(weight = "bold"),
      locations = cells_body(columns = p.adj.signif)
    ) %>%
    tab_options(
      table.font.names = c("Arial", "Helvetica", "sans-serif"),
      table.font.size = px(13),
      heading.title.font.size = px(18),
      heading.subtitle.font.size = px(13),
      row_group.font.size = px(13),
      row_group.font.weight = "bold",
      column_labels.font.weight = "bold",
      table.border.top.width = px(2),
      table.border.bottom.width = px(2),
      data_row.padding = px(5)
    ) %>%
    cols_align(
      align = "center",
      columns = c(n1, n2, statistic, p_display, p_adj_display, p.adj.signif)
    ) %>%
    cols_align(
      align = "left",
      columns = c(comparison, direction)
    )
  
  if (save_html) {
    gt::gtsave(
      data = gt_tbl,
      filename = output_html
    )
  }
  
  gt_tbl
}


```


# RNA Preprocessing

## RNA Identifier Harmonization

### Map Ensembl IDs to HUGO Symbols

```{r}

library(biomaRt)
library(dplyr)
library(tibble)

# -----------------------------
# 1. Clean Ensembl IDs
# -----------------------------
RNA <- vroom::vroom("input/tables/RNAseq_TCGA_TUMOR_MATCHED.csv")

sample_cols <- grep("^TCGA-", colnames(RNA), value = TRUE)

RNA_clean <- RNA %>%
  mutate(
    ensembl_id_original = ensembl_id,
    ensembl_id = sub("\\..*$", "", ensembl_id)
  )

ens_ids <- unique(RNA_clean$ensembl_id)

duplicated_ensembl_before <- RNA_clean %>%
  dplyr::count(ensembl_id, name = "n_rows") %>%
  filter(n_rows > 1) %>%
  arrange(desc(n_rows))

print(duplicated_ensembl_before)

RNA_ensembl_collapsed <- RNA_clean %>%
  select(ensembl_id, all_of(sample_cols)) %>%
  group_by(ensembl_id) %>%
  summarise(
    across(all_of(sample_cols), ~ sum(.x, na.rm = TRUE)),
    .groups = "drop"
  )

stopifnot(sum(duplicated(RNA_ensembl_collapsed$ensembl_id)) == 0)

ens_ids <- unique(RNA_ensembl_collapsed$ensembl_id)


# -----------------------------
# 2. Query BioMart
# -----------------------------

mart <- useEnsembl(
  biomart = "genes",
  dataset = "hsapiens_gene_ensembl",
  mirror = "useast"
)

gene_annot <- getBM(
  attributes = c(
    "ensembl_gene_id",
    "hgnc_symbol",
    "entrezgene_id",
    "gene_biotype"
  ),
  filters = "ensembl_gene_id",
  values = ens_ids,
  mart = mart
)

gene_annot <- gene_annot %>%
  rename(
    ensembl_id = ensembl_gene_id,
    gene_symbol = hgnc_symbol,
    entrez_id = entrezgene_id
  ) %>%
  distinct(ensembl_id, gene_symbol, entrez_id, gene_biotype)

duplicated_biomart_annotations <- gene_annot %>%
  dplyr::count(ensembl_id, name = "n_annot_rows") %>%
  filter(n_annot_rows > 1) %>%
  arrange(desc(n_annot_rows))

print(duplicated_biomart_annotations)

gene_annot_one <- gene_annot %>%
  mutate(
    has_symbol = !is.na(gene_symbol) & gene_symbol != "",
    has_entrez = !is.na(entrez_id),
    is_protein_coding = gene_biotype == "protein_coding"
  ) %>%
  arrange(
    ensembl_id,
    desc(has_symbol),
    desc(is_protein_coding),
    desc(has_entrez),
    gene_symbol,
    entrez_id
  ) %>%
  group_by(ensembl_id) %>%
  slice(1) %>%
  ungroup() %>%
  select(-has_symbol, -has_entrez, -is_protein_coding)

stopifnot(sum(duplicated(gene_annot_one$ensembl_id)) == 0)

RNA_annotated <- RNA_ensembl_collapsed %>%
  left_join(gene_annot_one, by = "ensembl_id")

stopifnot(nrow(RNA_annotated) == nrow(RNA_ensembl_collapsed))

RNA_annotated <- RNA_clean %>%
  left_join(gene_annot, by = "ensembl_id")

mapping_summary <- RNA_annotated %>%
  summarise(
    n_input_ensembl = n(),
    n_with_biomart_match = sum(!is.na(gene_biotype)),
    n_without_biomart_match = sum(is.na(gene_biotype)),
    n_with_gene_symbol = sum(!is.na(gene_symbol) & gene_symbol != ""),
    n_without_gene_symbol = sum(is.na(gene_symbol) | gene_symbol == "")
  )

mapping_summary

RNA_mapped <- RNA_annotated %>%
  filter(!is.na(gene_symbol), gene_symbol != "")

duplicated_gene_symbols <- RNA_mapped %>%
  dplyr::count(gene_symbol, name = "n_ensembl_ids") %>%
  filter(n_ensembl_ids > 1) %>%
  arrange(desc(n_ensembl_ids))

print(duplicated_gene_symbols)

duplicated_gene_symbol_mapping <- RNA_mapped %>%
  semi_join(duplicated_gene_symbols, by = "gene_symbol") %>%
  select(ensembl_id, gene_symbol, entrez_id, gene_biotype) %>%
  arrange(gene_symbol, ensembl_id)

duplicated_gene_symbol_mapping


RNA_gene_symbol <- RNA_mapped %>%
  select(gene_symbol, all_of(sample_cols)) %>%
  group_by(gene_symbol) %>%
  summarise(
    across(all_of(sample_cols), ~ sum(.x, na.rm = TRUE)),
    .groups = "drop"
  )

final_qc <- tibble(
  original_rows = nrow(RNA),
  ensembl_collapsed_rows = nrow(RNA_ensembl_collapsed),
  annotated_rows = nrow(RNA_annotated),
  mapped_rows_with_symbol = nrow(RNA_mapped),
  final_gene_symbol_rows = nrow(RNA_gene_symbol),
  n_samples = length(sample_cols),
  duplicated_ensembl_after_collapse = sum(duplicated(RNA_ensembl_collapsed$ensembl_id)),
  duplicated_gene_symbols_final = sum(duplicated(RNA_gene_symbol$gene_symbol)),
  missing_gene_symbols_final = sum(is.na(RNA_gene_symbol$gene_symbol)),
  empty_gene_symbols_final = sum(RNA_gene_symbol$gene_symbol == "")
)

print(final_qc)

stopifnot(final_qc$duplicated_ensembl_after_collapse == 0)
stopifnot(final_qc$duplicated_gene_symbols_final == 0)
stopifnot(final_qc$missing_gene_symbols_final == 0)
stopifnot(final_qc$empty_gene_symbols_final == 0)

RNA_gene_symbol |>
  fwrite(file = "input/tables/HUGO_RNAseq_TCGA_TUMOR_MATCHED.csv")

```

## RNA Expression Matrices

### Build CPM and logCPM Matrices

```{r}

set.seed(RANDOM_SEED)

rna_mat <- vroom("input/CLEAN_RNA_raw_counts_TCGA.csv") |>
  column_to_rownames("gene")

rna_cpm <- compute_cpm(rna_mat)
rna_logcpm <- compute_logcpm(rna_mat)

```

# C1-Core Scoring

## Transcriptome-Based C1-Core Scores

### Scoring Setup

```{r}

# ============================================================
# C1 CORE SCORING: singscore + GSVA + consensus
# ============================================================

library(dplyr)
library(tidyr)
library(tibble)
library(vroom)
library(ggplot2)
library(ggpubr)
library(forcats)
library(singscore)
library(GSVA)

set.seed(RANDOM_SEED)

dir.create("output/tables", recursive = TRUE, showWarnings = FALSE)
dir.create("output/figures", recursive = TRUE, showWarnings = FALSE)

theme_set(theme_pubclean(base_size = 12))

pastel_score_cols <- c(
  "singscore" = "#9ecae1",
  "GSVA" = "#fdae6b",
  "consensus" = "#a1d99b"
)

pastel_group_cols <- c(
  "Not-C1-like" = "#bcbddc",
  "Not-C1-intermediate" = "#c7e9c0",
  "C1-intermediate" = "#fdd0a2",
  "C1-like" = "#fcbba1"
)

class_from_z <- function(x) {
  case_when(
    x > 1 ~ "C1-like",
    x > 0 & x <= 1 ~ "C1-intermediate",
    x >= -1 & x <= 0 ~ "Not-C1-intermediate",
    x < -1 ~ "Not-C1-like",
    TRUE ~ NA_character_
  )
}

scale_numeric <- function(x) {
  as.numeric(scale(x))
}

c1_group_levels <- c(
  "Not-C1-like",
  "Not-C1-intermediate",
  "C1-intermediate",
  "C1-like"
)

```

### Prepare C1-Core Signature Inputs

```{r}

# ============================================================
# 1. Prepare expression matrix and C1-core genes
# ============================================================

c1_core <- vroom(
  "input/tables/C1_Core_signature.csv",
  show_col_types = FALSE
)

c1_core_genes <- c1_core |>
  filter(C1_Core_direction %in% c("UP", "DOWN")) |>
  mutate(
    C1_Core_direction = toupper(C1_Core_direction)
  ) |>
  distinct(gene, C1_Core_direction)

c1_core_up <- c1_core_genes |>
  filter(C1_Core_direction == "UP") |>
  pull(gene) |>
  unique()

c1_core_down <- c1_core_genes |>
  filter(C1_Core_direction == "DOWN") |>
  pull(gene) |>
  unique()

c1_core_up_found <- intersect(c1_core_up, rownames(rna_logcpm))
c1_core_down_found <- intersect(c1_core_down, rownames(rna_logcpm))

c1_gene_coverage <- tibble(
  set = c("UP", "DOWN"),
  requested = c(length(c1_core_up), length(c1_core_down)),
  found = c(length(c1_core_up_found), length(c1_core_down_found)),
  missing = c(
    paste(base::setdiff(c1_core_up, rownames(rna_logcpm)), collapse = ";"),
    paste(base::setdiff(c1_core_down, rownames(rna_logcpm)), collapse = ";")
  )
)

c1_gene_coverage

write.csv(
c1_gene_coverage,
  "output/tables/C1_core_gene_coverage.csv",
  row.names = FALSE
)

```

### singscore-Based C1-Core Scoring

```{r}

# ============================================================
# 2. singscore C1-core scoring
# ============================================================

rank_data <- rankGenes(rna_logcpm)

c1_singscore_raw <- simpleScore(
  rankData = rank_data,
  upSet = c1_core_up_found,
  downSet = c1_core_down_found
)

c1_singscore <- c1_singscore_raw |>
  as.data.frame() |>
  rownames_to_column("sample") |>
  transmute(
    sample,
    singscore_TotalScore = TotalScore,
    singscore_UpScore = UpScore,
    singscore_DownScore = DownScore,
    singscore_TotalScore_check = UpScore + DownScore,
    C1_singscore = scale_numeric(TotalScore)
  )

c1_singscore_check <- c1_singscore |>
  summarise(
    max_abs_diff_TotalScore = max(
      abs(singscore_TotalScore - singscore_TotalScore_check),
      na.rm = TRUE
    )
  )

c1_singscore_check

```

### GSVA-Based C1-Core Scoring

```{r}

# ============================================================
# 3. GSVA C1-core scoring
# ============================================================

c1_gene_sets <- list(
  C1_core_UP = c1_core_up_found,
  C1_core_DOWN = c1_core_down_found
)

gsva_par <- gsvaParam(
  exprData = rna_logcpm,
  geneSets = c1_gene_sets,
  kcdf = "Gaussian"
)

gsva_score_mat <- gsva(gsva_par)

c1_gsva <- gsva_score_mat |>
  as.data.frame() |>
  rownames_to_column("gene_set") |>
  pivot_longer(
    cols = -gene_set,
    names_to = "sample",
    values_to = "gsva_score"
  ) |>
  pivot_wider(
    names_from = gene_set,
    values_from = gsva_score
  ) |>
  mutate(
    GSVA_UP_z = scale_numeric(C1_core_UP),
    GSVA_DOWN_z = scale_numeric(C1_core_DOWN),
    C1_GSVA_raw = GSVA_UP_z - GSVA_DOWN_z,
    C1_GSVA = scale_numeric(C1_GSVA_raw)
  ) |>
  select(
    sample,
    GSVA_UP = C1_core_UP,
    GSVA_DOWN = C1_core_DOWN,
    GSVA_UP_z,
    GSVA_DOWN_z,
    C1_GSVA_raw,
    C1_GSVA
  )

```

### Consensus C1-Core Score

```{r}

# ============================================================
# 4. Consensus score
# ============================================================

c1_score_tbl <- c1_singscore |>
  inner_join(c1_gsva, by = "sample") |>
  mutate(
    C1_consensus = scale_numeric(rowMeans(
      cbind(C1_singscore, C1_GSVA),
      na.rm = TRUE
    )),
    C1_singscore_group = factor(class_from_z(C1_singscore), levels = c1_group_levels),
    C1_GSVA_group = factor(class_from_z(C1_GSVA), levels = c1_group_levels),
    C1_consensus_group = factor(class_from_z(C1_consensus), levels = c1_group_levels)
  ) |>
  as_tibble()

write.csv(
  c1_score_tbl,
  "output/tables/C1_core_singscore_GSVA_consensus_scores.csv",
  row.names = FALSE
)

c1_score_summary <- c1_score_tbl |>
  summarise(
    n_samples = n(),
    spearman_singscore_gsva = cor(C1_singscore, C1_GSVA, method = "spearman"),
    spearman_singscore_consensus = cor(C1_singscore, C1_consensus, method = "spearman"),
    spearman_gsva_consensus = cor(C1_GSVA, C1_consensus, method = "spearman")
  )

c1_score_summary

write.csv(
  c1_score_summary,
  "output/tables/C1_core_score_method_correlations.csv",
  row.names = FALSE
)

library(GGally)
p <- ggpairs(data = c1_score_tbl, columns = c("C1_singscore","C1_GSVA","C1_consensus"),
  lower = list(continuous = wrap("smooth_loess", se = T, color = "lightblue", alpha = 0.5)),
  upper = list(continuous = wrap("cor", size = 10,method = "spearman")),
  diag = list(continuous = wrap("densityDiag"))
) + 
  theme_bw(base_size = 15)

p

c1_score_tbl |>
  count(C1_singscore_group)

c1_score_tbl |>
  count(C1_GSVA_group)

c1_score_tbl |>
  count(C1_consensus_group)

```

# Additional Functional Readouts

## Metabolic Flux

### METAFlux

```{r}
set.seed(RANDOM_SEED)

library(METAFlux)

expr_metaflux <- log2(rna_cpm + 1)

# medium file for human derived samples
data("human_blood")
data("human_gem")
#Calculate mras for human sample data
scores<-calculate_reaction_score(expr_metaflux)
#Calculate flux for human sample data
flux<-compute_flux(mras=scores,medium=human_blood) 

#optional: flux scores cubic root normalization
cbrt <- function(x) {
    sign(x) * abs(x)^(1/3)
}

flux=cbrt(flux)

flux |>
  as.data.frame() |>
  rownames_to_column("flux") |>
  write_fst("output/tables/LUAD_METAFlux_Flux.fst")

flux <- read_fst("output/tables/LUAD_METAFlux_Flux.fst") |>
  column_to_rownames("flux")

data("nutrient_lookup_files")

library(dplyr)

subsystem_map <- human_gem |>
  select(ID, SUBSYSTEM) |>
  filter(
    ID %in% rownames(flux),
    !is.na(SUBSYSTEM),
    SUBSYSTEM != ""
  )

flux_sub <- flux[subsystem_map$ID, , drop = FALSE]

pathway_sum <- rowsum(
  x = abs(flux_sub),
  group = subsystem_map$SUBSYSTEM,
  reorder = FALSE
)

n_rxn <- table(subsystem_map$SUBSYSTEM)

pathway_score <- sweep(
  pathway_sum,
  1,
  n_rxn[rownames(pathway_sum)],
  "/"
)

pathway_score |>
  as.data.frame() |>
  rownames_to_column("flux") |>
  fwrite("output/tables/LUAD_METAFlux_pathway_score.csv")


```

## Transcription Factor Activity

### decoupleR

```{r}


set.seed(RANDOM_SEED)
net <- decoupleR::get_collectri(organism = 'human', 
                                split_complexes = FALSE)

net

sample_acts <- decoupleR::run_mlm(mat = rna_mat, 
                                  net = net, 
                                  .source = 'source', 
                                  .target = 'target',
                                  .mor = 'mor', 
                                  minsize = 5)

sample_acts

sample_acts_mat <- sample_acts %>%
  tidyr::pivot_wider(
    id_cols = source,
    names_from = condition,
    values_from = score
  )

fwrite(sample_acts_mat, "input/tables/TF_activity_TCGA_TUMOR_MATCHED.csv")

```

## Cytokine Activity

### SecAct

```{r}
library(SecAct)
set.seed(RANDOM_SEED)
# SecAct infers cytokine activity from the matched RNA-seq matrix.
SecAct_scores <- SecAct.activity.inference(rna_mat, sigMatrix = "SecAct-Lung-Adeno")$zscore |>
  t() |>
  as.data.frame() |>
  tibble::rownames_to_column("sample") |>
  as_tibble()

SecAct_scores

fwrite(SecAct_scores, "input/tables/TCGA_SecAct_scores.csv")

```


# RNA Deconvolution

## ConsensusTME

### Run ConsensusTME Across Methods

```{r}
library(ConsensusTME)

CTME_deconv_ssgsea <- consensusTMEAnalysis(rna_logcpm, cancer = "LUAD", statMethod = "ssgsea") |>
  as.data.frame()
head(CTME_deconv_ssgsea)

CTME_deconv_gsva <- consensusTMEAnalysis(rna_logcpm, cancer = "LUAD", statMethod = "gsva") |>
  as.data.frame()
head(CTME_deconv_gsva)

CTME_deconv_singScore <- consensusTMEAnalysis(rna_logcpm, cancer = "LUAD", statMethod = "singScore") |>
  as.data.frame()
head(CTME_deconv_singScore)

```


### Compare ConsensusTME Methods

```{r}

library(tidyverse)
library(patchwork)

# Extract matrices
CTME_gsva <- as.matrix(CTME_deconv_gsva)
CTME_ssgsea <- as.matrix(CTME_deconv_ssgsea)

CTME_sing <- if (is.list(CTME_deconv_singScore) && "Scores" %in% names(CTME_deconv_singScore)) {
  as.matrix(CTME_deconv_singScore$Scores)
} else {
  as.matrix(CTME_deconv_singScore)
}

# Fix sample names if singScore was converted with as.data.frame()
colnames(CTME_sing) <- gsub("^Scores\\.", "", colnames(CTME_sing))
colnames(CTME_sing) <- gsub("\\.", "-", colnames(CTME_sing))

# Match features and samples
common_celltypes <- Reduce(
  intersect,
  list(
    rownames(CTME_gsva),
    rownames(CTME_ssgsea),
    rownames(CTME_sing)
  )
)

common_samples <- Reduce(
  intersect,
  list(
    colnames(CTME_gsva),
    colnames(CTME_ssgsea),
    colnames(CTME_sing)
  )
)

CTME_gsva <- CTME_gsva[common_celltypes, common_samples, drop = FALSE]
CTME_ssgsea <- CTME_ssgsea[common_celltypes, common_samples, drop = FALSE]
CTME_sing <- CTME_sing[common_celltypes, common_samples, drop = FALSE]

same_celltype_cor <- tibble(
  cell_type = common_celltypes,
  gsva_vs_singScore = diag(cor(
    t(CTME_gsva),
    t(CTME_sing),
    method = "spearman",
    use = "pairwise.complete.obs"
  )),
  ssgsea_vs_singScore = diag(cor(
    t(CTME_ssgsea),
    t(CTME_sing),
    method = "spearman",
    use = "pairwise.complete.obs"
  )),
  ssgsea_vs_gsva = diag(cor(
    t(CTME_ssgsea),
    t(CTME_gsva),
    method = "spearman",
    use = "pairwise.complete.obs"
  ))
) |>
  pivot_longer(
    cols = -cell_type,
    names_to = "comparison",
    values_to = "spearman_rho"
  ) |>
  mutate(
    comparison = recode(
      comparison,
      gsva_vs_singScore = "GSVA vs singScore",
      ssgsea_vs_singScore = "ssGSEA vs singScore",
      ssgsea_vs_gsva = "ssGSEA vs GSVA"
    )
  )

same_celltype_cor

same_celltype_cor |>
  group_by(comparison) |>
  summarise(
    mean_rho = mean(spearman_rho, na.rm = TRUE),
    median_rho = median(spearman_rho, na.rm = TRUE),
    min_rho = min(spearman_rho, na.rm = TRUE),
    max_rho = max(spearman_rho, na.rm = TRUE),
    .groups = "drop"
  )

p_same <- ggplot(
  same_celltype_cor,
  aes(x = reorder(cell_type, spearman_rho), y = spearman_rho)
) +
  geom_col(width = 0.75) +
  coord_flip() +
  facet_wrap(~ comparison, nrow = 1) +
  ylim(0, 1) +
  labs(
    title = "Same-cell-type agreement across ConsensusTME scoring methods",
    x = NULL,
    y = "Spearman rho across samples"
  ) +
  theme_bw(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

p_same

```

### Export the ConsensusTME singScore Matrix

```{r}

CTME_deconv_singScore <- consensusTMEAnalysis(rna_logcpm, cancer = "LUAD", statMethod = "singScore") |>
  as.data.frame()

colnames(CTME_deconv_singScore) <- colnames(rna_logcpm)

CTME_deconv_singScore <- CTME_deconv_singScore |>
  rownames_to_column("consensusTME") |>
  as_tibble()

head(CTME_deconv_singScore)

vroom::vroom_write(x = CTME_deconv_singScore, file = "input/tables/consensusTME_TCGA_TUMOR_MATCHED.csv")


```

### ConsensusTME by C1-Core Group

```{r}

library(tidyverse)
library(ggsci)
library(ggpubr)
library(rstatix)

# Convert deconvolution matrix from wide to long format
ctme_long <- CTME_deconv_singScore %>%
  pivot_longer(
    cols = -consensusTME,
    names_to = "sample",
    values_to = "score"
  ) %>%
  left_join(
    c1_score_tbl %>%
      select(sample, C1_consensus, C1_consensus_group),
    by = "sample"
  ) %>%
  filter(!is.na(C1_consensus_group))

# Check group order
levels(ctme_long$C1_consensus_group)

ctme_long <- ctme_long %>%
  mutate(
    C1_consensus_group = factor(
      C1_consensus_group,
      levels = c(
        "Not-C1-like",
        "Not-C1-intermediate",
        "C1-intermediate",
        "C1-like"
      )
    )
  ) |>
    filter(consensusTME != "Immune_Score")

p_ctme_violin <- ggplot(
  ctme_long,
  aes(
    x = C1_consensus_group,
    y = score,
    fill = C1_consensus_group
  )
) +
  geom_violin(
    trim = TRUE,
    alpha = 0.75,
    color = "grey30",
    linewidth = 0.25
  ) +
  geom_boxplot(
    width = 0.16,
    outlier.shape = NA,
    alpha = 0.85,
    color = "black",
    linewidth = 0.3
  ) +
  facet_wrap(
    ~ consensusTME,
    scales = "free_y",
    ncol = 4
  ) +
  scale_fill_npg() +
  labs(
    title = "ConsensusTME singScore deconvolution by C1 consensus group",
    x = NULL,
    y = "ConsensusTME singScore",
    fill = "C1 consensus group"
  ) +
  stat_compare_means() +
  theme_pubclean(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

p_ctme_violin

ggsave(filename = "output/figures/TCGA_LUAD_consensusTME.png",plot = p_ctme_violin,width = 24,height = 20,dpi = 600)

```

### ConsensusTME Dunn Post-hoc Testing

```{r}


CTME_dunn <- ctme_long %>%
  group_by(consensusTME) %>%
  dunn_test(
    score ~ C1_consensus_group,
    p.adjust.method = "BH"
  ) %>%
  ungroup() %>%
  add_significance("p.adj") %>%
  filter(p.adj.signif != "ns") |>
  arrange(consensusTME, p.adj)

CTME_dunn

vroom::vroom_write(
  CTME_dunn,
  "output/tables/TCGA_LUAD_consensusTME_dunn_by_C1_group.csv"
)

consensusTME_dunn_gt <- plot_deconv_dunn_table(
  dunn_tbl = CTME_dunn,
  deconv_col = "consensusTME",
  title = "consensusTME immune cell differences across C1 consensus groups",
  subtitle = "Dunn post-hoc test after Kruskal-Wallis; BH-adjusted significant comparisons only",
  output_html = "consensusTME_Dunn_significant_comparisons.html",
  save_html = TRUE,
  table_id = "consensusTME_dunn_table"
)

consensusTME_dunn_gt

```

### ConsensusTME Extreme-Group Comparison

```{r}

ctme_extreme <- ctme_long %>%
  filter(C1_consensus_group %in% c("Not-C1-like", "C1-like")) %>%
  mutate(
    C1_consensus_group = factor(
      C1_consensus_group,
      levels = c("Not-C1-like", "C1-like")
    )
  )

p_ctme_extreme <- ggplot(
  ctme_extreme,
  aes(
    x = C1_consensus_group,
    y = score,
    fill = C1_consensus_group
  )
) +
  geom_violin(
    trim = FALSE,
    alpha = 0.75,
    color = "grey30",
    linewidth = 0.25
  ) +
  geom_boxplot(
    width = 0.18,
    outlier.shape = NA,
    alpha = 0.85,
    color = "black",
    linewidth = 0.3
  ) +
  stat_compare_means(
    size = 4
  ) +
  facet_wrap(
    ~ consensusTME,
    scales = "free_y",
    ncol = 4
  ) +
  scale_fill_npg() +
  labs(
    title = "ConsensusTME singScore deconvolution: Not-C1-like vs C1-like",
    x = NULL,
    y = "ConsensusTME singScore",
    fill = "C1 consensus group"
  ) +
  theme_pubclean(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

p_ctme_extreme

ggsave(
  filename = "output/figures/TCGA_LUAD_consensusTME_Not_C1_like_vs_C1_like.png",
  plot = p_ctme_extreme,
  width = 24,
  height = 20,
  dpi = 600
)

```


## CIBERSORTx

### Load CIBERSORTx Results

```{r}

CIBERSORTx <- vroom("input/CIBERSORTx_Job6_Results.csv") |>
  as.data.frame() |>
  column_to_rownames("Mixture") |>
  select(-`P-value`, -`Absolute score (sig.score)`, -RMSE, -Correlation)

```

### CIBERSORTx by C1-Core Group

```{r}

library(tidyverse)
library(ggsci)
library(ggpubr)
library(rstatix)

# Convert deconvolution matrix from wide to long format
ctme_long <- CIBERSORTx %>%
  t() |>
  as.data.frame() |>
  rownames_to_column("consensusTME") |>
  pivot_longer(
    cols = -consensusTME,
    names_to = "sample",
    values_to = "score"
  ) %>%
  left_join(
    c1_score_tbl %>%
      select(sample, C1_consensus, C1_consensus_group),
    by = "sample"
  ) %>%
  filter(!is.na(C1_consensus_group))

# Check group order
levels(ctme_long$C1_consensus_group)

ctme_long <- ctme_long %>%
  mutate(
    C1_consensus_group = factor(
      C1_consensus_group,
      levels = c(
        "Not-C1-like",
        "Not-C1-intermediate",
        "C1-intermediate",
        "C1-like"
      )
    )
  ) |>
    filter(consensusTME != "Immune_Score")

p_ctme_violin <- ggplot(
  ctme_long,
  aes(
    x = C1_consensus_group,
    y = score,
    fill = C1_consensus_group
  )
) +
  geom_violin(
    trim = TRUE,
    alpha = 0.75,
    color = "grey30",
    linewidth = 0.25
  ) +
  geom_boxplot(
    width = 0.16,
    outlier.shape = NA,
    alpha = 0.85,
    color = "black",
    linewidth = 0.3
  ) +
  facet_wrap(
    ~ consensusTME,
    scales = "free_y",
    ncol = 4
  ) +
  scale_fill_npg() +
  labs(
    title = "CIBERSORTx singScore deconvolution by C1 consensus group",
    x = NULL,
    y = "CIBERSORTx singScore",
    fill = "C1 consensus group"
  ) +
  stat_compare_means() +
  theme_pubclean(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

p_ctme_violin

ggsave(filename = "output/figures/TCGA_LUAD_CIBERSORTx.png",plot = p_ctme_violin,width = 24,height = 20,dpi = 600)

```

### CIBERSORTx Dunn Post-hoc Testing

```{r}

library(gt)
library(gtExtras)

CIBERSORTx_dunn <- ctme_long %>%
  dplyr::rename("CIBERSORTx" = "consensusTME")|>
  group_by(CIBERSORTx) %>%
  dunn_test(
    score ~ C1_consensus_group,
    p.adjust.method = "BH"
  ) %>%
  ungroup() %>%
  add_significance("p.adj") %>%
  filter(p.adj.signif != "ns") |>
  arrange(CIBERSORTx, p.adj)

CIBERSORTx_dunn

vroom::vroom_write(
  CIBERSORTx_dunn,
  "output/tables/TCGA_LUAD_CIBERSORTx_dunn_by_C1_group.csv"
)

library(tidyverse)
library(gt)
library(gtExtras)
library(scales)

plot_cibersortx_dunn_table <- function(
  ctme_dunn,
  title = "CIBERSORTx Dunn post-hoc comparisons across C1 consensus groups",
  subtitle = "Only BH-adjusted significant pairwise comparisons are shown",
  output_html = "CIBERSORTx_Dunn_significant_comparisons.html",
  save_html = TRUE
) {
  
  dunn_tbl <- ctme_dunn %>%
    mutate(
      comparison = paste(group1, "vs", group2),
      direction = case_when(
        statistic > 0 ~ paste0(group1, " higher"),
        statistic < 0 ~ paste0(group2, " higher"),
        TRUE ~ "No direction"
      ),
      p = as.numeric(p),
      p.adj = as.numeric(p.adj),
      statistic = round(statistic, 2),
      p_display = scales::pvalue(p, accuracy = 0.001, add_p = FALSE),
      p_adj_display = scales::pvalue(p.adj, accuracy = 0.001, add_p = FALSE),
      p.adj.signif = factor(
        p.adj.signif,
        levels = c("*", "**", "***", "****")
      )
    ) %>%
    arrange(CIBERSORTx, p.adj) %>%
    select(
      CIBERSORTx,
      comparison,
      direction,
      n1,
      n2,
      statistic,
      p,
      p.adj,
      p_display,
      p_adj_display,
      p.adj.signif
    )
  
  gt_tbl <- dunn_tbl %>%
    gt(groupname_col = "CIBERSORTx") %>%
    tab_header(
      title = md(paste0("**", title, "**")),
      subtitle = subtitle
    ) %>%
    cols_label(
      comparison = "Comparison",
      direction = "Direction",
      n1 = "n1",
      n2 = "n2",
      statistic = "Dunn Z",
      p_display = "p",
      p_adj_display = "BH-adjusted p",
      p.adj.signif = "Significance"
    ) %>%
    tab_spanner(
      label = "Groups",
      columns = c(comparison, direction, n1, n2)
    ) %>%
    tab_spanner(
      label = "Statistics",
      columns = c(statistic, p_display, p_adj_display, p.adj.signif)
    ) %>%
    data_color(
      columns = statistic,
      palette = "RdBu",
      reverse = TRUE
    ) %>%
    data_color(
      columns = p.adj,
      colors = scales::col_numeric(
        palette = c("#2166AC", "#F7F7F7"),
        domain = c(0, 0.05)
      )
    ) %>%
    cols_hide(columns = c(p, p.adj)) %>%
    gtExtras::gt_theme_538() %>%
    tab_style(
      style = cell_text(weight = "bold"),
      locations = cells_column_labels(everything())
    ) %>%
    tab_style(
      style = cell_text(weight = "bold"),
      locations = cells_row_groups()
    ) %>%
    tab_style(
      style = cell_fill(color = "#F3F4F6"),
      locations = cells_row_groups()
    ) %>%
    tab_style(
      style = cell_text(weight = "bold"),
      locations = cells_body(columns = p.adj.signif)
    ) %>%
    tab_options(
      table.font.names = c("Arial", "Helvetica", "sans-serif"),
      table.font.size = px(13),
      heading.title.font.size = px(18),
      heading.subtitle.font.size = px(13),
      row_group.font.size = px(13),
      row_group.font.weight = "bold",
      column_labels.font.weight = "bold",
      table.border.top.width = px(2),
      table.border.bottom.width = px(2),
      data_row.padding = px(5)
    ) %>%
    cols_align(
      align = "center",
      columns = c(n1, n2, statistic, p_display, p_adj_display, p.adj.signif)
    ) %>%
    cols_align(
      align = "left",
      columns = c(comparison, direction)
    )
  
  if (save_html) {
    gtsave(
      data = gt_tbl,
      filename = output_html
    )
  }
  
  gt_tbl
}

cibersortx_dunn_gt <- plot_cibersortx_dunn_table(
  ctme_dunn = CIBERSORTx_dunn,
  title = "CIBERSORTx immune cell differences across C1 consensus groups",
  subtitle = "Dunn post-hoc test after Kruskal-Wallis; BH-adjusted significant comparisons only",
  output_html = "CIBERSORTx_Dunn_significant_comparisons.html",
  save_html = TRUE
)

cibersortx_dunn_gt

```

### CIBERSORTx Extreme-Group Comparison

```{r}

ctme_extreme <- ctme_long %>%
  filter(C1_consensus_group %in% c("Not-C1-like", "C1-like")) %>%
  mutate(
    C1_consensus_group = factor(
      C1_consensus_group,
      levels = c("Not-C1-like", "C1-like")
    )
  )

p_ctme_extreme <- ggplot(
  ctme_extreme,
  aes(
    x = C1_consensus_group,
    y = score,
    fill = C1_consensus_group
  )
) +
  geom_violin(
    trim = FALSE,
    alpha = 0.75,
    color = "grey30",
    linewidth = 0.25
  ) +
  geom_boxplot(
    width = 0.18,
    outlier.shape = NA,
    alpha = 0.85,
    color = "black",
    linewidth = 0.3
  ) +
  stat_compare_means(
    size = 4
  ) +
  facet_wrap(
    ~ consensusTME,
    scales = "free_y",
    ncol = 4
  ) +
  scale_fill_npg() +
  labs(
    title = "CIBERSORTx singScore deconvolution: Not-C1-like vs C1-like",
    x = NULL,
    y = "CIBERSORTx singScore",
    fill = "C1 consensus group"
  ) +
  theme_pubclean(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    strip.text = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

p_ctme_extreme

ggsave(
  filename = "output/figures/TCGA_LUAD_CIBERSORTx_Not_C1_like_vs_C1_like.png",
  plot = p_ctme_extreme,
  width = 24,
  height = 20,
  dpi = 600
)


```

# DNA Methylation Deconvolution

## Methylation-Based Deconvolution Input

### Load the Methylation Matrix for Deconvolution

```{r}

meth_mvalue <- vroom::vroom(
  "input/tables/Methylation_TCGA_M_VALUES.csv",
  show_col_types = FALSE
) |>
  column_to_rownames("probe_id") |>
  as.matrix()

meth_beta <- plogis(meth_mvalue * log(2))

meth_deconv_samples <- intersect(colnames(meth_beta), C1_Core$sample)

meth_beta_deconv <- meth_beta[, meth_deconv_samples, drop = FALSE]

meth_deconv_input_summary <- tibble(
  n_probes = nrow(meth_beta_deconv),
  n_samples = ncol(meth_beta_deconv),
  min_value = min(meth_beta_deconv, na.rm = TRUE),
  max_value = max(meth_beta_deconv, na.rm = TRUE)
)

meth_deconv_input_summary


```

```{r}

library(EpiDISH)
data(centEpiFibIC.m)
data(centBloodSub.m)
frac.m <- hepidish(beta.m = meth_beta_deconv, ref1.m = centEpiFibIC.m, ref2.m = centBloodSub.m, h.CT.idx = 3, method = 'RPC')
frac.m

```

### Placeholder for Methylation Deconvolution Analyses

This section mirrors the RNA-based deconvolution blocks and provides a dedicated entry point for methylation-derived cell-fraction analyses.

# Tumor Purity and Deconvolution Heatmaps

## Build C1-Core and Purity Metadata

```{r}

library(TCGAbiolinks)

C1_Core <- vroom::vroom(
  "output/tables/C1_core_singscore_GSVA_consensus_scores.csv",
  show_col_types = FALSE
) |>
  dplyr::select(sample, C1_consensus, C1_consensus_group) |>
  dplyr::mutate(
    sample = as.character(sample),
    case_submitter_id = substr(sample, 1, 12),
    C1_consensus = as.numeric(C1_consensus),
    C1_consensus_group = factor(
      C1_consensus_group,
      levels = c(
        "Not-C1-like",
        "Not-C1-intermediate",
        "C1-intermediate",
        "C1-like"
      )
    )
  ) |>
  dplyr::distinct(sample, .keep_all = TRUE)

data("Tumor.purity", package = "TCGAbiolinks")

parse_purity <- function(x) {
  x <- as.character(x)
  x[x %in% c("NaN", "NA", "NULL", "", "nan")] <- NA_character_
  as.numeric(gsub(",", ".", x))
}

tumor_purity_luad <- TCGAbiolinks::Tumor.purity |>
  janitor::clean_names() |>
  dplyr::filter(cancer_type == "LUAD") |>
  dplyr::mutate(
    case_submitter_id = substr(as.character(sample_id), 1, 12),
    purity_cpe = parse_purity(cpe),
    purity_absolute = parse_purity(absolute),
    purity_estimate = parse_purity(estimate),
    purity_ihc = parse_purity(ihc),
    purity_primary = dplyr::coalesce(
      purity_cpe,
      purity_absolute,
      purity_estimate,
      purity_ihc
    )
  ) |>
  dplyr::select(case_submitter_id, purity_primary) |>
  dplyr::distinct(case_submitter_id, .keep_all = TRUE)

C1_Core <- C1_Core |>
  dplyr::left_join(
    tumor_purity_luad,
    by = "case_submitter_id"
  )

C1_Core

CTME_deconv_singScore <- CTME_deconv_singScore |>
  as.data.frame() |>
  column_to_rownames("consensusTME") |>
  t() |>
  as.data.frame()
  #filter(consensusTME != "Immune_Score")

C1_Core <- C1_Core |>
  as.data.frame() |>
  column_to_rownames("sample")

CIBERSORTx <- vroom("input/CIBERSORTx_Job6_Results.csv") |>
  as.data.frame() |>
  column_to_rownames("Mixture") |>
  select(-`P-value`,-`Absolute score (sig.score)`, -RMSE, -Correlation)

```

## Deconvolution Heatmap Helper

```{r}
#| label: deconv-complexheatmap-c1-purity
#| echo: false
#| warning: false
#| message: false
#| results: hide

library(tidyverse)
library(ComplexHeatmap)
library(circlize)
library(viridis)
library(grid)

plot_deconv_C1_purity_heatmap <- function(
  deconv_mat,
  metadata_tbl,
  deconv_title = "Deconvolution",
  sample_col = NULL,
  c1_score_col = "C1_consensus",
  c1_group_col = "C1_consensus_group",
  purity_col = "purity_primary",
  remove_immune_score = TRUE,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  clustering_method = "average",
  save_png = TRUE,
  filename = NULL,
  width = 14,
  height = 8,
  dpi = 600
) {
  row_zscore <- function(mat) {
    mat <- as.matrix(mat)

    row_mean <- rowMeans(mat, na.rm = TRUE)
    row_sd <- apply(mat, 1, sd, na.rm = TRUE)
    row_sd[is.na(row_sd) | row_sd == 0] <- 1

    mat_scaled <- sweep(mat, 1, row_mean, "-")
    mat_scaled <- sweep(mat_scaled, 1, row_sd, "/")
    mat_scaled[is.na(mat_scaled)] <- 0

    mat_scaled
  }

  pearson_row_hclust <- function(mat) {
    cor_mat <- cor(
      t(mat),
      method = "pearson",
      use = "pairwise.complete.obs"
    )

    cor_mat[is.na(cor_mat)] <- 0

    hclust(
      as.dist(1 - cor_mat),
      method = clustering_method
    )
  }

  pearson_column_hclust <- function(mat) {
    cor_mat <- cor(
      mat,
      method = "pearson",
      use = "pairwise.complete.obs"
    )

    cor_mat[is.na(cor_mat)] <- 0

    hclust(
      as.dist(1 - cor_mat),
      method = clustering_method
    )
  }

  if (is.null(filename)) {
    filename_safe <- deconv_title |>
      stringr::str_replace_all("[^A-Za-z0-9]+", "_") |>
      stringr::str_replace_all("_+$", "") |>
      stringr::str_replace_all("^_+", "")

    filename <- paste0(
      "TCGA_LUAD_",
      filename_safe,
      "_C1_purity_heatmap.png"
    )
  }

  deconv_df <- deconv_mat |>
    as.data.frame()

  if ("consensusTME" %in% colnames(deconv_df)) {
    rownames(deconv_df) <- deconv_df$consensusTME

    deconv_df <- deconv_df |>
      dplyr::select(-consensusTME)
  }

  deconv_numeric <- deconv_df |>
    dplyr::mutate(
      dplyr::across(
        dplyr::everything(),
        as.numeric
      )
    ) |>
    as.matrix()

  sample_like_rownames <- grepl("^TCGA-", rownames(deconv_numeric)[1])
  sample_like_colnames <- grepl("^TCGA-", colnames(deconv_numeric)[1])

  if (sample_like_rownames && !sample_like_colnames) {
    deconv_numeric <- t(deconv_numeric)
  }

  if (remove_immune_score) {
    deconv_numeric <- deconv_numeric[
      !rownames(deconv_numeric) %in% "Immune_Score",
      ,
      drop = FALSE
    ]
  }

  metadata_df <- metadata_tbl |>
    as.data.frame()

  if (is.null(sample_col)) {
    if ("sample" %in% colnames(metadata_df)) {
      sample_col <- "sample"
    } else {
      metadata_df$sample <- rownames(metadata_df)
      sample_col <- "sample"
    }
  }

  metadata_df <- metadata_df |>
    dplyr::mutate(
      sample = as.character(.data[[sample_col]]),
      C1_consensus = as.numeric(.data[[c1_score_col]]),
      C1_consensus_group = factor(
        .data[[c1_group_col]],
        levels = c(
          "Not-C1-like",
          "Not-C1-intermediate",
          "C1-intermediate",
          "C1-like"
        )
      ),
      purity_primary = as.numeric(.data[[purity_col]])
    ) |>
    dplyr::select(
      sample,
      C1_consensus,
      C1_consensus_group,
      purity_primary
    ) |>
    dplyr::distinct(sample, .keep_all = TRUE)

  matched_samples <- intersect(
    colnames(deconv_numeric),
    metadata_df$sample
  )

  if (length(matched_samples) == 0) {
    stop("No matching samples between deconvolution matrix and metadata.")
  }

  metadata_matched <- metadata_df |>
    dplyr::filter(sample %in% matched_samples) |>
    dplyr::arrange(
      C1_consensus_group,
      C1_consensus,
      purity_primary
    )

  sample_order <- metadata_matched$sample

  heat_mat <- deconv_numeric[
    ,
    sample_order,
    drop = FALSE
  ]

  heat_mat_scaled <- row_zscore(heat_mat)

  annotation_df <- metadata_matched |>
    dplyr::select(
      C1_consensus_group,
      C1_consensus,
      purity_primary
    ) |>
    as.data.frame()

  colnames(annotation_df) <- c(
    "C1 group",
    "C1 score",
    "Purity"
  )

  rownames(annotation_df) <- sample_order

  group_colors <- c(
    "Not-C1-like" = "#4D4D4D",
    "Not-C1-intermediate" = "#A6A6A6",
    "C1-intermediate" = "#FDB863",
    "C1-like" = "#D73027"
  )

  c1_score_range <- range(
    annotation_df[["C1 score"]],
    na.rm = TRUE
  )

  c1_score_col_fun <- circlize::colorRamp2(
    c(
      c1_score_range[1],
      0,
      c1_score_range[2]
    ),
    c(
      "#2166AC",
      "white",
      "#B2182B"
    )
  )

  purity_range <- range(
    annotation_df[["Purity"]],
    na.rm = TRUE
  )

  purity_mid <- median(
    annotation_df[["Purity"]],
    na.rm = TRUE
  )

  purity_col_fun <- circlize::colorRamp2(
    c(
      purity_range[1],
      purity_mid,
      purity_range[2]
    ),
    c(
      "#F7F7F7",
      "#BDBDBD",
      "#252525"
    )
  )

  ha_col <- ComplexHeatmap::HeatmapAnnotation(
    df = annotation_df,
    col = list(
      "C1 group" = group_colors,
      "C1 score" = c1_score_col_fun,
      "Purity" = purity_col_fun
    ),
    annotation_name_gp = grid::gpar(fontsize = 9),
    na_col = "grey85"
  )

  ht <- ComplexHeatmap::Heatmap(
    heat_mat_scaled,
    name = "row z-score",
    col = circlize::colorRamp2(
      c(-2, 0, 2),
      viridis::viridis(3)
    ),
    use_raster = TRUE,
    top_annotation = ha_col,

    cluster_rows = if (cluster_rows) pearson_row_hclust else FALSE,
    cluster_columns = if (cluster_columns) pearson_column_hclust else FALSE,

    column_split = annotation_df[["C1 group"]],
    cluster_column_slices = FALSE,

    show_row_names = TRUE,
    show_column_names = FALSE,

    row_names_gp = grid::gpar(fontsize = 9),

    column_title = paste0(
      deconv_title,
      " deconvolution ordered by C1 consensus group and tumor purity"
    ),
    column_title_gp = grid::gpar(
      fontsize = 12,
      fontface = "bold"
    ),

    heatmap_legend_param = list(
      title = "Cell type\nz-score"
    )
  )

  ComplexHeatmap::draw(
    ht,
    heatmap_legend_side = "right",
    annotation_legend_side = "right"
  )

  if (save_png) {
    png(
      filename = filename,
      width = width,
      height = height,
      units = "in",
      res = dpi
    )

    ComplexHeatmap::draw(
      ht,
      heatmap_legend_side = "right",
      annotation_legend_side = "right"
    )

    dev.off()
  }

  invisible(ht)
}

```

## Deconvolution Heatmaps Ordered by C1-Core and Purity

```{r}

plot_deconv_C1_purity_heatmap(
  deconv_mat = CTME_deconv_singScore,
  metadata_tbl = C1_Core,
  deconv_title = "ConsensusTME singScore",
  save_png = TRUE,
  filename = "output/figures/TCGA_LUAD_ConsensusTME_singScore_C1_purity_heatmap.png",
  width = 14,
  height = 8,
  dpi = 600
)

plot_deconv_C1_purity_heatmap(
  deconv_mat = CIBERSORTx,
  metadata_tbl = C1_Core,
  deconv_title = "CIBERSORTx",
  save_png = TRUE,
  filename = "output/figures/TCGA_LUAD_CIBERSORTx_C1_purity_heatmap.png",
  width = 14,
  height = 8,
  dpi = 600
)

```


# Deconvolution Association with C1-Core

## ConsensusTME Association Analysis

```{r}
#| label: c1-core-consensustme-factor-correlations

# Convert deconvolution matrix from wide to long format
ctme_long <- CTME_deconv_singScore %>%
  pivot_longer(
    cols = -consensusTME,
    names_to = "sample",
    values_to = "score"
  ) %>%
  left_join(
    c1_score_tbl %>%
      select(sample, C1_consensus, C1_consensus_group),
    by = "sample"
  ) %>%
  filter(!is.na(C1_consensus_group))

# Check group order
levels(ctme_long$C1_consensus_group)

ctme_long <- ctme_long %>%
  mutate(
    C1_consensus_group = factor(
      C1_consensus_group,
      levels = c(
        "Not-C1-like",
        "Not-C1-intermediate",
        "C1-intermediate",
        "C1-like"
      )
    )
  ) |>
    filter(consensusTME != "Immune_Score")

# Test each MOFA factor against the continuous C1-core score.
cor_tbl <- ctme_long |>
  dplyr::group_by(consensusTME) |>
  dplyr::summarise(
    n = sum(stats::complete.cases(score, C1_consensus)),

    spearman_rho = stats::cor(
      score,
      C1_consensus,
      method = "spearman",
      use = "complete.obs"
    ),

    pearson_r = stats::cor(
      score,
      C1_consensus,
      method = "pearson",
      use = "complete.obs"
    ),

    spearman_p = stats::cor.test(
      score,
      C1_consensus,
      method = "spearman",
      exact = FALSE
    )$p.value,

    pearson_p = stats::cor.test(
      score,
      C1_consensus,
      method = "pearson"
    )$p.value,

    .groups = "drop"
  ) |>
  dplyr::mutate(
    spearman_fdr = stats::p.adjust(spearman_p, method = "BH"),
    pearson_fdr = stats::p.adjust(pearson_p, method = "BH"),
    abs_spearman_rho = abs(spearman_rho)
  ) |>
  dplyr::arrange(dplyr::desc(abs_spearman_rho))

cor_tbl



```


### Correlation Summary

```{r}
#| label: c1-core-consensustme-correlation-barplot

p_cor_bar <- cor_tbl |>
  dplyr::mutate(
    consensusTME = forcats::fct_reorder(consensusTME, spearman_rho),
    significant = spearman_fdr < 0.05,
    direction = dplyr::if_else(spearman_rho >= 0, "Positive", "Negative"),
    label = dplyr::if_else(
      significant,
      sprintf("rho = %.2f", spearman_rho),
      ""
    ),
    label_y = spearman_rho + dplyr::if_else(spearman_rho >= 0, 0.035, -0.035),
    label_hjust = dplyr::if_else(spearman_rho >= 0, 0, 1)
  ) |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = consensusTME,
      y = spearman_rho,
      fill = direction,
      alpha = significant
    )
  ) +
  ggplot2::geom_hline(yintercept = 0, linewidth = 0.4, colour = "grey35") +
  ggplot2::geom_col(width = 0.72, colour = "black", linewidth = 0.15) +
  ggplot2::geom_text(
    ggplot2::aes(y = label_y, label = label, hjust = label_hjust),
    size = 3.1,
    lineheight = 0.9
  ) +
  ggplot2::coord_flip(clip = "off") +
  ggsci::scale_fill_npg(name = "Direction") +
  ggplot2::scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.35), guide = "none") +
  ggplot2::scale_y_continuous(
    limits = c(
      min(cor_tbl$spearman_rho, na.rm = TRUE) - 0.18,
      max(cor_tbl$spearman_rho, na.rm = TRUE) + 0.18
    ),
    breaks = scales::pretty_breaks(n = 6)
  ) +
  ggplot2::labs(
    title = "C1-Core-associated consensusTMEs",
    subtitle = "Spearman association between consensusTME and continuous C1-Core score",
    x = NULL,
    y = expression("Spearman " * rho),
    caption = "Labels are shown only for FDR-significant associations."
  ) +
  ggpubr::theme_pubclean(base_size = 12) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold", size = 14),
    plot.subtitle = ggplot2::element_text(size = 11, colour = "grey25"),
    plot.caption = ggplot2::element_text(size = 9, colour = "grey35", hjust = 0),
    axis.text.y = ggplot2::element_text(face = "bold", colour = "black"),
    axis.text.x = ggplot2::element_text(colour = "black"),
    axis.title.x = ggplot2::element_text(face = "bold"),
    legend.position = "right",
    legend.title = ggplot2::element_text(face = "bold"),
    plot.margin = ggplot2::margin(8, 35, 8, 8)
  )

p_cor_bar

ggsave(
  p_cor_bar,
  filename = "output/figures/consensusTME_correlation_barplot.png",
  width = 12,
  height = 8,
  dpi = 600
)

```


## CIBERSORTx Association Analysis

```{r}
#| label: c1-core-cibersortx-factor-correlations


# Convert deconvolution matrix from wide to long format
CIBERSORTx_dunn <- CIBERSORTx %>%
  t() |>
  as.data.frame() |>
  rownames_to_column("CIBERSORTx") |>
  pivot_longer(
    cols = -CIBERSORTx,
    names_to = "sample",
    values_to = "score"
  ) %>%
  left_join(
    c1_score_tbl %>%
      select(sample, C1_consensus, C1_consensus_group),
    by = "sample"
  ) %>%
  filter(!is.na(C1_consensus_group))

# Check group order
levels(CIBERSORTx_dunn$C1_consensus_group)

CIBERSORTx_dunn <- CIBERSORTx_dunn %>%
  mutate(
    C1_consensus_group = factor(
      C1_consensus_group,
      levels = c(
        "Not-C1-like",
        "Not-C1-intermediate",
        "C1-intermediate",
        "C1-like"
      )
    )
  )

# Test each MOFA factor against the continuous C1-core score.
cor_tbl <- CIBERSORTx_dunn |>
  dplyr::group_by(CIBERSORTx) |>
  dplyr::summarise(
    n = sum(stats::complete.cases(score, C1_consensus)),

    spearman_rho = stats::cor(
      score,
      C1_consensus,
      method = "spearman",
      use = "complete.obs"
    ),

    pearson_r = stats::cor(
      score,
      C1_consensus,
      method = "pearson",
      use = "complete.obs"
    ),

    spearman_p = stats::cor.test(
      score,
      C1_consensus,
      method = "spearman",
      exact = FALSE
    )$p.value,

    pearson_p = stats::cor.test(
      score,
      C1_consensus,
      method = "pearson"
    )$p.value,

    .groups = "drop"
  ) |>
  dplyr::mutate(
    spearman_fdr = stats::p.adjust(spearman_p, method = "BH"),
    pearson_fdr = stats::p.adjust(pearson_p, method = "BH"),
    abs_spearman_rho = abs(spearman_rho)
  ) |>
  dplyr::arrange(dplyr::desc(abs_spearman_rho))

cor_tbl



```


### Correlation Summary

```{r}
#| label: c1-core-cibersortx-correlation-barplot

p_cor_bar <- cor_tbl |>
  dplyr::mutate(
    CIBERSORTx = forcats::fct_reorder(CIBERSORTx, spearman_rho),
    significant = spearman_fdr < 0.05,
    direction = dplyr::if_else(spearman_rho >= 0, "Positive", "Negative"),
    label = dplyr::if_else(
      significant,
      sprintf("rho = %.2f", spearman_rho),
      ""
    ),
    label_y = spearman_rho + dplyr::if_else(spearman_rho >= 0, 0.035, -0.035),
    label_hjust = dplyr::if_else(spearman_rho >= 0, 0, 1)
  ) |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = CIBERSORTx,
      y = spearman_rho,
      fill = direction,
      alpha = significant
    )
  ) +
  ggplot2::geom_hline(yintercept = 0, linewidth = 0.4, colour = "grey35") +
  ggplot2::geom_col(width = 0.72, colour = "black", linewidth = 0.15) +
  ggplot2::geom_text(
    ggplot2::aes(y = label_y, label = label, hjust = label_hjust),
    size = 3.1,
    lineheight = 0.9
  ) +
  ggplot2::coord_flip(clip = "off") +
  ggsci::scale_fill_npg(name = "Direction") +
  ggplot2::scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.35), guide = "none") +
  ggplot2::scale_y_continuous(
    limits = c(
      min(cor_tbl$spearman_rho, na.rm = TRUE) - 0.18,
      max(cor_tbl$spearman_rho, na.rm = TRUE) + 0.18
    ),
    breaks = scales::pretty_breaks(n = 6)
  ) +
  ggplot2::labs(
    title = "C1-Core-associated CIBERSORTx",
    subtitle = "Spearman association between CIBERSORTx and continuous C1-Core score",
    x = NULL,
    y = expression("Spearman " * rho),
    caption = "Labels are shown only for FDR-significant associations."
  ) +
  ggpubr::theme_pubclean(base_size = 12) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold", size = 14),
    plot.subtitle = ggplot2::element_text(size = 11, colour = "grey25"),
    plot.caption = ggplot2::element_text(size = 9, colour = "grey35", hjust = 0),
    axis.text.y = ggplot2::element_text(face = "bold", colour = "black"),
    axis.text.x = ggplot2::element_text(colour = "black"),
    axis.title.x = ggplot2::element_text(face = "bold"),
    legend.position = "right",
    legend.title = ggplot2::element_text(face = "bold"),
    plot.margin = ggplot2::margin(8, 35, 8, 8)
  )

p_cor_bar

ggsave(
  p_cor_bar,
  filename = "output/figures/CIBERSORTx_correlation_barplot.png",
  width = 12,
  height = 8,
  dpi = 600
)

```
