---
title: "MAP 00 - Seurat QC and Integration"
format: html
---

# Section 1. Overview

This notebook builds the integrated CSS epithelial single-cell object used by the rest of the LUAD plasticity workflow.

**MAP role:** mandatory foundation notebook. Keep QC, doublet removal, integration, clustering, annotation, LISI, and object export. Deprioritize ad hoc exploratory plots unless they support QC or reproducibility decisions.

## Section 1.1. Reader Guide

- Inputs are the four raw 10x `.h5` matrices under `input/ROC-25-0074` through `input/ROC-25-0077`.
- The notebook performs sample-level QC, filtering, merging, normalization, Harmony integration, clustering, and annotation.
- Main outputs are QC figures, QC tables, RNA layer exports, integration diagnostics, and the processed Seurat object.
- Read this as the foundation step: every plasticity score, trajectory, velocity result, and module analysis downstream depends on this object.

## Section 1.2. MAP Section Boundaries

### Section 1.2.1. Notebook Flow

1. Setup and parameters.
2. Load four CSS samples.
3. Per-sample QC and doublet filtering.
4. Merge cleaned samples.
5. Normalize, reduce dimensions, integrate, cluster, and annotate.
6. Export RNA layers and processed objects.
7. Evaluate integration quality with LISI and final diagnostic plots.

```{r include=FALSE}
# Keep the original project-relative paths working from inside code_MAP/
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

## Section 1.3. Setup

This section loads the required packages and restores the Seurat object used as the starting point for downstream analyses.

```{r include=FALSE}

packages <- c(
  "tidyverse",
  "Seurat",
  "SeuratWrappers",
  "harmony",
  "patchwork",
  "data.table",
  "SingleR",
  "fst",
  "UCell",
  "CytoTRACE2",
  "slingshot",
  "tradeSeq",
  "viridis",
  "RColorBrewer",
  "scran",
  "qs2",
  "ggsci",
  "scCustomize", 
  "future",
  "ggpubr",
  "GGally"
)


purrr::walk(packages, library, character.only = TRUE)

plan(multicore, workers = 5)

```

# Section 2. Load and QC the Four CSS Samples

The four CSS samples are loaded independently, merged into a single object, and annotated with standard QC metrics before filtering.

```{r}



# PARAMETERS

set.seed(1234)
params <- list(
  mad_multiplier = 5,  # keep standard MAD (3 is correct baseline)
  output_dir = "output",
  plot_dir = "output/figures/qc_plots",
  qc_dir = "output/tables/qc_metrics"
)

dir.create(params$output_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(params$plot_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(params$qc_dir, showWarnings = FALSE, recursive = TRUE)


# LOAD SAMPLES


load_samples <- function() {
  cat("Loading samples...\n")

  S1 <- CreateSeuratObject(
    Read10X_h5("input/ROC-25-0074/input/filtered_feature_bc_matrix.h5"),
    project = "S1_CSS", min.cells = 3, min.features = 200
  )

  S2 <- CreateSeuratObject(
    Read10X_h5("input/ROC-25-0075/filtered_feature_bc_matrix.h5"),
    project = "S2_CSS", min.cells = 3, min.features = 200
  )

  S3 <- CreateSeuratObject(
    Read10X_h5("input/ROC-25-0076/filtered_feature_bc_matrix.h5"),
    project = "S3_CSS", min.cells = 3, min.features = 200
  )

  S4 <- CreateSeuratObject(
    Read10X_h5("input/ROC-25-0077/filtered_feature_bc_matrix.h5"),
    project = "S4_CSS", min.cells = 3, min.features = 200
  )

  list(Css1 = S1, Css2 = S2, Css3 = S3, Css4 = S4)
}


# QC METRICS


add_qc_metrics <- function(obj) {
  obj$percent.mt <- PercentageFeatureSet(obj, pattern = "^MT-")
  obj
}


# MAD THRESHOLDS (FIXED LOGIC)

# KEY FIX:
# - keep MAD
# - avoid over-trimming upper tails of RNA distributions

compute_mad_thresholds <- function(obj, k = 3) {

  med_feat <- median(obj$nFeature_RNA)
  mad_feat <- mad(obj$nFeature_RNA)

  med_count <- median(obj$nCount_RNA)
  mad_count <- mad(obj$nCount_RNA)

  med_mt <- median(obj$percent.mt)
  mad_mt <- mad(obj$percent.mt)

  thr <- list(

    # -------------------------
    # FEATURES (symmetric OK)
    # -------------------------
    nFeature_low  = max(0, med_feat - k * mad_feat),
    nFeature_high = med_feat + k * mad_feat,

    # -------------------------
    # COUNTS (ASYMMETRIC FIX)
    # preserve high-expression epithelial states
    # -------------------------
    nCount_low  = max(0, med_count - k * mad_count),
    nCount_high = med_count + (k * mad_count * 1.5),

    # -------------------------
    # MITO (UPPER ONLY, BIOLOGICAL FIXED STYLE)
    # -------------------------
    mt_high = med_mt + k * mad_mt,

    # diagnostics
    median_nFeature = med_feat,
    mad_nFeature = mad_feat,
    median_nCount = med_count,
    mad_nCount = mad_count,
    median_mt = med_mt,
    mad_mt = mad_mt,
    mad_multiplier = k
  )

  thr
}


# SAVE THRESHOLDS


save_thresholds <- function(thr, sample_name, qc_dir) {

  thr_df <- data.frame(
    sample = sample_name,
    mad_multiplier = thr$mad_multiplier,

    nFeature_low = thr$nFeature_low,
    nFeature_high = thr$nFeature_high,

    nCount_low = thr$nCount_low,
    nCount_high = thr$nCount_high,

    mt_high = thr$mt_high
  )

  write.table(
    thr_df,
    file = file.path(qc_dir, paste0(sample_name, "_mad_thresholds.tsv")),
    sep = "\t",
    row.names = FALSE,
    quote = FALSE
  )

  thr_df
}


# PLOTTING


plot_qc <- function(obj, sample_name, stage, outdir) {

  p_vln <- VlnPlot(
    obj,
    features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
    pt.size = 0.1,
    ncol = 3
  )

  ggsave(
    file.path(outdir, paste0(sample_name, "_", stage, "_vln.png")),
    p_vln, width = 14, height = 5, dpi = 600
  )

  p_scatter <- FeatureScatter(obj, "nCount_RNA", "nFeature_RNA", pt.size = 0.6)

  ggsave(
    file.path(outdir, paste0(sample_name, "_", stage, "_scatter.png")),
    p_scatter, width = 7, height = 6, dpi = 600
  )

  p_mt <- FeatureScatter(obj, "nCount_RNA", "percent.mt", pt.size = 0.6)

  ggsave(
    file.path(outdir, paste0(sample_name, "_", stage, "_mt.png")),
    p_mt, width = 7, height = 6, dpi = 600
  )
}


# PROCESS SAMPLE


process_sample <- function(obj, sample_name, params) {

  cat("\n============================\n")
  cat("Processing:", sample_name, "\n")
  cat("============================\n")

  obj$sample <- sample_name
  obj <- add_qc_metrics(obj)

  cat("Cells before:", ncol(obj), "\n")

  plot_qc(obj, sample_name, "preQC", params$plot_dir)

  # -------------------------
  # QC thresholds (MAD-based, corrected)
  # -------------------------
  thr <- compute_mad_thresholds(obj, k = params$mad_multiplier)
  save_thresholds(thr, sample_name, params$qc_dir)

  # -------------------------
  # FILTERING
  # -------------------------
  obj_qc <- subset(
    obj,
    subset =
      nFeature_RNA > thr$nFeature_low &
      nFeature_RNA < thr$nFeature_high &
      nCount_RNA > thr$nCount_low &
      nCount_RNA < thr$nCount_high &
      percent.mt < thr$mt_high
  )

  cat("Cells after QC:", ncol(obj_qc), "\n")

  # -------------------------
  # DOUBLETS
  # -------------------------
  sce <- as.SingleCellExperiment(obj_qc)
  sce <- scDblFinder(sce)

  obj_qc$scDblFinder.class <- colData(sce)$scDblFinder.class
  obj_qc$scDblFinder.score <- colData(sce)$scDblFinder.score

  obj_final <- subset(obj_qc, subset = scDblFinder.class == "singlet")

  cat("Final cells:", ncol(obj_final), "\n")

  plot_qc(obj_final, sample_name, "postQC", params$plot_dir)

  summary <- data.frame(
    sample = sample_name,
    cells_before = ncol(obj),
    cells_after_qc = ncol(obj_qc),
    cells_final = ncol(obj_final),
    pct_removed_qc = round(100 * (ncol(obj) - ncol(obj_qc)) / ncol(obj), 2),
    pct_removed_doublets = round(100 * (ncol(obj_qc) - ncol(obj_final)) / ncol(obj_qc), 2)
  )

  write.table(
    summary,
    file = file.path(params$qc_dir, paste0(sample_name, "_summary.tsv")),
    sep = "\t",
    row.names = FALSE,
    quote = FALSE
  )

  list(obj = obj_final, summary = summary)
}


```

# Section 3. Run Pipeline Per Sample
```{r}

set.seed(1234)
css.list <- load_samples()

results <- lapply(names(css.list), function(sample_name) {
  process_sample(css.list[[sample_name]], sample_name, params)
})

names(results) <- names(css.list)

css.list_clean <- lapply(results, `[[`, "obj")
names(css.list_clean) <- names(css.list)

qc_summary <- do.call(rbind, lapply(results, `[[`, "summary"))

write.table(
  qc_summary,
  file = file.path(params$qc_dir, "QC_summary_all_samples.tsv"),
  sep = "\t",
  row.names = FALSE,
  quote = FALSE
)

cat("\nPer-sample QC complete.\n")
print(qc_summary)


# MERGE CLEANED SAMPLES

cat("\nMerging cleaned samples...\n")
css_combined <- merge(
  css.list_clean[[1]],
  y = css.list_clean[2:4],
  add.cell.ids = names(css.list_clean)
)

# Optional: save merged object before downstream analysis
qs_save(css_combined, file = "output/objects/css_combined_postQC.qs2",nthreads = 10,compress_level = 9)

```

```{r}

# to try, but I prefer harmony 
css_combined <- IntegrateLayers(
  object = css_combined,
  method = RPCAIntegration,
  normalization.method = "SCT",
  verbose = T
)

```

```{r}

set.seed(1234)
options(future.globals.maxSize = 25 * 1024^3)  # 6 GB

css_combined <- SCTransform(css_combined)

css_combined <- RunPCA(css_combined)

ElbowPlot_scCustom(css_combined, ndims=50)
ggsave("output/figures/ElbowPlot_scCustom.png", plot=ElbowPlot_scCustom(css_combined, ndims=50), width=12, height=8, dpi = 600)

css_combined <- IntegrateLayers(
  object = css_combined, 
  method = HarmonyIntegration,
  orig.reduction = "pca", 
  normalization.method = "SCT",
  new.reduction = "harmony",
  verbose = TRUE
)

DimPlot_scCustom(seurat_object = css_combined, group.by = "orig.ident",reduction = "harmony")

PC_Plotting(css_combined, dim_number=1)
ggsave("output/figures/PC1_Plotting.png", plot=PC_Plotting(css_combined, dim_number=1), width=10, height=12, dpi = 600)
PC_Plotting(css_combined, dim_number=2)
ggsave("output/figures/PC2_Plotting.png", plot=PC_Plotting(css_combined, dim_number=2), width=10, height=12, dpi = 600)

cat("Finding neighbors and clusters...\n")
DefaultAssay(css_combined) <- "SCT"
css_combined <- FindNeighbors(css_combined, dims = 1:41, reduction = "harmony")
css_combined <- FindClusters(css_combined, resolution = 0.4,algorithm = 4, random.seed = 1234)

cat("Running UMAP...\n")
css_combined <- RunUMAP(css_combined, dims = 1:41,umap.method = "uwot",reduction = "harmony",seed.use = 1234)
css_combined <- RunTSNE(css_combined, dims = 1:41,,reduction = "harmony",seed.use = 1234)

DimPlot_scCustom(seurat_object = css_combined,label = T,label.size = 8,reduction = "tsne") + 
  umap_theme
DimPlot_scCustom(seurat_object = css_combined,label = T,label.size = 8) + 
  umap_theme

library(ggplot2)
library(ggsci)

n_clusters <- length(unique(css_combined$seurat_clusters))

p <- DimPlot_scCustom(
  seurat_object = css_combined,
  group.by = "seurat_clusters",
  colors_use = pal_frontiers()(n_clusters),
  label = TRUE,
  label.size = 7,
  label.box = TRUE
) +
  umap_theme() +
  NoLegend() +
  labs(
    title = "UMAP: Clusters identified using Leiden algorithm",x = "UMAP1", y = "UMAP2"
  )

p

ggsave(
  filename = "output/figures/umap_leiden_clusters.png",
  plot = p,
  width = 8,
  height = 6,
  dpi = 600
)

DimPlot_scCustom(seurat_object = css_combined, group.by = "orig.ident",reduction = "harmony")

DimPlot_scCustom(seurat_object = css_combined, group.by = "orig.ident",label.size = 8)

DimPlot_scCustom(seurat_object = css_combined, split.by = "orig.ident",label.size = 7)


# FINAL PLOTS

cat("Saving final plots...\n")

p_umap_sample <- DimPlot(css_combined, group.by = "orig.ident", label = FALSE) +
  ggtitle("UMAP by sample")

ggsave(
  filename = file.path(params$plot_dir, "UMAP_by_sample.png"),
  plot = p_umap_sample,
  width = 8,
  height = 6,
  dpi = 600
)

p_umap_clusters <- DimPlot(css_combined, label = TRUE) +
  ggtitle("UMAP by clusters")

ggsave(
  filename = file.path(params$plot_dir, "UMAP_clusters.png"),
  plot = p_umap_clusters,
  width = 8,
  height = 6,
  dpi = 600
)

p_final_scatter <- FeatureScatter(
  css_combined,
  feature1 = "nCount_RNA",
  feature2 = "nFeature_RNA",
  pt.size = 0.5
) +
  ggtitle("Final nCount_RNA vs nFeature_RNA")

ggsave(
  filename = file.path(params$plot_dir, "final_scatter.png"),
  plot = p_final_scatter,
  width = 7,
  height = 6,
  dpi = 600
)

p_final_mt <- FeatureScatter(
  css_combined,
  feature1 = "nCount_RNA",
  feature2 = "percent.mt",
  pt.size = 0.5
) +
  ggtitle("Final nCount_RNA vs percent.mt")

ggsave(
  filename = file.path(params$plot_dir, "final_mt_scatter.png"),
  plot = p_final_mt,
  width = 7,
  height = 6,
  dpi = 600
)


cat("\nDone.\n")

```

# Section 4. Export RNA Layers for Downstream Tools

```{r}

DefaultAssay(css_combined) <- "RNA"
css_combined <- NormalizeData(css_combined)
css_combined <- FindVariableFeatures(css_combined)
css_combined <- ScaleData(css_combined)
css_combined <- JoinLayers(css_combined)

DefaultAssay(css_combined) <- "SCT"

css_combined <- Add_Mito_Ribo(object = css_combined, species = "human")

qs_save(object = css_combined, file = "output/objects/css_combined_processed.qs2",compress_level = 10,nthreads = 10)

```


# Section 5. Batch Effect Evaluation with LISI

Batch mixing after Harmony is quantified with LISI and summarized both globally and per sample.

```{r}

set.seed(1234)
library(lisi)

emb <- Embeddings(css_combined, "harmony")

data <- data.frame(
  harmony_1 = emb[,1],
  harmony_2 = emb[,2]
)

meta <- data.frame(batch = css_combined$orig.ident)

lisi <- compute_lisi(data, meta, "batch") |>
  as.data.frame() |>
  rownames_to_column("cell")

fwrite(lisi, "output/tables/LISI_batch.csv")

skimr::skim(lisi)

summary(lisi)

df <- data.frame(
  H1 = data[,1],
  H2 = data[,2],
  LISI = lisi$batch,
  batch = css_combined$orig.ident
)

p1 <- ggplot(df, aes(H1, H2, color = LISI)) +
  geom_point(size = .4, alpha = .8) +
  scale_color_viridis_c(limits = c(1, length(unique(df$batch)))) +
  theme_bw(base_size = 12) +
  labs(title = "Harmony PCA by Local inverse Simpson’s Index (LISI)", x = "harmony 1", y = "harmony 2") +
  theme(legend.position = "bottom")

p1 <- ggMarginal(p1, type="histogram")

ggsave("output/figures/LISI_harmony.svg", plot=p1, width=12, height=8, device="svg", bg = "white")

library(patchwork)

med <- median(df$LISI, na.rm = TRUE)

p1 <- ggplot(df, aes(x = LISI)) +
  geom_histogram(bins = 50, fill = "grey85", color = "black") +
  geom_vline(xintercept = med, linetype = "dashed", color = "red") +
  annotate(
    "text",
    x = med,
    y = Inf,
    label = round(med, 2),
    vjust = 1.5,
    hjust = -0.5,
    color = "red",
    size = 6
  ) +
  theme_bw(base_size = 14) +
  labs(x = "LISI", y = "Counts of cells")

# violin per batch
p2 <- ggplot(df, aes(batch, LISI)) +
  geom_violin(trim = TRUE, fill = "white") +
  geom_boxplot(width = 0.1, outlier.shape = NA) +
  theme_bw(base_size = 14) +
  labs(x = "Batch", y = "LISI")

combined <- (p1 + p2) + 
  plot_annotation(tag_levels = 'A') +
  plot_annotation(title = "Local inverse Simpson’s Index (LISI) distribution & batch")

ggsave(
  "output/figures/LISI_harmony_distribution.svg",
  plot = combined,
  width = 12,
  height = 8,
  device = "svg",
  bg = "white"
)

DimPlot_scCustom(seurat_object = css_combined, group.by = "orig.ident", reduction = "harmony") +
  theme_bw(base_size = 12) +
  scale_color_npg() +
  theme(legend.position = "bottom") +
  labs( title = "")

ggsave(
  "output/figures/harmony_orig_ident.svg",
  width = 10,
  height = 6,
  device = "svg",
  bg = "white"
)

```
