---
title: "MAP 02 - C1 Plasticity Scoring"
format: html
---

# Section 1. Overview

This notebook creates the central C1 plasticity score used throughout the CSS LUAD epithelial plasticity workflow.

**MAP role:** mandatory C1 scoring notebook. Preserve the continuous `C1_score`, the UP/DOWN UCell components, CytoTRACE2 comparison, sample/cluster summaries, and cell-cycle checks.

## Section 1.1. Reader Guide

- Inputs are the processed Seurat object from step 01 and the curated signature table `input/Signatures/Core_genes_tidy.csv`.
- The notebook scores UP and DOWN plasticity signatures with UCell, smooths them on the Harmony graph, and combines them into `C1_score`.
- It cross-checks the score against CytoTRACE2, sample identity, clusters, annotation labels, and cell-cycle metadata.
- Main outputs are score plots, signature presence tables, annotation comparisons, and updated Seurat metadata.
- Read this as the step where the biological idea of plasticity becomes a numeric gradient.

## Section 1.2. MAP Section Boundaries

### Section 1.2.1. Notebook Flow

1. Setup and load processed object.
2. Score UP and DOWN C1 signatures.
3. Build the continuous `C1_score`.
4. Validate C1 against CytoTRACE2 and annotations.
5. Generate essential UMAP, violin, and distribution plots.
6. Export score summaries and updated object metadata.

## Section 1.3. Setup

This section loads the required packages and restores the Seurat object used as the starting point for downstream analyses.

```{r include=FALSE}

packages <- c(
  "tidyverse",
  "Seurat",
  "SeuratWrappers",
  "harmony",
  "patchwork",
  "data.table",
  "fst",
  "UCell",
  "CytoTRACE2",
  "slingshot",
  "tradeSeq",
  "viridis",
  "RColorBrewer",
  "scran",
  "qs2",
  "ggsci",
  "scCustomize", 
  "future",
  "ggpubr",
  "GGally"
)


purrr::walk(packages, library, character.only = TRUE)

plan(multicore, workers = 5)

umap_theme <- ggpubr::theme_classic2() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 11, color = "grey30"),

    axis.title = element_text(size = 13),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),

    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10)
  )


```

# Section 2. Plasticity Scoring with UCell

Plasticity-related signatures are scored with UCell, smoothed on the Harmony graph, and collapsed into a single `C1_score` used throughout the notebook.

```{r UCell score}

set.seed(1234)

library(UCell)
library(BiocParallel)

BPPARAM <- MulticoreParam(workers = 5)
register(BPPARAM)

Core_genes_tidy <- read_csv("input/Signatures/Core_genes_tidy.csv")

gene_list <- list(
  DOWN_Signature = na.omit(Core_genes_tidy$Core_genes_DOWN),
  UP_Signature   = na.omit(Core_genes_tidy$Core_genes_UP)
)

gene_list <- lapply(gene_list, function(g) intersect(g, rownames(css_combined)))

css_combined <- AddModuleScore_UCell(css_combined, features = gene_list)

css_combined <- SmoothKNN(
  css_combined,
  signature.names = paste0(names(gene_list), "_UCell"),
  reduction = "harmony"
)

up   <- css_combined$UP_Signature_UCell_kNN
down <- css_combined$DOWN_Signature_UCell_kNN

score <- scale(up) - scale(down)

css_combined$C1_score_raw <- as.numeric(score)
css_combined$C1_score_scaled <- as.numeric(scale(score))

scale01 <- function(x) {
  r <- range(x, na.rm = TRUE)
  if (r[1] == r[2]) return(rep(0, length(x)))
  (x - r[1]) / (r[2] - r[1])
}

css_combined$C1_score <- scale01(css_combined$C1_score_scaled)

md <- css_combined@meta.data

keep_cols <- c(
  colnames(md)[!grepl("_UCell$", colnames(md))],
  "C1_score_raw",
  "C1_score"
)

css_combined@meta.data <- md[, intersect(keep_cols, colnames(md)), drop = FALSE]

bpstop(BPPARAM)

```


# Section 3. Core Genes in CSS

Before interpreting the signature scores, the notebook checks how many curated core genes are actually present in the merged dataset.

```{r}

Core_genes_tidy <- read_csv("input/Signatures/Core_genes_tidy.csv")

Core_genes_tidy$Core_genes_UP_DOWN %in% rownames(css_combined)

table(Core_genes_tidy$Core_genes_UP_DOWN %in% rownames(css_combined))

gene_presence <- data.frame(
  Gene = Core_genes_tidy$Core_genes_UP_DOWN,
  Present = Core_genes_tidy$Core_genes_UP_DOWN %in% rownames(css_combined)
)

counts <- table(Core_genes_tidy$Core_genes_UP_DOWN %in% rownames(css_combined))
percent <- round(100 * counts / sum(counts), 1)

summary_table <- tibble(
  Present    = names(counts),
  Count      = as.numeric(counts),
  Percentage = percent
)

summary_table

write.table(
  summary_table,
  file = "output/tables/Core_genes_in_samples.csv",
  row.names = FALSE
)


```

# Section 4. UCell Plots

These plots summarize the spatial distribution of the UCell-derived signatures and compare the resulting `C1_score` across samples and Leiden clusters.

```{r UCell plots}

set.seed(1234)

FeaturePlot_scCustom(css_combined, na_cutoff = NA,
  features = c("UP_Signature_UCell_kNN","DOWN_Signature_UCell_kNN",
               "C1_score","C1_score_raw"),
  reduction = "umap") &
  scale_color_viridis_c() & 
  theme_bw(base_size = 12) &
  umap_theme

ggsave("output/figures/UCell_core_genes_UMAP.png", width=12, height=8)

FeaturePlot_scCustom(css_combined, na_cutoff = NA,
  features = "C1_score",
  reduction = "umap") +
  scale_color_viridis_c() +
    labs(
    x = "UMAP1",
    y = "UMAP2",
    title = "C1 score") +
  umap_theme

ggsave("output/figures/C1_score_UMAP.png", width=8, height=6,dpi = 600)

c1_umap_p <- FeaturePlot_scCustom(
  seurat_object = css_combined,
  features = "C1_score",
  reduction = "umap",
  na_cutoff = NA,
  order = TRUE
) +
  scale_color_viridis_c(
    option = "magma",
    name = "C1 score"
  ) +
  labs(
    x = "UMAP 1",
    y = "UMAP 2",
    title = "C1 trajectory score"
  ) +
  umap_theme

c1_umap_p

ggsave("output/figures/C1_score_raw_UMAP.png", width=8, height=6,dpi = 600)

FeaturePlot(css_combined,
  features = c("UP_Signature_UCell_kNN","DOWN_Signature_UCell_kNN",
               "C1_score","C1_score_raw"),
  reduction = "harmony") &
  scale_color_viridis_c() & 
  umap_theme

ggsave("output/figures/UCell_core_genes_PCA.png", width=12, height=8)

umap_df <- FetchData(css_combined, vars = c("C1_score", "orig.ident", "SCT_snn_res.0.1"))

p1 <- ggplot(umap_df, aes(x = orig.ident, y = C1_score, fill = orig.ident)) +
  geom_violin(trim = TRUE, alpha = 0.7, fill = "white") +
  geom_boxplot(width = 0.1, outlier.shape = NA, alpha = 0.5) +
  ggpubr::stat_compare_means() +
  umap_theme +
  labs(
    x = "Sample",
    y = "C1 score",
    title = "C1 score distribution across samples",
    subtitle = "Violin and boxplot comparison by sample identity"
  ) + 
  scale_fill_brewer(palette = "Set2") +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    legend.position = "none"
  )

p1

ggsave("output/figures/C1_score_violin.png", p1, width = 10, height = 6)

p2 <- ggplot(umap_df, aes(x = SCT_snn_res.0.1, y = C1_score, fill = SCT_snn_res.0.1)) +
  geom_violin(trim = TRUE, alpha = 0.7, fill = "white") +
  geom_boxplot(width = 0.1, outlier.shape = NA, alpha = 0.5) +
  ggpubr::stat_compare_means() +
  umap_theme +
  labs(
    x = "Leiden cluster",
    y = "C1 score",
    title = "C1 score across Leiden clusters",
    subtitle = "Distribution of C1 program activity by transcriptional state"
  ) + 
  scale_fill_brewer(palette = "Set2") +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    legend.position = "none"
  )

p2

ggsave("output/figures/leiden_paga_violin.png", p2, width = 10, height = 6)

```

# Section 5. Discrete Plasticity Strata from C1 Score

The continuous `C1_score` is converted into low, medium, and high plasticity strata using the 25th and 75th percentiles.

```{r plasticity strata}

set.seed(1234)
library(fitdistrplus)
library(diptest)

x <- css_combined$C1_score

svg("output/figures/C1_score_descdist.svg", width = 12, height = 8)
descdist(x, discrete = FALSE, boot = 1000)
dev.off()

plotdist(x, histo = TRUE, demp = TRUE)

svg("output/figures/C1_score_norm_fit.svg", width = 12, height = 8)
fw <- fitdist(x, "norm", discrete = FALSE)
plot(fw)
dev.off()

summary(fw)

dip.test(x)

q <- quantile(x, c(.25, .75), type = 8)
q_low  <- q[1]
q_high <- q[2]

css_combined$plasticity <- cut(
  x,
  breaks = c(-Inf, q_low, q_high, Inf),
  labels = c("low", "medium", "high"),
  ordered_result = TRUE
)

df <- data.frame(C1_score = x)

p <- ggplot(df, aes(x = C1_score)) +
  geom_density(linewidth = 1.2, fill = "steelblue", alpha = 0.3) +
  geom_vline(xintercept = q_low, linetype = "dashed", linewidth = 1, color = "red") +
  geom_vline(xintercept = q_high, linetype = "dashed", linewidth = 1, color = "red") +
  annotate("text", x = q_low, y = 0.3 * max(density(x)$y), label = "Q25", color = "red", hjust = 1.1) +
  annotate("text", x = q_high, y = 0.3 * max(density(x)$y), label = "Q75", color = "red", hjust = -0.1) +
  labs(
    title = "Distribution of C1 score (kNN-smoothed)",
    subtitle = "Dashed red lines indicate Q25 and Q75 thresholds",
    x = "C1 score",
    y = "Density"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text( size = 12)
  )

p

ggsave("output/figures/C1_score_distribution.png", plot = p, width = 12, height = 8)

```


# Section 6. CytoTRACE2 Potency Scoring

CytoTRACE2 is run on raw RNA counts to provide an orthogonal estimate of differentiation potency that can be compared against the UCell-derived plasticity score.

```{r}

set.seed(1234)
library(BiocParallel)
library(CytoTRACE2)

cts <- GetAssayData(css_combined, assay="RNA", layer="counts")

ct <- cytotrace2(cts, is_seurat=FALSE, species="human", seed=1234,ncores = 5)

css_combined$CytoTRACE2 <- ct$CytoTRACE2_Score
css_combined$CytoTRACE2_Potency <- factor(ct$CytoTRACE2_Potency)

#qs_save(object = css_combined,file = "output/objects/css_harmony.qs2",compress_level = 5, nthreads = 10)

```

# Section 7. Plot CytoTRACE

These figures assess agreement between CytoTRACE2 and `C1_score`, while also checking that CytoTRACE2 potency is not merely tracking sample identity.

```{r}

set.seed(1234)

ari <- mclust::adjustedRandIndex(css_combined$CytoTRACE2_Potency, css_combined$orig.ident)
print(paste("Adjusted Rand Index vs batch:", round(ari, 3)))

p1 <- FeaturePlot_scCustom(
  css_combined,
  features = c("CytoTRACE2", "C1_score"),
  reduction = "umap",
  na_cutoff = NA
) &
  scale_color_viridis_c() &
  theme_bw(base_size = 12)

p1

ggsave("output/figures/UCell_and_CytoTRACE2.png", plot = p1, width = 10, height = 6)

Idents(css_combined) <- "CytoTRACE2_Potency"
p2_feat <- FeaturePlot_scCustom(
  css_combined,
  features = "C1_score",
  reduction = "umap",
  na_cutoff = NA
) +
  scale_color_viridis_c(option = "magma") +
  ggtitle("C1 Score") +
  labs(subtitle = "UMAP projection of C1 score (continuous program activity)") +
  umap_theme

p2_cluster <- DimPlot_scCustom(
  css_combined,
  reduction = "umap",
  group.by = "CytoTRACE2_Potency",
  label = TRUE,
  label.size = 5,label.box = T
) +
  ggtitle("CytoTRACE2 Potency") +
  labs(subtitle = "Transcriptional potency inferred by CytoTRACE2")  +
  umap_theme

p2 <- (p2_feat + p2_cluster)

ggsave(
  "output/figures/UCell_and_CytoTRACE2_clusters.png",
  plot = p2,
  width = 12,
  height = 12
)


VlnPlot_scCustom(
  css_combined,
  features = "C1_score",
  group.by = "CytoTRACE2_Potency"
) + theme_bw() +
  theme(legend.position = "none") +
  ggpubr::stat_compare_means()


```

# Section 8. Correlation with C1 Score

The direct association between the UCell-derived plasticity score and CytoTRACE2 is evaluated with a Spearman correlation and pairwise density/scatter panels.

```{r}

cor.test(x = css_combined@meta.data$C1_score, css_combined@meta.data$CytoTRACE2,method = "spearman")

df <- data.frame(
  C1_score = css_combined$C1_score,
  CytoTRACE2 = css_combined$CytoTRACE2
)

library(GGally)
p <- ggpairs(
  df,columns = 1:2,
  lower = list(continuous = wrap("smooth_loess", se = T, color = "lightblue", alpha = 0.5)),
  upper = list(continuous = wrap("cor", size = 10,method = "spearman")),
  diag = list(continuous = wrap("densityDiag"))
) + 
  theme_bw(base_size = 15)

ggsave(plot = p,filename = "output/figures/UCell_and_CytoTRACE2_COR.png", width = 10, height = 6)

```



# Section 9. Reference-Based Cell Annotation

This section projects the integrated object onto the Azimuth lung reference, then tests how the resulting annotation states align with the plasticity score.
```{r}

set.seed(1234)

library(Azimuth)

#ref_data <- tibble(SeuratData::AvailableData())
#options(timeout = 50000)
#SeuratData::InstallData("lungref")

set.seed(12345)
plan(sequential)
css_combined <- RunAzimuth(css_combined, reference = "lungref")


DimPlot_scCustom(css_combined, group.by = "predicted.ann_finest_level") + theme_bw(base_size = 25)
table(css_combined$predicted.ann_finest_level)


```

# Section 10. Annotation-Level Statistical Comparisons
```{r}

library(ggpubr)
library(FSA)
library(rcompanion)

# data
meta <- css_combined@meta.data

# analysis of global differences (Kruskal-Wallis)
kruskal.test(C1_score ~ predicted.ann_finest_level, data = meta)

# analysis of pairwise differences (Wilcoxon)
pairwise.wilcox.test(
  meta$C1_score,
  meta$predicted.ann_finest_level,
  p.adjust.method = "BH"
)

# analysis of post-hoc Dunn test
dunn_res <- dunnTest(
  C1_score ~ predicted.ann_finest_level,
  data = meta,
  method = "bh"
)

dunn_df <- dunn_res$res

sig_dunn <- dunn_df %>%
  as_tibble() %>%
  separate(Comparison, into = c("group1", "group2"), sep = " - ") %>%
  filter(P.adj < 0.05) %>%
  arrange(P.adj) %>%
  mutate(direction = ifelse(Z > 0, "higher", "lower"))

write_csv(sig_dunn, "output/tables/C1_dunn_cells.csv")

# data prep for plotting (global)
meta <- meta %>%
  mutate(
    predicted.ann_finest_level = reorder(
      predicted.ann_finest_level,
      C1_score,
      FUN = max
    )
  )

# plot of C1 score across all cell types
p1 <- ggplot(meta, aes(
  x = predicted.ann_finest_level,
  y = C1_score,
  fill = predicted.ann_finest_level
)) +
  geom_violin(trim = TRUE, scale = "width", alpha = 0.8) +
  geom_boxplot(width = 0.12, outlier.shape = NA, fill = "white") +
  stat_compare_means(label = "p.signif", hide.ns = TRUE) +
  labs(
    x = "Azimuth annotation",
    y = "C1 score",
    title = "C1 program activity across cell states",
    subtitle = "Azimuth finest-level annotation"
  ) +
  coord_flip() +
  cowplot::theme_cowplot() +
  theme(
    plot.title = element_text(size = 15, face = "bold"),
    plot.subtitle = element_text(size = 11, color = "grey30"),
    legend.position = "none"
  )

ggsave("output/figures/C1_by_celltype.png", p1, width = 10, height = 6, dpi = 600)

# analysis subset for focused comparison
df_focus <- meta %>%
  filter(predicted.ann_finest_level %in%
           c("AT2 proliferating", "Suprabasal", "AT2", "Basal resting")) %>%
  mutate(predicted.ann_finest_level = factor(predicted.ann_finest_level))

# plot of selected epithelial states
ggplot(df_focus, aes(
  x = predicted.ann_finest_level,
  y = C1_score,
  fill = predicted.ann_finest_level
)) +
  geom_violin(trim = TRUE, alpha = 0.25, color = NA) +
  geom_boxplot(width = 0.15, fill = "white", color = "black",
               outlier.shape = NA, linewidth = 0.4) +
  scale_fill_brewer(palette = "Set2") +
  stat_compare_means(
    comparisons = list(
      c("AT2", "AT2 proliferating"),
      c("Suprabasal", "Basal resting")
    )
  ) +
  theme_bw(base_size = 12) +
  labs(
    x = "Cell type",
    y = "C1 score",
    title = "C1 score across selected cell types",
    subtitle = "Azimuth annotation"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    legend.position = "none"
  )

ggsave("output/figures/C1_by_celltype_AT2.png", width = 10, height = 6, dpi = 600)

# analysis of effect size (non-parametric)
df_focus2 <- meta %>%
  filter(predicted.ann_finest_level %in% c("Suprabasal", "Basal resting")) %>%
  mutate(predicted.ann_finest_level = factor(predicted.ann_finest_level))

rcompanion::wilcoxonRG(
  x = df_focus2$C1_score,
  g = df_focus2$predicted.ann_finest_level,
  verbose = TRUE
)

# analysis of effect size (non-parametric)
df_focus2 <- meta %>%
  filter(predicted.ann_finest_level %in% c("AT2", "AT2 proliferating")) %>%
  mutate(predicted.ann_finest_level = factor(predicted.ann_finest_level))

rcompanion::wilcoxonRG(
  x = df_focus2$C1_score,
  g = df_focus2$predicted.ann_finest_level,
  verbose = TRUE
)

```


# Section 11. Annotation and Cluster UMAP Views
```{r}

FeaturePlot_scCustom(css_combined, na_cutoff = NA,
  features = "C1_score",
  reduction = "umap") +
  scale_color_viridis_c() +
    labs(
    x = "UMAP1",
    y = "UMAP2",
    title = "C1 score") +
  umap_theme

ggsave("output/figures/C1_score_UMAP.png", width=8, height=6,dpi = 600)

library(scCustomize)
library(ggsci)
library(ggrepel)

# Colors
npg_cols <- pal_npg("nrc")(length(unique(css_combined$seurat_clusters)))
names(npg_cols) <- sort(unique(css_combined$seurat_clusters))

# Base UMAP
p <- DimPlot_scCustom(
  css_combined,
  group.by = "seurat_clusters",
  reduction = "umap",
  label = FALSE
) +
  scale_color_npg() +
  labs(x = "UMAP1", y = "UMAP2", title = "UMAP by leiden clusters") +
  umap_theme +
  theme(legend.position = "none")

# Centroids
centroids <- Embeddings(css_combined, "umap") %>%
  as.data.frame() %>%
  mutate(cluster = css_combined$seurat_clusters) %>%
  group_by(cluster) %>%
  summarise(
    UMAP_1 = median(umap_1),
    UMAP_2 = median(umap_2),
    .groups = "drop"
  )

# Plot + BIGGER labels
p <- p +
geom_label(
  data = centroids,
  aes(UMAP_1, UMAP_2, label = cluster, fill = cluster),
  color = "black",
  fontface = "bold",
  size = 4,
  label.size = 0.6,         # thickness of border
  label.padding = unit(0.4, "lines"),
  label.r = unit(0.35, "lines"),
  show.legend = FALSE
) +
  scale_fill_manual(values = npg_cols)


ggsave("output/figures/leiden_UMAP.png", width=8, height=6,dpi = 600)


library(dplyr)
library(ggrepel)

umap_df <- Embeddings(css_combined, "umap") %>%
  as.data.frame() %>%
  mutate(celltype = css_combined$predicted.ann_finest_level)

centroids <- umap_df %>%
  group_by(celltype) %>%
  summarise(
    UMAP_1 = median(umap_1),
    UMAP_2 = median(umap_2),
    n = n(),
    .groups = "drop"
  )
 
p <- DimPlot_scCustom(
  css_combined,
  group.by = "predicted.ann_finest_level",
  reduction = "umap",
  label = FALSE
) +
  labs(x = "UMAP1", y = "UMAP2", title = "Cell type annotation") +
  umap_theme +
  theme(legend.position = "none")

p +
  geom_label_repel(
    data = centroids,
    aes(
      UMAP_1,
      UMAP_2,
      label = paste0(celltype, "\n(n=", n, ")")
    ),
    size = 4.5,
    fontface = "bold",
    
    # box style
    fill = "white",
    label.size = 0.6,
    label.colour = "black",
    
    # repelling behavior (important)
    box.padding = 0.6,
    point.padding = 0.4,
    max.overlaps = Inf,
    
    # leader lines
    segment.color = "black",
    segment.size = 0.3,
    
    show.legend = FALSE
  )

p <- DimPlot_scCustom(
  css_combined,
  group.by = "predicted.ann_finest_level",
  reduction = "umap",
  label = F
) +
  labs(x = "UMAP1", y = "UMAP2", title = "Cell type annotation") +
  umap_theme 

p

ggsave("output/figures/annotation_UMAP.png", plot = p, width=8, height=6,dpi = 600)
 
```

# Section 12. Assign Cell-Cycle Scores
```{r}

DefaultAssay(css_combined) <- "RNA"

s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

css_combined <- CellCycleScoring(css_combined, s.features = s.genes, g2m.features = g2m.genes, set.ident = TRUE)

# view cell cycle scores and phase assignments
head(css_combined[[]])

```

# Section 13. Save Updated Metadata and Annotation Outputs

```{r}

Idents(css_combined) <- "seurat_clusters"

cols_to_remove <- c(
  # assay junk
  "nCount_SCT", "nFeature_SCT",
  "nCount_tfsulm", "nFeature_tfsulm",
  "nCount_pathwaysulm", "nFeature_pathwaysulm",
  
  # clustering redundancy
  "SCT_snn_res.0.1",
  
  # azimuth levels
  "predicted.ann_level_1", "predicted.ann_level_1.score",
  "predicted.ann_level_2", "predicted.ann_level_2.score",
  "predicted.ann_level_3", "predicted.ann_level_3.score",
  "predicted.ann_level_4", "predicted.ann_level_4.score",
  "predicted.ann_level_5", "predicted.ann_level_5.score",
  "predicted.ann_finest_level.score",
  "mapping.score"
)

css_combined@meta.data <- css_combined@meta.data[ , 
  !(colnames(css_combined@meta.data) %in% cols_to_remove)
]

assays_to_remove <- c(
  "prediction.score.ann_level_1",
  "prediction.score.ann_level_2",
  "prediction.score.ann_level_3",
  "prediction.score.ann_level_4",
  "prediction.score.ann_level_5",
  "prediction.score.ann_finest_level"
)

for (assay in assays_to_remove) {
  if (assay %in% names(css_combined@assays)) {
    css_combined[[assay]] <- NULL
  }
}

head(css_combined@meta.data)

qs_save(object = css_combined,file = "output/objects/css_combined_processed.qs2",compress_level = 10, nthreads = 10)

```
