---
title: "MAP 04 - C1 Gene Programs and tradeSeq"
format: html
---

# Section 1. Overview

This notebook turns the C1 plasticity gradient into a trajectory-style analysis and identifies genes that vary along that axis.

**MAP role:** mandatory gene-program notebook, but only the C1-associated gene evidence is core. Keep Spearman, GAM/tradeSeq, direction concordance, and compact enrichment. Treat Slingshot visualizations as support, not as the state definition.

## Section 1.1. Reader Guide

- Input is the processed Seurat object with C1 score, plasticity labels, and dimensional reductions.
- The notebook fits Slingshot pseudotime and uses tradeSeq-style models to find genes changing along C1.
- It also compares correlation-based and GAM-based gene evidence, exports dynamic-gene tables, and runs enrichment analyses.
- Read this as the gene-dynamics notebook: it asks which genes change as cells move along the plasticity axis.

## Section 1.2. MAP Section Boundaries

### Section 1.2.1. Notebook Flow

1. Setup and load scored object.
2. Fit C1-associated tradeSeq GAMs using the continuous C1 score as pseudotime.
3. Extract genes changing along the C1 axis.
4. Run Spearman gene-C1 correlations.
5. Merge GAM and Spearman evidence.
6. Compare C1-associated genes with core signatures and external marker context.
7. Keep enrichment compact and tied to the dynamic gene evidence.
8. Keep the self-GAM section at the end as a draft sensitivity layer.

## Section 1.3. Setup

This section loads the required packages and restores the Seurat object used as the starting point for downstream analyses.

```{r include=FALSE}

packages <- c(
  "tidyverse",
  "Seurat",
  "SeuratWrappers",
  "harmony",
  "patchwork",
  "data.table",
  "fst",
  "UCell",
  "CytoTRACE2",
  "slingshot",
  "tradeSeq",
  "viridis",
  "RColorBrewer",
  "scran",
  "qs2",
  "ggsci",
  "scCustomize", 
  "future",
  "ggpubr",
  "GGally"
)


purrr::walk(packages, library, character.only = TRUE)

plan(multicore, workers = 4)

umap_theme <- cowplot::theme_cowplot() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 11, color = "grey30"),
    axis.title = element_text(size = 13),
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10)
  )

css_combined <- qs_read(file = "output/objects/css_combined_processed.qs2",nthreads = 10)

```


# Section 2. Fit C1-Associated tradeSeq GAMs

This block treats the continuous C1 score as a one-lineage pseudotime axis. The goal is not to redefine cell states, but to test which genes vary smoothly across the C1 plasticity gradient.

```{r}

library(tradeSeq)
library(SingleCellExperiment)
library(BiocParallel)


# Parallel backend (real control)

#BPPARAM <- MulticoreParam(workers = 5)


# Raw counts only

DefaultAssay(css_combined) <- "RNA"
counts <- GetAssayData(css_combined, layer  = "counts")


# Pseudotime (1 lineage)

pt <- css_combined$C1_score

pseudotime_mat <- matrix(
  pt,
  ncol = 1,
  dimnames = list(colnames(css_combined), "lineage1")
)


# Cell weights

cellweights_mat <- matrix(
  1,
  nrow = ncol(css_combined),
  ncol = 1,
  dimnames = list(colnames(css_combined), "lineage1")
)


# Gene selection (IMPORTANT)

# keep expressed genes first
keep <- Matrix::rowSums(counts > 0) > 20
counts_filt <- counts[keep, ]

# then HVGs if available
genes_use <- VariableFeatures(css_combined,nfeatures = 3000)
genes_use <- intersect(genes_use, rownames(counts_filt))

# HARD CAP
genes_use <- head(genes_use, 2500)


# Fit GAM

library(BiocParallel)
#BPPARAM <- SnowParam(workers = 2, type = "SOCK")  # safer on big matrices

BPPARAM <- MulticoreParam(workers = 3)

set.seed(1234)
DEGs_along_pseudotime <- fitGAM(
  counts = counts[genes_use, ],
  pseudotime = pseudotime_mat,
  cellWeights = cellweights_mat,
  nknots = 5,
  verbose = TRUE,
  parallel = TRUE,
  BPPARAM = BPPARAM,
  family = "nb"
)

qs_save(object = DEGs_along_pseudotime,file = "output/objects/fitGAM_tradeSeq.qs",compress_level = 10,nthreads = 10)

```

# Section 3. Genes Changing Along Pseudotime

This section extracts significant dynamic genes from the fitted GAM object and visualizes the strongest changing genes across cells ordered by C1 score.

```{r}

library(ComplexHeatmap)
library(magick)
library(circlize)
library(viridis)

# This can be interpreted as testing whether the average gene expression is significantly changing along pseudotime.
assoRes <- startVsEndTest(DEGs_along_pseudotime) 
head(assoRes)

assoRes <- assoRes |>
  rownames_to_column("gene") |>
  mutate(FDR = p.adjust(pvalue, method = "BH")) |>
  filter(FDR < 0.05)

#assoRes <- assoRes[order(-abs(assoRes$logFClineage1)), ]

library(tradeSeq)

#yhat <- predictSmooth(DEGs_along_pseudotime, nPoints = 100, gene = 1:2500)

fwrite(x = assoRes, file = "output/tables/along_DEGs.csv")

along_DEGs <- read_csv("output/tables/along_DEGs.csv") 

top_genes <- along_DEGs %>% 
  arrange(desc(waldStat)) %>% 
  #arrange(FDR) |>
  pull(gene) %>% head(50)

# Order cells by pseudotime
cell_ord <- order(css_combined@meta.data$C1_score)

# Expression matrix
mat <- GetAssayData(css_combined, assay = "SCT", layer = "data")[top_genes, cell_ord]
mat_dense <- as.matrix(mat)

# Column annotation: numeric continuous
col_ha <- HeatmapAnnotation(
  "C1 score" = css_combined$C1_score[cell_ord],
  col = list(
    "C1 score" = colorRamp2(
      seq(min(css_combined$C1_score, na.rm = TRUE),
          max(css_combined$C1_score, na.rm = TRUE),
          length.out = 100),
      viridis(100)
    )
  ),
  annotation_height = unit(6, "mm")
)

# Row annotation (optional distribution)
row_ha <- rowAnnotation(
  ExprDist = anno_boxplot(mat_dense, type = "violin", gp = gpar(fill = "lightblue"))
)

colors <- rev(RColorBrewer::brewer.pal(n = 11, name = "RdBu"))
colors.use <- grDevices::colorRampPalette(colors = colors)(100)


# Scale rows (z-score per gene) like pheatmap
mat_scaled <- t(scale(t(mat_dense)))

# Symmetric color scale centered at 0 for z-scores
col_zscore <- colorRamp2(c(-2, 0, 2), c("#313695", "white", "#a50026"))

# PNG
png(
  "output/figures/top_50_genes_complex_logFC.png",
  width = 10,
  height = 7,
  units = "in",
  res = 600
)

ht <- Heatmap(
  mat_scaled,
  name = "Z-score",
  col = col_zscore,
  top_annotation = col_ha,
  left_annotation = row_ha,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  clustering_distance_rows = "pearson",
  show_column_names = FALSE,
  show_row_names = TRUE,
  use_raster = TRUE,
  row_names_gp = gpar(fontsize = 8),
  column_title = "Top 50 genes along trajectory (genes by FDR, cells by C1 score)",
  heatmap_legend_param = list(title = "Z-score")
)

draw(ht)

dev.off()

# PDF
  pdf("output/figures/top_50_genes_complex.pdf", width = 10, height = 8)
Heatmap(
  mat_scaled,
  name = "Z-score",
  col = col_zscore,
  top_annotation = col_ha,
  left_annotation = row_ha,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  show_column_names = FALSE,
  show_row_names = TRUE,
  use_raster = TRUE,
  row_names_gp = gpar(fontsize = 8),
  column_title = "Top 50 genes along trajectory",
  heatmap_legend_param = list(title = "Z-score")
)
dev.off()

```

# Section 4. Correlation of Genes Along the C1 Axis

The correlation layer provides a simpler monotonic check for the GAM results. Genes that agree by both methods are stronger C1-program candidates.

```{r}

css_combined <- qs_read(file = "output/objects/css_combined_processed.qs2",nthreads = 10)

```

```{r}


library(tradeSeq)
library(SingleCellExperiment)
library(BiocParallel)


# Parallel backend (real control)

#BPPARAM <- MulticoreParam(workers = 5)


# Use SCT
DefaultAssay(css_combined) <- "SCT"
mat <- GetAssayData(css_combined, layer = "data")

# Subset genes
genes_use <- VariableFeatures(css_combined,nfeatures = 3000)
mat <- mat[genes_use, ]

# Rank pseudotime once
pt_rank <- rank(pt)

# Rank expression matrix (row-wise)
mat_rank <- t(apply(mat, 1, rank))

library(matrixStats)

# Rank
mat_rank <- rowRanks(as.matrix(mat))
pt_rank <- rank(pt)

# Center
mat_centered <- mat_rank - rowMeans(mat_rank)
pt_centered <- pt_rank - mean(pt_rank)

# Correlation
rho <- as.vector(mat_centered %*% pt_centered) /
       sqrt(rowSums(mat_centered^2) * sum(pt_centered^2))

names(rho) <- rownames(mat)

# P-values
n <- length(pt)
tstat <- rho * sqrt((n - 2) / (1 - rho^2))
pvals <- 2 * pt(-abs(tstat), df = n - 2)

res <- tibble(
  gene = rownames(mat),
  rho = rho,
  pval = pvals ) |>
  mutate(FDR = p.adjust(pval, method = "BH")) |>
  filter(FDR < 0.05)

head(res)

fwrite(x = res, file = "output/tables/COR_along_C1.csv")

```

```{r}

library(Seurat)
library(ComplexHeatmap)
library(circlize)
library(viridis)
library(RColorBrewer)
library(readr)
library(dplyr)
library(grid)

dir.create("output/figures", recursive = TRUE, showWarnings = FALSE)

n_genes <- 50

along_DEGs <- readr::read_csv(
  "output/tables/tradeSeq_gam_on_genes_along_C1.csv",
  show_col_types = FALSE
)

# Select top genes along C1 trajectory
# Here genes are ranked by Wald statistic
top_genes <- along_DEGs %>%
  dplyr::arrange(desc(waldStat)) %>%
  dplyr::pull(gene) %>%
  unique() %>%
  head(n_genes)

# Keep only genes present in the SCT assay
expr_mat_all <- GetAssayData(
  css_combined,
  assay = "SCT",
  layer = "data"
)

top_genes_found <- intersect(top_genes, rownames(expr_mat_all))
top_genes_missing <- setdiff(top_genes, rownames(expr_mat_all))

if (length(top_genes_missing) > 0) {
  message(
    "Missing genes not plotted: ",
    paste(top_genes_missing, collapse = ", ")
  )
}

stopifnot(length(top_genes_found) > 1)

# Order cells by C1 score
cell_ord <- order(css_combined@meta.data$C1_score, na.last = NA)
cells_ordered <- rownames(css_combined@meta.data)[cell_ord]

# Extract expression matrix
# Rows = genes
# Columns = cells ordered by C1 score
mat_raw <- expr_mat_all[top_genes_found, cells_ordered, drop = FALSE]
mat_raw <- as.matrix(mat_raw)

# Row-scale genes
# Equivalent to pheatmap(scale = "row")
mat_scaled <- t(scale(t(mat_raw)))
mat_scaled[is.na(mat_scaled)] <- 0

# C1 annotation values
c1_values <- css_combined@meta.data[cells_ordered, "C1_score_raw"]

# Top annotation: continuous C1 score
col_ha <- HeatmapAnnotation(
  `C1 score` = c1_values,
  col = list(
    `C1 score` = colorRamp2(
      c(
        min(c1_values, na.rm = TRUE),
        median(c1_values, na.rm = TRUE),
        max(c1_values, na.rm = TRUE)
      ),
      viridis(3)
    )
  ),
  annotation_height = unit(5, "mm"),
  annotation_name_gp = gpar(fontsize = 8)
)

# Left annotation: raw expression distribution per gene
row_ha <- rowAnnotation(
  `Exp.` = anno_boxplot(
    mat_raw,
    which = "row",
    type = "violin",
    border = FALSE,
    outline = FALSE,
    width = unit(2.0, "cm"),
    gp = gpar(fill = "grey75", col = "grey35")
  ),
  annotation_name_rot = 0,
  annotation_name_gp = gpar(fontsize = 8)
)

# Color scale for gene z-scores
col_zscore <- colorRamp2(
  c(-2, -1, 0, 1, 2),
  c("#053061", "#4393C3", "#F7F7F7", "#D6604D", "#67001F")
)

# Heatmap title
heatmap_title <- paste0(
  "Top genes changing along C1 trajectory\n",
  nrow(mat_scaled),
  " genes displayed | cells ordered by C1 score | Pearson distance | average linkage"
)

ht_opt$TITLE_PADDING = unit(c(8, 8), "points")

ht <- Heatmap(
  mat_scaled,
  name = "Expression (SCT)",
  col = col_zscore,

  top_annotation = col_ha,
  left_annotation = row_ha,

  cluster_rows = TRUE,
  cluster_columns = FALSE,

  clustering_distance_rows = "pearson",
  clustering_method_rows = "average",

  show_row_dend = TRUE,
  show_column_dend = FALSE,

  show_column_names = FALSE,
  show_row_names = TRUE,

  use_raster = TRUE,
  raster_quality = 5,

  row_names_gp = gpar(fontsize = 8),

  row_title = "Genes",
  column_title = heatmap_title,

  column_title_gp = gpar(
    fontsize = 13,
    fontface = "bold",
    lineheight = 1.15
  ),
  row_title_gp = gpar(
    fontsize = 11,
    fontface = "bold"
  ),

  border = TRUE,
  rect_gp = gpar(col = NA),

  heatmap_legend_param = list(
    title = "Row z-score",
    at = c(-2, -1, 0, 1, 2),
    labels = c("Low", "-1", "0", "+1", "High")
  )
)

png(
  filename = "output/figures/top_50_genes_along_C1_heatmap.png",
  width = 12,
  height = 8,
  units = "in",
  res = 600,
  type = "cairo-png"
)

draw(
  ht,
  heatmap_legend_side = "right",
  annotation_legend_side = "right"
)

dev.off()

pdf(
  file = "output/figures/top_50_genes_along_C1_heatmap.pdf",
  width = 12,
  height = 14,
  useDingbats = FALSE
)

draw(
  ht,
  heatmap_legend_side = "right",
  annotation_legend_side = "right"
)

dev.off()

ht_opt(RESET = TRUE)

```

```{r}

library(Seurat)
library(ComplexHeatmap)
library(circlize)
library(viridis)
library(readr)
library(dplyr)
library(grid)

dir.create("output/figures", recursive = TRUE, showWarnings = FALSE)

n_genes <- 50

along_DEGs <- readr::read_csv(
  "output/tables/tradeSeq_gam_on_genes_along_C1.csv",
  show_col_types = FALSE
)

# Select top genes along C1 trajectory
# Here genes are ranked by Wald statistic
top_genes <- along_DEGs %>%
  dplyr::arrange(desc(waldStat)) %>%
  dplyr::pull(gene) %>%
  unique() %>%
  head(n_genes)

# Extract SCT expression matrix
# Rows = genes
# Columns = cells
expr_mat_all <- GetAssayData(
  css_combined,
  assay = "SCT",
  layer = "data"
)

# Keep only genes present in the SCT assay
top_genes_found <- intersect(top_genes, rownames(expr_mat_all))
top_genes_missing <- setdiff(top_genes, rownames(expr_mat_all))

if (length(top_genes_missing) > 0) {
  message(
    "Missing genes not plotted: ",
    paste(top_genes_missing, collapse = ", ")
  )
}

stopifnot(length(top_genes_found) > 1)

# Align metadata to expression matrix columns
meta <- css_combined@meta.data
meta <- meta[colnames(expr_mat_all), , drop = FALSE]

stopifnot(all(rownames(meta) == colnames(expr_mat_all)))

# Cluster labels
cluster <- as.character(meta$seurat_clusters)

# Count cells per cluster
cluster_counts <- table(cluster)

# Extract selected expression matrix
# Cells x genes for efficient rowsum aggregation
expr_cell_gene <- t(as.matrix(expr_mat_all[top_genes_found, , drop = FALSE]))

# Sum expression by cluster
cluster_sum_mat <- rowsum(
  expr_cell_gene,
  group = cluster,
  reorder = FALSE
)

# Convert cluster sums into cluster means
# Result: clusters x genes
cluster_mean_mat <- sweep(
  cluster_sum_mat,
  MARGIN = 1,
  STATS = as.numeric(cluster_counts[rownames(cluster_sum_mat)]),
  FUN = "/"
)

# Raw heatmap matrix
# Rows = genes
# Columns = clusters
plot_mat_raw <- t(cluster_mean_mat)

# Row-scale genes across clusters
# Equivalent to pheatmap(scale = "row")
plot_mat <- t(scale(t(plot_mat_raw)))
plot_mat[is.na(plot_mat)] <- 0

# Cluster-level annotations
cluster_annot <- meta %>%
  dplyr::group_by(seurat_clusters) %>%
  dplyr::summarise(
    mean_C1_score_raw = mean(C1_Core, na.rm = TRUE),
    n_cells = dplyr::n(),
    .groups = "drop"
  ) %>%
  as.data.frame()

cluster_annot$seurat_clusters <- as.character(cluster_annot$seurat_clusters)

# Match annotation order to heatmap columns
cluster_annot <- cluster_annot[
  match(colnames(plot_mat), cluster_annot$seurat_clusters),
  ,
  drop = FALSE
]

rownames(cluster_annot) <- cluster_annot$seurat_clusters

stopifnot(!any(is.na(cluster_annot$seurat_clusters)))
stopifnot(identical(rownames(cluster_annot), colnames(plot_mat)))

# Top annotation: mean C1 raw and cell count
c1_col_fun <- colorRamp2(
  c(
    min(cluster_annot$mean_C1_score_raw, na.rm = TRUE),
    median(cluster_annot$mean_C1_score_raw, na.rm = TRUE),
    max(cluster_annot$mean_C1_score_raw, na.rm = TRUE)
  ),
  c("#313695", "#F7F7F7", "#A50026")
)

top_ha <- HeatmapAnnotation(
  `Mean C1 raw` = cluster_annot$mean_C1_score_raw,
  `Cells` = anno_barplot(
    cluster_annot$n_cells,
    gp = gpar(fill = "grey40", col = NA),
    border = FALSE,
    height = unit(1.0, "cm")
  ),
  col = list(
    `Mean C1 raw` = c1_col_fun
  ),
  annotation_name_side = "left",
  annotation_name_gp = gpar(fontsize = 8),
  annotation_legend_param = list(
    `Mean C1 raw` = list(title = "Mean C1 raw")
  )
)

# Left annotation: raw cluster-level expression distribution per gene
row_ha <- rowAnnotation(
  `Exp.` = anno_boxplot(
    plot_mat_raw,
    which = "row",
    type = "violin",
    border = FALSE,
    outline = FALSE,
    width = unit(2.0, "cm"),
    gp = gpar(fill = "grey75", col = "grey35")
  ),
  annotation_name_rot = 0,
  annotation_name_gp = gpar(fontsize = 8)
)

# Color scale for gene z-scores
col_zscore <- colorRamp2(
  c(-2, -1, 0, 1, 2),
  c("#053061", "#4393C3", "#F7F7F7", "#D6604D", "#67001F")
)

# Heatmap title
heatmap_title <- paste0(
  "Top genes changing along C1 trajectory by cluster\n",
  nrow(plot_mat),
  " genes displayed | cluster mean SCT expression | row z-score | Pearson distance | average linkage"
)

ht_opt$TITLE_PADDING = unit(c(8, 8), "points")

ht <- Heatmap(
  plot_mat,
  name = "Expression (SCT)",
  col = col_zscore,

  top_annotation = top_ha,
  left_annotation = row_ha,

  cluster_rows = TRUE,
  cluster_columns = TRUE,

  clustering_distance_rows = "pearson",
  clustering_distance_columns = "pearson",
  clustering_method_rows = "average",
  clustering_method_columns = "average",

  show_row_dend = TRUE,
  show_column_dend = TRUE,

  show_column_names = TRUE,
  show_row_names = TRUE,

  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 0,

  row_title = "Genes",
  column_title = heatmap_title,

  column_title_gp = gpar(
    fontsize = 13,
    fontface = "bold",
    lineheight = 1.15
  ),
  row_title_gp = gpar(
    fontsize = 11,
    fontface = "bold"
  ),

  border = TRUE,
  rect_gp = gpar(col = "grey90", lwd = 0.3),

  heatmap_legend_param = list(
    title = "Row z-score",
    at = c(-2, -1, 0, 1, 2),
    labels = c("Low", "-1", "0", "+1", "High")
  )
)

png(
  filename = "output/figures/top_50_genes_along_C1_by_cluster_heatmap.png",
  width = 12,
  height = 8,
  units = "in",
  res = 600,
  type = "cairo-png"
)

draw(
  ht,
  heatmap_legend_side = "right",
  annotation_legend_side = "right"
)

dev.off()

pdf(
  file = "output/figures/top_50_genes_along_C1_by_cluster_heatmap.pdf",
  width = 12,
  height = 14,
  useDingbats = FALSE
)

draw(
  ht,
  heatmap_legend_side = "right",
  annotation_legend_side = "right"
)

dev.off()

ht_opt(RESET = TRUE)

```

# Section 5. Merge Correlation and GAM Results

This section joins the model-based and correlation-based evidence so each gene can be read with both dynamic and monotonic support.

```{r}

library(dplyr)
library(readr)

along_DEGs <- read_csv("output/tables/along_DEGs.csv") 
Corr_DEGs  <- read_csv("output/tables/COR_along_C1.csv") 

gam_df <- along_DEGs %>%
    dplyr::rename(
    waldStat_GAM = waldStat,
    df_GAM = df,
    pvalue_GAM = pvalue,
    logFC_GAM = logFClineage1,
    FDR_GAM = FDR
  )

corr_df <- Corr_DEGs %>%
    dplyr::rename(
    rho_corr = rho,
    pval_corr = pval,
    FDR_corr = FDR
  )

Corr_and_GAMs <- gam_df %>%
  left_join(corr_df, by = "gene")

Corr_and_GAMs <- Corr_and_GAMs %>%
  mutate(
    sig_GAM = FDR_GAM < 0.05,
    sig_corr = FDR_corr < 0.05,
    
    concordant = case_when(
      sig_GAM & sig_corr & sign(logFC_GAM) == sign(rho_corr) ~ TRUE,
      sig_GAM & sig_corr ~ FALSE,
      TRUE ~ NA
    )
  )

Corr_and_GAMs <- Corr_and_GAMs %>%
  mutate(
    score = -log10(FDR_GAM + 1e-300) * sign(logFC_GAM) +
            -log10(FDR_corr + 1e-300) * sign(rho_corr)
  )

fwrite(x = Corr_and_GAMs, file = "output/tables/Corr_and_GAMs_along_C1.csv")

```

# Section 6. Overlap Between C1-Associated and tradeSeq Genes

The overlap checks ask whether the C1-associated dynamic genes recover the original C1 signature and related gene sets.

```{r}

library(ggvenn)

# Load C1 core signature genes used for the overlap check.
Core_genes_tidy <- read_csv("input/Signatures/Core_genes_tidy.csv")

# Prepare gene sets for the Venn diagram.
x <- list(
  "C1 core genes UP"     = na.omit(Core_genes_tidy$Core_genes_UP),
  "C1 core genes DOWN"   = na.omit(Core_genes_tidy$Core_genes_DOWN),
  "DEGs along C1 score" = along_DEGs$gene
)

# Plot and display the overlap between the original C1 genes and dynamic genes.
p <- ggvenn(
  x,
  fill_color   = c("grey90", "red"),
  stroke_size  = 0.5,
  set_name_size = 5
)

p

# Save the Venn diagram as a high-resolution PNG.
ggsave(
  filename = "output/figures/DEGs_along_C1_ggvenn.png",
  plot     = p,
  width    = 6,
  height   = 6,
  dpi      = 600,
  bg       = "white"
)


# Prepare gene sets for enrichment tests.
up_genes      <- na.omit(Core_genes_tidy$Core_genes_UP)
down_genes    <- na.omit(Core_genes_tidy$Core_genes_DOWN)
merged_genes  <- unique(c(up_genes, down_genes))
degs          <- along_DEGs$gene

# Use all genes in the Seurat object as the enrichment universe.
universe <- rownames(css_combined)


# Fisher's Exact Test Function
run_fisher <- function(sig, degs, universe) {
  a <- sum(sig %in% degs)                         # in sig & in DEGs
  b <- sum(sig %in% universe & !(sig %in% degs)) # in sig & not DEGs
  c <- sum(!(universe %in% sig) & universe %in% degs) # not sig & in DEGs
  d <- sum(!(universe %in% sig) & !(universe %in% degs)) # not sig & not DEGs
  
  mat <- matrix(
    c(a, b, c, d),
    nrow = 2,
    dimnames = list(
      Signature = c("InSig", "NotInSig"),
      DEGs      = c("InDEGs", "NotInDEGs")
    )
  )
  
  fisher.test(mat)
}


# Run Fisher Tests
fisher_up      <- run_fisher(up_genes, degs, universe)
fisher_down    <- run_fisher(down_genes, degs, universe)


# Collect Results
results <- data.frame(
  Signature = c("UP", "DOWN"),
  logOddsRatio = c(fisher_up$estimate, fisher_down$estimate),
  Pvalue    = c(fisher_up$p.value, fisher_down$p.value)
)

# Log-transform odds ratios
results$logOddsRatio <- log(results$logOddsRatio)

# View results
results

library(dplyr)

results <- data.frame(
  Signature = c("UP", "DOWN"),
  logOddsRatio = c(fisher_up$estimate, fisher_down$estimate),
  Pvalue = c(fisher_up$p.value, fisher_down$p.value)
) %>%
  mutate(
    logOddsRatio = log(logOddsRatio),
    negLog10P = -log10(Pvalue),
    significance = case_when(
      Pvalue < 1e-10 ~ "****",
      Pvalue < 1e-5  ~ "***",
      Pvalue < 1e-3  ~ "**",
      Pvalue < 0.05  ~ "*",
      TRUE ~ "ns"
    )
  )

write_csv(x = results,file = "output/tables/330_along_C1.csv")

library(gt)
pretty_table <- results %>%
  select(Signature, logOddsRatio, Pvalue, significance) %>%
  gt() %>%
  
  tab_header(
    title = "Fisher Enrichment of Gene Signatures",
    subtitle = "Enrichment of UP and DOWN genes within DEG universe"
  ) %>%
  
  fmt_number(columns = logOddsRatio, decimals = 2) %>%
  fmt_scientific(columns = Pvalue, decimals = 2)


small_genes <- tibble(genes = c(
  "HOXB7",
  "THY1",
  "SCGB1A1",
  "TEAD4",
  "E2F4",
  "MCM6",
  "SF3B1",
  "E2F1",
  "RRM2",
  "NUDCD1",
  "HSPG2",
  "SCGB3A1",
  "NKX2-1",
  "SFTPC",
  "TFF1"
))

write_csv(x = small_genes,file = "output/tables/!0_genes.csv")


```


# Section 7. HPA-Based Marker Context

This section adds external marker context from Human Protein Atlas-derived annotations. Treat this as annotation support, not as a primary statistical filter.

```{r}

library(HPAanalyze)

hpaListParam()

data <- hpaDownload(
  downloadList = c("normal_tissue", "pathology", "subcellular_location")
)

hpaVis(
  data,
  targetGene = top_genes$gene,
  targetCancer = "lung cancer"
)


subset_data <- hpaSubset(
  data,
  targetGene = top_genes$gene,
  targetCancer = "lung cancer"
)

View(subset_data$subcellular_location)

top_genes <- read_csv("output/tables/C1_seudotime_DEGs_UP_ordered.csv")

library(biomaRt)

mart <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")

ids <- getBM(
  attributes = c("hgnc_symbol", "ensembl_gene_id"),
  filters = "hgnc_symbol",
  values = top_genes$gene,
  mart = mart
)

ids

library(HPAanalyze)
library(purrr)
library(dplyr)

antibody_list <- map2(
  ids$ensembl_gene_id,
  ids$hgnc_symbol,
  function(ensg, symbol) {
    
    message("Downloading: ", symbol, " (", ensg, ")")
    
    tryCatch({
      x <- hpaXml(ensg)
      ab <- x$Antibody
      
      if (is.null(ab) || nrow(ab) == 0) return(NULL)
      
      ab %>%
        mutate(
          ensembl_gene_id = ensg,
          hgnc_symbol = symbol
        )
      
    }, error = function(e) {
      message("Failed: ", symbol)
      NULL
    })
  }
)

antibodies_df <- bind_rows(antibody_list)

final_table <- antibodies_df %>%
  left_join(
    subset_data$subcellular_location,
    by = c("ensembl_gene_id" = "ensembl")
  )

write_csv(x = final_table,file = "output/tables/IHC_antibody_table.csv")

```


# Section 8. ORA for Genes Increasing Along Pseudotime

This compact enrichment step translates increasing C1 genes into pathway-level hypotheses.

```{r}

library(gprofiler2) 

along_DEGs <- read_csv("output/tables/along_DEGs.csv")  |>
  filter(FDR < 0.05)


# Run GSEA using gprofiler2

along_gsea <- gost(
  query = along_DEGs$gene,
  organism = "hsapiens",       # 
  ordered_query = TRUE,        #
  multi_query = FALSE,
  significant = TRUE,
  sources = c("GO:BP","KEGG","TF","MIRNA") # databases to query
)

qs_save(object = along_gsea, file = "output/objects/gProfiler_along.qs2", compress_level = 5, nthreads = 10)


# View results

along_gsea$result %>% head(20)


# Plot top enriched terms

gostplot(along_gsea, capped = TRUE, interactive = FALSE) +
  theme_bw()

fwrite(along_gsea$result, file = "output/tables/gprofiler2_along_gsea.csv")


```

# Section 9. GSEA for GO Biological Processes

This section tests whether ranked C1-dynamic genes are enriched for GO biological processes.

```{r}

library(clusterProfiler)
library(org.Hs.eg.db)
library(dplyr)

# Keep unique genes and filter NA waldStat
ranked_genes <- along_DEGs %>%
  filter(!is.na(waldStat)) %>%
  distinct(gene, .keep_all = TRUE) %>%
  arrange(desc(logFClineage1))   # or waldStat

# Create numeric vector for GSEA
gene_list <- ranked_genes$logFClineage1   # numeric vector
names(gene_list) <- ranked_genes$gene  # assign gene symbols as names

# Sort decreasing
gene_list <- sort(gene_list, decreasing = TRUE)

# Check
head(gene_list)


# Run GSEA on GO BP
gsea_res_GO <- gseGO(
  geneList = gene_list,
  OrgDb = org.Hs.eg.db,
  ont = "BP",
  keyType = "SYMBOL",
  pvalueCutoff = 0.05,
  minGSSize = 10,
  maxGSSize = 500,
  verbose = TRUE
)


qs_save(object = gsea_res_GO, file = "output/objects/clusterProfiler_GO_tradeSEQ.qs2", compress_level = 5, nthreads = 15)

dotplot(gsea_res_GO, showCategory = 15) +
  theme_bw(base_size = 11) +
  labs(title = "GSEA of BP for genes along the Pseudotime")

ggsave(filename = "output/figures/clusterProfiler_GO_tradeSEQ.svg", width = 9, height = 7,dpi = 600)

fwrite(as.data.frame(gsea_res_GO), "output/tables/GSEA_along_C1.csv")

```

# Section 10. GSEA for KEGG Pathways

This section repeats the ranked enrichment logic using KEGG pathways for a pathway-level view of the C1 gradient.

```{r}

library(clusterProfiler)
library(org.Hs.eg.db)
library(dplyr)
library(data.table)


# Keep unique genes and filter NA waldStat
ranked_genes <- along_DEGs %>%
  filter(!is.na(waldStat)) %>%
  distinct(gene, .keep_all = TRUE) %>%
  arrange(desc(logFClineage1))

# Create numeric vector for GSEA
gene_list <- ranked_genes$logFClineage1
names(gene_list) <- ranked_genes$gene
gene_list <- sort(gene_list, decreasing = TRUE)

# Map gene symbols to Entrez IDs for KEGG
gene_df <- bitr(names(gene_list), fromType = "SYMBOL", 
                toType = "ENTREZID", OrgDb = org.Hs.eg.db)
gene_list_entrez <- gene_list[gene_df$SYMBOL]
names(gene_list_entrez) <- gene_df$ENTREZID

# Run GSEA for KEGG
gsea_res_KEGG <- gseKEGG(
  geneList = gene_list_entrez,
  organism = "hsa",
  pvalueCutoff = 0.05,
  minGSSize = 10,
  maxGSSize = 500,
  verbose = TRUE
  #keyType = "ncbi-geneid",
)

# Save object
qs_save(object = gsea_res_KEGG, 
        file = "output/objects/clusterProfiler_KEGG_tradeSEQ.qs2", 
        compress_level = 5, nthreads = 10)

# Plot
dotplot(gsea_res_KEGG, showCategory = 15) +
  theme_bw(base_size = 11) +
  labs(title = "GSEA KEGG Pathways for genes along Pseudotime")

ggsave(filename = "output/figures/clusterProfiler_KEGG_tradeSEQ.svg", width = 9, height = 7)

# Save results as CSV
fwrite(as.data.frame(gsea_res_KEGG), "output/tables/tradeSeq_GSEA_KEGG_results.csv")


library(ReactomePA)

gsea_res_Reactome <- gsePathway(
  geneList = gene_list_entrez,
  organism = "human",
  pvalueCutoff = 0.05,
  minGSSize = 10,
  maxGSSize = 500,
  verbose = TRUE
)

# Convert Entrez IDs to symbols
gsea_res_Reactome <- setReadable(gsea_res_Reactome, OrgDb = org.Hs.eg.db)

dotplot(gsea_res_Reactome, showCategory = 15)  +
  theme_bw(base_size = 11) +
  labs(title = "GSEA ReactomePA Pathways for genes along Pseudotime")

ggsave(filename = "output/figures/clusterProfiler_ReactomePA_tradeSEQ.svg", width = 10, height = 8)

# Save results as CSV
fwrite(as.data.frame(gsea_res_Reactome), "output/tables/tradeSeq_GSEA_ReactomePA_results.csv")


```

# Section 11. Draft Sensitivity: Self-Implemented GAMs Over Genes

This draft section is intentionally placed after the main overlap and enrichment results. It implements a minimal in-house GAM layer over genes, independent of `tradeSeq`, as a transparent sensitivity check rather than as a replacement for the main `tradeSeq` result.

## Section 11.1. R or Python?

For this project, **R is the better default** for self-implemented gene-wise GAMs because `mgcv` is mature, well-tested, supports penalized splines, negative-binomial families, REML fitting, offsets, fixed covariates, and clean model summaries. Python can do GAM-like models through `pyGAM`, `statsmodels`, or custom `sklearn` pipelines, but the single-cell RNA-seq use case is less standardized for negative-binomial gene-wise models with robust spline inference. Python remains useful for AnnData handling and plotting; R is stronger for this specific statistical layer.

This draft should be biologically motivated, not just a generic "fit a curve to every gene" exercise. The key question is whether **C1-EpiCore/C1 biology explains expression after accounting for library size, sample structure, and major cell-cycle effects**. Therefore the model uses raw RNA counts, tests Seurat-selected highly variable genes directly, keeps a 4,000-gene cap for a realistic sensitivity run, models `C1_score` as the primary smooth term, includes a library-size offset, adds sample and optional cell-cycle covariates, and exports interpretable high-versus-low C1 effects.

For progress reporting, use `progressr` with `furrr`. For parallel execution, the robust default remains `future::multisession`; if `future.mirai`/`mirai` is installed, the notebook can use the mirai-backed future plan for cleaner worker management. This keeps the code notebook-friendly while making the parallel layer easy to upgrade.

```{r eval=FALSE}
library(mgcv)
library(furrr)
library(broom)
library(progressr)
library(future)
library(parallelly)

set.seed(1234)

# Parallel and progress setup
# Keep workers conservative in a notebook. GAM fitting is CPU-heavy and each
# worker receives the count matrix and metadata.
n_workers <- max(1, min(4, parallelly::availableCores(omit = 1)))

# Use a mirai-backed future plan when available. This is optional and guarded:
# the same code still runs with plain multisession on machines without mirai.
mirai_plan <- NULL
if (
  requireNamespace("future.mirai", quietly = TRUE) &&
  exists("mirai_multisession", envir = asNamespace("future.mirai"))
) {
  mirai_plan <- get("mirai_multisession", envir = asNamespace("future.mirai"))
}

if (!is.null(mirai_plan)) {
  future::plan(mirai_plan, workers = n_workers)
  parallel_backend <- "future.mirai::mirai_multisession"
} else {
  future::plan(future::multisession, workers = n_workers)
  parallel_backend <- "future::multisession"
}

# Use a visible progress bar in interactive sessions.
# In Quarto/terminal rendering, progressr will fall back to a text handler.
progressr::handlers(global = TRUE)
progressr::handlers("progress")

message("Self-GAM draft will use ", n_workers, " parallel workers via ", parallel_backend, ".")

DefaultAssay(css_combined) <- "RNA"
counts_mat <- GetAssayData(css_combined, assay = "RNA", slot = "counts")

# Gene set: Seurat HVGs, capped at 4,000
# This avoids testing arbitrary expressed genes. The sensitivity layer should
# focus on the variable transcriptome that Seurat already considers informative.
if (length(VariableFeatures(css_combined)) < 4000) {
  css_combined <- FindVariableFeatures(
    css_combined,
    assay = "RNA",
    selection.method = "vst",
    nfeatures = 4000,
    verbose = FALSE
  )
}

seurat_hvgs <- VariableFeatures(css_combined)

genes_to_test <- seurat_hvgs[seurat_hvgs %in% rownames(counts_mat)]

# Keep genes with enough detected cells for a negative-binomial smooth fit.
genes_to_test <- genes_to_test[Matrix::rowSums(counts_mat[genes_to_test, , drop = FALSE] > 0) >= 50]

# A 4K cap is appropriate here: broad enough to cover the variable transcriptome,
# small enough to keep an exploratory draft tractable on a laptop/workstation.
genes_to_test <- head(genes_to_test, 4000)

message("Genes selected for self-GAM draft from Seurat HVGs: ", length(genes_to_test))

# Metadata and biologically motivated covariates
meta_cols <- c(
  "C1_score",
  "nCount_RNA",
  "sample",
  "S.Score",
  "G2M.Score",
  "percent.mt"
)

meta_cols <- intersect(meta_cols, colnames(css_combined@meta.data))

gam_meta <- css_combined@meta.data |>
  dplyr::select(dplyr::all_of(meta_cols))

if (!"sample" %in% colnames(gam_meta)) {
  gam_meta$sample <- "all"
}

gam_meta <- gam_meta |>
  dplyr::mutate(
    C1_score = as.numeric(C1_score),
    log_library = log(pmax(nCount_RNA, 1)),
    sample = factor(sample)
  )

# Optional covariates are included only if present and informative. Sample is
# treated as a fixed effect because there are only four samples in this dataset.
linear_terms <- c()

if ("sample" %in% colnames(gam_meta) && nlevels(gam_meta$sample) > 1) {
  linear_terms <- c(linear_terms, "sample")
}

if ("S.Score" %in% colnames(gam_meta)) {
  linear_terms <- c(linear_terms, "S.Score")
}

if ("G2M.Score" %in% colnames(gam_meta)) {
  linear_terms <- c(linear_terms, "G2M.Score")
}

if ("percent.mt" %in% colnames(gam_meta)) {
  linear_terms <- c(linear_terms, "percent.mt")
}

rhs_terms <- c("s(C1_score, k = 6)", linear_terms, "offset(log_library)")
gam_formula <- stats::as.formula(paste("y ~", paste(rhs_terms, collapse = " + ")))

message("Self-GAM formula: ", deparse(gam_formula))

fit_one_gene_gam <- function(gene, progressor = NULL) {
  if (!is.null(progressor)) {
    progressor(sprintf("Fitting %s", gene))
  }

  y <- as.numeric(counts_mat[gene, ])

  df <- gam_meta |>
    dplyr::mutate(y = y) |>
    tidyr::drop_na(C1_score, log_library, y)

  if (sum(df$y > 0) < 50) {
    return(tibble(gene = gene, status = "too_sparse"))
  }

  # bam() is preferred here over gam() because this is a gene-wise screen.
  # nthreads = 1 prevents nested parallelism inside each worker.
  fit <- tryCatch(
    mgcv::bam(
      gam_formula,
      data = df,
      family = nb(),
      method = "fREML",
      discrete = TRUE,
      nthreads = 1
    ),
    error = function(e) NULL
  )

  if (is.null(fit)) {
    return(tibble(gene = gene, status = "fit_failed"))
  }

  low_df <- df
  low_df$C1_score <- stats::quantile(df$C1_score, 0.1, na.rm = TRUE)

  high_df <- df
  high_df$C1_score <- stats::quantile(df$C1_score, 0.9, na.rm = TRUE)

  pred_low <- mean(predict(fit, newdata = low_df, type = "response"), na.rm = TRUE)
  pred_high <- mean(predict(fit, newdata = high_df, type = "response"), na.rm = TRUE)

  smooth_table <- summary(fit)$s.table
  c1_row <- grep("^s\\(C1_score\\)", rownames(smooth_table), value = TRUE)

  if (length(c1_row) != 1) {
    return(tibble(gene = gene, status = "no_C1_smooth"))
  }

  tibble(
    gene = gene,
    status = "ok",
    mean_count = mean(df$y),
    pct_detected = mean(df$y > 0),
    smooth_p_C1 = smooth_table[c1_row, "p-value"],
    edf_C1 = smooth_table[c1_row, "edf"],
    deviance_explained = summary(fit)$dev.expl,
    pred_low_C1 = pred_low,
    pred_high_C1 = pred_high,
    logFC_high_vs_low = log2((pred_high + 1e-6) / (pred_low + 1e-6)),
    direction_at_high_C1 = dplyr::case_when(
      logFC_high_vs_low > 0 ~ "higher_at_high_C1",
      logFC_high_vs_low < 0 ~ "lower_at_high_C1",
      TRUE ~ "flat"
    )
  )
}

# Parallel model fitting with progress
# progressr::with_progress() gives a single progress bar for all parallel jobs.
# furrr_options(seed = TRUE) keeps any stochastic parts reproducible.
self_gam_started <- Sys.time()

self_gam_results <- progressr::with_progress({
  p <- progressr::progressor(along = genes_to_test)

  furrr::future_map_dfr(
    genes_to_test,
    ~ fit_one_gene_gam(.x, progressor = p),
    .options = furrr::furrr_options(seed = TRUE),
    .progress = FALSE
  )
}) |>
  dplyr::mutate(
    FDR_C1 = p.adjust(smooth_p_C1, method = "BH"),
    Direction = dplyr::case_when(
      status != "ok" ~ NA_character_,
      logFC_high_vs_low > 0 ~ "C1_positive",
      logFC_high_vs_low < 0 ~ "C1_negative",
      TRUE ~ "flat"
    )
  )

self_gam_runtime <- difftime(Sys.time(), self_gam_started, units = "mins")
message("Self-GAM draft completed in ", round(as.numeric(self_gam_runtime), 2), " minutes.")

# Compact run summary for notebook readability.
self_gam_summary <- self_gam_results |>
  dplyr::count(status, name = "n_genes") |>
  dplyr::arrange(desc(n_genes))

self_gam_summary

readr::write_csv(
  self_gam_results,
  "output/tables/MAP_self_GAM_HVG4000_along_C1_mgcv_draft.csv"
)

readr::write_csv(
  self_gam_summary,
  "output/tables/MAP_self_GAM_HVG4000_along_C1_mgcv_draft_summary.csv"
)

# Reset the future plan after the draft run so later notebook chunks do not
# unexpectedly inherit parallel workers.
future::plan(future::sequential)
```
