---
title: "MAP 06 - hdWGCNA Co-expression Modules"
format: html
---

# Section 1. Overview

This notebook builds a systems-level co-expression view of the CSS epithelial atlas using hdWGCNA.

**MAP role:** reduced module notebook. Keep module construction only if the modules support C1 programs, hub genes, or scATAC hypotheses. Avoid treating every module plot as a required deliverable.

## Section 1.1. Reader Guide

- Inputs are the processed Seurat object and/or the saved hdWGCNA object in `output/objects/hdWGCNA_object.qs2`.
- The notebook builds metacells, selects a soft threshold, constructs a co-expression network, identifies modules, ranks hub genes, and compares modules with C1-related traits.
- Main outputs are module plots, hub-gene tables, module eigengene tables, differential module results, and enrichment dot plots.
- Read modules as coordinated gene programs, not as single marker genes.

## Section 1.2. MAP Section Boundaries

### Section 1.2.1. Notebook Flow

1. Setup and load hdWGCNA-ready object.
2. Build metacells and choose soft threshold.
3. Construct modules and eigengenes.
4. Export hub genes and module eigengenes.
5. Compare modules with C1 and key metadata.
6. Keep enrichment plots compact and focused on C1-relevant modules.

## Section 1.3. Setup

This section loads the required packages and restores the Seurat object used as the starting point for downstream analyses.

```{r include=FALSE}

library(WGCNA)
library(hdWGCNA)

packages <- c(
  "tidyverse",
  "Seurat",
  "harmony",
  "patchwork",
  "data.table",
  "fst",
  "UCell",
  "viridis",
  "RColorBrewer",
  "qs2",
  "ggsci",
  "scCustomize", 
  "future",
  "ggpubr"
)

purrr::walk(packages, library, character.only = TRUE)

plan(multicore, workers = 4)

options(future.globals.maxSize = 20 * 1024^3)  # 6 GB

umap_theme <- cowplot::theme_cowplot() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 11, color = "grey30"),
    axis.title = element_text(size = 13),
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10)
  )

#css_combined <- qs_read(file = "output/objects/css_combined_processed.qs2",nthreads = 10)

seurat_obj <- qs_read(file = "output/objects/hdWGCNA_object.qs2", nthreads = 10)

```


# Section 2. Build the hdWGCNA Network

The workflow below follows the standard hdWGCNA order: define identities, build metacells, choose a soft-threshold, construct the network, and then interpret modules biologically.
# Section 3. Module Construction, Hub Genes, and Module UMAPs
```{r}

# 1. Define identity class (important for grouping)
DefaultAssay(css_combined) <- "SCT"

Idents(css_combined) <- css_combined$seurat_clusters


# 2. Initialize hdWGCNA object
#    IMPORTANT: explicitly pass features to avoid genes_use = 0

seurat_obj <- SetupForWGCNA(
  css_combined,
  features = VariableFeatures(css_combined),  # <- critical fix
  wgcna_name = "seurat_clusters"
)


# 3. Construct metacells

seurat_obj <- MetacellsByGroups(
  seurat_obj = seurat_obj,
  group.by = "seurat_clusters",
  reduction = "harmony",
  k = 25,
  max_shared = 10,
  ident.group = "seurat_clusters"
)



# 4. Normalize metacell expression matrix
seurat_obj <- NormalizeMetacells(seurat_obj)

# 5. Define expression matrix for WGCNA
#    (this is where cells + assay + layer must match correctly)

seurat_obj <- SetDatExpr(
  seurat_obj,
  group_name = c("1","2","3","4","5","6","7","8"),
  group.by = "seurat_clusters",
  assay = "SCT",
  layer = "data"
)

# 6. Soft-threshold selection

seurat_obj <- TestSoftPowers(
  seurat_obj,
  networkType = "signed hybrid" # signed hybrid
)

plot_list <- PlotSoftPowers(seurat_obj)

hybrid_plot <- wrap_plots(plot_list, ncol = 2) &
  theme_bw(base_size = 12)

hybrid_plot

ggsave(
  "output/figures/PlotSoftPowers_hdWGCNA.png",
  hybrid_plot,
  width = 12,
  height = 10,
  dpi = 600
)

power_table <- GetPowerTable(seurat_obj)
head(power_table)



# 7. Construct co-expression network

seurat_obj <- ConstructNetwork(
  seurat_obj,
  randomSeed = 1234,
  soft_power = 4,
  overwrite_tom = TRUE,
  tom_name = "seurat_clusters"
)

PlotDendrogram(seurat_obj, main = "seurat_clusters hdWGCNA Dendrogram")


library(harmony)
# 9. Compute module eigengenes

seurat_obj <- ModuleEigengenes(
  seurat_obj,
  assay = "SCT",
  group.by.vars = "seurat_clusters"
)

hMEs <- GetMEs(seurat_obj)  # harmonized MEs
head(hMEs, 25)

MEs  <- GetMEs(seurat_obj, harmonized = FALSE)
head(MEs, 25)

# 10. Compute module connectivity (kME)

seurat_obj <- ModuleConnectivity(
  seurat_obj,
  group.by = "seurat_clusters",
  group_name = c("2","4","1","3","8","5","6","7"),
)

# rename the modules
seurat_obj <- ResetModuleNames(
  seurat_obj,
  new_name = "Plasticity-M"
)

# 11. Plot kME-ranked genes per module

p <- PlotKMEs(seurat_obj, ncol = 3) 
p

ggsave(
  "output/figures/PlotKMEs_hdWGCNA.png",
  p,
  width = 12,
  height = 8,
  dpi = 600
)


# get the module assignment table:
modules <- GetModules(seurat_obj) %>% subset(module != 'grey')

# show the first 6 columns:
head(modules[,1:6],25)

# get hub genes
hub_df <- GetHubGenes(seurat_obj, n_hubs = 100) |>
  as_tibble()

print(hub_df, n = 50)

write_csv(x = hub_df,file = "output/tables/hdWGCNA_hub_genes.csv")

hdWGCNA_hub_genes <- vroom::vroom("output/tables/hdWGCNA_hub_genes.csv")
hdWGCNA_hub_genes

p1 <- ggplot(data = hdWGCNA_hub_genes, 
             aes(x = module, y = kME, fill = module)) +
  geom_violin(
    trim = TRUE,
    alpha = 0.7,
    color = NA
  ) +
  geom_boxplot(
    width = 0.1,
    outlier.shape = NA,
    alpha = 0.5,
    fill = "white"
  ) +
  ggpubr::stat_compare_means() +
  theme_bw(base_size = 12) +
  labs(
    x = "module",
    y = "kME",
    title = "hdWGCNA hub genes kME",
    subtitle = "Violin and boxplot comparison by module"
  ) + 
  scale_fill_brewer(palette = "Set3") +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    legend.position = "none"
  )

p1


# compute gene scoring for the top 25 hub genes by kME for each module
# with UCell method
library(UCell)
seurat_obj <- ModuleExprScore(
  seurat_obj,
  n_genes = 30,
  method='UCell'
)

# make a featureplot of hMEs for each module
plot_list <- ModuleFeaturePlot(
  seurat_obj,
  features='hMEs', # plot the hMEs
  order=TRUE # order so the points with highest hMEs are on top
)

# stitch together with patchwork
wrap_plots(plot_list, ncol=3)

ggsave(
  "output/figures/ModuleFeaturePlot_hdWGCNA.png",
  wrap_plots(plot_list, ncol=3),
  width = 12,
  height = 8,
  dpi = 600
)

# make a featureplot of hub scores for each module
plot_list <- ModuleFeaturePlot(
  seurat_obj,
  features='scores', # plot the hub gene scores
  order='shuffle', # order so cells are shuffled
  ucell = TRUE # depending on Seurat vs UCell for gene scoring
)

# stitch together with patchwork
wrap_plots(plot_list, ncol=3)

ggsave(
  "output/figures/ModuleFeaturePlot_scores_hdWGCNA.png",
  wrap_plots(plot_list, ncol=3),
  width = 12,
  height = 8,
  dpi = 600
)


#qs_save(object = seurat_obj, file = "output/objects/hdWGCNA_object.qs2", nthreads = 10,compress_level = 9)
#seurat_obj <- qs_read(file = "output/objects/hdWGCNA_object.qs2", nthreads = 10)
Radar_p <- ModuleRadarPlot(
  seurat_obj,
  group.by = 'seurat_clusters',
  barcodes = seurat_obj@meta.data %>% subset(seurat_clusters != "Uncertain") %>% rownames(),
  axis.label.size=4,
  grid.label.size=4
)
Radar_p

ggsave(
  "output/figures/ModuleRadarPlot_hdWGCNA.png",
  Radar_p,
  width = 10,
  height = 8,
  dpi = 600
)

options(future.globals.maxSize = 12 * 1024^3)  # 12 GB
# hubgene network
HubGeneNetworkPlot(
  seurat_obj,
  n_hubs = 5, n_other=8,
  edge_prop = 0.75,
  mods = 'all'
)

# get the list of modules:
modules <- GetModules(seurat_obj)
mods <- levels(modules$module); mods <- mods[mods != 'grey']

# hubgene network
HubGeneNetworkPlot(
  seurat_obj,
  n_hubs = 10, n_other=20,
  edge_prop = 0.75
)

seurat_obj <- RunModuleUMAP(
  seurat_obj,
  n_hubs = 20,
  n_neighbors=15,
  min_dist=0.1,
  supervised=TRUE,
  target_weight=0.5
)
# get the hub gene UMAP table from the seurat object
umap_df <- GetModuleUMAP(seurat_obj)

umap_df <- GetModuleUMAP(seurat_obj) %>%
  group_by(module) %>%
  mutate(is_hub = rank(-kME) <= 5) %>%
  ungroup()

p <- ggplot(umap_df, aes(UMAP1, UMAP2)) +
  geom_point(aes(color = module), size = 1, alpha = 0.4) +
  geom_point(
    data = subset(umap_df, is_hub),
    aes(fill = module),
    shape = 21,
    color = "black",
    size = 2.5,
    stroke = 0.4
  ) +
  geom_text_repel(
    data = subset(umap_df, is_hub),
    aes(label = gene),
    size = 4,
    box.padding = 0.3,
    point.padding = 0.2,
    segment.size = 0.3,
    max.overlaps = Inf
  ) +
  labs(title = "Module UMAP (Hub Genes Highlighted)") +
  #theme_ggprism_mod(base_size = 12) +
  umap_theme +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.line = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )

p

ggsave(
  "output/figures/module_umap_compact.png",
  plot = p,
  width = 10,
  height = 6,
  dpi = 600
)

ModuleUMAPPlot(
  seurat_obj,
  edge.alpha=0.25,
  sample_edges=TRUE,
  edge_prop=0.2, # proportion of edges to sample (20% here)
  label_hubs=4 ,# how many hub genes to plot per module?
  keep_grey_edges=FALSE)

group1 <- seurat_obj@meta.data %>% subset(seurat_clusters == '7') %>% rownames
group2 <- seurat_obj@meta.data %>% subset(seurat_clusters == '2') %>% rownames

DMEs_scores <- FindDMEs(
  seurat_obj,
  features = 'ModuleScores',
  barcodes1 = group1,
  barcodes2 = group2,
  #test.use='MAST',
  wgcna_name='seurat_clusters'
)

DMEs <- FindDMEs(
  seurat_obj,
  barcodes1 = group1,
  barcodes2 = group2,
  test.use='MAST',
  wgcna_name='seurat_clusters'
)


head(DMEs)


PlotDMEsVolcano(
  seurat_obj,
  DMEs,
  wgcna_name = 'seurat_clusters'
) + theme_bw(base_size = 12)

```

# Section 4. Differential Module Expression
```{r}


# 16. Differential module expression

## define groups
group1 <- seurat_obj@meta.data %>%
  subset(seurat_clusters == '7') %>%
  rownames()

group2 <- seurat_obj@meta.data %>%
  subset(seurat_clusters == '2') %>%
  rownames()

## DME analysis comparing two groups
DMEs <- FindDMEs(
  seurat_obj,
  barcodes1 = group1,
  barcodes2 = group2,
  test.use = 'MAST',
  wgcna_name = 'seurat_clusters'
)

## Using ModuleScores instead of MEs
DMEs_scores <- FindDMEs(
  seurat_obj,
  features = 'ModuleScores',
  barcodes1 = group1,
  barcodes2 = group2,
  test.use = 'MAST',
  wgcna_name = 'seurat_clusters'
)

head(DMEs)

PlotDMEsVolcano(
  seurat_obj,
  DMEs,
  wgcna_name = 'seurat_clusters'
)

## One-versus-all DME analysis
group.by = 'seurat_clusters'

DMEs_all <- FindAllDMEs(
  seurat_obj,
  #test.use = "MAST",
  group.by = 'seurat_clusters',
  wgcna_name = 'seurat_clusters'
)

head(DMEs_all)


p <- PlotDMEsVolcano(
  seurat_obj,
  DMEs_all,
  wgcna_name = 'seurat_clusters',
  plot_labels=T,
  show_cutoff=T
)

# facet wrap by each cell type
p <-  p +
  facet_wrap(~group, ncol = 5) +
  labs(
    title = "Differential Module Expression Across Seurat Clusters",
    subtitle = "Module score contrasts computed using FindAllDMEs"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 11),
    legend.position = "none"
  )

p

ggsave(
  "output/figures/PlotDMEsVolcano.png",
  plot = p,
  width = 18,
  height = 8,
  dpi = 600
)

```

# Section 5. Module Enrichment and Trait Association
```{r}

set.seed(1234)
# define the enrichr databases to test
dbs <- c('GO_Biological_Process_2025',"JASPAR_PWM_Human_2025","KEGG_2026","WikiPathways_2024_Human","TargetScan_microRNA_2017","RummaGEO_GenePerturbations_2025","Reactome_Pathways_2024","MSigDB_Oncogenic_Signatures","ClinVar_2025","Cancer_Cell_Line_Encyclopedia")

# perform enrichment tests
seurat_obj <- RunEnrichr(
  seurat_obj,
  dbs=dbs,
  max_genes = 500 # use max_genes = Inf to choose all genes
)

# retrieve the output table
enrich_df <- GetEnrichrTable(seurat_obj)

# look at the results
head(enrich_df)


```

```{r}


# enrichr dotplot
EnrichrDotPlot(
  seurat_obj,
  mods = "all", # use all modules (default)
  database = "Cancer_Cell_Line_Encyclopedia", # this must match one of the dbs used previously
  n_terms=10, # number of terms per module
  term_size=8, # font size for the terms
  p_adj = TRUE # show the p-val or adjusted p-val?
)  + scale_color_stepsn(colors=rev(viridis::magma(256))) + theme_bw(base_size = 12)

ggsave(
  "output/figures/EnrichrDotPlot_Cancer_Cell_Line_Encyclopedia.png",
  width = 15,
  height = 12,
  dpi = 600
)

# enrichr dotplot
EnrichrDotPlot(
  seurat_obj,
  mods = "all", # use all modules (default)
  database = "ClinVar_2025", # this must match one of the dbs used previously
  n_terms=10, # number of terms per module
  term_size=8, # font size for the terms
  p_adj = TRUE # show the p-val or adjusted p-val?
)  + scale_color_stepsn(colors=rev(viridis::magma(256))) + theme_bw(base_size = 12)

ggsave(
  "output/figures/EnrichrDotPlot_ClinVar_2025.png",
  width = 15,
  height = 12,
  dpi = 600
)

# enrichr dotplot
EnrichrDotPlot(
  seurat_obj,
  mods = "all", # use all modules (default)
  database = "MSigDB_Oncogenic_Signatures", # this must match one of the dbs used previously
  n_terms=10, # number of terms per module
  term_size=8, # font size for the terms
  p_adj = TRUE # show the p-val or adjusted p-val?
)  + scale_color_stepsn(colors=rev(viridis::magma(256))) + theme_bw(base_size = 12)

ggsave(
  "output/figures/EnrichrDotPlot_MSigDB_Oncogenic_Signatures.png",
  width = 15,
  height = 12,
  dpi = 600
)


# enrichr dotplot
EnrichrDotPlot(
  seurat_obj,
  mods = "all", # use all modules (default)
  database = "TargetScan_microRNA_2017", # this must match one of the dbs used previously
  n_terms=10, # number of terms per module
  term_size=8, # font size for the terms
  p_adj = TRUE # show the p-val or adjusted p-val?
)  + scale_color_stepsn(colors=rev(viridis::magma(256))) + theme_bw(base_size = 12)

ggsave(
  "output/figures/EnrichrDotPlot_TargetScan_microRNA_2017.png",
  width = 15,
  height = 12,
  dpi = 600
)

# enrichr dotplot
EnrichrDotPlot(
  seurat_obj,
  mods = "all", # use all modules (default)
  database = "JASPAR_PWM_Human_2025", # this must match one of the dbs used previously
  n_terms=10, # number of terms per module
  term_size=8, # font size for the terms
  p_adj = TRUE # show the p-val or adjusted p-val?
)  + scale_color_stepsn(colors=rev(viridis::magma(256))) + theme_bw(base_size = 12)

ggsave(
  "output/figures/EnrichrDotPlot_JASPAR_PWM_Human_2025.png",
  width = 15,
  height = 12,
  dpi = 600
)

# enrichr dotplot
EnrichrDotPlot(
  seurat_obj,
  mods = "all", # use all modules (default)
  database = "Reactome_Pathways_2024", # this must match one of the dbs used previously
  n_terms=10, # number of terms per module
  term_size=8, # font size for the terms
  p_adj = TRUE # show the p-val or adjusted p-val?
)  + scale_color_stepsn(colors=rev(viridis::magma(256))) + theme_bw(base_size = 12)

ggsave(
  "output/figures/EnrichrDotPlot_Reactome_Pathways_2024.png",
  width = 15,
  height = 12,
  dpi = 600
)

# enrichr dotplot
EnrichrDotPlot(
  seurat_obj,
  mods = "all", # use all modules (default)
  database = "RummaGEO_GenePerturbations_2025", # this must match one of the dbs used previously
  n_terms=10, # number of terms per module
  term_size=8, # font size for the terms
  p_adj = TRUE # show the p-val or adjusted p-val?
)  + scale_color_stepsn(colors=rev(viridis::magma(256))) + theme_bw(base_size = 12)

ggsave(
  "output/figures/EnrichrDotPlot_RummaGEO_GenePerturbations_2025.png",
  width = 15,
  height = 12,
  dpi = 600
)

# enrichr dotplot
EnrichrDotPlot(
  seurat_obj,
  mods = "all", # use all modules (default)
  database = "KEGG_2026", # this must match one of the dbs used previously
  n_terms=10, # number of terms per module
  term_size=8, # font size for the terms
  p_adj = TRUE # show the p-val or adjusted p-val?
)  + scale_color_stepsn(colors=rev(viridis::magma(256))) + theme_bw(base_size = 12)

ggsave(
  "output/figures/EnrichrDotPlot_KEGG_2026.png",
  width = 15,
  height = 12,
  dpi = 600
)

# enrichr dotplot
EnrichrDotPlot(
  seurat_obj,
  mods = "all", # use all modules (default)
  database = "GO_Biological_Process_2025", # this must match one of the dbs used previously
  n_terms=5, # number of terms per module
  term_size=8, # font size for the terms
  p_adj = TRUE # show the p-val or adjusted p-val?
)  + scale_color_stepsn(colors=rev(viridis::magma(256))) + theme_bw(base_size = 12)

ggsave(
  "output/figures/EnrichrDotPlot_GO_Biological_Process_2025.png",
  width = 15,
  height = 12,
  dpi = 600
)

# enrichr dotplot
EnrichrDotPlot(
  seurat_obj,
  mods = "all", # use all modules (default)
  database = "WikiPathways_2024_Human", # this must match one of the dbs used previously
  n_terms=5, # number of terms per module
  term_size=8, # font size for the terms
  p_adj = TRUE # show the p-val or adjusted p-val?
)  + scale_color_stepsn(colors=rev(viridis::magma(256))) + theme_bw(base_size = 12)

ggsave(
  "output/figures/EnrichrDotPlot_WikiPathways_2024_Human.png",
  width = 15,
  height = 12,
  dpi = 600
)

```

```{r}

hMEs <- GetMEs(seurat_obj)  # harmonized MEs
head(hMEs, 25)

hMEs_clean <- hMEs |>
  dplyr::select(-grey) |>
  rownames_to_column("cells")

write_csv(x = hMEs_clean,file = "output/tables/hMEs_clean.csv")

traits <- seurat_obj@meta.data %>%
  dplyr::select(C1_score)

cor_mat <- WGCNA::cor(hMEs, traits, method = "spearman")
p_mat <- WGCNA::corPvalueStudent(cor_mat, nSamples = nrow(hMEs))

cor_mat
p_mat

module_trait_results <- data.frame(
  module = rownames(cor_mat),
  correlation_C1_score = cor_mat[, 1],
  p_value = p_mat[, 1]
) %>%
  dplyr::mutate(
    padj = p.adjust(p_value, method = "BH")
  ) %>%
  dplyr::arrange(padj) |>
  filter(module != "grey")

write_csv(x = module_trait_results,file = "output/tables/CORR_hMEs_C1.csv")

plot_wgcna_module_C1 <- module_trait_results %>%
  dplyr::mutate(
    module = factor(module, levels = module[order(correlation_C1_score)])
  ) %>%
  ggplot2::ggplot(
    ggplot2::aes(
      x = correlation_C1_score,
      y = module,
      fill = correlation_C1_score
    )
  ) +
  ggplot2::geom_col(width = 0.7) +
  ggplot2::geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  scale_fill_viridis_c(option = "D") +  # 👈 FIXED
  ggplot2::labs(
    title = "WGCNA Module Association With C1 Score",
    subtitle = "Spearman correlation between module eigengenes and C1 score",
    x = "Correlation with C1 score",
    y = "WGCNA module",
    fill = "Spearman rho"
  ) +
  ggplot2::theme_classic(base_size = 13) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold"),
    axis.text = ggplot2::element_text(color = "black")
  )

plot_wgcna_module_C1

ggsave(
  "output/figures/plot_wgcna_module_C1.png",
  plot = plot_wgcna_module_C1,
  width = 10,
  height = 6,
  dpi = 600
)
 
```
