---
title: "Biomarker model reporting"
format:
  html:
    toc: true
    code-fold: true
execute:
  freeze: false
---

> Status: `draft`
>
> Template class: SOURCE-BACKED WORKFLOW

## Purpose

Read frozen model outputs and produce compact audit/presentation tables, a
probability plot, and a feature-expression heatmap. This notebook does not
refit a model, select features, tune thresholds, or re-evaluate performance.

::: {.callout-note}
## Source-backed reporting workflow

The reporting layout preserves the useful `gt`/`gtExtras` companion workflow
from the source analysis while replacing fixed clinical and biomarker names
with dynamic columns from the model-output contract.
:::

## Inputs and parameters

The complete and essential TSVs are written by
`binary_classification_brglm2.qmd`. The essential table contains
the sample, split, truth, prediction, probability, threshold, and selected
feature columns. The complete table may additionally contain metadata.

```{r}
# Input contract
complete_input <- "input/biomarker_train_test_complete.tsv"
essential_input <- "input/biomarker_train_test_essential.tsv"
output_table_dir <- "output/tables"
output_figure_dir <- "output/figures"

sample_id_column <- "sample_id"
split_column <- "split"
truth_column <- "truth"
predicted_column <- "predicted_class"
probability_column <- "probability_positive"
threshold_column <- "threshold"
feature_prefix <- "feature_"

dir.create(output_table_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(output_figure_dir, recursive = TRUE, showWarnings = FALSE)
```

## Frozen-output checks

```{r}
library(tidyverse)
library(gt)
library(gtExtras)
library(ComplexHeatmap)
library(circlize)

complete_tbl <- readr::read_tsv(complete_input, show_col_types = FALSE)
essential_tbl <- readr::read_tsv(essential_input, show_col_types = FALSE)

required_columns <- c(
  sample_id_column,
  split_column,
  truth_column,
  predicted_column,
  probability_column,
  threshold_column
)
stopifnot(
  all(required_columns %in% names(essential_tbl)),
  !anyDuplicated(essential_tbl[[sample_id_column]]),
  all(essential_tbl[[sample_id_column]] %in% complete_tbl[[sample_id_column]])
)

feature_columns <- names(essential_tbl)[startsWith(names(essential_tbl), feature_prefix)]
stopifnot(
  length(feature_columns) > 0L,
  all(feature_columns %in% names(complete_tbl))
)

model_columns <- c(
  split_column,
  truth_column,
  predicted_column,
  probability_column,
  threshold_column
)

label_map <- c(
  setNames("Sample", sample_id_column),
  setNames("Set", split_column),
  setNames("True class", truth_column),
  setNames("Predicted class", predicted_column),
  setNames("Positive probability", probability_column),
  setNames("Threshold", threshold_column)
)
feature_labels <- stats::setNames(
  sub(paste0("^", feature_prefix), "", feature_columns),
  feature_columns
)
```

## Complete audit table

The complete table keeps all available metadata and selected-feature values.
It is an HTML presentation artifact; the frozen numerical source remains the
TSV produced by the modeling notebook.

```{r}
complete_gt <- complete_tbl |>
  gt::gt(id = "complete_biomarker_model_summary") |>
  gtExtras::gt_theme_538() |>
  gt::tab_header(
    title = gt::md("**Complete biomarker model summary**"),
    subtitle = gt::md(
      "Frozen predictions, observed classes, thresholds, metadata, and selected features"
    )
  ) |>
  gt::cols_label(!!!as.list(c(label_map, feature_labels))) |>
  gt::tab_spanner(
    label = "Model output",
    columns = dplyr::all_of(model_columns)
  ) |>
  gt::tab_spanner(
    label = "Selected features",
    columns = dplyr::all_of(feature_columns)
  ) |>
  gt::fmt_number(
    columns = dplyr::all_of(c(probability_column, threshold_column)),
    decimals = 3
  ) |>
  gt::data_color(
    columns = dplyr::all_of(probability_column),
    fn = scales::col_numeric(c("#E8F5E9", "#FFCDD2"), domain = c(0, 1))
  )

complete_gt
gt::gtsave(
  complete_gt,
  file.path(output_table_dir, "biomarker_complete_model_summary.html")
)
```

## Essential presentation table

The essential table is intentionally narrower but retains the split and
selected-feature columns needed to interpret the displayed predictions.

```{r}
essential_gt <- essential_tbl |>
  gt::gt(id = "essential_biomarker_model_summary") |>
  gtExtras::gt_theme_538() |>
  gt::tab_header(
    title = gt::md("**Essential biomarker model summary**"),
    subtitle = gt::md("Train/test annotations and frozen model predictions")
  ) |>
  gt::cols_label(!!!as.list(c(label_map, feature_labels))) |>
  gt::tab_spanner(
    label = "Model output",
    columns = dplyr::all_of(model_columns)
  ) |>
  gt::tab_spanner(
    label = "Selected features",
    columns = dplyr::all_of(feature_columns)
  ) |>
  gt::fmt_number(
    columns = dplyr::all_of(c(probability_column, threshold_column)),
    decimals = 3
  ) |>
  gt::data_color(
    columns = dplyr::all_of(probability_column),
    fn = scales::col_numeric(c("#E8F5E9", "#FFCDD2"), domain = c(0, 1))
  ) |>
  gtExtras::gt_plt_bar(
    column = probability_column,
    keep_column = TRUE,
    scale_type = "number",
    width = 60
  )

essential_gt
gt::gtsave(
  essential_gt,
  file.path(output_table_dir, "biomarker_essential_model_summary.html")
)
```

## Model probability plot

```{r}
probability_plot <- complete_tbl |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = .data[[probability_column]],
      fill = .data[[split_column]],
      colour = .data[[truth_column]]
    )
  ) +
  ggplot2::geom_density(alpha = 0.25) +
  ggplot2::geom_vline(
    ggplot2::aes(xintercept = .data[[threshold_column]]),
    linetype = "dashed",
    colour = "grey30"
  ) +
  ggplot2::scale_x_continuous(limits = c(0, 1)) +
  ggplot2::labs(
    title = "Frozen model probabilities",
    x = "Positive-class probability",
    y = "Density",
    fill = "Set",
    colour = "True class"
  ) +
  ggplot2::theme_bw()

probability_plot
ggplot2::ggsave(
  file.path(output_figure_dir, "model_probability_plot.png"),
  probability_plot,
  width = 8,
  height = 5,
  dpi = 300
)
```

## Selected-feature heatmap

The heatmap standardizes each selected feature across the rows present in the
frozen essential table. Zero-variance features are excluded because their
standard deviation is undefined. This is a visualization transform, not a
modeling or re-normalization step.

```{r}
feature_matrix <- essential_tbl |>
  dplyr::select(dplyr::all_of(feature_columns)) |>
  as.matrix()
rownames(feature_matrix) <- essential_tbl[[sample_id_column]]

feature_sd <- apply(feature_matrix, 2, stats::sd, na.rm = TRUE)
keep_features <- is.finite(feature_sd) & feature_sd > 0
stopifnot(any(keep_features))

feature_z <- scale(feature_matrix[, keep_features, drop = FALSE])
feature_z <- t(feature_z)

split_values <- as.character(essential_tbl[[split_column]])
truth_values <- as.character(essential_tbl[[truth_column]])
split_levels <- unique(split_values)
truth_levels <- unique(truth_values)
split_colours <- stats::setNames(scales::hue_pal()(length(split_levels)), split_levels)
truth_colours <- stats::setNames(scales::hue_pal()(length(truth_levels)), truth_levels)

annotation <- ComplexHeatmap::HeatmapAnnotation(
  Set = split_values,
  Truth = truth_values,
  col = list(Set = split_colours, Truth = truth_colours),
  annotation_name_side = "left"
)

heatmap <- ComplexHeatmap::Heatmap(
  feature_z,
  name = "Feature z-score",
  top_annotation = annotation,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  show_column_names = FALSE,
  column_title = "Samples",
  row_title = "Selected features",
  col = circlize::colorRamp2(c(-2, 0, 2), c("#2166AC", "white", "#B2182B"))
)

heatmap
png(file.path(output_figure_dir, "selected_feature_heatmap.png"), width = 1800, height = 1100, res = 200)
ComplexHeatmap::draw(heatmap)
dev.off()
```

## Method notes

- The companion notebook consumes frozen outputs and does not refit or tune.
- Train/test membership, truth, predicted class, probability, and threshold
  remain visible in both table contracts.
- Complete metadata are displayed when present, but metadata are not assumed
  to be model predictors.
- The heatmap z-scores are dataset-relative visualization values and should not
  be interpreted as model coefficients or transferable measurements.
- Consult current `gt`, `gtExtras`, `ComplexHeatmap`, and `circlize`
  documentation for advanced table styling and annotation options.
