---
title: "Biomarker binary classification with bias-reduced logistic regression"
format:
  html:
    toc: true
    code-fold: true
execute:
  freeze: false
---

> Status: `draft`
>
> Template class: SOURCE-BACKED WORKFLOW

## Purpose

Fit a small-cohort binary biomarker classifier with a fixed train/test split,
training-only preprocessing, supervised feature selection inside resampling,
class weighting, bias-reduced logistic regression, and explicit held-out
evaluation. The reduced stable-feature model is a secondary sensitivity
analysis, not a replacement for the primary model.

::: {.callout-note}
## Source-backed workflow

This notebook preserves the implemented modeling route and reporting contracts
from the source biomarker analysis. Project labels, identifiers, and paths are
replaced by explicit inputs and parameters.
:::

## Inputs and parameters

The expression file is a wide CSV with one feature identifier column and one
sample per remaining column. The split and signature files are TSV files. The
split file must contain one row per sample and the columns named below.

```{r}
# Input contract
expression_path <- "input/feature_expression.csv"
split_path <- "input/train_test_split.tsv"
signature_path <- "input/signature_features.tsv"
metadata_path <- NULL
feature_id_column <- 1L

sample_id_column <- "sample_id"
outcome_column <- "outcome"
split_column <- "train_test"
train_label <- "train"
test_label <- "test"
positive_class <- "positive"
negative_class <- "negative"

# Modeling decisions
class_weight_alpha <- 0.5
mc_prop <- 0.70
mc_times <- 200L
top_n_features <- 10L
threshold_grid <- seq(0.45, 0.80, by = 0.05)
stability_cutoff <- 0.50
reduced_panel_size <- 4L
run_reduced_panel <- TRUE
seed <- 1234L

# Output contract
output_model <- "output/models/biomarker_brglm2_fit_bundle.qs2"
output_reduced_model <- "output/models/biomarker_brglm2_reduced_fit_bundle.qs2"
output_table_dir <- "output/tables"
output_figure_dir <- "output/figures"
dir.create(dirname(output_model), recursive = TRUE, showWarnings = FALSE)
dir.create(output_table_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(output_figure_dir, recursive = TRUE, showWarnings = FALSE)
```

## Analysis

```{r}
library(tidyverse)
library(tidymodels)
library(tailor)
library(filtro)
library(important)
library(desirability2)
library(brglm2)
library(bundle)
library(qs2)

tidymodels_prefer()
set.seed(seed)

split_raw <- readr::read_tsv(split_path, show_col_types = FALSE)
stopifnot(all(c(sample_id_column, outcome_column, split_column) %in% names(split_raw)))

split_tbl <- split_raw |>
  dplyr::transmute(
    sample_id = .data[[sample_id_column]],
    outcome = .data[[outcome_column]],
    train_test = .data[[split_column]]
  )

stopifnot(
  !anyNA(split_tbl$sample_id),
  !anyDuplicated(split_tbl$sample_id),
  all(split_tbl$train_test %in% c(train_label, test_label)),
  all(split_tbl$outcome %in% c(positive_class, negative_class))
)

split_tbl <- split_tbl |>
  dplyr::mutate(
    outcome = factor(outcome, levels = c(positive_class, negative_class))
  )

signature_tbl <- readr::read_tsv(signature_path, show_col_types = FALSE)
signature_features <- signature_tbl[[1]] |>
  as.character() |>
  unique()
stopifnot(length(signature_features) > 0L, !anyNA(signature_features))

expression_wide <- readr::read_csv(expression_path, show_col_types = FALSE)
names(expression_wide)[feature_id_column] <- "feature_id"
expression_sample_ids <- setdiff(names(expression_wide), "feature_id")

stopifnot(
  !anyDuplicated(expression_wide$feature_id),
  !anyNA(expression_wide$feature_id),
  !anyDuplicated(expression_sample_ids),
  all(split_tbl$sample_id %in% expression_sample_ids),
  all(signature_features %in% expression_wide$feature_id)
)

expression_matrix <- expression_wide |>
  dplyr::filter(feature_id %in% signature_features) |>
  tidyr::pivot_longer(
    -feature_id,
    names_to = "sample_id",
    values_to = "expression"
  ) |>
  tidyr::pivot_wider(
    names_from = feature_id,
    values_from = expression
  )

original_features <- setdiff(names(expression_matrix), "sample_id")
predictor_names <- make.names(original_features, unique = TRUE)
predictor_map <- tibble::tibble(
  original_feature = original_features,
  predictor = predictor_names
)
names(expression_matrix)[-1] <- predictor_names

model_data <- split_tbl |>
  dplyr::inner_join(expression_matrix, by = "sample_id")
stopifnot(nrow(model_data) == nrow(split_tbl))

train_data <- model_data |>
  dplyr::filter(train_test == train_label) |>
  dplyr::select(-train_test)
test_data <- model_data |>
  dplyr::filter(train_test == test_label) |>
  dplyr::select(-train_test)

stopifnot(
  nrow(train_data) > 1L,
  nrow(test_data) > 0L,
  dplyr::n_distinct(train_data$outcome) == 2L,
  dplyr::n_distinct(test_data$outcome) == 2L,
  !anyDuplicated(train_data$sample_id),
  !anyDuplicated(test_data$sample_id),
  !anyDuplicated(intersect(train_data$sample_id, test_data$sample_id)),
  top_n_features > 0L,
  top_n_features <= length(predictor_names)
)

class_weight_tbl <- train_data |>
  dplyr::count(outcome) |>
  dplyr::mutate(
    class_weight = (sum(n) / (dplyr::n_distinct(outcome) * n)) ^ class_weight_alpha,
    class_weight = class_weight / stats::weighted.mean(class_weight, n)
  ) |>
  dplyr::select(outcome, class_weight)

train_data <- train_data |>
  dplyr::left_join(class_weight_tbl, by = "outcome") |>
  dplyr::mutate(
    class_weight = hardhat::importance_weights(as.numeric(class_weight))
  )

readr::write_tsv(
  class_weight_tbl,
  file.path(output_table_dir, "class_weights.tsv")
)
```

The recipe is estimated separately inside every resampling split. The
supervised filter therefore cannot use assessment or test observations when it
selects features. Importance weights are represented as a typed case-weight
column rather than as an ordinary biomarker predictor.

```{r}
filter_goals <- desirability2::desirability(
  maximize(roc_auc),
  maximize(aov_fstat)
)

filter_prop <- top_n_features / length(predictor_names)

model_recipe <- recipes::recipe(outcome ~ ., data = train_data) |>
  recipes::update_role(sample_id, new_role = "id") |>
  recipes::step_zv(recipes::all_predictors()) |>
  important::step_predictor_desirability(
    recipes::all_predictors(),
    score = filter_goals,
    prop_terms = filter_prop
  ) |>
  recipes::step_normalize(recipes::all_predictors())

training_balance <- model_recipe |>
  recipes::prep() |>
  recipes::bake(new_data = NULL) |>
  dplyr::count(outcome, name = "n")

readr::write_tsv(
  training_balance,
  file.path(output_table_dir, "training_balance_after_preprocessing.tsv")
)

cv_folds <- rsample::mc_cv(
  train_data,
  prop = mc_prop,
  times = mc_times,
  strata = outcome
)

tune_metrics <- yardstick::metric_set(
  yardstick::bal_accuracy,
  yardstick::kap,
  yardstick::f_meas,
  yardstick::roc_auc,
  yardstick::accuracy,
  yardstick::sens,
  yardstick::spec,
  yardstick::recall,
  yardstick::mcc
)

extract_selected_features <- function(x) {
  selected_predictors <- workflows::extract_mold(x)$predictors |>
    names() |>
    intersect(predictor_names)

  tibble::tibble(predictor = selected_predictors)
}

brglm_spec <- parsnip::logistic_reg() |>
  parsnip::set_engine(
    "glm",
    family = stats::binomial(),
    method = brglm2::brglm_fit,
    type = "AS_mean"
  ) |>
  parsnip::set_mode("classification")

threshold_tailor <- tailor::tailor() |>
  tailor::adjust_probability_threshold(threshold = tune::tune())

brglm_workflow <- workflows::workflow() |>
  workflows::add_recipe(model_recipe) |>
  workflows::add_model(brglm_spec) |>
  workflows::add_tailor(threshold_tailor)

brglm_tune <- tune::tune_grid(
  brglm_workflow,
  resamples = cv_folds,
  grid = tibble::tibble(threshold = threshold_grid),
  metrics = tune_metrics,
  control = tune::control_grid(
    save_pred = TRUE,
    extract = extract_selected_features
  )
)

best_brglm <- tune::select_best(
  brglm_tune,
  metric = "bal_accuracy"
)

brglm_fit <- tune::finalize_workflow(brglm_workflow, best_brglm) |>
  parsnip::fit(data = train_data)

brglm_fit_bundle <- bundle::bundle(brglm_fit)
qs2::qs_save(brglm_fit_bundle, output_model)

readr::write_tsv(
  tune::collect_metrics(brglm_tune),
  file.path(output_table_dir, "tuning_metrics.tsv")
)
```

## Held-out diagnostics

The test set is used once for final evaluation. ROC-AUC and PR-AUC describe
probability ranking; threshold-dependent metrics describe the selected cutoff.
Undefined metrics remain `NA` when a class is absent from the predictions.

```{r}
test_predictions <- test_data |>
  dplyr::select(sample_id, outcome) |>
  dplyr::bind_cols(
    predict(brglm_fit, new_data = test_data, type = "prob"),
    predict(brglm_fit, new_data = test_data, type = "class")
  ) |>
  dplyr::mutate(
    truth = factor(outcome, levels = c(positive_class, negative_class)),
    .pred_class = factor(.pred_class, levels = c(positive_class, negative_class))
  )

test_metrics <- dplyr::bind_rows(
  tune_metrics(
    test_predictions,
    truth = truth,
    estimate = .pred_class,
    .pred_positive,
    event_level = "first"
  ),
  yardstick::pr_auc(
    test_predictions,
    truth = truth,
    .pred_positive,
    event_level = "first"
  ),
  yardstick::precision(
    test_predictions,
    truth = truth,
    estimate = .pred_class,
    event_level = "first"
  ),
  yardstick::npv(
    test_predictions,
    truth = truth,
    estimate = .pred_class,
    event_level = "first"
  ),
  yardstick::j_index(
    test_predictions,
    truth = truth,
    estimate = .pred_class,
    event_level = "first"
  )
) |>
  dplyr::mutate(
    model = "bias-reduced logistic regression",
    evaluation = paste0(
      "held-out test; threshold = ",
      round(best_brglm$threshold, 3)
    )
  )

confusion_table <- test_predictions |>
  dplyr::count(truth, prediction = .pred_class, name = "n")

roc_curve <- yardstick::roc_curve(
  test_predictions,
  truth = truth,
  .pred_positive,
  event_level = "first"
)
pr_curve <- yardstick::pr_curve(
  test_predictions,
  truth = truth,
  .pred_positive,
  event_level = "first"
)

readr::write_tsv(test_predictions, file.path(output_table_dir, "test_predictions.tsv"))
readr::write_tsv(test_metrics, file.path(output_table_dir, "test_metrics.tsv"))
readr::write_tsv(confusion_table, file.path(output_table_dir, "test_confusion_matrix.tsv"))
readr::write_tsv(roc_curve, file.path(output_table_dir, "test_roc_curve.tsv"))
readr::write_tsv(pr_curve, file.path(output_table_dir, "test_pr_curve.tsv"))

roc_plot <- autoplot(roc_curve) +
  ggplot2::labs(title = "Held-out ROC curve")
pr_plot <- autoplot(pr_curve) +
  ggplot2::labs(title = "Held-out precision-recall curve")
ggplot2::ggsave(
  file.path(output_figure_dir, "test_roc_curve.png"),
  roc_plot,
  width = 7,
  height = 5,
  dpi = 300
)
ggplot2::ggsave(
  file.path(output_figure_dir, "test_pr_curve.png"),
  pr_plot,
  width = 7,
  height = 5,
  dpi = 300
)
```

## Feature stability

Feature stability is calculated from one selected-feature record per Monte
Carlo resample. The final model's selected features and the full stability
table are both retained.

```{r}
selected_features_by_resample <- tune::collect_extracts(brglm_tune) |>
  dplyr::select(id, .extracts) |>
  dplyr::distinct(id, .keep_all = TRUE) |>
  tidyr::unnest(.extracts)

n_resamples <- dplyr::n_distinct(selected_features_by_resample$id)

feature_stability <- selected_features_by_resample |>
  dplyr::group_by(predictor) |>
  dplyr::summarise(
    n_selected = dplyr::n_distinct(id),
    selection_frequency = n_selected / n_resamples,
    .groups = "drop"
  ) |>
  dplyr::left_join(predictor_map, by = "predictor") |>
  dplyr::mutate(
    feature = dplyr::coalesce(original_feature, predictor),
    stability_class = dplyr::case_when(
      selection_frequency >= 0.70 ~ "highly stable",
      selection_frequency >= stability_cutoff ~ "stable",
      TRUE ~ "unstable"
    )
  ) |>
  dplyr::select(feature, predictor, n_selected, selection_frequency, stability_class) |>
  dplyr::arrange(dplyr::desc(selection_frequency), feature)

final_predictors <- workflows::extract_mold(brglm_fit)$predictors |>
  names() |>
  intersect(predictor_names)

final_feature_stability <- feature_stability |>
  dplyr::filter(predictor %in% final_predictors)

readr::write_tsv(
  feature_stability,
  file.path(output_table_dir, "feature_stability.tsv")
)
readr::write_tsv(
  final_feature_stability,
  file.path(output_table_dir, "final_feature_stability.tsv")
)

stability_plot <- feature_stability |>
  dplyr::slice_max(selection_frequency, n = 20, with_ties = FALSE) |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = selection_frequency,
      y = forcats::fct_reorder(feature, selection_frequency),
      fill = stability_class
    )
  ) +
  ggplot2::geom_col() +
  ggplot2::scale_x_continuous(limits = c(0, 1)) +
  ggplot2::labs(
    title = "Feature-selection stability",
    x = "Selection frequency",
    y = NULL,
    fill = "Stability"
  ) +
  ggplot2::theme_bw()

ggplot2::ggsave(
  file.path(output_figure_dir, "feature_stability.png"),
  stability_plot,
  width = 8,
  height = 6,
  dpi = 300
)
```

## Secondary reduced stable-feature model

This sensitivity model uses only features that meet the declared stability
cutoff. Its feature panel is selected from training resamples before the
held-out test evaluation. It must not replace the primary model silently.

```{r}
if (run_reduced_panel) {
  stable_predictors <- feature_stability |>
    dplyr::filter(selection_frequency >= stability_cutoff) |>
    dplyr::arrange(dplyr::desc(selection_frequency), feature) |>
    dplyr::slice_head(n = reduced_panel_size) |>
    dplyr::pull(predictor)

  stopifnot(length(stable_predictors) == reduced_panel_size)

  reduced_recipe <- recipes::recipe(outcome ~ ., data = train_data) |>
    recipes::update_role(sample_id, new_role = "id") |>
    recipes::step_rm(
      tidyselect::all_of(setdiff(predictor_names, stable_predictors))
    ) |>
    recipes::step_zv(recipes::all_predictors()) |>
    recipes::step_normalize(recipes::all_predictors())

  reduced_workflow <- workflows::workflow() |>
    workflows::add_recipe(reduced_recipe) |>
    workflows::add_model(brglm_spec) |>
    workflows::add_tailor(threshold_tailor)

  reduced_tune <- tune::tune_grid(
    reduced_workflow,
    resamples = cv_folds,
    grid = tibble::tibble(threshold = threshold_grid),
    metrics = tune_metrics,
    control = tune::control_grid(save_pred = TRUE)
  )
  reduced_best <- tune::select_best(reduced_tune, metric = "bal_accuracy")
  reduced_fit <- tune::finalize_workflow(reduced_workflow, reduced_best) |>
    parsnip::fit(data = train_data)

  reduced_fit_bundle <- bundle::bundle(reduced_fit)
  qs2::qs_save(reduced_fit_bundle, output_reduced_model)

  reduced_test_predictions <- test_data |>
    dplyr::select(sample_id, outcome) |>
    dplyr::bind_cols(
      predict(reduced_fit, new_data = test_data, type = "prob"),
      predict(reduced_fit, new_data = test_data, type = "class")
    ) |>
    dplyr::mutate(
      truth = factor(outcome, levels = c(positive_class, negative_class)),
      .pred_class = factor(.pred_class, levels = c(positive_class, negative_class))
    )

  reduced_test_metrics <- dplyr::bind_rows(
    tune_metrics(
      reduced_test_predictions,
      truth = truth,
      estimate = .pred_class,
      .pred_positive,
      event_level = "first"
    ),
    yardstick::pr_auc(
      reduced_test_predictions,
      truth = truth,
      .pred_positive,
      event_level = "first"
    ),
    yardstick::precision(
      reduced_test_predictions,
      truth = truth,
      estimate = .pred_class,
      event_level = "first"
    ),
    yardstick::npv(
      reduced_test_predictions,
      truth = truth,
      estimate = .pred_class,
      event_level = "first"
    ),
    yardstick::j_index(
      reduced_test_predictions,
      truth = truth,
      estimate = .pred_class,
      event_level = "first"
    )
  ) |>
    dplyr::mutate(
      model = paste0("reduced stable-feature panel (n=", length(stable_predictors), ")"),
      evaluation = paste0(
        "held-out test; threshold = ",
        round(reduced_best$threshold, 3)
      )
    )

  reduced_confusion_table <- reduced_test_predictions |>
    dplyr::count(truth, prediction = .pred_class, name = "n")
  reduced_roc_curve <- yardstick::roc_curve(
    reduced_test_predictions,
    truth = truth,
    .pred_positive,
    event_level = "first"
  )
  reduced_pr_curve <- yardstick::pr_curve(
    reduced_test_predictions,
    truth = truth,
    .pred_positive,
    event_level = "first"
  )

  readr::write_tsv(
    feature_stability |>
      dplyr::filter(predictor %in% stable_predictors) |>
      dplyr::slice_head(n = reduced_panel_size),
    file.path(output_table_dir, "reduced_stable_features.tsv")
  )
  readr::write_tsv(
    reduced_test_predictions,
    file.path(output_table_dir, "reduced_test_predictions.tsv")
  )
  readr::write_tsv(
    reduced_test_metrics,
    file.path(output_table_dir, "reduced_test_metrics.tsv")
  )
  readr::write_tsv(
    reduced_confusion_table,
    file.path(output_table_dir, "reduced_test_confusion_matrix.tsv")
  )
  readr::write_tsv(
    reduced_roc_curve,
    file.path(output_table_dir, "reduced_test_roc_curve.tsv")
  )
  readr::write_tsv(
    reduced_pr_curve,
    file.path(output_table_dir, "reduced_test_pr_curve.tsv")
  )
}
```

## Frozen prediction summaries

These tables are the input contract for
`biomarker_model_reporting.qmd`. Feature columns are prefixed with
`feature_`, while the complete table may additionally contain metadata joined
after model fitting. Metadata are never used by the classifier unless they are
explicitly added to the expression input and recipe.

```{r}
feature_output_names <- paste0("feature_", final_predictors)
feature_values <- model_data |>
  dplyr::select(sample_id, dplyr::all_of(final_predictors))
names(feature_values)[-1] <- feature_output_names

all_predictions <- model_data |>
  dplyr::select(sample_id, train_test, outcome) |>
  dplyr::bind_cols(
    predict(brglm_fit, new_data = model_data, type = "prob"),
    predict(brglm_fit, new_data = model_data, type = "class")
  ) |>
  dplyr::transmute(
    sample_id,
    split = train_test,
    truth = as.character(outcome),
    predicted_class = as.character(.pred_class),
    probability_positive = .pred_positive,
    probability_negative = .pred_negative,
    threshold = best_brglm$threshold
  ) |>
  dplyr::left_join(feature_values, by = "sample_id")

complete_summary <- all_predictions

if (!is.null(metadata_path)) {
  metadata <- readr::read_tsv(metadata_path, show_col_types = FALSE)
  stopifnot(sample_id_column %in% names(metadata))
  names(metadata)[names(metadata) == sample_id_column] <- "sample_id"
  stopifnot(!anyDuplicated(metadata$sample_id))
  complete_summary <- complete_summary |>
    dplyr::left_join(metadata, by = "sample_id")
}

essential_columns <- c(
  "sample_id",
  "split",
  "truth",
  "predicted_class",
  "probability_positive",
  "probability_negative",
  "threshold",
  feature_output_names
)
essential_summary <- complete_summary |>
  dplyr::select(dplyr::all_of(essential_columns))

readr::write_tsv(
  complete_summary,
  file.path(output_table_dir, "biomarker_train_test_complete.tsv")
)
readr::write_tsv(
  essential_summary,
  file.path(output_table_dir, "biomarker_train_test_essential.tsv")
)

if (requireNamespace("sessioninfo", quietly = TRUE)) {
  capture.output(sessioninfo::session_info(), file = "output/session_info.txt")
}
```

## Method notes

- The train/test split is supplied, not inferred from the feature matrix.
- Feature selection, normalization, and threshold tuning occur using training
  data and resampling only; the test set remains held out until the final
  evaluation.
- The supervised top-feature filter and class weighting are retained from the
  source workflow. The number of Monte Carlo resamples is exposed because it
  affects stability and runtime.
- Bias-reduced logistic regression is a modeling choice from the source
  workflow, not a guarantee against small-sample optimism or separation.
- ROC-AUC, PR-AUC, threshold metrics, confusion counts, coefficient summaries,
  and feature stability answer different questions and should be reviewed
  together.
- The reduced panel is a training-derived sensitivity analysis. Panel size and
  stability cutoff are scientific decisions and should not be selected from
  held-out performance.

::: {.callout-warning}
## Replication and validation warning

Rows in this notebook are observations for the classifier. If multiple rows
come from the same patient, donor, or experimental unit, a row-level split can
overstate generalization. Use grouped splitting or external validation when
the biological replicate is not the individual row.
:::

Before using this template, verify the current `tidymodels`, `tailor`,
`filtro`, `desirability2`, `brglm2`, and `bundle` APIs, reproduce the held-out
metrics with representative inputs, inspect undefined metrics, and confirm
that the model bundle can be reloaded with `qs2::qs_read()`.
