---
title: "MOFA Template: Project Setup"
format: html
editor: source
---

# Overview

This notebook defines the shared contract for the distilled MOFA workflow:

1. Build matched multi-omics inputs under `input/tables/`.
2. Create score-augmented metadata and aligned matrices.
3. Fit or inspect the MOFA model under `output/objects/`.
4. Keep project-specific feature choices and biological interpretation inline.

# Setup

```{r}
set.seed(1234)

Sys.setenv(
  OPENBLAS_NUM_THREADS = 1,
  OMP_NUM_THREADS = 1,
  MKL_NUM_THREADS = 1,
  VECLIB_MAXIMUM_THREADS = 1,
  NUMEXPR_NUM_THREADS = 1
)

options(future.globals.maxSize = 25 * 1024^3)

paths <- list(
  input_tables = "input/tables",
  input_objects = "input/objects",
  output_tables = "output/tables",
  output_figures = "output/figures",
  output_objects = "output/objects",
  helpers = "helpers/helpers_from_MOFA.R"
)

invisible(lapply(paths, dir.create, recursive = TRUE, showWarnings = FALSE))

project <- list(
  project_id = "PROJECT_ID",
  cohort_label = "COHORT_LABEL",
  sample_id_length = 16,
  score_column = "C1_score",
  class_column = "C1_class"
)
```

# Notebook Order

```{r}
workflow_order <- c(
  "00_project_setup.qmd",
  "01_multiomics_input_prep.qmd",
  "02_signature_scoring.qmd",
  "03_mofa_model.qmd"
)
workflow_order
```

# Object Contract

```{r}
object_contract <- list(
  matched_inputs = file.path(paths$input_tables, paste0(project$project_id, "_matched_multiomics_manifest.tsv")),
  scored_metadata = file.path(paths$output_tables, paste0(project$project_id, "_scored_metadata.tsv")),
  mofa_input = file.path(paths$output_objects, paste0(project$project_id, "_mofa_input.qs2")),
  mofa_model = file.path(paths$output_objects, paste0(project$project_id, "_mofa_model.hdf5")),
  mofa_factors = file.path(paths$output_tables, paste0(project$project_id, "_mofa_factors.tsv"))
)

knitr::kable(
  data.frame(name = names(object_contract), path = unlist(object_contract)),
  caption = "Shared MOFA template outputs"
)
```

# Notes

- Keep assay-specific file names and download queries visible in the prep notebook.
- Keep score thresholds and subtype labels visible in the scoring notebook.
- Keep model dimensions, view selection, and factor interpretation visible in the MOFA notebook.
