---
title: "MOFA Template: Project Setup"
editor: source
lang: en
format:
  html:
    theme:
      light: cosmo
      dark: darkly
    respect-user-color-scheme: true
    toc: true
    toc-title: "Contents"
    toc-depth: 4
    toc-location: right
    toc-expand: 2
    number-sections: true
    number-depth: 3
    smooth-scroll: true
    anchor-sections: true
    code-fold: true
    code-summary: "Show analysis code"
    code-copy: true
    code-overflow: wrap
    code-line-numbers: false
    fig-width: 16
    fig-height: 12
    fig-dpi: 300
    fig-format: png
    fig-align: center
    fig-responsive: true
    fig-cap-location: bottom
    tbl-cap-location: top
    df-print: paged
    crossrefs-hover: true
    footnotes-hover: true
    citations-hover: true
    embed-resources: true
    self-contained-math: true
    page-layout: article
    grid:
      sidebar-width: 250px
      body-width: 1050px
      margin-width: 300px
      gutter-width: 1.5rem
    link-external-newwindow: true
    link-external-icon: true
lightbox:
  match: auto
  effect: fade
  desc-position: bottom
  loop: false
execute:
  echo: true
  warning: false
  message: false
  error: false
knitr:
  opts_chunk:
    fig.align: center
    out.width: "100%"
    collapse: true
    comment: "#>"
crossref:
  fig-title: "Figure"
  tbl-title: "Table"
  sec-prefix: "Section"
  fig-prefix: "Figure"
  tbl-prefix: "Table"
  ref-hyperlink: true
---

## Purpose

This notebook defines the shared contract for the distilled MOFA workflow:

1. Build matched multi-omics inputs under `input/tables/`.
2. Create score-augmented metadata and aligned matrices.
3. Fit or inspect the MOFA model under `output/objects/`.
4. Keep project-specific feature choices and biological interpretation inline.

::: {.callout-note}
## Archive status

This is a legacy notebook kept for reference. For current reusable workflows,
see the active templates in the repository.
:::

## Inputs and setup

Set the project-level paths and identifiers used by the later MOFA notebooks.

```{r}
#| label: mofa-project-setup
set.seed(1234)

Sys.setenv(
  OPENBLAS_NUM_THREADS = 1,
  OMP_NUM_THREADS = 1,
  MKL_NUM_THREADS = 1,
  VECLIB_MAXIMUM_THREADS = 1,
  NUMEXPR_NUM_THREADS = 1
)

options(future.globals.maxSize = 25 * 1024^3)

paths <- list(
  input_tables = "input/tables",
  input_objects = "input/objects",
  output_tables = "output/tables",
  output_figures = "output/figures",
  output_objects = "output/objects",
  helpers = "helpers/helpers_from_MOFA.R"
)

invisible(lapply(paths, dir.create, recursive = TRUE, showWarnings = FALSE))

project <- list(
  project_id = "PROJECT_ID",
  cohort_label = "COHORT_LABEL",
  sample_id_length = 16,
  score_column = "C1_score",
  class_column = "C1_class"
)
```

## Workflow order

This short list records the intended order of the four legacy notebooks.

```{r}
#| label: mofa-workflow-order
workflow_order <- c(
  "00_project_setup.qmd",
  "01_multiomics_input_prep.qmd",
  "02_signature_scoring.qmd",
  "03_mofa_model.qmd"
)
workflow_order
```

## Outputs

The object contract below names the files passed between the notebook steps.

```{r}
#| label: tbl-mofa-output-contract
#| tbl-cap: "Shared outputs for the MOFA workflow."
object_contract <- list(
  matched_inputs = file.path(paths$input_tables, paste0(project$project_id, "_matched_multiomics_manifest.tsv")),
  scored_metadata = file.path(paths$output_tables, paste0(project$project_id, "_scored_metadata.tsv")),
  mofa_input = file.path(paths$output_objects, paste0(project$project_id, "_mofa_input.qs2")),
  mofa_model = file.path(paths$output_objects, paste0(project$project_id, "_mofa_model.hdf5")),
  mofa_factors = file.path(paths$output_tables, paste0(project$project_id, "_mofa_factors.tsv"))
)

knitr::kable(
  data.frame(name = names(object_contract), path = unlist(object_contract)),
  caption = "Shared MOFA template outputs"
)
```

## Notes

- Keep assay-specific file names and download queries visible in the prep notebook.
- Keep score thresholds and subtype labels visible in the scoring notebook.
- Keep model dimensions, view selection, and factor interpretation visible in the MOFA notebook.

## Reproducibility {.appendix}

This final chunk records the R session used when the notebook is rendered.

```{r}
#| label: session-information
#| echo: false
#| results: markup
sessionInfo()
```
