---
title: "MOFA Template: Multi-Omics Input Preparation"
editor: source
lang: en
format:
  html:
    theme:
      light: cosmo
      dark: darkly
    respect-user-color-scheme: false
    toc: true
    toc-title: "Contents"
    toc-depth: 4
    toc-location: right
    toc-expand: 2
    number-sections: true
    number-depth: 3
    smooth-scroll: true
    anchor-sections: true
    code-fold: true
    code-summary: "Show analysis code"
    code-copy: true
    code-overflow: wrap
    code-line-numbers: false
    fig-width: 16
    fig-height: 12
    fig-dpi: 300
    fig-format: png
    fig-align: center
    fig-responsive: true
    fig-cap-location: bottom
    tbl-cap-location: top
    df-print: paged
    crossrefs-hover: true
    footnotes-hover: true
    citations-hover: true
    embed-resources: true
    self-contained-math: true
    page-layout: article
    grid:
      sidebar-width: 250px
      body-width: 1050px
      margin-width: 300px
      gutter-width: 1.5rem
    link-external-newwindow: true
    link-external-icon: true
lightbox:
  match: auto
  effect: fade
  desc-position: bottom
  loop: false
execute:
  echo: true
  warning: false
  message: false
  error: false
knitr:
  opts_chunk:
    fig.align: center
    out.width: "100%"
    collapse: true
    comment: "#>"
crossref:
  fig-title: "Figure"
  tbl-title: "Table"
  sec-prefix: "Section"
  fig-prefix: "Figure"
  tbl-prefix: "Table"
  ref-hyperlink: true
---

## Purpose

This notebook prepares the matched matrices consumed by the scoring and MOFA notebooks.

::: {.callout-note}
## Archive status

This is a legacy notebook kept for reference. For current reusable workflows,
see the active templates in the repository.
:::

## Inputs and setup

Set the project identifiers and input table locations before matching samples.

```{r}
#| label: mofa-input-setup
library(dplyr)
library(readr)
library(tibble)
library(purrr)
library(qs2)

source("helpers/helpers_from_MOFA.R")

project_id <- "PROJECT_ID"
sample_id_length <- 16

dir.create("input/tables", recursive = TRUE, showWarnings = FALSE)
dir.create("output/tables", recursive = TRUE, showWarnings = FALSE)

input_files <- list(
  rna = "input/tables/RNAseq_raw.tsv",
  mirna = "input/tables/miRNA_raw.tsv",
  methylation = "input/tables/methylation_raw.tsv",
  mutation = "input/tables/mutation_raw.tsv",
  rppa = "input/tables/RPPA_raw.tsv"
)
```

## Normalize sample IDs

Normalize sample columns before comparing the available sample sets.

```{r}
#| label: mofa-normalize-sample-ids
normalize_sample_cols <- function(df, feature_col) {
  sample_cols <- setdiff(colnames(df), feature_col)
  renamed <- stats::setNames(sample_cols, tcga_sample_barcode(sample_cols))
  colnames(df) <- c(feature_col, renamed)
  df
}
```

## Build matched tables

Read each modality, align its sample identifiers, and write the matched-input manifest.

```{r}
#| label: mofa-build-matched-tables
# Replace these example reads with the concrete assay exports for a new project.
rna <- readr::read_tsv(input_files$rna, show_col_types = FALSE)
mirna <- readr::read_tsv(input_files$mirna, show_col_types = FALSE)
methylation <- readr::read_tsv(input_files$methylation, show_col_types = FALSE)
mutation <- readr::read_tsv(input_files$mutation, show_col_types = FALSE)
rppa <- readr::read_tsv(input_files$rppa, show_col_types = FALSE)

rna <- normalize_sample_cols(rna, feature_col = colnames(rna)[1])
mirna <- normalize_sample_cols(mirna, feature_col = colnames(mirna)[1])
methylation <- normalize_sample_cols(methylation, feature_col = colnames(methylation)[1])
mutation <- normalize_sample_cols(mutation, feature_col = colnames(mutation)[1])
rppa <- normalize_sample_cols(rppa, feature_col = colnames(rppa)[1])

sample_sets <- list(
  setdiff(colnames(rna), colnames(rna)[1]),
  setdiff(colnames(mirna), colnames(mirna)[1]),
  setdiff(colnames(methylation), colnames(methylation)[1]),
  setdiff(colnames(mutation), colnames(mutation)[1]),
  setdiff(colnames(rppa), colnames(rppa)[1])
)
matched_samples <- Reduce(intersect, sample_sets)
stopifnot(length(matched_samples) > 0)

write_output_manifest <- function() {
  manifest <- tibble::tibble(
    assay = c("rna", "mirna", "methylation", "mutation", "rppa"),
    n_samples = purrr::map_int(sample_sets, length),
    n_matched = length(matched_samples)
  )
  readr::write_tsv(
    manifest,
    file.path("input/tables", paste0(project_id, "_matched_multiomics_manifest.tsv"))
  )
}

write_output_manifest()
```

## Outputs

Save one matched table per modality for the downstream scoring and MOFA steps.

```{r}
#| label: mofa-save-matched-inputs
matched_tables <- list(
  rna = rna[, c(1, match(matched_samples, colnames(rna))), drop = FALSE],
  mirna = mirna[, c(1, match(matched_samples, colnames(mirna))), drop = FALSE],
  methylation = methylation[, c(1, match(matched_samples, colnames(methylation))), drop = FALSE],
  mutation = mutation[, c(1, match(matched_samples, colnames(mutation))), drop = FALSE],
  rppa = rppa[, c(1, match(matched_samples, colnames(rppa))), drop = FALSE]
)

purrr::iwalk(matched_tables, function(tbl, nm) {
  readr::write_tsv(tbl, file.path("input/tables", paste0(toupper(nm), "_MATCHED.tsv")))
})
```

## Notes

- Keep project-specific download code, GDC query logic, and collapse rules visible here.
- Do not hide assay-specific cleanup decisions in helpers.

## Reproducibility {.appendix}

This final chunk records the R session used when the notebook is rendered.

```{r}
#| label: session-information
#| echo: false
#| results: markup
sessionInfo()
```
