---
title: "MOFA Template: Signature Scoring and Metadata"
editor: source
lang: en
format:
  html:
    theme:
      light: cosmo
      dark: darkly
    respect-user-color-scheme: true
    toc: true
    toc-title: "Contents"
    toc-depth: 4
    toc-location: right
    toc-expand: 2
    number-sections: true
    number-depth: 3
    smooth-scroll: true
    anchor-sections: true
    code-fold: true
    code-summary: "Show analysis code"
    code-copy: true
    code-overflow: wrap
    code-line-numbers: false
    fig-width: 16
    fig-height: 12
    fig-dpi: 300
    fig-format: png
    fig-align: center
    fig-responsive: true
    fig-cap-location: bottom
    tbl-cap-location: top
    df-print: paged
    crossrefs-hover: true
    footnotes-hover: true
    citations-hover: true
    embed-resources: true
    self-contained-math: true
    page-layout: article
    grid:
      sidebar-width: 250px
      body-width: 1050px
      margin-width: 300px
      gutter-width: 1.5rem
    link-external-newwindow: true
    link-external-icon: true
lightbox:
  match: auto
  effect: fade
  desc-position: bottom
  loop: false
execute:
  echo: true
  warning: false
  message: false
  error: false
knitr:
  opts_chunk:
    fig.align: center
    out.width: "100%"
    collapse: true
    comment: "#>"
crossref:
  fig-title: "Figure"
  tbl-title: "Table"
  sec-prefix: "Section"
  fig-prefix: "Figure"
  tbl-prefix: "Table"
  ref-hyperlink: true
---

## Purpose

This notebook computes the phenotype or signature scores that will later be tested against MOFA factors.

::: {.callout-note}
## Archive status

This is a legacy notebook kept for reference. For current reusable workflows,
see the active templates in the repository.
:::

## Inputs and setup

Load the matched RNA table, metadata, and signature definition used for scoring.

```{r}
#| label: mofa-signature-setup
library(dplyr)
library(readr)
library(tibble)
library(ggplot2)
library(edgeR)

source("helpers/helpers_from_MOFA.R")

project_id <- "PROJECT_ID"
score_column <- "C1_score"
class_column <- "C1_class"

dir.create("output/tables", recursive = TRUE, showWarnings = FALSE)
dir.create("output/figures", recursive = TRUE, showWarnings = FALSE)

rna <- readr::read_tsv("input/tables/RNA_MATCHED.tsv", show_col_types = FALSE)
metadata <- readr::read_tsv("input/tables/sample_metadata.tsv", show_col_types = FALSE)
signature_tbl <- readr::read_csv("input/tables/signature_genes.csv", show_col_types = FALSE)
```

## Score construction

Calculate the signature score, derive its class labels, and save the scored metadata.

```{r}
#| label: mofa-score-construction
feature_col <- colnames(rna)[1]
sample_ids <- setdiff(colnames(rna), feature_col)
count_mat <- as.matrix(rna[, sample_ids, drop = FALSE])
rownames(count_mat) <- rna[[feature_col]]

logcpm <- compute_logcpm(count_mat)
signature_genes <- intersect(signature_tbl$gene, rownames(logcpm))
stopifnot(length(signature_genes) > 0)

score_vec <- colMeans(logcpm[signature_genes, , drop = FALSE], na.rm = TRUE)
score_tbl <- tibble::tibble(
  sample_id = names(score_vec),
  !!score_column := as.numeric(scale(score_vec))
)

score_tbl[[class_column]] <- class_from_z(score_tbl[[score_column]])
metadata_scored <- dplyr::left_join(metadata, score_tbl, by = "sample_id")

readr::write_tsv(
  metadata_scored,
  file.path("output/tables", paste0(project_id, "_scored_metadata.tsv"))
)
```

## Diagnostics

Inspect the score distribution across the configured class labels.

```{r}
#| label: mofa-score-diagnostics
ggplot(metadata_scored, aes(x = .data[[class_column]], y = .data[[score_column]], fill = .data[[class_column]])) +
  geom_violin(alpha = 0.5) +
  geom_jitter(width = 0.15, alpha = 0.2, size = 0.4) +
  theme_bw(11)
```

## Notes

- Keep alternate score definitions, deconvolution summaries, and purity correlations inline here when needed.
- Do not hide threshold choices; the class cutoffs should stay visible.

## Reproducibility {.appendix}

This final chunk records the R session used when the notebook is rendered.

```{r}
#| label: session-information
#| echo: false
#| results: markup
sessionInfo()
```
