---
title: "MOFA Template: Multi-Omics Model"
format: html
editor: source
---

# Overview

This notebook builds the aligned MOFA input list, trains the model, and keeps the central downstream analysis in one place:

1. factor-level summaries and score association
2. sample heatmaps across factors
3. top-feature heatmaps within each view
4. group-wise median and mutation-frequency differences for the leading factor

# Setup

```{r}
library(dplyr)
library(readr)
library(qs2)
library(MOFA2)
library(ggplot2)
library(tidyr)
library(tibble)
library(purrr)

source("helpers/helpers_from_MOFA.R")

project_id <- "PROJECT_ID"
top_features_per_view <- 5000
num_factors <- 10
focus_score_column <- "C1_score"
focus_group_column <- "C1_class"
focus_group_high <- "high"
focus_group_low <- "low"
heatmap_factors <- 1:5
top_features_for_heatmap <- 30

FIG_DIR <- "output/figures"
TABLE_DIR <- "output/tables"
OBJECT_DIR <- "output/objects"
dir.create(FIG_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(TABLE_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(OBJECT_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(FIG_DIR, "mofa_heatmaps"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(TABLE_DIR, "mofa_group_summaries"), recursive = TRUE, showWarnings = FALSE)

metadata_scored <- readr::read_tsv(
  file.path(TABLE_DIR, paste0(project_id, "_scored_metadata.tsv")),
  show_col_types = FALSE
)
```

# Helper Functions

```{r}
load_view_matrix <- function(path) {
  tbl <- readr::read_tsv(path, show_col_types = FALSE)
  feature_col <- colnames(tbl)[1]
  sample_ids <- setdiff(colnames(tbl), feature_col)
  mat <- as.matrix(tbl[, sample_ids, drop = FALSE])
  rownames(mat) <- tbl[[feature_col]]
  storage.mode(mat) <- "numeric"
  mat
}

row_zscore <- function(mat) {
  z <- t(scale(t(mat)))
  z[is.na(z)] <- 0
  z
}

factor_tbl_from_fit <- function(mofa_fit) {
  MOFA2::get_factors(mofa_fit, factors = "all", as.data.frame = TRUE) |>
    dplyr::rename(sample_id = sample, factor = factor, value = value)
}

weight_tbl_from_fit <- function(mofa_fit, view_name) {
  MOFA2::get_weights(mofa_fit, views = view_name, factors = "all", as.data.frame = TRUE) |>
    dplyr::rename(feature = feature, factor = factor, value = value)
}

plot_factor_heatmap <- function(factor_wide, annotation_cols) {
  long_df <- factor_wide |>
    tidyr::pivot_longer(
      cols = dplyr::starts_with("Factor"),
      names_to = "factor",
      values_to = "value"
    )

  ggplot2::ggplot(long_df, ggplot2::aes(x = factor, y = sample_id, fill = value)) +
    ggplot2::geom_tile() +
    ggplot2::scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B") +
    ggplot2::theme_minimal(base_size = 10) +
    ggplot2::theme(axis.text.y = ggplot2::element_blank()) +
    ggplot2::labs(x = NULL, y = "Samples", fill = "Factor")
}

plot_view_weight_heatmap <- function(weight_tbl, view_name, factor_id, n_features = 30) {
  factor_label <- paste0("Factor", factor_id)
  top_tbl <- weight_tbl |>
    dplyr::filter(factor == factor_label) |>
    dplyr::slice_max(order_by = abs(value), n = n_features, with_ties = FALSE) |>
    dplyr::mutate(feature = factor(feature, levels = rev(feature)))

  ggplot2::ggplot(top_tbl, ggplot2::aes(x = factor, y = feature, fill = value)) +
    ggplot2::geom_tile() +
    ggplot2::scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B") +
    ggplot2::theme_minimal(base_size = 10) +
    ggplot2::labs(
      title = paste(view_name, factor_label, "top weights"),
      x = NULL,
      y = NULL,
      fill = "Weight"
    )
}

extract_weights_with_group_medians <- function(weight_tbl, view_mat, metadata_tbl, factor_id, group_col, high_label, low_label, n_features = 50) {
  factor_label <- paste0("Factor", factor_id)
  top_tbl <- weight_tbl |>
    dplyr::filter(factor == factor_label) |>
    dplyr::slice_max(order_by = abs(value), n = n_features, with_ties = FALSE) |>
    dplyr::rename(weight = value)

  high_samples <- metadata_tbl |>
    dplyr::filter(.data[[group_col]] == high_label) |>
    dplyr::pull(sample_id)
  low_samples <- metadata_tbl |>
    dplyr::filter(.data[[group_col]] == low_label) |>
    dplyr::pull(sample_id)

  common_high <- intersect(high_samples, colnames(view_mat))
  common_low <- intersect(low_samples, colnames(view_mat))

  tibble::tibble(feature = rownames(view_mat)) |>
    dplyr::filter(feature %in% top_tbl$feature) |>
    dplyr::mutate(
      median_high = apply(view_mat[feature, common_high, drop = FALSE], 1, median, na.rm = TRUE),
      median_low = apply(view_mat[feature, common_low, drop = FALSE], 1, median, na.rm = TRUE),
      median_diff = median_high - median_low
    ) |>
    dplyr::left_join(top_tbl, by = "feature") |>
    dplyr::arrange(dplyr::desc(abs(weight)))
}

extract_mutation_weights_with_group_frequency <- function(weight_tbl, mutation_mat, metadata_tbl, factor_id, group_col, high_label, low_label, n_features = 50) {
  factor_label <- paste0("Factor", factor_id)
  top_tbl <- weight_tbl |>
    dplyr::filter(factor == factor_label) |>
    dplyr::slice_max(order_by = abs(value), n = n_features, with_ties = FALSE) |>
    dplyr::rename(weight = value)

  high_samples <- metadata_tbl |>
    dplyr::filter(.data[[group_col]] == high_label) |>
    dplyr::pull(sample_id)
  low_samples <- metadata_tbl |>
    dplyr::filter(.data[[group_col]] == low_label) |>
    dplyr::pull(sample_id)

  common_high <- intersect(high_samples, colnames(mutation_mat))
  common_low <- intersect(low_samples, colnames(mutation_mat))

  tibble::tibble(feature = rownames(mutation_mat)) |>
    dplyr::filter(feature %in% top_tbl$feature) |>
    dplyr::mutate(
      frequency_high = apply(mutation_mat[feature, common_high, drop = FALSE] > 0, 1, mean, na.rm = TRUE),
      frequency_low = apply(mutation_mat[feature, common_low, drop = FALSE] > 0, 1, mean, na.rm = TRUE),
      frequency_diff = frequency_high - frequency_low
    ) |>
    dplyr::left_join(top_tbl, by = "feature") |>
    dplyr::arrange(dplyr::desc(abs(weight)))
}
```

# Load Views

```{r}
views <- list(
  RNA = compute_logcpm(load_view_matrix("input/tables/RNA_MATCHED.tsv")),
  miRNA = compute_logcpm(load_view_matrix("input/tables/MIRNA_MATCHED.tsv")),
  methylation = load_view_matrix("input/tables/METHYLATION_MATCHED.tsv"),
  mutation = load_view_matrix("input/tables/MUTATION_MATCHED.tsv"),
  RPPA = load_view_matrix("input/tables/RPPA_MATCHED.tsv")
)
```

# Build MOFA Input

```{r}
view_mats <- lapply(views, function(mat) {
  keep_most_variable_dt(mat, top_n = top_features_per_view)
})

common_samples <- Reduce(intersect, lapply(view_mats, colnames))
stopifnot(length(common_samples) > 0)

view_mats <- lapply(view_mats, function(mat) mat[, common_samples, drop = FALSE])
metadata_scored <- metadata_scored |>
  dplyr::filter(sample_id %in% common_samples)

qs2::qs_save(view_mats, file.path(OBJECT_DIR, paste0(project_id, "_mofa_input.qs2")))

mofa_obj <- MOFA2::create_mofa(view_mats)
model_opts <- MOFA2::get_default_model_options(mofa_obj)
train_opts <- MOFA2::get_default_training_options(mofa_obj)
model_opts$num_factors <- num_factors

mofa_obj <- MOFA2::prepare_mofa(
  object = mofa_obj,
  model_options = model_opts,
  training_options = train_opts
)
```

# Train and Export

```{r}
outfile <- file.path(OBJECT_DIR, paste0(project_id, "_mofa_model.hdf5"))
mofa_fit <- MOFA2::run_mofa(mofa_obj, outfile = outfile)

factor_tbl <- factor_tbl_from_fit(mofa_fit)
readr::write_tsv(factor_tbl, file.path(TABLE_DIR, paste0(project_id, "_mofa_factors.tsv")))
```

# Factor Association Summary

```{r}
factor_wide <- factor_tbl |>
  tidyr::pivot_wider(names_from = factor, values_from = value) |>
  dplyr::left_join(metadata_scored, by = "sample_id")

association_tbl <- purrr::map_dfr(
  grep("^Factor", colnames(factor_wide), value = TRUE),
  function(factor_name) {
    stat <- suppressWarnings(
      stats::cor.test(
        factor_wide[[factor_name]],
        factor_wide[[focus_score_column]],
        method = "spearman"
      )
    )
    tibble::tibble(
      factor = factor_name,
      spearman_rho = unname(stat$estimate),
      p_value = stat$p.value
    )
  }
) |>
  dplyr::arrange(p_value)

readr::write_tsv(
  association_tbl,
  file.path(TABLE_DIR, paste0(project_id, "_factor_score_associations.tsv"))
)

association_tbl
```

# Sample-Level Factor Heatmap

```{r}
heatmap_cols <- intersect(
  c(paste0("Factor", heatmap_factors), focus_score_column, focus_group_column),
  colnames(factor_wide)
)

factor_heatmap_input <- factor_wide |>
  dplyr::select(sample_id, dplyr::all_of(heatmap_cols)) |>
  dplyr::arrange(dplyr::desc(.data[[focus_score_column]]))

plot_factor_heatmap(factor_heatmap_input, annotation_cols = c(focus_score_column, focus_group_column))
```

# Factor Distributions by Group

```{r}
factor_violin_tbl <- factor_wide |>
  tidyr::pivot_longer(
    cols = dplyr::starts_with("Factor"),
    names_to = "factor",
    values_to = "value"
  )

ggplot2::ggplot(
  factor_violin_tbl,
  ggplot2::aes(x = .data[[focus_group_column]], y = value, fill = .data[[focus_group_column]])
) +
  ggplot2::geom_violin(alpha = 0.5) +
  ggplot2::geom_jitter(width = 0.15, alpha = 0.15, size = 0.25) +
  ggplot2::facet_wrap(~ factor, scales = "free_y") +
  ggplot2::theme_bw(11)
```

# View Weight Heatmaps

```{r}
lead_factors <- association_tbl |>
  dplyr::slice_head(n = min(length(heatmap_factors), nrow(association_tbl))) |>
  dplyr::pull(factor)

weight_tbls <- purrr::map(names(view_mats), function(view_name) {
  weight_tbl_from_fit(mofa_fit, view_name) |>
    dplyr::mutate(view = view_name)
})
names(weight_tbls) <- names(view_mats)

purrr::walk(names(weight_tbls), function(view_name) {
  purrr::walk(lead_factors, function(factor_name) {
    factor_id <- as.integer(gsub("Factor", "", factor_name))
    p <- plot_view_weight_heatmap(
      weight_tbl = weight_tbls[[view_name]],
      view_name = view_name,
      factor_id = factor_id,
      n_features = top_features_for_heatmap
    )
    print(p)
    ggplot2::ggsave(
      filename = file.path(FIG_DIR, "mofa_heatmaps", paste0(view_name, "_", factor_name, "_weights_heatmap.png")),
      plot = p,
      width = 8,
      height = 10,
      dpi = 300
    )
  })
})
```

# Group Median and Frequency Differences

```{r}
target_factor <- as.integer(gsub("Factor", "", lead_factors[[1]]))

continuous_views <- c("RNA", "miRNA", "methylation", "RPPA")
continuous_summaries <- purrr::map(continuous_views, function(view_name) {
  extract_weights_with_group_medians(
    weight_tbl = weight_tbls[[view_name]],
    view_mat = view_mats[[view_name]],
    metadata_tbl = metadata_scored,
    factor_id = target_factor,
    group_col = focus_group_column,
    high_label = focus_group_high,
    low_label = focus_group_low,
    n_features = 50
  ) |>
    dplyr::mutate(view = view_name)
})
names(continuous_summaries) <- continuous_views

purrr::iwalk(continuous_summaries, function(tbl, view_name) {
  readr::write_tsv(
    tbl,
    file.path(
      TABLE_DIR,
      "mofa_group_summaries",
      paste0(project_id, "_", view_name, "_Factor", target_factor, "_median_differences.tsv")
    )
  )
})

mutation_summary <- extract_mutation_weights_with_group_frequency(
  weight_tbl = weight_tbls[["mutation"]],
  mutation_mat = view_mats[["mutation"]],
  metadata_tbl = metadata_scored,
  factor_id = target_factor,
  group_col = focus_group_column,
  high_label = focus_group_high,
  low_label = focus_group_low,
  n_features = 50
)

readr::write_tsv(
  mutation_summary,
  file.path(
    TABLE_DIR,
    "mofa_group_summaries",
    paste0(project_id, "_mutation_Factor", target_factor, "_frequency_differences.tsv")
  )
)
```

# Quick Checks

```{r}
variance_explained <- MOFA2::calculate_variance_explained(mofa_fit)
print(variance_explained)

continuous_summaries[[1]] |>
  dplyr::slice_head(n = 10)

mutation_summary |>
  dplyr::slice_head(n = 10)
```



### Heatmap Helper

Plot row-scaled top-weight features for selected views and factors, ordered by C1-core group.

```{r}
#| label: custom-complexheatmap
#| echo: false
#| warning: false
#| message: false
#| results: hide

library(MOFA2)
library(dplyr)
library(tibble)
library(ComplexHeatmap)
library(circlize)
library(viridis)
library(grid)

plot_mofa_view_heatmap <- function(
  MOFAobject,
  factor_id = 1,
  view_id = "RNA",
  group_id = "group1",
  n_features = 100,
  feature_selection = "balanced",
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  clustering_method = "average",
  save_png = TRUE,
  width = 14,
  height = 10
) {
  row_zscore <- function(mat) {
    mat <- as.matrix(mat)

    row_median <- apply(mat, 1, median, na.rm = TRUE)
    row_mad <- apply(mat, 1, mad, na.rm = TRUE)

    row_mad[is.na(row_mad) | row_mad == 0] <- 1

    mat_scaled <- sweep(mat, 1, row_median, "-")
    mat_scaled <- sweep(mat_scaled, 1, row_mad, "/")
    mat_scaled[is.na(mat_scaled)] <- 0

    mat_scaled
  }

  pearson_row_hclust <- function(mat) {
    cor_mat <- cor(
      t(mat),
      method = "pearson",
      use = "pairwise.complete.obs"
    )

    cor_mat[is.na(cor_mat)] <- 0

    hclust(
      as.dist(1 - cor_mat),
      method = clustering_method
    )
  }

  pearson_column_hclust <- function(mat) {
    cor_mat <- cor(
      mat,
      method = "pearson",
      use = "pairwise.complete.obs"
    )

    cor_mat[is.na(cor_mat)] <- 0

    hclust(
      as.dist(1 - cor_mat),
      method = clustering_method
    )
  }

  data_list <- get_data(
    MOFAobject,
    views = view_id,
    as.data.frame = FALSE
  )

  view_mat <- data_list[[view_id]][[group_id]]
  mofa_samples <- samples_names(MOFAobject)[[group_id]]

  if (all(mofa_samples %in% rownames(view_mat))) {
    view_mat <- t(view_mat)
  }

  if (!all(mofa_samples %in% colnames(view_mat))) {
    stop("MOFA samples are not in the columns of the selected view matrix.")
  }

  weights_df <- get_weights(
    MOFAobject,
    views = view_id,
    factors = factor_id,
    as.data.frame = TRUE
  )

  weights_view <- weights_df |>
    dplyr::filter(view == view_id) |>
    dplyr::mutate(abs_weight = abs(value))

  if (!feature_selection %in% c("balanced", "absolute")) {
    stop("feature_selection must be either 'balanced' or 'absolute'.")
  }

  if (feature_selection == "absolute") {
    top_features <- weights_view |>
      dplyr::arrange(dplyr::desc(abs_weight)) |>
      dplyr::slice_head(n = n_features) |>
      dplyr::pull(feature)
  }

  if (feature_selection == "balanced") {
    n_negative <- floor(n_features / 2)
    n_positive <- n_features - n_negative

    top_negative_features <- weights_view |>
      dplyr::filter(value < 0) |>
      dplyr::arrange(value) |>
      dplyr::slice_head(n = n_negative) |>
      dplyr::pull(feature)

    top_positive_features <- weights_view |>
      dplyr::filter(value > 0) |>
      dplyr::arrange(dplyr::desc(value)) |>
      dplyr::slice_head(n = n_positive) |>
      dplyr::pull(feature)

    top_features <- c(
      top_negative_features,
      top_positive_features
    )
  }

  top_features <- intersect(top_features, rownames(view_mat))

  if (length(top_features) == 0) {
    stop("No top features matched the rownames of the selected view matrix.")
  }

  metadata_mofa <- samples_metadata(MOFAobject) |>
    tibble::as_tibble() |>
    dplyr::mutate(
      sample = as.character(sample),
      C1_Core_group = factor(
        C1_Core_group,
        levels = c(
          "Not-C1-like",
          "Not-C1-intermediate",
          "C1-intermediate",
          "C1-like"
        )
      ),
      purity_annotation = dplyr::coalesce(
        as.numeric(purity_primary),
        as.numeric(purity_cpe),
        as.numeric(purity_absolute),
        as.numeric(purity_estimate),
        as.numeric(purity_ihc)
      )
    )

  sample_order <- metadata_mofa |>
    dplyr::filter(sample %in% colnames(view_mat)) |>
    dplyr::arrange(C1_Core_group, C1_Core) |>
    dplyr::pull(sample)

  if (length(sample_order) == 0) {
    stop("No MOFA samples matched the metadata sample column.")
  }

  heat_mat <- view_mat[top_features, sample_order, drop = FALSE]
  heat_mat_scaled <- row_zscore(heat_mat)

  annotation_df <- metadata_mofa |>
    dplyr::filter(sample %in% sample_order) |>
    dplyr::arrange(match(sample, sample_order)) |>
    dplyr::select(
      C1_Core_group,
      C1_Core,
      purity_annotation
    ) |>
    as.data.frame()

  colnames(annotation_df) <- c(
    "C1 group",
    "C1 score",
    "Purity"
  )

  rownames(annotation_df) <- sample_order

  group_colors <- c(
    "Not-C1-like" = "#4D4D4D",
    "Not-C1-intermediate" = "#A6A6A6",
    "C1-intermediate" = "#FDB863",
    "C1-like" = "#D73027"
  )

  c1_score_col_fun <- circlize::colorRamp2(
    c(
      min(annotation_df[["C1 score"]], na.rm = TRUE),
      0,
      max(annotation_df[["C1 score"]], na.rm = TRUE)
    ),
    c("#2166AC", "white", "#B2182B")
  )

  purity_col_fun <- circlize::colorRamp2(
    c(
      min(annotation_df[["Purity"]], na.rm = TRUE),
      median(annotation_df[["Purity"]], na.rm = TRUE),
      max(annotation_df[["Purity"]], na.rm = TRUE)
    ),
    c("#542788", "white", "#238443")
  )

  ha_col <- ComplexHeatmap::HeatmapAnnotation(
    df = annotation_df,
    col = list(
      "C1 group" = group_colors,
      "C1 score" = c1_score_col_fun,
      "Purity" = purity_col_fun
    ),
    annotation_name_gp = grid::gpar(fontsize = 9),
    na_col = "grey85"
  )

  ht <- ComplexHeatmap::Heatmap(
    heat_mat_scaled,
    name = "row z-score",
    col = circlize::colorRamp2(
      c(-2, 0, 2),
      viridis::viridis(3)
    ),
    use_raster = TRUE,
    top_annotation = ha_col,

    cluster_rows = if (cluster_rows) pearson_row_hclust else FALSE,
    cluster_columns = if (cluster_columns) pearson_column_hclust else FALSE,

    column_split = annotation_df[["C1 group"]],
    cluster_column_slices = FALSE,

    show_row_names = TRUE,
    show_column_names = FALSE,

    row_names_gp = grid::gpar(fontsize = 8),

    column_title = paste0(
      "MOFA Factor ", factor_id,
      ": ", feature_selection, " top ", length(top_features), " ",
      view_id, " features"
    ),
    column_title_gp = grid::gpar(fontsize = 12, fontface = "bold"),

    heatmap_legend_param = list(
      title = "Feature\nz-score"
    )
  )

  ComplexHeatmap::draw(
    ht,
    heatmap_legend_side = "right",
    annotation_legend_side = "right"
  )

  if (save_png) {
    png(
      filename = file.path(
        FIG_DIR,
        paste0(
          "MOFA_complexheatmap_",
          view_id,
          "_factor",
          factor_id,
          "_",
          feature_selection,
          "_top",
          length(top_features),
          "_C1_purity.png"
        )
      ),
      width = width,
      height = height,
      units = "in",
      res = FIG_DPI
    )

    ComplexHeatmap::draw(
      ht,
      heatmap_legend_side = "right",
      annotation_legend_side = "right"
    )

    dev.off()
  }

  invisible(ht)
}

```

### miRNA Heatmaps

```{r}
#| label: plot-mirna-mofa-heatmaps
#| echo: false
#| warning: false
#| message: false

view_id <- "miRNA"

plot_mofa_view_heatmap(
  MOFAobject = MOFAobject,
  factor_id = 1,
  view_id = view_id,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  clustering_method = "average"
)


plot_mofa_view_heatmap(
  MOFAobject = MOFAobject,
  factor_id = 10,
  view_id = view_id,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  clustering_method = "average"
)


```


### RNA Heatmaps

```{r}
#| label: plot-rna-mofa-heatmaps
#| echo: false
#| warning: false
#| message: false

view_id <- "RNA"

plot_mofa_view_heatmap(
  MOFAobject = MOFAobject,
  factor_id = 1,
  n_features = 100,
  view_id = view_id,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  clustering_method = "average"
)

plot_mofa_view_heatmap(
  MOFAobject = MOFAobject,
  factor_id = 10,
  view_id = view_id,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  clustering_method = "average"
)

```


### mutation Heatmaps

```{r}
#| label: plot-mutation-mofa-heatmaps
#| echo: false
#| warning: false
#| message: false

view_id <- "mutation"

plot_mofa_view_heatmap(
  MOFAobject = MOFAobject,
  factor_id = 1,
  view_id = view_id,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  clustering_method = "average"
)

plot_mofa_view_heatmap(
  MOFAobject = MOFAobject,
  factor_id = 10,
  view_id = view_id,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  clustering_method = "average"
)

```


### RPPA Heatmaps

```{r}
#| label: plot-RPPA-mofa-heatmaps
#| echo: false
#| warning: false
#| message: false

view_id <- "RPPA"

plot_mofa_view_heatmap(
  MOFAobject = MOFAobject,
  factor_id = 1,
  view_id = view_id,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  clustering_method = "average"
)


plot_mofa_view_heatmap(
  MOFAobject = MOFAobject,
  factor_id = 10,
  view_id = view_id,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  clustering_method = "average"
)


```

# Expansion Space

The following section anchors are intentionally empty and can be expanded later without disturbing the core MOFA workflow.

```{r}

library(MOFA2)
library(dplyr)
library(tibble)
library(readr)
library(matrixStats)

# Cleaning MOFA feature names
clean_mofa_feature_name <- function(x) {
  sub("_(RNA|RPPA|miRNA|mutation)$", "", x)
}


# Extracting MOFA weights and C1/Not-C1 medians for RNA, miRNA, and RPPA
extract_mofa_weights_with_group_medians <- function(
  MOFAobject,
  mofa_data,
  factor_id = 1,
  view_id = "RNA",
  feature_col = "feature",
  c1_group_col = "C1_Core_group",
  c1_label = "C1-like",
  not_c1_label = "Not-C1-like"
) {
  view_mat <- mofa_data[[view_id]]

  if (is.null(view_mat)) {
    stop("The selected view_id is not present in mofa_data.")
  }

  view_mat <- as.matrix(view_mat)
  storage.mode(view_mat) <- "numeric"

  metadata_tbl <- MOFA2::samples_metadata(MOFAobject) |>
    tibble::as_tibble() |>
    dplyr::mutate(sample = as.character(sample))

  if (!c1_group_col %in% colnames(metadata_tbl)) {
    stop("The selected C1 group column is not present in sample metadata.")
  }

  c1_samples <- metadata_tbl |>
    dplyr::filter(.data[[c1_group_col]] == c1_label) |>
    dplyr::pull(sample) |>
    intersect(colnames(view_mat))

  not_c1_samples <- metadata_tbl |>
    dplyr::filter(.data[[c1_group_col]] == not_c1_label) |>
    dplyr::pull(sample) |>
    intersect(colnames(view_mat))

  if (length(c1_samples) == 0) {
    stop("No C1-like samples matched the selected omics matrix.")
  }

  if (length(not_c1_samples) == 0) {
    stop("No Not-C1-like samples matched the selected omics matrix.")
  }

  weights_tbl <- MOFA2::get_weights(
    MOFAobject,
    views = view_id,
    factors = factor_id,
    as.data.frame = TRUE
  ) |>
    dplyr::filter(view == view_id) |>
    dplyr::transmute(
      feature_original = feature,
      feature = clean_mofa_feature_name(feature),
      weight = value
    )

  median_tbl <- tibble::tibble(
    feature = rownames(view_mat),
    median_C1 = matrixStats::rowMedians(
      view_mat[, c1_samples, drop = FALSE],
      na.rm = TRUE
    ),
    median_NotC1 = matrixStats::rowMedians(
      view_mat[, not_c1_samples, drop = FALSE],
      na.rm = TRUE
    )
  ) |>
    dplyr::mutate(
      delta_median_C1_minus_NotC1 = median_C1 - median_NotC1
    )

  weights_tbl |>
    dplyr::left_join(
      median_tbl,
      by = "feature"
    ) |>
    dplyr::mutate(
      agreement = dplyr::case_when(
        weight > 0 & delta_median_C1_minus_NotC1 > 0 ~ "Concordant",
        weight < 0 & delta_median_C1_minus_NotC1 < 0 ~ "Concordant",
        weight > 0 & delta_median_C1_minus_NotC1 < 0 ~ "Discordant",
        weight < 0 & delta_median_C1_minus_NotC1 > 0 ~ "Discordant",
        TRUE ~ "Neutral"
      ),
      association = dplyr::case_when(
        weight > 0 ~ "C1",
        weight < 0 ~ "NotC1",
        TRUE ~ "Neutral"
      ),
      agreement = factor(
        agreement,
        levels = c("Concordant", "Discordant", "Neutral")
      )
    ) |>
    dplyr::arrange(
      agreement,
      dplyr::desc(abs(weight))
    ) |>
    dplyr::select(
      feature_original,
      feature,
      weight,
      median_C1,
      median_NotC1,
      delta_median_C1_minus_NotC1,
      association,
      agreement
    ) |>
    dplyr::rename(
      !!feature_col := feature
    )
}


# Extracting MOFA mutation weights and C1/Not-C1 frequencies
extract_mofa_mutation_weights_with_group_frequency <- function(
  MOFAobject,
  mofa_data,
  factor_id = 1,
  view_id = "mutation",
  c1_group_col = "C1_Core_group",
  c1_label = "C1-like",
  not_c1_label = "Not-C1-like"
) {
  mutation_mat <- mofa_data[[view_id]]

  if (is.null(mutation_mat)) {
    stop("The selected mutation view is not present in mofa_data.")
  }

  mutation_mat <- as.matrix(mutation_mat)
  storage.mode(mutation_mat) <- "numeric"

  metadata_tbl <- MOFA2::samples_metadata(MOFAobject) |>
    tibble::as_tibble() |>
    dplyr::mutate(sample = as.character(sample))

  if (!c1_group_col %in% colnames(metadata_tbl)) {
    stop("The selected C1 group column is not present in sample metadata.")
  }

  c1_samples <- metadata_tbl |>
    dplyr::filter(.data[[c1_group_col]] == c1_label) |>
    dplyr::pull(sample) |>
    intersect(colnames(mutation_mat))

  not_c1_samples <- metadata_tbl |>
    dplyr::filter(.data[[c1_group_col]] == not_c1_label) |>
    dplyr::pull(sample) |>
    intersect(colnames(mutation_mat))

  if (length(c1_samples) == 0) {
    stop("No C1-like samples matched the mutation matrix.")
  }

  if (length(not_c1_samples) == 0) {
    stop("No Not-C1-like samples matched the mutation matrix.")
  }

  weights_tbl <- MOFA2::get_weights(
    MOFAobject,
    views = view_id,
    factors = factor_id,
    as.data.frame = TRUE
  ) |>
    dplyr::filter(view == view_id) |>
    dplyr::transmute(
      mutation_original = feature,
      mutation = clean_mofa_feature_name(feature),
      weight = value
    )

  frequency_tbl <- tibble::tibble(
    mutation = rownames(mutation_mat),
    frequency_C1 = matrixStats::rowMeans2(
      mutation_mat[, c1_samples, drop = FALSE] > 0,
      na.rm = TRUE
    ),
    frequency_NotC1 = matrixStats::rowMeans2(
      mutation_mat[, not_c1_samples, drop = FALSE] > 0,
      na.rm = TRUE
    )
  ) |>
    dplyr::mutate(
      delta_frequency_C1_minus_NotC1 = frequency_C1 - frequency_NotC1
    )

  weights_tbl |>
    dplyr::left_join(
      frequency_tbl,
      by = "mutation"
    ) |>
    dplyr::mutate(
      agreement = dplyr::case_when(
        weight > 0 & delta_frequency_C1_minus_NotC1 > 0 ~ "Concordant",
        weight < 0 & delta_frequency_C1_minus_NotC1 < 0 ~ "Concordant",
        weight > 0 & delta_frequency_C1_minus_NotC1 < 0 ~ "Discordant",
        weight < 0 & delta_frequency_C1_minus_NotC1 > 0 ~ "Discordant",
        TRUE ~ "Neutral"
      ),
      association = dplyr::case_when(
        weight > 0 ~ "C1",
        weight < 0 ~ "NotC1",
        TRUE ~ "Neutral"
      ),
      agreement = factor(
        agreement,
        levels = c("Concordant", "Discordant", "Neutral")
      )
    ) |>
    dplyr::arrange(
      agreement,
      dplyr::desc(abs(weight))
    ) |>
    dplyr::select(
      mutation_original,
      mutation,
      weight,
      frequency_C1,
      frequency_NotC1,
      delta_frequency_C1_minus_NotC1,
      association,
      agreement
    )
}

```

```{r}

TABLE_DIR <- "output/tables"
dir.create(TABLE_DIR, recursive = TRUE, showWarnings = FALSE)


# RNA
rna_factor1_weights_medians <- extract_mofa_weights_with_group_medians(
  MOFAobject = MOFAobject,
  mofa_data = mofa_data,
  factor_id = 1,
  view_id = "RNA",
  feature_col = "gene"
)

readr::write_tsv(
  rna_factor1_weights_medians,
  file.path(TABLE_DIR, "MOFA_factor1_RNA_weights_C1_NotC1_medians.tsv")
)


# miRNA
mirna_factor1_weights_medians <- extract_mofa_weights_with_group_medians(
  MOFAobject = MOFAobject,
  mofa_data = mofa_data,
  factor_id = 1,
  view_id = "miRNA",
  feature_col = "miRNA"
)

readr::write_tsv(
  mirna_factor1_weights_medians,
  file.path(TABLE_DIR, "MOFA_factor1_miRNA_weights_C1_NotC1_medians.tsv")
)


# RPPA
protein_factor1_weights_medians <- extract_mofa_weights_with_group_medians(
  MOFAobject = MOFAobject,
  mofa_data = mofa_data,
  factor_id = 1,
  view_id = "RPPA",
  feature_col = "protein"
)

readr::write_tsv(
  protein_factor1_weights_medians,
  file.path(TABLE_DIR, "MOFA_factor1_RPPA_weights_C1_NotC1_medians.tsv")
)


# Mutation
mutation_factor1_weights_frequency <- extract_mofa_mutation_weights_with_group_frequency(
  MOFAobject = MOFAobject,
  mofa_data = mofa_data,
  factor_id = 1,
  view_id = "mutation"
)

readr::write_tsv(
  mutation_factor1_weights_frequency,
  file.path(TABLE_DIR, "MOFA_factor1_mutation_weights_C1_NotC1_frequency.tsv")
)

```

```{r}

rna_factor1_weights_medians |>
  dplyr::filter(agreement == "Concordant") |>
  dplyr::filter(weight > 0) |>
  dplyr::slice_max(abs(weight), n = 50)

```

# Notes

- Keep the factor heatmaps and group-difference exports in this notebook; they are central, not optional add-ons.
- Change `focus_score_column`, `focus_group_column`, and the group labels at the top when the project is not a C1-style analysis.
- Promote additional per-factor or per-view interpretation only after the core heatmap and difference summaries are stable.
