---
title: "TRENTO miRNA responder model"
author: "Thomas Sirchi"
date: today
date-format: "DD MMMM YYYY"

format-links:
  - pdf

format:
  html:
    output-file: "TRENTO_miRNA_responder_modeling.html"

    theme:
      light: cosmo
      dark: darkly
    respect-user-color-scheme: true

    # Navigation and section numbering.
    toc: true
    toc-title: "Contents"
    toc-depth: 4
    toc-location: right
    number-sections: true
    number-depth: 3
    smooth-scroll: true
    anchor-sections: true

    # Reader-friendly code display.
    code-fold: true
    code-summary: "Show analysis code"
    code-copy: true
    code-overflow: wrap

    # Figure and page layout.
    fig-responsive: true
    embed-resources: true
    page-layout: article
    grid:
      sidebar-width: 250px
      body-width: 1050px
      margin-width: 300px
      gutter-width: 1.5rem

    link-external-newwindow: true

    title-block-banner: true
    df-print: paged
    self-contained-math: true
    fig-align: center
    fig-width: 9
    fig-height: 6
    fig-format: retina
    fig-dpi: 300
    fig-cap-location: bottom
    tbl-cap-location: top
    callout-appearance: simple
    keep-md: false
    keep-ipynb: false

  pdf:
    output-file: "TRENTO_miRNA_responder_modeling.pdf"

    pdf-engine: xelatex

    toc: true
    toc-title: "Contents"
    toc-depth: 3

    number-sections: true
    number-depth: 3

    documentclass: article
    fontsize: 10pt
    geometry:
      - margin=1in

    fig-align: center
    fig-width: 7
    fig-height: 5
    fig-format: pdf
    fig-dpi: 300
    fig-cap-location: bottom

    tbl-cap-location: top

    colorlinks: true

    keep-tex: false
    keep-md: false
    keep-ipynb: false

execute:
  echo: true
  warning: false
  message: false
  error: false

# Click figures to inspect them at full size.
lightbox:
  match: auto
  effect: fade
  desc-position: bottom
  loop: false

# Sensible defaults for knitr figures and code blocks.
knitr:
  opts_chunk:
    fig.align: center
    fig.width: 16
    fig.height: 12
    dpi: 300
    out.width: "100%"
    collapse: true
    comment: "#>"

editor: source
---


```{r}
#| echo: false
#| message: false
#| warning: false
#| label: setup
#| include: false
library(tidyverse)
library(tidymodels)
library(vip)
library(themis)
library(tune)
library(finetune)
library(filtro)
library(important)
library(desirability2)
library(brglm2)
library(bundle)
library(GGally)

tidymodels_prefer()

RANDOM_SEED <- 1234
set.seed(RANDOM_SEED)
theme_set(theme_minimal(base_size = 12))

dir.create("output/tables", recursive = TRUE, showWarnings = FALSE)
dir.create("output/figures", recursive = TRUE, showWarnings = FALSE)
dir.create("output/models", recursive = TRUE, showWarnings = FALSE)

extract_selected_features <- function(x) {

  selected_predictors <- workflows::extract_mold(x)$predictors |>
    names() |>
    intersect(predictor_names)

  tibble::tibble(
    predictor = selected_predictors
  )
}

```

## Overview

This report fits a binary miRNA classifier for TRENTO responder status. It uses the fixed split, the union signature, supervised top-10 selection, and bias-reduced logistic regression. The threshold is tuned by Monte Carlo cross-validation and evaluated on the held-out test set.

It covers the data split, model performance, feature stability, selected-miRNA expression, and sample-level results.

## Data and split

The notebook joins the fixed split labels with filtered miRNA expression and keeps shared samples. Predictor names are made model-safe; original miRNA names are retained for reporting.

The test set is held out from resampling, feature selection, threshold tuning, and fitting.

```{r}
#| message: false
#| warning: false
#| label: load-data
#| echo: false
#| include: false

set.seed(RANDOM_SEED)

# --- Load metadata ------------------------------------------------------------

split_tbl <- read_tsv(
  "input/all_samples_train_test_split_big_test.tsv",
  show_col_types = FALSE
)

# --- Load signature -----------------------------------------------------------

signature_mirnas <- read_tsv(
  "input/mirna_union_signature.tsv",
  show_col_types = FALSE
) |>
  pull(1)

# --- Prepare miRNA expression -------------------------------------------------

mirna_expr <- read_csv(
  "input/TRENTO_Full_miRNAseq_LogCPM_filtered_ready.csv",
  show_col_types = FALSE
) |>
  rename(miRNA = 1) |>
  filter(miRNA %in% signature_mirnas) |>
  pivot_longer(
    -miRNA,
    names_to = "sample_id",
    values_to = "expression"
  ) |>
  pivot_wider(
    names_from = miRNA,
    values_from = expression
  )

# Make miRNA names model-safe
mirna_cols <- setdiff(names(mirna_expr), "sample_id")
predictor_names <- make.names(mirna_cols, unique = TRUE)

predictor_map <- tibble(
  original_name = mirna_cols,
  predictor = predictor_names
)

names(mirna_expr)[-1] <- predictor_names

# --- Build modelling dataset --------------------------------------------------

model_data <- split_tbl |>
  transmute(
    sample_id = sample,
    outcome = responder_2class,
    train_test
  ) |>
  inner_join(mirna_expr, by = "sample_id") |>
  drop_na(train_test, outcome) |>
  mutate(
    outcome = factor(
      outcome,
      levels = c("responder", "non-responder")
    )
  )

train_data <- model_data |>
  filter(train_test == "train") |>
  select(-train_test)

test_data <- model_data |>
  filter(train_test == "test") |>
  select(-train_test)

# --- Class weights ------------------------------------------------------------

weight_alpha <- 0.5

class_weight_tbl <- train_data |>
  count(outcome) |>
  mutate(
    class_weight = (sum(n) / (n_distinct(outcome) * n)) ^ weight_alpha,
    class_weight = class_weight /
      weighted.mean(class_weight, n)
  ) |>
  select(outcome, class_weight)

train_data <- train_data |>
  left_join(class_weight_tbl, by = "outcome") |>
  mutate(
    class_weight = hardhat::importance_weights(class_weight)
  )

positive_class <- levels(train_data$outcome)[1]

```

### Class balance

Class counts are shown before modeling. Because the classes are imbalanced, balanced accuracy, sensitivity, specificity, ROC AUC, and PR AUC are more informative than accuracy alone.

```{r}
#| message: false
#| warning: false
#| label: tbl-class-balance
#| echo: false
#| tbl-cap: "Class counts and proportions in the training and held-out test sets."

set.seed(RANDOM_SEED)

bind_rows(
  train_data |> count(outcome, name = "n") |> mutate(set = "train"),
  test_data |> count(outcome, name = "n") |> mutate(set = "test")
) |>
  select(set, outcome, n) |>
  group_by(set) |>
  mutate(prop = n / sum(n)) |>
  ungroup()

class_weight_tbl

```

## Model

The recipe removes zero-variance predictors, selects the top 10 miRNAs with supervised desirability scores, and normalizes them within each resample. All preprocessing is learned from training data.

### Cross-validation

Monte Carlo cross-validation uses 200 stratified resamples, with 70% of the training data in each analysis set. Tuning uses these folds only; the test set remains untouched. The resamples estimate performance variability and feature-selection stability; they do not add patients.

The fitted preprocessing is applied unchanged to validation and test data.

### Class weights

Class weights give the smaller class more influence during training without changing outcome labels. Test metrics use the observed class counts.

```{r}
#| message: false
#| warning: false
#| label: tbl-model-training-balance
#| echo: false
#| tbl-cap: "Training-class counts after preprocessing."

set.seed(RANDOM_SEED)

# Desirability function for supervised miRNA filtering.
# Here we keep predictors that have high univariate ROC AUC and high ANOVA signal.
filter_goals <- desirability(
  maximize(roc_auc),
  maximize(aov_fstat),
)

# There are 45 miRNA predictors. Keep only the top 10 miRNA predictors inside each CV fold.
top_n_predictors <- 10
filter_prop <- top_n_predictors / length(predictor_names)

model_recipe <- recipe(outcome ~ ., data = train_data) |>
  update_role(sample_id, new_role = "id") |>
  # Remove predictors with zero variance.
  step_zv(all_predictors()) |>
  # Supervised filter. This is estimated separately inside each resampling fold. With filter_prop = 10 / n_predictors, only the top 10 miRNA is retained.
  step_predictor_desirability(
    all_predictors(),
    score = filter_goals,
    prop_terms = filter_prop
  ) |>
  # Normalize predictors after filtering and resampling.
  step_normalize(all_predictors())

# Check the final training balance after preprocessing.
training <- model_recipe |>
  prep() |>
  bake(new_data = NULL) |>
  count(outcome, name = "training")

training

# Monte Carlo CV lets you choose a more realistic train/assessment ratio repeatedly.
cv_folds <- mc_cv(
  train_data,
  prop = 0.7,
  times = 200,
  strata = outcome
)

# Metrics used during model tuning.
# Do not use precision for tuning here because it can become NA when no responders are predicted.
model_metrics <- metric_set(
  bal_accuracy,
  kap,
  f_meas,
  roc_auc,
  accuracy,
  sens,
  spec,
  recall,
  mcc
)

```

## Model fit

The final model is a bias-reduced logistic regression using `brglm2::brglm_fit`. Only the probability threshold is tuned; the recipe and model specification stay fixed. Coefficients are saved and plotted.

```{r}
#| message: false
#| warning: false
#| label: glm-fit
#| echo: false
#| include: false

set.seed(RANDOM_SEED)
br_logreg_spec <- logistic_reg() |>
  set_engine("glm", family = stats::binomial(), method = brglm2::brglm_fit, type = "AS_mean") |>
  set_mode("classification")

# Tailor postprocessor
# This tunes the probability cutoff used to convert probabilities into classes
br_logreg_tailor <- tailor() |>
  #adjust_probability_calibration(method = "isotonic") |>
  adjust_probability_threshold(threshold = tune())

br_logreg_workflow <- workflow() |>
  add_recipe(model_recipe) |>
  add_model(br_logreg_spec) |>
  add_tailor(br_logreg_tailor)

# Tune only the probability threshold for the bias-reduced logistic regression model.
br_logreg_grid <- tibble(
  threshold = seq(0.45, 0.80, by = 0.05)
)

br_logreg_tune <- tune_grid(
  br_logreg_workflow,
  resamples = cv_folds,
  grid = br_logreg_grid,
  metrics = model_metrics,
  control = control_grid(
    save_pred = TRUE,
    extract = extract_selected_features
  )
)

best_br_logreg <- select_best(br_logreg_tune, metric = "bal_accuracy")

br_logreg_fit <- finalize_workflow(br_logreg_workflow, best_br_logreg) |>
  fit(data = train_data)

br_logreg_fit_bundle <- bundle(br_logreg_fit)
saveRDS(br_logreg_fit_bundle, "output/models/br_logreg_fit_bundle.rds")

```

```{r}
#| message: false
#| warning: false
#| label: fig-glm-coefficients
#| echo: false
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "Coefficient summaries for the primary bias-reduced logistic-regression model."

set.seed(RANDOM_SEED)

collect_metrics(br_logreg_tune) |>
  filter(.metric %in% c("roc_auc", "bal_accuracy", "sens", "spec", "f_meas")) |>
  arrange(.metric, desc(mean))

br_logreg_coefficients <- extract_fit_parsnip(br_logreg_fit)$fit |>
  broom::tidy() |>
  filter(term != "(Intercept)", estimate != 0) |>
  left_join(predictor_map, by = c("term" = "predictor")) |>
  mutate(term = coalesce(original_name, term)) |>
  arrange(desc(abs(estimate)))

readr::write_tsv(br_logreg_coefficients, "output/tables/br_logreg_coefficients.tsv")

br_logreg_coef_plot <- br_logreg_coefficients |>
  slice_max(abs(estimate), n = 20) |>
  mutate(term = fct_reorder(term, estimate)) |>
  ggplot(aes(x = estimate, y = term, fill = estimate > 0)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("#D95F02", "#1B9E77")) +
  labs(
    title = "Model coefficients",
    x = "Coefficient",
    y = NULL
  )

ggsave("output/figures/br_logreg_coefficients.png", br_logreg_coef_plot, width = 7, height = 5, dpi = 300)

br_logreg_coef_plot

br_logreg_coef_plot2 <- ggcoef(
  extract_fit_parsnip(br_logreg_fit),
  exponentiate = TRUE,
  exclude_intercept = TRUE,
  conf.int = TRUE,
  sort = "ascending",
  color = "darkgreen"
) +
  labs(
    title = "Model coefficients (odds ratios)",
    subtitle = "OR > 1 = higher odds of response",
    x = "Odds ratio",
    y = NULL
  ) +
  theme_bw(base_size = 12)

br_logreg_coef_plot2

ggsave(
  "output/figures/br_logreg_coef_plot2.png",
  plot = br_logreg_coef_plot2,
  width = 10,
  height = 8
)

```

## Test performance

The finalized model is evaluated on held-out test samples using the tuned threshold. Predictions and threshold-dependent metrics are saved.

ROC AUC summarizes ranking; PR AUC is useful with class imbalance; sensitivity, specificity, and balanced accuracy describe threshold performance.

```{r}
#| message: false
#| warning: false
#| label: glm-test-evaluation
#| echo: false

set.seed(RANDOM_SEED)

# Predict probabilities and classes on the independent test set.
# The class prediction uses the threshold selected by tailor.
br_logreg_test_predictions <- test_data |>
  select(sample_id, outcome) |>
  bind_cols(
    predict(br_logreg_fit, new_data = test_data, type = "prob"),
    predict(br_logreg_fit, new_data = test_data, type = "class")
  ) |>
  mutate(
    truth = factor(outcome, levels = c("responder", "non-responder")),
    .pred_class = factor(.pred_class, levels = c("responder", "non-responder"))
  )

br_logreg_test_predictions |>
  ggplot(aes(x = .pred_responder, fill = truth)) +
  geom_density(alpha = 0.4) +
  geom_vline(xintercept = best_br_logreg$threshold, linetype = "dashed") +
  labs(
    title = "Predicted responder probabilities",
    x = "Predicted probability of responder",
    y = "Density"
  )

# Test-set metrics
br_logreg_test_metrics <- model_metrics(
  br_logreg_test_predictions,
  truth = truth,
  estimate = .pred_class,
  .pred_responder,
  event_level = "first"
) |>
  mutate(
    model = "bias-reduced logistic regression",
    evaluation = paste0("test set; threshold = ", round(best_br_logreg$threshold, 3))
  )

# Test-set confusion matrix
br_logreg_test_conf_mat <- br_logreg_test_predictions |>
  conf_mat(
    truth = truth,
    estimate = .pred_class
  )

# Save outputs
readr::write_tsv(
  br_logreg_test_predictions,
  "output/tables/br_logreg_test_predictions.tsv"
)

readr::write_tsv(
  br_logreg_test_metrics,
  "output/tables/br_logreg_test_metrics.tsv"
)

br_logreg_test_metrics
br_logreg_test_conf_mat

```


## Confusion matrix

The confusion matrix shows correct and incorrect test-set calls at the tuned threshold. Counts and the heatmap are saved for the misclassification review.

```{r}
#| message: false
#| warning: false
#| label: fig-glm-test-confusion
#| echo: false
#| fig-width: 7
#| fig-height: 6
#| fig-cap: "Confusion matrix for the primary model on the independent test set at the tuned classification threshold."

set.seed(RANDOM_SEED)

# Confusion matrix
br_logreg_conf_mat <- br_logreg_test_predictions |>
  yardstick::conf_mat(
    truth = truth,
    estimate = .pred_class
  )

br_logreg_conf_mat

# Confusion matrix plot
br_logreg_conf_mat_plot <- br_logreg_conf_mat |>
  autoplot(type = "heatmap") +
  labs(
    title = "Test-set confusion matrix",
    subtitle = paste0(
      "Independent test set; threshold = ",
      round(best_br_logreg$threshold, 3)
    ),
    x = "Predicted class",
    y = "True class"
  ) +
  scale_fill_gradient(low = "white", high = "#3B528B")

ggsave(
  "output/figures/br_logreg_test_confusion_matrix.png",
  plot = br_logreg_conf_mat_plot,
  width = 7,
  height = 6,
  dpi = 300
)

br_logreg_conf_mat_plot

# Save confusion matrix counts
br_logreg_conf_mat_table <- br_logreg_test_predictions |>
  count(truth, prediction = .pred_class, name = "n")

readr::write_tsv(
  br_logreg_conf_mat_table,
  "output/tables/br_logreg_test_confusion_matrix.tsv"
)
```

## Misclassified samples

The TSV lists misclassified held-out samples with predictions and metadata. The compact GT table highlights key clinical and immune fields.

```{r}
#| message: false
#| warning: false
#| label: tbl-glm-misclassified-metadata
#| echo: false
#| tbl-cap: "Clinical and immune metadata for misclassified held-out test samples."

br_logreg_metadata <- readr::read_csv(
  "input/TRENTO_metadata_with_ifng_scores.csv",
  show_col_types = FALSE
) |>
  dplyr::rename(sample_id = sample)

br_logreg_misclassified <- br_logreg_test_predictions |>
  filter(truth != .pred_class) |>
  arrange(desc(.pred_responder)) |>
  left_join(br_logreg_metadata, by = "sample_id") |>
  select(
    sample_id,
    truth,
    .pred_responder,
    .pred_class,
    .pred_non_responder = `.pred_non-responder`,
    everything()
  )

readr::write_tsv(
  br_logreg_misclassified,
  "output/tables/br_logreg_misclassified.tsv"
)

br_logreg_misclassified_main <- br_logreg_misclassified |>
  transmute(
    sample_id,
    truth = as.character(truth),
    predicted_class = as.character(.pred_class),
    responder_probability = .pred_responder,
    histology = histology_2class,
    pdl1 = dplyr::case_when(
      pdl1_class == "PD-L1 high (≥50%)" ~ "High (>=50%)",
      pdl1_class == "PD-L1 low/intermediate (1–49%)" ~ "Low/intermediate (1-49%)",
      pdl1_class == "PD-L1 negative (<1%)" ~ "Negative (<1%)",
      TRUE ~ pdl1_class
    ),
    pdl1_percent,
    cd274 = cd274_group,
    ifng = Signature_IFNG_class,
    age = age_diagnosis,
    sex,
    smoking = smoking_2class,
    treatment = treatment_type
  )

br_logreg_misclassified_gt <- br_logreg_misclassified_main |>
  gt::gt(id = "br_logreg_misclassified") |>
  gtExtras::gt_theme_538() |>
  gt::tab_header(
    title = gt::md("**Misclassified test samples**"),
    subtitle = gt::md(
      "Clinical and immune metadata"
    )
  ) |>
  gt::cols_label(
    sample_id = "Sample",
    truth = "True class",
    predicted_class = "Predicted class",
    responder_probability = "P(responder)",
    histology = "Histology",
    pdl1 = "PD-L1",
    pdl1_percent = "PD-L1 (%)",
    cd274 = "CD274",
    ifng = "IFNG",
    age = "Age",
    sex = "Sex",
    smoking = "Smoking",
    treatment = "Treatment"
  ) |>
  gt::fmt_number(columns = pdl1_percent, decimals = 1) |>
  gt::fmt_number(columns = responder_probability, decimals = 3) |>
  gt::cols_align(align = "center", columns = c(
    truth, predicted_class, responder_probability, pdl1_percent, age, sex
  )) |>
  gt::tab_options(
    table.font.size = gt::px(10),
    data_row.padding = gt::px(4),
    column_labels.font.weight = "bold",
    heading.title.font.size = gt::px(16),
    heading.subtitle.font.size = gt::px(11)
  ) |>
  gt::tab_source_note(
    source_note = gt::md(
      "**Note:** only test-set samples with truth != predicted class are shown."
    )
  )

br_logreg_misclassified_gt

if (!knitr::is_latex_output()) {
  br_logreg_misclassified_gt <- br_logreg_misclassified_gt |>
    gt::data_color(
      columns = responder_probability,
      fn = scales::col_numeric(
        palette = c("#E8F5E9", "#FFCDD2"),
        domain = c(0, 1)
      )
    ) |>
    gtExtras::gt_plt_bar(
      column = responder_probability,
      keep_column = TRUE,
      scale_type = "number",
      width = 60
    )
}

if (knitr::is_latex_output()) {
  br_logreg_misclassified_pdf_gt <- br_logreg_misclassified_gt |>
    gt::cols_hide(columns = c(
      pdl1_percent, cd274, ifng, age, sex, smoking, treatment
    ))
  knitr::asis_output(gt::as_latex(br_logreg_misclassified_pdf_gt))
} else {
  br_logreg_misclassified_gt
  gt::gtsave(
    br_logreg_misclassified_gt,
    "output/tables/br_logreg_misclassified_main_metadata.html"
  )
}
```

### ROC and precision-recall curves

ROC-AUC summarizes ranking discrimination across classification thresholds. PR-AUC summarizes precision versus recall for responders and should be read against the responder-prevalence baseline. Figure @fig-glm-test-curves shows both views for the independent test set.

```{r}
#| message: false
#| warning: false
#| label: fig-glm-test-curves
#| echo: false
#| fig-width: 14
#| fig-height: 6
#| fig-cap: "ROC and precision-recall curves for the primary bias-reduced logistic-regression model on the independent test set. The dashed line in the precision-recall panel marks responder prevalence."

set.seed(RANDOM_SEED)

# Test-set ROC AUC
br_logreg_roc_auc_value <- br_logreg_test_predictions |>
  yardstick::roc_auc(
    truth = truth,
    .pred_responder,
    event_level = "first"
  ) |>
  pull(.estimate)

# ROC curve
br_logreg_roc_plot <- br_logreg_test_predictions |>
  yardstick::roc_curve(
    truth = truth,
    .pred_responder,
    event_level = "first"
  ) |>
  autoplot() +
  labs(
    title = "Test-set ROC curve",
    subtitle = paste0(
      "AUC = ", sprintf("%.2f", br_logreg_roc_auc_value),
      "; threshold = ", sprintf("%.2f", best_br_logreg$threshold)
    ),
    x = "1 - Specificity",
    y = "Sensitivity"
  )

ggsave(
  "output/figures/br_logreg_test_roc_curve.png",
  plot = br_logreg_roc_plot,
  width = 7,
  height = 6,
  dpi = 300
)

# Test-set PR AUC
br_logreg_pr_auc_value <- br_logreg_test_predictions |>
  yardstick::pr_auc(
    truth = truth,
    .pred_responder,
    event_level = "first"
  ) |>
  pull(.estimate)

# Random PR baseline = responder prevalence in the test set
br_logreg_random_pr_baseline <- mean(
  br_logreg_test_predictions$truth == "responder",
  na.rm = TRUE
)

# PR curve
br_logreg_pr_plot <- br_logreg_test_predictions |>
  yardstick::pr_curve(
    truth = truth,
    .pred_responder,
    event_level = "first"
  ) |>
  autoplot() +
  geom_hline(
    yintercept = br_logreg_random_pr_baseline,
    linetype = "dashed",
    linewidth = 0.8
  ) +
  labs(
    title = "Test-set precision-recall curve",
    subtitle = paste0(
      "AUC = ", sprintf("%.2f", br_logreg_pr_auc_value),
      "; threshold = ", sprintf("%.2f", best_br_logreg$threshold),
      "; random baseline = ", sprintf("%.2f", br_logreg_random_pr_baseline)
    ),
    x = "Recall / Sensitivity",
    y = "Precision"
  )

ggsave(
  "output/figures/br_logreg_test_pr_curve.png",
  plot = br_logreg_pr_plot,
  width = 7,
  height = 6,
  dpi = 300
)

br_logreg_test_curves_patchwork <- patchwork::wrap_plots(
  br_logreg_roc_plot,
  br_logreg_pr_plot,
  ncol = 2
)

ggsave(
  "output/figures/br_logreg_test_roc_pr_patchwork.png",
  plot = br_logreg_test_curves_patchwork,
  width = 14,
  height = 6,
  dpi = 300
)

br_logreg_test_curves_patchwork
```

## Feature stability

Feature stability is measured across the 200 Monte Carlo resamples. Each resample selects 10 miRNAs with supervised filtering; selection frequency is the proportion of resamples containing each miRNA. Final-model features are matched to these frequencies.

Features selected in at least 70% of resamples are highly stable, 50–70% are stable, and fewer than 50% are unstable. Frequency reflects selection robustness, not effect size or significance.

```{r}
#| label: fig-br-logreg-feature-stability
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-height: 8
#| fig-cap: "Selection frequency of final-model miRNAs across Monte Carlo cross-validation resamples."

br_logreg_extracts <- collect_extracts(br_logreg_tune)

# Keep one feature-selection result per Monte Carlo resample.
# All threshold configurations contain the same 10 selected predictors.
br_logreg_selected_features <- br_logreg_extracts |>
  dplyr::select(id, .extracts) |>
  dplyr::distinct(id, .keep_all = TRUE) |>
  tidyr::unnest(.extracts)

br_logreg_selected_features

n_resamples <- dplyr::n_distinct(br_logreg_selected_features$id)

br_logreg_stability_tbl <- br_logreg_selected_features |>
  dplyr::group_by(predictor) |>
  dplyr::summarise(
    n_selected = dplyr::n_distinct(id),
    selection_frequency = n_selected / n_resamples,
    .groups = "drop"
  ) |>
  dplyr::left_join(
    predictor_map,
    by = "predictor"
  ) |>
  dplyr::mutate(
    miRNA = dplyr::coalesce(original_name, predictor),
    stability_class = dplyr::case_when(
      selection_frequency >= 0.70 ~ "Highly stable",
      selection_frequency >= 0.50 ~ "Stable",
      TRUE ~ "Unstable"
    )
  ) |>
  dplyr::select(
    miRNA,
    predictor,
    n_selected,
    selection_frequency,
    stability_class
  ) |>
  dplyr::arrange(
    dplyr::desc(selection_frequency),
    miRNA
  )

br_logreg_stability_tbl

readr::write_tsv(
  br_logreg_stability_tbl,
  "output/tables/br_logreg_feature_stability.tsv"
)

# Extract the miRNAs retained in the final fitted model.
br_logreg_best_features <- workflows::extract_mold(
  br_logreg_fit
)$predictors |>
  names() |>
  intersect(predictor_names)

br_logreg_best_features

br_logreg_best_feature_stability <- tibble::tibble(
  predictor = br_logreg_best_features
) |>
  dplyr::left_join(
    br_logreg_stability_tbl,
    by = "predictor"
  ) |>
  dplyr::arrange(
    dplyr::desc(selection_frequency)
  )

br_logreg_best_feature_stability

readr::write_tsv(
  br_logreg_best_feature_stability,
  "output/tables/br_logreg_best_model_feature_stability.tsv"
)

br_logreg_stability_plot <- br_logreg_best_feature_stability |>
  dplyr::slice_max(
    order_by = selection_frequency,
    n = 20,
    with_ties = FALSE
  ) |>
  dplyr::mutate(
    miRNA = forcats::fct_reorder(miRNA, selection_frequency)
  ) |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = selection_frequency,
      y = miRNA,
      fill = stability_class
    )
  ) +
  ggsci::scale_fill_npg() +
  ggplot2::geom_col(width = 0.7) +
  ggplot2::geom_vline(
    xintercept = 0.50,
    linetype = "dashed"
  ) +
  ggplot2::scale_x_continuous(
    limits = c(0, 1),
    labels = scales::percent_format(accuracy = 1)
  ) +
  ggplot2::labs(
    title = "Feature stability",
    subtitle = paste0(
      "Selection frequency across ",
      n_resamples,
      " Monte Carlo CV resamples | dashed line at 50% selection"
    ),
    x = "Selection frequency",
    y = NULL,
    fill = "Selection frequency"
  ) +
  ggplot2::theme_bw(base_size = 12) +
  ggplot2::theme(
    panel.grid.major.y = ggplot2::element_blank()
  )

br_logreg_stability_plot

ggplot2::ggsave(
  "output/figures/br_logreg_feature_stability.png",
  plot = br_logreg_stability_plot,
  width = 9,
  height = 7,
  dpi = 300
)

```

## Selected miRNA expression

This raincloud plot shows expression of the 10 final-model miRNAs by responder class.

```{r}
#| label: fig-br-logreg-model-mirna-expression-raincloud
#| echo: false
#| message: false
#| warning: false
#| fig-width: 11
#| fig-height: 9
#| fig-cap: "Expression of the final-model miRNAs by responder class across all modeled samples."

br_logreg_raincloud_features <- workflows::extract_mold(
  br_logreg_fit
)$predictors |>
  names() |>
  intersect(predictor_names)

stopifnot(length(br_logreg_raincloud_features) == top_n_predictors)

br_logreg_model_expression_long <- model_data |>
  dplyr::select(
    sample_id,
    outcome,
    dplyr::all_of(br_logreg_raincloud_features)
  ) |>
  tidyr::pivot_longer(
    cols = dplyr::all_of(br_logreg_raincloud_features),
    names_to = "predictor",
    values_to = "expression"
  ) |>
  dplyr::left_join(predictor_map, by = "predictor") |>
  dplyr::mutate(
    miRNA = factor(
      original_name,
      levels = predictor_map$original_name[
        match(br_logreg_raincloud_features, predictor_map$predictor)
      ]
    ),
    outcome = factor(
      outcome,
      levels = c("responder", "non-responder")
    )
  )

stopifnot(
  dplyr::n_distinct(br_logreg_model_expression_long$miRNA) == top_n_predictors,
  !anyNA(br_logreg_model_expression_long$expression)
)

br_logreg_model_mirna_expression_raincloud <- ggplot2::ggplot(
  br_logreg_model_expression_long,
  ggplot2::aes(
    x = outcome,
    y = expression,
    fill = outcome,
    color = outcome
  )
) +
  ggrain::geom_rain(
    rain.side = "r",
    alpha = 0.6
  ) +
  ggplot2::facet_wrap(
    ~ miRNA,
    scales = "free_y",
    ncol = 2
  ) +
  ggsci::scale_fill_npg() +
  ggsci::scale_color_npg() +
  ggplot2::labs(
    title = "Expression of final-model miRNAs",
    x = NULL,
    y = "Expression (logCPM)",
    fill = NULL,
    color = NULL
  ) +
  ggpubr::stat_compare_means(label.x.npc = "center", label.y.npc = "bottom", size = 3) +
  ggplot2::theme_bw(base_size = 12) +
  ggplot2::theme(
    legend.position = "top"
  )

ggplot2::ggsave(
  "output/figures/br_logreg_model_miRNA_expression_raincloud.png",
  plot = br_logreg_model_mirna_expression_raincloud,
  width = 16,
  height = 12,
  dpi = 300
)

br_logreg_model_mirna_expression_raincloud
```

## Final sample summary

This summary applies the final model to all samples in `model_data`. It records split, truth, predictions, probabilities, threshold, metadata, and final-model miRNA expression. Complete and compact TSVs are saved.

```{r}
#| label: br-logreg-train-test-summary-data
#| echo: false
#| include: false
#| message: false
#| warning: false

br_logreg_prediction_input <- model_data |>
  dplyr::select(sample_id, dplyr::all_of(predictor_names))

br_logreg_train_test_model_data <- model_data |>
  dplyr::select(sample_id, outcome, train_test) |>
  dplyr::bind_cols(
    predict(
      br_logreg_fit,
      new_data = br_logreg_prediction_input,
      type = "prob"
    ),
    predict(
      br_logreg_fit,
      new_data = br_logreg_prediction_input,
      type = "class"
    )
  ) |>
  dplyr::transmute(
    sample_id,
    train_test,
    split = dplyr::if_else(train_test == "test", "Test", "Train"),
    model = "bias-reduced logistic regression",
    true_class = as.character(outcome),
    predicted_class = as.character(.pred_class),
    responder_probability = .pred_responder,
    non_responder_probability = `.pred_non-responder`,
    model_threshold = best_br_logreg$threshold
  ) |>
  dplyr::arrange(
    factor(train_test, levels = c("test", "train")),
    sample_id
  )

br_logreg_model_mirnas <- mirna_expr |>
  dplyr::select(sample_id, dplyr::all_of(br_logreg_best_features)) |>
  dplyr::rename_with(
    ~ paste0(
      "miRNA_",
      predictor_map$original_name[
        match(.x, predictor_map$predictor)
      ]
    ),
    dplyr::all_of(br_logreg_best_features)
  )

br_logreg_model_mirna_cols <- setdiff(
  names(br_logreg_model_mirnas),
  "sample_id"
)

stopifnot(
  length(br_logreg_model_mirna_cols) == length(br_logreg_best_features),
  !anyNA(br_logreg_model_mirna_cols),
  !anyDuplicated(br_logreg_model_mirna_cols)
)

br_logreg_summary_metadata <- br_logreg_metadata |>
  dplyr::select(sample_id, dplyr::everything())

br_logreg_train_test_complete <- br_logreg_train_test_model_data |>
  dplyr::left_join(br_logreg_summary_metadata, by = "sample_id") |>
  dplyr::left_join(br_logreg_model_mirnas, by = "sample_id")

essential_metadata_cols <- c(
  "histology_2class",
  "pdl1_class",
  "pdl1_percent",
  "cd274_group",
  "Signature_IFNG_class",
  "age_diagnosis",
  "sex",
  "smoking_2class",
  "treatment_type",
  "ecog_group"
)

br_logreg_train_test_essential <- br_logreg_train_test_model_data |>
  dplyr::left_join(
    br_logreg_summary_metadata |>
      dplyr::select(
        sample_id,
        dplyr::any_of(essential_metadata_cols)
      ),
    by = "sample_id"
  ) |>
  dplyr::left_join(br_logreg_model_mirnas, by = "sample_id")

stopifnot(
  nrow(br_logreg_train_test_complete) == nrow(br_logreg_train_test_model_data),
  nrow(br_logreg_train_test_essential) == nrow(br_logreg_train_test_model_data),
  identical(
    names(br_logreg_train_test_complete)[
      (ncol(br_logreg_train_test_complete) - length(br_logreg_model_mirna_cols) + 1):
        ncol(br_logreg_train_test_complete)
    ],
    br_logreg_model_mirna_cols
  ),
  identical(
    names(br_logreg_train_test_essential)[
      (ncol(br_logreg_train_test_essential) - length(br_logreg_model_mirna_cols) + 1):
        ncol(br_logreg_train_test_essential)
    ],
    br_logreg_model_mirna_cols
  )
)

readr::write_tsv(
  br_logreg_train_test_complete,
  "output/tables/br_logreg_train_test_complete.tsv"
)

readr::write_tsv(
  br_logreg_train_test_essential,
  "output/tables/br_logreg_train_test_essential.tsv"
)
```

The GT table groups train and test samples. Training calls are in-sample; test calls are held out. The PDF uses a compact view, while the TSV and HTML retain the full values.

```{r}
#| label: tbl-br-logreg-train-test-summary
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Final-model predictions, selected-miRNA expression, and essential metadata by split."

br_logreg_model_gt_cols <- c(
  "train_test",
  "model",
  "true_class",
  "predicted_class",
  "responder_probability",
  "non_responder_probability",
  "model_threshold"
)

br_logreg_metadata_gt_cols <- intersect(
  essential_metadata_cols,
  names(br_logreg_train_test_essential)
)

br_logreg_train_test_gt <- br_logreg_train_test_essential |>
  gt::gt(
    groupname_col = "split",
    id = "br_logreg_train_test_summary"
  ) |>
  gtExtras::gt_theme_538() |>
  gt::tab_header(
    title = gt::md("**Final sample summary**"),
    subtitle = gt::md(
      "Predictions and selected-miRNA expression"
    )
  ) |>
  gt::cols_label(
    sample_id = "Sample",
    train_test = "Split",
    model = "Model",
    true_class = "True class",
    predicted_class = "Predicted class",
    responder_probability = "P(responder)",
    non_responder_probability = "P(non-responder)",
    model_threshold = "Threshold",
    histology_2class = "Histology",
    pdl1_class = "PD-L1",
    pdl1_percent = "PD-L1 (%)",
    cd274_group = "CD274",
    Signature_IFNG_class = "IFNG",
    age_diagnosis = "Age",
    sex = "Sex",
    smoking_2class = "Smoking",
    treatment_type = "Treatment",
    ecog_group = "ECOG"
  ) |>
  gt::tab_spanner(
    label = "Model data",
    columns = dplyr::all_of(br_logreg_model_gt_cols)
  ) |>
  gt::tab_spanner(
    label = "Essential metadata",
    columns = dplyr::all_of(br_logreg_metadata_gt_cols)
  ) |>
  gt::tab_spanner(
    label = "Final-model miRNA expression",
    columns = dplyr::all_of(br_logreg_model_mirna_cols)
  ) |>
  gt::fmt_number(
    columns = dplyr::all_of(c(
      "responder_probability",
      "non_responder_probability",
      "model_threshold",
      "pdl1_percent",
      br_logreg_model_mirna_cols
    )),
    decimals = 3
  ) |>
  gt::cols_align(
    align = "center",
    columns = dplyr::all_of(c(
      "train_test",
      "true_class",
      "predicted_class",
      "responder_probability",
      "non_responder_probability",
      "model_threshold",
      br_logreg_model_mirna_cols
    ))
  ) |>
  gt::tab_style(
    style = list(
      gt::cell_fill(color = "#E8F1FB"),
      gt::cell_text(weight = "bold")
    ),
    locations = gt::cells_row_groups(groups = "Test")
  ) |>
  gt::tab_style(
    style = list(
      gt::cell_fill(color = "#EAF5EA"),
      gt::cell_text(weight = "bold")
    ),
    locations = gt::cells_row_groups(groups = "Train")
  ) |>
  gt::tab_options(
    table.font.size = gt::px(9),
    data_row.padding = gt::px(3),
    column_labels.font.weight = "bold",
    row_group.font.weight = "bold",
    heading.title.font.size = gt::px(16),
    heading.subtitle.font.size = gt::px(11)
  )

br_logreg_train_test_gt

if (knitr::is_latex_output()) {
  br_logreg_train_test_pdf_gt <- br_logreg_train_test_gt |>
    gt::cols_hide(
      columns = dplyr::all_of(c(
        "model",
        "non_responder_probability",
        "model_threshold",
        br_logreg_metadata_gt_cols,
        br_logreg_model_mirna_cols
      ))
    ) |>
    gt::tab_header(
      title = gt::md("**Final sample summary**"),
      subtitle = gt::md(
        "Compact PDF view; full values are in the TSV and HTML table"
      )
    )
  knitr::asis_output(gt::as_latex(br_logreg_train_test_pdf_gt))
} else {
  br_logreg_train_test_gt <- br_logreg_train_test_gt |>
    gt::tab_source_note(
      source_note = gt::md(
        "**Note:** train rows are in-sample evaluations; test rows are independent held-out evaluations."
      )
    )
  br_logreg_train_test_gt
  gt::gtsave(
    br_logreg_train_test_gt,
    "output/tables/br_logreg_train_test_summary.html"
  )
}


```

## Top 4 stable miRNA model

This secondary analysis retains only the four miRNAs with the highest selection frequency in the training-derived stability table. The primary 10-miRNA model above is unchanged. The reduced model uses the same class weights, Monte Carlo resamples, bias-reduced logistic-regression specification, threshold grid, and held-out test set.

```{r}
#| label: top4-stable-mirna-model
#| echo: false
#| message: false
#| warning: false

top4_stable_mirnas <- br_logreg_stability_tbl |>
  dplyr::filter(stability_class != "Unstable") |>
  dplyr::arrange(dplyr::desc(selection_frequency), miRNA) |>
  dplyr::slice_head(n = 4)

stopifnot(nrow(top4_stable_mirnas) == 4L)

top4_predictors <- top4_stable_mirnas$predictor

top4_stable_mirnas

readr::write_tsv(
  top4_stable_mirnas,
  "output/tables/br_logreg_top4_stable_mirnas.tsv"
)

top4_recipe <- recipe(outcome ~ ., data = train_data) |>
  update_role(sample_id, new_role = "id") |>
  step_rm(dplyr::all_of(setdiff(predictor_names, top4_predictors))) |>
  step_zv(all_predictors()) |>
  step_normalize(all_predictors())

top4_workflow <- workflow() |>
  add_recipe(top4_recipe) |>
  add_model(br_logreg_spec) |>
  add_tailor(br_logreg_tailor)

top4_tune <- tune_grid(
  top4_workflow,
  resamples = cv_folds,
  grid = br_logreg_grid,
  metrics = model_metrics,
  control = control_grid(save_pred = TRUE)
)

top4_best <- select_best(top4_tune, metric = "bal_accuracy")

top4_fit <- finalize_workflow(top4_workflow, top4_best) |>
  fit(data = train_data)

top4_fit_bundle <- bundle(top4_fit)
saveRDS(
  top4_fit_bundle,
  "output/models/br_logreg_top4_fit_bundle.rds"
)

top4_test_predictions <- test_data |>
  select(sample_id, outcome) |>
  bind_cols(
    predict(top4_fit, new_data = test_data, type = "prob"),
    predict(top4_fit, new_data = test_data, type = "class")
  ) |>
  mutate(
    truth = factor(outcome, levels = c("responder", "non-responder")),
    .pred_class = factor(.pred_class, levels = c("responder", "non-responder"))
  )

top4_test_metrics <- bind_rows(
  model_metrics(
    top4_test_predictions,
    truth = truth,
    estimate = .pred_class,
    .pred_responder,
    event_level = "first"
  ),
  yardstick::precision(
    top4_test_predictions,
    truth = truth,
    estimate = .pred_class,
    event_level = "first"
  ),
  yardstick::npv(
    top4_test_predictions,
    truth = truth,
    estimate = .pred_class,
    event_level = "first"
  ),
  yardstick::j_index(
    top4_test_predictions,
    truth = truth,
    estimate = .pred_class,
    event_level = "first"
  ),
  yardstick::pr_auc(
    top4_test_predictions,
    truth = truth,
    .pred_responder,
    event_level = "first"
  )
) |>
  mutate(
    model = "top 4 stable-miRNA bias-reduced logistic regression",
    evaluation = paste0("test set; threshold = ", round(top4_best$threshold, 3))
  )

readr::write_tsv(
  top4_test_predictions,
  "output/tables/br_logreg_top4_test_predictions.tsv"
)

readr::write_tsv(
  top4_test_metrics,
  "output/tables/br_logreg_top4_test_metrics.tsv"
)

top4_test_metrics |>
  select(.metric, .estimate, model, evaluation)
```

### Top 4 test ROC and precision–recall curves

Figure @fig-top4-test-curves provides a brief held-out performance view for the reduced model. The ROC and precision–recall calculations use the unchanged test samples and responder probability predictions from `top4_fit`.

```{r}
#| label: fig-top4-test-curves
#| echo: false
#| message: false
#| warning: false
#| fig-width: 14
#| fig-height: 6
#| fig-cap: "ROC and precision-recall curves for the top-4 stable-miRNA model on the independent test set. The dashed line in the precision-recall panel marks responder prevalence."

top4_roc_auc_value <- top4_test_metrics |>
  filter(.metric == "roc_auc") |>
  pull(.estimate)

top4_roc_plot <- top4_test_predictions |>
  yardstick::roc_curve(
    truth = truth,
    .pred_responder,
    event_level = "first"
  ) |>
  autoplot() +
  labs(
    title = "Test ROC",
    subtitle = paste0("AUC = ", sprintf("%.2f", top4_roc_auc_value)),
    x = "1 - Specificity",
    y = "Sensitivity"
  )

ggsave(
  "output/figures/br_logreg_top4_test_roc_curve.png",
  plot = top4_roc_plot,
  width = 7,
  height = 6,
  dpi = 300
)

top4_pr_auc_value <- top4_test_metrics |>
  filter(.metric == "pr_auc") |>
  pull(.estimate)

top4_random_pr_baseline <- mean(
  top4_test_predictions$truth == "responder",
  na.rm = TRUE
)

top4_pr_plot <- top4_test_predictions |>
  yardstick::pr_curve(
    truth = truth,
    .pred_responder,
    event_level = "first"
  ) |>
  autoplot() +
  geom_hline(
    yintercept = top4_random_pr_baseline,
    linetype = "dashed",
    linewidth = 0.8
  ) +
  labs(
    title = "Test precision-recall",
    subtitle = paste0(
      "AUC = ", sprintf("%.2f", top4_pr_auc_value),
      "; random baseline = ", sprintf("%.2f", top4_random_pr_baseline)
    ),
    x = "Recall / Sensitivity",
    y = "Precision"
  )

ggsave(
  "output/figures/br_logreg_top4_test_pr_curve.png",
  plot = top4_pr_plot,
  width = 7,
  height = 6,
  dpi = 300
)

top4_test_curves_patchwork <- patchwork::wrap_plots(
  top4_roc_plot,
  top4_pr_plot,
  ncol = 2
)

ggsave(
  "output/figures/br_logreg_top4_test_roc_pr_patchwork.png",
  plot = top4_test_curves_patchwork,
  width = 14,
  height = 6,
  dpi = 300
)

top4_test_curves_patchwork
```

The corresponding ROC-AUC and PR-AUC values are listed below.

```{r}
#| label: tbl-top4-test-curves
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "ROC-AUC and PR-AUC for the top-4 stable-miRNA model on the independent test set."

top4_test_metrics |>
  filter(.metric %in% c("roc_auc", "pr_auc")) |>
  select(.metric, .estimate)
```

### Top 4 test confusion matrix

The confusion matrix reports the held-out test-set counts at the threshold selected by balanced accuracy during training resampling.

```{r}
#| label: fig-top4-test-confusion
#| echo: false
#| message: false
#| warning: false
#| fig-width: 7
#| fig-height: 6
#| fig-cap: "Confusion matrix for the top-4 stable-miRNA model on the independent test set."

top4_test_conf_mat <- top4_test_predictions |>
  yardstick::conf_mat(
    truth = truth,
    estimate = .pred_class
  )

top4_test_conf_mat_plot <- top4_test_conf_mat |>
  autoplot(type = "heatmap") +
  labs(
    title = "Top 4 stable-miRNA model: test confusion matrix",
    subtitle = paste0(
      "Independent test set; threshold = ",
      round(top4_best$threshold, 3)
    ),
    x = "Predicted class",
    y = "True class"
  ) +
  scale_fill_gradient(low = "white", high = "#3B528B")

ggsave(
  "output/figures/br_logreg_top4_test_confusion_matrix.png",
  plot = top4_test_conf_mat_plot,
  width = 7,
  height = 6,
  dpi = 300
)

top4_test_conf_mat_table <- top4_test_predictions |>
  count(truth, prediction = .pred_class, name = "n")

readr::write_tsv(
  top4_test_conf_mat_table,
  "output/tables/br_logreg_top4_test_confusion_matrix.tsv"
)

top4_test_conf_mat
top4_test_conf_mat_plot
```


## Query and print information about the current R session
```{r}

sessioninfo::session_info()

```
