---
title: "MAP 00 ATAC - Signac QC"
format: html
---

# Overview

## Aim

This notebook processes the independent scATAC-seq peak matrices for the five CSS samples, runs explicit per-sample Signac/fragtk QC, writes QC plots and CSV summaries, and saves one filtered ATAC object per sample. Consensus peak quantification, dimensional reduction, integration, and label transfer are left inactive for a later notebook pass.

## Setup

This section loads packages and sets shared execution options for the QC run.
The global future plan is useful for plotting and package internals, but
expensive fragment requantification is kept out of the active QC path.

```{r include=FALSE}

packages <- c(
  "tidyverse",
  "Seurat",
  "SeuratWrappers",
  "harmony",
  "patchwork",
  "data.table",
  "qs2",
  "future",
  "Signac",
  "GenomicRanges",
  "GenomeInfoDb",
  "EnsDb.Hsapiens.v86",
  "AnnotationDbi"
)

library(Seurat)
library(scCustomize)
library(ggplot2)

purrr::walk(packages, library, character.only = TRUE)

options(future.globals.maxSize = 25 * 1024^3)
plan(multicore, workers = 5, maxSizeOfObjects = 25 * 1024^3)

umap_theme <- theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 11, color = "grey30"),

    # Remove background grid
    panel.grid = element_blank(),

    # Keep axis titles
    axis.title = element_text(size = 13),

    # Remove axis numbers, ticks, and lines
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),

    # Legend formatting
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10)
  )

```

# Per-Sample ATAC QC

## Configure Inputs and Outputs

```{r}

set.seed(1234)

paths <- list(
  input_dir = "input/GSE274934_RAW",
  output_dir = "output/scATAC",
  object_dir = "output/scATAC/objects",
  figure_dir = "output/scATAC/figures",
  table_dir = "output/scATAC/tables",
  atac_plot_dir = "output/scATAC/figures/atac_qc_plots",
  atac_qc_dir = "output/scATAC/tables/atac_qc_metrics",
  rna_object_dir = "output/scRNA/objects"
)

# Cell-level QC thresholds used by every sample. These thresholds are written
# into each per-sample summary CSV so downstream runs can be audited.
atac_params <- list(
  min_peak_counts = 1000,
  min_passed_filters = 3000,
  min_pct_reads_in_peaks = 25,
  max_blacklist_ratio = 0.05,
  max_nucleosome_signal = 1.5,
  min_tss_enrichment = 6
)

# Create every output directory used by this QC notebook before helper execution.
dir.create(paths$output_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(paths$object_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(paths$figure_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(paths$table_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(paths$atac_plot_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(paths$atac_qc_dir, showWarnings = FALSE, recursive = TRUE)

samples <- c("AL05", "AL08", "AL10", "AL11", "AL12")

# Input file maps are keyed by sample name. The QC helper uses these maps to
# keep explicit per-sample execution concise without hiding sample identity.
atac_files <- c(
  AL05 = "GSM8462143_AL05_ATAC_filtered_peak_bc_matrix.h5",
  AL08 = "GSM8462144_AL08_ATAC_filtered_peak_bc_matrix.h5",
  AL10 = "GSM8462145_AL10_ATAC_filtered_peak_bc_matrix.h5",
  AL11 = "GSM8462146_AL11_ATAC_filtered_peak_bc_matrix.h5",
  AL12 = "GSM8462147_AL12_ATAC_filtered_peak_bc_matrix.h5"
)

atac_peak_files <- c(
  AL05 = "GSM8462143_AL05_ATAC_peaks.bed.gz",
  AL08 = "GSM8462144_AL08_ATAC_peaks.bed.gz",
  AL10 = "GSM8462145_AL10_ATAC_peaks.bed.gz",
  AL11 = "GSM8462146_AL11_ATAC_peaks.bed.gz",
  AL12 = "GSM8462147_AL12_ATAC_peaks.bed.gz"
)

frag_files <- c(
  AL05 = "GSM8462143_AL05_ATAC_fragments.tsv.gz",
  AL08 = "GSM8462144_AL08_ATAC_fragments.tsv.gz",
  AL10 = "GSM8462145_AL10_ATAC_fragments.tsv.gz",
  AL11 = "GSM8462146_AL11_ATAC_fragments.tsv.gz",
  AL12 = "GSM8462147_AL12_ATAC_fragments.tsv.gz"
)

atac_singlecell_files <- c(
  AL05 = "GSM8462143_AL05_ATAC_singlecell.csv.gz",
  AL08 = "GSM8462144_AL08_ATAC_singlecell.csv.gz",
  AL10 = "GSM8462145_AL10_ATAC_singlecell.csv.gz",
  AL11 = "GSM8462146_AL11_ATAC_singlecell.csv.gz",
  AL12 = "GSM8462147_AL12_ATAC_singlecell.csv.gz"
)

### Gene annotation helper

build_hg38_gene_annotations <- function() {
  
  annotations <- GetGRangesFromEnsDb(ensdb = EnsDb.Hsapiens.v86)
  GenomeInfoDb::seqlevelsStyle(annotations) <- "UCSC"
  GenomeInfoDb::genome(annotations) <- "hg38"
  annotations
}

gene_annotations <- build_hg38_gene_annotations()

```

# scATAC QC

The ATAC files are processed as an independent modality because the GEX and ATAC barcodes have minimal exact overlap. This section performs per-sample cell QC only. Consensus peak construction and fragment requantification are intentionally deferred until the integration stage is redesigned.

## Define ATAC QC Helpers

### Numeric and Peak Helpers

These small helpers support ratio calculations and later inactive common-peak construction.

```{r}

### Numeric and peak helpers

# Safe division for Cell Ranger-derived ratios. Failed denominators are kept as
# NA so filtering can treat missing metrics explicitly.

safe_ratio <- function(numerator, denominator, scale = 1) {
  ifelse(
    is.na(denominator) | denominator <= 0,
    NA_real_,
    scale * numerator / denominator
  )
}

read_peak_bed <- function(peak_path) {
  
  bed <- data.table::fread(peak_path, header = FALSE) |>
    as.data.frame()
  bed <- bed[, 1:3]
  colnames(bed) <- c("chrom", "start", "end")
  
  GenomicRanges::GRanges(
    seqnames = bed$chrom,
    ranges = IRanges::IRanges(start = bed$start + 1, end = bed$end)
  )
}

# Load a sample peak BED as GRanges. This is not used by the active QC run, but
# is available for the inactive downstream common-peak requantification section.
load_atac_peaks <- function(sample_name, paths) {
  
  peaks <- read_peak_bed(file.path(paths$input_dir, atac_peak_files[[sample_name]]))
  GenomeInfoDb::keepStandardChromosomes(
    peaks,
    pruning.mode = "coarse"
  )
}

# Build a reduced union peak set for later Signac-style merging. The active QC
# workflow does not call this helper.
build_unified_atac_peaks <- function(atac_peak_list) {
  
  cat("Building unified ATAC peak set...\n")
  
  combined_peaks <- GenomicRanges::GRangesList(atac_peak_list) |>
    unlist(use.names = FALSE)
  combined_peaks <- GenomicRanges::reduce(combined_peaks)
  combined_peaks <- sort(combined_peaks)
  
  cat("Unified ATAC peaks:", length(combined_peaks), "\n")
  
  combined_peaks
}

```

### Cell Ranger ATAC Metadata

This helper reads the Cell Ranger single-cell metrics and aligns them to the cells present in the peak matrix.

```{r}

### Cell Ranger ATAC metadata helper

# Read the Cell Ranger single-cell metrics and align them to the input peak
# matrix columns before adding them to the Seurat object.

load_atac_metrics <- function(sample_name, paths, atac_counts) {
  
  singlecell_path <- file.path(paths$input_dir, atac_singlecell_files[[sample_name]])
  
  singlecell_metrics <- data.table::fread(singlecell_path) |>
    as.data.frame()
  
  singlecell_metrics <- singlecell_metrics[singlecell_metrics$barcode != "NO_BARCODE", , drop = FALSE]
  rownames(singlecell_metrics) <- singlecell_metrics$barcode
  
  singlecell_metrics$pct_reads_in_peaks <- safe_ratio(
    singlecell_metrics$peak_region_fragments,
    singlecell_metrics$passed_filters,
    scale = 100
  )
  singlecell_metrics$pct_reads_in_DNase <- safe_ratio(
    singlecell_metrics$DNase_sensitive_region_fragments,
    singlecell_metrics$passed_filters,
    scale = 100
  )
  singlecell_metrics$blacklist_ratio <- safe_ratio(
    singlecell_metrics$blacklist_region_fragments,
    singlecell_metrics$peak_region_fragments
  )
  singlecell_metrics$promoter_ratio <- safe_ratio(
    singlecell_metrics$promoter_region_fragments,
    singlecell_metrics$passed_filters
  )
  singlecell_metrics$tss_fragment_ratio <- safe_ratio(
    singlecell_metrics$TSS_fragments,
    singlecell_metrics$passed_filters
  )
  
  matched_metrics <- singlecell_metrics[colnames(atac_counts), , drop = FALSE]
  matched_metrics$nCount_ATAC <- Matrix::colSums(atac_counts)
  matched_metrics$nFeature_ATAC <- Matrix::colSums(atac_counts > 0)
  matched_metrics$sample <- sample_name
  
  matched_metrics
}

```

### Metadata Cleanup

These helpers normalize duplicated metadata names from Cell Ranger and Signac, then remove exact duplicate or uninformative columns.

```{r}

### Metadata cleanup helpers

# Metadata comes from both Cell Ranger and Signac::ATACqc(). These helpers keep
# canonical QC column names stable and prevent duplicate/constant columns from
# cluttering plots, summaries, and exported CSVs.

metadata_columns_equal <- function(x, y) {
  isTRUE(all.equal(x, y, check.attributes = FALSE))
}

rename_atac_metadata_columns <- function(meta, sample_name = NULL, verbose = TRUE) {
  
  rename_map <- c(
    Nucleosome_signal = "nucleosome_signal",
    TSS_enrichment = "TSS.enrichment",
    is__cell_barcode = "is_cell_barcode",
    FRiP = "frip_signac"
  )
  
  for (old_name in names(rename_map)) {
    new_name <- unname(rename_map[[old_name]])
    
    if (old_name %in% colnames(meta) && new_name %in% colnames(meta)) {
      if (!metadata_columns_equal(meta[[old_name]], meta[[new_name]])) {
        stop(
          "Conflicting ATAC metadata columns",
          if (!is.null(sample_name)) paste0(" for ", sample_name) else "",
          ": ",
          old_name,
          " and ",
          new_name,
          " are both present but differ."
        )
      }
      
      if (verbose) {
        cat("Dropping duplicated ATAC metadata alias:", old_name, "\n")
      }
      
      meta <- meta[, setdiff(colnames(meta), old_name), drop = FALSE]
    } else if (old_name %in% colnames(meta)) {
      colnames(meta)[colnames(meta) == old_name] <- new_name
    }
  }
  
  meta
}

# Treat all-NA, single-value, and all-zero numeric columns as uninformative.
is_constant_metadata_column <- function(x) {
  x_no_na <- stats::na.omit(x)
  
  length(x_no_na) == 0 ||
    length(unique(x_no_na)) <= 1 ||
    (is.numeric(x_no_na) && all(x_no_na == 0))
}

# Drop exact duplicates only outside protected QC/provenance columns.
drop_exact_duplicate_metadata <- function(meta, protected_columns, verbose = TRUE) {
  
  drop_columns <- character()
  
  for (column_index in seq_along(colnames(meta))) {
    column <- colnames(meta)[[column_index]]
    
    if (column %in% protected_columns) {
      next
    }
    
    prior_columns <- setdiff(colnames(meta)[seq_len(column_index - 1)], drop_columns)
    
    for (prior_column in prior_columns) {
      if (metadata_columns_equal(meta[[column]], meta[[prior_column]])) {
        drop_columns <- c(drop_columns, column)
        break
      }
    }
  }
  
  drop_columns <- unique(drop_columns)
  
  if (length(drop_columns) > 0) {
    if (verbose) {
      cat("Dropping exact duplicate ATAC metadata columns:", paste(drop_columns, collapse = ", "), "\n")
    }
    
    meta <- meta[, setdiff(colnames(meta), drop_columns), drop = FALSE]
  }
  
  meta
}

clean_atac_metadata <- function(obj, sample_name = NULL, verbose = TRUE) {
  
  protected_columns <- c(
    "orig.ident",
    "sample",
    "pass_atac_qc",
    "is_cell_barcode",
    "excluded_reason",
    "nCount_ATAC",
    "nFeature_ATAC",
    "passed_filters",
    "pct_reads_in_peaks",
    "blacklist_ratio",
    "nucleosome_signal",
    "TSS.enrichment"
  )
  
  meta <- obj@meta.data
  meta <- rename_atac_metadata_columns(meta, sample_name = sample_name, verbose = verbose)
  meta <- drop_exact_duplicate_metadata(meta, protected_columns = protected_columns, verbose = verbose)
  
  candidate_columns <- setdiff(colnames(meta), protected_columns)
  constant_columns <- candidate_columns[
    vapply(
      candidate_columns,
      function(column) is_constant_metadata_column(meta[[column]]),
      logical(1)
    )
  ]
  
  if (length(constant_columns) > 0) {
    if (verbose) {
      cat("Dropping constant ATAC metadata columns:", paste(constant_columns, collapse = ", "), "\n")
    }
    
    meta <- meta[, setdiff(colnames(meta), constant_columns), drop = FALSE]
  }
  
  obj@meta.data <- meta
  obj
}

```

### QC Plot Helpers

The plot helpers are shared by pre-filter and post-filter QC plots. They skip missing or constant metrics cleanly and add threshold lines from `atac_params`.

```{r}

### QC plot helpers

# A metric needs more than one non-NA value to support violin or scatter plots.

metric_has_signal <- function(df, metric) {
  metric %in% colnames(df) &&
    length(unique(stats::na.omit(df[[metric]]))) > 1
}

# Generic scatter writer used by the individual QC plots below. Missing and
# constant metrics are skipped with a console message rather than failing the run.
save_atac_scatter <- function(qc_df, sample_name, stage, outdir, x, y, filename_suffix,
                              x_label, y_label, vline = NULL, hline = NULL) {
  
  required_metrics <- c(x, y)
  missing_metrics <- setdiff(required_metrics, colnames(qc_df))
  
  if (length(missing_metrics) > 0) {
    cat(
      "Skipping ATAC QC scatter for",
      sample_name,
      stage,
      filename_suffix,
      "because metrics are missing:",
      paste(missing_metrics, collapse = ", "),
      "\n"
    )
    return(invisible(NULL))
  }
  
  if (!metric_has_signal(qc_df, x) || !metric_has_signal(qc_df, y)) {
    cat(
      "Skipping ATAC QC scatter for",
      sample_name,
      stage,
      filename_suffix,
      "because one or both metrics are constant:",
      paste(required_metrics, collapse = ", "),
      "\n"
    )
    return(invisible(NULL))
  }
  
  p <- ggplot(qc_df, aes(x = .data[[x]], y = .data[[y]])) +
    geom_point(size = 0.4, alpha = 0.5) +
    theme_bw(base_size = 12) +
    labs(x = x_label, y = y_label)
  
  if (!is.null(vline)) {
    p <- p + geom_vline(xintercept = vline, linetype = "dashed", color = "firebrick")
  }
  
  if (!is.null(hline)) {
    p <- p + geom_hline(yintercept = hline, linetype = "dashed", color = "firebrick")
  }
  
  ggsave(
    file.path(outdir, paste0(sample_name, "_", stage, "_", filename_suffix, ".png")),
    p,
    width = 7,
    height = 6,
    dpi = 600
  )
  
  invisible(NULL)
}

# Write pre- and post-filter QC plots. Threshold lines use atac_params so each
# figure records the same cutoffs used by the filter.
plot_atac_qc <- function(obj, sample_name, stage, outdir, atac_params) {
  
  DefaultAssay(obj) <- "ATAC"
  
  atac_features <- intersect(
    c(
      "nCount_ATAC",
      "nFeature_ATAC",
      "passed_filters",
      "peak_region_fragments",
      "pct_reads_in_peaks",
      "pct_reads_in_DNase",
      "blacklist_ratio",
      "nucleosome_signal",
      "TSS.enrichment"
    ),
    colnames(obj@meta.data)
  )
  
  atac_vln_features <- atac_features[
    vapply(atac_features, function(feature) metric_has_signal(obj@meta.data, feature), logical(1))
  ]
  
  skipped_features <- setdiff(atac_features, atac_vln_features)
  if (length(skipped_features) > 0) {
    cat(
      "Skipping constant ATAC QC violin features for",
      sample_name,
      stage,
      ":",
      paste(skipped_features, collapse = ", "),
      "\n"
    )
  }
  
  old_plan <- future::plan()
  future::plan(future::sequential)
  on.exit(future::plan(old_plan), add = TRUE)
  
  if (length(atac_vln_features) > 0) {
    p_vln <- VlnPlot(
      obj,
      features = atac_vln_features,
      layer = "counts",
      pt.size = 0.1,
      ncol = 4
    )
    
    ggsave(
      file.path(outdir, paste0(sample_name, "_", stage, "_ATAC_vln.png")),
      p_vln,
      width = 16,
      height = 7,
      dpi = 600
    )
  }
  
  qc_df <- as.data.frame(obj@meta.data)
  colnames(qc_df) <- make.unique(colnames(qc_df))
  qc_df <- tibble::rownames_to_column(qc_df, "cell")

  density_x <- dplyr::case_when(
    "nCount_peaks" %in% colnames(obj@meta.data) ~ "nCount_peaks",
    "nCount_ATAC" %in% colnames(obj@meta.data) ~ "nCount_ATAC",
    TRUE ~ NA_character_
  )

  if (!is.na(density_x) && metric_has_signal(qc_df, density_x) && metric_has_signal(qc_df, "TSS.enrichment")) {
    p_tss_density <- DensityScatter(
      obj,
      x = density_x,
      y = "TSS.enrichment",
      log_x = TRUE,
      quantiles = TRUE
    ) +
      geom_hline(
        yintercept = atac_params$min_tss_enrichment,
        linetype = "dashed",
        color = "firebrick"
      ) +
      labs(x = density_x, y = "TSS enrichment")
    
    if (identical(density_x, "nCount_ATAC")) {
      p_tss_density <- p_tss_density +
        geom_vline(
          xintercept = atac_params$min_peak_counts,
          linetype = "dashed",
          color = "firebrick"
        )
    }

    ggsave(
      file.path(outdir, paste0(sample_name, "_", stage, "_ATAC_TSS_density_scatter.png")),
      p_tss_density,
      width = 7,
      height = 6,
      dpi = 600
    )
  } else {
    cat(
      "Skipping ATAC TSS density scatter for",
      sample_name,
      stage,
      "because required metrics are missing or constant.\n"
    )
  }
  
  save_atac_scatter(
    qc_df,
    sample_name,
    stage,
    outdir,
    x = "nCount_ATAC",
    y = "pct_reads_in_peaks",
    filename_suffix = "ATAC_frip_scatter",
    x_label = "ATAC peak counts",
    y_label = "Reads in peaks (%)",
    vline = atac_params$min_peak_counts,
    hline = atac_params$min_pct_reads_in_peaks
  )
  
  save_atac_scatter(
    qc_df,
    sample_name,
    stage,
    outdir,
    x = "nCount_ATAC",
    y = "passed_filters",
    filename_suffix = "ATAC_passed_filters_scatter",
    x_label = "ATAC peak counts",
    y_label = "Passed filter fragments",
    vline = atac_params$min_peak_counts,
    hline = atac_params$min_passed_filters
  )
  
  save_atac_scatter(
    qc_df,
    sample_name,
    stage,
    outdir,
    x = "nCount_ATAC",
    y = "blacklist_ratio",
    filename_suffix = "ATAC_blacklist_scatter",
    x_label = "ATAC peak counts",
    y_label = "Blacklist ratio",
    vline = atac_params$min_peak_counts,
    hline = atac_params$max_blacklist_ratio
  )
  
  save_atac_scatter(
    qc_df,
    sample_name,
    stage,
    outdir,
    x = "nCount_ATAC",
    y = "nucleosome_signal",
    filename_suffix = "ATAC_nucleosome_scatter",
    x_label = "ATAC peak counts",
    y_label = "Nucleosome signal",
    vline = atac_params$min_peak_counts,
    hline = atac_params$max_nucleosome_signal
  )
}

```

### Signac Fragtk QC

This helper wraps `Signac::ATACqc()` and adds the returned nucleosome and TSS metrics back to the Seurat metadata.

```{r}

### Signac fragtk QC helper

# Run Signac/fragtk QC on the raw sample ATAC object. This adds the
# nucleosome_signal and TSS.enrichment metrics used by qc_atac_sample().
run_signac_atacqc <- function(obj, sample_name, paths) {
  
  DefaultAssay(obj) <- "ATAC"
  qc_outdir <- file.path(paths$atac_qc_dir, paste0(sample_name, "_fragtk_qc"))
  dir.create(qc_outdir, showWarnings = FALSE, recursive = TRUE)
  
  fragtk_path <- Sys.which("fragtk")
  
  if (!nzchar(fragtk_path)) {
    stop(
      "fragtk not found on PATH. Install fragtk or add it to PATH before running ",
      "Signac::ATACqc() for ",
      sample_name,
      "."
    )
  }
  
  cat("Computing ATAC QC metrics with Signac::ATACqc:", sample_name, "\n")
  atac_qc_metadata <- Signac::ATACqc(
    object = obj[["ATAC"]],
    annotations = gene_annotations,
    fragtk.path = unname(fragtk_path),
    outdir = qc_outdir,
    cleanup = TRUE,
    verbose = TRUE
  )
  
  atac_qc_metadata <- as.data.frame(atac_qc_metadata)
  colnames(atac_qc_metadata) <- make.unique(colnames(atac_qc_metadata))
  atac_qc_metadata <- rename_atac_metadata_columns(
    atac_qc_metadata,
    sample_name = sample_name
  )
  
  required_atac_qc_metrics <- c("nucleosome_signal", "TSS.enrichment")
  missing_atac_qc_metrics <- setdiff(required_atac_qc_metrics, colnames(atac_qc_metadata))
  if (length(missing_atac_qc_metrics) > 0) {
    stop(
      "Signac::ATACqc() did not return required mapped metrics for ",
      sample_name,
      ": ",
      paste(missing_atac_qc_metrics, collapse = ", "),
      ". Returned columns: ",
      paste(colnames(atac_qc_metadata), collapse = ", ")
    )
  }
  
  matched_cells <- intersect(colnames(obj), rownames(atac_qc_metadata))
  if (length(matched_cells) == 0) {
    stop("No ATACqc cell barcodes matched the Seurat object for ", sample_name, ".")
  }
  
  if (length(matched_cells) < ncol(obj)) {
    cat(
      "ATACqc metrics matched",
      length(matched_cells),
      "of",
      ncol(obj),
      "ATAC cells for",
      sample_name,
      "\n"
    )
  }
  
  duplicated_metadata <- intersect(
    colnames(atac_qc_metadata),
    colnames(obj@meta.data)
  )
  
  if (length(duplicated_metadata) > 0) {
    cat(
      "Dropping duplicated ATACqc metadata columns:",
      paste(duplicated_metadata, collapse = ", "),
      "\n"
    )
    atac_qc_metadata <- atac_qc_metadata[
      ,
      setdiff(colnames(atac_qc_metadata), duplicated_metadata),
      drop = FALSE
      ]
  }
  
  atac_qc_full_metadata <- data.frame(row.names = colnames(obj))
  atac_qc_full_metadata[
    matched_cells,
    colnames(atac_qc_metadata)
  ] <- atac_qc_metadata[matched_cells, , drop = FALSE]
  
  obj <- AddMetaData(obj, metadata = atac_qc_full_metadata)
  obj
}

```

### Per-Sample QC

Active QC entry point. It writes pre-filter metadata, post-filter metadata, retained barcodes, summary CSVs, QC plots, and returns the filtered sample-specific object.

```{r}

### Per-sample ATAC QC helper

# Export cell-level metadata at defined checkpoints:
# - pre_filter includes all cells plus pass_atac_qc and excluded_reason
# - post_filter includes only retained cells
write_atac_metadata_csv <- function(obj, sample_name, stage, paths) {
  
  metadata_table <- as.data.frame(obj@meta.data)
  colnames(metadata_table) <- make.unique(colnames(metadata_table))
  metadata_table <- tibble::rownames_to_column(metadata_table, "cell")
  
  data.table::fwrite(
    metadata_table,
    file.path(paths$atac_qc_dir, paste0(sample_name, "_ATAC_cell_qc_metrics_", stage, ".csv"))
  )
  
  invisible(metadata_table)
}

# Convert per-metric failure flags into an auditable semicolon-separated reason.
# Retained cells are labelled "retained".
build_atac_excluded_reason <- function(qc_failures) {
  
  reasons <- apply(
    qc_failures,
    1,
    function(x) {
      failed_metrics <- names(x)[x %in% TRUE]
      
      if (length(failed_metrics) == 0) {
        "retained"
      } else {
        paste(failed_metrics, collapse = ";")
      }
    }
  )
  
  unname(reasons)
}

# Main active QC entry point. It filters cells in the original Cell Ranger peak
# space and returns a sample-specific filtered Seurat object. It does not build
# common peaks and does not call FeatureMatrix().
qc_atac_sample <- function(sample_name, paths, atac_params) {
  
  cat("\n============================\n")
  cat("Processing ATAC:", sample_name, "\n")
  cat("============================\n")
  
  atac_path <- file.path(paths$input_dir, atac_files[[sample_name]])
  fragment_path <- file.path(paths$input_dir, frag_files[[sample_name]])
  
  # Start from the original Cell Ranger peak matrix. Per-barcode metrics are
  # aligned to matrix columns before being added to metadata.
  raw_atac_counts <- Read10X_h5(atac_path)
  atac_metrics <- load_atac_metrics(sample_name, paths, raw_atac_counts)
  
  raw_chromatin_assay <- CreateChromatinAssay(
    counts = raw_atac_counts,
    sep = c(":", "-"),
    fragments = fragment_path,
    genome = "hg38",
    annotation = gene_annotations,
    min.cells = 1,
    min.features = 1,
    validate.fragments = FALSE
  )
  
  obj <- CreateSeuratObject(
    counts = raw_chromatin_assay,
    assay = "ATAC",
    project = sample_name
  )
  
  obj <- AddMetaData(obj, metadata = atac_metrics[colnames(obj), , drop = FALSE])
  obj$sample <- sample_name
  atac_cells_before <- ncol(obj)
  
  cat("ATAC cells before QC:", atac_cells_before, "\n")
  cat("Computing Signac ATAC QC metrics:", sample_name, "\n")
  
  # Signac::ATACqc adds nucleosome and TSS metrics used by the filter below.
  # The metadata cleanup step immediately normalizes names and removes clutter.
  options(future.globals.maxSize = 25 * 1024^3)
  obj <- run_signac_atacqc(obj, sample_name, paths)
  obj <- clean_atac_metadata(obj, sample_name = sample_name)
  
  required_atac_qc_metrics <- c("nucleosome_signal", "TSS.enrichment")
  missing_atac_qc_metrics <- setdiff(required_atac_qc_metrics, colnames(obj@meta.data))
  if (length(missing_atac_qc_metrics) > 0) {
    stop(
      "Missing required ATAC QC metrics after Signac QC for ",
      sample_name,
      ": ",
      paste(missing_atac_qc_metrics, collapse = ", ")
    )
  }
  
  # Build explicit failure flags first so rejected cells keep an interpretable
  # excluded_reason in the pre-filter CSV.
  qc_failures <- data.frame(
    low_peak_counts = is.na(obj$nCount_ATAC) | obj$nCount_ATAC < atac_params$min_peak_counts,
    low_passed_filters = is.na(obj$passed_filters) | obj$passed_filters < atac_params$min_passed_filters,
    low_pct_reads_in_peaks = is.na(obj$pct_reads_in_peaks) | obj$pct_reads_in_peaks < atac_params$min_pct_reads_in_peaks,
    high_blacklist_ratio = !is.na(obj$blacklist_ratio) & obj$blacklist_ratio > atac_params$max_blacklist_ratio,
    high_nucleosome_signal = is.na(obj$nucleosome_signal) | obj$nucleosome_signal > atac_params$max_nucleosome_signal,
    low_tss_enrichment = is.na(obj$TSS.enrichment) | obj$TSS.enrichment < atac_params$min_tss_enrichment,
    row.names = colnames(obj)
  )
  
  obj$pass_atac_qc <- rowSums(qc_failures) == 0
  obj$excluded_reason <- build_atac_excluded_reason(qc_failures)
  
  write_atac_metadata_csv(
    obj,
    sample_name = sample_name,
    stage = "pre_filter",
    paths = paths
  )
  
  plot_atac_qc(obj, sample_name, "preQC", paths$atac_plot_dir, atac_params)
  
  # From this point onward, only retained cells are carried into the returned
  # object and saved barcode list.
  qc_barcodes <- rownames(obj@meta.data)[obj$pass_atac_qc %in% TRUE]
  cat("ATAC cells after QC:", length(qc_barcodes), "\n")
  
  if (length(qc_barcodes) == 0) {
    stop(
      "No ATAC cells passed QC for ",
      sample_name,
      ". Review thresholds and pre-filter QC metrics before continuing."
    )
  }
  
  retained_barcodes <- data.frame(cell = qc_barcodes)
  data.table::fwrite(
    retained_barcodes,
    file.path(paths$atac_qc_dir, paste0(sample_name, "_ATAC_retained_barcodes.csv"))
  )
  
  obj_qc <- subset(obj, cells = qc_barcodes)
  
  obj_qc$sample <- sample_name
  obj_qc$pass_atac_qc <- TRUE
  obj_qc$excluded_reason <- "retained"
  obj_qc <- clean_atac_metadata(obj_qc, sample_name = sample_name)
  
  write_atac_metadata_csv(
    obj_qc,
    sample_name = sample_name,
    stage = "post_filter",
    paths = paths
  )
  
  plot_atac_qc(obj_qc, sample_name, "postQC", paths$atac_plot_dir, atac_params)
  
  # Summary statistics are written as CSV; the returned value remains Seurat only.
  # Thresholds are included so each summary is self-contained.
  summary <- data.frame(
    sample = sample_name,
    atac_cells_before = atac_cells_before,
    atac_cells_after_qc = ncol(obj_qc),
    pct_removed_atac_qc = round(100 * (atac_cells_before - ncol(obj_qc)) / atac_cells_before, 2),
    min_peak_counts = atac_params$min_peak_counts,
    min_passed_filters = atac_params$min_passed_filters,
    min_pct_reads_in_peaks = atac_params$min_pct_reads_in_peaks,
    max_blacklist_ratio = atac_params$max_blacklist_ratio,
    max_nucleosome_signal = atac_params$max_nucleosome_signal,
    min_tss_enrichment = atac_params$min_tss_enrichment,
    median_nCount_ATAC = median(obj_qc$nCount_ATAC, na.rm = TRUE),
    median_nFeature_ATAC = median(obj_qc$nFeature_ATAC, na.rm = TRUE),
    median_pct_reads_in_peaks = median(obj_qc$pct_reads_in_peaks, na.rm = TRUE),
    median_blacklist_ratio = median(obj_qc$blacklist_ratio, na.rm = TRUE),
    median_nucleosome_signal = median(obj_qc$nucleosome_signal, na.rm = TRUE),
    median_TSS_enrichment = median(obj_qc$TSS.enrichment, na.rm = TRUE)
  )
  
  data.table::fwrite(
    summary,
    file.path(paths$atac_qc_dir, paste0(sample_name, "_ATAC_summary.csv"))
  )
  
  obj_qc
}

```

### Downstream Integration Helpers

These helpers are defined here for later use, but are not called by the active QC section. They implement the Signac-style common-peak requantification before merging.

```{r}

### Downstream integration helpers

# Used only by the inactive downstream integration section. Per-sample QC does
# not call FeatureMatrix or requantify fragments. This follows the Signac merge
# pattern: requantify retained cells on the same common peak set before merging.
requantify_atac_qc_object <- function(obj, sample_name, paths, common_peaks,
                                      process_n = 2000) {
  
  cat("Requantifying retained ATAC cells on common peaks:", sample_name, "\n")
  
  retained_cells <- colnames(obj)
  fragment_path <- file.path(paths$input_dir, frag_files[[sample_name]])
  
  if (!file.exists(fragment_path)) {
    stop("Missing ATAC fragment file for ", sample_name, ": ", fragment_path)
  }
  
  fragment <- CreateFragmentObject(
    path = fragment_path,
    cells = retained_cells,
    validate.fragments = FALSE
  )
  
  # FeatureMatrix can be memory intensive. Keep this step sequential by default
  # so process_n has predictable memory behavior during later integration work.
  old_plan <- future::plan()
  future::plan(future::sequential)
  on.exit(future::plan(old_plan), add = TRUE)
  
  common_counts <- FeatureMatrix(
    fragments = fragment,
    features = common_peaks,
    cells = retained_cells,
    process_n = process_n
  )
  
  # Rebuild the ChromatinAssay with common peak rows. Metadata comes from the
  # already-filtered QC object, so no rejected cells are reintroduced here.
  common_assay <- CreateChromatinAssay(
    counts = common_counts,
    ranges = common_peaks,
    fragments = fragment,
    genome = "hg38",
    annotation = gene_annotations,
    min.cells = 0,
    min.features = 0,
    validate.fragments = FALSE
  )
  
  metadata <- obj@meta.data[colnames(common_counts), , drop = FALSE]
  
  obj_common <- CreateSeuratObject(
    counts = common_assay,
    assay = "ATAC",
    project = sample_name,
    meta.data = metadata
  )
  
  obj_common$sample <- sample_name
  obj_common
}

# Merge filtered QC objects only after each sample has been requantified on the
# same common peak set. If common_peaks is not provided, build a reduced union
# from the supplied sample peak BED files.
merge_atac_qc_objects <- function(..., paths, common_peaks = NULL, process_n = 2000) {
  
  atac_qc_list <- list(...)
  
  stopifnot(length(atac_qc_list) > 0)
  stopifnot(!is.null(names(atac_qc_list)))
  
  missing_names <- names(atac_qc_list) == "" | is.na(names(atac_qc_list))
  if (any(missing_names)) {
    stop("All ATAC QC objects passed to merge_atac_qc_objects() must be named by sample.")
  }
  
  if (is.null(common_peaks)) {
    # The common peak set is intentionally constructed here, not during QC.
    atac_peak_list <- lapply(
      names(atac_qc_list),
      load_atac_peaks,
      paths = paths
    )
    names(atac_peak_list) <- names(atac_qc_list)
    common_peaks <- build_unified_atac_peaks(atac_peak_list)
  }
  
  qs2::qs_save(
    common_peaks,
    file = file.path(paths$object_dir, "GSE274934_common_ATAC_peaks.qs2")
  )
  
  # Requantify every filtered sample object independently before merging.
  atac_common_list <- mapply(
    FUN = requantify_atac_qc_object,
    obj = atac_qc_list,
    sample_name = names(atac_qc_list),
    MoreArgs = list(
      paths = paths,
      common_peaks = common_peaks,
      process_n = process_n
    ),
    SIMPLIFY = FALSE
  )
  
  if (length(atac_common_list) == 1) {
    return(atac_common_list[[1]])
  }
  
  merge(
    atac_common_list[[1]],
    y = atac_common_list[-1],
    add.cell.ids = names(atac_common_list)
  )
}

```

### Summary Helpers

Read the per-sample summaries written by `qc_atac_sample()` and combine them into one QC review table.

```{r}

### Summary CSV helper

# Read per-sample summary files and write a single table for quick QC review.
read_all_atac_summaries <- function(sample_names, paths) {
  
  summary_files <- file.path(
    paths$atac_qc_dir,
    paste0(sample_names, "_ATAC_summary.csv")
  )
  
  names(summary_files) <- sample_names
  
  missing_summary_files <- summary_files[!file.exists(summary_files)]
  if (length(missing_summary_files) > 0) {
    stop(
      "Missing ATAC summary CSV files: ",
      paste(missing_summary_files, collapse = ", ")
    )
  }
  
  summary_df <- do.call(
    rbind,
    lapply(summary_files, data.table::fread)
  )
  
  data.table::fwrite(
    summary_df,
    file.path(paths$atac_qc_dir, "all_samples_ATAC_summary.csv")
  )
  
  summary_df
}

```

### Inactive Silhouette Diagnostic Helper

This diagnostic is reserved for the later integration pass. It is intentionally separated from the active QC helpers.

```{r}

### Inactive integration diagnostic helper

# Reserved for the later integration pass. Not called by the active QC workflow.
compute_silhouette_resolution <- function(
  object,
  cluster_col,
  embedding,
  n_per_cluster = 1000,
  seed = 1234
) {
  set.seed(seed)
  
  clusters <- object[[cluster_col]][, 1]
  clusters <- as.factor(clusters)
  names(clusters) <- colnames(object)
  
  if (length(levels(clusters)) < 2) {
    return(
      data.frame(
        resolution = cluster_col,
        n_clusters = length(levels(clusters)),
        n_cells_used = NA_integer_,
        mean_silhouette = NA_real_,
        median_silhouette = NA_real_,
        frac_negative = NA_real_
      )
    )
  }
  
  cells_use <- unlist(
    lapply(levels(clusters), function(cl) {
      cells_cl <- names(clusters)[clusters == cl]
      sample(cells_cl, min(length(cells_cl), n_per_cluster))
    })
  )
  
  emb_sub <- embedding[cells_use, , drop = FALSE]
  clusters_sub <- clusters[cells_use]
  
  sil <- silhouette(
    x = as.integer(clusters_sub),
    dist = dist(emb_sub)
  )
  
  sil_df <- as.data.frame(sil)
  
  data.frame(
    resolution = cluster_col,
    n_clusters = length(levels(clusters)),
    n_cells_used = length(cells_use),
    mean_silhouette = mean(sil_df$sil_width),
    median_silhouette = median(sil_df$sil_width),
    frac_negative = mean(sil_df$sil_width < 0)
  )
}

```

## Run ATAC QC Per Sample

Consensus peak construction and fragment requantification are not part of this QC pass. They will be planned with the downstream integration workflow.

```{r}

cat("Run modular ATAC QC per sample...\n")

```

### AL05 ATAC QC

```{r}

# Run one sample and save the filtered object immediately.
AL05_ATAC <- qc_atac_sample(
  sample_name = "AL05",
  paths = paths,
  atac_params = atac_params
)

# Memory checkpoint before serializing the object.
gc()

qs2::qs_save(
  AL05_ATAC,
  file = file.path(paths$object_dir, "AL05_ATAC_QC.qs2")
)

```

### AL08 ATAC QC

```{r}

# Run one sample and save the filtered object immediately.
AL08_ATAC <- qc_atac_sample(
  sample_name = "AL08",
  paths = paths,
  atac_params = atac_params
)

# Memory checkpoint before serializing the object.
gc()

qs2::qs_save(
  AL08_ATAC,
  file = file.path(paths$object_dir, "AL08_ATAC_QC.qs2")
)

```

### AL10 ATAC QC

```{r}

# Run one sample and save the filtered object immediately.
AL10_ATAC <- qc_atac_sample(
  sample_name = "AL10",
  paths = paths,
  atac_params = atac_params
)

qs2::qs_save(
  AL10_ATAC,
  file = file.path(paths$object_dir, "AL10_ATAC_QC.qs2")
)

```

### AL11 ATAC QC

```{r}

# Run one sample and save the filtered object immediately.
AL11_ATAC <- qc_atac_sample(
  sample_name = "AL11",
  paths = paths,
  atac_params = atac_params
)

qs2::qs_save(
  AL11_ATAC,
  file = file.path(paths$object_dir, "AL11_ATAC_QC.qs2")
)

```

### AL12 ATAC QC

```{r}

# Run one sample and save the filtered object immediately.
AL12_ATAC <- qc_atac_sample(
  sample_name = "AL12",
  paths = paths,
  atac_params = atac_params
)

qs2::qs_save(
  AL12_ATAC,
  file = file.path(paths$object_dir, "AL12_ATAC_QC.qs2")
)

```

```{r}

### Combine the per-sample CSV summaries

# This expects every explicit sample block above to have written
# {sample}_ATAC_summary.csv.
atac_summary_df <- read_all_atac_summaries(samples, paths)

```

# scATAC Harmony Integration

This section is intentionally inactive. Consensus peak selection, fragment requantification, dimensional reduction, Harmony integration, and RNA label transfer will be redesigned after the per-sample QC outputs are reviewed.

## Integrate ATAC Across Samples

```{r eval=FALSE}

### Merge the retained per-sample Seurat objects
library(qs2)
AL05_ATAC <- qs_read(file = "output/scATAC/objects/AL05_ATAC_QC.qs2",nthreads = 10)
AL08_ATAC <- qs_read(file = "output/scATAC/objects/AL08_ATAC_QC.qs2",nthreads = 10)
AL10_ATAC <- qs_read(file = "output/scATAC/objects/AL10_ATAC_QC.qs2",nthreads = 10)
AL11_ATAC <- qs_read(file = "output/scATAC/objects/AL11_ATAC_QC.qs2",nthreads = 10)
AL12_ATAC <- qs_read(file = "output/scATAC/objects/AL12_ATAC_QC.qs2",nthreads = 10)



# Later integration starts here. The merge helper will requantify each filtered
# QC object on common peaks before merging, following the Signac merge workflow.
GSE274934_ATAC <- merge_atac_qc_objects(
  AL05 = AL05_ATAC,
  AL08 = AL08_ATAC,
  AL10 = AL10_ATAC,
  AL11 = AL11_ATAC,
  AL12 = AL12_ATAC,
  paths = paths,
  process_n = 2000
)

DefaultAssay(GSE274934_ATAC) <- "ATAC"

```

### LSI and Unintegrated ATAC Space

```{r eval=FALSE}

### Run LSI and inspect the unintegrated ATAC space

GSE274934_ATAC <- RunTFIDF(GSE274934_ATAC)
GSE274934_ATAC <- FindTopFeatures(GSE274934_ATAC, min.cutoff = "q5")
GSE274934_ATAC <- RunSVD(GSE274934_ATAC)

ElbowPlot_scCustom(GSE274934_ATAC,reduction = "lsi")

GSE274934_ATAC <- RunUMAP(
  GSE274934_ATAC,
  reduction = "lsi",
  umap.method = "uwot2",
  dims = 1:39
)

p_atac_unintegrated <- DimPlot_scCustom(
  GSE274934_ATAC,
  group.by = "sample"
) +
  labs(title = "Unintegrated ATAC UMAP by sample", x = "UMAP1", y = "UMAP2") +
  umap_theme

p_atac_unintegrated

ggsave(
  file.path(paths$figure_dir, "GSE274934_ATAC_unintegrated_umap_by_sample.png"),
  plot = p_atac_unintegrated,
  width = 12,
  height = 8,
  dpi = 600
)

```

### Harmony Integration

```{r eval=FALSE}

### Integrate with Harmony and cluster the integrated ATAC space
ElbowPlot_scCustom(GSE274934_ATAC,reduction = "lsi")

GSE274934_ATAC <- RunHarmony(
  object = GSE274934_ATAC,
  group.by.vars = "sample",
  reduction.use = "lsi",
  dims.use = 1:39,
  reduction.save = "harmony.atac",
  project.dim = FALSE
)

GSE274934_ATAC <- RunUMAP(
  GSE274934_ATAC,
  reduction = "harmony.atac",
  umap.method = "uwot2",
  dims = 1:39,
  seed.use = 1234
)

GSE274934_ATAC <- FindNeighbors(
  GSE274934_ATAC,
  reduction = "harmony.atac",
  dims = 1:39
)

```

### Resolution Scan and Silhouette Diagnostics

```{r eval=FALSE}

# Candidate resolutions to compare before choosing a final ATAC clustering.
resolutions <- seq(0.2, 1.5, by = 0.1)

# Use the Harmony embedding for both clustering and silhouette scoring.
reduction_use <- "harmony.atac"
dims_use <- 1:39

emb <- Embeddings(GSE274934_ATAC, reduction = reduction_use)[, dims_use, drop = FALSE]

# Store each temporary clustering in its own metadata column.
for (res in resolutions) {
  GSE274934_ATAC <- FindClusters(
    object = GSE274934_ATAC,
    resolution = res,
    algorithm = 4,
    random.seed = 1234,
    cluster.name = paste0("ATAC_snn_res.", res)
  )
}
cluster_cols <- paste0("ATAC_snn_res.", resolutions)

# Score each candidate clustering on a capped per-cluster cell sample.
library(cluster)
silhouette_results <- map_dfr(
  cluster_cols,
  ~ compute_silhouette_resolution(
    object = GSE274934_ATAC,
    cluster_col = .x,
    embedding = emb,
    n_per_cluster = 1000,
    seed = 1234
  )
)

# Convert metadata column names back to numeric resolution values for plotting.
silhouette_results <- silhouette_results %>%
  mutate(
    resolution_value = as.numeric(gsub("ATAC_snn_res\\.", "", resolution))
  ) %>%
  arrange(resolution_value) |>
  as_tibble()

silhouette_results

p_silhouette <- ggplot(
  silhouette_results,
  aes(x = resolution_value, y = mean_silhouette)
) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  geom_text(
    aes(label = n_clusters),
    vjust = -1,
    size = 3
  ) +
  theme_bw(base_size = 13) +
  labs(
    title = "Silhouette score across clustering resolutions",
    subtitle = "Numbers above points indicate number of clusters",
    x = "Clustering resolution",
    y = "Mean silhouette score"
  )

p_silhouette


# Export the diagnostic figure and table for later manual resolution choice.
ggsave(
  filename = file.path(paths$figure_dir, "ATAC_silhouette_by_resolution.png"),
  plot = p_silhouette,
  width = 7,
  height = 5,
  dpi = 600
)

write_tsv(
  silhouette_results,
  file = file.path(paths$table_dir, "ATAC_silhouette_by_resolution.tsv")
)

```

### Final ATAC Clustering and Integration Outputs

```{r eval=FALSE}

# Keep selected resolution as final cluster annotation
GSE274934_ATAC$atac_clusters_res0.2 <- factor(GSE274934_ATAC$ATAC_snn_res.0.2)

# Set active identities to resolution 0.4
Idents(GSE274934_ATAC) <- "atac_clusters_res0."

# Reset seurat_clusters to match resolution 0.4
GSE274934_ATAC$seurat_clusters <- Idents(GSE274934_ATAC)

# Remove all temporary ATAC_snn_res.* columns
res_cols <- grep("^ATAC_snn_res\\.", colnames(GSE274934_ATAC@meta.data), value = TRUE)

GSE274934_ATAC@meta.data <- GSE274934_ATAC@meta.data[
  ,
  !colnames(GSE274934_ATAC@meta.data) %in% res_cols,
  drop = FALSE
]

p_atac_sample <- DimPlot_scCustom(
  GSE274934_ATAC,
  reduction = "umap",
  group.by = "sample"
) +
  labs(title = "ATAC UMAP by sample", x = "UMAP1", y = "UMAP2") +
  umap_theme

p_atac_sample

ggsave(
  file.path(paths$figure_dir, "GSE274934_ATAC_umap_by_sample.png"),
  plot = p_atac_sample,
  width = 12,
  height = 8,
  dpi = 600
)

p_atac_clusters <- DimPlot_scCustom(
  GSE274934_ATAC,
  reduction = "umap",
  label = TRUE,label.box = T
) +
  labs(title = "ATAC UMAP by clusters", x = "UMAP1", y = "UMAP2") +
  umap_theme +
  NoLegend()

p_atac_clusters

ggsave(
  file.path(paths$figure_dir, "GSE274934_ATAC_umap_by_clusters.png"),
  plot = p_atac_clusters,
  width = 12,
  height = 8,
  dpi = 600
)


atac_harmony_summary <- GSE274934_ATAC@meta.data |>
  as.data.frame() |>
  dplyr::group_by(sample) |>
  dplyr::summarise(
    cells_after_qc = dplyr::n(),
    clusters = dplyr::n_distinct(seurat_clusters),
    median_nCount_ATAC = median(nCount_ATAC, na.rm = TRUE),
    median_nFeature_ATAC = median(nFeature_ATAC, na.rm = TRUE),
    median_pct_reads_in_peaks = median(pct_reads_in_peaks, na.rm = TRUE),
    median_blacklist_ratio = median(blacklist_ratio, na.rm = TRUE),
    median_nucleosome_signal = median(nucleosome_signal, na.rm = TRUE),
    median_TSS_enrichment = median(TSS.enrichment, na.rm = TRUE),
    .groups = "drop"
  )

atac_harmony_summary

write_tsv(
  atac_harmony_summary,
  file.path(paths$atac_qc_dir, "all_samples_ATAC_signac_harmony_summary.tsv")
)


gene_activity <- GeneActivity(GSE274934_ATAC)
GSE274934_ATAC[["ACTIVITY"]] <- CreateAssayObject(counts = gene_activity)
plan(sequential)
DefaultAssay(GSE274934_ATAC) <- "ACTIVITY"
GSE274934_ATAC <- NormalizeData(GSE274934_ATAC)
GSE274934_ATAC <- ScaleData(GSE274934_ATAC)

qs2::qs_save(
  GSE274934_ATAC,
  file = file.path(paths$object_dir, "GSE274934_ATAC.qs2"),
  nthreads = 10,
  compress_level = 10
)



```

# Plots of Interest

Exploratory coverage and feature plots for later review. This section is inactive during the QC-only run.

```{r eval=FALSE}


DefaultAssay(GSE274934_ATAC) <- "ATAC"

# Coverage plots for individual genes of interest after ATAC integration.
YAP1 <- CoveragePlot(
  object = GSE274934_ATAC,
  region = "YAP1",
  annotation = FALSE,
  peaks = FALSE
) +
  ggsci::scale_color_npg() +
  labs(title = "YAP1")

TEAD4 <- CoveragePlot(
  object = GSE274934_ATAC,
  region = "TEAD4",
  annotation = FALSE,
  peaks = FALSE
) +
  ggsci::scale_color_npg() +
  labs(title = "TEAD4")


FeaturePlot_scCustom(seurat_object = GSE274934_ATAC,reduction = "umap", features = c("HMGB1","HMGA2","YAP1","TEAD4")) &
  umap_theme



HMGB1 <- CoveragePlot(
  object = GSE274934_ATAC,
  region = "HMGB1",
  annotation = FALSE,
  peaks = FALSE
) +
  ggsci::scale_color_npg() +
  labs(title = "HMGB1")

HMGB2 <- CoveragePlot(
  object = GSE274934_ATAC,
  region = "HMGB2",
  annotation = FALSE,
  peaks = FALSE
) +
  ggsci::scale_color_npg() +
  labs(title = "HMGB2")

HMGA2 <- CoveragePlot(
  object = GSE274934_ATAC,
  region = "HMGA2",
  annotation = FALSE,
  peaks = FALSE
) +
  ggsci::scale_color_npg() +
  labs(title = "HMGA2")


library(patchwork)

HMGB1 + HMGB2 + YAP1 + TEAD4

```