---
title: "MAP 01 - Metadata Bridge"
format: html
---

# Section 1. Overview

This bridge notebook restores patient-level metadata that was not carried forward in the first integrated Seurat object, then checks how those labels relate to the plasticity score.

**MAP role:** metadata bridge. Keep this notebook small and use it only to restore sample or patient fields that are required for downstream state summaries, statistics, or multiome matching.

## Section 1.1. Reader Guide

- Inputs are the processed Seurat object and the manually entered sample metadata table.
- The notebook joins sample-level labels to each cell through `orig.ident`.
- The immediate check is a sex-stratified comparison of `C1_score`.
- Read this as a metadata repair and sanity-check step, not as the main statistical interpretation notebook.

## Section 1.2. MAP Section Boundaries

### Section 1.2.1. Notebook Flow

1. Load processed object.
2. Define and join sample-level metadata.
3. Run only minimal metadata sanity checks.
4. Save the updated object if the metadata fields are needed downstream.

## Section 1.3. Setup
```{r include=FALSE}

library(Seurat)
library(tidyverse)
library(ggpubr)
library(scCustomize)
library(RColorBrewer)
library(future)
library(qs2)

#plan(multicore, workers = 5)
#purrr::walk(packages, pak)

css_combined <- qs_read(file="output/objects/css_combined_processed.qs2", nthreads = 10)
#qs_save(object = css_combined, file = "output/objects/css_combined_processed.qs2", nthreads = 10,compress_level = 10)

```

# Section 2. Add Patient-Level Metadata
```{r}

df <- tribble(
  ~`Negedia code`, ~`Line primarie`, ~`Stage TNM (8th ed)`, ~`Classification from Oncogene`, ~Sex, ~`Sample type`,
  "ROC-25-0075",   "FD",             "IB-IIA",              "C1",                                                "M",  "Serum",
  "ROC-25-0074",   "CV",             "IIIA",                "C1",                                              "F",  "Serum",
  "ROC-25-0077",   "BB",             "IIIB",                "Not_determine",                                                  "M",  "Serum",
  "ROC-25-0076",   "MC",             "IIB",                 "nonC1",                                       "F",  "Serum"
) |>
    janitor::clean_names() |>
    mutate(orig.ident = c("S2_CSS", "S1_CSS", "S4_CSS", "S3_CSS"))

df

meta <- css_combined@meta.data

head(meta) 


meta_joined <- meta %>%
  left_join(df, by = "orig.ident")

head(meta_joined)

```


# Section 3. Quick Comparison by Sex
```{r}

meta_joined$sex <- factor(meta_joined$sex, levels = c("M","F"))

rcompanion::wilcoxonRG(x = meta_joined$C1_score, g = meta_joined$sex,verbose = T)

meta_joined %>%
  group_by(orig.ident, sex) %>%
  summarise(
    n_cells = n(),
    mean_C1 = mean(C1_score, na.rm = TRUE),
    median_C1 = median(C1_score, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(meta_joined, aes(x = orig.ident, y = C1_score, fill = sex)) +
  geom_violin(alpha = 0.3, trim = TRUE, linewidth = 0) +
  geom_boxplot(width = 0.15, fill = "white", color = "black",
               outlier.shape = NA, linewidth = 0.4) +
  scale_fill_brewer(palette = "Set2") +
  stat_compare_means() +
  theme_bw(base_size = 14)

```
