---
title: "MAP 02 - C1 Plasticity Scoring"
format: html
---

# Section 1. Overview

This notebook creates the central C1 plasticity score used throughout the CSS LUAD epithelial plasticity workflow.

**MAP role:** mandatory C1 scoring notebook. Preserve the continuous `C1_score`, the UP/DOWN UCell components, CytoTRACE2 comparison, sample/cluster summaries, and cell-cycle checks.

## Section 1.1. Reader Guide

- Inputs are the processed Seurat object from step 01 and the curated signature table `input/Signatures/Core_genes_tidy.csv`.
- The notebook scores UP and DOWN plasticity signatures with UCell, smooths them on the Harmony graph, and combines them into `C1_score`.
- It cross-checks the score against CytoTRACE2, sample identity, clusters, annotation labels, and cell-cycle metadata.
- Main outputs are score plots, signature presence tables, annotation comparisons, and updated Seurat metadata.
- Read this as the step where the biological idea of plasticity becomes a numeric gradient.

## Section 1.2. MAP Section Boundaries

### Section 1.2.1. Notebook Flow

1. Setup and load processed object.
2. Score UP and DOWN C1 signatures.
3. Build the continuous `C1_score`.
4. Validate C1 against CytoTRACE2 and annotations.
5. Generate essential UMAP, violin, and distribution plots.
6. Export score summaries and updated object metadata.

## Section 1.3. Setup

This section loads the required packages and restores the Seurat object used as the starting point for downstream analyses.

```{r include=FALSE}

packages <- c(
  "tidyverse",
  "Seurat",
  "SeuratWrappers",
  "harmony",
  "patchwork",
  "data.table",
  "fst",
  "UCell",
  "CytoTRACE2",
  "slingshot",
  "tradeSeq",
  "viridis",
  "RColorBrewer",
  "scran",
  "qs2",
  "ggsci",
  "scCustomize", 
  "future",
  "ggpubr",
  "GGally"
)

purrr::walk(packages, library, character.only = TRUE)

plan(multicore, workers = 5)

umap_theme <- ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(size = 16, face = "bold"),
    plot.subtitle = ggplot2::element_text(size = 11, color = "grey30"),
    panel.grid = ggplot2::element_blank(),
    axis.title = ggplot2::element_text(size = 13),
    axis.text = ggplot2::element_blank(),
    axis.ticks = ggplot2::element_blank(),
    axis.line = ggplot2::element_blank(),
    legend.title = ggplot2::element_text(size = 11),
    legend.text = ggplot2::element_text(size = 10)
  )

GSE274934 <- qs_read(file = "output/scRNA/objects/GSE274934_merged_RNA_QC.qs2",nthreads = 10)

```

# Cell-Cycle Scoring
```{r}

# A list of cell cycle markers, from Tirosh et al, 2015, is loaded with Seurat.  We can
# segregate this list into markers of G2/M phase and markers of S phase
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

GSE274934 <- CellCycleScoring(GSE274934, s.features = s.genes, g2m.features = g2m.genes, set.ident = F)

# Phase
p1 <- DimPlot_scCustom(
  GSE274934,
  group.by = c("Phase"),
  reduction = "umap",
) +
  umap_theme +
  labs(x = "UMAP1", y = "UMAP2", title = "Cell-Cycle Phase", subtitle = "Cell-Cycle Scoring")

p1

ggsave("output/scRNA/figures/CC_Phase.png", plot = p1, width = 12, height = 8)

p1 <- FeaturePlot_scCustom(
  GSE274934,
  features = c("G2M.Score","S.Score"),
  reduction = "umap",
  na_cutoff = NA
) &
  scale_color_viridis_c(option = "magma") &
  umap_theme &
  labs(x = "UMAP1", y = "UMAP2")

p1

ggsave("output/scRNA/figures/G2M_S_score.png", plot = p1, width = 16, height = 10)

```


# Section 2. Plasticity Scoring with UCell

Plasticity-related signatures are scored with UCell, smoothed on the Harmony graph, and collapsed into a single `C1_score` used throughout the notebook.

C1_Core_UCell_native     = native UCell directional UP+/DOWN- score
C1_Core_UCell_native_kNN = smoothed native UCell directional score
C1_Core                  = raw signed z(UP) - z(DOWN) contrast
C1_Core_kNN              = smoothed signed z(UP_kNN) - z(DOWN_kNN) contrast

```{r UCell score}

set.seed(1234)

library(dplyr)
library(UCell)
library(BiocParallel)
library(qs)

BPPARAM <- MulticoreParam(workers = 9)
register(BPPARAM)

zscore <- function(x) as.numeric(scale(x))

class_from_z <- function(x) {
  case_when(
    x > 1 ~ "C1-like",
    x > 0 & x <= 1 ~ "C1-intermediate",
    x >= -1 & x <= 0 ~ "Not-C1-intermediate",
    x < -1 ~ "Not-C1-like",
    TRUE ~ NA_character_
  )
}

check_signature <- function(gene_vector, object_genes, signature_name) {
  genes_input <- unique(na.omit(gene_vector))
  genes_found <- intersect(genes_input, object_genes)
  genes_lost  <- setdiff(genes_input, object_genes)

  message(
    signature_name, ": ",
    length(genes_found), "/", length(genes_input),
    " genes found in object."
  )

  if (length(genes_found) < 5) {
    stop(signature_name, " has fewer than 5 genes present in the object.")
  }

  list(genes_found = genes_found, genes_lost = genes_lost)
}

C1_Core_signature <- vroom::vroom(
  "input/C1_Core_signature.csv",
  show_col_types = FALSE
)

C1_Core_UP <- C1_Core_signature |>
  filter(C1_Core_direction == "UP")

C1_Core_DOWN <- C1_Core_signature |>
  filter(C1_Core_direction == "DOWN")

object_genes <- rownames(GSE274934)

up_checked <- check_signature(
  gene_vector = C1_Core_UP$gene,
  object_genes = object_genes,
  signature_name = "C1_Core_UP"
)

down_checked <- check_signature(
  gene_vector = C1_Core_DOWN$gene,
  object_genes = object_genes,
  signature_name = "C1_Core_DOWN"
)

missing_signature_genes <- list(
  C1_Core_UP_missing = up_checked$genes_lost,
  C1_Core_DOWN_missing = down_checked$genes_lost
)

missing_signature_genes

gene_list_split <- list(
  C1_Core_UP = up_checked$genes_found,
  C1_Core_DOWN = down_checked$genes_found
)

gene_list_native <- list(
  C1_Core_UCell_native = c(
    paste0(up_checked$genes_found, "+"),
    paste0(down_checked$genes_found, "-")
  )
)

GSE274934 <- AddModuleScore_UCell(
  obj = GSE274934,
  features = gene_list_split,
  BPPARAM = BPPARAM,
  name = NULL
)

GSE274934 <- AddModuleScore_UCell(
  obj = GSE274934,
  features = gene_list_native,
  BPPARAM = BPPARAM,
  name = NULL
)

GSE274934$C1_Core <- zscore(GSE274934$C1_Core_UP_UCell) -
  zscore(GSE274934$C1_Core_DOWN_UCell)

GSE274934 <- SmoothKNN(
  obj = GSE274934,
  signature.names = c(
    "C1_Core_UCell_native",
    "C1_Core_UP_UCell",
    "C1_Core_DOWN_UCell"
  ),
  reduction = "harmony"
)

GSE274934$C1_Core_kNN <- zscore(GSE274934$C1_Core_UP_UCell_kNN) -
  zscore(GSE274934$C1_Core_DOWN_UCell_kNN)

c1_group_levels <- c(
  "Not_C1_like",
  "Not_C1_intermediate",
  "C1_intermediate",
  "C1_like"
)

GSE274934$C1_Core_Class <- factor(
  class_from_z(GSE274934$C1_Core_kNN),
  levels = c1_group_levels
)

cols_remove <- c(
  "C1_Core_UP_UCell",
  "C1_Core_DOWN_UCell",
  "C1_Core_UP_UCell_kNN",
  "C1_Core_DOWN_UCell_kNN"
)

GSE274934@meta.data <- GSE274934@meta.data[
  ,
  setdiff(colnames(GSE274934@meta.data), cols_remove),
  drop = FALSE
]

GSE274934$C1_Core_2Class <- case_when(
  GSE274934$C1_Core_kNN > 1 ~ "C1_like",
  TRUE ~ "C1_not_enriched"
)

GSE274934@misc$C1_Core_missing_genes <- missing_signature_genes

bpstop(BPPARAM)

qs_save(
  GSE274934,
  file = file.path("output/scRNA/objects/GSE274934_merged_RNA_QC.qs2"),
  nthreads = 10,
  compress_level = 10
)

```
standardized difference between smoothed UP-UCell and smoothed DOWN-UCell. Derived z-contrast

```{r}

hlca_epithelial_finest <- c(
  "AT1",
  "AT2",
  "AT2 proliferating",
  "Transitional Club-AT2",
  "Suprabasal",
  "Basal resting",
  "Multiciliated (non-nasal)",
  "Multiciliated (nasal)",
  "Deuterosomal",
  "Goblet (nasal)",
  "Goblet (bronchial)",
  "Goblet (subsegmental)",
  "Club (nasal)",
  "Club (non-nasal)",
  "Neuroendocrine",
  "Ionocyte",
  "Tuft",
  "SMG serous (bronchial)",
  "SMG serous (nasal)",
  "SMG mucous",
  "SMG duct"
)

condense_C1_labels <- function(
  obj,
  sample_name,
  epithelial_finest_labels = hlca_epithelial_finest
) {
  
  cat("\n==============================\n")
  cat("C1 Core epithelial filtering:", sample_name, "\n")
  cat("==============================\n")
  
  required_cols <- c("predicted.ann_finest_level", "C1_Core_Class")
  missing_cols <- setdiff(required_cols, colnames(obj@meta.data))
  
  if (length(missing_cols) > 0) {
    stop("Missing metadata columns: ", paste(missing_cols, collapse = ", "))
  }
  
  original_class <- as.character(obj@meta.data[["C1_Core_Class"]])
  finest_label <- as.character(obj@meta.data[["predicted.ann_finest_level"]])
  
  is_epithelial <- finest_label %in% epithelial_finest_labels
  is_epithelial[is.na(is_epithelial)] <- FALSE
  
  obj@meta.data[["C1_Core_Class"]] <- dplyr::case_when(
    !is_epithelial ~ "filtered",
    TRUE ~ original_class
  )
  
  obj@meta.data[["C1_Core_Class"]] <- as.character(obj@meta.data[["C1_Core_Class"]])
  
  cat("\nFinal C1 Core labels:\n")
  print(table(obj@meta.data[["C1_Core_Class"]], useNA = "ifany"))
  
  cat("\nFinal labels by finest annotation:\n")
  print(table(
    C1_Core_Class = obj@meta.data[["C1_Core_Class"]],
    finest = obj@meta.data[["predicted.ann_finest_level"]],
    useNA = "ifany"
  ))
  
  cat("\nDone:", sample_name, "\n")
  
  return(obj)
}

GSE274934 <- condense_C1_labels(GSE274934, "GSE274934")

add_C1_Core_Epi <- function(
  obj,
  sample_name,
  epithelial_finest_labels = hlca_epithelial_finest
) {
  
  cat("\n==============================\n")
  cat("C1 Core continuous epithelial filtering:", sample_name, "\n")
  cat("==============================\n")
  
  required_cols <- c("predicted.ann_finest_level", "C1_Core")
  missing_cols <- setdiff(required_cols, colnames(obj@meta.data))
  
  if (length(missing_cols) > 0) {
    stop("Missing metadata columns: ", paste(missing_cols, collapse = ", "))
  }
  
  finest_label <- as.character(obj@meta.data[["predicted.ann_finest_level"]])
  
  is_epithelial <- finest_label %in% epithelial_finest_labels
  is_epithelial[is.na(is_epithelial)] <- FALSE
  
  original_score <- as.numeric(obj@meta.data[["C1_Core"]])
  
  obj@meta.data[["C1_Core_Epi"]] <- dplyr::case_when(
    !is_epithelial ~ NA_real_,
    TRUE ~ original_score
  )
  
  cat("\nEpithelial status:\n")
  print(table(is_epithelial, useNA = "ifany"))
  
  cat("\nC1_Core summary, original:\n")
  print(summary(obj@meta.data[["C1_Core"]]))
  
  cat("\nC1_Core_Epi summary, epithelial only:\n")
  print(summary(obj@meta.data[["C1_Core_Epi"]]))
  
  cat("\nNumber of NA in C1_Core_Epi:\n")
  print(sum(is.na(obj@meta.data[["C1_Core_Epi"]])))
  
  cat("\nC1_Core_Epi NA status by finest annotation:\n")
  print(table(
    C1_Core_Epi_is_NA = is.na(obj@meta.data[["C1_Core_Epi"]]),
    finest = obj@meta.data[["predicted.ann_finest_level"]],
    useNA = "ifany"
  ))
  
  cat("\nDone:", sample_name, "\n")
  
  return(obj)
}

GSE274934 <- add_C1_Core_Epi(GSE274934, "GSE274934")

```


# Section 4. UCell Plots

These plots summarize the spatial distribution of the UCell-derived signatures and compare the resulting `C1_score` across samples and Leiden clusters.

```{r UCell plots}

set.seed(1234)

library(ggsci)

# Extract NPG colors for the non-filtered classes
classes <- levels(factor(GSE274934$C1_Core_Class))
classes_non_filtered <- setdiff(classes, "filtered")

npg_cols <- pal_npg("nrc")(length(classes_non_filtered))

c1_cols <- c(
  filtered = "grey80",
  setNames(npg_cols, classes_non_filtered)
)

p1 <- DimPlot_scCustom(
  GSE274934,
  group.by = "C1_Core_Class",
  reduction = "umap"
) &
  scale_color_manual(values = c1_cols) &
  umap_theme &
  theme(legend.position = "bottom") &
  labs(
    x = "UMAP1",
    y = "UMAP2",
    title = "C1 Core Class",
    subtitle = "113 core genes"
  )

p1

ggsave(
  "output/scRNA/figures/C1_Core_Class.png",
  plot = p1,
  width = 12,
  height = 8
)

p1 <- FeaturePlot_scCustom(
  GSE274934,
  features = c("C1_Core"),
  reduction = "umap",
  na_cutoff = NA
) +
  scale_color_viridis_c(option = "magma") +
  umap_theme +
  labs(x = "UMAP1", y = "UMAP2", title = "C1 Core score", subtitle = "113 core genes")

p1

ggsave("output/scRNA/figures/CytoTRACE2_score.png", plot = p1, width = 12, height = 8)

p1 <- FeaturePlot_scCustom(
  GSE274934,
  features = "C1_Core_Epi",
  reduction = "umap"
) &
    scale_color_viridis_c(option = "magma") &
  umap_theme &
  labs(
    x = "UMAP1",
    y = "UMAP2",
    title = "C1 Core score",
    subtitle = "Epithelial cells only; non-epithelial cells set to NA"
  )

p1

ggsave("output/scRNA/figures/C1_Core_Epi.png", plot = p1, width = 12, height = 8)

umap_df <- FetchData(GSE274934, vars = c("C1_score", "orig.ident", "seurat_clusters"))

p1 <- ggplot(umap_df, aes(x = orig.ident, y = C1_score, fill = orig.ident)) +
  geom_violin(trim = TRUE, alpha = 0.7, fill = "white") +
  geom_boxplot(width = 0.1, outlier.shape = NA, alpha = 0.5) +
  ggpubr::stat_compare_means() +
  umap_theme +
  labs(
    x = "Sample",
    y = "C1 score",
    title = "C1 score distribution across samples",
    subtitle = "Violin and boxplot comparison by sample identity"
  ) + 
  scale_fill_brewer(palette = "Set2") +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    legend.position = "none"
  )

p1

ggsave("output/figures/C1_score_violin.png", p1, width = 10, height = 6)

p2 <- ggplot(umap_df, aes(x = SCT_snn_res.0.1, y = C1_score, fill = SCT_snn_res.0.1)) +
  geom_violin(trim = TRUE, alpha = 0.7, fill = "white") +
  geom_boxplot(width = 0.1, outlier.shape = NA, alpha = 0.5) +
  ggpubr::stat_compare_means() +
  umap_theme +
  labs(
    x = "Leiden cluster",
    y = "C1 score",
    title = "C1 score across Leiden clusters",
    subtitle = "Distribution of C1 program activity by transcriptional state"
  ) + 
  scale_fill_brewer(palette = "Set2") +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    legend.position = "none"
  )

p2

ggsave("output/figures/leiden_paga_violin.png", p2, width = 10, height = 6)

```

# Section 5. Discrete Plasticity Strata from C1 Score

The continuous `C1_score` is converted into low, medium, and high plasticity strata using the 25th and 75th percentiles.

```{r plasticity strata}

set.seed(1234)
library(fitdistrplus)
library(diptest)

x <- GSE274934$C1_Core

descdist(x, discrete = FALSE, boot = 1000)

plotdist(x, histo = TRUE, demp = TRUE)

fw <- fitdist(x, "norm", discrete = FALSE)
plot(fw)

summary(fw)

dip.test(x)

q <- quantile(x, c(.25, .75), type = 8)
q_low  <- q[1]
q_high <- q[2]

GSE274934$plasticity <- cut(
  x,
  breaks = c(-Inf, q_low, q_high, Inf),
  labels = c("low", "medium", "high"),
  ordered_result = TRUE
)

df <- data.frame(C1_Core = x)

p <- ggplot(df, aes(x = C1_Core)) +
  geom_density(linewidth = 1.2, fill = "steelblue", alpha = 0.3) +
  geom_vline(xintercept = q_low, linetype = "dashed", linewidth = 1, color = "red") +
  geom_vline(xintercept = q_high, linetype = "dashed", linewidth = 1, color = "red") +
  annotate("text", x = q_low, y = 0.3 * max(density(x)$y), label = "Q25", color = "red", hjust = 1.1) +
  annotate("text", x = q_high, y = 0.3 * max(density(x)$y), label = "Q75", color = "red", hjust = -0.1) +
  labs(
    title = "Distribution of C1 score (kNN-smoothed)",
    subtitle = "Dashed red lines indicate Q25 and Q75 thresholds",
    x = "C1 score",
    y = "Density"
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text( size = 12)
  )

p

ggsave("output/scRNA/figures/C1_Core_distribution.png", plot = p, width = 10, height = 6)



```


# Section 6. CytoTRACE2 Potency Scoring

CytoTRACE2 is run on raw RNA counts to provide an orthogonal estimate of differentiation potency that can be compared against the UCell-derived plasticity score.

```{r}

set.seed(1234)
library(BiocParallel)
library(CytoTRACE2)

cts <- GetAssayData(GSE274934, assay="RNA", layer="counts")

ct <- cytotrace2(cts, is_seurat=FALSE, species="human", seed=1234,ncores = 5)

GSE274934$CytoTRACE2 <- ct$CytoTRACE2_Score
GSE274934$CytoTRACE2_Potency <- factor(ct$CytoTRACE2_Potency)

library(janitor)
GSE274934@meta.data %>%
  tabyl(CytoTRACE2_Potency,SCEVAN_class) %>%
  adorn_totals("col") %>%
  adorn_percentages("row") %>%
  adorn_pct_formatting(digits = 2) %>%
  adorn_ns() %>%
  adorn_title()

qs_save(GSE274934, file = file.path("output/scRNA/objects/GSE274934_merged_RNA_QC.qs2"), nthreads = 10, compress_level = 10)

```

# Section 7. Plot CytoTRACE

These figures assess agreement between CytoTRACE2 and `C1_score`, while also checking that CytoTRACE2 potency is not merely tracking sample identity.

```{r}

set.seed(1234)

print(paste("Adjusted Rand Index CytoTRACE2 vs batch:", round(mclust::adjustedRandIndex(GSE274934$CytoTRACE2_Potency, GSE274934$orig.ident), 3)))
chisq.test(table(GSE274934$CytoTRACE2_Potency, GSE274934$orig.ident))
rstatix::cramer_v(table(GSE274934$CytoTRACE2_Potency, GSE274934$orig.ident), correct = TRUE)

print(paste("Adjusted Rand Index CytoTRACE2 vs Cell-Cycle phase:", round(mclust::adjustedRandIndex(GSE274934$CytoTRACE2_Potency, GSE274934$Phase), 3)))
chisq.test(table(GSE274934$CytoTRACE2_Potency, GSE274934$Phase))
rstatix::cramer_v(table(GSE274934$CytoTRACE2_Potency, GSE274934$Phase), correct = TRUE)

print(paste("Adjusted Rand Index CytoTRACE2 vs C1 Core Group:", round(mclust::adjustedRandIndex(GSE274934$CytoTRACE2_Potency, GSE274934$C1_Core_Class), 3)))
chisq.test(table(GSE274934$CytoTRACE2_Potency, GSE274934$C1_Core_Class))
rstatix::cramer_v(table(GSE274934$CytoTRACE2_Potency, GSE274934$C1_Core_Class), correct = TRUE)

# SCEVAN_class
p1 <- DimPlot_scCustom(
  GSE274934,
  group.by = c("SCEVAN_class"),
  reduction = "umap",
) +
  umap_theme +
  labs(x = "UMAP1", y = "UMAP2", title = "SCEVAN class", subtitle = "Single CEll Variational Aneuploidy aNalysis on epithelial cells")

p1

ggsave("output/scRNA/figures/SCEVAN_class.png", plot = p1, width = 12, height = 8)

# SCEVAN_class_and_CytoTRACE2
p1 <- DimPlot_scCustom(
  GSE274934,
  group.by = c("CytoTRACE2_Potency", "SCEVAN_class"),
  reduction = "umap",
) &
  umap_theme &
  theme(legend.position = "bottom") &
  labs(x = "UMAP1", y = "UMAP2")

p1

ggsave("output/scRNA/figures/SCEVAN_class_and_CytoTRACE2.png", plot = p1, width = 16, height = 10)

p1 <- FeaturePlot_scCustom(
  GSE274934,
  features = c("CytoTRACE2"),
  reduction = "umap",
  na_cutoff = NA
) +
  scale_color_viridis_c(option = "magma") +
  umap_theme +
  labs(x = "UMAP1", y = "UMAP2")

p1

ggsave("output/scRNA/figures/CytoTRACE2_score.png", plot = p1, width = 12, height = 8)



p1 <- FeaturePlot_scCustom(
  GSE274934,
  features = c("CytoTRACE2", "C1_score"),
  reduction = "umap",
  na_cutoff = NA
) +
  scale_color_viridis_c() +
  umap_theme +
  labs(x = "UMAP1", y = "UMAP2")

p1

ggsave("output/figures/UCell_and_CytoTRACE2.png", plot = p1, width = 10, height = 6)

Idents(GSE274934) <- "CytoTRACE2_Potency"
p2_feat <- FeaturePlot_scCustom(
  GSE274934,
  features = "C1_score",
  reduction = "umap",
  na_cutoff = NA
) +
  scale_color_viridis_c(option = "magma") +
  ggtitle("C1 Score") +
  labs(subtitle = "UMAP projection of C1 score (continuous program activity)") +
  umap_theme

p2_cluster <- DimPlot_scCustom(
  GSE274934,
  reduction = "umap",
  group.by = "CytoTRACE2_Potency",
  label = TRUE,
  label.size = 5,label.box = T
) +
  ggtitle("CytoTRACE2 Potency") +
  labs(subtitle = "Transcriptional potency inferred by CytoTRACE2")  +
  umap_theme

p2 <- (p2_feat + p2_cluster)

ggsave(
  "output/figures/UCell_and_CytoTRACE2_clusters.png",
  plot = p2,
  width = 12,
  height = 12
)


VlnPlot_scCustom(
  GSE274934,
  features = "C1_score",
  group.by = "CytoTRACE2_Potency"
) + theme_bw() +
  theme(legend.position = "none") +
  ggpubr::stat_compare_means()


```

CytoTRACE2 potency categories showed statistically significant associations with batch, cell-cycle phase, and C1 core class. However, adjusted Rand indices were low, indicating limited concordance between CytoTRACE2 potency labels and these categorical variables. Cramér’s V values were in the weak-to-moderate range, suggesting that batch and cell cycle may contribute to CytoTRACE2 variation but are unlikely to fully explain the potency stratification.

# Section 8. Correlation with C1 Score

The direct association between the UCell-derived plasticity score and CytoTRACE2 is evaluated with a Spearman correlation and pairwise density/scatter panels.

```{r}

cor.test(x = GSE274934@meta.data$C1_score, GSE274934@meta.data$CytoTRACE2,method = "spearman")

df <- data.frame(
  C1_score = GSE274934$C1_score,
  CytoTRACE2 = GSE274934$CytoTRACE2
)

library(GGally)
p <- ggpairs(
  df,columns = 1:2,
  lower = list(continuous = wrap("smooth_loess", se = T, color = "lightblue", alpha = 0.5)),
  upper = list(continuous = wrap("cor", size = 10,method = "spearman")),
  diag = list(continuous = wrap("densityDiag"))
) + 
  theme_bw(base_size = 15)

ggsave(plot = p,filename = "output/figures/UCell_and_CytoTRACE2_COR.png", width = 10, height = 6)

```



# Section 9. Annotation-Level Statistical Comparisons
```{r}

library(ggpubr)
library(FSA)
library(rcompanion)

# data
meta <- GSE274934@meta.data

# analysis of global differences (Kruskal-Wallis)
kruskal.test(C1_score ~ predicted.ann_finest_level, data = meta)

# analysis of pairwise differences (Wilcoxon)
pairwise.wilcox.test(
  meta$C1_score,
  meta$predicted.ann_finest_level,
  p.adjust.method = "BH"
)

# analysis of post-hoc Dunn test
dunn_res <- dunnTest(
  C1_score ~ predicted.ann_finest_level,
  data = meta,
  method = "bh"
)

dunn_df <- dunn_res$res

sig_dunn <- dunn_df %>%
  as_tibble() %>%
  separate(Comparison, into = c("group1", "group2"), sep = " - ") %>%
  filter(P.adj < 0.05) %>%
  arrange(P.adj) %>%
  mutate(direction = ifelse(Z > 0, "higher", "lower"))

write_csv(sig_dunn, "output/tables/C1_dunn_cells.csv")

# data prep for plotting (global)
meta <- meta %>%
  mutate(
    predicted.ann_finest_level = reorder(
      predicted.ann_finest_level,
      C1_score,
      FUN = max
    )
  )

# plot of C1 score across all cell types
p1 <- ggplot(meta, aes(
  x = predicted.ann_finest_level,
  y = C1_score,
  fill = predicted.ann_finest_level
)) +
  geom_violin(trim = TRUE, scale = "width", alpha = 0.8) +
  geom_boxplot(width = 0.12, outlier.shape = NA, fill = "white") +
  stat_compare_means(label = "p.signif", hide.ns = TRUE) +
  labs(
    x = "Azimuth annotation",
    y = "C1 score",
    title = "C1 program activity across cell states",
    subtitle = "Azimuth finest-level annotation"
  ) +
  coord_flip() +
  cowplot::theme_cowplot() +
  theme(
    plot.title = element_text(size = 15, face = "bold"),
    plot.subtitle = element_text(size = 11, color = "grey30"),
    legend.position = "none"
  )

ggsave("output/figures/C1_by_celltype.png", p1, width = 10, height = 6, dpi = 600)

# analysis subset for focused comparison
df_focus <- meta %>%
  filter(predicted.ann_finest_level %in%
           c("AT2 proliferating", "Suprabasal", "AT2", "Basal resting")) %>%
  mutate(predicted.ann_finest_level = factor(predicted.ann_finest_level))

# plot of selected epithelial states
ggplot(df_focus, aes(
  x = predicted.ann_finest_level,
  y = C1_score,
  fill = predicted.ann_finest_level
)) +
  geom_violin(trim = TRUE, alpha = 0.25, color = NA) +
  geom_boxplot(width = 0.15, fill = "white", color = "black",
               outlier.shape = NA, linewidth = 0.4) +
  scale_fill_brewer(palette = "Set2") +
  stat_compare_means(
    comparisons = list(
      c("AT2", "AT2 proliferating"),
      c("Suprabasal", "Basal resting")
    )
  ) +
  theme_bw(base_size = 12) +
  labs(
    x = "Cell type",
    y = "C1 score",
    title = "C1 score across selected cell types",
    subtitle = "Azimuth annotation"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    legend.position = "none"
  )

ggsave("output/figures/C1_by_celltype_AT2.png", width = 10, height = 6, dpi = 600)

# analysis of effect size (non-parametric)
df_focus2 <- meta %>%
  filter(predicted.ann_finest_level %in% c("Suprabasal", "Basal resting")) %>%
  mutate(predicted.ann_finest_level = factor(predicted.ann_finest_level))

rcompanion::wilcoxonRG(
  x = df_focus2$C1_score,
  g = df_focus2$predicted.ann_finest_level,
  verbose = TRUE
)

# analysis of effect size (non-parametric)
df_focus2 <- meta %>%
  filter(predicted.ann_finest_level %in% c("AT2", "AT2 proliferating")) %>%
  mutate(predicted.ann_finest_level = factor(predicted.ann_finest_level))

rcompanion::wilcoxonRG(
  x = df_focus2$C1_score,
  g = df_focus2$predicted.ann_finest_level,
  verbose = TRUE
)

```


# Section 11. Annotation and Cluster UMAP Views
```{r}

FeaturePlot_scCustom(GSE274934, na_cutoff = NA,
  features = "C1_score",
  reduction = "umap") +
  scale_color_viridis_c() +
    labs(
    x = "UMAP1",
    y = "UMAP2",
    title = "C1 score") +
  umap_theme

ggsave("output/figures/C1_score_UMAP.png", width=8, height=6,dpi = 600)

library(scCustomize)
library(ggsci)
library(ggrepel)

# Colors
npg_cols <- pal_npg("nrc")(length(unique(GSE274934$seurat_clusters)))
names(npg_cols) <- sort(unique(GSE274934$seurat_clusters))

# Base UMAP
p <- DimPlot_scCustom(
  GSE274934,
  group.by = "seurat_clusters",
  reduction = "umap",
  label = FALSE
) +
  scale_color_npg() +
  labs(x = "UMAP1", y = "UMAP2", title = "UMAP by leiden clusters") +
  umap_theme +
  theme(legend.position = "none")

# Centroids
centroids <- Embeddings(GSE274934, "umap") %>%
  as.data.frame() %>%
  mutate(cluster = GSE274934$seurat_clusters) %>%
  group_by(cluster) %>%
  summarise(
    UMAP_1 = median(umap_1),
    UMAP_2 = median(umap_2),
    .groups = "drop"
  )

# Plot + BIGGER labels
p <- p +
geom_label(
  data = centroids,
  aes(UMAP_1, UMAP_2, label = cluster, fill = cluster),
  color = "black",
  fontface = "bold",
  size = 4,
  label.size = 0.6,         # thickness of border
  label.padding = unit(0.4, "lines"),
  label.r = unit(0.35, "lines"),
  show.legend = FALSE
) +
  scale_fill_manual(values = npg_cols)


ggsave("output/figures/leiden_UMAP.png", width=8, height=6,dpi = 600)


library(dplyr)
library(ggrepel)

umap_df <- Embeddings(GSE274934, "umap") %>%
  as.data.frame() %>%
  mutate(celltype = GSE274934$predicted.ann_finest_level)

centroids <- umap_df %>%
  group_by(celltype) %>%
  summarise(
    UMAP_1 = median(umap_1),
    UMAP_2 = median(umap_2),
    n = n(),
    .groups = "drop"
  )
 
p <- DimPlot_scCustom(
  GSE274934,
  group.by = "predicted.ann_finest_level",
  reduction = "umap",
  label = FALSE
) +
  labs(x = "UMAP1", y = "UMAP2", title = "Cell type annotation") +
  umap_theme +
  theme(legend.position = "none")

p +
  geom_label_repel(
    data = centroids,
    aes(
      UMAP_1,
      UMAP_2,
      label = paste0(celltype, "\n(n=", n, ")")
    ),
    size = 4.5,
    fontface = "bold",
    
    # box style
    fill = "white",
    label.size = 0.6,
    label.colour = "black",
    
    # repelling behavior (important)
    box.padding = 0.6,
    point.padding = 0.4,
    max.overlaps = Inf,
    
    # leader lines
    segment.color = "black",
    segment.size = 0.3,
    
    show.legend = FALSE
  )

p <- DimPlot_scCustom(
  GSE274934,
  group.by = "predicted.ann_finest_level",
  reduction = "umap",
  label = F
) +
  labs(x = "UMAP1", y = "UMAP2", title = "Cell type annotation") +
  umap_theme 

p

ggsave("output/figures/annotation_UMAP.png", plot = p, width=8, height=6,dpi = 600)
 
```


# Section 13. Save Updated Metadata and Annotation Outputs

```{r}

Idents(GSE274934) <- "seurat_clusters"

cols_to_remove <- c(
  # assay junk
  "nCount_SCT", "nFeature_SCT",
  "nCount_tfsulm", "nFeature_tfsulm",
  "nCount_pathwaysulm", "nFeature_pathwaysulm",
  
  # clustering redundancy
  "SCT_snn_res.0.1",
  
  # azimuth levels
  "predicted.ann_level_1", "predicted.ann_level_1.score",
  "predicted.ann_level_2", "predicted.ann_level_2.score",
  "predicted.ann_level_3", "predicted.ann_level_3.score",
  "predicted.ann_level_4", "predicted.ann_level_4.score",
  "predicted.ann_level_5", "predicted.ann_level_5.score",
  "predicted.ann_finest_level.score",
  "mapping.score"
)

GSE274934@meta.data <- GSE274934@meta.data[ , 
  !(colnames(GSE274934@meta.data) %in% cols_to_remove)
]

assays_to_remove <- c(
  "prediction.score.ann_level_1",
  "prediction.score.ann_level_2",
  "prediction.score.ann_level_3",
  "prediction.score.ann_level_4",
  "prediction.score.ann_level_5",
  "prediction.score.ann_finest_level"
)

for (assay in assays_to_remove) {
  if (assay %in% names(GSE274934@assays)) {
    GSE274934[[assay]] <- NULL
  }
}

head(GSE274934@meta.data)

```


```{r}



library(dplyr)
library(ggplot2)
library(ggpubr)
library(RColorBrewer)
library(patchwork)

dir.create("output/figures", recursive = TRUE, showWarnings = FALSE)

epithelial_labels <- c(
  "AT1",
  "AT2",
  "AT2 proliferating",
  "Basal resting",
  "Club (nasal)",
  "Club (non-nasal)",
  "Goblet (nasal)",
  "Ionocyte",
  "Multiciliated (non-nasal)",
  "SMG duct",
  "SMG mucous",
  "SMG serous (bronchial)",
  "Suprabasal",
  "Transitional Club-AT2",
  "Tuft"
)


df <- GSE274934@meta.data |>
  as_tibble() |>
  filter(predicted.ann_finest_level %in% epithelial_labels) |>
  mutate(
    predicted.ann_finest_level = factor(predicted.ann_finest_level)
  )


library(dplyr)
library(ggplot2)
library(ggpubr)
library(RColorBrewer)
library(patchwork)

dir.create("output/figures", recursive = TRUE, showWarnings = FALSE)

cell_levels <- df |>
  group_by(predicted.ann_finest_level) |>
  summarise(med_C1 = median(C1_Core, na.rm = TRUE), .groups = "drop") |>
  arrange(med_C1) |>
  pull(predicted.ann_finest_level)

df <- df |>
  mutate(predicted.ann_finest_level = factor(predicted.ann_finest_level, levels = cell_levels))

pastel_cols <- setNames(
  colorRampPalette(brewer.pal(9, "Pastel1"))(length(cell_levels)),
  cell_levels
)

base_theme <- theme_pubclean(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 11, color = "grey30"),
    legend.position = "none"
  )

p1 <- ggplot(
  df,
  aes(
    x = predicted.ann_finest_level,
    y = C1_Core,
    fill = predicted.ann_finest_level,
    color = predicted.ann_finest_level
  )
) +
  geom_violin(trim = TRUE, scale = "width", alpha = 0.75, linewidth = 0.25) +
  geom_boxplot(
    width = 0.13,
    outlier.shape = NA,
    fill = "white",
    color = "black",
    linewidth = 0.35
  ) +
  scale_fill_manual(values = pastel_cols, drop = FALSE) +
  scale_color_manual(values = pastel_cols, drop = FALSE) +
  coord_flip() +
  labs(
    x = "Cell type",
    y = "C1-Core",
    title = "All cell types",
    subtitle = "Azimuth::Lung v2 (HLCA) | Kruskal–Wallis p < 2e-16"
  ) +
  base_theme

focus_cells <- c("AT2", "AT2 proliferating", "Basal resting", "Suprabasal")

df_focus <- df |>
  filter(predicted.ann_finest_level %in% focus_cells) |>
  mutate(predicted.ann_finest_level = factor(predicted.ann_finest_level, levels = focus_cells))

p2 <- ggplot(
  df_focus,
  aes(
    x = predicted.ann_finest_level,
    y = C1_Core,
    fill = predicted.ann_finest_level,
    color = predicted.ann_finest_level
  )
) +
  geom_violin(trim = TRUE, scale = "width", alpha = 0.75, linewidth = 0.25) +
  geom_boxplot(
    width = 0.15,
    outlier.shape = NA,
    fill = "white",
    color = "black",
    linewidth = 0.35
  ) +
  stat_compare_means(
    comparisons = list(
      c("AT2", "AT2 proliferating"),
      c("Basal resting", "Suprabasal")
    ),
    method = "wilcox.test",
    label = "p.format",
    size = 4
  ) +
  scale_fill_manual(values = pastel_cols, drop = FALSE) +
  scale_color_manual(values = pastel_cols, drop = FALSE) +
  labs(
    x = "Cell type",
    y = "C1-Core",
    title = "Focused epithelial populations",
    subtitle = "Wilcoxon tests between selected groups"
  ) +
  base_theme

comb <- (p1 | p2) +
  plot_layout(widths = c(1.35, 1)) +
  plot_annotation(
    title = "C1-Core distribution across Azimuth lung cell annotations",
    theme = theme(
      plot.title = element_text(face = "bold", size = 17),
      plot.subtitle = element_text(size = 12, color = "grey30")
    )
  )

p1
p2
comb

ggsave("output/figures/C1_Core_by_celltype.png", p1, width = 12, height = 8, dpi = 600)
ggsave("output/figures/C1_Core_by_celltype_focus.png", p2, width = 10, height = 6, dpi = 600)
ggsave("output/figures/C1_Core_by_celltype_Combined.png", comb, width = 15, height = 8, dpi = 600)


```

# Checkpoint

This checkpoint saves the Seurat object after adding regulatory, pathway, secreted-protein, and Hallmark activity layers.

```{r}

DefaultAssay(GSE274934) <- "RNA"
qs_save(GSE274934, file = file.path("output/scRNA/objects/GSE274934_merged_RNA_QC.qs2"), nthreads = 10, compress_level = 10)

```
