---
title: "MAP 03 - State Markers, Statistics, and Export"
format: html
---

# Section 1. Overview

This notebook collects downstream interpretation and statistical checks for the CSS plasticity analysis.

**MAP role:** mandatory state, marker, statistics, and export notebook. Any miscellaneous plots, covariate checks, score tests, regression models, or random statistical diagnostics should live here instead of being scattered across the MAP workflow.

## Section 1.1. Reader Guide

- Input is the processed Seurat object with scores, clusters, annotations, and metadata from earlier steps.
- The notebook discovers cluster markers, summarizes C1 by sample and cluster, runs non-parametric and linear-model checks, and evaluates possible confounders.
- Main outputs are marker tables, sample and cluster summaries, residual plots, correlation checks, and the Python-ready `.h5ad` export.
- Read this as the R-side interpretation and quality-control bridge before the velocity and CellRank branch.

## Section 1.2. MAP Section Boundaries

### Section 1.2.1. Notebook Flow

1. Load scored object.
2. Compute cluster markers and essential marker plots.
3. Optionally re-import scVelo/CellRank2 metadata when those outputs already exist.
4. Run essential sample, annotation, and cluster summaries.
5. Run only core Wilcoxon/Kruskal checks needed to read the result.
6. Compute residual and confounder checks.
7. Save the updated object and export AnnData for Python velocity/metabolism.

Extensive post-hoc tests, adjusted models, permutations, and sensitivity analyses belong in `10_MAP_extensive_statistics.qmd`.

## Section 1.3. Setup

This section loads the required packages and restores the Seurat object used as the starting point for downstream analyses.

```{r include=FALSE}

packages <- c(
  "tidyverse",
  "Seurat",
  "patchwork",
  "data.table",
  "fst",
  "UCell",
  "viridis",
  "RColorBrewer",
  "qs2",
  "ggsci",
  "scCustomize", 
  "future",
  "ggpubr",
  "GGally"
)


purrr::walk(packages, library, character.only = TRUE)

plan(multicore, workers = 4)

options(future.globals.maxSize = 15 * 1024^3)  # 6 GB

umap_theme <- cowplot::theme_cowplot() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 11, color = "grey30"),
    axis.title = element_text(size = 13),
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10)
  )

css_combined <- qs_read(file = "output/objects/css_combined_processed.qs2",nthreads = 10)

```


# Section 2. Cluster Biomarker Discovery

This first analysis block identifies broad cluster markers that will be reused in later interpretation plots and summary tables.
```{r }
#| label: all_markers
#| warning: false

# Default assay is SCT
#DefaultAssay(css_combined) <- "SCT"
# Prepare SCT models for DE testing
#css_combined <- PrepSCTFindMarkers(css_combined, assay = "SCT")

set.seed(1234)
DefaultAssay(css_combined) <- "RNA"

# Then run DE
combined_markers <- FindAllMarkers(
  css_combined,
  #test.use = "MAST",
  only.pos = FALSE,
  group.by = "seurat_clusters",
  min.pct = 0.25,
  logfc.threshold = 0.25#, latent.vars = c("orig.ident")
)

fwrite(combined_markers,"output/tables/CSS_combined_markers.csv",row.names = T)


top3 <- combined_markers %>% group_by(cluster) %>% top_n(n = 4, wt = avg_log2FC)
DoHeatmap(css_combined, features = top3$gene) + NoLegend()
rm(top3)
 


```

# Section 3. Marker Table Exports

This section splits the all-marker table into cluster-specific files so later notebooks and reports can reuse the marker calls without rerunning differential expression.

```{r}

head(combined_markers)

# split markers into one data frame per cluster
cluster_markers_list <- combined_markers |>
  split(~ cluster)

# give the list clean names like Cluster1_markers, Cluster2_markers, ...
names(cluster_markers_list) <- paste0("Cluster", names(cluster_markers_list), "_markers")

# write one csv per cluster
iwalk(
  cluster_markers_list,
  ~ write_csv(.x, file = file.path("output/tables", paste0(.y, ".csv")))
)

```

# Section 4. ORA

```{r}

library(gprofiler2)
library(data.table)
library(ggplot2)
library(dplyr)
library(purrr)

run_gprofiler_ordered <- function(
  genes,
  object_name,
  sources = c("GO:BP", "KEGG", "REAC", "WP", "TF", "MIRNA")
) {
  message("Running g:Profiler for ", object_name)
  
  genes <- genes |>
    na.omit() |>
    unique()
  
  if (length(genes) < 5) {
    message("  Too few genes")
    return(NULL)
  }
  
  gp_res <- tryCatch(
    gost(
      query = genes,
      organism = "hsapiens",
      ordered_query = TRUE,
      multi_query = FALSE,
      significant = TRUE,
      sources = sources
    ),
    error = function(e) {
      message("  gost failed: ", e$message)
      NULL
    }
  )
  
  if (is.null(gp_res) || is.null(gp_res$result) || nrow(gp_res$result) == 0) {
    message("  No significant terms")
    return(gp_res)
  }
  
  object_stub <- gsub("[^A-Za-z0-9_]+", "_", object_name)
  
  fwrite(
    gp_res$result,
    file = file.path("output/tables", paste0(object_stub, "_gprofiler_results.csv"))
  )
  
  p <- gostplot(gp_res, capped = TRUE, interactive = FALSE) +
    theme_bw(base_size = 11) +
    labs(title = paste("g:Profiler ordered enrichment -", object_name))
  
  ggsave(
    plot = p,
    filename = file.path("output/figures", paste0(object_stub, "_gprofiler_plot.png")),
    width = 16,
    height = 8,
    dpi = 600
  )
  
  message("  Saved CSV and plot")
  gp_res
}

extract_ranked_genes <- function(df, rank_col = "avg_log2FC") {
  df |>
    filter(!is.na(gene), !is.na(.data[[rank_col]])) |>
    distinct(gene, .keep_all = TRUE) |>
    arrange(desc(.data[[rank_col]])) |>
    pull(gene)
}

n_clusters <- length(cluster_markers_list)

gprofiler_results_list <- imap(
  cluster_markers_list,
  function(df, object_name) {
    i <- match(object_name, names(cluster_markers_list))
    message("[", i, "/", n_clusters, "] g:Profiler for ", object_name)
    
    genes <- extract_ranked_genes(df, rank_col = "avg_log2FC")
    
    run_gprofiler_ordered(
      genes = genes,
      object_name = object_name,
      sources = c("GO:BP", "KEGG", "REAC", "WP", "TF", "MIRNA")
    )
  }
)

```

# Section 5. GSEA for GO Biological Processes

```{r}

library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(dplyr)
library(data.table)
library(ggplot2)
library(purrr)

run_go_gsea <- function(df, object_name, rank_col = "logFClineage1", ont = "BP") {
  
  message("Running GO GSEA for ", object_name)
  
  ranked_genes <- df %>%
    filter(!is.na(gene), !is.na(.data[[rank_col]])) %>%
    distinct(gene, .keep_all = TRUE) %>%
    arrange(desc(.data[[rank_col]]))
  
  gene_list <- ranked_genes[[rank_col]]
  names(gene_list) <- ranked_genes$gene
  gene_list <- sort(gene_list, decreasing = TRUE)
  
  if (length(gene_list) < 10) {
    message("  Too few genes for GSEA")
    return(NULL)
  }
  
  gsea_res <- tryCatch(
    gseGO(
      geneList      = gene_list,
      OrgDb         = org.Hs.eg.db,
      ont           = ont,
      keyType       = "SYMBOL",
      pvalueCutoff  = 0.05,
      minGSSize     = 10,
      maxGSSize     = 500,
      eps           = 0,
      verbose       = FALSE
    ),
    error = function(e) {
      message("  gseGO failed: ", e$message)
      NULL
    }
  )
  
  if (is.null(gsea_res)) return(NULL)
  
  gsea_df <- as.data.frame(gsea_res)
  object_stub <- gsub("[^A-Za-z0-9_]+", "_", object_name)
  
  fwrite(
    gsea_df,
    file = file.path("output/tables", paste0(object_stub, "_GO_", ont, "_gsea_results.csv"))
  )
  
  if (nrow(gsea_df) == 0) {
    message("  No significant GO terms")
    return(gsea_res)
  }
  
  p <- enrichplot::dotplot(gsea_res, showCategory = 25) +
    theme_bw(base_size = 11) +
    labs(title = paste("GO", ont, "GSEA -", object_name))
  
  ggsave(
    filename = file.path("output/figures", paste0(object_stub, "_GO_", ont, "_gsea_barplot.png")),
    plot = p,
    width = 14,
    height = 10,
    dpi = 600
  )
  
  message("  Saved CSV and barplot")
  gsea_res
}

n_clusters <- length(cluster_markers_list)

go_results_list <- imap(
  cluster_markers_list,
  function(df, object_name) {
    i <- match(object_name, names(cluster_markers_list))
    message("[", i, "/", n_clusters, "] GO GSEA for ", object_name)
    run_go_gsea(df, object_name, rank_col = "avg_log2FC", ont = "BP")
  }
)

```

# Section 6. GSEA for KEGG Pathways

```{r}

cluster_markers_list <- combined_markers |>
  split(~ cluster)

names(cluster_markers_list) <- paste0("Cluster", names(cluster_markers_list), "_markers")

all_cluster_genes <- combined_markers |>
  dplyr::filter(!is.na(gene)) |>
  dplyr::distinct(gene) |>
  dplyr::pull(gene)

gene_map <- bitr(
  all_cluster_genes,
  fromType = "SYMBOL",
  toType   = "ENTREZID",
  OrgDb    = org.Hs.eg.db
) |>
  dplyr::distinct(SYMBOL, .keep_all = TRUE)

```

```{r}

library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(dplyr)
library(data.table)
library(ggplot2)
library(purrr)


run_kegg_gsea <- function(df, cluster_name, gene_map, rank_col = "avg_log2FC") {
  
  ranked_genes <- df |>
    filter(!is.na(gene), !is.na(.data[[rank_col]])) |>
    distinct(gene, .keep_all = TRUE) |>
    arrange(desc(.data[[rank_col]]))
  
  gene_list <- ranked_genes[[rank_col]]
  names(gene_list) <- ranked_genes$gene
  gene_list <- sort(gene_list, decreasing = TRUE)
  
  map_sub <- gene_map |>
    filter(SYMBOL %in% names(gene_list))
  
  if (nrow(map_sub) == 0) {
    message("  No mapped genes found")
    return(NULL)
  }
  
  gene_list_entrez <- gene_list[map_sub$SYMBOL]
  names(gene_list_entrez) <- map_sub$ENTREZID
  
  gene_list_entrez <- gene_list_entrez[!duplicated(names(gene_list_entrez))]
  gene_list_entrez <- sort(gene_list_entrez, decreasing = TRUE)
  
  if (length(gene_list_entrez) < 10) {
    message("  Too few mapped genes for GSEA")
    return(NULL)
  }
  
  gsea_res <- tryCatch(
    gseKEGG(
      geneList = gene_list_entrez,
      organism = "hsa",
      pvalueCutoff = 0.05,
      minGSSize = 10,
      maxGSSize = 500,
      eps = 0,
      verbose = FALSE
    ),
    error = function(e) {
      message("  gseKEGG failed: ", e$message)
      NULL
    }
  )
  
  if (is.null(gsea_res)) return(NULL)
  
  gsea_df <- as.data.frame(gsea_res)
  cluster_stub <- gsub("[^A-Za-z0-9_]+", "_", cluster_name)
  
  fwrite(
    gsea_df,
    file = file.path("output/tables", paste0(cluster_stub, "_KEGG_gsea_results.csv"))
  )
  
  if (nrow(gsea_df) == 0) {
    message("  No significant KEGG terms")
    return(gsea_res)
  }
  
  p <- enrichplot::dotplot(gsea_res, showCategory = 25) +
    theme_bw(base_size = 12) +
    labs(title = paste("KEGG GSEA -", cluster_name))
  
  ggsave(
    filename = file.path("output/figures", paste0(cluster_stub, "_KEGG_gsea_barplot.png")),
    plot = p,
    width = 14,
    height = 10,
    dpi = 600
  )
  
  message("  Saved CSV and barplot")
  gsea_res
}

n_clusters <- length(cluster_markers_list)

kegg_results_list <- imap(
  cluster_markers_list,
  function(df, cluster_name) {
    i <- match(cluster_name, names(cluster_markers_list))
    message("[", i, "/", n_clusters, "] Running KEGG for ", cluster_name)
    run_kegg_gsea(df, cluster_name, gene_map = gene_map, rank_col = "avg_log2FC")
  }
)

```

# Section 7. ReactomePA
```{r}

library(clusterProfiler)
library(org.Hs.eg.db)
library(dplyr)
library(data.table)


# Keep unique genes and filter NA waldStat
ranked_genes <- along_DEGs %>%
  filter(!is.na(waldStat)) %>%
  distinct(gene, .keep_all = TRUE) %>%
  arrange(desc(logFClineage1))

# Create numeric vector for GSEA
gene_list <- ranked_genes$logFClineage1
names(gene_list) <- ranked_genes$gene
gene_list <- sort(gene_list, decreasing = TRUE)

# Map gene symbols to Entrez IDs for KEGG
gene_df <- bitr(names(gene_list), fromType = "SYMBOL", 
                toType = "ENTREZID", OrgDb = org.Hs.eg.db)
gene_list_entrez <- gene_list[gene_df$SYMBOL]
names(gene_list_entrez) <- gene_df$ENTREZID

library(ReactomePA)

gsea_res_Reactome <- gsePathway(
  geneList = gene_list_entrez,
  organism = "human",
  pvalueCutoff = 0.05,
  minGSSize = 10,
  maxGSSize = 500,
  verbose = TRUE
)

# Convert Entrez IDs to symbols
gsea_res_Reactome <- setReadable(gsea_res_Reactome, OrgDb = org.Hs.eg.db)

dotplot(gsea_res_Reactome, showCategory = 15)  +
  theme_bw(base_size = 11) +
  labs(title = "GSEA ReactomePA Pathways for genes along Pseudotime")

ggsave(filename = "output/figures/clusterProfiler_ReactomePA_tradeSEQ.svg", width = 10, height = 8)

# Save results as CSV
fwrite(as.data.frame(gsea_res_Reactome), "output/tables/tradeSeq_GSEA_ReactomePA_results.csv")

```

# Section 8. Marker Visualization

```{r}

DefaultAssay(css_combined) <- "SCT"
library(SeuratExtend)
a <- FeaturePlot3(
  css_combined,
  reduction = "umap",
  color = "ryb",
  feature.1 = "DLK2",
  feature.2 = "RRM2",
  feature.3 = "SAA2",
  pt.size = 0.5
)
a 

ggsave(
  filename = "output/figures/markers_fate.jpg",
  plot = a,
  width = 12,
  height = 8)


```


# Section 9. Optional Re-import of scVelo and CellRank2 Metadata

Run this section after `07_MAP_velocity_scvelo_cellrank.qmd` if the velocity metadata should be available inside the Seurat object for R-side plotting, violin tests, summaries, or export.

```{r}

velocity_metadata_path <- "output/tables/MAP_velocity_cellrank_metadata_for_seurat.csv"

if (file.exists(velocity_metadata_path)) {
  velocity_metadata <- readr::read_csv(velocity_metadata_path, show_col_types = FALSE)

  velocity_metadata <- velocity_metadata |>
    dplyr::mutate(cell_id = as.character(cell_id)) |>
    tibble::column_to_rownames("cell_id")

  velocity_cols_to_add <- c(
    "velocity_pseudotime",
    "latent_time",
    "velocity_length",
    "velocity_confidence",
    "velocity_confidence_transition",
    "root_cells",
    "end_points",
    "harmony_1",
    "harmony_2"
  )

  velocity_cols_to_add <- intersect(velocity_cols_to_add, colnames(velocity_metadata))

  common_cells <- intersect(colnames(css_combined), rownames(velocity_metadata))

  css_combined <- AddMetaData(
    object = css_combined,
    metadata = velocity_metadata[common_cells, velocity_cols_to_add, drop = FALSE]
  )

  qs_save(
    object = css_combined,
    file = "output/objects/css_combined_processed_with_velocity.qs2",
    compress_level = 10,
    nthreads = 10
  )
}

```

# Section 10. Metadata-Level Statistical Tests

These summary tables keep the C1 score readable at sample, cluster, and sample-by-cluster levels before the notebook moves into formal tests.

```{r}

# Set a reproducible seed for the metadata statistics.
set.seed(1234)

# Libraries
library(dplyr)
library(readr)
library(rstatix)
library(e1071)

# Metadata
meta <- css_combined@meta.data
meta$seurat_clusters <- factor(meta$seurat_clusters)

# BASIC SUMMARY STATS PER SAMPLE
Summary_C1 <- meta %>%
  group_by(sample) %>%
  summarise(
    n_cells     = n(),

    mean_C1     = mean(C1_score, na.rm = TRUE),
    median_C1   = median(C1_score, na.rm = TRUE),

    sd_C1       = sd(C1_score, na.rm = TRUE),
    var_C1      = var(C1_score, na.rm = TRUE),

    min_C1      = min(C1_score, na.rm = TRUE),
    max_C1      = max(C1_score, na.rm = TRUE),

    q25_C1      = quantile(C1_score, 0.25, na.rm = TRUE),
    q75_C1      = quantile(C1_score, 0.75, na.rm = TRUE),
    IQR_C1      = IQR(C1_score, na.rm = TRUE),

    skewness_C1 = skewness(C1_score, na.rm = TRUE),
    kurtosis_C1 = kurtosis(C1_score, na.rm = TRUE),

    sem_C1      = sd_C1 / sqrt(n_cells),

    ci95_low    = mean_C1 - 1.96 * sem_C1,
    ci95_high   = mean_C1 + 1.96 * sem_C1,

    pct_high_C1 = mean(C1_score > 0.5) * 100
  ) %>%
  arrange(desc(mean_C1))

Summary_C1

# Save
write_csv(
  Summary_C1,
  file = "output/tables/Summary_of_C1_per_sample.csv"
)


Cluster_Summary <- meta %>%
  group_by(seurat_clusters) %>%
  summarise(
    n_cells     = n(),

    mean_C1     = mean(C1_score, na.rm = TRUE),
    median_C1   = median(C1_score, na.rm = TRUE),

    sd_C1       = sd(C1_score, na.rm = TRUE),
    var_C1      = var(C1_score, na.rm = TRUE),

    min_C1      = min(C1_score, na.rm = TRUE),
    max_C1      = max(C1_score, na.rm = TRUE),

    q25_C1      = quantile(C1_score, 0.25, na.rm = TRUE),
    q75_C1      = quantile(C1_score, 0.75, na.rm = TRUE),
    IQR_C1      = IQR(C1_score, na.rm = TRUE),

    skewness_C1 = skewness(C1_score, na.rm = TRUE),
    kurtosis_C1 = kurtosis(C1_score, na.rm = TRUE),

    sem_C1      = sd_C1 / sqrt(n_cells),

    ci95_low    = mean_C1 - 1.96 * sem_C1,
    ci95_high   = mean_C1 + 1.96 * sem_C1,

    pct_high_C1 = mean(C1_score > 0.5) * 100
  ) %>%
  arrange(desc(mean_C1))

Cluster_Summary

write_csv(
  Cluster_Summary,
  file = "output/tables/Cluster_Summary_of_C1.csv"
)

Sample_Cluster_Summary <- meta %>%
  group_by(sample, seurat_clusters) %>%
  summarise(
    n_cells     = n(),

    mean_C1     = mean(C1_score, na.rm = TRUE),
    median_C1   = median(C1_score, na.rm = TRUE),

    sd_C1       = sd(C1_score, na.rm = TRUE),
    var_C1      = var(C1_score, na.rm = TRUE),

    min_C1      = min(C1_score, na.rm = TRUE),
    max_C1      = max(C1_score, na.rm = TRUE),

    q25_C1      = quantile(C1_score, 0.25, na.rm = TRUE),
    q75_C1      = quantile(C1_score, 0.75, na.rm = TRUE),
    IQR_C1      = IQR(C1_score, na.rm = TRUE),

    skewness_C1 = skewness(C1_score, na.rm = TRUE),
    kurtosis_C1 = kurtosis(C1_score, na.rm = TRUE),

    sem_C1      = sd_C1 / sqrt(n_cells),

    ci95_low    = mean_C1 - 1.96 * sem_C1,
    ci95_high   = mean_C1 + 1.96 * sem_C1,

    pct_high_C1 = mean(C1_score > 0.5) * 100
  ) %>%
  arrange(desc(mean_C1))

Sample_Cluster_Summary

write_csv(
  Sample_Cluster_Summary,
  file = "output/tables/Sample_Cluster_Summary_of_C1.csv"
)

```

```{r}

# KRUSKAL-WALLIS TEST
kruskal.test(C1_score ~ sample, data = meta)

# PAIRWISE WILCOXON TESTS
pairwise_res <- pairwise.wilcox.test(
  x = meta$C1_score,
  g = meta$sample,
  p.adjust.method = "BH"
)

pairwise_res

# EFFECT SIZE
effect_sizes <- meta %>%
  wilcox_effsize(C1_score ~ sample)

```

# Section 11. Linear Models for Sample and Annotation Effects
```{r}

library(broom)
library(jtools)

# Pseudotime model (main model)
m_traj <- lm(
  C1_score ~ S.Score + G2M.Score + percent.mt + percent_mito_ribo + sample,
  data = meta
)

tidy(m_traj)

summ(
  m_traj,
  scale = TRUE,
  confint = TRUE,
  robust = "HC3"
)

library(car)

Anova(m_traj, type = 2)

# Cell Type Model
m_celltype <- lm(C1_score ~ predicted.ann_finest_level, data = meta)

tidy(m_celltype)

summ(m_celltype, scale = TRUE, confint = TRUE, robust = TRUE)
Anova(m_celltype, type = 2)


```


# Section 12. Cluster-Level Summary Statistics

This block tests whether C1 differs across clusters and estimates the size of the cluster effect after basic covariate adjustment.

```{r}


# Cluster-level summary statistics
cluster_summary <- meta %>%
  dplyr::group_by(seurat_clusters) %>%
  dplyr::summarise(
    mean_C1   = mean(C1_score),
    median_C1 = median(C1_score),
    sd_C1     = sd(C1_score),
    n         = n()
  )

# Global non-parametric test across clusters
kruskal.test(C1_score ~ seurat_clusters, data = meta)

library(FSA)

# Post-hoc pairwise cluster comparisons (non-parametric)
dunnTest(
  C1_score ~ seurat_clusters,
  data = meta,
  method = "bh"
)

# Covariate-adjusted linear model
cluster_lm <- lm(
  C1_score ~ seurat_clusters + S.Score + G2M.Score +
    sample + percent_mito_ribo,
  data = meta
)

tidy(cluster_lm)

# Model summary
summary(cluster_lm)

# Clean regression table with diagnostics
summ(
  cluster_lm,
  scale = TRUE,
  confint = TRUE,
  vifs = TRUE,
  robust = "HC3"
)

# ANOVA-style decomposition of variance
cluster_aov <- aov(
  C1_score ~ seurat_clusters + S.Score + G2M.Score +
    sample + percent_mito_ribo,
  data = meta
)

cluster_aov

# Post-hoc cluster comparisons (parametric)
TukeyHSD(cluster_aov, "seurat_clusters")

library(effectsize)

eta_squared(cluster_aov, partial = TRUE)

library(emmeans)

emm <- emmeans(cluster_lm, ~ seurat_clusters)

pairs(emm, adjust = "tukey")

emm2 <- emmeans(cluster_lm, ~ seurat_clusters | sample)

pairs(emm2)


```

# Section 13. Cluster-Level Visualization

These plots turn the cluster-level statistics into compact figures for quick review.

```{r}

library(ggplot2)

ggplot(meta, aes(x = seurat_clusters, y = C1_score, fill = seurat_clusters)) +
  geom_violin(fill = "grey90", color = NA) +
  geom_boxplot(width = 0.15, outlier.size = 0.2) +
    stat_compare_means(
    label = "p.signif",
    hide.ns = TRUE
  ) +
  theme_pubr(base_size = 12) +
  scale_color_npg() +
  labs(
    title = "C1 activity varies across transcriptional clusters",
    subtitle = "Non-parametric comparison with Kruskal–Wallis test tests",
    x = "Seurat clusters",
    y = "C1 score"
  ) +
    theme(legend.position = "none")

ggsave("output/figures/C1_cluster_distribution.png",
       width = 10,
       height = 6,
       dpi = 600)

library(emmeans)

emm_df <- as.data.frame(emmeans(cluster_lm, ~ seurat_clusters))

ggplot(emm_df, aes(x = seurat_clusters, y = emmean, color = seurat_clusters)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2) +
  theme_ggprism_mod(base_size = 12) +
  scale_color_npg() +
  labs(
    title = "Covariate-adjusted C1 score per cluster",
    y = "Estimated marginal mean (C1 score)",
    x = "Seurat clusters"
  )  +
    theme(legend.position = "none")

library(effectsize)

eta <- eta_squared(cluster_aov, partial = TRUE)

ggplot(eta, aes(x = Parameter, y = Eta2_partial, fill = Parameter)) +
  geom_col() +
  theme_ggprism_mod(base_size = 12) +
  scale_fill_npg() +
  coord_flip() +
  labs(
    title = "Variance explained by model components",
    x = NULL,
    y = "Partial eta²"
  ) +
    theme(legend.position = "none")

```

# Section 14. Correlation Analysis Across Progression Metrics
```{r}

library(correlation)

vars <- c(
  "C1_score",
  "CytoTRACE2",
  "slingPseudotime_1",
  "velocity_pseudotime",
  "velocity_length",
  "S.Score",
  "G2M.Score"
)

df <- meta[, vars]

cor_results <- correlation(df, method = "spearman", p_adjust = "BH")

summary(cor_results)

```

# Section 15. Cell-Cycle Confounding Check
```{r}

library(ggplot2)
library(viridis)
library(ggpubr)


# 1. Plot: C1 score vs G2M


p1 <- ggplot(meta, aes(C1_score, G2M.Score, color = CytoTRACE2)) +
  geom_point(size = 0.6, alpha = 0.6) +
  geom_smooth(
    method = "gam",
    formula = y ~ s(x),
    color = "black"
  ) +
  stat_cor(method = "spearman", color = "black") +
  scale_color_viridis_c() +
  theme_bw(base_size = 12) +
  labs(
    title = "Association between C1 score and G2M score",
    subtitle = "Colored by CytoTRACE2 (differentiation state)",
    x = "C1 score",
    y = "G2M score",
    color = "CytoTRACE2"
  )

# Display plot
p1

# Save plot
ggsave("output/figures/C1_vs_G2M.png", p1, width = 8, height = 6, dpi = 600)



# 2. Compute residuals (regress out confounders)


# Linear model: remove cell cycle + mitochondrial effects
model <- lm(C1_score ~ G2M.Score + S.Score + percent_mito_ribo, data = meta)

# Scaled residuals
meta$C1_residual <- scale(residuals(model))

# Add to Seurat object metadata
css_combined@meta.data$C1_residual <- meta$C1_residual



# 3. Plot: Residual C1 vs G2M
p2 <- ggplot(meta, aes(C1_residual, G2M.Score, color = CytoTRACE2)) +
  geom_point(size = 0.6, alpha = 0.6) +
  geom_smooth(
    method = "gam",
    formula = y ~ s(x),
    color = "black"
  ) +
  stat_cor(method = "spearman", color = "black") +
  scale_color_viridis_c() +
  theme_bw(base_size = 12) +
  labs(
    title = "Residual C1 score vs G2M score",
    subtitle = "C1 score corrected for cell cycle, mitochondrial and  ribosomal content",
    x = "C1 residual",
    y = "G2M score",
    color = "CytoTRACE2"
  )

p2

ggsave("output/figures/C1residual_vs_G2M.png", p2, width = 8, height = 6, dpi = 600)



# 4. Plot: Residual C1 vs original C1

p3 <- ggplot(meta, aes(C1_residual, C1_score, color = CytoTRACE2)) +
  geom_point(size = 0.6, alpha = 0.6) +
  geom_smooth(
    method = "gam",
    formula = y ~ s(x),
    color = "black"
  ) +
  stat_cor(method = "spearman", color = "black") +
  scale_color_viridis_c() +
  theme_bw(base_size = 12) +
  labs(
    title = "Residual vs original C1 score",
    subtitle = "Assessing effect of regression on C1 signal",
    x = "C1 residual",
    y = "C1 score",
    color = "CytoTRACE2"
  )

p3

ggsave("output/figures/C1residual_vs_C1.png", p3, width = 8, height = 6, dpi = 600)


p_tot <- ((p1 / p2) | p3) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "right")  # or "bottom"


p_tot

ggsave("output/figures/p_tot_C1residual_vs_C1.png", p_tot, width = 14, height = 10, dpi = 600)


```

# Section 16. Save Updated Object

```{r}

qs_save(object = css_combined,file = "output/objects/css_combined_processed.qs2",compress_level = 10, nthreads = 10)

```

# Section 17. Export Processed Object for Python
```{r}

DefaultAssay(object = css_combined) <- "RNA"

as.anndata(x = css_combined, file_path = "output/objects/", file_name = "css_combined_processed.h5ad")

```
