---
title: "MAP 04 - decoupler Bridge"
format: html
---

# Section 1. Overview

This notebook bridges the R/Seurat C1-scored checkpoint into a Python `decoupler` run and restores the resulting activity scores back into Seurat assays for the downstream regulatory notebook.

## Section 1.1. Reader Guide

- Input is the C1-scored Seurat object from `01_MAP_c1_scoring.qmd`.
- The bridge exports a stable matrix and metadata payload under `output/scRNA/objects/decoupler_bridge/`.
- Python rebuilds an AnnData object, runs `decoupler` ULM scoring for TFs and pathways, and writes stable cell-by-feature score tables.
- R re-imports those tables, validates cell identity and ordering, creates the `tfsulm` and `pathwaysulm` assays, and saves a new regulatory-ready checkpoint.

## Section 1.2. Notebook Flow

1. Load the scored Seurat object.
2. Export normalized RNA data and required metadata for Python.
3. Build the AnnData handoff and run `decoupler` in Python.
4. Re-import TF and pathway score tables into Seurat assays.
5. Save the updated object for `05_MAP_regulatory_programs.qmd`.

# Section 2. R Setup and Seurat Export

```{r}

set.seed(1234)

packages <- c(
  "tidyverse",
  "Matrix",
  "Seurat",
  "qs2",
  "data.table"
)

invisible(purrr::walk(packages, library, character.only = TRUE))

bridge_object_dir <- file.path("output", "scRNA", "objects", "decoupler_bridge")
bridge_table_dir <- file.path("output", "scRNA", "tables", "decoupler")

dir.create(bridge_object_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(bridge_table_dir, recursive = TRUE, showWarnings = FALSE)

input_object_path <- file.path("output", "scRNA", "objects", "GSE274934_merged_RNA_QC.qs2")
output_object_path <- file.path("output", "scRNA", "objects", "GSE274934_regulatory_ready.qs2")

matrix_path <- file.path(bridge_object_dir, "GSE274934_decoupler_input_matrix.mtx")
obs_path <- file.path(bridge_object_dir, "GSE274934_decoupler_input_obs.csv")
var_path <- file.path(bridge_object_dir, "GSE274934_decoupler_input_var.csv")
input_h5ad_path <- file.path(bridge_object_dir, "GSE274934_decoupler_input.h5ad")
output_h5ad_path <- file.path(bridge_object_dir, "GSE274934_decoupler_output.h5ad")
tf_score_path <- file.path(bridge_table_dir, "tf_ulm_scores.csv")
tf_padj_path <- file.path(bridge_table_dir, "tf_ulm_padj.csv")
pathway_score_path <- file.path(bridge_table_dir, "pathway_ulm_scores.csv")
pathway_padj_path <- file.path(bridge_table_dir, "pathway_ulm_padj.csv")
collectri_network_path <- file.path(bridge_table_dir, "collectri_human_network.csv")
progeny_network_path <- file.path(bridge_table_dir, "progeny_human_network.csv")

required_meta <- c(
  "sample",
  "orig.ident",
  "seurat_clusters",
  "predicted.ann_level_1",
  "predicted.ann_finest_level",
  "SCEVAN_class"
)

first_present <- function(candidates, meta_df, label) {
  hit <- candidates[candidates %in% colnames(meta_df)][1]

  if (is.na(hit)) {
    stop(
      "Missing required metadata for ", label, ". Checked: ",
      paste(candidates, collapse = ", ")
    )
  }

  hit
}

write_bridge_payload <- function(object, matrix_path, obs_path, var_path) {
  DefaultAssay(object) <- "RNA"
  expr_mat <- GetAssayData(object = object, assay = "RNA", slot = "data")

  if (!inherits(expr_mat, "dgCMatrix")) {
    expr_mat <- as(expr_mat, "dgCMatrix")
  }

  if (nrow(expr_mat) == 0 || ncol(expr_mat) == 0) {
    stop("RNA normalized matrix is empty. Cannot build decoupler bridge payload.")
  }

  missing_meta <- setdiff(required_meta, colnames(object@meta.data))
  if (length(missing_meta) > 0) {
    stop(
      "Missing required metadata columns in Seurat object: ",
      paste(missing_meta, collapse = ", ")
    )
  }

  c1_col <- first_present(c("C1_Core", "C1_score"), object@meta.data, "C1 score")

  obs <- object@meta.data |>
    tibble::rownames_to_column("cell") |>
    dplyr::select(
      cell,
      sample,
      orig.ident,
      seurat_clusters,
      dplyr::all_of(c1_col),
      predicted.ann_level_1,
      predicted.ann_finest_level,
      SCEVAN_class
    )

  colnames(obs)[colnames(obs) == c1_col] <- c1_col

  if (!("C1_Core" %in% colnames(obs)) && "C1_score" %in% colnames(obs)) {
    obs$C1_Core <- obs$C1_score
  }

  if (!("C1_score" %in% colnames(obs)) && "C1_Core" %in% colnames(obs)) {
    obs$C1_score <- obs$C1_Core
  }

  obs <- obs |>
    dplyr::select(
      cell,
      sample,
      orig.ident,
      seurat_clusters,
      C1_Core,
      C1_score,
      predicted.ann_level_1,
      predicted.ann_finest_level,
      SCEVAN_class
    )

  stopifnot(identical(obs$cell, colnames(expr_mat)))

  var_tbl <- tibble::tibble(gene = rownames(expr_mat))

  Matrix::writeMM(obj = expr_mat, file = matrix_path)
  data.table::fwrite(obs, file = obs_path)
  data.table::fwrite(var_tbl, file = var_path)

  cat(
    "Bridge export complete:",
    nrow(var_tbl), "genes and", nrow(obs), "cells written for Python.\n"
  )
}

GSE274934 <- qs_read(file = input_object_path, nthreads = 10)

write_bridge_payload(
  object = GSE274934,
  matrix_path = matrix_path,
  obs_path = obs_path,
  var_path = var_path
)

```

# Section 3. Python decoupler Run

```{python}

from pathlib import Path

import anndata as ad
import decoupler as dc
import pandas as pd
from scipy.io import mmread
from scipy import sparse

bridge_object_dir = Path("output/scRNA/objects/decoupler_bridge")
bridge_table_dir = Path("output/scRNA/tables/decoupler")

matrix_path = bridge_object_dir / "GSE274934_decoupler_input_matrix.mtx"
obs_path = bridge_object_dir / "GSE274934_decoupler_input_obs.csv"
var_path = bridge_object_dir / "GSE274934_decoupler_input_var.csv"
input_h5ad_path = bridge_object_dir / "GSE274934_decoupler_input.h5ad"
output_h5ad_path = bridge_object_dir / "GSE274934_decoupler_output.h5ad"
tf_score_path = bridge_table_dir / "tf_ulm_scores.csv"
tf_padj_path = bridge_table_dir / "tf_ulm_padj.csv"
pathway_score_path = bridge_table_dir / "pathway_ulm_scores.csv"
pathway_padj_path = bridge_table_dir / "pathway_ulm_padj.csv"
collectri_network_path = bridge_table_dir / "collectri_human_network.csv"
progeny_network_path = bridge_table_dir / "progeny_human_network.csv"


def obsm_to_df(adata, key):
    extracted = dc.pp.get_obsm(adata=adata, key=key)
    if isinstance(extracted, pd.DataFrame):
        return extracted.copy()
    if hasattr(extracted, "to_df"):
        return extracted.to_df()
    return pd.DataFrame(extracted, index=adata.obs_names)


obs = pd.read_csv(obs_path)
var = pd.read_csv(var_path)
expr = mmread(matrix_path)

if not sparse.issparse(expr):
    expr = sparse.csr_matrix(expr)
else:
    expr = expr.tocsr()

adata = ad.AnnData(
    X=expr.T,
    obs=obs.set_index("cell"),
    var=var.set_index("gene"),
)

adata.obs_names_make_unique()
adata.var_names_make_unique()
adata.write_h5ad(input_h5ad_path)

collectri = dc.get_network("collecTRI", collection="human")
collectri = collectri.drop_duplicates().reset_index(drop=True)
collectri.to_csv(collectri_network_path, index=False)

dc.mt.ulm(data=adata, net=collectri)

tf_scores = obsm_to_df(adata, "score_ulm")
tf_scores.index.name = "cell"
tf_scores.reset_index().to_csv(tf_score_path, index=False)

if "padj_ulm" in adata.obsm:
    tf_padj = obsm_to_df(adata, "padj_ulm")
    tf_padj.index.name = "cell"
    tf_padj.reset_index().to_csv(tf_padj_path, index=False)

pathway_adata = adata.copy()
progeny = dc.op.progeny(organism="human")
progeny = progeny.drop_duplicates().reset_index(drop=True)
progeny.to_csv(progeny_network_path, index=False)

dc.mt.ulm(data=pathway_adata, net=progeny)

pathway_scores = obsm_to_df(pathway_adata, "score_ulm")
pathway_scores.index.name = "cell"
pathway_scores.reset_index().to_csv(pathway_score_path, index=False)

if "padj_ulm" in pathway_adata.obsm:
    pathway_padj = obsm_to_df(pathway_adata, "padj_ulm")
    pathway_padj.index.name = "cell"
    pathway_padj.reset_index().to_csv(pathway_padj_path, index=False)

adata.obsm["pathway_score_ulm"] = pathway_scores.loc[adata.obs_names]
if "padj_ulm" in pathway_adata.obsm:
    adata.obsm["pathway_padj_ulm"] = obsm_to_df(pathway_adata, "padj_ulm").loc[adata.obs_names]

adata.write_h5ad(output_h5ad_path)

print(
    "decoupler outputs written:",
    f"TF scores={tf_scores.shape}, pathway scores={pathway_scores.shape}"
)

```

# Section 4. R Re-import into Seurat

```{r}

read_activity_matrix <- function(csv_path) {
  score_tbl <- data.table::fread(csv_path)

  if (!("cell" %in% colnames(score_tbl))) {
    stop("Activity table is missing the required 'cell' column: ", csv_path)
  }

  cell_ids <- score_tbl$cell
  score_tbl$cell <- NULL

  if (anyDuplicated(cell_ids) > 0) {
    stop("Duplicated cell barcodes detected in ", csv_path)
  }

  if (anyDuplicated(colnames(score_tbl)) > 0) {
    stop("Duplicated feature names detected in ", csv_path)
  }

  score_mat <- as.matrix(score_tbl)
  rownames(score_mat) <- cell_ids

  t(score_mat)
}

align_activity_matrix <- function(feature_by_cell, seurat_cells, label) {
  if (anyDuplicated(colnames(feature_by_cell)) > 0) {
    stop("Duplicated cells detected in ", label, " activity matrix.")
  }

  missing_cells <- setdiff(seurat_cells, colnames(feature_by_cell))
  extra_cells <- setdiff(colnames(feature_by_cell), seurat_cells)

  if (length(missing_cells) > 0 || length(extra_cells) > 0) {
    stop(
      label, " cell mismatch between Seurat and imported score table. ",
      "Missing in table: ", length(missing_cells),
      ". Extra in table: ", length(extra_cells), "."
    )
  }

  aligned <- feature_by_cell[, seurat_cells, drop = FALSE]

  stopifnot(identical(colnames(aligned), seurat_cells))
  aligned
}

row_zscore <- function(feature_by_cell) {
  scaled <- t(scale(t(feature_by_cell)))
  scaled[is.na(scaled)] <- 0
  scaled
}

add_activity_assay <- function(object, assay_name, feature_by_cell) {
  assay_obj <- CreateAssayObject(data = feature_by_cell)
  object[[assay_name]] <- assay_obj
  object <- SetAssayData(
    object = object,
    assay = assay_name,
    slot = "scale.data",
    new.data = row_zscore(feature_by_cell)
  )

  object
}

tf_scores <- read_activity_matrix(tf_score_path)
pathway_scores <- read_activity_matrix(pathway_score_path)

tf_scores <- align_activity_matrix(
  feature_by_cell = tf_scores,
  seurat_cells = colnames(GSE274934),
  label = "TF"
)

pathway_scores <- align_activity_matrix(
  feature_by_cell = pathway_scores,
  seurat_cells = colnames(GSE274934),
  label = "Pathway"
)

GSE274934 <- add_activity_assay(
  object = GSE274934,
  assay_name = "tfsulm",
  feature_by_cell = tf_scores
)

GSE274934 <- add_activity_assay(
  object = GSE274934,
  assay_name = "pathwaysulm",
  feature_by_cell = pathway_scores
)

DefaultAssay(GSE274934) <- "tfsulm"

stopifnot(ncol(GSE274934) == ncol(tf_scores))
stopifnot(identical(colnames(GSE274934), colnames(tf_scores)))
stopifnot(identical(colnames(GSE274934), colnames(pathway_scores)))
stopifnot(nrow(GetAssayData(GSE274934, assay = "tfsulm", slot = "data")) > 0)
stopifnot(nrow(GetAssayData(GSE274934, assay = "pathwaysulm", slot = "data")) > 0)

cat(
  "Regulatory-ready object contains",
  nrow(GetAssayData(GSE274934, assay = "tfsulm", slot = "data")),
  "TF features and",
  nrow(GetAssayData(GSE274934, assay = "pathwaysulm", slot = "data")),
  "pathway features across",
  ncol(GSE274934),
  "cells.\n"
)

qs_save(
  GSE274934,
  file = output_object_path,
  nthreads = 10,
  compress_level = 10
)

```
