---
title: "MAP 05 - Regulatory Programs"
format: html
---

# Section 1. Overview

This notebook interprets the C1 plasticity axis through regulatory, pathway, surface-marker, secreted-protein, and hallmark-style activity programs.

**MAP role:** reduced interpretation notebook. Keep TF activity, compact pathway/Hallmark summaries, and surface or secreted candidates that feed the multiome hypotheses. Do not expand this into broad exploratory enrichment unless it changes the regulator shortlist.

## Section 1.1. Reader Guide

- Input is the processed Seurat object carrying the C1 score, clusters, and cell-state annotations.
- The notebook estimates TF activity, pathway activity, surface-marker associations, secreted-protein activity, and hallmark-style scores.
- Main outputs are regulatory plots, heatmaps, correlation tables, marker tables, and activity assays stored in the working object.
- Read this as the mechanistic interpretation layer: it asks which regulators and programs track with the plasticity gradient.

## Section 1.2. MAP Section Boundaries

### Section 1.2.1. Notebook Flow

1. Setup and load scored object.
2. Infer TF activity and correlate TFs with C1.
3. Summarize selected target-expression maps and TF heatmaps.
4. Infer compact pathway and Hallmark programs.
5. Keep surface and secreted markers only when they support follow-up hypotheses.
6. Import MAP 08 metabolic scores only for lightweight cross-layer checks.
7. Export regulator, pathway, secreted-protein, Hallmark, and metabolic-link tables for scATAC hypothesis testing.
8. Save the regulator-enriched Seurat object checkpoint here.

## Section 1.3. Setup

This section loads the required packages and restores the Seurat object used as the starting point for downstream analyses.


```{r include=FALSE}

set.seed(1234)

# ==============================================================================
# Session setup
# ==============================================================================

# Reduce OpenMP/BLAS resource warnings on macOS and avoid oversubscription.
# Run this before loading heavy packages.
Sys.setenv(
  OMP_NUM_THREADS = "1",
  OPENBLAS_NUM_THREADS = "1",
  MKL_NUM_THREADS = "1",
  VECLIB_MAXIMUM_THREADS = "1",
  NUMEXPR_NUM_THREADS = "1"
)

if (requireNamespace("RhpcBLASctl", quietly = TRUE)) {
  RhpcBLASctl::blas_set_num_threads(1)
  RhpcBLASctl::omp_set_num_threads(1)
}

# ==============================================================================
# Packages
# ==============================================================================

packages <- c(
  "tidyverse",
  "Matrix",
  "Seurat",
  "SeuratWrappers",
  "patchwork",
  "data.table",
  "viridis",
  "RColorBrewer",
  "qs2",
  "ggsci",
  "scCustomize",
  "future",
  "ggpubr",
  "ggExtra",
  "skimr",
  "lisi",
  "SeuratData"
)

invisible(purrr::walk(packages, library, character.only = TRUE))

# Prefer scrapper replacements when working directly with SingleCellExperiment.
# Old deprecated calls:
#   normalizeCounts(sce)
#   librarySizeFactors(sce)
#
# Updated calls:
#   sce <- scrapper::normalizeCounts(sce)
#   sce <- scrapper::centerSizeFactors(sce)

# ==============================================================================
# Parallelization
# ==============================================================================

options(future.globals.maxSize = 25 * 1024^3)

future::plan(
  strategy = future::multicore,
  workers = 5
)

# ==============================================================================
# Shared plotting theme
# ==============================================================================

umap_theme <- ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(size = 16, face = "bold"),
    plot.subtitle = ggplot2::element_text(size = 11, color = "grey30"),
    panel.grid = ggplot2::element_blank(),
    axis.title = ggplot2::element_text(size = 13),
    axis.text = ggplot2::element_blank(),
    axis.ticks = ggplot2::element_blank(),
    axis.line = ggplot2::element_blank(),
    legend.title = ggplot2::element_text(size = 11),
    legend.text = ggplot2::element_text(size = 10)
  )

regulatory_ready_path <- file.path(
  "output", "scRNA", "objects", "GSE274934_regulatory_ready.qs2"
)

collectri_network_path <- file.path(
  "output", "scRNA", "tables", "decoupler", "collectri_human_network.csv"
)

GSE274934 <- qs_read(file = regulatory_ready_path, nthreads = 10)

if (!("tfsulm" %in% names(GSE274934@assays))) {
  stop(
    "The regulatory-ready checkpoint is missing the 'tfsulm' assay. ",
    "Run 04_MAP_decoupler_bridge.qmd before MAP 05."
  )
}

if (!file.exists(collectri_network_path)) {
  stop(
    "Missing exported CollecTRI network table at ",
    collectri_network_path,
    ". Run 04_MAP_decoupler_bridge.qmd before MAP 05."
  )
}

net <- vroom::vroom(collectri_network_path, show_col_types = FALSE)

```

# Section 1.4. Bridge Handoff Check
```{r}

cat(
  "Loaded regulatory-ready checkpoint with",
  ncol(GSE274934),
  "cells,",
  nrow(GSE274934@assays$tfsulm),
  "TF activity features, and",
  if ("pathwaysulm" %in% names(GSE274934@assays)) nrow(GSE274934@assays$pathwaysulm) else 0,
  "pathway activity features.\n"
)

```

# Section 2. Regulatory Activity Inference with decoupler Bridge Output



# Section 4. Target Expression Maps for Selected TFs
```{r}

DefaultAssay(GSE274934) <- "RNA"
HOXB7 <- net |>
  filter(source == "HOXB7")

p <- FeaturePlot_scCustom(
  seurat_object = GSE274934,
  features = HOXB7$target,num_columns = 5,
  reduction = "umap",
  #ncol = 4,              # controls wrapping
  combine = TRUE
) & theme_bw() 

p

ggsave("output/figures/umap_HOXB7_targets.png", p, width = 18, height = 12,dpi = 600)
ggsave("output/figures/umap_HOXB7_targets.svg", p, width = 18, height = 12,dpi = 600,device = "svg")

DefaultAssay(GSE274934) <- "RNA"
TEAD4 <- net |>
  filter(source == "TEAD4")

p <- FeaturePlot_scCustom(
  seurat_object = GSE274934,
  features = TEAD4$target,num_columns = 3,
  #na_cutoff = NA,
  reduction = "umap",
  #ncol = 4,              # controls wrapping
  combine = TRUE
) & theme_bw() & umap_theme 

p

ggsave("output/figures/umap_TEAD4_targets.png", p, width = 15, height = 10,dpi = 600)

```

# Section 5. TF Activity and Expression Comparison Plots
```{r}

DefaultAssay(GSE274934) <- "tfsulm"

p_tead4_org <- FeaturePlot_scCustom(
  GSE274934,
  features = "TEAD4",
  split.by = "orig.ident",
  num_columns = 2
) &
  umap_theme

ggsave("output/figures/TEAD4_org.png", p_tead4_org, width = 12, height = 8, dpi = 600, bg = "white")

p_tead4_activity <- FeaturePlot_scCustom(
  GSE274934,
  features = "TEAD4",
  na_cutoff = NA
) +
  ggtitle("TEAD4 activity") +
  labs(subtitle = "Regulon-based activity score") +
  umap_theme

ggsave(
  "output/figures/TEAD4_activity.png",
  p_tead4_activity,
  width = 7,
  height = 6,
  dpi = 600
)

DefaultAssay(GSE274934) <- "SCT"

p_tead4_expression <- FeaturePlot_scCustom(
  GSE274934,
  features = "TEAD4"
) +
  ggtitle("TEAD4 expression") +
  labs(subtitle = "RNA expression on UMAP") +
  umap_theme

ggsave(
  "output/figures/TEAD4_expression.png",
  p_tead4_expression,
  width = 7,
  height = 6,
  dpi = 600
)

```

# Section 6. Cluster-Level TF Heatmap
```{r}

library(ComplexHeatmap)
library(circlize)
library(dplyr)
library(grid)

n_tfs <- 50

dir.create("output/figures", recursive = TRUE, showWarnings = FALSE)

# Extract TF activity matrix
# Result: cells x TFs
acts_mat <- t(as.matrix(GSE274934@assays$tfsulm@data))

# Align metadata to the TF activity matrix
meta <- GSE274934@meta.data
meta <- meta[rownames(acts_mat), , drop = FALSE]

stopifnot(all(rownames(acts_mat) == rownames(meta)))

# Cluster labels
cluster <- as.character(meta$seurat_clusters)

# Count cells per cluster
cluster_counts <- table(cluster)

# Sum TF activity by cluster
cluster_sum_mat <- rowsum(
  acts_mat,
  group = cluster,
  reorder = FALSE
)

# Convert cluster sums into cluster means
# Result: clusters x TFs
cluster_mean_mat <- sweep(
  cluster_sum_mat,
  MARGIN = 1,
  STATS = as.numeric(cluster_counts[rownames(cluster_sum_mat)]),
  FUN = "/"
)

# Select TFs with strongest variation across clusters
tf_sd <- apply(cluster_mean_mat, 2, sd, na.rm = TRUE)

top_tfs <- names(sort(tf_sd, decreasing = TRUE))[
  seq_len(min(n_tfs, length(tf_sd)))
]

cluster_mean_mat <- cluster_mean_mat[, top_tfs, drop = FALSE]

# Scale each TF across clusters
# Since cluster_mean_mat is clusters x TFs, scale() scales each TF column
# This is equivalent to pheatmap(scale = "row") after transposing
cluster_mean_scaled <- scale(cluster_mean_mat)
cluster_mean_scaled[is.na(cluster_mean_scaled)] <- 0

# Heatmap matrix
# Rows = TFs, columns = clusters
plot_mat <- t(cluster_mean_scaled)

# Unscaled matrix used only for the TF activity distribution annotation
plot_mat_unscaled <- t(cluster_mean_mat)

# Cluster-level annotations
cluster_annot <- meta %>%
  dplyr::group_by(seurat_clusters) %>%
  dplyr::summarise(
    mean_C1_Core = mean(C1_Core, na.rm = TRUE),
    n_cells = dplyr::n(),
    .groups = "drop"
  ) %>%
  as.data.frame()

cluster_annot$seurat_clusters <- as.character(cluster_annot$seurat_clusters)

# Match annotation order to the heatmap columns safely
cluster_annot <- cluster_annot[
  match(colnames(plot_mat), cluster_annot$seurat_clusters),
  ,
  drop = FALSE
]

rownames(cluster_annot) <- cluster_annot$seurat_clusters

stopifnot(!any(is.na(cluster_annot$seurat_clusters)))
stopifnot(identical(rownames(cluster_annot), colnames(plot_mat)))

# Color scale for scaled TF activity
tf_col_fun <- colorRamp2(
  c(-2, -1, 0, 1, 2),
  c("#053061", "#4393C3", "#F7F7F7", "#D6604D", "#67001F")
)

# Color scale for mean C1_Core
c1_col_fun <- colorRamp2(
  c(
    min(cluster_annot$mean_C1_Core, na.rm = TRUE),
    median(cluster_annot$mean_C1_Core, na.rm = TRUE),
    max(cluster_annot$mean_C1_Core, na.rm = TRUE)
  ),
  c("#313695", "#F7F7F7", "#A50026")
)

# Top annotations for clusters
top_ha <- HeatmapAnnotation(
  `Mean C1 raw` = cluster_annot$mean_C1_Core,
  `Cells` = anno_barplot(
    cluster_annot$n_cells,
    gp = gpar(fill = "grey40", col = NA),
    border = FALSE,
    height = unit(1, "cm")
  ),
  col = list(
    `Mean C1 raw` = c1_col_fun
  ),
  annotation_name_side = "left",
  annotation_name_gp = gpar(fontsize = 8),
  annotation_legend_param = list(
    `Mean C1 raw` = list(title = "Mean C1-Core")
  )
)

# Left annotation for TFs
# Boxplots show the distribution of unscaled cluster-level mean activity
left_ha <- rowAnnotation(
  `Activity distribution` = anno_boxplot(
    plot_mat_unscaled,
    which = "row",
    border = FALSE,
    outline = FALSE,
    axis = TRUE,
    width = unit(2.3, "cm"),
    gp = gpar(fill = "grey70", col = "grey40")
  ),
  annotation_name_rot = 90,
  annotation_name_gp = gpar(fontsize = 9)
)

# Title and subtitle
main_title <- "Cluster-level TF activity heatmap"

subtitle <- paste0(
  nrow(plot_mat),
  " TFs displayed | Pearson distance | average linkage"
)
heatmap_title <- paste0(
  "Cluster-level TF activity heatmap\n",
  nrow(plot_mat),
  " TFs displayed | row z-score | Pearson distance | average linkage"
)

ht_opt$TITLE_PADDING = unit(c(10, 10), "points")

png(
  filename = "output/figures/TF_heatmap_simple.png",
  width = 8,
  height = 10,
  units = "in",
  res = 600
)

ht <- Heatmap(
  plot_mat,
  name = "TF activity",
  col = tf_col_fun,

  # left_annotation = left_ha,
  top_annotation = top_ha,

  cluster_rows = TRUE,
  cluster_columns = TRUE,

  clustering_distance_rows = "pearson",
  clustering_distance_columns = "pearson",
  clustering_method_rows = "average",
  clustering_method_columns = "average",

  show_row_dend = TRUE,
  show_column_dend = TRUE,

  row_title = "TFs",
  column_title = heatmap_title,

  column_title_gp = gpar(
    fontsize = 13,
    fontface = "bold",
    lineheight = 1.15
  ),
  row_title_gp = gpar(
    fontsize = 11,
    fontface = "bold"
  ),

  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 0,

  border = TRUE,
  rect_gp = gpar(col = "grey90", lwd = 0.3),

  heatmap_legend_param = list(
    title = "Row z-score",
    at = c(-2, -1, 0, 1, 2),
    labels = c("Low", "-1", "0", "+1", "High")
  )
)

draw(
  ht,
  heatmap_legend_side = "right",
  annotation_legend_side = "right"
)

dev.off()


pdf(
  file = "output/figures/TF_heatmap_simple.pdf",
  width = 10,
  height = 12,
  useDingbats = FALSE
)

ht <- Heatmap(
  plot_mat,
  name = "TF activity",
  col = tf_col_fun,

  # left_annotation = left_ha,
  top_annotation = top_ha,

  cluster_rows = TRUE,
  cluster_columns = TRUE,

  clustering_distance_rows = "pearson",
  clustering_distance_columns = "pearson",
  clustering_method_rows = "average",
  clustering_method_columns = "average",

  show_row_dend = TRUE,
  show_column_dend = TRUE,

  row_title = "TFs",
  column_title = heatmap_title,

  column_title_gp = gpar(
    fontsize = 13,
    fontface = "bold",
    lineheight = 1.15
  ),
  row_title_gp = gpar(
    fontsize = 11,
    fontface = "bold"
  ),

  row_names_gp = gpar(fontsize = 6),
  column_names_gp = gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 0,

  border = TRUE,
  rect_gp = gpar(col = "grey90", lwd = 0.3),

  heatmap_legend_param = list(
    title = "Row z-score",
    at = c(-2, -1, 0, 1, 2),
    labels = c("Low", "-1", "0", "+1", "High")
  )
)

draw(
  ht,
  heatmap_legend_side = "right",
  annotation_legend_side = "right"
)

dev.off()

ht_opt(RESET = TRUE)

```

# Section 7. TF Correlation Along the C1 Axis
```{r}

library(future)
library(progressr)       # ← main package
library(future.apply)

plan(multicore, workers = 5)     # or multicore on Linux/Mac

# Activate a nice default progress bar style
handlers(global = TRUE)                  # so you don't need to pass handler every time
handlers("cli")                     # clean text bar (default)
#handlers("cli")

tf_activity <- as.matrix(GSE274934@assays$tfsulm@scale.data)
tf_names <- rownames(tf_activity)
C1_score <- GSE274934@meta.data$C1_Core

with_progress({
  # Create a progressor object (tells how many steps = how many TFs)
  p <- progressor(along = tf_names)         
  
  tf_pseudotime_cor <- future_lapply(tf_names, function(tf) {
    p(sprintf("TF %s", tf))                  # updates the bar + shows current TF
    
    r <- cor.test(tf_activity[tf, ], C1_score, method = "spearman",exact = F)
    
    data.frame(
      TF       = tf,
      rho      = r$estimate[["rho"]],
      p.value  = r$p.value,
      stringsAsFactors = FALSE
    )
  }, future.seed = TRUE) |> 
    bind_rows()
})

# Add FDR
tf_pseudotime_cor <- tf_pseudotime_cor |>
  mutate(fdr = p.adjust(p.value, method = "fdr"))

sig_tfs <- tf_pseudotime_cor |>
  #dplyr::filter(abs(rho) >= 0.3, fdr <= 0.05) |>
  dplyr::filter(fdr <= 0.05) |>
  dplyr::arrange(-rho) |>
  dplyr::mutate(
    direction = dplyr::if_else(rho > 0, "increasing", "decreasing")
  )

fwrite(x = sig_tfs, file = "output/tables/cor_spearman_TF_along_C1.csv")

print(sig_tfs |> slice_head(n = 15))

n_each <- 8

top_pos <- sig_tfs |>
  dplyr::arrange(desc(rho)) |>
  dplyr::slice_head(n = n_each)

top_neg <- sig_tfs |>
  dplyr::arrange(rho) |>
  dplyr::slice_head(n = n_each)

top_tfs <- dplyr::bind_rows(top_pos, top_neg) |>
  dplyr::pull(TF)


plot_data <- tibble(
  cell       = colnames(tf_activity),
  pseudotime = C1_score
) |>
  bind_cols(as_tibble(t(tf_activity[top_tfs, , drop = FALSE]))) |>
  pivot_longer(
    cols      = all_of(top_tfs),
    names_to  = "TF",
    values_to = "activity"
  ) |>
  mutate(TF = factor(TF, levels = sig_tfs$TF))

# Plot (same as before)
tf_plot_8 <- ggplot(plot_data, aes(x = pseudotime, y = activity)) +
  scattermore::geom_scattermore( color = "grey30") + #alpha = 0.08, size = 0.7,
  geom_smooth(
    method  = "gam",
    formula = y ~ s(x, bs = "cs"),
    color   = "#d73027", fill = "#d73027", alpha = 0.12,
    linewidth = 0.9, se = TRUE
  ) +
  facet_wrap(~ TF, scales = "free_y", ncol = 4) +
  labs(
    title    = "Top TF activity trends along C1-Core",
    subtitle = sprintf("Spearman |ρ|, FDR ≤ 0.05"),#  (top %d up + %d down)", n_each, n_each),
    x = "C1 score", y = "TF activity"
  ) +
  theme_bw(base_size = 12) +
  ggpubr::stat_cor(method = "spearman") +
  theme(
    strip.text = element_text(face = "bold", size = 10),
    panel.grid.minor = element_blank()
  )
tf_plot_8

ggsave(plot = tf_plot_8,filename = "output/figures/cor_spearman_TF_along_C1_top_16.png", width = 14, height = 10,dpi = 600)

library(dplyr)

cor_TF <- vroom::vroom("output/tables/cor_spearman_TF_along_C1.csv")

# Top 20 pro-plasticity (positive rho)
top_pos <- cor_TF %>%
  filter(fdr < 0.05, rho > 0) %>%
  arrange(desc(rho)) %>%
  slice_head(n = 50) %>%
  mutate(direction = "pro-plasticity")

# Top 20 pro-differentiation (negative rho)
top_neg <- cor_TF %>%
  filter(fdr < 0.05, rho < 0) %>%
  arrange(rho) %>%
  slice_head(n = 50) %>%
  mutate(direction = "pro-differentiation")

# Final table
tf_table <- bind_rows(top_pos, top_neg) %>%
  select(TF, `rho vs C1` = rho, direction, FDR = fdr)

fwrite(x = tf_table, file = "output/tables/cor_spearman_TF_alg_C1_top_100.csv")

```


# Section 11. Surface Markers of Plastic versus Differentiated States

```{r}


# Set default assay
DefaultAssay(GSE274934) <- "SCT"

# Get list of surface proteins (from CellMarker or GO:0009986)
surface_proteins <- msigdbr(species = "Homo sapiens", collection = "C5") %>%
  filter(gs_name == "GOCC_CELL_SURFACE") %>%
  pull(gene_symbol) %>%
  unique()

# Intersect with expressed genes in Seurat object
expressed_surface <- intersect(surface_proteins, rownames(GSE274934))

library(future)
library(progressr)
library(future.apply)
library(dplyr)

plan(multicore, workers = 4)
handlers(global = TRUE)
handlers("progress")

surface_genes <- expressed_surface
plasticity_score <- as.numeric(GSE274934$C1_score)
surface_expr <- GSE274934@assays$SCT@data[expressed_surface, ]

with_progress({
  p <- progressor(along = surface_genes)
  
  surface_marker_analysis <- future_lapply(seq_along(surface_genes), function(i) {
    gene <- surface_genes[i]
    p(sprintf("Gene %s", gene))
    
    # Extract sparse row inside the worker (avoid copying whole matrix)
    gene_vec <- as.numeric(surface_expr[gene, ])
    
    r <- cor.test(gene_vec, plasticity_score, method = "spearman", exact = FALSE)
    
    data.frame(
      gene        = gene,
      correlation = r$estimate[["rho"]],
      p.value     = r$p.value,
      stringsAsFactors = FALSE
    )
  }, future.seed = TRUE) |> bind_rows()
})


# Adjust p-values and filter significant markers
surface_markers <- surface_marker_analysis %>%
  mutate(fdr = p.adjust(p.value, method = "BH")) %>%
  filter(fdr < 0.05, abs(correlation) > 0.4) %>%
  arrange(-correlation)

print(tibble(surface_markers) |> slice_head(n = 15))

fwrite(x = surface_markers, file = "output/tables/surface_markers_alg_pseudotime.csv")

# Top markers for HIGH and LOW plasticity
high_plasticity_markers <- surface_markers %>% filter(correlation > 0.5) %>% head(5)
low_plasticity_markers  <- surface_markers %>% filter(correlation < -0.5) %>% head(5)

cat("HIGH PLASTICITY surface markers:\n")
print(high_plasticity_markers)

cat("LOW PLASTICITY surface markers:\n")
print(low_plasticity_markers)

n_each <- 16
top_surface_markers <- surface_markers |>
  slice_head(n = n_each) |>
  ungroup() |>
  pull(gene) |>
  unique()

plot_data <- tibble(
  cell       = colnames(surface_expr),
  plasticity_score = plasticity_score
) |>
  bind_cols(as_tibble(t(surface_expr[top_surface_markers, , drop = FALSE]))) |>
  pivot_longer(
    cols      = all_of(top_surface_markers),
    names_to  = "gene",
    values_to = "expression"
  ) |>
  mutate(gene = factor(gene, levels = surface_markers$gene))

# Plot (same as before)
FACS_8 <- ggplot(plot_data, aes(x = plasticity_score, y = expression)) +
  scattermore::geom_scattermore( color = "grey30") + #alpha = 0.08, size = 0.7,
  geom_smooth(
    method  = "gam",
    formula = y ~ s(x, bs = "cs"),
    color   = "#d73027", fill = "#d73027", alpha = 0.12,
    linewidth = 0.9, se = TRUE
  ) +
  facet_wrap(~ gene, scales = "free_y", ncol = 4) +
  ggpubr::stat_cor(method = "spearman", size = 3) +
  labs(
    title    = "Top gene expression trends along plasticity index",
    subtitle = sprintf("Spearman |ρ| ≥ 0.4, FDR ≤ 0.05"),#  (top %d up + %d down)", n_each, n_each),
    x = "Plasticity index", y = "Gene expression"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold", size = 10),
    panel.grid.minor = element_blank()
  )
FACS_8

ggsave(plot = FACS_8,filename = "output/figures/FACS_8.png", width = 12, height = 8,dpi = 600)
ggsave(plot = FACS_8,filename = "output/figures/FACS_8.svg", width = 12, height = 8,device = "svg")


surface_markers_high <- vroom::vroom("output/tables/surface_markers_alg_pseudotime.csv") %>% 
  filter(correlation > 0.4)


fwrite(x = surface_markers_high, file = "output/tables/surface_markers_high.csv")

DotPlot_scCustom(GSE274934 = GSE274934, features = surface_markers_high$gene)

DoHeatmap(object = GSE274934, features = surface_markers_high$gene)


library(SeuratExtend)
toplot <- CalcStats(GSE274934, features = surface_markers_high$gene, method = "zscore", order = "p")
# Create a basic heatmap
Heatmap(toplot, lab_fill = "zscore")

high_FACS <- combined_markers_high|>
  filter(...1 %in% surface_proteins)

high_FACS

```

# Section 12. Secreted Protein Activity Inference

This block adds SecAct as a secreted-protein signaling layer. The initial output is a compact top-protein table; the next subsections turn the activity matrix into cluster-level heatmaps.

```{r}

library(SecAct)
GSE274934 <- SecAct.activity.inference.scRNAseq(GSE274934, cellType_meta="predicted.ann_finest_level")

# SecretedProteinActivity$zscore stores activity
act <- GSE274934@misc$SecAct_output$SecretedProteinActivity$zscore

# show activity
head(act)

# select top secreted proteins
n <- 50
SPs.1 <- names(sort(act[,1],decreasing=T))[1:n]
SPs.2 <- names(sort(act[,2],decreasing=T))[1:n]
SPs.3 <- names(sort(act[,3],decreasing=T))[1:n]

SPs <- unique(c(SPs.1, SPs.2, SPs.3))
fg.mat <- act[SPs,]

fg.mat |>
  as.data.frame() |>
  rownames_to_column("Secreted_Protein") |>
  write_csv(file = "output/scRNA/tables/TOP_50_SecretedProtein_SectAct.csv")

```

## Section 12.1. SecAct Top Active Proteins Per Cluster

This heatmap asks which secreted proteins are strongest within each epithelial cluster.
```{r}

library(ComplexHeatmap)
library(circlize)
library(dplyr)
library(grid)

n_per_cluster <- 20
max_sps <- 100

dir.create("output/figures", recursive = TRUE, showWarnings = FALSE)

# Full SecAct activity matrix
# Rows = secreted proteins
# Columns = clusters
act_mat <- as.matrix(act)
colnames(act_mat) <- as.character(colnames(act_mat))

# Select top active SPs per cluster
top_sps_by_cluster <- lapply(
  colnames(act_mat),
  function(cl) {
    names(sort(act_mat[, cl], decreasing = TRUE))[
      seq_len(min(n_per_cluster, nrow(act_mat)))
    ]
  }
)

names(top_sps_by_cluster) <- colnames(act_mat)

# Union of top SPs across all clusters
selected_sps <- unique(unlist(top_sps_by_cluster))

# Foreground matrix
fg.mat <- act_mat[selected_sps, , drop = FALSE]

# Remove proteins with zero or missing variation across clusters
keep_var <- apply(fg.mat, 1, sd, na.rm = TRUE) > 0
fg.mat <- fg.mat[keep_var, , drop = FALSE]

# If too many SPs are selected, keep the most variable ones
if (nrow(fg.mat) > max_sps) {
  sp_sd <- apply(fg.mat, 1, sd, na.rm = TRUE)

  keep_sps <- names(sort(sp_sd, decreasing = TRUE))[
    seq_len(max_sps)
  ]

  fg.mat <- fg.mat[keep_sps, , drop = FALSE]
}

# Final matrix to plot
# Assumed already score/z-score-like, so no extra scaling
plot_mat <- fg.mat

# Cluster-level annotations
meta <- GSE274934@meta.data
meta$predicted.ann_finest_level <- as.character(meta$predicted.ann_finest_level)

cluster_annot <- meta %>%
  dplyr::group_by(predicted.ann_finest_level) %>%
  dplyr::summarise(
    mean_C1_Core = mean(C1_Core, na.rm = TRUE),
    n_cells = dplyr::n(),
    .groups = "drop"
  ) %>%
  as.data.frame()

cluster_annot$predicted.ann_finest_level <- as.character(cluster_annot$predicted.ann_finest_level)

# Match annotation order to heatmap columns
cluster_annot <- cluster_annot[
  match(colnames(plot_mat), cluster_annot$predicted.ann_finest_level),
  ,
  drop = FALSE
]

rownames(cluster_annot) <- cluster_annot$predicted.ann_finest_level

stopifnot(!any(is.na(cluster_annot$predicted.ann_finest_level)))
stopifnot(identical(rownames(cluster_annot), colnames(plot_mat)))

# Robust symmetric color scale
# Useful because SecAct values can be very large
lim <- quantile(abs(as.numeric(plot_mat)), probs = 0.98, na.rm = TRUE)

col_fun <- colorRamp2(
  c(-lim, 0, lim),
  c("#053061", "#F7F7F7", "#67001F")
)

# Color scale for mean C1_Core
c1_col_fun <- colorRamp2(
  c(
    min(cluster_annot$mean_C1_Core, na.rm = TRUE),
    median(cluster_annot$mean_C1_Core, na.rm = TRUE),
    max(cluster_annot$mean_C1_Core, na.rm = TRUE)
  ),
  c("#313695", "#F7F7F7", "#A50026")
)

# Top annotations for clusters
top_ha <- HeatmapAnnotation(
  `Mean C1 raw` = cluster_annot$mean_C1_Core,
  `Cells` = anno_barplot(
    cluster_annot$n_cells,
    gp = gpar(fill = "grey40", col = NA),
    border = FALSE,
    height = unit(1.0, "cm")
  ),
  col = list(
    `Mean C1 raw` = c1_col_fun
  ),
  annotation_name_side = "left",
  annotation_name_gp = gpar(fontsize = 8),
  annotation_legend_param = list(
    `Mean C1 raw` = list(title = "Mean C1 Core")
  )
)

# Heatmap title
heatmap_title <- paste0(
  "SecAct: top active secreted proteins (SPs)\n",
  nrow(plot_mat),
  " SPs displayed | top ",
  n_per_cluster,
  " per cluster",
  ifelse(nrow(plot_mat) == max_sps, paste0(" | capped at ", max_sps), ""),
  " | Pearson distance | average linkage"
)

ht_opt$TITLE_PADDING = unit(c(8, 8), "points")

png(
  filename = "output/scRNA/figures/SecAct_top_active_secreted_proteins_heatmap.png",
  width = 16,
  height = 12,
  units = "in",
  res = 600,
  type = "cairo-png"
)

ht <- ComplexHeatmap::Heatmap(
  plot_mat,
  name = "SecAct",
  col = col_fun,

  top_annotation = top_ha,

  cluster_rows = TRUE,
  cluster_columns = TRUE,

  clustering_distance_rows = "pearson",
  clustering_distance_columns = "pearson",
  clustering_method_rows = "average",
  clustering_method_columns = "average",

  show_row_dend = TRUE,
  show_column_dend = TRUE,

  row_title = "Secreted proteins",
  column_title = heatmap_title,

  column_title_gp = gpar(
    fontsize = 13,
    fontface = "bold",
    lineheight = 1.15
  ),
  row_title_gp = gpar(
    fontsize = 11,
    fontface = "bold"
  ),

  row_names_gp = gpar(fontsize = 5.5),
  column_names_gp = gpar(fontsize = 5, fontface = "bold"),
  column_names_rot = 90,

  border = TRUE,
  rect_gp = gpar(col = "grey90", lwd = 0.25),

  heatmap_legend_param = list(
    title = "SecAct score",
    at = c(-lim, 0, lim),
    labels = c(
      paste0("-", round(lim, 1)),
      "0",
      paste0("+", round(lim, 1))
    )
  )
)

draw(
  ht,
  heatmap_legend_side = "right",
  annotation_legend_side = "right"
)

dev.off()

pdf(
  file = "output/figures/SecAct_top_active_secreted_proteins_heatmap.pdf",
  width = 10,
  height = 12
)

ht <- ComplexHeatmap::Heatmap(
  plot_mat,
  name = "SecAct",
  col = col_fun,

  top_annotation = top_ha,

  cluster_rows = TRUE,
  cluster_columns = TRUE,

  clustering_distance_rows = "pearson",
  clustering_distance_columns = "pearson",
  clustering_method_rows = "average",
  clustering_method_columns = "average",

  show_row_dend = TRUE,
  show_column_dend = TRUE,

  row_title = "Secreted proteins",
  column_title = heatmap_title,

  column_title_gp = gpar(
    fontsize = 13,
    fontface = "bold",
    lineheight = 1.15
  ),
  row_title_gp = gpar(
    fontsize = 11,
    fontface = "bold"
  ),

  row_names_gp = gpar(fontsize = 5.5),
  column_names_gp = gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 0,

  border = TRUE,
  rect_gp = gpar(col = "grey90", lwd = 0.25),

  heatmap_legend_param = list(
    title = "SecAct score",
    at = c(-lim, 0, lim),
    labels = c(
      paste0("-", round(lim, 1)),
      "0",
      paste0("+", round(lim, 1))
    )
  )
)

draw(
  ht,
  heatmap_legend_side = "right",
  annotation_legend_side = "right"
)

dev.off()

ht_opt(RESET = TRUE)

```

## Section 12.2. SecAct Most Variable Proteins

This companion heatmap keeps the proteins with the strongest between-cluster variation, independent of which cluster contributed them.
```{r}

library(ComplexHeatmap)
library(circlize)
library(dplyr)
library(grid)

n_sps <- 80

dir.create("output/figures", recursive = TRUE, showWarnings = FALSE)

# Full SecAct activity matrix
# Rows = secreted proteins
# Columns = clusters
act_mat <- as.matrix(act)
colnames(act_mat) <- as.character(colnames(act_mat))

# Remove proteins with zero or missing variation
sp_sd <- apply(act_mat, 1, sd, na.rm = TRUE)
sp_sd <- sp_sd[is.finite(sp_sd) & sp_sd > 0]

# Select top variable secreted proteins across clusters
top_sps <- names(sort(sp_sd, decreasing = TRUE))[
  seq_len(min(n_sps, length(sp_sd)))
]

plot_mat <- act_mat[top_sps, , drop = FALSE]

# Cluster-level annotations
meta <- GSE274934@meta.data
meta$predicted.ann_finest_level <- as.character(meta$predicted.ann_finest_level)

cluster_annot <- meta %>%
  dplyr::group_by(predicted.ann_finest_level) %>%
  dplyr::summarise(
    mean_C1_Core = mean(C1_Core, na.rm = TRUE),
    n_cells = dplyr::n(),
    .groups = "drop"
  ) %>%
  as.data.frame()

cluster_annot$predicted.ann_finest_level <- as.character(cluster_annot$predicted.ann_finest_level)

# Match annotation order to heatmap columns
cluster_annot <- cluster_annot[
  match(colnames(plot_mat), cluster_annot$predicted.ann_finest_level),
  ,
  drop = FALSE
]

rownames(cluster_annot) <- cluster_annot$predicted.ann_finest_level

stopifnot(!any(is.na(cluster_annot$predicted.ann_finest_level)))
stopifnot(identical(rownames(cluster_annot), colnames(plot_mat)))

# Robust symmetric color scale
# The data are assumed already score/z-score-like, so no extra scaling is applied
lim <- quantile(abs(as.numeric(plot_mat)), probs = 0.98, na.rm = TRUE)

col_fun <- colorRamp2(
  c(-lim, 0, lim),
  c("#053061", "#F7F7F7", "#67001F")
)

# Color scale for mean C1_Core
c1_col_fun <- colorRamp2(
  c(
    min(cluster_annot$mean_C1_Core, na.rm = TRUE),
    median(cluster_annot$mean_C1_Core, na.rm = TRUE),
    max(cluster_annot$mean_C1_Core, na.rm = TRUE)
  ),
  c("#313695", "#F7F7F7", "#A50026")
)

# Top annotations for clusters
top_ha <- HeatmapAnnotation(
  `Mean C1 raw` = cluster_annot$mean_C1_Core,
  `Cells` = anno_barplot(
    cluster_annot$n_cells,
    gp = gpar(fill = "grey40", col = NA),
    border = FALSE,
    height = unit(1.0, "cm")
  ),
  col = list(
    `Mean C1 raw` = c1_col_fun
  ),
  annotation_name_side = "left",
  annotation_name_gp = gpar(fontsize = 8),
  annotation_legend_param = list(
    `Mean C1 raw` = list(title = "Mean C1 Core")
  )
)

# Heatmap title
heatmap_title <- paste0(
  "SecAct: most variable secreted proteins (SPs)\n",
  nrow(plot_mat),
  " SPs displayed | selected by SD across clusters | Pearson distance | average linkage"
)

ht_opt$TITLE_PADDING = unit(c(8, 8), "points")

png(
  filename = "output/scRNA/figures/SecAct_variable_secreted_proteins_heatmap.png",
  width = 12,
  height = 12,
  units = "in",
  res = 800
)

ht <- ComplexHeatmap::Heatmap(
  plot_mat,
  name = "SecAct",
  col = col_fun,

  top_annotation = top_ha,

  cluster_rows = TRUE,
  cluster_columns = TRUE,

  clustering_distance_rows = "pearson",
  clustering_distance_columns = "pearson",
  clustering_method_rows = "average",
  clustering_method_columns = "average",

  show_row_dend = TRUE,
  show_column_dend = TRUE,

  row_title = "Secreted proteins",
  column_title = heatmap_title,

  column_title_gp = gpar(
    fontsize = 13,
    fontface = "bold",
    lineheight = 1.15
  ),
  row_title_gp = gpar(
    fontsize = 11,
    fontface = "bold"
  ),

  row_names_gp = gpar(fontsize = 6),
  column_names_gp = gpar(fontsize = 6, fontface = "bold"),
  column_names_rot = 90,

  border = TRUE,
  rect_gp = gpar(col = "grey90", lwd = 0.25),

  heatmap_legend_param = list(
    title = "SecAct score",
    at = c(-lim, 0, lim),
    labels = c(
      paste0("-", round(lim, 1)),
      "0",
      paste0("+", round(lim, 1))
    )
  )
)

draw(
  ht,
  heatmap_legend_side = "right",
  annotation_legend_side = "right"
)

dev.off()

pdf(
  file = "output/figures/SecAct_variable_secreted_proteins_heatmap.pdf",
  width = 12,
  height = 12
)

ht <- ComplexHeatmap::Heatmap(
  plot_mat,
  name = "SecAct",
  col = col_fun,

  top_annotation = top_ha,

  cluster_rows = TRUE,
  cluster_columns = TRUE,

  clustering_distance_rows = "pearson",
  clustering_distance_columns = "pearson",
  clustering_method_rows = "average",
  clustering_method_columns = "average",

  show_row_dend = TRUE,
  show_column_dend = TRUE,

  row_title = "Secreted proteins",
  column_title = heatmap_title,

  column_title_gp = gpar(
    fontsize = 13,
    fontface = "bold",
    lineheight = 1.15
  ),
  row_title_gp = gpar(
    fontsize = 11,
    fontface = "bold"
  ),

  row_names_gp = gpar(fontsize = 6),
  column_names_gp = gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 0,

  border = TRUE,
  rect_gp = gpar(col = "grey90", lwd = 0.25),

  heatmap_legend_param = list(
    title = "SecAct score",
    at = c(-lim, 0, lim),
    labels = c(
      paste0("-", round(lim, 1)),
      "0",
      paste0("+", round(lim, 1))
    )
  )
)

draw(
  ht,
  heatmap_legend_side = "right",
  annotation_legend_side = "right"
)

dev.off()

ht_opt(RESET = TRUE)

```

# Section 13. Hallmark Pathway Activity with AUCell

This section scores MSigDB Hallmark programs per cell and summarizes how pathway activity differs across epithelial clusters.

```{r}

library(SeuratExtend)
plan(multicore, workers = 5)
DefaultAssay(GSE274934) <- "RNA"
GSE274934 <- GeneSetAnalysis(GSE274934, genesets = hall50$human)

matr <- as.matrix(GSE274934@misc$AUCell$genesets)
auc <- t(matr)
meta <- GSE274934@meta.data

# ensure same cells order
auc <- auc[rownames(meta), ]

H50_stats <- CalcStats(matr, f = GSE274934@meta.data$seurat_clusters)

desc <- dplyr::desc

WaterfallPlot(matr, f = GSE274934$seurat_clusters, ident.1 = "7", ident.2 = "2", top.n = 20, len.threshold = 2,color_theme = "D")  +
  theme_pubclean(base_size = 12)

ggsave(
  filename = "output/figures/HALLMARK_50_7vs2.png",
  width = 10,
  height = 8
)

```

## Section 13.1. Hallmark Cluster Heatmap

This heatmap gives the compact cluster-level pathway view used for interpretation. The C1 annotation above the heatmap helps separate plasticity-linked Hallmarks from cluster-only effects.

```{r}

library(ComplexHeatmap)
library(circlize)
library(grid)
library(dplyr)

dir.create("output/figures", recursive = TRUE, showWarnings = FALSE)

# H50_stats is already assumed to be a z-score matrix
# Rows = Hallmark pathways
# Columns = clusters
plot_mat <- as.matrix(H50_stats)

# Make sure cluster names are character
colnames(plot_mat) <- as.character(colnames(plot_mat))

# Optional: clean pathway names for plotting
rownames(plot_mat) <- gsub("^HALLMARK_", "", rownames(plot_mat))
rownames(plot_mat) <- gsub("_", " ", rownames(plot_mat))

# Cluster-level annotations from metadata
meta <- GSE274934@meta.data
meta$seurat_clusters <- as.character(meta$seurat_clusters)

cluster_annot <- meta %>%
  dplyr::group_by(seurat_clusters) %>%
  dplyr::summarise(
    mean_C1_Core = mean(C1_Core, na.rm = TRUE),
    n_cells = dplyr::n(),
    .groups = "drop"
  ) %>%
  as.data.frame()

cluster_annot$seurat_clusters <- as.character(cluster_annot$seurat_clusters)

# Match annotation order to heatmap columns
cluster_annot <- cluster_annot[
  match(colnames(plot_mat), cluster_annot$seurat_clusters),
  ,
  drop = FALSE
]

rownames(cluster_annot) <- cluster_annot$seurat_clusters

stopifnot(!any(is.na(cluster_annot$seurat_clusters)))
stopifnot(identical(rownames(cluster_annot), colnames(plot_mat)))

# Color scale for pathway z-scores
col_fun <- colorRamp2(
  c(-2, -1, 0, 1, 2),
  c("#053061", "#4393C3", "#F7F7F7", "#D6604D", "#67001F")
)

# Color scale for mean C1_Core
c1_col_fun <- colorRamp2(
  c(
    min(cluster_annot$mean_C1_Core, na.rm = TRUE),
    median(cluster_annot$mean_C1_Core, na.rm = TRUE),
    max(cluster_annot$mean_C1_Core, na.rm = TRUE)
  ),
  c("#313695", "#F7F7F7", "#A50026")
)

# Top annotations for clusters
top_ha <- HeatmapAnnotation(
  `Mean C1 raw` = cluster_annot$mean_C1_Core,
  `Cells` = anno_barplot(
    cluster_annot$n_cells,
    gp = gpar(fill = "grey40", col = NA),
    border = FALSE,
    height = unit(1.0, "cm")
  ),
  col = list(
    `Mean C1 raw` = c1_col_fun
  ),
  annotation_name_side = "left",
  annotation_name_gp = gpar(fontsize = 8),
  annotation_legend_param = list(
    `Mean C1 raw` = list(title = "Mean C1 Core")
  )
)

# Title
heatmap_title <- paste0(
  "Hallmark pathway activity across clusters\n",
  nrow(plot_mat),
  " pathways displayed | z-score | Pearson distance | average linkage"
)

ht_opt$TITLE_PADDING = unit(c(8, 8), "points")

png(
  filename = "output/scRNA/figures/H50_hallmark_heatmap_C1_cells.png",
  width = 12,
  height = 8,
  units = "in",
  res = 600
)

ht <- ComplexHeatmap::Heatmap(
  plot_mat,
  name = "Pathway activity",
  col = col_fun,

  top_annotation = top_ha,

  cluster_rows = TRUE,
  cluster_columns = TRUE,

  clustering_distance_rows = "pearson",
  clustering_distance_columns = "pearson",
  clustering_method_rows = "average",
  clustering_method_columns = "average",

  show_row_dend = TRUE,
  show_column_dend = TRUE,

  row_title = "Hallmark pathways",
  column_title = heatmap_title,

  column_title_gp = gpar(
    fontsize = 13,
    fontface = "bold",
    lineheight = 1.15
  ),
  row_title_gp = gpar(
    fontsize = 11,
    fontface = "bold"
  ),

  row_names_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 0,

  border = TRUE,
  rect_gp = gpar(col = "grey90", lwd = 0.3),

  heatmap_legend_param = list(
    title = "Z-score",
    at = c(-2, -1, 0, 1, 2),
    labels = c("Low", "-1", "0", "+1", "High")
  )
)

draw(
  ht,
  heatmap_legend_side = "right",
  annotation_legend_side = "right"
)

dev.off()

ht_opt(RESET = TRUE)

```

# Section 14. Epithelial only -  Hallmark-C1 Correlation Summary

This final Hallmark block correlates each pathway score with the continuous C1 score and exports the ranked table used by downstream MAP interpretation.

```{r}

library(dplyr)

HALL_50_corr <- t(GSE274934@misc$AUCell$genesets)|>
  as.data.frame() |>
  mutate(C1_score = GSE274934@meta.data$C1_Core)

c1 <- meta$C1_Core

hall_results <- apply(auc, 2, function(x) {
  test <- cor.test(x, c1, method = "spearman",exact = FALSE)

  c(
    rho = unname(test$estimate),
    p = test$p.value
  )
})

hall_results <- as.data.frame(t(hall_results))

hall_results$Hallmark <- rownames(hall_results)
hall_results$FDR <- p.adjust(hall_results$p, method = "BH")

hallmark_trait_results <- hall_results %>%
  arrange(desc(abs(rho))) %>%
  mutate(Direction = ifelse(rho > 0, "Positive", "Negative")) %>%
  as_tibble()

hallmark_trait_results

hallmark_trait_results |>
  write_csv(file = "output/tables/HALLMARK_50_COR_C1.csv")

plot_hallmark_C1 <- hallmark_trait_results %>%
  dplyr::mutate(
    Hallmark_clean = gsub("^HALLMARK_", "", Hallmark),
    Hallmark_clean = gsub("_", " ", Hallmark_clean),
    Hallmark_clean = factor(Hallmark_clean, levels = Hallmark_clean[order(rho)])
  ) %>%
  ggplot2::ggplot(
    ggplot2::aes(
      x = rho,
      y = Hallmark_clean,
      fill = rho
    )
  ) +
  ggplot2::geom_col(width = 0.7) +
  ggplot2::geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  scale_fill_viridis_c(option = "D") +  # 👈 FIXED
  ggplot2::labs(
    title = "Hallmark Pathway Association With C1 Score",
    subtitle = "Spearman correlation between Hallmark scores and C1 score",
    x = "Spearman rho with C1 score",
    y = "Hallmark pathway",
    fill = "Direction"
  ) +
  theme_classic2(base_size = 13) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold"),
    axis.text = ggplot2::element_text(color = "black"),
    legend.position = "none"
  )

plot_hallmark_C1

ggsave(
  "output/figures/plot_hallmark_C1.png",
  plot = plot_hallmark_C1,
  width = 12,
  height = 10,
  dpi = 600
)

```


# Section 15. Metabolism Hand-Off and Lightweight Checks

Metabolism is intentionally handled in `08_MAP_metabolic_activity_scCellFie.qmd`. The code below only imports the already computed metabolic task scores into the Seurat object and performs lightweight cross-layer checks that help connect regulatory programs with the MAP 08 output.

```{r}

metabolic_tasks <- fread(
  "output/tables/metabolic_tasks_scores.csv",
  check.names = FALSE
) |>
  as.data.frame() |>
  column_to_rownames("V1")

head(metabolic_tasks)

all(rownames(metabolic_tasks) == colnames(GSE274934))

# Check alignment
stopifnot(all(rownames(metabolic_tasks) == colnames(GSE274934)))

# Transpose: Seurat assays require features × cells
metabolic_mat <- t(as.matrix(metabolic_tasks))

# Create assay
GSE274934[["metabolic"]] <- CreateAssayObject(
  data = metabolic_mat
)

# Set as default assay when working with metabolic tasks
DefaultAssay(GSE274934) <- "metabolic"

```

## Section 15.1. Metabolic heatmap for cell type

```{r}

cat("to do...")

```

## Section 15.1. Epithelial only - Metabolic Task Correlation with C1 Score

This section ranks metabolic task scores by their Spearman association with the continuous C1 score.
```{r}

library(tidyverse)
library(data.table)
library(future)
library(progressr)
library(future.apply)
library(ggpubr)
library(scattermore)


DefaultAssay(GSE274934) <- "metabolic"

# Use data slot because these are already task activity/scores
metabolic_activity <- as.matrix(GetAssayData(
  object = GSE274934,
  assay = "metabolic",
  layer = "data"
))

metabolic_task_names <- rownames(metabolic_activity)

C1_score <- GSE274934@meta.data$C1_Core

stopifnot(length(C1_score) == ncol(metabolic_activity))
stopifnot(all(colnames(metabolic_activity) == colnames(GSE274934)))

plan(multicore, workers = 5)

handlers(global = TRUE)
handlers("cli")

with_progress({
  p <- progressor(along = metabolic_task_names)

  metabolic_pseudotime_cor <- future_lapply(
    metabolic_task_names,
    function(task) {
      p(sprintf("Metabolic task: %s", task))

      x <- metabolic_activity[task, ]

      keep <- is.finite(x) & is.finite(C1_score)

      r <- cor.test(
        x[keep],
        C1_score[keep],
        method = "spearman",
        exact = FALSE
      )

      data.frame(
        metabolic_task = task,
        rho = unname(r$estimate[["rho"]]),
        p.value = r$p.value,
        stringsAsFactors = FALSE
      )
    },
    future.seed = TRUE
  ) |>
    bind_rows()
})

metabolic_pseudotime_cor <- metabolic_pseudotime_cor |>
  mutate(fdr = p.adjust(p.value, method = "fdr"))

sig_metabolic_tasks <- metabolic_pseudotime_cor |>
  filter(fdr <= 0.05) |>
  arrange(desc(rho)) |>
  mutate(
    direction = if_else(
      rho > 0,
      "increasing with C1",
      "decreasing with C1"
    )
  )

sig_metabolic_tasks

fwrite(
  x = sig_metabolic_tasks,
  file = "output/tables/cor_spearman_metabolic_tasks_along_C1.csv"
)

print(sig_metabolic_tasks |> slice_head(n = 15))

n_each <- 8

top_pos <- sig_metabolic_tasks |>
  arrange(desc(rho)) |>
  slice_head(n = n_each)

top_neg <- sig_metabolic_tasks |>
  arrange(rho) |>
  slice_head(n = n_each)

top_metabolic_tasks <- bind_rows(top_pos, top_neg) |>
  pull(metabolic_task)

plot_data <- tibble(
  cell = colnames(metabolic_activity),
  C1_score = C1_score
) |>
  bind_cols(
    as_tibble(
      t(metabolic_activity[top_metabolic_tasks, , drop = FALSE]),
      .name_repair = "minimal"
    )
  ) |>
  pivot_longer(
    cols = all_of(top_metabolic_tasks),
    names_to = "metabolic_task",
    values_to = "activity"
  ) |>
  mutate(
    metabolic_task = factor(
      metabolic_task,
      levels = top_metabolic_tasks
    )
  )

metabolic_plot_16 <- ggplot(plot_data, aes(x = C1_score, y = activity)) +
  scattermore::geom_scattermore(
    color = "grey30",
    alpha = 0.8
  ) +
  geom_smooth(
    method = "gam",
    formula = y ~ s(x, bs = "cs"),
    color = "#d73027",
    fill = "#d73027",
    alpha = 0.12,
    linewidth = 0.9,
    se = TRUE
  ) +
  facet_wrap(
    ~ metabolic_task,
    scales = "free_y",
    ncol = 4
  ) +
  ggpubr::stat_cor(method = "spearman") +
  labs(
    title = "Top metabolic task trends along C1",
    subtitle = "Spearman correlation, FDR ≤ 0.05",
    x = "C1 score",
    y = "Metabolic task activity"
  ) +
  theme_bw(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold", size = 8),
    panel.grid.minor = element_blank()
  )

metabolic_plot_16

ggsave(
  plot = metabolic_plot_16,
  filename = "output/figures/cor_spearman_metabolic_tasks_along_C1_top_16.png",
  width = 16,
  height = 10,
  dpi = 600
)

```

## Section 15.2. Epithelial only - C1-Ordered Metabolic Task Heatmap

This heatmap displays the top positive and negative metabolic correlations across cells ordered by C1 score.

```{r}

library(Seurat)
library(tidyverse)
library(ComplexHeatmap)
library(circlize)
library(grid)
library(data.table)

DefaultAssay(GSE274934) <- "metabolic"

# Extract metabolic task matrix: tasks × cells
metabolic_activity <- as.matrix(GetAssayData(
  object = GSE274934,
  assay = "metabolic",
  layer = "data"
))

# C1 score
C1_score <- GSE274934@meta.data$C1_Core
names(C1_score) <- colnames(GSE274934)

stopifnot(all(colnames(metabolic_activity) == colnames(GSE274934)))

# Select top metabolic tasks from previous correlation table
sig_metabolic_tasks <- fread(
  "output/tables/cor_spearman_metabolic_tasks_along_C1.csv"
) |>
  as_tibble()

n_each <- 25

top_pos <- sig_metabolic_tasks |>
  filter(fdr <= 0.05, rho > 0) |>
  arrange(desc(rho)) |>
  slice_head(n = n_each)

top_neg <- sig_metabolic_tasks |>
  filter(fdr <= 0.05, rho < 0) |>
  arrange(rho) |>
  slice_head(n = n_each)

top_metabolic_tasks <- bind_rows(top_pos, top_neg) |>
  pull(metabolic_task)

# Keep only tasks present in the assay
top_metabolic_tasks <- intersect(top_metabolic_tasks, rownames(metabolic_activity))

# Subset matrix
plot_mat <- metabolic_activity[top_metabolic_tasks, , drop = FALSE]

# Order cells by C1
cell_order <- names(sort(C1_score, decreasing = FALSE))

plot_mat <- plot_mat[, cell_order, drop = FALSE]
C1_ordered <- C1_score[cell_order]

# Row-scale metabolic task activity
plot_mat_scaled <- t(scale(t(plot_mat)))
plot_mat_scaled[is.na(plot_mat_scaled)] <- 0

# Color functions
met_col_fun <- circlize::colorRamp2(
  c(-2, 0, 2),
  c("#2166AC", "white", "#B2182B")
)

C1_col_fun <- circlize::colorRamp2(
  c(
    min(C1_ordered, na.rm = TRUE),
    median(C1_ordered, na.rm = TRUE),
    max(C1_ordered, na.rm = TRUE)
  ),
  c("#313695", "white", "#A50026")
)

# Top annotation
top_ha <- HeatmapAnnotation(
  C1_score = C1_ordered,
  col = list(C1_score = C1_col_fun),
  annotation_name_gp = gpar(fontsize = 9, fontface = "bold"),
  simple_anno_size = unit(4, "mm")
)

# Optional row split by correlation direction
task_direction <- sig_metabolic_tasks |>
  filter(metabolic_task %in% top_metabolic_tasks) |>
  mutate(direction = if_else(rho > 0, "Positive rho", "Negative rho")) |>
  select(metabolic_task, direction)

row_split <- task_direction$direction[
  match(rownames(plot_mat_scaled), task_direction$metabolic_task)
]

# Heatmap
ht_c1 <- Heatmap(
  plot_mat_scaled,
  name = "Z-score",
  col = met_col_fun,
  
  top_annotation = top_ha,
  row_split = row_split,
  
  cluster_rows = TRUE,
  cluster_columns = FALSE,

  clustering_distance_rows = "pearson",
  clustering_distance_columns = "pearson",
  clustering_method_rows = "average",
  clustering_method_columns = "average",

  show_row_dend = TRUE,
  show_column_dend = TRUE,
  
  show_column_names = FALSE,
  show_row_names = TRUE,
  
  row_names_gp = gpar(fontsize = 7),
  column_title = "Cells ordered by C1 score",
  row_title = "Metabolic tasks",
  
  column_title_gp = gpar(fontsize = 13, fontface = "bold"),
  row_title_gp = gpar(fontsize = 11, fontface = "bold"),
  
  use_raster = TRUE,
  raster_quality = 5
)

pdf(
  file = "output/figures/metabolic_tasks_heatmap_cells_ordered_by_C1.pdf",
  width = 11,
  height = 12
)

draw(
  ht_c1,
  heatmap_legend_side = "right",
  annotation_legend_side = "right"
)

dev.off()

```

## Section 15.3. Epithelial only - Cluster-Level Metabolic Task Heatmap

This view aggregates metabolic tasks by cluster so the metabolic signal can be compared with cluster-level C1 annotations.

```{r}

library(ComplexHeatmap)
library(circlize)
library(dplyr)
library(grid)
library(Seurat)

n_tasks <- 80

dir.create("output/figures", recursive = TRUE, showWarnings = FALSE)

DefaultAssay(GSE274934) <- "metabolic"

# Extract metabolic task activity matrix
# Seurat assay is tasks x cells, so transpose to cells x tasks
acts_mat <- t(as.matrix(GetAssayData(
  object = GSE274934,
  assay = "metabolic",
  layer = "data"
)))

# Align metadata to metabolic activity matrix
meta <- GSE274934@meta.data
meta <- meta[rownames(acts_mat), , drop = FALSE]

stopifnot(all(rownames(acts_mat) == rownames(meta)))

# Cluster labels
cluster <- as.character(meta$seurat_clusters)

# Count cells per cluster
cluster_counts <- table(cluster)

# Sum metabolic task activity by cluster
cluster_sum_mat <- rowsum(
  acts_mat,
  group = cluster,
  reorder = FALSE
)

# Convert cluster sums into cluster means
# Result: clusters x metabolic tasks
cluster_mean_mat <- sweep(
  cluster_sum_mat,
  MARGIN = 1,
  STATS = as.numeric(cluster_counts[rownames(cluster_sum_mat)]),
  FUN = "/"
)

# Select metabolic tasks with strongest variation across clusters
task_sd <- apply(cluster_mean_mat, 2, sd, na.rm = TRUE)

top_tasks <- names(sort(task_sd, decreasing = TRUE))[
  seq_len(min(n_tasks, length(task_sd)))
]

cluster_mean_mat <- cluster_mean_mat[, top_tasks, drop = FALSE]

# Scale each metabolic task across clusters
# cluster_mean_mat is clusters x tasks, so scale() scales each task column
cluster_mean_scaled <- scale(cluster_mean_mat)
cluster_mean_scaled[is.na(cluster_mean_scaled)] <- 0

# Heatmap matrix
# Rows = metabolic tasks, columns = clusters
plot_mat <- t(cluster_mean_scaled)

# Unscaled matrix, useful for optional row annotation
plot_mat_unscaled <- t(cluster_mean_mat)

# Cluster-level annotations
cluster_annot <- meta %>%
  dplyr::group_by(seurat_clusters) %>%
  dplyr::summarise(
    mean_C1_Core = mean(C1_Core, na.rm = TRUE),
    n_cells = dplyr::n(),
    .groups = "drop"
  ) %>%
  as.data.frame()

cluster_annot$seurat_clusters <- as.character(cluster_annot$seurat_clusters)

# Match annotation order to heatmap columns
cluster_annot <- cluster_annot[
  match(colnames(plot_mat), cluster_annot$seurat_clusters),
  ,
  drop = FALSE
]

rownames(cluster_annot) <- cluster_annot$seurat_clusters

stopifnot(!any(is.na(cluster_annot$seurat_clusters)))
stopifnot(identical(rownames(cluster_annot), colnames(plot_mat)))

# Color scale for scaled metabolic activity
met_col_fun <- colorRamp2(
  c(-2, -1, 0, 1, 2),
  c("#053061", "#4393C3", "#F7F7F7", "#D6604D", "#67001F")
)

# Color scale for mean C1_Core
c1_col_fun <- colorRamp2(
  c(
    min(cluster_annot$mean_C1_Core, na.rm = TRUE),
    median(cluster_annot$mean_C1_Core, na.rm = TRUE),
    max(cluster_annot$mean_C1_Core, na.rm = TRUE)
  ),
  c("#313695", "#F7F7F7", "#A50026")
)

# Top annotations for clusters
top_ha <- HeatmapAnnotation(
  `Mean C1 raw` = cluster_annot$mean_C1_Core,
  `Cells` = anno_barplot(
    cluster_annot$n_cells,
    gp = gpar(fill = "grey40", col = NA),
    border = FALSE,
    height = unit(1, "cm")
  ),
  col = list(
    `Mean C1 raw` = c1_col_fun
  ),
  annotation_name_side = "left",
  annotation_name_gp = gpar(fontsize = 8),
  annotation_legend_param = list(
    `Mean C1 raw` = list(title = "Mean C1 raw")
  )
)

# Optional left annotation
# Shows distribution of unscaled cluster-level mean metabolic activity
left_ha <- rowAnnotation(
  `Activity distribution` = anno_boxplot(
    plot_mat_unscaled,
    which = "row",
    border = FALSE,
    outline = FALSE,
    axis = TRUE,
    width = unit(2.3, "cm"),
    gp = gpar(fill = "grey70", col = "grey40")
  ),
  annotation_name_rot = 90,
  annotation_name_gp = gpar(fontsize = 9)
)

heatmap_title <- paste0(
  "Cluster-level metabolic task activity heatmap\n",
  nrow(plot_mat),
  " tasks displayed | row z-score | Pearson distance | average linkage"
)

ht_opt$TITLE_PADDING = unit(c(10, 10), "points")

png(
  filename = "output/figures/metabolic_tasks_heatmap_cluster_level.png",
  width = 9,
  height = 11,
  units = "in",
  res = 600
)

ht <- Heatmap(
  plot_mat,
  name = "Metabolic\nactivity",
  col = met_col_fun,

  # left_annotation = left_ha,
  top_annotation = top_ha,

  cluster_rows = TRUE,
  cluster_columns = TRUE,

  clustering_distance_rows = "pearson",
  clustering_distance_columns = "pearson",
  clustering_method_rows = "average",
  clustering_method_columns = "average",

  show_row_dend = TRUE,
  show_column_dend = TRUE,

  row_title = "Metabolic tasks",
  column_title = heatmap_title,

  column_title_gp = gpar(
    fontsize = 13,
    fontface = "bold",
    lineheight = 1.15
  ),
  row_title_gp = gpar(
    fontsize = 11,
    fontface = "bold"
  ),

  row_names_gp = gpar(fontsize = 6),
  column_names_gp = gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 0,

  border = TRUE,
  rect_gp = gpar(col = "grey90", lwd = 0.3),

  heatmap_legend_param = list(
    title = "Row z-score",
    at = c(-2, -1, 0, 1, 2),
    labels = c("Low", "-1", "0", "+1", "High")
  )
)

draw(
  ht,
  heatmap_legend_side = "right",
  annotation_legend_side = "right"
)

dev.off()


pdf(
  file = "output/figures/metabolic_tasks_heatmap_cluster_level.pdf",
  width = 10,
  height = 12,
  useDingbats = FALSE
)

ht <- Heatmap(
  plot_mat,
  name = "Metabolic\nactivity",
  col = met_col_fun,

  # left_annotation = left_ha,
  top_annotation = top_ha,

  cluster_rows = TRUE,
  cluster_columns = TRUE,

  clustering_distance_rows = "pearson",
  clustering_distance_columns = "pearson",
  clustering_method_rows = "average",
  clustering_method_columns = "average",

  show_row_dend = TRUE,
  show_column_dend = TRUE,

  row_title = "Metabolic tasks",
  column_title = heatmap_title,

  column_title_gp = gpar(
    fontsize = 13,
    fontface = "bold",
    lineheight = 1.15
  ),
  row_title_gp = gpar(
    fontsize = 11,
    fontface = "bold"
  ),

  row_names_gp = gpar(fontsize = 6),
  column_names_gp = gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 0,

  border = TRUE,
  rect_gp = gpar(col = "grey90", lwd = 0.3),

  heatmap_legend_param = list(
    title = "Row z-score",
    at = c(-2, -1, 0, 1, 2),
    labels = c("Low", "-1", "0", "+1", "High")
  )
)

draw(
  ht,
  heatmap_legend_side = "right",
  annotation_legend_side = "right"
)

dev.off()

ht_opt(RESET = TRUE)

```

## Section 15.5. Curated Metabolic UMAP Panels

These panels group interpretable task families so metabolism can be reviewed without scanning every scCellFie feature.

```{r}

metabolic_features <- rownames(GSE274934)

sugar_tasks <- c(
  "ATP generation from glucose (hypoxic conditions) - glycolysis",
  "ATP regeneration from glucose (normoxic conditions) - glycolysis + krebs cycle",
  "Glucose to lactate conversion",
  "Gluconeogenesis from Lactate",
  "Gluconeogenesis from pyruvate",
  "Fructose degradation (to glucose-3-phosphate)",
  "Fructose to glucose conversion (via fructose-6-phosphate)",
  "Glycogen biosynthesis",
  "Glycogen degradation",
  "Synthesis of ribose-5-phosphate"
)

fat_tasks <- c(
  "Palmitate synthesis",
  "Palmitate degradation",
  "Palmitolate synthesis",
  "Palmitolate degradation",
  "Synthesis of palmitoyl-CoA",
  "Cholesterol synthesis",
  "Mevalonate synthesis",
  "Triacylglycerol synthesis",
  "Ceramide synthesis",
  "Sphingomyelin synthesis",
  "Cardiolipin synthesis",
  "Arachidonate synthesis",
  "Arachidonate degradation",
  "Linoleate degradation",
  "Linolenate degradation",
  "cis-vaccenic acid synthesis",
  "cis-vaccenic acid degradation"
)

protein_tasks <- c(
  "Alanine synthesis",
  "Alanine degradation",
  "Glutamine synthesis",
  "Glutamine degradation",
  "Glutaminolysis (glutamine to lactate)",
  "Glutamate synthesis",
  "Glutamate degradation",
  "Serine synthesis",
  "Serine degradation",
  "Glycine synthesis",
  "Glycine degradation",
  "Proline synthesis",
  "Proline degradation",
  "Arginine synthesis",
  "Arginine degradation",
  "Aspartate synthesis",
  "Aspartate degradation",
  "Valine degradation",
  "Leucine degradation",
  "Isoleucine degradation",
  "Lysine degradation",
  "Methionine degradation",
  "Cysteine degradation",
  "Tryptophan degradation",
  "Tyrosine degradation",
  "Phenylalanine degradation"
)

mito_tasks <- c(
  "Krebs cycle - NADH generation",
  "Krebs cycle - oxidative decarboxylation of pyruvate",
  "Oxidative phosphorylation via succinate-coenzyme Q oxidoreductase (COMPLEX II)",
  "Malate to pyruvate conversion",
  "Conversion of GABA into succinate",
  "Conversion of leucine to acetyl-coA",
  "Valine to succinyl-coA",
  "Tyrosine to acetoacetate and fumarate",
  "Synthesis of ubiquinone from tyrosine",
  "Cardiolipin synthesis",
  "FAD synthesis",
  "NAD synthesis from nicotinamide"
)

ros_tasks <- c(
  "Reactive oxygen species detoxification (H2O2 to H2O)",
  "Glutathionate synthesis",
  "Presence of the thioredoxin system through the thioredoxin reductase activity",
  "Disulfide bond formation",
  "FAD synthesis",
  "NAD synthesis from nicotinamide",
  "Cysteine synthesis (need serine and methionine)",
  "Homocysteine degradation",
  "Homocysteine synthesis (need methionine)"
)

plot_metabolic_umaps <- function(tasks, filename_prefix, ncol = 4) {
  
  tasks <- intersect(tasks, metabolic_features)
  
  message("Plotting ", length(tasks), " tasks:")
  print(tasks)
  
  plots <- lapply(tasks, function(task) {
    FeaturePlot_scCustom(
      seurat_object = GSE274934,
      features = task
    ) +
      umap_theme +
      scale_color_viridis_c(option = "magma") + # 
      ggtitle(str_wrap(task, width = 35)) +
      labs(x = "UMAP1", y = "UMAP2") +
      theme(
        plot.title = element_text(
          size = 6.5,
          face = "bold",
          lineheight = 0.9
        )
      )
  })
  
  panel <- wrap_plots(plots, ncol = ncol)
  
  ggsave(
    filename = paste0("output/figures/", filename_prefix, ".pdf"),
    plot = panel,
    width = 18,
    height = ceiling(length(tasks) / ncol) * 3.2
  )
  
  
  panel
}

sugar_panel <- plot_metabolic_umaps(
  sugar_tasks,
  filename_prefix = "metabolic_umap_sugar_glycolysis_lactate",
  ncol = 4
)

fat_panel <- plot_metabolic_umaps(
  fat_tasks,
  filename_prefix = "metabolic_umap_fat_lipid_adipose",
  ncol = 4
)

protein_panel <- plot_metabolic_umaps(
  protein_tasks,
  filename_prefix = "metabolic_umap_protein_amino_acids",
  ncol = 4
)

mito_panel <- plot_metabolic_umaps(
  mito_tasks,
  filename_prefix = "metabolic_umap_mitochondrial_tca_oxphos",
  ncol = 4
)

ros_panel <- plot_metabolic_umaps(
  ros_tasks,
  filename_prefix = "metabolic_umap_ros_redox",
  ncol = 4
)

```

```{r}

metabolic_features <- rownames(GSE274934)

sugar_tasks <- c(
  "ATP generation from glucose (hypoxic conditions) - glycolysis",
  "ATP regeneration from glucose (normoxic conditions) - glycolysis + krebs cycle",
  "Glucose to lactate conversion",
  "Gluconeogenesis from Lactate",
  "Gluconeogenesis from pyruvate",
  "Fructose degradation (to glucose-3-phosphate)",
  "Fructose to glucose conversion (via fructose-6-phosphate)",
  "Glycogen biosynthesis",
  "Glycogen degradation",
  "Synthesis of ribose-5-phosphate"
)

fat_tasks <- c(
  "Palmitate synthesis",
  "Palmitate degradation",
  "Palmitolate synthesis",
  "Palmitolate degradation",
  "Synthesis of palmitoyl-CoA",
  "Cholesterol synthesis",
  "Mevalonate synthesis",
  "Triacylglycerol synthesis",
  "Ceramide synthesis",
  "Sphingomyelin synthesis",
  "Cardiolipin synthesis",
  "Arachidonate synthesis",
  "Arachidonate degradation",
  "Linoleate degradation",
  "Linolenate degradation",
  "cis-vaccenic acid synthesis",
  "cis-vaccenic acid degradation"
)

protein_tasks <- c(
  "Alanine synthesis",
  "Alanine degradation",
  "Glutamine synthesis",
  "Glutamine degradation",
  "Glutaminolysis (glutamine to lactate)",
  "Glutamate synthesis",
  "Glutamate degradation",
  "Serine synthesis",
  "Serine degradation",
  "Glycine synthesis",
  "Glycine degradation",
  "Proline synthesis",
  "Proline degradation",
  "Arginine synthesis",
  "Arginine degradation",
  "Aspartate synthesis",
  "Aspartate degradation",
  "Valine degradation",
  "Leucine degradation",
  "Isoleucine degradation",
  "Lysine degradation",
  "Methionine degradation",
  "Cysteine degradation",
  "Tryptophan degradation",
  "Tyrosine degradation",
  "Phenylalanine degradation"
)

mito_tasks <- c(
  "Krebs cycle - NADH generation",
  "Krebs cycle - oxidative decarboxylation of pyruvate",
  "Oxidative phosphorylation via succinate-coenzyme Q oxidoreductase (COMPLEX II)",
  "Malate to pyruvate conversion",
  "Conversion of GABA into succinate",
  "Conversion of leucine to acetyl-coA",
  "Valine to succinyl-coA",
  "Tyrosine to acetoacetate and fumarate",
  "Synthesis of ubiquinone from tyrosine",
  "Cardiolipin synthesis",
  "FAD synthesis",
  "NAD synthesis from nicotinamide"
)

ros_tasks <- c(
  "Reactive oxygen species detoxification (H2O2 to H2O)",
  "Glutathionate synthesis",
  "Presence of the thioredoxin system through the thioredoxin reductase activity",
  "Disulfide bond formation",
  "FAD synthesis",
  "NAD synthesis from nicotinamide",
  "Cysteine synthesis (need serine and methionine)",
  "Homocysteine degradation",
  "Homocysteine synthesis (need methionine)"
)

library(ComplexHeatmap)
library(circlize)
library(dplyr)
library(grid)
library(Seurat)
library(RColorBrewer)

# Create output directory
dir.create("output/figures", recursive = TRUE, showWarnings = FALSE)

# Use metabolic assay
DefaultAssay(GSE274934) <- "metabolic"

# Define selected metabolic tasks by category
selected_metabolic_tasks <- list(
  Sugar = sugar_tasks,
  Fat = fat_tasks,
  Protein = protein_tasks,
  Mitochondrial = mito_tasks,
  ROS = ros_tasks
)

# Create task annotation table
metabolic_task_annotation <- bind_rows(
  lapply(names(selected_metabolic_tasks), function(task_category) {
    data.frame(
      metabolic_task = selected_metabolic_tasks[[task_category]],
      task_category = task_category,
      stringsAsFactors = FALSE
    )
  })
) |>
  distinct(metabolic_task, .keep_all = TRUE)

# Preserve task category order
metabolic_task_annotation$task_category <- factor(
  metabolic_task_annotation$task_category,
  levels = c("Sugar", "Fat", "Protein", "Mitochondrial", "ROS")
)

# Extract metabolic task activity matrix
# Seurat assay is tasks x cells, so transpose to cells x tasks
metabolic_activity_matrix <- t(as.matrix(GetAssayData(
  object = GSE274934,
  assay = "metabolic",
  layer = "data"
)))

# Keep only selected tasks present in the metabolic assay
available_metabolic_tasks <- intersect(
  metabolic_task_annotation$metabolic_task,
  colnames(metabolic_activity_matrix)
)

# Report selected tasks missing from the metabolic assay
missing_metabolic_tasks <- setdiff(
  metabolic_task_annotation$metabolic_task,
  colnames(metabolic_activity_matrix)
)

missing_metabolic_tasks

# Stop if no selected tasks are present
stopifnot(length(available_metabolic_tasks) > 0)

# Filter annotation to available selected tasks
metabolic_task_annotation <- metabolic_task_annotation |>
  filter(metabolic_task %in% available_metabolic_tasks)

# Keep activity matrix with available selected tasks only
metabolic_activity_matrix <- metabolic_activity_matrix[
  ,
  metabolic_task_annotation$metabolic_task,
  drop = FALSE
]

# Align metadata to metabolic activity matrix
metadata_df <- GSE274934@meta.data
metadata_df <- metadata_df[rownames(metabolic_activity_matrix), , drop = FALSE]

# Check cell order
stopifnot(all(rownames(metabolic_activity_matrix) == rownames(metadata_df)))

# Define cluster labels
cluster_labels <- as.character(metadata_df$seurat_clusters)

# Count cells per cluster
cluster_cell_counts <- table(cluster_labels)

# Sum metabolic activity by cluster
cluster_sum_matrix <- rowsum(
  metabolic_activity_matrix,
  group = cluster_labels,
  reorder = FALSE
)

# Convert cluster sums to cluster means
cluster_mean_matrix <- sweep(
  cluster_sum_matrix,
  MARGIN = 1,
  STATS = as.numeric(cluster_cell_counts[rownames(cluster_sum_matrix)]),
  FUN = "/"
)

# Keep only tasks present after aggregation
available_ordered_tasks <- metabolic_task_annotation$metabolic_task[
  metabolic_task_annotation$metabolic_task %in% colnames(cluster_mean_matrix)
]

# Report selected tasks missing after aggregation
missing_after_aggregation_tasks <- setdiff(
  metabolic_task_annotation$metabolic_task,
  colnames(cluster_mean_matrix)
)

missing_after_aggregation_tasks

# Stop if no selected tasks remain
stopifnot(length(available_ordered_tasks) > 0)

# Reorder cluster mean matrix using available selected tasks only
cluster_mean_matrix <- cluster_mean_matrix[
  ,
  available_ordered_tasks,
  drop = FALSE
]

# Keep annotation only for plotted tasks
metabolic_task_annotation <- metabolic_task_annotation |>
  filter(metabolic_task %in% available_ordered_tasks)

# Match annotation order to heatmap task order
metabolic_task_annotation <- metabolic_task_annotation[
  match(colnames(cluster_mean_matrix), metabolic_task_annotation$metabolic_task),
  ,
  drop = FALSE
]

# Check annotation order
stopifnot(identical(metabolic_task_annotation$metabolic_task, colnames(cluster_mean_matrix)))

# Scale each metabolic task across clusters
cluster_mean_scaled_matrix <- scale(cluster_mean_matrix)
cluster_mean_scaled_matrix[is.na(cluster_mean_scaled_matrix)] <- 0

# Create heatmap matrix
# Rows = metabolic tasks, columns = clusters
heatmap_matrix <- t(cluster_mean_scaled_matrix)

# Create unscaled matrix for optional activity distribution annotation
heatmap_matrix_unscaled <- t(cluster_mean_matrix)

# Create cluster-level annotation table
cluster_annotation_df <- metadata_df |>
  group_by(seurat_clusters) |>
  summarise(
    mean_C1_Core = mean(C1_EpiCore, na.rm = TRUE),
    n_cells = n(),
    .groups = "drop"
  ) |>
  as.data.frame()

# Convert cluster labels to character
cluster_annotation_df$seurat_clusters <- as.character(cluster_annotation_df$seurat_clusters)

# Match annotation order to heatmap columns
cluster_annotation_df <- cluster_annotation_df[
  match(colnames(heatmap_matrix), cluster_annotation_df$seurat_clusters),
  ,
  drop = FALSE
]

# Set rownames for annotation
rownames(cluster_annotation_df) <- cluster_annotation_df$seurat_clusters

# Check cluster annotation order
stopifnot(!any(is.na(cluster_annotation_df$seurat_clusters)))
stopifnot(identical(rownames(cluster_annotation_df), colnames(heatmap_matrix)))

# Match row task category to heatmap rows
row_task_category <- metabolic_task_annotation$task_category[
  match(rownames(heatmap_matrix), metabolic_task_annotation$metabolic_task)
]

# Check row annotation
stopifnot(!any(is.na(row_task_category)))

# Define heatmap color scale
metabolic_activity_color_fun <- colorRamp2(
  c(-2, -1, 0, 1, 2),
  c("#053061", "#4393C3", "#F7F7F7", "#D6604D", "#67001F")
)

# Define C1 score color scale
c1_score_color_fun <- colorRamp2(
  c(
    min(cluster_annotation_df$mean_C1_Core, na.rm = TRUE),
    median(cluster_annotation_df$mean_C1_Core, na.rm = TRUE),
    max(cluster_annotation_df$mean_C1_Core, na.rm = TRUE)
  ),
  c("#313695", "#F7F7F7", "#A50026")
)

# Define task category colors
task_category_colors <- c(
  Sugar = "#E64B35FF",
  Fat = "#4DBBD5FF",
  Protein = "#00A087FF",
  Mitochondrial = "#3C5488FF",
  ROS = "#F39B7FFF"
)

# Create top cluster annotations
top_cluster_annotation <- HeatmapAnnotation(
  `Mean C1 raw` = cluster_annotation_df$mean_C1_Core,
  `Cells` = anno_barplot(
    cluster_annotation_df$n_cells,
    gp = gpar(fill = "grey40", col = NA),
    border = FALSE,
    height = unit(1, "cm")
  ),
  col = list(
    `Mean C1 raw` = c1_score_color_fun
  ),
  annotation_name_side = "left",
  annotation_name_gp = gpar(fontsize = 8),
  annotation_legend_param = list(
    `Mean C1 raw` = list(title = "Mean C1-EpiCore")
  )
)

# Create row task family annotation
left_task_annotation <- rowAnnotation(
  `Task family` = row_task_category,
  col = list(
    `Task family` = task_category_colors
  ),
  annotation_name_gp = gpar(fontsize = 8),
  annotation_legend_param = list(
    `Task family` = list(title = "Task family")
  )
)

# Optional row annotation for unscaled activity distribution
left_activity_annotation <- rowAnnotation(
  `Task family` = row_task_category,
  `Activity distribution` = anno_boxplot(
    heatmap_matrix_unscaled,
    which = "row",
    border = FALSE,
    outline = FALSE,
    axis = TRUE,
    width = unit(2.3, "cm"),
    gp = gpar(fill = "grey70", col = "grey40")
  ),
  col = list(
    `Task family` = task_category_colors
  ),
  annotation_name_gp = gpar(fontsize = 8),
  annotation_legend_param = list(
    `Task family` = list(title = "Task family")
  )
)

# Define heatmap title
heatmap_title <- paste0(
  "Cluster-level metabolic task activity heatmap\n",
  nrow(heatmap_matrix),
  " selected tasks | row z-score | Pearson distance | average linkage"
)

# Add title padding
ht_opt$TITLE_PADDING <- unit(c(10, 10), "points")

# Save heatmap as PNG
png(
  filename = "output/figures/metabolic_selected_tasks_heatmap_cluster_level.png",
  width = 12,
  height = 14,
  units = "in",
  res = 600
)

selected_task_heatmap <- Heatmap(
  heatmap_matrix,
  name = "Metabolic\nactivity",
  col = metabolic_activity_color_fun,

  left_annotation = left_task_annotation,
  top_annotation = top_cluster_annotation,

  cluster_rows = TRUE,
  cluster_columns = TRUE,

  clustering_distance_rows = "pearson",
  clustering_distance_columns = "pearson",
  clustering_method_rows = "average",
  clustering_method_columns = "average",

  show_row_dend = TRUE,
  show_column_dend = TRUE,

  row_title = "Selected metabolic tasks",
  column_title = heatmap_title,

  column_title_gp = gpar(
    fontsize = 13,
    fontface = "bold",
    lineheight = 1.15
  ),
  row_title_gp = gpar(
    fontsize = 11,
    fontface = "bold"
  ),

  row_names_gp = gpar(fontsize = 6),
  column_names_gp = gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 0,

  border = TRUE,
  rect_gp = gpar(col = "grey90", lwd = 0.3),

  heatmap_legend_param = list(
    title = "Row z-score",
    at = c(-2, -1, 0, 1, 2),
    labels = c("Low", "-1", "0", "+1", "High")
  )
)

draw(
  selected_task_heatmap,
  heatmap_legend_side = "right",
  annotation_legend_side = "right"
)

dev.off()

# Save heatmap as PDF
pdf(
  file = "output/figures/Metabolic_selected_tasks_heatmap_cluster_level.pdf",
  width = 12,
  height = 14,
  useDingbats = FALSE
)

selected_task_heatmap <- Heatmap(
  heatmap_matrix,
  name = "Metabolic\nactivity",
  col = metabolic_activity_color_fun,

  left_annotation = left_task_annotation,
  top_annotation = top_cluster_annotation,

  cluster_rows = TRUE,
  cluster_columns = TRUE,

  clustering_distance_rows = "pearson",
  clustering_distance_columns = "pearson",
  clustering_method_rows = "average",
  clustering_method_columns = "average",

  show_row_dend = TRUE,
  show_column_dend = TRUE,

  row_title = "Selected metabolic tasks",
  column_title = heatmap_title,

  column_title_gp = gpar(
    fontsize = 13,
    fontface = "bold",
    lineheight = 1.15
  ),
  row_title_gp = gpar(
    fontsize = 11,
    fontface = "bold"
  ),

  row_names_gp = gpar(fontsize = 6),
  column_names_gp = gpar(fontsize = 10, fontface = "bold"),
  column_names_rot = 0,

  border = TRUE,
  rect_gp = gpar(col = "grey90", lwd = 0.3),

  heatmap_legend_param = list(
    title = "Row z-score",
    at = c(-2, -1, 0, 1, 2),
    labels = c("Low", "-1", "0", "+1", "High")
  )
)

draw(
  selected_task_heatmap,
  heatmap_legend_side = "right",
  annotation_legend_side = "right"
)

dev.off()

# Reset ComplexHeatmap options
ht_opt(RESET = TRUE)

```

# Section 16. Checkpoint

This checkpoint saves the Seurat object after adding regulatory, pathway, secreted-protein, and Hallmark activity layers.

```{r}

DefaultAssay(GSE274934) <- "RNA"
qs_save(
  GSE274934,
  file = file.path("output/scRNA/objects/GSE274934_regulatory_ready.qs2"),
  nthreads = 10,
  compress_level = 10
)

```
