---
title: "MAP 09 - C1-Core Refined Signature"
format: html
---

# Section 1. Overview

This notebook is the authoritative C1 shortlist and `C1-Core` signature refinement step for the MAP workflow.

**MAP role:** step 09. This notebook runs after the metabolic layer and before the broad statistics notebook because it builds the final `C1-Core` score that downstream statistical checks can use.

## Section 1.1. Reader Guide

- Inputs are precomputed result tables from Spearman correlation, tradeSeq/GAM, TF activity, hdWGCNA hubs, kinetic velocity, and the original core C1 genes.
- The notebook standardizes source tables, defines overlap sets, exports joined evidence tables, and plots compact intersections.
- Main outputs are `C1_shortlist_all_source_evidence.csv`, `C1_Core_UP.csv`, `C1_Core_DOWN.csv`, the `C1_Core` score, and focused score-validation summaries.
- Read this as the trusted C1-Core consolidation layer: each candidate gene should remain traceable to the evidence source that supported it.

## Section 1.2. MAP Section Boundaries

### Section 1.2.1. Notebook Flow

1. Load all C1 evidence sources.
2. Build one source-aware evidence table.
3. Plot only compact C1 evidence intersections.
4. Build/refine UP and DOWN `C1-Core` signatures.
5. Score the atlas with `C1-Core`.
6. Export focused validation tables for the step 10 statistics notebook.

## Section 1.3. Setup

Load libraries, define output folders, and keep the plotting style consistent with the rest of the CSS_samples workflow.

```{r}

packages <- c(
  "tidyverse",
  "Seurat",
  "harmony",
  "patchwork",
  "data.table",
  "fst",
  "UCell",
  "viridis",
  "RColorBrewer",
  "qs2",
  "ggsci",
  "scCustomize", 
  "future",
  "ComplexUpset"
)

library(knitr)
library(gt)
library(gtExtras)

purrr::walk(packages, library, character.only = TRUE)

outdir_tables <- "output/tables/"
outdir_figures <- "output/figures/"
dir.create(outdir_tables, recursive = TRUE, showWarnings = FALSE)
dir.create(outdir_figures, recursive = TRUE, showWarnings = FALSE)


plan(multicore, workers = 4)
options(future.globals.maxSize = 20 * 1024^3)

umap_theme <- theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 11, color = "grey30"),

    # Remove background grid
    panel.grid = element_blank(),

    # Keep axis titles
    axis.title = element_text(size = 13),

    # Remove axis numbers, ticks, and lines
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),

    # Legend formatting
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10)
  )
css_combined <- qs_read(file = "output/objects/css_combined_processed.qs2", nthreads = 10)

core_genes <- vroom::vroom("input/Signatures/Core_genes_tidy.csv")
print(core_genes, n = 20)

```

# Section 2. Load C1 Evidence Sources

Read the significant gene or TF tables generated by earlier notebooks. Each table is kept as a separate object so the source of each C1 shortlist signal remains explicit.

```{r}

# hdWGCNA hub genes provide network-level support.
hdWGCNA_hub_genes <- vroom::vroom("output/tables/hdWGCNA_hub_genes.csv") |>
  mutate(
    dataset = "hdWGCNA_hub_genes"
  ) |>
    rename(gene = gene_name)

hdWGCNA_hub_genes

# Spearman genes capture monotonic association with the C1 score.
corr_genes <- vroom::vroom(
  "output/tables/corr_spearman_genes_along_C1.csv",
  show_col_types = FALSE
) |>
  mutate(
    Direction = ifelse(rho > 0, "Positive", "Negative"),
    dataset = "Spearman_genes_C1"
  ) |>
  select(gene, rho, FDR, Direction, dataset) |>
  filter(FDR < 0.05)

# tradeSeq genes capture dynamic changes along the C1 trajectory model.
tradeSeq_genes <- vroom::vroom(
  "output/tables/tradeSeq_gam_on_genes_along_C1.csv",
  show_col_types = FALSE
) |>
  mutate(
    Direction = ifelse(logFClineage1 > 0, "Up", "Down"),
    dataset = "tradeSeq_genes_C1"
  ) |>
  select(gene, waldStat, logFClineage1, FDR, Direction, dataset) |>
  filter(FDR < 0.05)

# TF activity correlations add regulatory-level evidence from decoupleR.
tf_corr <- vroom::vroom(
  "output/tables/corr_spearman_transcription_factor_along_C1_decoupleR.csv",
  show_col_types = FALSE
) |>
  mutate(
    Direction = ifelse(rho > 0, "Positive", "Negative"),
    FDR = fdr,
    gene = TF,
    dataset = "TF_activity_C1"
  ) |>
  select(gene, rho, FDR, Direction, dataset) |>
  filter(FDR < 0.05)

Differential_kinetic_evidence <- vroom::vroom("output/tables/Velocity_differential_kinetic_test.csv") |>
  rename(gene = "...1") |>
  select(-fit_diff_kinetics) |>
  mutate(
    fdr_kinetics = p.adjust(fit_pval_kinetics, method = "BH"),
    dataset = "Differential_kinetic_velocity"
  ) |>
  dplyr::rename(
    pval_kinetics = fit_pval_kinetics
  )

```

# Section 3. C1 Evidence Summary and Gene Sets

This section is the handoff between raw evidence loading and shortlist construction. It counts the evidence sources first, then builds the exact gene-set lists used by the tables and the single compact UpSet plot.

## Section 3.1. Evidence Counts and Gene Sets

Summarize how many significant genes come from each evidence source, then define the exact gene-set combinations used by the downstream tables and plots.

```{r}

# Summarize significant genes by method and direction.
# Each source keeps its own dataset label so the overlap evidence remains traceable.

summary_table <- bind_rows(
  corr_genes         |> dplyr::select(gene, FDR, Direction, dataset),
  tradeSeq_genes     |> dplyr::select(gene, FDR, Direction, dataset),
  tf_corr            |> dplyr::select(gene, FDR, Direction, dataset),
  hdWGCNA_hub_genes  |> dplyr::select(gene, module, kME, dataset),
  Differential_kinetic_evidence |> dplyr::select(gene, dataset)
) |>
  dplyr::mutate(Direction = tidyr::replace_na(Direction, "Not direction-coded")) |>
  dplyr::count(dataset, Direction, name = "n_genes") |>
  dplyr::arrange(dataset, Direction)

summary_table

# Define gene sets for overlap analysis.
# These lists support progressively broader overlap comparisons.

# Basic comparison: tradeSeq genes and hdWGCNA hubs.
gene_sets_basic <- list(
  tradeSeq_C1       = tradeSeq_genes$gene,
  hdWGCNA_hub_genes = hdWGCNA_hub_genes$gene
)

# Comparison: tradeSeq genes, hdWGCNA hubs and old 330.
gene_sets_original <- list(
  tradeSeq_C1       = tradeSeq_genes$gene,
  hdWGCNA_hub_genes = hdWGCNA_hub_genes$gene,
  core_genes = core_genes$Core_genes_UP_DOWN
)

# Add Spearman-correlated genes.
gene_sets_with_spearman <- list(
  Spearman_C1       = corr_genes$gene,
  tradeSeq_C1       = tradeSeq_genes$gene,
  hdWGCNA_hub_genes = hdWGCNA_hub_genes$gene
)

# Add TF activity genes.
gene_sets_with_tf <- list(
  Spearman_C1       = corr_genes$gene,
  tradeSeq_C1       = tradeSeq_genes$gene,
  TF_activity_C1    = tf_corr$gene,
  hdWGCNA_hub_genes = hdWGCNA_hub_genes$gene
)

# Full comparison including core_genes.
gene_sets_with_full <- list(
  Spearman_C1       = corr_genes$gene,
  tradeSeq_C1       = tradeSeq_genes$gene,
  TF_activity_C1    = tf_corr$gene,
  hdWGCNA_hub_genes = hdWGCNA_hub_genes$gene,
  core_genes = core_genes$Core_genes_UP_DOWN,
  Differential_kinetic = Differential_kinetic_evidence$gene
)


```

# Section 4. Evidence Table and Shortlist Statistics

This section creates the consultation tables before any plots are drawn. The first table keeps all source-level evidence in one union view, and the later exports save the genes shared across each plotted comparison.

```{r}

# Keep source-specific columns readable after joining multiple evidence tables.
tradeSeq_evidence <- tradeSeq_genes |>
  dplyr::distinct(gene, .keep_all = TRUE) |>
  dplyr::select(-dataset) |>
  dplyr::rename(
    tradeseq_wald_stat = waldStat,
    tradeseq_logfc_lineage1 = logFClineage1,
    tradeseq_fdr = FDR,
    tradeseq_direction = Direction
  )

spearman_evidence <- corr_genes |>
  dplyr::distinct(gene, .keep_all = TRUE) |>
  dplyr::select(-dataset) |>
  dplyr::rename(
    spearman_rho = rho,
    spearman_fdr = FDR,
    spearman_direction = Direction
  )

tf_activity_evidence <- tf_corr |>
  dplyr::distinct(gene, .keep_all = TRUE) |>
  dplyr::select(-dataset) |>
  dplyr::rename(
    tf_activity_rho = rho,
    tf_activity_fdr = FDR,
    tf_activity_direction = Direction
  )

hdwgcna_evidence <- hdWGCNA_hub_genes |>
  dplyr::distinct(gene, .keep_all = TRUE) |>
  dplyr::select(-dataset) |>
  dplyr::rename(
    hdwgcna_module = module,
    hdwgcna_kme = kME
  )

Differential_kinetic_evidence

```

```{r}


library(dplyr)
library(tibble)

# Set effect-size thresholds used to assign an effective direction.
logfc_min  <- 0.25
rho_min    <- 0.10
tf_rho_min <- 0.10

# Union-level evidence table across all sources
all_evidence_table <- tibble(
  gene = unique(unlist(gene_sets_with_full, use.names = FALSE))
) |>
  left_join(tradeSeq_evidence, by = "gene") |>
  left_join(spearman_evidence, by = "gene") |>
  left_join(tf_activity_evidence, by = "gene") |>
  left_join(hdwgcna_evidence, by = "gene") |>
  left_join(Differential_kinetic_evidence, by = "gene") |>
  mutate(
    # effective directions based on sensible effect-size thresholds
    tradeSeq_dir_effective = case_when(
      !is.na(tradeseq_logfc_lineage1) & tradeseq_logfc_lineage1 >=  logfc_min ~ "Up",
      !is.na(tradeseq_logfc_lineage1) & tradeseq_logfc_lineage1 <= -logfc_min ~ "Down",
      TRUE ~ NA_character_
    ),

    spearman_dir_effective = case_when(
      !is.na(spearman_rho) & spearman_rho >=  rho_min ~ "Up",
      !is.na(spearman_rho) & spearman_rho <= -rho_min ~ "Down",
      TRUE ~ NA_character_
    ),

    tf_dir_effective = case_when(
      !is.na(tf_activity_rho) & tf_activity_rho >=  tf_rho_min ~ "Up",
      !is.na(tf_activity_rho) & tf_activity_rho <= -tf_rho_min ~ "Down",
      TRUE ~ NA_character_
    ),

    # source membership
    in_tradeSeq_C1 = !is.na(tradeseq_fdr),
    in_Spearman_C1 = !is.na(spearman_fdr),
    in_TF_activity_C1 = !is.na(tf_activity_fdr),
    in_hdWGCNA_hub_genes = !is.na(hdwgcna_module),
    in_differential_kinetic = !is.na(fdr_kinetics),

    # number of supporting sources
    n_supporting_sources = rowSums(
      pick(
        in_tradeSeq_C1,
        in_Spearman_C1,
        in_TF_activity_C1,
        in_hdWGCNA_hub_genes,
        in_differential_kinetic
      ),
      na.rm = TRUE
    ),

    # final direction with fallback:
    # tradeSeq logFC -> Spearman -> TF activity -> Unresolved
    direction_final = case_when(
      !is.na(tradeSeq_dir_effective) ~ tradeSeq_dir_effective,
      is.na(tradeSeq_dir_effective) & !is.na(spearman_dir_effective) ~ spearman_dir_effective,
      is.na(tradeSeq_dir_effective) & is.na(spearman_dir_effective) & !is.na(tf_dir_effective) ~ tf_dir_effective,
      TRUE ~ "Unresolved"
    ),

    direction_source = case_when(
      !is.na(tradeSeq_dir_effective) ~ "tradeSeq_logFC",
      is.na(tradeSeq_dir_effective) & !is.na(spearman_dir_effective) ~ "Spearman",
      is.na(tradeSeq_dir_effective) & is.na(spearman_dir_effective) & !is.na(tf_dir_effective) ~ "TF_activity",
      TRUE ~ "None"
    ),

    # whether secondary sources support the chosen direction
    direction_support = case_when(
      direction_final == "Up" &
        (is.na(spearman_dir_effective) | spearman_dir_effective == "Up") &
        (is.na(tf_dir_effective) | tf_dir_effective == "Up") ~ "Supported_up",

      direction_final == "Down" &
        (is.na(spearman_dir_effective) | spearman_dir_effective == "Down") &
        (is.na(tf_dir_effective) | tf_dir_effective == "Down") ~ "Supported_down",

      direction_final == "Unresolved" ~ "Unresolved",

      TRUE ~ "Conflicting_or_partial"
    ),

    evidence_tier = case_when(
      n_supporting_sources >= 4 ~ "Tier_1_multi_evidence",
      n_supporting_sources == 3 ~ "Tier_2_strong",
      n_supporting_sources == 2 ~ "Tier_3_supported",
      n_supporting_sources == 1 ~ "Tier_4_single_source",
      TRUE ~ "Tier_5_no_support"
    )
  ) |>
  arrange(
    desc(n_supporting_sources),
    desc(direction_support %in% c("Supported_up", "Supported_down")),
    desc(abs(tradeseq_logfc_lineage1)),
    desc(abs(spearman_rho)),
    desc(abs(tf_activity_rho)),
    desc(hdwgcna_kme)
  )

head(all_evidence_table)

# Split genes by final direction for quick inspection.
UP_genes <- all_evidence_table |>
  filter(direction_final == "Up") |>
  pull(gene)

DOWN_genes <- all_evidence_table |>
  filter(direction_final == "Down") |>
  pull(gene)

UNRESOLVED_genes <- all_evidence_table |>
  filter(direction_final == "Unresolved") |>
  pull(gene)

# Summarize direction calls and evidence tiers before export.
all_evidence_table |>
  count(direction_final, direction_source, sort = TRUE)

all_evidence_table |>
  count(evidence_tier, direction_support, sort = TRUE)

readr::write_csv(
  all_evidence_table,
  file.path("output/tables/C1_shortlist_all_source_evidence.csv")
)

if (requireNamespace("openxlsx", quietly = TRUE)) {
  openxlsx::write.xlsx(
    all_evidence_table,
    file.path("output/tables/C1_shortlist_all_source_evidence.xlsx")
  )
}

```

# Section 5. Compact ComplexUpset Plot

ComplexUpset is used here because it shows exact intersection sizes as bars while retaining the set membership matrix.

```{r}

# Convert a named list of gene vectors into the logical matrix required by ComplexUpset.
make_upset_input <- function(gene_sets) {
  all_genes <- unique(unlist(gene_sets, use.names = FALSE))

  upset_data <- tibble::tibble(gene = all_genes)

  for (set_name in names(gene_sets)) {
    upset_data[[set_name]] <- all_genes %in% unique(gene_sets[[set_name]])
  }

  upset_data
}

# Build one ComplexUpset plot with shared styling across all C1 overlap views.
plot_c1_upset <- function(gene_sets, title, subtitle) {
  upset_data <- make_upset_input(gene_sets)

  ComplexUpset::upset(
    upset_data,
    intersect = names(gene_sets),
    name = "Genes",
    min_size = 10,
    width_ratio = 0.2,
    sort_intersections_by = "cardinality",

    base_annotations = list(
      "Intersection size" = ComplexUpset::intersection_size(counts = TRUE)
    ),

    queries = list(
      ComplexUpset::upset_query(
        intersect = names(gene_sets),   # full intersection across all sets
        fill = "red"
      )
    )
  ) +
    ggplot2::labs(
      title = title,
      subtitle = subtitle
    ) +
    ggplot2::theme_minimal() +
    ggplot2::theme(
      plot.title = ggplot2::element_text(size = 15, face = "bold"),
      plot.subtitle = ggplot2::element_text(size = 10, color = "grey30")
    )
}

# One complete overlap figure replaces the previous repeated partial-overlap
# figures. It is easier to read and keeps the notebook lean.
p_upset_full <- plot_c1_upset(
  gene_sets_with_full,
  title = "C1-Core evidence overlap",
  subtitle = "Core C1 genes, Spearman, tradeSeq, TF activity, hdWGCNA hubs, and kinetic velocity evidence (min 10)"
)

p_upset_full

ggsave(
  filename = file.path(outdir_figures, "C1_Core_evidence_upset.png"),
  plot     = p_upset_full,
  width    = 24,
  height   = 14,
  dpi      = 600,
  bg       = "white"
)

```

# Section 6. C1-Core Signature and UCell Scoring

Plasticity-related signatures are scored with UCell, smoothed on the Harmony graph, and collapsed into a single `C1_Core` score used throughout the notebook.

```{r}

genes_up <- tibble(gene = c(
  "HOXB7",
  #"THY1",
  "E2F4",
  "MCM6",
  "SF3B1",
  "E2F1",
  "RRM2"
))

genes_down <- tibble(gene = c(
  "SCGB1A1",
  "HSPG2",
  "SCGB3A1",
  "NKX2-1",
  #"SFTPC",
  "TFF1"
))


OLD_core_genes <- vroom::vroom("input/Signatures/Core_genes_tidy.csv") |>
  select(-Core_genes_UP_DOWN)

core_genes_long <- bind_rows(
  OLD_core_genes %>%
    select(gene = Core_genes_DOWN) %>%
    mutate(direction = "Down"),

  OLD_core_genes %>%
    select(gene = Core_genes_UP) %>%
    mutate(direction = "Up")
) %>%
  drop_na(gene)


C1_Core_UP <- all_evidence_table |>
  filter(gene %in% OLD_core_genes$Core_genes_UP) |>
  filter(direction_support == "Supported_up") |>
  #filter(direction_support != "Supported_down") |>
  #filter(!hdwgcna_module %in% c("Plasticity-M6", "Plasticity-M3")) |>
  filter(!hdwgcna_module %in% c("Plasticity-M3")) |>
  bind_rows(genes_up) |>
  distinct(gene, .keep_all = TRUE)

table(C1_Core_UP$hdwgcna_module)
nrow(C1_Core_UP)

C1_Core_DOWN <- all_evidence_table |>
  filter(gene %in% OLD_core_genes$Core_genes_DOWN) |>
  filter(direction_support == "Supported_down") |>
  #filter(direction_support != "Supported_up") |>
  #filter(!hdwgcna_module %in% c("Plasticity-M6", "Plasticity-M3")) |>
  filter(!hdwgcna_module %in% c("Plasticity-M3")) |>
  bind_rows(genes_down)|>
  distinct(gene, .keep_all = T)

table(C1_Core_DOWN$hdwgcna_module)
nrow(C1_Core_DOWN)

nrow(C1_Core_DOWN) + nrow(C1_Core_UP)

C1_Core_signature <- bind_rows(
  C1_Core_UP |> mutate(C1_Core_direction = "UP"),
  C1_Core_DOWN |> mutate(C1_Core_direction = "DOWN")
) |>
  distinct(gene, C1_Core_direction, .keep_all = TRUE)

readr::write_csv(
  C1_Core_signature,
  file.path(outdir_tables, "C1_Core_signature.csv")
)

C1_Core_signature <- vroom::vroom("output/tables/C1_Core_signature.csv")
nrow(C1_Core_signature)

```

```{r UCell score}

# Score the refined UP and DOWN signatures.
# The final exported score is C1_Core; raw UCell helper columns are removed after smoothing.

# C1 EpiCore scoring with UCell

# Set a reproducible seed for scoring.
set.seed(1234)

# Load the scoring packages used in this block.
library(UCell)
library(BiocParallel)

# Register the BiocParallel backend for UCell scoring.

BPPARAM <- MulticoreParam(workers = 9)
register(BPPARAM)

# Define small helpers for z-scoring and signature checks.

zscore <- function(x) {
  as.numeric(scale(x))
}

check_signature <- function(gene_vector, object_genes, signature_name) {
  genes_input <- unique(na.omit(gene_vector))
  genes_found <- intersect(genes_input, object_genes)
  genes_lost  <- setdiff(genes_input, object_genes)

  message(
    signature_name, ": ",
    length(genes_found), "/", length(genes_input),
    " genes found in object."
  )

  if (length(genes_found) < 5) {
    stop(
      signature_name,
      " has fewer than 5 genes present in the object. ",
      "The score would be unstable."
    )
  }

  list(
    genes_found = genes_found,
    genes_lost  = genes_lost
  )
}

# Build C1 EpiCore signatures

object_genes <- rownames(css_combined)

down_checked <- check_signature(
  gene_vector    = C1_Core_DOWN$gene,
  object_genes   = object_genes,
  signature_name = "DOWN_Signature"
)

up_checked <- check_signature(
  gene_vector    = C1_Core_UP$gene,
  object_genes   = object_genes,
  signature_name = "UP_Signature"
)

gene_list <- list(
  DOWN_Signature = down_checked$genes_found,
  UP_Signature   = up_checked$genes_found
)

# Optional report of genes missing from the Seurat object
missing_signature_genes <- list(
  DOWN_Signature_missing = down_checked$genes_lost,
  UP_Signature_missing   = up_checked$genes_lost
)

missing_signature_genes

# Add UCell scores

css_combined <- AddModuleScore_UCell(
  obj      = css_combined,
  features = gene_list,
  BPPARAM  = BPPARAM
)

# Smooth UCell scores in Harmony space
# Smoothing is performed in Harmony space, not UMAP space.
# This gives a neighborhood-level estimate of UP and DOWN activity.

css_combined <- SmoothKNN(
  obj             = css_combined,
  signature.names = c("UP_Signature_UCell", "DOWN_Signature_UCell"),
  reduction       = "harmony"
)

# Compute smoothed C1 EpiCore score
# Formula:
#   C1_Core = scale(scale(UP UCell kNN) - scale(DOWN UCell kNN))
#
# Interpretation:
#   - Higher values = stronger C1-like epithelial state
#   - Lower values  = stronger retained differentiated / lineage-like state
#
# Note:
#   This is a smoothed neighborhood-level score.
#   It is ideal for visualization, gradients, and trajectory/state analysis.

md <- css_combined@meta.data

up_z <- zscore(md$UP_Signature_UCell_kNN)
down_z <- zscore(md$DOWN_Signature_UCell_kNN)

css_combined$C1_Core_raw <- up_z - down_z
css_combined$C1_Core <- zscore(up_z - down_z)

# Clean temporary UCell columns
# Keep only the final C1_Core score.
# Remove both raw UCell and smoothed kNN UCell intermediate columns.

cols_remove <- c(
  "UP_Signature_UCell",
  "DOWN_Signature_UCell",
  "UP_Signature_UCell_kNN",
  "DOWN_Signature_UCell_kNN"
)

cols_remove <- intersect(cols_remove, colnames(css_combined@meta.data))

css_combined@meta.data <- css_combined@meta.data[
  ,
  setdiff(colnames(css_combined@meta.data), cols_remove),
  drop = FALSE
]

# Stop backend

bpstop(BPPARAM)

```

# Section 7. Focused C1-Core Validation

The C1 notebook keeps only the checks needed to read the refined score. Broad pairwise tests, adjusted models, and permutation checks belong in `10_MAP_extensive_statistics.qmd`.

```{r}

# Plot CytoTRACE2 potency labels on UMAP
Anno_umap_plot <- DimPlot_scCustom(
  seurat_object = css_combined,
  group.by = "predicted.ann_finest_level"
)  +
  labs(x = "UMAP1", y = "UMAP2", title = "Annotataion HLCA") &
  umap_theme

Anno_umap_plot

# Save main score UMAP
ggsave(
  filename = file.path(outdir_figures, "Anno_umap_plot.png"),
  plot = Anno_umap_plot,
  width = 10,
  height = 6,
  dpi = 600,
  bg = "white"
)

# Plot main continuous scores on UMAP
score_umap_plot <- FeaturePlot_scCustom(
  css_combined,
  features = c("C1_Core"),
  reduction = "umap",
  na_cutoff = NA
) &
  scale_color_viridis_c(option = "magma") &
  labs(x = "UMAP1", y = "UMAP2") &
  umap_theme

score_umap_plot

# Save main score UMAP
ggsave(
  filename = file.path(outdir_figures, "Anno_umap_plot.png"),
  plot = score_umap_plot,
  width = 16,
  height = 6,
  dpi = 600,
  bg = "white"
)

# Plot main continuous scores on UMAP
score_umap_plot <- FeaturePlot_scCustom(
  css_combined,
  features = c("C1_Core", "C1_Core_raw", "CytoTRACE2", "G2M.Score"),
  reduction = "umap",
  na_cutoff = NA
) &
  scale_color_viridis_c(option = "magma") &
  labs(x = "UMAP1", y = "UMAP2") &
  umap_theme

score_umap_plot

# Save main score UMAP
ggsave(
  filename = file.path(outdir_figures, "C1_Core_plots.png"),
  plot = score_umap_plot,
  width = 12,
  height = 8,
  dpi = 600,
  bg = "white"
)

# Plot main continuous scores on UMAP
score_umap_plot <- FeaturePlot_scCustom(
  css_combined,
  features = "C1_Core",
  split.by = "orig.ident",
  num_columns = 2,
  reduction = "umap",
  na_cutoff = NA
) &
  scale_color_viridis_c(option = "magma") &
  labs(x = "UMAP1", y = "UMAP2") &
  umap_theme

score_umap_plot

# Save main score UMAP
ggsave(
  filename = file.path(outdir_figures, "C1_Core_SAMPLE.png"),
  plot = score_umap_plot,
  width = 12,
  height = 8,
  dpi = 600,
  bg = "white"
)

# Plot C1-Core alone
C1_Core_umap_plot <- FeaturePlot_scCustom(
  css_combined,
  features = "C1_Core",
  reduction = "umap",
  na_cutoff = NA
) +
  scale_color_viridis_c(option = "magma") +
  labs(x = "UMAP1", y = "UMAP2", color = "C1-Core", title = "C1-Core", subtitle = "Refined version of the C1 score, retaing 113 core genes ") +
  umap_theme

C1_Core_umap_plot

# Save C1-Core UMAP
ggsave(
  filename = file.path(outdir_figures, "C1_Core_umap.png"),
  plot = C1_Core_umap_plot,
  width = 8,
  height = 6,
  dpi = 600,
  bg = "white"
)

# Plot continuous trajectory and cell-cycle scores
trajectory_score_umap_plot <- FeaturePlot_scCustom(
  css_combined,
  features = c("C1_Core", "latent_time", "G2M.Score", "S.Score", "CytoTRACE2"),
  reduction = "umap",
  na_cutoff = NA
) &
  scale_color_viridis_c(option = "magma") &
  labs(x = "UMAP1", y = "UMAP2") &
  umap_theme

# Extract NPG colors for CytoTRACE2 potency groups
cytotrace_potency_levels <- levels(factor(css_combined$CytoTRACE2_Potency))
cytotrace_potency_colors <- ggsci::pal_npg("nrc")(length(cytotrace_potency_levels))
names(cytotrace_potency_colors) <- cytotrace_potency_levels

# Plot CytoTRACE2 potency labels on UMAP
cytotrace_potency_umap_plot <- DimPlot_scCustom(
  seurat_object = css_combined,
  group.by = "CytoTRACE2_Potency",
  label = TRUE,
  repel = TRUE,
  label.box = TRUE,
  colors_use = cytotrace_potency_colors
) +
  umap_theme +
  NoLegend()

# Combine continuous scores and potency labels
combined_score_potency_umap_plot <- trajectory_score_umap_plot + cytotrace_potency_umap_plot

combined_score_potency_umap_plot

# Save combined UMAP panel
ggsave(
  filename = file.path(outdir_figures, "C1_Core_CytoTRACE2_Potency_umap.png"),
  plot = combined_score_potency_umap_plot,
  width = 14,
  height = 12,
  dpi = 600,
  bg = "white"
)

# Prepare metadata
metadata_df <- css_combined@meta.data |>
  mutate(
    seurat_clusters = factor(seurat_clusters),
    predicted.ann_finest_level = factor(predicted.ann_finest_level)
  )

# Define candidate scores for correlation analysis
candidate_score_columns <- c(
  "C1_score_raw",
  "C1_Core",
  "CytoTRACE2",
  "G2M.Score",
  "S.Score",
  "harmony_1",
  "harmony_2",
  "velocity_confidence_transition",
  "velocity_confidence",
  "velocity_length",
  "latent_time",
  "velocity_self_transition"
)

# Keep only scores present in metadata
available_score_columns <- intersect(candidate_score_columns, colnames(metadata_df))

# Compute Spearman correlations between scores
score_spearman_correlations <- cor(
  metadata_df[, available_score_columns, drop = FALSE],
  method = "spearman",
  use = "pairwise.complete.obs"
) |>
  as.data.frame() |>
  rownames_to_column("score")

score_spearman_correlations

# Save score correlation table
readr::write_csv(
  score_spearman_correlations,
  file.path(outdir_tables, "C1_Core_score_spearman_correlations.csv")
)

# Summarise C1-Core by Seurat cluster
C1_Core_cluster_summary <- metadata_df |>
  group_by(seurat_clusters) |>
  summarise(
    n = sum(!is.na(C1_Core)),
    mean_C1_Core = mean(C1_Core, na.rm = TRUE),
    median_C1_Core = median(C1_Core, na.rm = TRUE),
    sd_C1_Core = sd(C1_Core, na.rm = TRUE),
    mean_C1_score_raw = mean(C1_score_raw, na.rm = TRUE),
    median_C1_score_raw = median(C1_score_raw, na.rm = TRUE),
    sd_C1_score_raw = sd(C1_score_raw, na.rm = TRUE),
    mean_latent_time = mean(latent_time, na.rm = TRUE),
    median_latent_time = median(latent_time, na.rm = TRUE),
    sd_latent_time = sd(latent_time, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(median_C1_Core))

C1_Core_cluster_summary

# Save cluster summary table
readr::write_csv(
  C1_Core_cluster_summary,
  file.path(outdir_tables, "C1_Core_cluster_summary.csv")
)

# Test C1-Core differences across Seurat clusters
C1_Core_cluster_kruskal <- metadata_df |>
  filter(!is.na(C1_Core), !is.na(seurat_clusters)) |>
  rstatix::kruskal_test(C1_Core ~ seurat_clusters)

C1_Core_cluster_kruskal


# Plot C1-Core distribution by Seurat cluster
C1_Core_cluster_violin_plot <- ggplot(
  metadata_df,
  aes(x = seurat_clusters, y = C1_Core, fill = seurat_clusters)
) +
  geom_violin(scale = "width", trim = TRUE, alpha = 0.5) +
  geom_boxplot(width = 0.14, outlier.shape = NA, alpha = 0.85, fill = "white") +
  ggpubr::stat_compare_means(method = "kruskal.test") +
  ggsci::scale_fill_npg() +
  theme_bw(base_size = 12) +
  labs(
    title = "C1-Core by Seurat cluster",
    x = "Seurat cluster",
    y = "C1-Core score"
  ) +
  theme(legend.position = "none")

C1_Core_cluster_violin_plot

# Save C1-Core cluster violin plot
ggsave(
  filename = file.path(outdir_figures, "C1_Core_by_cluster_violin.png"),
  plot = C1_Core_cluster_violin_plot,
  width = 10,
  height = 6,
  dpi = 600,
  bg = "white"
)

```

```{r}


c1_cluster_table <- table(
  css_combined$seurat_clusters,
  css_combined$orig.ident
)


# 1. Save raw counts


c1_cluster_counts_df <- as.data.frame(as.table(c1_cluster_table)) |>
  as_tibble() |>
  rename(
    seurat_cluster = Var1,
    sample = Var2,
    n_cells = Freq
  ) |>
  arrange(seurat_cluster, sample)

c1_cluster_counts_df


# 2. Save within-cluster proportion

c1_cluster_prop <- prop.table(c1_cluster_table, margin = 1)

c1_cluster_prop_df <- as.data.frame(as.table(c1_cluster_prop)) |>
  as_tibble() |>
  rename(
    seurat_cluster = Var1,
    C1_Core_state = Var2,
    cluster_fraction = Freq
  ) |>
  arrange(seurat_cluster, C1_Core_state)

readr::write_csv(
  c1_cluster_prop_df,
  file.path(outdir_tables, "C1_Core_state_by_cluster_proportions.csv")
)


# 3. Save global proportions


c1_global_prop <- prop.table(c1_cluster_table)

c1_global_prop_df <- as.data.frame(as.table(c1_global_prop)) |>
  as_tibble() |>
  rename(
    seurat_cluster = Var1,
    C1_Core_state = Var2,
    global_fraction = Freq
  ) |>
  arrange(seurat_cluster, C1_Core_state)

readr::write_csv(
  c1_global_prop_df,
  file.path(outdir_tables, "C1_Core_state_by_cluster_global_proportions.csv")
)


# 4. Chi-square test and Cramer's V


chisq_res <- chisq.test(c1_cluster_table)

cramer_v <- rcompanion::cramerV(
  css_combined$seurat_clusters,
  css_combined$orig.ident
)

chisq_summary_df <- tibble(
  test = "Pearson Chi-squared test",
  statistic = unname(chisq_res$statistic),
  df = unname(chisq_res$parameter),
  p_value = chisq_res$p.value,
  cramer_v = as.numeric(cramer_v)
)

# Prepare metadata
metadata_df <- css_combined@meta.data |>
  mutate(
    seurat_clusters = factor(seurat_clusters),
    predicted.ann_finest_level = factor(predicted.ann_finest_level)
  )

library(dplyr)
library(ggplot2)
library(ggsci)
library(scales)

# Make sure cluster and sample variables are factors
metadata_df <- metadata_df %>%
  mutate(
    orig.ident = factor(orig.ident),
    seurat_clusters = factor(seurat_clusters)
  )

# Count cells per cluster and sample
cluster_counts <- metadata_df %>%
  count(seurat_clusters, orig.ident, name = "n_cells")

cluster_counts

p_counts <- ggplot(
  cluster_counts,
  aes(
    x = seurat_clusters,
    y = n_cells,
    fill = orig.ident
  )
) +
  geom_col(color = "black", linewidth = 0.2) +
  theme_bw(base_size = 12) +
  scale_fill_npg() +
  labs(
    x = "Cluster",
    y = "Number of cells",
    fill = "Sample",
    title = "Sample contribution to each cluster",
    subtitle = "Absolute number of cells from each sample within each cluster"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12)
  )

p_counts

cluster_proportions <- cluster_counts %>%
  group_by(seurat_clusters) %>%
  mutate(
    total_cells_cluster = sum(n_cells),
    proportion = n_cells / total_cells_cluster,
    percent = proportion * 100
  ) %>%
  ungroup()

cluster_proportions

p_props <- ggplot(
  cluster_proportions,
  aes(
    x = seurat_clusters,
    y = percent,
    fill = orig.ident
  )
) +
  geom_col(color = "black", linewidth = 0.2) +
  theme_bw(base_size = 12) +
  scale_fill_npg() +
  labs(
    x = "Cluster",
    y = "Cells (%)",
    fill = "Sample",
    title = "Sample composition within each cluster",
    subtitle = "Percentage of cells from each sample inside each cluster"
  ) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12)
  )

p_props

ggsave(
  filename = "output/figures/sample_contribution_per_cluster_counts.png",
  plot = p_counts,
  width = 8,
  height = 6,
  dpi = 600
)

ggsave(
  filename = "output/figures/sample_composition_per_cluster_percent.png",
  plot = p_props,
  width = 8,
  height = 6,
  dpi = 600
)

```

```{r}

library(dplyr)
library(tidyr)
library(ggplot2)
library(ggsci)
library(scales)
library(rstatix)

metadata_df <- metadata_df %>%
  mutate(
    orig.ident = factor(orig.ident),
    seurat_clusters = factor(seurat_clusters)
  )

# Count cells per cluster and sample
cluster_counts <- metadata_df %>%
  count(seurat_clusters, orig.ident, name = "n_cells")

# Contingency table
tab <- table(metadata_df$seurat_clusters, metadata_df$orig.ident)

# Global test: are clusters and samples independent?
chisq_global <- chisq.test(tab)

chisq_global

# Effect size
cramer_v <- rstatix::cramer_v(tab,correct = T)

cramer_v

residual_df <- as.data.frame(chisq_global$stdres) %>%
  rename(
    seurat_clusters = Var1,
    orig.ident = Var2,
    std_residual = Freq
  ) %>%
  mutate(
    seurat_clusters = factor(seurat_clusters, levels = levels(metadata_df$seurat_clusters)),
    orig.ident = factor(orig.ident, levels = levels(metadata_df$orig.ident)),
    direction = case_when(
      std_residual > 2 ~ "enriched",
      std_residual < -2 ~ "depleted",
      TRUE ~ "not_strong"
    )
  )

residual_df %>%
  arrange(desc(abs(std_residual)))

cluster7_bias <- residual_df %>%
  filter(seurat_clusters == "7") %>%
  arrange(desc(std_residual))

cluster7_bias

cluster7_counts <- cluster_counts %>%
  filter(seurat_clusters == "7") %>%
  complete(orig.ident, fill = list(n_cells = 0))

chisq_cluster7 <- chisq.test(cluster7_counts$n_cells)

chisq_cluster7

cluster_bias_tests <- cluster_counts %>%
  complete(seurat_clusters, orig.ident, fill = list(n_cells = 0)) %>%
  group_by(seurat_clusters) %>%
  summarise(
    p_value = chisq.test(n_cells)$p.value,
    .groups = "drop"
  ) %>%
  mutate(
    p_adj = p.adjust(p_value, method = "BH"),
    biased = p_adj < 0.05
  ) %>%
  arrange(p_adj)

cluster_bias_tests

cluster_proportions <- cluster_counts %>%
  group_by(seurat_clusters) %>%
  mutate(
    total_cells_cluster = sum(n_cells),
    percent = 100 * n_cells / total_cells_cluster
  ) %>%
  ungroup() %>%
  left_join(
    residual_df,
    by = c("seurat_clusters", "orig.ident")
  ) %>%
  mutate(
    label = ifelse(abs(std_residual) > 2, "*", "")
  )

p_props <- ggplot(
  cluster_proportions,
  aes(
    x = seurat_clusters,
    y = percent,
    fill = orig.ident
  )
) +
  geom_col(color = "black", linewidth = 0.2) +
  theme_bw(base_size = 12) +
  scale_fill_npg() +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  labs(
    x = "Cluster",
    y = "Cells (%)",
    fill = "Sample",
    title = "Sample composition within each cluster",
    subtitle = paste0(
      "Global chi-square p-value < 2.2e-16",
      "; Cramér's V = ",
      signif(cramer_v, 3)
    )
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 11)
  )

p_props

ggsave(
  filename = "output/figures/sample_composition_per_cluster_percent.png",
  plot = p_props,
  width = 8,
  height = 6,
  dpi = 600
)

```


# Section 8. Save C1-Core Object

```{r}
qs_save(
  object = css_combined,
  file = "output/objects/css_combined_processed.qs2",
  compress_level = 10,
  nthreads = 10
)

sessionInfo()
```


# RANDOM PLOTS
```{r}

# Plot main continuous scores on UMAP
score_umap_plot <- FeaturePlot_scCustom(
  css_combined,
  features = c("C1_Core", "CytoTRACE2","G2M.Score","S.Score"),
  reduction = "umap",
  na_cutoff = NA
) &
  scale_color_viridis_c(option = "magma") &
  labs(x = "UMAP1", y = "UMAP2") &
  umap_theme

score_umap_plot

ggsave(
  filename = "output/figures/score_umap_plot.png",
  plot = score_umap_plot,
  dpi = 600,
  width = 12,
  height = 8
)


DimPlot_scCustom( css_combined, reduction = "umap",group.by = "predicted.ann_finest_level")  +
  labs(x = "UMAP1", y = "UMAP2", title = "Annotation (HLCA)") +
  umap_theme

ggsave(
  filename = "output/figures/AnnotationLV1.png",
  dpi = 600,
  width = 12,
  height = 8
)

```