---
title: "Hallmark 50 pathway activity in MGI organoids"
subtitle: "Single-sample gene-set scoring, pathway heatmaps, and paired TME-addition tests"
date: today
date-format: "DD MM YYYY"

format:
  html:
    page-layout: full

    toc: true
    toc-title: "Notebook sections"
    toc-location: left
    toc-depth: 4

    number-sections: true
    number-depth: 4

    theme:
      light: lux
      dark: darkly

    title-block-banner: true

    code-fold: true
    code-summary: "Show analysis code"
    code-tools: true
    code-copy: true
    code-overflow: wrap

    df-print: paged

    smooth-scroll: true
    anchor-sections: true
    link-external-newwindow: true

    embed-resources: true
    self-contained-math: true

    fig-align: center
    fig-width: 12
    fig-height: 8
    fig-dpi: 600

    callout-appearance: simple

execute:
  echo: false
  warning: false
  message: false
  error: false
  freeze: false

editor: source
---

## Analysis Overview

This notebook scores the 50 MSigDB Hallmark gene sets in the MGI ovarian organoid expression matrix and asks whether pathway activity changes with TME addition within IEO grade groups. The workflow is intentionally explicit: load the expression matrix and metadata, check which Hallmark genes are measurable, compute single-sample pathway scores, visualize score patterns, and test paired TME-addition differences.

The **Hallmark 50** collection is a curated MSigDB resource containing 50 gene sets that represent broad, recurrent biological programs. Examples include proliferation, hypoxia, interferon response, epithelial-mesenchymal transition, inflammatory signaling, and apoptosis. Hallmark gene sets are useful here because they compress gene-level expression into interpretable pathway-level activity scores.

**ssGSEA** means single-sample Gene Set Enrichment Analysis. It estimates gene-set enrichment separately for each sample, so every sample receives one score per gene set before any group comparison is performed. This is useful when the goal is to compare pathway activity across samples rather than only report group-level enrichment.

**singscore** is another single-sample gene-set scoring method. It ranks genes inside each sample and asks whether genes from a signature tend to sit high or low in that ranked expression profile. In this notebook, singscore provides one pathway score per Hallmark signature and sample.

A **Z-score** is a standardized value: the mean is subtracted and the result is divided by the standard deviation. Here, Z-scores are used for heatmap visualization only. They put signatures with different raw score ranges on a common color scale, making relative high and low activity easier to compare across samples.


# Setup

## Libraries and inputs

## 1. Inputs and Setup

This section loads the packages, metadata, expression matrix, and Hallmark gene sets used by the analysis. The expression matrix is kept at the gene level and the metadata provides the sample-level variables used later for grouping and paired comparisons.

```{r}
#| label: setup-libraries-and-inputs

library(tidyverse)
library(patchwork)
library(vroom)
library(msigdbr)
library(BiocParallel)
library(qs2)
library(future)

plan(multicore, workers = 5)

metadata <- vroom("input/metadata_clean.tsv", show_col_types = F) |>
  dplyr::filter(use == TRUE)

rna_logcpm <- vroom("input/MGI_logCPM_dedup_nonzero.tsv") |>
  column_to_rownames("gene")


hallmark_signatures <- msigdbr(species = "human", collection = "H") |>
  dplyr::distinct(gs_name, gene_symbol) |>
  dplyr::group_by(gs_name) |>
  dplyr::summarise(
    genes = list(gene_symbol),
    .groups = "drop"
  ) |>
  tibble::deframe()

names(hallmark_signatures) <- stringr::str_remove(
  names(hallmark_signatures),
  "^HALLMARK_"
)

hallmark_signatures

```

# Check signature gene coverage

## 2. Hallmark Gene-Set Coverage

Before scoring pathways, this section checks how many genes from each Hallmark signature are present in the expression matrix. This matters because a score is only interpretable when enough signature genes are actually measured in the dataset.

```{r}

# ============================================================
# Check Hallmark gene-set coverage in the expression matrix
# ============================================================

expression_genes <- rownames(rna_logcpm)

hallmark_gene_coverage <- purrr::imap_dfr(
  hallmark_signatures,
  \(signature_genes, signature_name) {

    found_genes <- intersect(signature_genes, expression_genes)
    missing_genes <- setdiff(signature_genes, expression_genes)

    tibble::tibble(
      signature = signature_name,
      requested = length(signature_genes),
      found = length(found_genes),
      missing_n = length(missing_genes),
      coverage = length(found_genes) / length(signature_genes),
      missing_genes = paste(missing_genes, collapse = ";")
    )
  }
) |>
  dplyr::arrange(coverage, signature)

hallmark_gene_coverage

write_tsv(
  hallmark_gene_coverage,
  "output/tables/hallmark_gene_coverage.tsv"
)

hallmark_gene_coverage |>
  dplyr::filter(coverage < 0.8)

```

## 3. Coverage Visualization

This section plots the Hallmark gene-set coverage results. The plot is a quick quality-control view: signatures with poor coverage would be weaker candidates for downstream interpretation.

```{r}

# ============================================================
# Signature gene coverage
# ============================================================

coverage_plot <- hallmark_gene_coverage |>
  mutate(
    signature = stringr::str_remove(signature, "^HALLMARK_"),
    signature = stringr::str_replace_all(signature, "_", " "),
    signature = forcats::fct_reorder(signature, coverage)
  ) |>
  ggplot(aes(x = coverage, y = signature)) +
  geom_col(width = 0.75) +
  geom_text(
    aes(label = paste0(found, "/", requested)),
    hjust = -0.1,
    size = 3
  ) +
  scale_x_continuous(
    labels = scales::label_percent(accuracy = 1),
    limits = c(0, 1.12),
    expand = expansion(mult = c(0, 0))
  ) +
  labs(
    x = "Genes detected in expression matrix",
    y = NULL,
    title = "Hallmark signature gene coverage"
  ) +
  theme_bw(base_size = 10)

coverage_plot

```

# Retain genes found in the expression matrix

## 4. Keep Measured Hallmark Genes

This section filters each Hallmark gene set to the genes found in the expression matrix. The scoring methods should only receive genes that are actually present, otherwise missing genes would add noise or generate avoidable warnings.

```{r}

# ============================================================
# Keep only genes present in rna_logcpm
# ============================================================

hallmark_signatures_found <- purrr::map(
  hallmark_signatures,
  \(signature_genes) {
    intersect(signature_genes, expression_genes)
  }
)

lengths(hallmark_signatures_found)

```

# Score all Hallmark signatures with singscore
## Rank expression data once

## 5. Prepare Expression Data for Scoring

This section prepares the expression matrix for single-sample pathway scoring. The matrix is already logCPM, so the goal here is not normalization, but making sure the object has the expected orientation and gene identifiers for the scoring functions.

```{r}

rank_data <- singscore::rankGenes(rna_logcpm)

```

## Score every signature:

## 6. ssGSEA Hallmark Scores

This section computes ssGSEA scores for the Hallmark signatures. ssGSEA produces a pathway activity score for every sample and every Hallmark gene set, allowing pathway-level comparisons across the organoid samples.

```{r}

# ============================================================
# Score all Hallmark signatures
# ============================================================

hallmark_singscore_long <- purrr::imap_dfr(
  hallmark_signatures_found,
  \(signature_genes, signature_name) {

    score_result <- singscore::simpleScore(
      rankData = rank_data,
      upSet = signature_genes
    )

    score_result |>
      as.data.frame() |>
      tibble::rownames_to_column("sample") |>
      tibble::as_tibble() |>
      dplyr::transmute(
        sample,
        signature = signature_name,
        score = TotalScore
      )
  }
)

hallmark_singscore_long

hallmark_singscore <- hallmark_singscore_long |>
  tidyr::pivot_wider(
    names_from = signature,
    values_from = score
  ) |>
  dplyr::arrange(match(sample, colnames(rna_logcpm)))

hallmark_singscore

write_tsv(
  hallmark_singscore,
  "output/tables/hallmark_singscore_scores.tsv"
)

```


## 7. Hallmark Score Heatmap

This section visualizes Hallmark activity across samples. Scores are transformed to row-wise Z-scores for display so that each pathway is centered and scaled across samples; this highlights relative high and low activity patterns without changing the original score table used for statistical testing.

```{r}

# ============================================================
# Hallmark singscore heatmap
# ============================================================

library(ComplexHeatmap)
library(circlize)
library(grid)

# Match metadata to scored samples
heatmap_metadata <- metadata |>
  filter(sample %in% hallmark_singscore$sample) |>
  distinct(sample, .keep_all = TRUE) |>
  arrange(match(sample, hallmark_singscore$sample))

stopifnot(
  identical(
    heatmap_metadata$sample,
    hallmark_singscore$sample
  )
)

# Signature score matrix: signatures × samples
score_mat <- hallmark_singscore |>
  column_to_rownames("sample") |>
  as.matrix() |>
  t()

# Remove HALLMARK_ prefix for cleaner row labels
rownames(score_mat) <- stringr::str_remove(
  rownames(score_mat),
  "^HALLMARK_"
)

# Row-wise z-score for visualization
score_mat_z <- t(scale(t(score_mat)))

# Remove signatures with zero variance, if any
score_mat_z <- score_mat_z[
  apply(score_mat_z, 1, \(x) all(is.finite(x))),
  ,
  drop = FALSE
]

# Column annotations
column_ha <- HeatmapAnnotation(
  `TME addition` = heatmap_metadata$tme_addition,
  `IEO grade` = heatmap_metadata$ieo_grade,
  col = list(
    `TME addition` = c(
      "Yes" = "#D55E00",
      "No" = "#0072B2"
    ),
    `IEO grade` = c(
      "High" = "#B2182B",
      "Low" = "#2166AC",
      "Unknown" = "#BDBDBD"
    )
  ),
  annotation_name_side = "left",
  na_col = "#E5E5E5"
)

# Symmetric colour scale around zero
z_limit <- max(abs(score_mat_z), na.rm = TRUE)

score_col_fun <- circlize::colorRamp2(
  c(-z_limit, 0, z_limit),
  c("#2166AC", "white", "#B2182B")
)

hallmark_heatmap <- Heatmap(
  score_mat_z,
  name = "Row Z-score",
  col = score_col_fun,

  top_annotation = column_ha,

  cluster_rows = TRUE,
  cluster_columns =TRUE,

  show_row_names = TRUE,
  show_column_names = TRUE,

  row_names_gp = grid::gpar(fontsize = 7),
  column_names_gp = grid::gpar(fontsize = 8),
  column_names_rot = 45,

  column_title = "Hallmark signature activity",
  row_title = "Hallmark signatures",

  border = TRUE,
  use_raster = TRUE,
  raster_quality = 3,

  heatmap_legend_param = list(
    title = "Row\nZ-score",
    at = c(-z_limit, 0, z_limit),
    labels = round(c(-z_limit, 0, z_limit), 1)
  )
)

draw(
  hallmark_heatmap,
  heatmap_legend_side = "right",
  annotation_legend_side = "right",
  merge_legends = TRUE
)

png("output/figures/hallmark_heatmap_50.png", width = 2200, height = 1600, res = 200)
draw(
  hallmark_heatmap,
  heatmap_legend_side = "right",
  annotation_legend_side = "right",
  merge_legends = TRUE
)
dev.off()

```


## 8. singscore Hallmark Scores

This section computes singscore values for the same Hallmark signatures. singscore is rank-based within each sample, so it is robust to differences in absolute expression scale and gives another single-sample view of pathway activity.

```{r}

# ============================================================
# Hallmark singscore heatmaps by grade
# Columns ordered by TME addition; no column clustering
# ============================================================

library(ComplexHeatmap)
library(circlize)
library(grid)
library(dplyr)
library(tibble)
library(stringr)

dir.create(
  "output/figures",
  recursive = TRUE,
  showWarnings = FALSE
)

# ------------------------------------------------------------
# Function to generate one grade-specific heatmap
# ------------------------------------------------------------

make_grade_hallmark_heatmap <- function(
  grade,
  output_file,
  width = 2200,
  height = 1600,
  res = 200
) {

  # Select metadata for the requested grade and order by TME
  grade_metadata <- metadata |>
    filter(
      ieo_grade == grade,
      sample %in% hallmark_singscore$sample
    ) |>
    distinct(sample, .keep_all = TRUE) |>
    mutate(
      tme_addition = factor(
        tme_addition,
        levels = c("No", "Yes")
      )
    ) |>
    arrange(
      tme_addition,
      generic_name,
      sample
    )

  if (nrow(grade_metadata) == 0) {
    stop("No samples found for grade: ", grade)
  }

  # Extract scores and enforce metadata-defined sample order
  grade_scores <- hallmark_singscore |>
    filter(sample %in% grade_metadata$sample) |>
    arrange(match(sample, grade_metadata$sample))

  stopifnot(
    identical(
      grade_scores$sample,
      grade_metadata$sample
    )
  )

  # Signature × sample matrix
  score_mat <- grade_scores |>
    column_to_rownames("sample") |>
    as.matrix() |>
    t()

  rownames(score_mat) <- rownames(score_mat) |>
    str_remove("^HALLMARK_") |>
    str_replace_all("_", " ")

  # Row-wise scaling within the selected grade
  score_mat_z <- t(scale(t(score_mat)))

  # Remove signatures with zero variance within this grade
  keep_signatures <- apply(
    score_mat_z,
    1,
    \(x) all(is.finite(x))
  )

  score_mat_z <- score_mat_z[
    keep_signatures,
    ,
    drop = FALSE
  ]

  if (nrow(score_mat_z) == 0) {
    stop(
      "No signatures with non-zero variance for grade: ",
      grade
    )
  }

  # Top annotation
  column_ha <- HeatmapAnnotation(
    `TME addition` = grade_metadata$tme_addition,
    col = list(
      `TME addition` = c(
        "No" = "#0072B2",
        "Yes" = "#D55E00"
      )
    ),
    annotation_name_side = "left",
    na_col = "#E5E5E5"
  )

  # Symmetric colour scale
  z_limit <- max(abs(score_mat_z), na.rm = TRUE)

  score_col_fun <- circlize::colorRamp2(
    c(-z_limit, 0, z_limit),
    c("#2166AC", "white", "#B2182B")
  )


  hallmark_heatmap <- Heatmap(
    score_mat_z,
    name = "Row Z-score",
    col = score_col_fun,

    top_annotation = column_ha,

    cluster_rows = TRUE,
    cluster_columns = FALSE,

    # Visually separate No and Yes samples
    #column_split = grade_metadata$tme_addition,
    #cluster_column_slices = FALSE,
    #column_gap = grid::unit(3, "mm"),

    show_row_names = TRUE,
    show_column_names = TRUE,

    row_names_gp = grid::gpar(fontsize = 7),
    column_names_gp = grid::gpar(fontsize = 8),
    column_names_rot = 90,

    column_title = paste0(
      grade,
      "-grade Hallmark signature activity"
    ),
    row_title = "Hallmark signatures",

    border = TRUE,
    use_raster = TRUE,
    raster_quality = 3,

    heatmap_legend_param = list(
      title = "Row\nZ-score",
      at = c(-z_limit, 0, z_limit),
      labels = round(c(-z_limit, 0, z_limit), 1)
    )
  )

  # Display
  draw(
    hallmark_heatmap,
    heatmap_legend_side = "right",
    annotation_legend_side = "right",
    merge_legends = TRUE
  )

  # Save
  png(
    output_file,
    width = width,
    height = height,
    res = res
  )

  draw(
    hallmark_heatmap,
    heatmap_legend_side = "right",
    annotation_legend_side = "right",
    merge_legends = TRUE
  )

  dev.off()

  invisible(hallmark_heatmap)
}


# ============================================================
# High-grade heatmap
# ============================================================

hallmark_heatmap_high <- make_grade_hallmark_heatmap(
  grade = "High",
  output_file = "output/figures/hallmark_heatmap_high_grade.png"
)


# ============================================================
# Low-grade heatmap
# ============================================================

hallmark_heatmap_low <- make_grade_hallmark_heatmap(
  grade = "Low",
  output_file = "output/figures/hallmark_heatmap_low_grade.png"
)

```


## 9. Paired TME-Addition Comparisons

This section compares matched biological samples with and without TME addition within each IEO grade group. The paired structure is important: each difference is calculated within the same biological sample identity, reducing confounding from baseline sample-to-sample variability.

```{r}

library(tidyverse)
library(rstatix)

# ============================================================
# 1. Prepare Hallmark scores and metadata
# ============================================================

meta_hallmark <- hallmark_singscore |>
  left_join(
    metadata,
    by = "sample"
  )

signature_cols <- intersect(
  names(hallmark_signatures),
  names(meta_hallmark)
)

stopifnot(length(signature_cols) > 0)


# ============================================================
# 2. Convert Hallmark scores to long format
# ============================================================

hallmark_long <- meta_hallmark |>
  select(
    sample,
    generic_name,
    tme_addition,
    ieo_grade,
    all_of(signature_cols)
  ) |>
  pivot_longer(
    cols = all_of(signature_cols),
    names_to = "signature",
    values_to = "score"
  ) |>
  filter(
    !is.na(score),
    !is.na(generic_name),
    tme_addition %in% c("Yes", "No"),
    ieo_grade %in% c("Low", "High")
  ) |>
  mutate(
    tme_addition = factor(
      tme_addition,
      levels = c("Yes", "No")
    ),
    ieo_grade = factor(
      ieo_grade,
      levels = c("Low", "High")
    )
  )


# ============================================================
# 3. Build explicit paired differences
# ============================================================

hallmark_paired <- hallmark_long |>
  select(
    ieo_grade,
    signature,
    generic_name,
    tme_addition,
    score
  ) |>
  distinct() |>
  group_by(
    ieo_grade,
    signature,
    generic_name
  ) |>
  filter(
    all(c("Yes", "No") %in% tme_addition)
  ) |>
  ungroup() |>
  arrange(
    ieo_grade,
    signature,
    generic_name,
    tme_addition
  )

hallmark_differences <- hallmark_paired |>
  pivot_wider(
    names_from = tme_addition,
    values_from = score
  ) |>
  mutate(
    difference = Yes - No
  )

# Number of complete biological pairs in each grade
hallmark_pair_counts <- hallmark_differences |>
  distinct(
    ieo_grade,
    generic_name
  ) |>
  count(
    ieo_grade,
    name = "n_pairs"
  )

hallmark_pair_counts


# ============================================================
# 4. Normality of paired differences
# ============================================================
hallmark_normality <- hallmark_paired |>
  select(
    ieo_grade,
    signature,
    generic_name,
    tme_addition,
    score
  ) |>
  pivot_wider(
    names_from = tme_addition,
    values_from = score
  ) |>
  mutate(
    difference = Yes - No
  ) |>
  group_by(
    ieo_grade,
    signature
  ) |>
  shapiro_test(difference) |>
  ungroup() |>
  rename(
    shapiro_w = statistic,
    shapiro_p = p
  ) |>
  mutate(
    normality = if_else(
      shapiro_p > 0.05,
      "Approximately normal",
      "Non-normal"
    )
  )

hallmark_normality

table(hallmark_normality$normality)


```


## 10. Statistical Test Selection and Result Tables

This section checks whether paired differences are approximately normal and then reports paired t-test or Wilcoxon results as appropriate. The goal is to keep the statistical choice transparent while preserving the complete paired-score table for interpretation.

```{r}

# ============================================================
# Run both paired tests
# ============================================================

hallmark_ttest <- hallmark_paired |>
  group_by(ieo_grade, signature) |>
  t_test(
    score ~ tme_addition,
    paired = TRUE,
    detailed = TRUE
  ) |>
  ungroup() |>
  select(
    ieo_grade,
    signature,
    group1,
    group2,
    estimate,
    conf.low,
    conf.high,
    statistic,
    df,
    p
  ) |>
  rename(
    t_estimate = estimate,
    t_conf_low = conf.low,
    t_conf_high = conf.high,
    t_statistic = statistic,
    t_df = df,
    t_p = p
  )


hallmark_wilcox <- hallmark_paired |>
  group_by(ieo_grade, signature) |>
  wilcox_test(
    score ~ tme_addition,
    paired = TRUE,
    exact = FALSE,
    detailed = TRUE
  ) |>
  ungroup() |>
  select(
    ieo_grade,
    signature,
    statistic,
    p
  ) |>
  rename(
    wilcox_statistic = statistic,
    wilcox_p = p
  )

# ============================================================
# Select t-test or Wilcoxon according to normality
# ============================================================

hallmark_selected_tests <- hallmark_normality |>
  left_join(
    hallmark_ttest,
    by = c("ieo_grade", "signature")
  ) |>
  left_join(
    hallmark_wilcox,
    by = c("ieo_grade", "signature")
  ) |>
  mutate(
    test = if_else(
      normality == "Approximately normal",
      "Paired t-test",
      "Paired Wilcoxon"
    ),
    statistic = if_else(
      test == "Paired t-test",
      t_statistic,
      wilcox_statistic
    ),
    p = if_else(
      test == "Paired t-test",
      t_p,
      wilcox_p
    )
  ) |>
  group_by(ieo_grade) |>
  adjust_pvalue(
    p.col = "p",
    output.col = "p.adj",
    method = "BH"
  ) |>
  add_significance(
    p.col = "p.adj"
  ) |>
  ungroup() |>
  arrange(
    ieo_grade,
    p.adj
  )

hallmark_summary <- hallmark_paired |>
  select(
    ieo_grade,
    signature,
    generic_name,
    tme_addition,
    score
  ) |>
  pivot_wider(
    names_from = tme_addition,
    values_from = score
  ) |>
  mutate(
    difference = Yes - No
  ) |>
  group_by(ieo_grade, signature) |>
  summarise(
    n_pairs = n(),
    mean_no = mean(No),
    mean_yes = mean(Yes),
    median_no = median(No),
    median_yes = median(Yes),
    mean_difference = mean(difference),
    median_difference = median(difference),
    .groups = "drop"
  ) |>
  mutate(
    direction = case_when(
      mean_difference > 0 ~ "Higher with TME",
      mean_difference < 0 ~ "Lower with TME",
      TRUE ~ "No change"
    )
  )

hallmark_results <- hallmark_selected_tests |>
  left_join(
    hallmark_summary,
    by = c("ieo_grade", "signature")
  ) |>
  select(
    ieo_grade,
    signature,
    n_pairs,
    normality,
    shapiro_w,
    shapiro_p,
    test,
    group1,
    group2,
    mean_no,
    mean_yes,
    mean_difference,
    median_difference,
    direction,
    statistic,
    p,
    p.adj,
    p.adj.signif
  ) |>
  arrange(
    ieo_grade,
    p.adj
  )

hallmark_results |>
  filter(p.adj < 0.05)

high_hallmark_results <- hallmark_results |>
  filter(p.adj < 0.05, ieo_grade == "High")
high_hallmark_results

low_hallmark_results <- hallmark_results |>
  filter(p.adj < 0.05, ieo_grade == "Low")
low_hallmark_results

```

## 11. Final Plots and Exported Outputs

This section produces the final pathway-level summaries, figures, and result tables. These outputs are the files to use for review, reporting, and downstream interpretation of Hallmark activity changes.

```{r}


library(tidyverse)
library(ComplexHeatmap)
library(circlize)
library(grid)

make_grade_hallmark_heatmap <- function(
  grade,
  hallmark_results,
  output_file
) {

  # Colour-blind-friendly annotation palettes
  effect_colors <- c(
    "Higher with TME" = "#ff6f00",
    "Lower with TME"  = "#6200b2",
    "Not significant" = "#D3D3D3"
  )

  tme_colors <- c(
    "No"  = "#676767",
    "Yes" = "#009E73"
  )

  # Prepare and order sample metadata
  grade_metadata <- metadata |>
    filter(
      ieo_grade == grade,
      sample %in% hallmark_singscore$sample
    ) |>
    distinct(sample, .keep_all = TRUE) |>
    mutate(
      tme_addition = factor(
        tme_addition,
        levels = c("No", "Yes")
      )
    ) |>
    arrange(tme_addition, generic_name, sample)

  # Construct the signature-by-sample matrix
  score_mat <- hallmark_singscore |>
    filter(sample %in% grade_metadata$sample) |>
    arrange(match(sample, grade_metadata$sample)) |>
    column_to_rownames("sample") |>
    as.matrix() |>
    t()

  # Remove the Hallmark prefix
  rownames(score_mat) <- str_remove(
    rownames(score_mat),
    "^HALLMARK_"
  )

  # Standardize each signature across samples
  score_mat_z <- t(scale(t(score_mat)))

  # Remove signatures with undefined row Z-scores
  score_mat_z <- score_mat_z[
    apply(score_mat_z, 1, \(x) all(is.finite(x))),
    ,
    drop = FALSE
  ]

  # Format Hallmark names for display
  hallmark_labels <- rownames(score_mat_z) |>
    str_replace_all("_", " ")

  # Align statistical results to heatmap rows
  row_statistics <- tibble(
    signature = rownames(score_mat_z)
  ) |>
    left_join(
      hallmark_results |>
        filter(ieo_grade == grade) |>
        mutate(
          signature = str_remove(
            signature,
            "^HALLMARK_"
          )
        ) |>
        select(
          signature,
          mean_difference,
          p.adj
        ),
      by = "signature"
    ) |>
    mutate(
      `TME effect` = case_when(
        !is.na(p.adj) &
          p.adj < 0.05 &
          mean_difference > 0 ~ "Higher with TME",

        !is.na(p.adj) &
          p.adj < 0.05 &
          mean_difference < 0 ~ "Lower with TME",

        TRUE ~ "Not significant"
      ),
      `TME effect` = factor(
        `TME effect`,
        levels = c(
          "Higher with TME",
          "Lower with TME",
          "Not significant"
        )
      )
    )

  # Row annotation indicating direction of the TME effect
  effect_ha <- rowAnnotation(
    `TME effect` = row_statistics$`TME effect`,
    col = list(
      `TME effect` = effect_colors
    ),
    simple_anno_size = unit(4, "mm"),
    annotation_legend_param = list(
      `TME effect` = list(
        at = c(
          "Higher with TME",
          "Lower with TME",
          "Not significant"
        ),
        ncol = 1
      )
    )
  )

  # Column annotation indicating TME addition status
  top_ha <- HeatmapAnnotation(
    `TME addition` = grade_metadata$tme_addition,
    col = list(
      `TME addition` = tme_colors
    ),
    simple_anno_size = unit(4, "mm"),
    annotation_legend_param = list(
      `TME addition` = list(
        at = c("No", "Yes"),
        ncol = 1
      )
    )
  )

  # Use a symmetric colour scale around zero
  z_limit <- max(abs(score_mat_z), na.rm = TRUE)

  ht <- Heatmap(
    score_mat_z,
    name = "Row Z-score",
    col = colorRamp2(
      c(-z_limit, 0, z_limit),
      c("#3B75AF", "#F7F7F7", "#C94C4C")
    ),
    top_annotation = top_ha,
    right_annotation = effect_ha,
    cluster_rows = TRUE,
    cluster_columns = FALSE,
    row_labels = hallmark_labels,
    row_names_side = "right",
    row_names_gp = gpar(fontsize = 7),
    column_names_rot = 90,
    column_names_gp = gpar(fontsize = 7),
    column_title = paste0(
      grade,
      "-grade Hallmark signature activity"
    ),
    border = TRUE,
    heatmap_legend_param = list(
      color_bar = "continuous",
      legend_direction = "vertical"
    )
  )

  # Display in the active graphics device
  draw(
    ht,
    heatmap_legend_side = "right",
    annotation_legend_side = "right",
    merge_legends = TRUE,
    legend_grouping = "original"
  )

  # Save the heatmap
  png(
    filename = output_file,
    width = 2200,
    height = 1600,
    res = 200
  )

  draw(
    ht,
    heatmap_legend_side = "right",
    annotation_legend_side = "right",
    merge_legends = TRUE,
    legend_grouping = "original"
  )

  dev.off()

  invisible(ht)
}

make_grade_hallmark_heatmap(
  grade = "High",
  hallmark_results = hallmark_results,
  output_file = "output/figures/hallmark_heatmap_high_grade.png"
)

make_grade_hallmark_heatmap(
  grade = "Low",
  hallmark_results = hallmark_results,
  output_file = "output/figures/hallmark_heatmap_low_grade.png"
)

```
