---
title: "Create a Signac scATAC Object"
format:
  html:
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
---

> **Status:** `draft`<br>
> **Template class:** `SOURCE-BACKED WORKFLOW`

## Purpose

Create one Seurat object containing a Signac `ChromatinAssay` from a peak-by-
cell count matrix, aligned cell metadata, a tabix-indexed fragments file, and
compatible genome annotation.

## Inputs

This template deliberately supports one input contract: a qs2-saved R matrix
with peaks in rows and cells in columns, plus a metadata TSV with a `cell_id`
column. It does not try to discover every 10x, Cell Ranger, or count-file
layout.

```{r}
suppressPackageStartupMessages({
  library(Seurat)
  library(Signac)
  library(qs2)
  library(readr)
  library(tibble)
})

counts_path <- "input/peak_counts.qs2"
metadata_path <- "input/cell_metadata.tsv"
fragments_path <- "input/fragments.tsv.gz"
annotation_path <- "input/gene_annotation.qs2"
output_object <- "output/scatac_created.qs2"
alignment_table <- "output/scatac_cell_alignment.tsv"

assay_name <- "ATAC"
genome_build <- "hg38"
peak_separator <- c(":", "-")
min_cells <- 1L
min_features <- 1L
```

The default `hg38` is a valid example, not a universal choice. Peak
coordinates, the fragment file, gene annotations, sequence genome, and later
motif resources must use compatible builds and chromosome naming.

## Analysis

```{r}
counts <- qs2::qs_read(file = counts_path)
metadata <- readr::read_tsv(metadata_path, show_col_types = FALSE)
annotation <- qs2::qs_read(file = annotation_path)

stopifnot(is.matrix(counts) || inherits(counts, "Matrix"))
stopifnot(!is.null(rownames(counts)), !is.null(colnames(counts)))
stopifnot(!anyDuplicated(rownames(counts)), !anyDuplicated(colnames(counts)))
stopifnot(all(counts >= 0), all(is.finite(counts)))
stopifnot("cell_id" %in% colnames(metadata), !anyDuplicated(metadata$cell_id))
stopifnot(file.exists(fragments_path), file.exists(paste0(fragments_path, ".tbi")))

shared_cells <- intersect(colnames(counts), metadata$cell_id)
stopifnot(length(shared_cells) > 0)

alignment <- tibble::tibble(
  cell_id = c(setdiff(colnames(counts), shared_cells),
              setdiff(metadata$cell_id, shared_cells)),
  source = c(rep("count_matrix_only", length(setdiff(colnames(counts), shared_cells))),
             rep("metadata_only", length(setdiff(metadata$cell_id, shared_cells))))
)
readr::write_tsv(alignment, alignment_table)

counts <- counts[, shared_cells, drop = FALSE]
metadata <- metadata[match(shared_cells, metadata$cell_id), , drop = FALSE]
rownames(metadata) <- metadata$cell_id

fragment_object <- Signac::CreateFragmentObject(
  path = fragments_path,
  cells = shared_cells,
  validate.fragments = TRUE
)

chromatin_assay <- Signac::CreateChromatinAssay(
  counts = counts,
  sep = peak_separator,
  genome = genome_build,
  annotation = annotation,
  fragments = fragment_object,
  min.cells = min_cells,
  min.features = min_features,
  validate.fragments = TRUE
)

object <- Seurat::CreateSeuratObject(
  counts = chromatin_assay,
  assay = assay_name,
  meta.data = metadata,
  min.features = min_features
)
DefaultAssay(object) <- assay_name

stopifnot(identical(colnames(object), shared_cells))
stopifnot(inherits(object[[assay_name]], "ChromatinAssay"))
stopifnot(length(Signac::Fragments(object[[assay_name]])) > 0)

readr::write_tsv(
  tibble::tibble(
    cells = ncol(object),
    peaks = nrow(object[[assay_name]]),
    genome_build = genome_build,
    fragments = fragments_path
  ),
  "output/scatac_created_summary.tsv"
)
qs2::qs_save(object, file = output_object)
```

## Method notes

The matrix is explicitly feature-by-cell: peak identifiers are rows and cell
barcodes are columns. Metadata row names must match the retained cell names
exactly. The intersection and alignment TSV make dropped barcodes visible;
unrelated barcodes are not silently unioned. A fragment file should be bgzip
compressed and tabix indexed, with fragment barcodes matching the retained
cells. The source workflow used `validate.fragments = FALSE` for a project
input; this reusable template keeps validation enabled so an alignment problem
is visible before downstream QC.
