---
title: "scATAC TF-IDF, LSI, and Clustering"
format:
  html:
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
---

> **Status:** `draft`<br>
> **Template class:** `SOURCE-BACKED WORKFLOW`

## Purpose

Run the source-backed unintegrated Signac representation and clustering path:

```text
filtered Signac object → TF-IDF → top features → SVD → diagnostics
                         → neighbors → clusters → UMAP
```

TF-IDF/LSI is a representation and dimension-reduction workflow. It is not
batch correction.

## Parameters

```{r}
suppressPackageStartupMessages({
  library(Seurat)
  library(Signac)
  library(qs2)
  library(readr)
  library(tibble)
})

input_object <- "input/scatac_qc_filtered.qs2"
output_object <- "output/scatac_lsi_clusters.qs2"
lsi_table <- "output/scatac_lsi_embeddings.tsv"
cluster_table <- "output/scatac_cluster_counts.tsv"
lsi_figure <- "output/scatac_lsi_depth_correlation.pdf"

assay_name <- "ATAC"
min_feature_cutoff <- "q5"
n_svd <- 50L
lsi_dims <- 1:30
neighbors_k <- 20L
cluster_resolution <- 0.5
umap_seed <- 1L
```

## Analysis

```{r}
object <- qs2::qs_read(file = input_object)
stopifnot(inherits(object, "Seurat"), assay_name %in% Seurat::Assays(object))
DefaultAssay(object) <- assay_name

object <- Signac::RunTFIDF(object)
object <- Signac::FindTopFeatures(object, min.cutoff = min_feature_cutoff)
object <- Signac::RunSVD(object, n = n_svd)

pdf(lsi_figure, width = 7, height = 5)
print(Signac::DepthCor(object, reduction = "lsi"))
dev.off()

object <- Seurat::FindNeighbors(
  object, reduction = "lsi", dims = lsi_dims, k.param = neighbors_k,
  graph.name = c("atac_nn", "atac_snn")
)
object <- Seurat::FindClusters(
  object, graph.name = "atac_snn", resolution = cluster_resolution,
  random.seed = umap_seed
)
cluster_column <- paste0("atac_snn_res.", cluster_resolution)
object <- Seurat::RunUMAP(
  object, reduction = "lsi", dims = lsi_dims, seed.use = umap_seed,
  reduction.name = "umap_atac", reduction.key = "atacUMAP_"
)

lsi_embeddings <- Embeddings(object, "lsi") |>
  as.data.frame() |>
  tibble::rownames_to_column("cell_id")
readr::write_tsv(lsi_embeddings, lsi_table)
readr::write_tsv(
  tibble::tibble(cluster = object[[]][[cluster_column]]) |>
    dplyr::count(cluster, name = "n_cells"),
  cluster_table
)
qs2::qs_save(object, file = output_object)
```

## Diagnostics and method notes

The source used `min.cutoff = "q5"`; this is exposed rather than hidden. The
default dimension vector includes LSI1. Inspect its relationship with depth
and QC in `DepthCor` before deciding whether to use `2:30` or another range.
Discarding LSI1 automatically can remove a depth-related axis, but retaining
it blindly can also make depth dominate neighborhoods. That is a visible
representation decision, not an integration setting.
