---
title: "scATAC Motif Enrichment with FindMotifs"
format:
  html:
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
---

> **Status:** `draft`<br>
> **Template class:** `SOURCE-BACKED WORKFLOW`

## Purpose

Add JASPAR motif matches to a Signac assay and test enrichment in a selected
foreground peak set against an explicit background universe. This preserves the
real `002_ATAC_MOTIF.qmd` pattern without retaining project clusters or TF
interpretation.

## Inputs and parameters

```{r}
suppressPackageStartupMessages({
  library(Seurat)
  library(Signac)
  library(qs2)
  library(readr)
  library(tibble)
  library(TFBSTools)
  library(JASPAR2024)
  library(RSQLite)
  library(BSgenome.Hsapiens.UCSC.hg38)
})

input_object <- "input/scatac_lsi_clusters.qs2"
foreground_path <- "input/foreground_peaks.tsv"
background_path <- "input/background_peaks.tsv"
output_object <- "output/scatac_motifs.qs2"
motif_table_path <- "output/scatac_findmotifs.tsv"
motif_figure_path <- "output/scatac_motif_plot.pdf"

assay_name <- "ATAC"
peak_id_column <- "peak_id"
genome_build <- "hg38"
jaspar_database <- "JASPAR2024"
species_tax_id <- 9606L
top_motifs_to_plot <- 6L
stopifnot(genome_build == "hg38", jaspar_database == "JASPAR2024")
```

## Analysis

```{r}
object <- qs2::qs_read(file = input_object)
foreground <- readr::read_tsv(foreground_path, show_col_types = FALSE)
background <- readr::read_tsv(background_path, show_col_types = FALSE)
stopifnot(inherits(object, "Seurat"), assay_name %in% Assays(object))
stopifnot(all(c(peak_id_column) %in% colnames(foreground)))
stopifnot(all(c(peak_id_column) %in% colnames(background)))
stopifnot(!anyDuplicated(foreground[[peak_id_column]]))
stopifnot(!anyDuplicated(background[[peak_id_column]]))
stopifnot(all(foreground[[peak_id_column]] %in% rownames(object[[assay_name]])))
stopifnot(all(background[[peak_id_column]] %in% rownames(object[[assay_name]])))

jaspar_resource <- JASPAR2024::JASPAR2024()
jaspar_db <- RSQLite::dbConnect(RSQLite::SQLite(), JASPAR2024::db(jaspar_resource))
pfm <- TFBSTools::getMatrixSet(
  jaspar_db,
  opts = list(
    collection = "CORE",
    tax_group = "vertebrates",
    species = species_tax_id,
    all_versions = FALSE
  )
)
RSQLite::dbDisconnect(jaspar_db)

DefaultAssay(object) <- assay_name
object <- Signac::AddMotifs(
  object = object,
  genome = BSgenome.Hsapiens.UCSC.hg38,
  pfm = pfm,
  assay = assay_name
)

foreground_peaks <- foreground[[peak_id_column]]
background_peaks <- background[[peak_id_column]]
motif_results <- Signac::FindMotifs(
  object = object,
  features = foreground_peaks,
  background = background_peaks,
  assay = assay_name,
  p.adjust.method = "BH"
)

readr::write_tsv(
  tibble::rownames_to_column(as.data.frame(motif_results), "motif_id"),
  motif_table_path
)

pdf(motif_figure_path, width = 9, height = 7)
if (nrow(motif_results) > 0) {
  print(Signac::MotifPlot(object, motifs = head(rownames(motif_results), top_motifs_to_plot)))
}
dev.off()
qs2::qs_save(object, file = output_object)
```

## Method notes

The foreground definition is intentionally external and visible: it may be a
selected or differential peak set, but the selection rule belongs in the
upstream analysis. The background is also explicit rather than silently using
an arbitrary default. Peak coordinates, `genome_build`, the BSgenome sequence,
and JASPAR motif resources must agree. The source used JASPAR2024, CORE
vertebrate motifs, human taxon 9606, and current-version-only matrices.

`FindMotifs` is peak-set enrichment. chromVAR is a separate per-cell deviation
analysis; motif enrichment p-values are not chromVAR activity scores.
