---
title: "scATAC Quality Control"
format:
  html:
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
---

> **Status:** `draft`<br>
> **Template class:** `SOURCE-BACKED WORKFLOW`

## Purpose

Compute fragment-aware ATAC QC metrics, inspect them by sample, apply an
explicit baseline filter, and save the filtered object. The source workflow
combined Cell Ranger per-barcode metrics with Signac/fragtk metrics and kept
before/after accounting.

## Inputs and parameters

The metrics TSV should contain `cell_id`, `passed_filters`,
`peak_region_fragments`, and `blacklist_region_fragments`. The latter two are
used to calculate FRiP and blacklist ratio when those columns are not already
present.

```{r}
suppressPackageStartupMessages({
  library(Seurat)
  library(Signac)
  library(qs2)
  library(readr)
  library(dplyr)
  library(ggplot2)
})

input_object <- "input/scatac_created.qs2"
metrics_path <- "input/singlecell_metrics.tsv"
output_object <- "output/scatac_qc_filtered.qs2"
qc_table_path <- "output/scatac_qc_per_cell.tsv"
sample_summary_path <- "output/scatac_qc_by_sample.tsv"
figure_path <- "output/scatac_qc_distributions.pdf"

assay_name <- "ATAC"
sample_column <- "sample_id"

# These defaults mirror the active source workflow, but are not biological constants.
min_peak_counts <- 1000
min_passed_filters <- 3000
min_pct_reads_in_peaks <- 25
max_blacklist_ratio <- 0.05
max_nucleosome_signal <- 1.5
min_tss_enrichment <- 6
```

## Analysis

```{r}
object <- qs2::qs_read(file = input_object)
metrics <- readr::read_tsv(metrics_path, show_col_types = FALSE)
stopifnot(inherits(object, "Seurat"))
stopifnot(all(c(sample_column) %in% colnames(object[[]])))
peak_count_column <- paste0("nCount_", assay_name)
stopifnot(peak_count_column %in% colnames(object[[]]))
stopifnot(all(c("cell_id", "passed_filters", "peak_region_fragments",
                "blacklist_region_fragments") %in% colnames(metrics)))
stopifnot(!anyDuplicated(metrics$cell_id))
fragment_path <- Signac::Fragments(object[[assay_name]])[[1]]@path
stopifnot(file.exists(paste0(fragment_path, ".tbi")))

metrics <- metrics[match(colnames(object), metrics$cell_id), , drop = FALSE]
stopifnot(all(metrics$cell_id == colnames(object)))
metrics$pct_reads_in_peaks <- 100 * metrics$peak_region_fragments / metrics$passed_filters
metrics$blacklist_ratio <- metrics$blacklist_region_fragments / metrics$peak_region_fragments
rownames(metrics) <- metrics$cell_id

metric_columns <- setdiff(colnames(metrics), colnames(object[[]]))
object <- Seurat::AddMetaData(object, metadata = metrics[, metric_columns, drop = FALSE])
DefaultAssay(object) <- assay_name
object <- Signac::NucleosomeSignal(object, assay = assay_name)
object <- Signac::TSSEnrichment(object, assay = assay_name, fast = TRUE)

qc_table <- object[[]] |>
  tibble::rownames_to_column("cell_id") |>
  mutate(
    qc_pass =
      .data[[peak_count_column]] >= min_peak_counts &
      passed_filters >= min_passed_filters &
      pct_reads_in_peaks >= min_pct_reads_in_peaks &
      blacklist_ratio <= max_blacklist_ratio &
      nucleosome_signal <= max_nucleosome_signal &
      TSS.enrichment >= min_tss_enrichment
  )

before_by_sample <- qc_table |>
  count(.data[[sample_column]], name = "cells_before")
after_by_sample <- qc_table |>
  filter(qc_pass) |>
  count(.data[[sample_column]], name = "cells_after")
sample_summary <- full_join(before_by_sample, after_by_sample, by = sample_column) |>
  mutate(cells_after = coalesce(cells_after, 0L))

readr::write_tsv(qc_table, qc_table_path)
readr::write_tsv(sample_summary, sample_summary_path)

object$qc_pass <- qc_table$qc_pass[match(colnames(object), qc_table$cell_id)]
cells_before <- ncol(object)
keep_cells <- qc_table$cell_id[which(qc_table$qc_pass)]
object <- subset(object, cells = keep_cells)
cells_after <- ncol(object)

pdf(figure_path, width = 11, height = 8)
print(VlnPlot(object, features = c(peak_count_column,
                                  paste0("nFeature_", assay_name),
                                  "nucleosome_signal", "TSS.enrichment"),
              group.by = sample_column, pt.size = 0, ncol = 2))
dev.off()

readr::write_tsv(
  tibble::tibble(cells_before = cells_before, cells_after = cells_after,
                 peaks = nrow(object[[assay_name]])),
  "output/scatac_qc_before_after.tsv"
)
qs2::qs_save(object, file = output_object)
```

## Diagnostics and method notes

The source workflow used passed-filter fragments, fraction of reads in peaks,
blacklist ratio, peak counts, nucleosome signal, and TSS enrichment together.
The PDF and TSV outputs preserve distributions, per-cell flags, per-sample
counts, and before/after accounting. Missing or non-finite values should be
investigated rather than treated as high-quality cells.

::: {.callout-note}
## Workflow preference

I prefer inspecting QC distributions separately by sample and looking at
several metrics together. A pooled threshold can hide sample-specific behavior.
Fixed cutoffs are transparent and useful as a baseline, but they are not
universal constants; protocol, library depth, and biological context matter.
An unusual sample or cell is a reason to investigate, not an automatic deletion
rule. Doublet detection should be considered alongside, not hidden inside, QC.
:::

The source also contains distribution-aware/MAD ideas in other workflows, but
their exact multipliers are not made universal here. If a MAD policy is used,
make its direction, multiplier, and sample scope explicit before filtering.
