---
title: "Create a Seurat object"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
---

> Status: `draft`
>
> Template class: SOURCE-BACKED WORKFLOW

## Purpose

Create one RNA Seurat object from a feature-by-cell count matrix and cell metadata. The matrix orientation and cell identifiers are checked explicitly.

## Inputs

Set the paths to an R-serialized count matrix and a tab-separated metadata table. The metadata table must contain one row per cell.

```{r}
library(Seurat)
library(qs2)
library(readr)

counts_path <- "input/counts.qs2"
metadata_path <- "input/metadata.tsv"
cell_id_column <- "cell_id"
assay_name <- "RNA"
min_cells <- 3
min_features <- 200
mitochondrial_pattern <- "^MT-"
output_object <- "output/seurat_object.qs2"
output_metadata <- "output/seurat_metadata.tsv"
```

## Analysis

```{r}
counts <- qs2::qs_read(counts_path)
metadata <- read_tsv(metadata_path, show_col_types = FALSE)
metadata <- as.data.frame(metadata)

stopifnot(is.matrix(counts) || inherits(counts, "Matrix"))
stopifnot(!is.null(rownames(counts)), !is.null(colnames(counts)))
stopifnot(cell_id_column %in% colnames(metadata))
stopifnot(!anyNA(metadata[[cell_id_column]]))
stopifnot(!anyDuplicated(colnames(counts)))
stopifnot(!anyDuplicated(metadata[[cell_id_column]]))

rownames(metadata) <- metadata[[cell_id_column]]
metadata[[cell_id_column]] <- NULL

stopifnot(setequal(colnames(counts), rownames(metadata)))
metadata <- metadata[colnames(counts), , drop = FALSE]

object <- CreateSeuratObject(
  counts = counts,
  assay = assay_name,
  meta.data = metadata,
  min.cells = min_cells,
  min.features = min_features
)

object[["percent.mt"]] <- PercentageFeatureSet(
  object,
  assay = assay_name,
  pattern = mitochondrial_pattern
)
```

## Diagnostics

```{r}
print(object)
feature_column <- paste0("nFeature_", assay_name)
print(summary(object[[]][[feature_column]]))
print(summary(object[[]][["percent.mt"]]))
```

## Outputs

```{r}
dir.create(dirname(output_object), recursive = TRUE, showWarnings = FALSE)
dir.create(dirname(output_metadata), recursive = TRUE, showWarnings = FALSE)

qs2::qs_save(object, output_object)
write_tsv(
  cbind(cell_id = rownames(object[[]]), object[[]]),
  output_metadata
)
```

## Method notes

The object-construction filters are not biological quality thresholds. Apply
the QC template after checking the resulting distributions. Change the
mitochondrial pattern for the organism and genome annotation.

::: {.callout-note}
## Practical note

Seurat expects features in rows and cells in columns. Metadata rows must have
the same unique cell names as the count-matrix columns; matching by position is
not a safe substitute for exact identifier matching.
:::
