---
title: "CytoTRACE2 cell potency"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
---

> Status: `draft`
>
> Template class: SOURCE-BACKED WORKFLOW

## Purpose

Run CytoTRACE2 on raw RNA counts, attach the potency score and categorical
potency output to a Seurat object, and check cell-ID alignment.

## Inputs and parameters

```{r}
library(Seurat)
library(CytoTRACE2)
library(qs2)
library(readr)
library(ggplot2)

input_object <- "input/seurat_object_qc.qs2"
output_object <- "output/seurat_object_cytotrace2.qs2"
output_scores <- "output/cytotrace2_scores.tsv"
output_plot <- "output/cytotrace2_score.png"

assay_name <- "RNA"
counts_layer <- "counts"
species <- "human"
seed <- 1234
ncores <- 1
reduction_name <- "umap"
```

## Analysis

```{r}
object <- qs2::qs_read(input_object)
counts <- GetAssayData(object, assay = assay_name, layer = counts_layer)
stopifnot(ncol(counts) == ncol(object))
stopifnot(identical(colnames(counts), colnames(object)))

set.seed(seed)
cytotrace_result <- cytotrace2(
  counts,
  is_seurat = FALSE,
  species = species,
  seed = seed,
  ncores = ncores
)

score <- cytotrace_result$CytoTRACE2_Score
potency <- cytotrace_result$CytoTRACE2_Potency
stopifnot(!is.null(names(score)), !is.null(names(potency)))
stopifnot(all(colnames(object) %in% names(score)))
stopifnot(all(colnames(object) %in% names(potency)))

object[["cytotrace2_score"]] <- score[colnames(object)]
object[["cytotrace2_potency"]] <- potency[colnames(object)]
scores <- data.frame(
  cell_id = colnames(object),
  cytotrace2_score = object$cytotrace2_score,
  cytotrace2_potency = object$cytotrace2_potency
)
```

## Diagnostics

```{r}
dir.create(dirname(output_plot), recursive = TRUE, showWarnings = FALSE)
print(summary(scores$cytotrace2_score))
print(table(scores$cytotrace2_potency, useNA = "ifany"))
if (reduction_name %in% Reductions(object)) {
  p <- FeaturePlot(object, features = "cytotrace2_score", reduction = reduction_name)
  ggsave(output_plot, p, width = 7, height = 5, dpi = 150)
}
```

## Outputs

```{r}
dir.create(dirname(output_object), recursive = TRUE, showWarnings = FALSE)
dir.create(dirname(output_scores), recursive = TRUE, showWarnings = FALSE)
write_tsv(scores, output_scores)
qs2::qs_save(object, output_object)
```

## Method notes

The source workflow applies CytoTRACE2 to raw RNA counts as an orthogonal
potency estimate. Do not pass normalized or scaled values without confirming
the package contract. The score is not a replacement for UCell, cell-cycle,
CNV analysis, or trajectory inference. Record package/API versions and gene
identifier conventions when validating this template.

::: {.callout-note}
## Validation required

Verify the installed CytoTRACE2 API, the expected raw-count orientation,
species resource, and returned cell-ID names on a representative object before
using this template for analysis.
:::
