---
title: "Single-cell doublet detection with scDblFinder"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
---

> Status: `draft`
>
> Template class: SOURCE-BACKED WORKFLOW

## Purpose

Call doublets with scDblFinder, attach the calls to a Seurat object, and make
the exclusion policy explicit. The default output retains all cells and
records the calls.

## Inputs and parameters

```{r}
library(Seurat)
library(SingleCellExperiment)
library(scDblFinder)
library(qs2)
library(readr)

input_object <- "input/seurat_object_qc.qs2"
output_object <- "output/seurat_object_doublet_calls.qs2"
output_calls <- "output/doublet_calls.tsv"

assay_name <- "RNA"
sample_column <- "sample_id"
remove_doublets <- FALSE
```

## Analysis

```{r}
object <- qs2::qs_read(input_object)
DefaultAssay(object) <- assay_name
stopifnot(sample_column %in% colnames(object[[]]))

sce <- as.SingleCellExperiment(object, assay = assay_name)
sce <- scDblFinder(
  sce,
  samples = colData(sce)[[sample_column]]
)

calls <- data.frame(
  cell_id = colnames(sce),
  doublet_class = colData(sce)$scDblFinder.class,
  doublet_score = colData(sce)$scDblFinder.score,
  sample_id = colData(sce)[[sample_column]],
  row.names = colnames(sce)
)

object[["scDblFinder.class"]] <- calls[colnames(object), "doublet_class"]
object[["scDblFinder.score"]] <- calls[colnames(object), "doublet_score"]

if (remove_doublets) {
  object <- subset(object, subset = scDblFinder.class == "singlet")
}
```

## Diagnostics

```{r}
print(table(calls$doublet_class, calls$sample_id, useNA = "ifany"))
print(summary(calls$doublet_score))
```

## Outputs

```{r}
dir.create(dirname(output_object), recursive = TRUE, showWarnings = FALSE)
dir.create(dirname(output_calls), recursive = TRUE, showWarnings = FALSE)
write_tsv(calls, output_calls)
qs2::qs_save(object, output_object)
```

## Method notes

The source workflows apply doublet handling in a sample-aware context. Confirm
whether calls should be made before or after merging samples for the study
design. The `sample_column` should represent an independent capture/library
loading experiment; unrelated libraries should not be silently pooled as one
capture. A doublet score is not a biological state.

::: {.callout-warning}
## Capture-library contract

Doublet detection is conditioned on the cells that could have co-occurred in a
capture. Use a sample/library column that reflects that experimental unit and
review the per-library call counts before deciding whether to remove cells.
:::
