---
title: "hdWGCNA co-expression modules"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
---

> Status: `draft`
>
> Template class: SOURCE-BACKED WORKFLOW

## Purpose

Construct co-expression modules and module eigengenes with hdWGCNA from a
processed Seurat object. This is network/module inference, not predefined
gene-set scoring.

## Inputs and parameters

The object must contain an expression assay, a reduction for metacell
construction, a sample column, and a biological grouping column.

```{r}
library(Seurat)
library(hdWGCNA)
library(qs2)
library(patchwork)
library(ggplot2)
library(readr)

input_object <- "input/seurat_object_processed.qs2"
output_object <- "output/seurat_object_hdwgcna.qs2"
output_modules <- "output/hdwgcna_modules.tsv"
output_hubs <- "output/hdwgcna_hub_genes.tsv"
output_power <- "output/hdwgcna_power_table.tsv"
output_power_plot <- "output/hdwgcna_soft_powers.png"
output_kme_plot <- "output/hdwgcna_kme.png"

assay_name <- "SCT"
expression_layer <- "data"
metacell_reduction <- "harmony"
metacell_group_columns <- c("sample_id", "cluster")
module_group_column <- "cluster"
eigengene_batch_column <- "sample_id"
metacell_k <- 25
max_shared <- 10
wgcna_name <- "wgcna"
network_type <- "signed hybrid"
soft_power <- 4
n_variable_features <- 3000
seed <- 1234
```

## Analysis

```{r}
object <- qs2::qs_read(input_object)
DefaultAssay(object) <- assay_name
stopifnot(all(metacell_group_columns %in% colnames(object[[]])))
stopifnot(metacell_reduction %in% Reductions(object))

features <- head(VariableFeatures(object), n_variable_features)
stopifnot(length(features) > 0)

object <- SetupForWGCNA(
  object,
  features = features,
  wgcna_name = wgcna_name
)
object <- MetacellsByGroups(
  seurat_obj = object,
  group.by = metacell_group_columns,
  reduction = metacell_reduction,
  k = metacell_k,
  max_shared = max_shared,
  ident.group = module_group_column
)
object <- NormalizeMetacells(object)

module_groups <- unique(as.character(object[[module_group_column]][, 1]))
object <- SetDatExpr(
  object,
  group_name = module_groups,
  group.by = module_group_column,
  assay = assay_name,
  layer = expression_layer
)
object <- TestSoftPowers(object, networkType = network_type)
power_table <- GetPowerTable(object)
object <- ConstructNetwork(
  object,
  randomSeed = seed,
  soft_power = soft_power,
  overwrite_tom = TRUE,
  tom_name = wgcna_name
)
object <- ModuleEigengenes(
  object,
  assay = assay_name,
  group.by.vars = eigengene_batch_column
)
object <- ModuleConnectivity(
  object,
  group.by = module_group_column,
  group_name = module_groups
)

modules <- GetModules(object)
modules <- modules[modules$module != "grey", , drop = FALSE]
hub_genes <- modules[order(-abs(modules$kME_max), na.last = NA), , drop = FALSE]
```

## Diagnostics

```{r}
dir.create(dirname(output_power_plot), recursive = TRUE, showWarnings = FALSE)
power_plot <- wrap_plots(PlotSoftPowers(object), ncol = 2)
ggsave(output_power_plot, power_plot, width = 10, height = 7, dpi = 150)
print(power_table)
print(head(modules))
print(head(hub_genes))
kme_plot <- PlotKMEs(object, ncol = 3)
ggsave(output_kme_plot, kme_plot, width = 10, height = 7, dpi = 150)
```

## Outputs

```{r}
dir.create(dirname(output_object), recursive = TRUE, showWarnings = FALSE)
dir.create(dirname(output_modules), recursive = TRUE, showWarnings = FALSE)
qs2::qs_save(object, output_object)
write_tsv(as.data.frame(power_table), output_power)
write_tsv(as.data.frame(modules), output_modules)
write_tsv(as.data.frame(hub_genes), output_hubs)
```

## Method notes

Metacell grouping, sample grouping, network type, soft power, and hub ranking
are scientific decisions. The source used cluster-focused grouping; this
template exposes sample and cluster grouping so replicate structure is not
silently discarded. A soft-power value should be justified from diagnostics.
AddModuleScore-style scoring is not this network analysis.
