---
title: "Single-cell marker comparison"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
---

> Status: `draft`
>
> Template class: SOURCE-BACKED WORKFLOW

## Purpose

Compare two explicitly named cell groups with Seurat FindMarkers and export a
compact marker table. The printed group sizes are a basic diagnostic. This is a
group comparison, not an all-cluster marker
screen.

## Inputs and parameters

```{r}
library(Seurat)
library(qs2)
library(readr)
library(ggplot2)

input_object <- "input/seurat_object_clustered.qs2"
output_markers <- "output/markers_group_1_vs_group_2.tsv"
output_plot <- "output/marker_dotplot.png"

assay_name <- "RNA"
group_column <- "cluster"
group_1 <- "group_1"
group_2 <- "group_2"
test_use <- "wilcox"
min_pct <- 0.10
logfc_threshold <- 0.25
only_positive <- FALSE
n_plot_genes <- 12
```

## Analysis

```{r}
object <- qs2::qs_read(input_object)
DefaultAssay(object) <- assay_name
stopifnot(group_column %in% colnames(object[[]]))
Idents(object) <- object[[group_column, drop = TRUE]]
group_counts <- table(Idents(object))
stopifnot(all(c(group_1, group_2) %in% names(group_counts)))
print(group_counts[c(group_1, group_2)])

markers <- FindMarkers(
  object,
  ident.1 = group_1,
  ident.2 = group_2,
  assay = assay_name,
  test.use = test_use,
  min.pct = min_pct,
  logfc.threshold = logfc_threshold,
  only.pos = only_positive
)
markers <- data.frame(gene = rownames(markers), markers, row.names = NULL)
```

## Diagnostics

```{r}
dir.create(dirname(output_plot), recursive = TRUE, showWarnings = FALSE)
print(head(markers, 20))
print(nrow(markers))
if (nrow(markers) > 0) {
  top_genes <- head(markers$gene, n_plot_genes)
  p <- DotPlot(object, features = top_genes, group.by = group_column) + RotatedAxis()
  ggsave(output_plot, p, width = 10, height = 5, dpi = 150)
}
```

## Outputs

```{r}
dir.create(dirname(output_markers), recursive = TRUE, showWarnings = FALSE)
write_tsv(markers, output_markers)
```

## Method notes

The test, minimum detection fraction, and log-fold-change threshold remain
visible. FindMarkers is not a substitute for a replicated pseudobulk or
design-based differential-expression analysis. Add an all-cluster marker
template only if that is a separate supported use case.

::: {.callout-warning}
## Biological replication warning

`FindMarkers()` here is a cell-level exploratory comparison. When patients or
samples are the biological replicates, treating thousands of cells from a few
samples as independent replicates can produce overconfident inference. Use a
sample-aware design or pseudobulk analysis for replicated claims; that workflow
is intentionally outside this template.
:::
