---
title: "SCTransform normalization"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
---

> Status: `draft`
>
> Template class: SOURCE-BACKED WORKFLOW

## Purpose

Normalize a filtered Seurat object with SCTransform. The assay, covariates
regressed, number of variable features, and retention of all genes are
explicit.

## Inputs and parameters

```{r}
library(Seurat)
library(qs2)
library(ggplot2)

input_object <- "input/seurat_object_qc.qs2"
output_object <- "output/seurat_object_sct.qs2"
output_plot <- "output/sct_variable_features.png"

assay_name <- "RNA"
vars_to_regress <- character()
n_variable_features <- 3000
return_only_variable_genes <- FALSE
n_pcs <- 30
pca_reduction <- "pca"
```

## Analysis

```{r}
object <- qs2::qs_read(input_object)
DefaultAssay(object) <- assay_name
regressors <- if (length(vars_to_regress) == 0) NULL else vars_to_regress

object <- SCTransform(
  object,
  assay = assay_name,
  vars.to.regress = regressors,
  variable.features.n = n_variable_features,
  return.only.var.genes = return_only_variable_genes,
  verbose = FALSE
)
object <- RunPCA(
  object,
  assay = "SCT",
  features = VariableFeatures(object, assay = "SCT"),
  npcs = n_pcs,
  reduction.name = pca_reduction,
  verbose = FALSE
)
```

## Diagnostics

```{r}
dir.create(dirname(output_plot), recursive = TRUE, showWarnings = FALSE)
print(object)
print(head(VariableFeatures(object), 20))
print(length(VariableFeatures(object)))
p <- VariableFeaturePlot(object, assay = "SCT")
ggsave(output_plot, p, width = 7, height = 5, dpi = 150)
```

## Outputs

```{r}
dir.create(dirname(output_object), recursive = TRUE, showWarnings = FALSE)
qs2::qs_save(object, output_object)
```

## Method notes

SCTransform is not log normalization with different syntax. Regression
variables and the resulting assay are scientific decisions. The downstream
Harmony template expects PCA computed on the assay selected there. The output
is PCA-ready and contains the SCT assay plus its PCA reduction. Harmony is kept
separate so the choice between log normalization and SCTransform remains
visible.
