---
title: "Single-cell quality control"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
---

> Status: `draft`
>
> Template class: SOURCE-BACKED WORKFLOW

## Purpose

Apply transparent cell-level QC thresholds to an existing Seurat object and
save the filtered object. Thresholds are visible and are not universal
defaults.

## Inputs and parameters

```{r}
library(Seurat)
library(ggplot2)
library(readr)
library(dplyr)

input_object <- "input/seurat_object.qs2"
output_object <- "output/seurat_object_qc.qs2"
output_metadata <- "output/qc_metadata.tsv"
output_plot <- "output/qc_distributions.png"
output_sample_plot <- "output/qc_distributions_by_sample.png"

assay_name <- "RNA"
sample_column <- "sample_id"
mitochondrial_pattern <- "^MT-"
mitochondrial_column <- "percent.mt"
min_features <- 200
max_features <- Inf
min_counts <- 0
max_counts <- Inf
```

## Analysis

```{r}
object <- qs2::qs_read(input_object)
DefaultAssay(object) <- assay_name

feature_column <- paste0("nFeature_", assay_name)
count_column <- paste0("nCount_", assay_name)
stopifnot(all(c(feature_column, count_column) %in% colnames(object[[]])))

if (!mitochondrial_column %in% colnames(object[[]])) {
  object[[mitochondrial_column]] <- PercentageFeatureSet(
    object,
    assay = assay_name,
    pattern = mitochondrial_pattern
  )
}

metadata <- object[[]]
metadata$qc_pass <- with(
  metadata,
  !is.na(metadata[[feature_column]]) &
    !is.na(metadata[[count_column]]) &
    !is.na(metadata[[mitochondrial_column]]) &
    metadata[[feature_column]] >= min_features &
    metadata[[feature_column]] <= max_features &
    metadata[[count_column]] >= min_counts &
    metadata[[count_column]] <= max_counts
)

object_qc <- subset(object, cells = rownames(metadata)[metadata$qc_pass])
```

## Diagnostics

```{r}
dir.create(dirname(output_plot), recursive = TRUE, showWarnings = FALSE)
qc_summary <- data.frame(
  cells_before = ncol(object),
  cells_after = ncol(object_qc),
  cells_removed = ncol(object) - ncol(object_qc),
  fraction_pass = ncol(object_qc) / ncol(object)
)
print(qc_summary)
print(table(metadata$qc_pass, useNA = "ifany"))

p <- VlnPlot(
  object,
  features = c(feature_column, count_column, mitochondrial_column),
  group.by = if (sample_column %in% colnames(object[[]])) sample_column else NULL,
  ncol = 3,
  pt.size = 0
)
ggsave(output_plot, p, width = 11, height = 4, dpi = 150)

if (sample_column %in% colnames(metadata)) {
  sample_summary <- metadata |>
    group_by(sample = .data[[sample_column]]) |>
    summarise(
      cells_before = n(),
      cells_after = sum(qc_pass),
      fraction_pass = cells_after / cells_before,
      median_nFeature = median(.data[[feature_column]], na.rm = TRUE),
      median_nCount = median(.data[[count_column]], na.rm = TRUE),
      median_percent_mt = median(.data[[mitochondrial_column]], na.rm = TRUE),
      .groups = "drop"
    )
  print(sample_summary)
  p_sample <- VlnPlot(
    object,
    features = c(feature_column, count_column, mitochondrial_column),
    group.by = sample_column,
    ncol = 3,
    pt.size = 0
  )
  ggsave(output_sample_plot, p_sample, width = 12, height = 5, dpi = 150)
}
```

## Outputs

```{r}
dir.create(dirname(output_object), recursive = TRUE, showWarnings = FALSE)
dir.create(dirname(output_metadata), recursive = TRUE, showWarnings = FALSE)
write_tsv(cbind(cell_id = rownames(metadata), metadata), output_metadata)
qs2::qs_save(object_qc, output_object)
```

## Method notes

The source workflows use both fixed and MAD-derived thresholds. This template
implements fixed thresholds only; a MAD policy should be a separate method.
The saved object is the filtered object.

::: {.callout-note}
## Workflow preference

I generally inspect QC distributions separately by sample before filtering,
because a pooled threshold can hide sample-specific behavior. Fixed thresholds
are transparent, but they are not biological constants. Consider feature count,
total counts, mitochondrial fraction, doublet results, and biological context
together. The source notebooks also used distribution-aware/MAD ideas, but the
exact multiplier is deliberately left to the project rather than silently
choosing between incompatible policies.
:::
