---
title: "Slingshot trajectory inference"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
---

> Status: `draft`
>
> Template class: API / TUTORIAL REMINDER

## What it does

Slingshot infers a cluster-level lineage structure from a reduced-dimensional
representation and fits smooth principal curves through the inferred lineages.
It is useful when a continuous transition or branching process is biologically
plausible and clustering has already been performed.

::: {.callout-note}
## API reminder

This is an API/tutorial reminder constructed from the current official
Slingshot API, not an executed analysis pipeline. Consult the current package
documentation and vignette before applying it to real data.
:::

## Inputs and parameters

The input Seurat object must contain a reduced-dimensional embedding and a
cell-level cluster column. The cluster resolution and the choice of reduction
are scientific decisions, not merely plotting settings.

```{r}
library(Seurat)
library(SingleCellExperiment)
library(slingshot)
library(qs2)
library(readr)

input_object <- "input/seurat_object_clustered.qs2"
output_object <- "output/slingshot_sce.qs2"
output_pseudotime <- "output/slingshot_pseudotime.tsv"
output_weights <- "output/slingshot_curve_weights.tsv"
output_plot <- "output/slingshot_trajectories.png"

assay_name <- "RNA"
reduction_name <- "pca"
cluster_column <- "cluster"
start_cluster <- NULL
end_clusters <- NULL
```

## Minimal API

```{r}
object <- qs2::qs_read(input_object)
stopifnot(reduction_name %in% Reductions(object))
stopifnot(cluster_column %in% colnames(object[[]]))
stopifnot(identical(rownames(Embeddings(object, reduction_name)), colnames(object)))

sce <- as.SingleCellExperiment(object, assay = assay_name)
stopifnot(identical(colnames(sce), colnames(object)))
reducedDim(sce, reduction_name) <- Embeddings(object, reduction_name)
colData(sce)[[cluster_column]] <- factor(object[[cluster_column, drop = TRUE]])

sce <- slingshot(
  sce,
  clusterLabels = cluster_column,
  reducedDim = reduction_name,
  start.clus = start_cluster,
  end.clus = end_clusters
)

pseudotime <- slingPseudotime(sce)
curve_weights <- slingCurveWeights(sce)
```

## Outputs and diagnostic

```{r}
print(dim(pseudotime))
print(colSums(!is.na(pseudotime)))

dir.create(dirname(output_object), recursive = TRUE, showWarnings = FALSE)
png(output_plot, width = 900, height = 700)
plot(
  reducedDim(sce, reduction_name)[, 1:2],
  col = as.integer(colData(sce)[[cluster_column]]),
  pch = 16,
  asp = 1,
  xlab = reduction_name,
  ylab = paste0(reduction_name, " 2")
)
lines(SlingshotDataSet(sce), lwd = 2)
dev.off()

qs2::qs_save(sce, output_object)
write_tsv(
  data.frame(cell_id = rownames(pseudotime), pseudotime),
  output_pseudotime
)
write_tsv(
  data.frame(cell_id = rownames(curve_weights), curve_weights),
  output_weights
)
```

## Method notes

Review the reduction, cluster resolution, root/starting cluster, terminal
constraints, branching plausibility, and curve fit before interpreting
pseudotime. The pseudotime matrix and curve weights can be passed to tradeSeq;
tradeSeq is a separate gene-level GAM analysis.

For current arguments, branching constraints, lineage diagnostics, and
version-specific behavior, consult the [official Slingshot package
documentation](https://bioconductor.org/packages/release/bioc/html/slingshot.html)
and [vignette](https://github.com/kstreet13/slingshot/blob/main/vignettes/vignette.Rmd).
