---
title: "TCGA-LUAD MOFA2 Multi-Omics Model"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    theme: lux
    fig-width: 12
    fig-height: 8
    fig-align: center
    fig-responsive: true
    df-print: paged
    self-contained: true
execute:
  echo: false
  warning: false
  message: false
knitr:
  opts_chunk:
    fig.width: 12
    fig.height: 8
    fig.align: "center"
    out.width: "100%"
    dpi: 600
    dev: "ragg_png"
editor: source
---

# Setup

```{r}
#| label: setup
#| echo: false
#| warning: false
#| message: false
#| include: false

library(data.table)
library(MOFA2)
library(future)
library(tidyverse)
library(ggpubr)
library(ggsci)
library(matrixStats)
library(patchwork)
library(rcompanion)


# Keep choices explicit because these packages share common
# function names used later in the notebook.
conflicted::conflicts_prefer(
  dplyr::filter,
  dplyr::select,
  dplyr::first,
  janitor::clean_names,
  base::intersect
)

N_WORKERS <- min(10, future::availableCores())
if (N_WORKERS > 1) {
  plan(multisession, workers = N_WORKERS)
} else {
  plan(sequential)
}
cat("Number of workers:", nbrOfWorkers(), "\n")

RANDOM_SEED <- 12345

FIG_DIR <- "output/figures"
OBJECT_DIR <- "output/objects"
FIG_DPI <- 600
dir.create(FIG_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(OBJECT_DIR, recursive = TRUE, showWarnings = FALSE)

save_plot_png <- function(plot, filename, width = 10, height = 6) {
  ggplot2::ggsave(
    filename = file.path(FIG_DIR, filename),
    plot = plot,
    width = width,
    height = height,
    units = "in",
    dpi = FIG_DPI,
    bg = "white"
  )
}

compute_logcpm <- function(count_mat) {
  # MOFA receives normalized log-CPM miRNA values, not raw counts.
  dge <- edgeR::DGEList(counts = count_mat)
  keep <- rowSums(edgeR::cpm(dge) > 1) >= 3
  dge <- dge[keep, , keep.lib.sizes = FALSE]
  dge <- edgeR::normLibSizes(dge, method = "TMMwsp")
  edgeR::cpm(dge, log = TRUE, prior.count = 1)
}

keep_most_variable_dt <- function(mat, top_n = 5000, min_non_na_frac = 0.8) {
  mat <- as.matrix(mat)
  storage.mode(mat) <- "numeric"

  keep <- matrixStats::rowMeans2(!is.na(mat)) >= min_non_na_frac
  mat <- mat[keep, , drop = FALSE]

  vars <- matrixStats::rowVars(mat, na.rm = TRUE)

  var_dt <- data.table::data.table(
    feature = rownames(mat),
    variance = vars
  ) |>
    dplyr::filter(is.finite(variance)) |>
    dplyr::arrange(dplyr::desc(variance))

  keep_features <- var_dt |>
    dplyr::slice_head(n = min(top_n, nrow(var_dt))) |>
    dplyr::pull(feature)

  mat[keep_features, , drop = FALSE]
}

to_numeric_matrix <- function(x) {
  x <- as.matrix(x)
  storage.mode(x) <- "numeric"
  x
}

remove_zero_var <- function(x) {
  x <- to_numeric_matrix(x)
  keep <- apply(x, 1, function(z) var(z, na.rm = TRUE) > 0)
  x[keep, , drop = FALSE]
}
clinical_luad <- TCGAbiolinks::GDCquery_clinic(
  project = "TCGA-LUAD",
  type = "clinical"
) |>
  janitor::clean_names() |>
  dplyr::mutate(
    submitter_id = as.character(submitter_id)
  ) |>
  dplyr::distinct(submitter_id, .keep_all = TRUE)


```

# Build MOFA Input

## Load Multi-Omics Inputs


Load the matched TCGA-LUAD molecular views that will be used as MOFA input matrices.

```{r}
#| label: load-multiomics-inputs

miRNA <- vroom::vroom("input/tables/miRNA_TCGA_TUMOR_MATCHED.csv",show_col_types = FALSE) |>
  na.omit() |>
  column_to_rownames("miRNA") |>
  compute_logcpm() |>
  as.data.frame()

nrow(miRNA)

RNA <- vroom::vroom("input/tables/HUGO_RNAseq_TCGA_TUMOR_MATCHED.csv",show_col_types = FALSE) |>
  na.omit() |>
  column_to_rownames("gene_symbol") |>
  compute_logcpm() |>
  as.data.frame() |>
  keep_most_variable_dt(top_n = 5000)

nrow(RNA)

RPPA <- vroom::vroom("input/tables/RPPA_TCGA_TUMOR_MATCHED.csv",show_col_types = FALSE) |>
  column_to_rownames("protein")

nrow(RPPA)

mutation_min_freq <- 0.05
mutation_max_freq <- 0.97

mutation <- vroom::vroom(
  "input/tables/Mutation_TCGA_TUMOR_MATCHED.csv",
  show_col_types = FALSE
) |>
  column_to_rownames("gene") |>
  as.matrix()

storage.mode(mutation) <- "numeric"
mutation[is.na(mutation)] <- 0
mutation <- +(mutation > 0)

mutation_burden_all <- tibble(
  sample = colnames(mutation),
  mutation_burden_all = colSums(mutation),
  mutation_burden_all_z = as.numeric(scale(mutation_burden_all))
)

mutation_freq <- rowMeans(mutation)

mutation <- mutation[
  mutation_freq >= mutation_min_freq & mutation_freq <= mutation_max_freq,
  ,
  drop = FALSE
]

nrow(mutation)

```

## Harmonize Samples Across Modalities

Restrict all views to the same ordered sample set before constructing the MOFA object.

```{r}
#| label: harmonize-samples-across-modalities

sample_lists <- list(
  RNA = colnames(RNA),
  miRNA = colnames(miRNA),
  RPPA = colnames(RPPA),
  mutation = colnames(mutation)
)

common_samples <- Reduce(intersect, sample_lists)

RNA <- RNA[, common_samples, drop = FALSE]
miRNA <- miRNA[, common_samples, drop = FALSE]
RPPA <- RPPA[, common_samples, drop = FALSE]
mutation <- mutation[, common_samples, drop = FALSE]

```

## Prepare MOFA Input Matrices

Convert all views to numeric matrices, remove zero-variance features, scale continuous views, and keep mutation as binary.

```{r}
#| label: prepare-mofa-input-matrices

mofa_data <- list(
  RNA = RNA,
  miRNA = miRNA,
  RPPA = RPPA,
  mutation = mutation
)

mofa_data <- lapply(mofa_data, to_numeric_matrix)
mofa_data <- lapply(mofa_data, remove_zero_var)

scale_rows <- function(x) {
  x <- t(scale(t(x)))
  x[is.na(x)] <- 0
  x
}

continuous_views <- c(
  "RNA",
  "miRNA",
  "RPPA"
)

mofa_data[continuous_views] <- lapply(
  mofa_data[continuous_views],
  scale_rows
)

mofa_data$mutation <- ifelse(mofa_data$mutation > 0, 1, 0)
storage.mode(mofa_data$mutation) <- "numeric"

mutation_freq <- rowMeans(mofa_data$mutation, na.rm = TRUE)

mofa_data$mutation <- mofa_data$mutation[
  mutation_freq >= 0.03 & mutation_freq <= 0.97,
  ,
  drop = FALSE
]

stopifnot(
  is.list(mofa_data),
  all(vapply(mofa_data, is.matrix, logical(1))),
  all(vapply(mofa_data, is.numeric, logical(1))),
  length(unique(vapply(mofa_data, ncol, integer(1)))) == 1,
  all(vapply(
    mofa_data,
    function(x) identical(colnames(x), common_samples),
    logical(1)
  )),
  all(mofa_data$mutation %in% c(0, 1))
)

qs2::qs_save(
  object = mofa_data,
  file = file.path(OBJECT_DIR, "mofa_data_list.qs2"),
  compress_level = 10,
  nthreads = 10
)

mofa_qc <- data.frame(
  view = names(mofa_data),
  n_features = vapply(mofa_data, nrow, integer(1)),
  n_samples = vapply(mofa_data, ncol, integer(1)),
  missing_fraction = vapply(mofa_data, function(x) mean(is.na(x)), numeric(1)),
  mean_feature_mean = vapply(mofa_data, function(x) mean(rowMeans(x, na.rm = TRUE), na.rm = TRUE), numeric(1)),
  mean_feature_sd = vapply(mofa_data, function(x) mean(apply(x, 1, sd, na.rm = TRUE), na.rm = TRUE), numeric(1)),
  median_feature_sd = vapply(mofa_data, function(x) median(apply(x, 1, sd, na.rm = TRUE), na.rm = TRUE), numeric(1)),
  zero_fraction = vapply(mofa_data, function(x) mean(x == 0, na.rm = TRUE), numeric(1)),
  constant_features = vapply(mofa_data, function(x) sum(apply(x, 1, sd, na.rm = TRUE) == 0, na.rm = TRUE), integer(1))
)

mofa_qc


```

# Train MOFA Model

Create the MOFA object, set the model likelihoods, and train the mixed-likelihood model.

```{r}
#| label: train-mofa-model

# MOFA2 delegates model fitting to mofapy2 through reticulate.
reticulate::py_require("mofapy2")


MOFAobject <- create_mofa(mofa_data)


mofa_samples <- samples_names(MOFAobject)[[1]]

# Load continuous and categorical C1-Core scores used for downstream factor interpretation.

C1_Core <- vroom::vroom(
  "output/tables/C1_core_singscore_GSVA_consensus_scores.csv",
  show_col_types = FALSE
) |>
  dplyr::select(sample, C1_consensus, C1_consensus_group) |>
  dplyr::mutate(
    sample = as.character(sample),
    C1_consensus = as.numeric(C1_consensus),
    C1_consensus_group = factor(
      C1_consensus_group,
      levels = c(
        "Not-C1-like",
        "Not-C1-intermediate",
        "C1-intermediate",
        "C1-like"
      )
    )
  ) |>
  dplyr::distinct(sample, .keep_all = TRUE)

# Add tumor-purity estimates from TCGAbiolinks and standardize mixed string/numeric encodings.

data("Tumor.purity", package = "TCGAbiolinks")

parse_purity <- function(x) {
  x <- as.character(x)
  x[x %in% c("NaN", "NA", "NULL", "", "nan")] <- NA_character_
  as.numeric(gsub(",", ".", x))
}

tumor_purity_luad <- TCGAbiolinks::Tumor.purity |>
  janitor::clean_names() |>
  dplyr::filter(cancer_type == "LUAD") |>
  dplyr::mutate(
    sample = as.character(sample_id),
    case_submitter_id = substr(sample, 1, 12),
    purity_estimate = parse_purity(estimate),
    purity_absolute = parse_purity(absolute),
    purity_lump = parse_purity(lump),
    purity_ihc = parse_purity(ihc),
    purity_cpe = parse_purity(cpe)
  ) |>
  dplyr::select(
    case_submitter_id,
    purity_estimate,
    purity_absolute,
    purity_lump,
    purity_ihc,
    purity_cpe
  ) |>
  dplyr::distinct(case_submitter_id, .keep_all = TRUE)

# Assemble the final MOFA metadata table in the same sample order as the model object.

metadata_mofa <- tibble::tibble(
  sample = mofa_samples,
  case_submitter_id = substr(sample, 1, 12),
  sample_type_code = substr(sample, 14, 15),
  vial = substr(sample, 16, 16)
) |>
  dplyr::left_join(
    clinical_luad,
    by = c("case_submitter_id" = "submitter_id")
  ) |>
  dplyr::left_join(
    tumor_purity_luad,
    by = "case_submitter_id"
  ) |>
  dplyr::left_join(
    C1_Core,
    by = "sample"
  ) |>
  dplyr::mutate(
    C1_Core = as.numeric(C1_consensus),
    C1_Core_group = factor(
      C1_consensus_group,
      levels = c(
        "Not-C1-like",
        "Not-C1-intermediate",
        "C1-intermediate",
        "C1-like"
      )
    ),
    purity_primary = dplyr::coalesce(
      purity_cpe,
      purity_absolute,
      purity_estimate,
      purity_ihc
    ),
    purity_primary_method = dplyr::case_when(
      !is.na(purity_cpe) ~ "CPE",
      is.na(purity_cpe) & !is.na(purity_absolute) ~ "ABSOLUTE",
      is.na(purity_cpe) & is.na(purity_absolute) & !is.na(purity_estimate) ~ "ESTIMATE",
      is.na(purity_cpe) & is.na(purity_absolute) & is.na(purity_estimate) & !is.na(purity_ihc) ~ "IHC",
      TRUE ~ NA_character_
    ),
    purity_primary_z = as.numeric(scale(purity_primary)),
    purity_cpe_z = as.numeric(scale(purity_cpe)),
    purity_absolute_z = as.numeric(scale(purity_absolute)),
    purity_estimate_z = as.numeric(scale(purity_estimate)),
    age_at_diagnosis_years = dplyr::coalesce(
      as.numeric(age_at_diagnosis) / 365.25,
      as.numeric(age_at_index)
    ),
    os_time_days = dplyr::case_when(
      !is.na(days_to_death) ~ as.numeric(days_to_death),
      !is.na(days_to_last_follow_up) ~ as.numeric(days_to_last_follow_up),
      TRUE ~ NA_real_
    ),
    os_event = dplyr::case_when(
      vital_status == "Dead" ~ 1,
      vital_status == "Alive" ~ 0,
      TRUE ~ NA_real_
    ),
    tumor_stage = ajcc_pathologic_stage,
    tumor_t = ajcc_pathologic_t,
    tumor_n = ajcc_pathologic_n,
    tumor_m = ajcc_pathologic_m,
    sex = gender,
    smoking_status = tobacco_smoking_status,
    sample_type = dplyr::case_when(
      sample_type_code == "01" ~ "Primary Tumor",
      sample_type_code == "06" ~ "Metastatic",
      sample_type_code == "11" ~ "Solid Tissue Normal",
      TRUE ~ sample_type_code
    )
  ) |>
  dplyr::select(
    sample,
    case_submitter_id,
    sample_type_code,
    sample_type,
    vial,
    C1_Core,
    C1_Core_group,
    purity_primary,
    purity_primary_z,
    purity_primary_method,
    purity_cpe,
    purity_cpe_z,
    purity_absolute,
    purity_absolute_z,
    purity_estimate,
    purity_estimate_z,
    purity_lump,
    purity_ihc,
    sex,
    race,
    ethnicity,
    age_at_diagnosis_years,
    vital_status,
    os_time_days,
    os_event,
    tumor_stage,
    tumor_t,
    tumor_n,
    tumor_m,
    pack_years_smoked,
    smoking_status,
    tobacco_smoking_quit_year,
    tobacco_smoking_onset_year,
    tissue_or_organ_of_origin,
    laterality,
    primary_diagnosis,
    morphology,
    residual_disease,
    prior_malignancy,
    prior_treatment,
    synchronous_malignancy,
    treatments_pharmaceutical_treatment_or_therapy,
    treatments_radiation_treatment_or_therapy,
    dplyr::everything()
  )

stopifnot(
  identical(metadata_mofa$sample, mofa_samples),
  !anyNA(metadata_mofa$C1_Core),
  !anyNA(metadata_mofa$C1_Core_group)
)

samples_metadata(MOFAobject) <- metadata_mofa

# Summarize missingness and key metadata distributions before model fitting.

metadata_qc <- metadata_mofa |>
  dplyr::summarise(
    dplyr::across(
      dplyr::everything(),
      ~ sum(is.na(.x))
    )
  ) |>
  tidyr::pivot_longer(
    dplyr::everything(),
    names_to = "field",
    values_to = "n_missing"
  ) |>
  dplyr::arrange(dplyr::desc(n_missing))

metadata_qc

metadata_mofa |>
  dplyr::count(sample_type, sample_type_code)

metadata_mofa |>
  dplyr::count(C1_Core_group)

metadata_mofa |>
  dplyr::count(purity_primary_method)

metadata_mofa |>
  dplyr::summarise(
    n_samples = dplyr::n(),
    n_cases = dplyr::n_distinct(case_submitter_id),
    n_missing_clinical = sum(is.na(vital_status)),
    n_missing_C1 = sum(is.na(C1_Core)),
    n_missing_os = sum(is.na(os_time_days)),
    n_missing_purity_primary = sum(is.na(purity_primary)),
    n_missing_purity_cpe = sum(is.na(purity_cpe)),
    n_missing_purity_absolute = sum(is.na(purity_absolute)),
    n_missing_purity_estimate = sum(is.na(purity_estimate)),
    n_dead = sum(os_event == 1, na.rm = TRUE),
    n_alive = sum(os_event == 0, na.rm = TRUE)
  )

head(samples_metadata(MOFAobject))


```

```{r}

#| label: train-mofa-model

p_mofa_train_data_overview <- plot_data_overview(MOFAobject) +
  ggplot2::ggtitle("MOFA direct-omics training data overview")

p_mofa_train_data_overview

save_plot_png(
  p_mofa_train_data_overview,
  filename = "MOFA_direct_omics_train_data_overview.png",
  width = 10,
  height = 6
)

mofa_outdir <- "output/mofa"
dir.create(mofa_outdir, recursive = TRUE, showWarnings = FALSE)

mofa_outfile <- file.path(
  mofa_outdir,
  "LUAD_TCGA_MOFA_scaled_direct_omics.hdf5"
)

if (file.exists(mofa_outfile)) {
  warning("MOFA output file already exists and will be overwritten: ", mofa_outfile)
}

data_opts <- get_default_data_options(MOFAobject)
data_opts$scale_views <- FALSE
data_opts$use_float32 <- TRUE

model_opts <- get_default_model_options(MOFAobject)
model_opts$num_factors <- 15

likelihoods <- c(
  RNA = "gaussian",
  miRNA = "gaussian",
  RPPA = "gaussian",
  mutation = "bernoulli"
)

available_views <- views_names(MOFAobject)

missing_likelihoods <- setdiff(available_views, names(likelihoods))
extra_likelihoods <- setdiff(names(likelihoods), available_views)

if (length(missing_likelihoods) > 0) {
  stop(
    "Missing likelihoods for these MOFA views: ",
    paste(missing_likelihoods, collapse = ", ")
  )
}

if (length(extra_likelihoods) > 0) {
  warning(
    "Likelihoods were provided for views not present in MOFAobject: ",
    paste(extra_likelihoods, collapse = ", ")
  )
}

model_opts$likelihoods <- likelihoods[available_views]
model_opts$spikeslab_weights <- TRUE
model_opts$ard_weights <- TRUE
model_opts$ard_factors <- FALSE

train_opts <- get_default_training_options(MOFAobject)
train_opts$seed <- RANDOM_SEED
train_opts$convergence_mode <- "slow"
train_opts$maxiter <- 1500
train_opts$verbose <- TRUE
train_opts$outfile <- mofa_outfile
train_opts$drop_factor_threshold <- 0.01

MOFAobject <- prepare_mofa(
  object = MOFAobject,
  data_options = data_opts,
  model_options = model_opts,
  training_options = train_opts
)

MOFAobject <- run_mofa(
  MOFAobject,
  use_basilisk = FALSE
)

```

# Interpret MOFA Model

## MOFA Model Overview

Reload the trained model and inspect the fitted factors, variance explained, and factor correlations.

```{r}
#| label: mofa-model-overview

MOFAobject <- load_model("output/mofa/LUAD_TCGA_MOFA_scaled_direct_omics.hdf5")

mofa_samples <- samples_names(MOFAobject)[[1]]
C1_Core_mofa <- samples_metadata(MOFAobject)

p_mofa_variance_explained <- plot_variance_explained(MOFAobject) +
  ggtitle("MOFA variance explained")

p_mofa_variance_explained
save_plot_png(
  p_mofa_variance_explained,
  filename = "MOFA_variance_explained.png",
  width = 10,
  height = 6
)

png(
  file.path(FIG_DIR, "MOFA_factor_spearman_correlation.png"),
  width = 10,
  height = 10,
  units = "in",
  res = FIG_DPI
)

plot_factor_cor(
  MOFAobject,
  method = "spearman"
)

title("MOFA factor Spearman correlation")

dev.off()


```

```{r}
#| label: extract-factors-and-variance

factors <- get_factors(MOFAobject, factors = "all")[[1]]

factors_df <- as.data.frame(factors)
factors_df$sample <- rownames(factors_df)

head(factors_df)

variance_explained <- get_variance_explained(MOFAobject)

r2_per_factor <- variance_explained$r2_per_factor
r2_total <- variance_explained$r2_total

r2_per_factor
r2_total

```

## C1-Core Association Analysis

```{r}
#| label: c1-core-factor-correlations

factors <- MOFA2::get_factors(MOFAobject, factors = "all")[[1]]

factors_df <- factors |>
  as.data.frame() |>
  tibble::rownames_to_column("sample") |>
  dplyr::as_tibble()

factor_c1_df <- factors_df |>
  dplyr::inner_join(C1_Core_mofa, by = "sample")

factor_c1_long <- factor_c1_df |>
  tidyr::pivot_longer(
    cols = dplyr::starts_with("Factor"),
    names_to = "factor",
    values_to = "factor_value"
  )

# Test each MOFA factor against the continuous C1-core score.
cor_tbl <- factor_c1_long |>
  dplyr::group_by(factor) |>
  dplyr::summarise(
    n = sum(stats::complete.cases(factor_value, C1_Core)),
    spearman_rho = cor(
      factor_value,
      C1_Core,
      method = "spearman",
      use = "complete.obs"
    ),
    pearson_r = cor(
      factor_value,
      C1_Core,
      method = "pearson",
      use = "complete.obs"
    ),
    spearman_p = cor.test(
      factor_value,
      C1_Core,
      method = "spearman",
      exact = FALSE
    )$p.value,
    pearson_p = cor.test(
      factor_value,
      C1_Core,
      method = "pearson"
    )$p.value,
    .groups = "drop"
  ) |>
  dplyr::mutate(
    spearman_fdr = p.adjust(spearman_p, method = "BH"),
    pearson_fdr = p.adjust(pearson_p, method = "BH"),
    abs_spearman_rho = abs(spearman_rho)
  ) |>
  dplyr::arrange(dplyr::desc(abs_spearman_rho))

cor_tbl


```

## C1-Core Plots

Summarize the strongest C1-core/factor relationships with matrix, scatter, and correlation-bar plots.

### Correlation Summary

```{r}
#| label: c1-core-correlation-barplot

p_cor_bar <- cor_tbl |>
  dplyr::mutate(
    factor = forcats::fct_reorder(factor, spearman_rho),
    significant = spearman_fdr < 0.05,
    direction = dplyr::if_else(spearman_rho >= 0, "Positive", "Negative"),
    label = dplyr::if_else(
      significant,
      sprintf("rho = %.2f\nFDR = %.2g", spearman_rho, spearman_fdr),
      ""
    ),
    label_y = spearman_rho + dplyr::if_else(spearman_rho >= 0, 0.035, -0.035),
    label_hjust = dplyr::if_else(spearman_rho >= 0, 0, 1)
  ) |>
  ggplot2::ggplot(
    ggplot2::aes(
      x = factor,
      y = spearman_rho,
      fill = direction,
      alpha = significant
    )
  ) +
  ggplot2::geom_hline(yintercept = 0, linewidth = 0.4, colour = "grey35") +
  ggplot2::geom_col(width = 0.72, colour = "black", linewidth = 0.15) +
  ggplot2::geom_text(
    ggplot2::aes(y = label_y, label = label, hjust = label_hjust),
    size = 3.1,
    lineheight = 0.9
  ) +
  ggplot2::coord_flip(clip = "off") +
  ggsci::scale_fill_npg(name = "Direction") +
  ggplot2::scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.35), guide = "none") +
  ggplot2::scale_y_continuous(
    limits = c(
      min(cor_tbl$spearman_rho, na.rm = TRUE) - 0.18,
      max(cor_tbl$spearman_rho, na.rm = TRUE) + 0.18
    ),
    breaks = scales::pretty_breaks(n = 6)
  ) +
  ggplot2::labs(
    title = "C1-Core-associated MOFA factors",
    subtitle = "Spearman association between MOFA latent factors and continuous C1-Core score",
    x = NULL,
    y = expression("Spearman " * rho),
    caption = "Labels are shown only for FDR-significant associations."
  ) +
  ggpubr::theme_pubclean(base_size = 12) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(face = "bold", size = 14),
    plot.subtitle = ggplot2::element_text(size = 11, colour = "grey25"),
    plot.caption = ggplot2::element_text(size = 9, colour = "grey35", hjust = 0),
    axis.text.y = ggplot2::element_text(face = "bold", colour = "black"),
    axis.text.x = ggplot2::element_text(colour = "black"),
    axis.title.x = ggplot2::element_text(face = "bold"),
    legend.position = "right",
    legend.title = ggplot2::element_text(face = "bold"),
    plot.margin = ggplot2::margin(8, 35, 8, 8)
  )

p_cor_bar

save_plot_png(
  p_cor_bar,
  "MOFA_C1_factor_correlation_barplot.png",
  width = 10,
  height = 6
)

```

### Top-Factor Correlation Matrix

```{r}
#| label: c1-core-ggpairs

library(GGally)

top_factors <- cor_tbl |>
  dplyr::slice_max(abs_spearman_rho, n = 4) |>
  dplyr::pull(factor)

c1_mofa_corr_tbl <- factor_c1_df |>
  dplyr::select(
    sample,
    C1_Core,
    C1_Core_group,
    dplyr::all_of(top_factors)
  )

p_c1_mofa_ggpairs <- GGally::ggpairs(
  data = c1_mofa_corr_tbl,
  columns = c("C1_Core", top_factors),
  mapping = ggplot2::aes(color = C1_Core_group, alpha = 0.75),
  lower = list(
    continuous = GGally::wrap(
      "smooth_loess",
      se = TRUE,
      color = "lightblue",
      alpha = 0.5
    )
  ),
  upper = list(
    continuous = GGally::wrap(
      "cor",
      size = 5,
      method = "spearman"
    )
  ),
  diag = list(
    continuous = GGally::wrap("densityDiag")
  )
) +
  ggplot2::theme_bw(base_size = 13) +
  ggplot2::theme(
    strip.background = ggplot2::element_rect(fill = "grey90", color = NA)
  )

p_c1_mofa_ggpairs

save_plot_png(
  p_c1_mofa_ggpairs,
  "MOFA_C1_top_factor_ggpairs.png",
  width = 20,
  height = 10
)

```

### Top-Factor Scatterplots

```{r}
#| label: c1-core-top-factor-scatterplots

top_factors <- cor_tbl |>
  dplyr::slice_max(abs_spearman_rho, n = 6) |>
  dplyr::pull(factor)

plot_factor_c1 <- function(factor_name) {
  plot_df <- factor_c1_long |>
    dplyr::filter(factor == factor_name)
  
  cor_info <- cor_tbl |>
    dplyr::filter(factor == factor_name)
  
  ggplot2::ggplot(plot_df, ggplot2::aes(x = factor_value, y = C1_Core)) +
    ggplot2::geom_point(
      ggplot2::aes(color = C1_Core_group),
      size = 2.3,
      alpha = 0.85
    ) +
    ggplot2::geom_smooth(
      method = "lm",
      se = TRUE,
      linewidth = 0.8,
      color = "black"
    ) +
    ggplot2::geom_smooth(
      method = "gam",
      formula = y ~ s(x, bs = "cs"),
      se = FALSE,
      linewidth = 0.8,
      linetype = "dashed",
      color = "grey30"
    ) +
    ggpubr::stat_cor(method = "spearman")+
    ggplot2::theme_bw(base_size = 12) +
    ggplot2::labs(
      title = factor_name,
      x = "MOFA factor value",
      y = "C1-Core",
      color = "C1 group"
    )
}

factor_plots <- lapply(top_factors, plot_factor_c1)

p_factor_c1_patchwork <- patchwork::wrap_plots(factor_plots, ncol = 3) +
  patchwork::plot_annotation(
    title = "Top MOFA factors associated with C1-Core",
    subtitle = "Solid line = linear fit; dashed line = GAM smoother"
  ) + 
    plot_annotation(tag_levels = 'A') +
    plot_layout(guides = "collect") &
    theme(legend.position = "bottom")


p_factor_c1_patchwork

save_plot_png(
  p_factor_c1_patchwork,
  "MOFA_C1_top_factor_scatterplots.png",
  width = 20,
  height = 10
)

```

## MOFA Feature Weights

Inspect view-specific feature weights for selected MOFA factors.

### Weight Plot Helpers

```{r}
#| label: define-mofa-weight-plot-helper
#| echo: false
#| warning: false
#| message: false
#| results: hide

plot_mofa_weights <- function(view_id, factor_id, n_features = 20, save_png = TRUE) {
  p <- MOFA2::plot_weights(
    MOFAobject,
    view = view_id,
    factor = factor_id,
    nfeatures = n_features,
    scale = TRUE
  ) +
    ggplot2::labs(
      title = paste0("MOFA Factor ", factor_id, " weights: ", view_id),
      x = "Feature",
      y = "Scaled weight"
    )
  
  if (save_png) {
    save_plot_png(
      p,
      paste0("MOFA_weights_", view_id, "_factor", factor_id, ".png"),
      width = 12,
      height = 8
    )
  }
  
  p
}

```

### C1-Associated Factor Weights

```{r}
#| label: plot-mofa-feature-weights
#| echo: false
#| warning: false
#| message: false
#| fig-align: center

dir.create("output/figures/mofa_weights", recursive = TRUE, showWarnings = FALSE)

annot <- plot_annotation(
  tag_levels = "A",
  title = "MOFA weights for C1-core-associated factors",
  subtitle = "Factors 1 positively correlate with C1-core; factors 10 negatively correlate with C1-core"
)

# mutation
a <- plot_mofa_weights(view_id = "mutation", factor_id = 1, n_features = 20, save_png = FALSE)
b <- plot_mofa_weights(view_id = "mutation", factor_id = 10, n_features = 20, save_png = FALSE)

comb_mut <- ((a + b)) + annot

ggsave(
  filename = "output/figures/mofa_weights/MOFA_weights_mutation.png",
  plot = comb_mut,
  width = 12,
  height = 8,
  dpi = 600,
  bg = "white"
)


# RNA
a <- plot_mofa_weights(view_id = "RNA", factor_id = 1, n_features = 20, save_png = FALSE)
b <- plot_mofa_weights(view_id = "RNA", factor_id = 10, n_features = 20, save_png = FALSE)

comb_rna <- ((a + b)) + annot

ggsave(
  filename = "output/figures/mofa_weights/MOFA_weights_RNA.png",
  plot = comb_rna,
  width = 12,
  height = 8,
  dpi = 600,
  bg = "white"
)


# miRNA
a <- plot_mofa_weights(view_id = "miRNA", factor_id = 1, n_features = 20, save_png = FALSE)
b <- plot_mofa_weights(view_id = "miRNA", factor_id = 10, n_features = 20, save_png = FALSE)

comb_mirna <- ((a + b)) + annot

ggsave(
  filename = "output/figures/mofa_weights/MOFA_weights_miRNA.png",
  plot = comb_mirna,
  width = 12,
  height = 8,
  dpi = 600,
  bg = "white"
)


# RPPA
a <- plot_mofa_weights(view_id = "RPPA", factor_id = 1, n_features = 20, save_png = FALSE)
b <- plot_mofa_weights(view_id = "RPPA", factor_id = 10, n_features = 20, save_png = FALSE)

comb_rppa <- ((a + b)) + annot


ggsave(
  filename = "output/figures/mofa_weights/MOFA_weights_RPPA.png",
  plot = comb_rppa,
  width = 12,
  height = 8,
  dpi = 600,
  bg = "white"
)

```

### Top Weights Across Views

```{r}
#| label: define-mofa-top-weight-plot-helper
#| echo: false
#| warning: false
#| message: false
#| results: hide

plot_mofa_top_weights <- function(view_id, factor_id, n_features = 20, save_png = TRUE) {
  p <- MOFA2::plot_top_weights(
    MOFAobject,
    view = view_id,
    factor = factor_id,
    nfeatures = n_features,
    scale = TRUE
  ) +
    ggplot2::labs(
      title = paste0("MOFA Factor ", factor_id, " top weights: ", view_id),
      x = "Feature",
      y = "Scaled weight"
    )
  
  if (save_png) {
    save_plot_png(
      p,
      paste0("MOFA_top_weights_", view_id, "_factor", factor_id, ".png"),
      width = 6,
      height = 4
    )
  }
  
  p
}

```

### Selected Top-Weight Plots

```{r}
#| label: plot-mofa-top-feature-weights
#| echo: false
#| warning: false
#| message: false
#| fig-width: 6
#| fig-height: 4
#| fig-align: center

dir.create("output/figures/mofa_weights", recursive = TRUE, showWarnings = FALSE)

annot <- plot_annotation(
  tag_levels = "A",
  title = "MOFA weights for C1-core-associated factors",
  subtitle = "Factors 1 positively correlate with C1-core; factors 10 negatively correlate with C1-core"
)

# mutation
a <- plot_mofa_top_weights(view_id = "mutation", factor_id = 1, n_features = 50, save_png = FALSE)
b <- plot_mofa_top_weights(view_id = "mutation", factor_id = 10, n_features = 50, save_png = FALSE)

comb_mut <- ((a + b)) + annot

ggsave(
  filename = "output/figures/mofa_weights/MOFA_top_weights_mutation.png",
  plot = comb_mut,
  width = 16,
  height = 14,
  dpi = 600,
  bg = "white"
)


# RNA
a <- plot_mofa_top_weights(view_id = "RNA", factor_id = 1, n_features = 50, save_png = FALSE)
b <- plot_mofa_top_weights(view_id = "RNA", factor_id = 10, n_features = 50, save_png = FALSE)

comb_rna <- ((a + b)) + annot

ggsave(
  filename = "output/figures/mofa_weights/MOFA_top_weights_RNA.png",
  plot = comb_rna,
  width = 16,
  height = 14,
  dpi = 600,
  bg = "white"
)


# miRNA
a <- plot_mofa_top_weights(view_id = "miRNA", factor_id = 1, n_features = 50, save_png = FALSE)
b <- plot_mofa_top_weights(view_id = "miRNA", factor_id = 10, n_features = 50, save_png = FALSE)

comb_mirna <- ((a + b)) + annot

ggsave(
  filename = "output/figures/mofa_weights/MOFA_top_weights_miRNA.png",
  plot = comb_mirna,
  width = 16,
  height = 14,
  dpi = 600,
  bg = "white"
)


# RPPA
a <- plot_mofa_top_weights(view_id = "RPPA", factor_id = 1, n_features = 50, save_png = FALSE)
b <- plot_mofa_top_weights(view_id = "RPPA", factor_id = 10, n_features = 50, save_png = FALSE)

comb_rppa <- ((a + b)) + annot

ggsave(
  filename = "output/figures/mofa_weights/MOFA_top_weights_RPPA.png",
  plot = comb_rppa,
  width = 16,
  height = 14,
  dpi = 600,
  bg = "white"
)

```

### Factor Distributions by C1-Core Group

```{r}
#| label: plot-factor-violin-by-c1-core
#| echo: false
#| warning: false
#| message: false

violin_factor <- plot_factor(
  MOFAobject, 
  add_boxplot = TRUE,
  factors = c(1, 10),
  color_by = "C1_Core_group",
  dot_size = 2,
  dodge = TRUE,
  legend = TRUE,
  add_violin = TRUE,
  violin_alpha = 0.25
) +
  stat_compare_means() +
  scale_fill_npg() +
  labs(title = "MOFA factor weights")

violin_factor

save_plot_png(
  violin_factor,
  filename = "MOFA_violin_factor.png",
  width = 12,
  height = 8
)

```

### Heatmap Helper

Plot row-scaled top-weight features for selected views and factors, ordered by C1-core group.

```{r}
#| label: custom-complexheatmap
#| echo: false
#| warning: false
#| message: false
#| results: hide

library(MOFA2)
library(dplyr)
library(tibble)
library(ComplexHeatmap)
library(circlize)
library(viridis)
library(grid)

plot_mofa_view_heatmap <- function(
  MOFAobject,
  factor_id = 1,
  view_id = "RNA",
  group_id = "group1",
  n_features = 100,
  feature_selection = "balanced",
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  clustering_method = "average",
  save_png = TRUE,
  width = 14,
  height = 10
) {
  row_zscore <- function(mat) {
    mat <- as.matrix(mat)

    row_median <- apply(mat, 1, median, na.rm = TRUE)
    row_mad <- apply(mat, 1, mad, na.rm = TRUE)

    row_mad[is.na(row_mad) | row_mad == 0] <- 1

    mat_scaled <- sweep(mat, 1, row_median, "-")
    mat_scaled <- sweep(mat_scaled, 1, row_mad, "/")
    mat_scaled[is.na(mat_scaled)] <- 0

    mat_scaled
  }

  pearson_row_hclust <- function(mat) {
    cor_mat <- cor(
      t(mat),
      method = "pearson",
      use = "pairwise.complete.obs"
    )

    cor_mat[is.na(cor_mat)] <- 0

    hclust(
      as.dist(1 - cor_mat),
      method = clustering_method
    )
  }

  pearson_column_hclust <- function(mat) {
    cor_mat <- cor(
      mat,
      method = "pearson",
      use = "pairwise.complete.obs"
    )

    cor_mat[is.na(cor_mat)] <- 0

    hclust(
      as.dist(1 - cor_mat),
      method = clustering_method
    )
  }

  data_list <- get_data(
    MOFAobject,
    views = view_id,
    as.data.frame = FALSE
  )

  view_mat <- data_list[[view_id]][[group_id]]
  mofa_samples <- samples_names(MOFAobject)[[group_id]]

  if (all(mofa_samples %in% rownames(view_mat))) {
    view_mat <- t(view_mat)
  }

  if (!all(mofa_samples %in% colnames(view_mat))) {
    stop("MOFA samples are not in the columns of the selected view matrix.")
  }

  weights_df <- get_weights(
    MOFAobject,
    views = view_id,
    factors = factor_id,
    as.data.frame = TRUE
  )

  weights_view <- weights_df |>
    dplyr::filter(view == view_id) |>
    dplyr::mutate(abs_weight = abs(value))

  if (!feature_selection %in% c("balanced", "absolute")) {
    stop("feature_selection must be either 'balanced' or 'absolute'.")
  }

  if (feature_selection == "absolute") {
    top_features <- weights_view |>
      dplyr::arrange(dplyr::desc(abs_weight)) |>
      dplyr::slice_head(n = n_features) |>
      dplyr::pull(feature)
  }

  if (feature_selection == "balanced") {
    n_negative <- floor(n_features / 2)
    n_positive <- n_features - n_negative

    top_negative_features <- weights_view |>
      dplyr::filter(value < 0) |>
      dplyr::arrange(value) |>
      dplyr::slice_head(n = n_negative) |>
      dplyr::pull(feature)

    top_positive_features <- weights_view |>
      dplyr::filter(value > 0) |>
      dplyr::arrange(dplyr::desc(value)) |>
      dplyr::slice_head(n = n_positive) |>
      dplyr::pull(feature)

    top_features <- c(
      top_negative_features,
      top_positive_features
    )
  }

  top_features <- intersect(top_features, rownames(view_mat))

  if (length(top_features) == 0) {
    stop("No top features matched the rownames of the selected view matrix.")
  }

  metadata_mofa <- samples_metadata(MOFAobject) |>
    tibble::as_tibble() |>
    dplyr::mutate(
      sample = as.character(sample),
      C1_Core_group = factor(
        C1_Core_group,
        levels = c(
          "Not-C1-like",
          "Not-C1-intermediate",
          "C1-intermediate",
          "C1-like"
        )
      ),
      purity_annotation = dplyr::coalesce(
        as.numeric(purity_primary),
        as.numeric(purity_cpe),
        as.numeric(purity_absolute),
        as.numeric(purity_estimate),
        as.numeric(purity_ihc)
      )
    )

  sample_order <- metadata_mofa |>
    dplyr::filter(sample %in% colnames(view_mat)) |>
    dplyr::arrange(C1_Core_group, C1_Core) |>
    dplyr::pull(sample)

  if (length(sample_order) == 0) {
    stop("No MOFA samples matched the metadata sample column.")
  }

  heat_mat <- view_mat[top_features, sample_order, drop = FALSE]
  heat_mat_scaled <- row_zscore(heat_mat)

  annotation_df <- metadata_mofa |>
    dplyr::filter(sample %in% sample_order) |>
    dplyr::arrange(match(sample, sample_order)) |>
    dplyr::select(
      C1_Core_group,
      C1_Core,
      purity_annotation
    ) |>
    as.data.frame()

  colnames(annotation_df) <- c(
    "C1 group",
    "C1 score",
    "Purity"
  )

  rownames(annotation_df) <- sample_order

  group_colors <- c(
    "Not-C1-like" = "#4D4D4D",
    "Not-C1-intermediate" = "#A6A6A6",
    "C1-intermediate" = "#FDB863",
    "C1-like" = "#D73027"
  )

  c1_score_col_fun <- circlize::colorRamp2(
    c(
      min(annotation_df[["C1 score"]], na.rm = TRUE),
      0,
      max(annotation_df[["C1 score"]], na.rm = TRUE)
    ),
    c("#2166AC", "white", "#B2182B")
  )

  purity_col_fun <- circlize::colorRamp2(
    c(
      min(annotation_df[["Purity"]], na.rm = TRUE),
      median(annotation_df[["Purity"]], na.rm = TRUE),
      max(annotation_df[["Purity"]], na.rm = TRUE)
    ),
    c("#542788", "white", "#238443")
  )

  ha_col <- ComplexHeatmap::HeatmapAnnotation(
    df = annotation_df,
    col = list(
      "C1 group" = group_colors,
      "C1 score" = c1_score_col_fun,
      "Purity" = purity_col_fun
    ),
    annotation_name_gp = grid::gpar(fontsize = 9),
    na_col = "grey85"
  )

  ht <- ComplexHeatmap::Heatmap(
    heat_mat_scaled,
    name = "row z-score",
    col = circlize::colorRamp2(
      c(-2, 0, 2),
      viridis::viridis(3)
    ),
    use_raster = TRUE,
    top_annotation = ha_col,

    cluster_rows = if (cluster_rows) pearson_row_hclust else FALSE,
    cluster_columns = if (cluster_columns) pearson_column_hclust else FALSE,

    column_split = annotation_df[["C1 group"]],
    cluster_column_slices = FALSE,

    show_row_names = TRUE,
    show_column_names = FALSE,

    row_names_gp = grid::gpar(fontsize = 8),

    column_title = paste0(
      "MOFA Factor ", factor_id,
      ": ", feature_selection, " top ", length(top_features), " ",
      view_id, " features"
    ),
    column_title_gp = grid::gpar(fontsize = 12, fontface = "bold"),

    heatmap_legend_param = list(
      title = "Feature\nz-score"
    )
  )

  ComplexHeatmap::draw(
    ht,
    heatmap_legend_side = "right",
    annotation_legend_side = "right"
  )

  if (save_png) {
    png(
      filename = file.path(
        FIG_DIR,
        paste0(
          "MOFA_complexheatmap_",
          view_id,
          "_factor",
          factor_id,
          "_",
          feature_selection,
          "_top",
          length(top_features),
          "_C1_purity.png"
        )
      ),
      width = width,
      height = height,
      units = "in",
      res = FIG_DPI
    )

    ComplexHeatmap::draw(
      ht,
      heatmap_legend_side = "right",
      annotation_legend_side = "right"
    )

    dev.off()
  }

  invisible(ht)
}

```

### miRNA Heatmaps

```{r}
#| label: plot-mirna-mofa-heatmaps
#| echo: false
#| warning: false
#| message: false

view_id <- "miRNA"

plot_mofa_view_heatmap(
  MOFAobject = MOFAobject,
  factor_id = 1,
  view_id = view_id,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  clustering_method = "average"
)


plot_mofa_view_heatmap(
  MOFAobject = MOFAobject,
  factor_id = 10,
  view_id = view_id,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  clustering_method = "average"
)


```


### RNA Heatmaps

```{r}
#| label: plot-rna-mofa-heatmaps
#| echo: false
#| warning: false
#| message: false

view_id <- "RNA"

plot_mofa_view_heatmap(
  MOFAobject = MOFAobject,
  factor_id = 1,
  n_features = 100,
  view_id = view_id,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  clustering_method = "average"
)

plot_mofa_view_heatmap(
  MOFAobject = MOFAobject,
  factor_id = 10,
  view_id = view_id,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  clustering_method = "average"
)

```


### mutation Heatmaps

```{r}
#| label: plot-mutation-mofa-heatmaps
#| echo: false
#| warning: false
#| message: false

view_id <- "mutation"

plot_mofa_view_heatmap(
  MOFAobject = MOFAobject,
  factor_id = 1,
  view_id = view_id,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  clustering_method = "average"
)

plot_mofa_view_heatmap(
  MOFAobject = MOFAobject,
  factor_id = 10,
  view_id = view_id,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  clustering_method = "average"
)

```


### RPPA Heatmaps

```{r}
#| label: plot-RPPA-mofa-heatmaps
#| echo: false
#| warning: false
#| message: false

view_id <- "RPPA"

plot_mofa_view_heatmap(
  MOFAobject = MOFAobject,
  factor_id = 1,
  view_id = view_id,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  clustering_method = "average"
)


plot_mofa_view_heatmap(
  MOFAobject = MOFAobject,
  factor_id = 10,
  view_id = view_id,
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  clustering_method = "average"
)


```

# Expansion Space

The following section anchors are intentionally empty and can be expanded later without disturbing the core MOFA workflow.

```{r}

library(MOFA2)
library(dplyr)
library(tibble)
library(readr)
library(matrixStats)

# Cleaning MOFA feature names
clean_mofa_feature_name <- function(x) {
  sub("_(RNA|RPPA|miRNA|mutation)$", "", x)
}


# Extracting MOFA weights and C1/Not-C1 medians for RNA, miRNA, and RPPA
extract_mofa_weights_with_group_medians <- function(
  MOFAobject,
  mofa_data,
  factor_id = 1,
  view_id = "RNA",
  feature_col = "feature",
  c1_group_col = "C1_Core_group",
  c1_label = "C1-like",
  not_c1_label = "Not-C1-like"
) {
  view_mat <- mofa_data[[view_id]]

  if (is.null(view_mat)) {
    stop("The selected view_id is not present in mofa_data.")
  }

  view_mat <- as.matrix(view_mat)
  storage.mode(view_mat) <- "numeric"

  metadata_tbl <- MOFA2::samples_metadata(MOFAobject) |>
    tibble::as_tibble() |>
    dplyr::mutate(sample = as.character(sample))

  if (!c1_group_col %in% colnames(metadata_tbl)) {
    stop("The selected C1 group column is not present in sample metadata.")
  }

  c1_samples <- metadata_tbl |>
    dplyr::filter(.data[[c1_group_col]] == c1_label) |>
    dplyr::pull(sample) |>
    intersect(colnames(view_mat))

  not_c1_samples <- metadata_tbl |>
    dplyr::filter(.data[[c1_group_col]] == not_c1_label) |>
    dplyr::pull(sample) |>
    intersect(colnames(view_mat))

  if (length(c1_samples) == 0) {
    stop("No C1-like samples matched the selected omics matrix.")
  }

  if (length(not_c1_samples) == 0) {
    stop("No Not-C1-like samples matched the selected omics matrix.")
  }

  weights_tbl <- MOFA2::get_weights(
    MOFAobject,
    views = view_id,
    factors = factor_id,
    as.data.frame = TRUE
  ) |>
    dplyr::filter(view == view_id) |>
    dplyr::transmute(
      feature_original = feature,
      feature = clean_mofa_feature_name(feature),
      weight = value
    )

  median_tbl <- tibble::tibble(
    feature = rownames(view_mat),
    median_C1 = matrixStats::rowMedians(
      view_mat[, c1_samples, drop = FALSE],
      na.rm = TRUE
    ),
    median_NotC1 = matrixStats::rowMedians(
      view_mat[, not_c1_samples, drop = FALSE],
      na.rm = TRUE
    )
  ) |>
    dplyr::mutate(
      delta_median_C1_minus_NotC1 = median_C1 - median_NotC1
    )

  weights_tbl |>
    dplyr::left_join(
      median_tbl,
      by = "feature"
    ) |>
    dplyr::mutate(
      agreement = dplyr::case_when(
        weight > 0 & delta_median_C1_minus_NotC1 > 0 ~ "Concordant",
        weight < 0 & delta_median_C1_minus_NotC1 < 0 ~ "Concordant",
        weight > 0 & delta_median_C1_minus_NotC1 < 0 ~ "Discordant",
        weight < 0 & delta_median_C1_minus_NotC1 > 0 ~ "Discordant",
        TRUE ~ "Neutral"
      ),
      association = dplyr::case_when(
        weight > 0 ~ "C1",
        weight < 0 ~ "NotC1",
        TRUE ~ "Neutral"
      ),
      agreement = factor(
        agreement,
        levels = c("Concordant", "Discordant", "Neutral")
      )
    ) |>
    dplyr::arrange(
      agreement,
      dplyr::desc(abs(weight))
    ) |>
    dplyr::select(
      feature_original,
      feature,
      weight,
      median_C1,
      median_NotC1,
      delta_median_C1_minus_NotC1,
      association,
      agreement
    ) |>
    dplyr::rename(
      !!feature_col := feature
    )
}


# Extracting MOFA mutation weights and C1/Not-C1 frequencies
extract_mofa_mutation_weights_with_group_frequency <- function(
  MOFAobject,
  mofa_data,
  factor_id = 1,
  view_id = "mutation",
  c1_group_col = "C1_Core_group",
  c1_label = "C1-like",
  not_c1_label = "Not-C1-like"
) {
  mutation_mat <- mofa_data[[view_id]]

  if (is.null(mutation_mat)) {
    stop("The selected mutation view is not present in mofa_data.")
  }

  mutation_mat <- as.matrix(mutation_mat)
  storage.mode(mutation_mat) <- "numeric"

  metadata_tbl <- MOFA2::samples_metadata(MOFAobject) |>
    tibble::as_tibble() |>
    dplyr::mutate(sample = as.character(sample))

  if (!c1_group_col %in% colnames(metadata_tbl)) {
    stop("The selected C1 group column is not present in sample metadata.")
  }

  c1_samples <- metadata_tbl |>
    dplyr::filter(.data[[c1_group_col]] == c1_label) |>
    dplyr::pull(sample) |>
    intersect(colnames(mutation_mat))

  not_c1_samples <- metadata_tbl |>
    dplyr::filter(.data[[c1_group_col]] == not_c1_label) |>
    dplyr::pull(sample) |>
    intersect(colnames(mutation_mat))

  if (length(c1_samples) == 0) {
    stop("No C1-like samples matched the mutation matrix.")
  }

  if (length(not_c1_samples) == 0) {
    stop("No Not-C1-like samples matched the mutation matrix.")
  }

  weights_tbl <- MOFA2::get_weights(
    MOFAobject,
    views = view_id,
    factors = factor_id,
    as.data.frame = TRUE
  ) |>
    dplyr::filter(view == view_id) |>
    dplyr::transmute(
      mutation_original = feature,
      mutation = clean_mofa_feature_name(feature),
      weight = value
    )

  frequency_tbl <- tibble::tibble(
    mutation = rownames(mutation_mat),
    frequency_C1 = matrixStats::rowMeans2(
      mutation_mat[, c1_samples, drop = FALSE] > 0,
      na.rm = TRUE
    ),
    frequency_NotC1 = matrixStats::rowMeans2(
      mutation_mat[, not_c1_samples, drop = FALSE] > 0,
      na.rm = TRUE
    )
  ) |>
    dplyr::mutate(
      delta_frequency_C1_minus_NotC1 = frequency_C1 - frequency_NotC1
    )

  weights_tbl |>
    dplyr::left_join(
      frequency_tbl,
      by = "mutation"
    ) |>
    dplyr::mutate(
      agreement = dplyr::case_when(
        weight > 0 & delta_frequency_C1_minus_NotC1 > 0 ~ "Concordant",
        weight < 0 & delta_frequency_C1_minus_NotC1 < 0 ~ "Concordant",
        weight > 0 & delta_frequency_C1_minus_NotC1 < 0 ~ "Discordant",
        weight < 0 & delta_frequency_C1_minus_NotC1 > 0 ~ "Discordant",
        TRUE ~ "Neutral"
      ),
      association = dplyr::case_when(
        weight > 0 ~ "C1",
        weight < 0 ~ "NotC1",
        TRUE ~ "Neutral"
      ),
      agreement = factor(
        agreement,
        levels = c("Concordant", "Discordant", "Neutral")
      )
    ) |>
    dplyr::arrange(
      agreement,
      dplyr::desc(abs(weight))
    ) |>
    dplyr::select(
      mutation_original,
      mutation,
      weight,
      frequency_C1,
      frequency_NotC1,
      delta_frequency_C1_minus_NotC1,
      association,
      agreement
    )
}

```

```{r}

TABLE_DIR <- "output/tables"
dir.create(TABLE_DIR, recursive = TRUE, showWarnings = FALSE)


# RNA
rna_factor1_weights_medians <- extract_mofa_weights_with_group_medians(
  MOFAobject = MOFAobject,
  mofa_data = mofa_data,
  factor_id = 1,
  view_id = "RNA",
  feature_col = "gene"
)

readr::write_tsv(
  rna_factor1_weights_medians,
  file.path(TABLE_DIR, "MOFA_factor1_RNA_weights_C1_NotC1_medians.tsv")
)


# miRNA
mirna_factor1_weights_medians <- extract_mofa_weights_with_group_medians(
  MOFAobject = MOFAobject,
  mofa_data = mofa_data,
  factor_id = 1,
  view_id = "miRNA",
  feature_col = "miRNA"
)

readr::write_tsv(
  mirna_factor1_weights_medians,
  file.path(TABLE_DIR, "MOFA_factor1_miRNA_weights_C1_NotC1_medians.tsv")
)


# RPPA
protein_factor1_weights_medians <- extract_mofa_weights_with_group_medians(
  MOFAobject = MOFAobject,
  mofa_data = mofa_data,
  factor_id = 1,
  view_id = "RPPA",
  feature_col = "protein"
)

readr::write_tsv(
  protein_factor1_weights_medians,
  file.path(TABLE_DIR, "MOFA_factor1_RPPA_weights_C1_NotC1_medians.tsv")
)


# Mutation
mutation_factor1_weights_frequency <- extract_mofa_mutation_weights_with_group_frequency(
  MOFAobject = MOFAobject,
  mofa_data = mofa_data,
  factor_id = 1,
  view_id = "mutation"
)

readr::write_tsv(
  mutation_factor1_weights_frequency,
  file.path(TABLE_DIR, "MOFA_factor1_mutation_weights_C1_NotC1_frequency.tsv")
)

```

```{r}

rna_factor1_weights_medians |>
  dplyr::filter(agreement == "Concordant") |>
  dplyr::filter(weight > 0) |>
  dplyr::slice_max(abs(weight), n = 50)

```

## Clinical Metadata Associations

## Survival Associations

## Immune Marker Associations

## Factor-Specific View Dissection

## Export Tables
