---
title: "Create a Paired RNA/ATAC Object"
format:
  html:
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
---

> **Status:** `draft`<br>
> **Template class:** `SOURCE-BACKED WORKFLOW`

## Purpose

Create one Seurat object with matched RNA and Signac ATAC assays. This page
handles object construction and cell matching only; modality-specific QC and
reductions belong to their respective canonical templates.

## Inputs and parameters

The supported contract is one qs2 R matrix for RNA, one qs2 R matrix for ATAC
peaks, a metadata TSV with `cell_id`, a tabix-indexed fragments file, and a qs2
gene-annotation object. Both matrices are feature-by-cell.

```{r}
suppressPackageStartupMessages({
  library(Seurat)
  library(SeuratObject)
  library(Signac)
  library(qs2)
  library(readr)
  library(tibble)
})

rna_counts_path <- "input/rna_counts.qs2"
atac_counts_path <- "input/atac_peak_counts.qs2"
metadata_path <- "input/cell_metadata.tsv"
fragments_path <- "input/fragments.tsv.gz"
annotation_path <- "input/gene_annotation.qs2"
output_object <- "output/multiome_created.qs2"
alignment_table <- "output/multiome_cell_alignment.tsv"

rna_assay <- "RNA"
atac_assay <- "ATAC"
genome_build <- "hg38"
peak_separator <- c(":", "-")
min_cells <- 1L
# Preserve all deliberately matched cells; filter cells in the QC stage.
min_features <- 0L
```

The example `hg38` build must be replaced when the data use another valid
build. Peak coordinates, fragments, annotations, and downstream sequence
resources must use compatible builds and chromosome naming.

## Analysis

```{r}
rna_counts <- qs2::qs_read(file = rna_counts_path)
atac_counts <- qs2::qs_read(file = atac_counts_path)
metadata <- readr::read_tsv(metadata_path, show_col_types = FALSE)
annotation <- qs2::qs_read(file = annotation_path)

stopifnot(is.matrix(rna_counts) || inherits(rna_counts, "Matrix"))
stopifnot(is.matrix(atac_counts) || inherits(atac_counts, "Matrix"))
stopifnot(!is.null(colnames(rna_counts)), !is.null(colnames(atac_counts)))
stopifnot(!anyDuplicated(colnames(rna_counts)), !anyDuplicated(colnames(atac_counts)))
stopifnot(!anyDuplicated(rownames(rna_counts)), !anyDuplicated(rownames(atac_counts)))
stopifnot("cell_id" %in% colnames(metadata), !anyDuplicated(metadata$cell_id))
stopifnot(file.exists(fragments_path), file.exists(paste0(fragments_path, ".tbi")))

all_cells <- unique(c(colnames(rna_counts), colnames(atac_counts), metadata$cell_id))
alignment <- tibble::tibble(
  cell_id = all_cells,
  in_rna = cell_id %in% colnames(rna_counts),
  in_atac = cell_id %in% colnames(atac_counts),
  in_metadata = cell_id %in% metadata$cell_id
)
readr::write_tsv(alignment, alignment_table)

shared_cells <- all_cells[alignment$in_rna & alignment$in_atac & alignment$in_metadata]
stopifnot(length(shared_cells) > 0)
rna_counts <- rna_counts[, shared_cells, drop = FALSE]
atac_counts <- atac_counts[, shared_cells, drop = FALSE]
metadata <- metadata[match(shared_cells, metadata$cell_id), , drop = FALSE]
rownames(metadata) <- metadata$cell_id

fragment_object <- Signac::CreateFragmentObject(
  path = fragments_path,
  cells = shared_cells,
  validate.fragments = TRUE
)
chromatin_assay <- Signac::CreateChromatinAssay(
  counts = atac_counts,
  sep = peak_separator,
  genome = genome_build,
  annotation = annotation,
  fragments = fragment_object,
  min.cells = min_cells,
  min.features = min_features,
  validate.fragments = TRUE
)

object <- Seurat::CreateSeuratObject(
  counts = rna_counts,
  assay = rna_assay,
  meta.data = metadata,
  min.features = min_features
)
object[[atac_assay]] <- chromatin_assay
DefaultAssay(object) <- rna_assay

stopifnot(identical(colnames(object), shared_cells))
stopifnot(inherits(object[[atac_assay]], "ChromatinAssay"))
stopifnot(length(Signac::Fragments(object[[atac_assay]])) > 0)

readr::write_tsv(
  tibble::tibble(cells = ncol(object), genes = nrow(object[[rna_assay]]),
                 peaks = nrow(object[[atac_assay]]), genome_build = genome_build),
  "output/multiome_created_summary.tsv"
)
qs2::qs_save(object, file = output_object)
```

## Method notes

Paired multiome data require deliberate matching of cells across modalities.
The alignment TSV records cells missing from RNA, ATAC, or metadata; the code
intersects them and never silently unions unmatched barcodes. Fragments must be
indexed and contain the retained barcodes. Construction keeps matched cells;
cell filtering belongs to the QC stage. No modality QC is hidden here.
