---
title: "RNA-to-ATAC Label Transfer"
format:
  html:
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
---

> **Status:** `draft`<br>
> **Template class:** `SOURCE-BACKED WORKFLOW`

## Purpose

Transfer labels from an RNA reference to an ATAC query by creating a
GeneActivity assay in the query and finding anchors in shared gene space. The
reference/query roles are directional. This is label transfer, not generic
batch integration.

::: {.callout-warning}
## GeneActivity is not measured RNA

GeneActivity is an accessibility-derived proxy summarized in gene space. It is
useful for mapping an ATAC query to an RNA reference, but it is not observed
RNA expression and should not be interpreted as replacing the RNA assay.
:::

## Parameters

```{r}
suppressPackageStartupMessages({
  library(Seurat)
  library(SeuratObject)
  library(Signac)
  library(qs2)
  library(readr)
  library(tibble)
})

reference_object_path <- "input/rna_reference.qs2"
query_object_path <- "input/atac_query.qs2"
output_query_object <- "output/atac_query_label_transferred.qs2"
prediction_table <- "output/atac_label_transfer_predictions.tsv"

reference_assay <- "RNA"
query_atac_assay <- "ATAC"
activity_assay <- "ACTIVITY"
reference_label_column <- "cell_type"
query_atac_reduction <- "lsi"
anchor_reduction <- "rpca"
anchor_dims <- 1:30
weight_dims <- 1:39
```

## Analysis

```{r}
reference <- qs2::qs_read(file = reference_object_path)
query <- qs2::qs_read(file = query_object_path)
stopifnot(inherits(reference, "Seurat"), inherits(query, "Seurat"))
stopifnot(reference_assay %in% Assays(reference), query_atac_assay %in% Assays(query))
stopifnot(reference_label_column %in% colnames(reference[[]]))
stopifnot(query_atac_reduction %in% Reductions(query))
stopifnot(anchor_reduction == "rpca")
stopifnot("pca" %in% Reductions(reference))
stopifnot(max(weight_dims) <= ncol(Embeddings(query, query_atac_reduction)))
stopifnot(!activity_assay %in% Assays(query))

transfer_features <- VariableFeatures(reference, assay = reference_assay)
stopifnot(length(transfer_features) > 0)
gene_activities <- Signac::GeneActivity(
  query, assay = query_atac_assay, features = transfer_features
)
query[[activity_assay]] <- SeuratObject::CreateAssay5Object(counts = gene_activities)
query <- NormalizeData(query, assay = activity_assay)
query <- ScaleData(query, assay = activity_assay, features = rownames(query[[activity_assay]]))

shared_features <- intersect(transfer_features, rownames(query[[activity_assay]]))
stopifnot(length(shared_features) > 0)
anchors <- FindTransferAnchors(
  reference = reference,
  query = query,
  reference.assay = reference_assay,
  query.assay = activity_assay,
  features = shared_features,
  reduction = anchor_reduction,
  dims = anchor_dims
)

predictions <- TransferData(
  anchorset = anchors,
  refdata = setNames(reference[[]][[reference_label_column]], rownames(reference[[]])),
  query = query,
  weight.reduction = query[[query_atac_reduction]],
  dims = weight_dims
)
stopifnot(identical(rownames(predictions), colnames(query)))
query <- AddMetaData(query, metadata = predictions)
prediction_table_data <- tibble::rownames_to_column(as.data.frame(predictions), "cell_id")
readr::write_tsv(prediction_table_data, prediction_table)
qs2::qs_save(query, file = output_query_object)
```

## Diagnostics and method notes

The prediction table retains predicted labels and confidence columns returned
by the current Seurat API. Review score distributions, cell-level identities,
and reference/query compatibility; there is no universal acceptance threshold.
The source workflow used `reduction = "rpca"` and weighted transfer using the
query LSI reduction, which are retained explicitly here. RPCA anchors in this
context are a transfer choice, not a claim that the query was batch-integrated.
