---
title: "RNA–ATAC Peak–Gene Linkage"
format:
  html:
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
---

> **Status:** `draft`<br>
> **Template class:** `SOURCE-BACKED WORKFLOW`

## Purpose

Use Signac `LinkPeaks` to estimate statistical associations between RNA
expression and nearby ATAC peaks in a paired object. Links are regulatory
evidence for follow-up, not proof of enhancer causality.

## Parameters

```{r}
suppressPackageStartupMessages({
  library(Seurat)
  library(Signac)
  library(qs2)
  library(readr)
  library(tibble)
})

input_object <- "input/multiome_wnn.qs2"
gene_coordinates_path <- "input/gene_coordinates.qs2"
output_object <- "output/multiome_rna_atac_links.qs2"
links_table <- "output/multiome_rna_atac_links.tsv"
coverage_figure <- "output/multiome_coverage_links.pdf"

rna_assay <- "RNA"
atac_assay <- "ATAC"
genes_to_link <- NULL
link_distance <- 500000L
min_link_cells <- 10L
link_method <- "pearson"
coverage_region <- NULL
```

## Analysis

```{r}
object <- qs2::qs_read(file = input_object)
gene_coordinates <- qs2::qs_read(file = gene_coordinates_path)
stopifnot(inherits(object, "Seurat"))
stopifnot(all(c(rna_assay, atac_assay) %in% Assays(object)))
stopifnot(identical(colnames(object[[rna_assay]]), colnames(object[[atac_assay]])))
stopifnot(inherits(gene_coordinates, "GRanges"))

DefaultAssay(object) <- atac_assay
object <- Signac::LinkPeaks(
  object = object,
  peak.assay = atac_assay,
  expression.assay = rna_assay,
  expression.slot = "data",
  peak.slot = "counts",
  gene.coords = gene_coordinates,
  genes.use = genes_to_link,
  distance = link_distance,
  min.cells = min_link_cells,
  method = link_method
)
links <- Signac::Links(object[[atac_assay]])
link_table <- tibble::as_tibble(as.data.frame(links))
readr::write_tsv(link_table, links_table)

pdf(coverage_figure, width = 9, height = 4)
if (!is.null(coverage_region)) {
  print(Signac::CoveragePlot(
    object, region = coverage_region, assay = atac_assay,
    expression.assay = rna_assay
  ))
}
dev.off()
qs2::qs_save(object, file = output_object)
```

## Method notes

The distance window, gene coordinate object, expression/peak assays, and model
method are explicit. `LinkPeaks` relies on compatible genomic coordinates and
an appropriate expression representation; inspect its returned scores and
correlations rather than treating every nearby peak as linked. Set
`coverage_region` to a real region only after checking the coordinate build.
