---
title: "Paired RNA/ATAC Weighted Nearest Neighbors"
format:
  html:
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
---

> **Status:** `draft`<br>
> **Template class:** `SOURCE-BACKED WORKFLOW`

## Purpose

Build RNA and ATAC representations, combine them with Seurat weighted nearest
neighbors, and cluster/visualize the paired object. This does not repeat
fragment-aware QC.

::: {.callout-warning}
## WNN is not batch correction

Weighted nearest neighbors combine modalities for cell-cell neighborhood
construction. WNN does not, by itself, remove technical batch effects between
samples.
:::

## Parameters

```{r}
suppressPackageStartupMessages({
  library(Seurat)
  library(Signac)
  library(qs2)
  library(readr)
  library(tibble)
})

input_object <- "input/multiome_created.qs2"
output_object <- "output/multiome_wnn.qs2"
metadata_table <- "output/multiome_wnn_metadata.tsv"
cluster_table <- "output/multiome_wnn_clusters.tsv"

rna_assay <- "RNA"
atac_assay <- "ATAC"
rna_reduction <- "pca"
atac_reduction <- "lsi"
wnn_reduction <- "wnn.umap"
weighted_nn_name <- "weighted.nn"
knn_graph_name <- "wknn"
snn_graph_name <- "wsnn"
modality_weight_name <- "RNA.weight"
n_variable_features <- 2000L
n_pcs <- 30L
n_svd <- 50L
rna_dims <- 1:n_pcs
atac_dims <- 1:30
neighbors_k <- 20L
cluster_resolution <- 0.5
seed <- 1L
```

## Analysis

```{r}
object <- qs2::qs_read(file = input_object)
stopifnot(inherits(object, "Seurat"))
stopifnot(all(c(rna_assay, atac_assay) %in% Assays(object)))
stopifnot(identical(colnames(object[[rna_assay]]), colnames(object[[atac_assay]])))
set.seed(seed)

DefaultAssay(object) <- rna_assay
object <- NormalizeData(object, assay = rna_assay)
object <- FindVariableFeatures(object, assay = rna_assay, nfeatures = n_variable_features)
object <- ScaleData(object, assay = rna_assay, features = VariableFeatures(object, assay = rna_assay))
object <- RunPCA(object, assay = rna_assay, npcs = n_pcs, reduction.name = rna_reduction)

DefaultAssay(object) <- atac_assay
object <- Signac::RunTFIDF(object, assay = atac_assay)
object <- Signac::FindTopFeatures(object, assay = atac_assay, min.cutoff = "q0")
object <- Signac::RunSVD(object, assay = atac_assay, n = n_svd, reduction.name = atac_reduction)

object <- FindMultiModalNeighbors(
  object,
  reduction.list = list(rna_reduction, atac_reduction),
  dims.list = list(rna_dims, atac_dims),
  k.nn = neighbors_k,
  knn.graph.name = knn_graph_name,
  snn.graph.name = snn_graph_name,
  weighted.nn.name = weighted_nn_name,
  modality.weight.name = modality_weight_name
)
object <- RunUMAP(
  object,
  nn.name = weighted_nn_name,
  reduction.name = wnn_reduction,
  reduction.key = "wnnUMAP_",
  seed.use = seed
)
object <- FindClusters(
  object, graph.name = snn_graph_name, resolution = cluster_resolution,
  random.seed = seed
)

cluster_column <- paste0(snn_graph_name, "_res.", cluster_resolution)
readr::write_tsv(
  object[[]] |>
    tibble::rownames_to_column("cell_id"),
  metadata_table
)
readr::write_tsv(
  tibble::tibble(cluster = object[[]][[cluster_column]]) |>
    dplyr::count(cluster, name = "n_cells"),
  cluster_table
)
qs2::qs_save(object, file = output_object)
```

## Method notes

RNA PCA and ATAC LSI are distinct modality reductions. The dimension vectors,
neighbor/graph names, k, resolution, and seed are exposed because they affect
the neighborhood result. The default ATAC vector includes LSI1; inspect its
depth relationship before changing it.
