---
title: "scRNA Trajectory and Gene Programs"
lang: en
format:
  html:
    theme:
      light: cosmo
      dark: darkly
    respect-user-color-scheme: true
    toc: true
    toc-title: "Contents"
    toc-depth: 4
    toc-location: right
    toc-expand: 2
    number-sections: true
    number-depth: 3
    smooth-scroll: true
    anchor-sections: true
    code-fold: true
    code-summary: "Show analysis code"
    code-copy: true
    code-overflow: wrap
    code-line-numbers: false
    fig-width: 16
    fig-height: 12
    fig-dpi: 300
    fig-format: png
    fig-align: center
    fig-responsive: true
    fig-cap-location: bottom
    tbl-cap-location: top
    df-print: paged
    crossrefs-hover: true
    footnotes-hover: true
    citations-hover: true
    embed-resources: true
    self-contained-math: true
    page-layout: article
    grid:
      sidebar-width: 250px
      body-width: 1050px
      margin-width: 300px
      gutter-width: 1.5rem
    link-external-newwindow: true
    link-external-icon: true
lightbox:
  match: auto
  effect: fade
  desc-position: bottom
  loop: false
execute:
  echo: true
  warning: false
  message: false
  error: false
knitr:
  opts_chunk:
    fig.align: center
    out.width: "100%"
    collapse: true
    comment: "#>"
crossref:
  fig-title: "Figure"
  tbl-title: "Table"
  sec-prefix: "Section"
  fig-prefix: "Figure"
  tbl-prefix: "Table"
  ref-hyperlink: true
---

## Purpose

Fit trajectory-aware gene programs and store pseudotime summaries in a Seurat object.

::: {.callout-note}
## Archive status

This is a legacy notebook kept for reference. For current reusable workflows,
see the active templates in the repository.
:::

## Packages

Load the packages needed for this notebook.

```{r}
#| label: tpl-scrna-06-trajectory-gene-programs-packages
#| echo: true
#| warning: false
#| message: false
# Load only the packages used in this template.
suppressPackageStartupMessages({
  library(Seurat)
  library(qs2)
  library(readr)
  library(dplyr)
})
```

## Parameters

Set the parameters that control the analysis and output locations.

```{r}
#| label: tpl-scrna-06-trajectory-gene-programs-user-config
#| echo: true
#| warning: false
#| message: false
# USER INPUT:
input_object <- "path/to/seurat_clustered.qs"
output_object <- "path/to/seurat_trajectory_programs.qs"
results_dir <- "results/trajectory_gene_programs"
figures_dir <- "figures/trajectory_gene_programs"
assay_name <- "RNA"
reduction_name <- "umap"
sample_column <- "sample_column"
group_column <- "group_column"
pseudotime_column <- "pseudotime"
features_of_interest <- c("GENE1", "GENE2")
n_bins <- 20
seed <- 1
n_threads <- 4
compress_level <- 6
```

## Inputs

Load the input object or tables used by the workflow.

```{r}
#| label: tpl-scrna-06-trajectory-gene-programs-input-loading
#| echo: true
#| warning: false
#| message: false
# Create output folders and load configured inputs.
set.seed(seed)
dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(figures_dir, recursive = TRUE, showWarnings = FALSE)
object <- qs2::qs_read(file = input_object, nthreads = n_threads)
DefaultAssay(object) <- assay_name
```

## Validation

Check the required inputs before running the analysis.

```{r}
#| label: tpl-scrna-06-trajectory-gene-programs-validation
#| echo: true
#| warning: false
#| message: false
# Stop early when required inputs or columns are missing.
stopifnot(inherits(object, "Seurat"))
stopifnot(all(c(sample_column, group_column, pseudotime_column) %in% colnames(object[[]])))
stopifnot(reduction_name %in% names(object@reductions))
features_of_interest <- intersect(features_of_interest, rownames(object))
```

## Input Summary

Record a compact summary of the inputs for later review.

```{r}
#| label: tbl-scrna-trajectory-input-summary
#| tbl-cap: "Summary of the trajectory input object."
#| echo: true
#| warning: false
#| message: false
# Write a small TSV summary of the loaded input.
input_summary <- object[[]] |> tibble::as_tibble() |> summarise(cells = n(), pseudotime_min = min(.data[[pseudotime_column]], na.rm = TRUE), pseudotime_max = max(.data[[pseudotime_column]], na.rm = TRUE))
readr::write_tsv(input_summary, file.path(results_dir, "trajectory_input_summary.tsv"))
input_summary
```

## Analysis

Run the main analysis with the parameters defined above.

```{r}
#| label: tpl-scrna-06-trajectory-gene-programs-main-analysis
#| echo: true
#| warning: false
#| message: false
# Run the reusable analysis step for this template.
expr <- FetchData(object, vars = c(features_of_interest, pseudotime_column, group_column))
expr$pseudotime_bin <- cut(expr[[pseudotime_column]], breaks = n_bins, include.lowest = TRUE)
program_tbl <- expr |>
  group_by(pseudotime_bin, .data[[group_column]]) |>
  summarise(across(all_of(features_of_interest), mean, na.rm = TRUE), .groups = "drop")
```

## Diagnostics

Inspect the checks that help assess the analysis output.

```{r}
#| label: tpl-scrna-06-trajectory-gene-programs-qc-checks
#| echo: true
#| warning: false
#| message: false
# Export checks that make failures easy to inspect.
bin_counts <- expr |> count(pseudotime_bin, .data[[group_column]], name = "n_cells")
readr::write_tsv(bin_counts, file.path(results_dir, "pseudotime_bin_counts.tsv"))
```

## Visualization

Save the main diagnostic or result figures to the output directory.

```{r}
#| label: tpl-scrna-06-trajectory-gene-programs-visualization
#| echo: true
#| warning: false
#| message: false
# Save compact figures for the main outputs.
pdf(file.path(figures_dir, "trajectory_features.pdf"), width = 7, height = 5)
if (length(features_of_interest) > 0) print(FeaturePlot(object, features = features_of_interest, reduction = reduction_name))
dev.off()
```

## Object metadata

Store the derived metadata alongside the analysis object.

```{r}
#| label: tpl-scrna-06-trajectory-gene-programs-object-result-storage
#| echo: true
#| warning: false
#| message: false
# Store reusable analysis metadata inside the object.
object@misc$trajectory_gene_programs <- list(features = features_of_interest, pseudotime_column = pseudotime_column, n_bins = n_bins)
```

## File outputs

Write the tabular outputs produced by this workflow.

```{r}
#| label: tpl-scrna-06-trajectory-gene-programs-result-export
#| echo: true
#| warning: false
#| message: false
# Export the main reusable tables as TSV files.
readr::write_tsv(program_tbl, file.path(results_dir, "trajectory_gene_programs.tsv"))
```

## Save

Save the resulting object in the format used by this notebook.

```{r}
#| label: tpl-scrna-06-trajectory-gene-programs-final-save
#| echo: true
#| warning: false
#| message: false
# Save the updated object for the next template step.
qs2::qs_save(object, file = output_object, nthreads = n_threads, compress_level = compress_level)
```

## Output summary

Summarize the files and objects written by the workflow.

```{r}
#| label: tpl-scrna-06-trajectory-gene-programs-output-summary
#| echo: true
#| warning: false
#| message: false
# Print a compact summary of generated outputs.
tibble::tibble(output_object = output_object, features_used = length(features_of_interest))
```

## Reproducibility {.appendix}

This final chunk records the R session used when the notebook is rendered.

```{r}
#| label: session-information
#| echo: false
#| results: markup
sessionInfo()
```
