---
title: "MOFA2: model diagnostics"
format: html
editor: source
---

> **Status:** `draft`<br>
> **Template class:** `SOURCE-BACKED WORKFLOW`

## Purpose

Load the fitted native MOFA2 model and check whether its latent structure is
sensible before assigning biological meaning. This page does not refit the
model or change its options.

## Inputs and parameters

```{r}
#| label: parameters

model_path <- "output/mofa/mofa_model.hdf5"
variance_total_output <- "output/mofa/variance_explained_total.tsv"
variance_factor_output <- "output/mofa/variance_explained_per_factor.tsv"
factor_scores_output <- "output/mofa/factor_scores_diagnostics.tsv"
factor_correlation_output <- "output/mofa/factor_correlation.tsv"
top_weights_output <- "output/mofa/top_weights_diagnostic.tsv"
technical_association_output <- "output/mofa/technical_factor_associations.tsv"

# Optional sample metadata. Technical columns should be numeric quality or
# processing variables; do not turn this section into an outcome screen.
metadata_path <- NULL
metadata_sample_column <- "sample_id"
technical_columns <- character()
top_features_per_factor <- 10L
```

::: {.callout-note}
## Practical note

Before interpreting a latent factor biologically, I first check whether it is
strongly associated with obvious technical variables or sample-quality effects.
A technically driven factor is still a real pattern in the data, but it should
not be relabeled as biology simply because it explains substantial variance.
:::

::: {.callout-warning}
## Factor numbers are not biological identities

`Factor1` in one fit is not guaranteed to represent the same pattern as
`Factor1` in another fit. Factor order and sign can change between models.
Interpret factors through their scores, weights, variance explained and
associations rather than by their numeric label alone.
:::

## Load the native model

```{r}
#| label: load-model

model <- MOFA2::load_model(model_path, verbose = FALSE)
```

## Variance explained and factor-level summaries

```{r}
#| label: variance-explained

variance_explained <- MOFA2::get_variance_explained(
  model,
  as.data.frame = TRUE
)
readr::write_tsv(variance_explained$r2_total, variance_total_output)
readr::write_tsv(variance_explained$r2_per_factor, variance_factor_output)

factor_variance <- variance_explained$r2_per_factor
factor_summary <- aggregate(
  value ~ factor,
  data = factor_variance,
  FUN = sum
)
names(factor_summary)[names(factor_summary) == "value"] <-
  "sum_view_specific_variance_explained"
print(factor_summary)

variance_plot <- MOFA2::plot_variance_explained(
  model,
  plot_total = TRUE
)
print(variance_plot)
```

Variance explained is view- and factor-specific. A factor that matters for one
view may contribute little to another; low total variance is not proof of
biological irrelevance, and high variance is not proof of biological importance.

## Factor scores, correlation and weight sanity checks

```{r}
#| label: factor-and-weight-diagnostics

factor_scores <- MOFA2::get_factors(
  model,
  groups = "all",
  factors = "all",
  as.data.frame = TRUE
)
names(factor_scores)[names(factor_scores) == "sample"] <- "sample_id"
readr::write_tsv(factor_scores, factor_scores_output)

factor_list <- MOFA2::get_factors(model, groups = "all", factors = "all")
if (length(factor_list) != 1L) {
  stop("This compact diagnostic expects one matched sample group.")
}
factor_matrix <- factor_list[[1L]]
factor_correlation <- stats::cor(
  factor_matrix,
  method = "spearman",
  use = "pairwise.complete.obs"
)
readr::write_tsv(
  tibble::as_tibble(factor_correlation, rownames = "factor"),
  factor_correlation_output
)
boxplot(
  as.data.frame(factor_matrix),
  las = 2,
  main = "MOFA factor score distributions",
  ylab = "Factor score"
)

weights <- MOFA2::get_weights(
  model,
  views = "all",
  factors = "all",
  scale = FALSE,
  as.data.frame = TRUE
)
weights$abs_weight <- abs(weights$value)
weight_groups <- split(
  seq_len(nrow(weights)),
  list(weights$view, weights$factor),
  drop = TRUE
)
top_weight_indices <- unlist(lapply(weight_groups, function(indices) {
  ordered <- indices[order(weights$abs_weight[indices], decreasing = TRUE)]
  head(ordered, top_features_per_factor)
}))
top_weights <- weights[top_weight_indices, , drop = FALSE]
readr::write_tsv(top_weights, top_weights_output)
```

The weight check is a compact sanity view of high absolute weights. It keeps
the sign in the exported `value` column; absolute weight is used only for
ranking.

## Optional technical associations

```{r}
#| label: technical-associations

if (!is.null(metadata_path) && length(technical_columns) > 0L) {
  metadata <- readr::read_tsv(metadata_path, show_col_types = FALSE)
  if (!metadata_sample_column %in% names(metadata)) {
    stop("metadata_sample_column is absent from the metadata table.")
  }
  if (anyDuplicated(metadata[[metadata_sample_column]])) {
    stop("Metadata sample IDs must be unique.")
  }
  if (any(!rownames(factor_matrix) %in% metadata[[metadata_sample_column]])) {
    stop("All model samples must be present in the metadata table.")
  }
  metadata <- metadata[
    match(rownames(factor_matrix), metadata[[metadata_sample_column]]),
    ,
    drop = FALSE
  ]

  association_rows <- list()
  association_index <- 0L
  for (technical_column in technical_columns) {
    if (!technical_column %in% names(metadata) ||
        !is.numeric(metadata[[technical_column]])) {
      stop("technical_columns must name numeric metadata columns.")
    }
    for (factor_name in colnames(factor_matrix)) {
      keep <- stats::complete.cases(
        metadata[[technical_column]],
        factor_matrix[, factor_name]
      )
      if (sum(keep) >= 3L &&
          length(unique(metadata[[technical_column]][keep])) > 1L &&
          length(unique(factor_matrix[keep, factor_name])) > 1L) {
        test <- suppressWarnings(stats::cor.test(
          metadata[[technical_column]][keep],
          factor_matrix[keep, factor_name],
          method = "spearman",
          exact = FALSE
        ))
        association_index <- association_index + 1L
        association_rows[[association_index]] <- data.frame(
          metadata_column = technical_column,
          factor = factor_name,
          n_samples = sum(keep),
          rho = unname(test$estimate),
          p_value = test$p.value,
          stringsAsFactors = FALSE
        )
      }
    }
  }
  if (length(association_rows) > 0L) {
    technical_associations <- do.call(rbind, association_rows)
    technical_associations$adjusted_p_value <- stats::p.adjust(
      technical_associations$p_value,
      method = "BH"
    )
    readr::write_tsv(technical_associations, technical_association_output)
  }
}
```

Technical associations are diagnostics, not proof that a factor is technical
or biological. Multiple factor-variable tests are adjusted together, and one
sample remains one observation regardless of how many views contributed to the
model.

## Output contract

This page consumes `mofa_model.hdf5` and exports variance-explained tables,
factor scores, factor correlations and a compact high-weight diagnostic. It
does not refit or rewrite the model. Technical association TSV output is created
only when numeric metadata columns are explicitly supplied.
