---
title: "MOFA2: fit a multi-view factor model"
format: html
editor: source
---

> **Status:** `draft`<br>
> **Template class:** `SOURCE-BACKED WORKFLOW`

## Purpose

Train a MOFA2 model from the prepared named view list. This page stops at the
native trained model and records the choices needed to reproduce the fit.
Diagnostics and biological interpretation belong in the downstream pages.

## Inputs and parameters

```{r}
#| label: parameters

prepared_views_path <- "output/mofa/prepared_views.qs2"
model_output <- "output/mofa/mofa_model.hdf5"
model_options_output <- "output/mofa/mofa_model_options.tsv"
training_options_output <- "output/mofa/mofa_training_options.tsv"

num_factors <- 10L
scale_views <- FALSE
seed <- 1234L
convergence_mode <- "slow"
max_iterations <- 1500L
drop_factor_threshold <- 0.01
use_basilisk <- FALSE
save_data <- TRUE
verbose <- TRUE

# Declare one current MOFA2 likelihood for every view. Do not leave this
# implicit: use gaussian for prepared continuous values, bernoulli for binary
# presence/absence values, or poisson only when the count-scale model is
# intentionally appropriate. Example:
# likelihoods <- c(transcriptome = "gaussian", mutation = "bernoulli")
likelihoods <- c()
```

::: {.callout-important}
## MOFA is unsupervised

Train the model from the omics views, not from the biological outcome that you
later hope to interpret. Clinical groups, response labels, signatures and
phenotypes may be associated with learned factors afterward in
`interpretation.qmd`. Do not use a target phenotype to choose features or tune
this model while describing it as unsupervised.
:::

## Validate the prepared object

```{r}
#| label: validate-prepared-views

views <- qs2::qs_read(prepared_views_path)
if (!is.list(views) || is.null(names(views)) || length(views) < 2L) {
  stop("The prepared object must be a named list containing at least two views.")
}
if (anyDuplicated(names(views))) {
  stop("Prepared view names must be unique.")
}
if (any(!vapply(views, is.matrix, logical(1))) ||
    any(!vapply(views, is.numeric, logical(1)))) {
  stop("Every prepared view must be a numeric matrix.")
}
if (any(vapply(views, function(x) nrow(x) == 0L || ncol(x) == 0L,
            logical(1)))) {
  stop("Prepared views must have non-zero dimensions.")
}
if (any(vapply(views, function(x) is.null(rownames(x)) ||
                 anyDuplicated(rownames(x)) > 0L, logical(1)))) {
  stop("Prepared views need valid unique feature names.")
}
if (any(vapply(views, function(x) is.null(colnames(x)) ||
                 anyDuplicated(colnames(x)) > 0L, logical(1)))) {
  stop("Prepared views need valid unique sample IDs.")
}

sample_ids <- colnames(views[[1L]])
if (any(!vapply(views, function(x) identical(colnames(x), sample_ids),
               logical(1)))) {
  stop("All prepared views must use the same sample order.")
}
if (any(vapply(views, function(x) {
  feature_sd <- apply(x, 1L, stats::sd, na.rm = TRUE)
  all(!is.finite(feature_sd) | feature_sd == 0)
}, logical(1)))) {
  stop("A view has no variable feature available for fitting.")
}
if (any(vapply(views, function(x) any(!is.na(x) & !is.finite(x)),
             logical(1)))) {
  stop("Prepared views contain Inf or NaN values.")
}

view_summary <- data.frame(
  view = names(views),
  features = vapply(views, nrow, integer(1)),
  samples = vapply(views, ncol, integer(1)),
  missing_values = vapply(views, function(x) sum(is.na(x)), integer(1)),
  stringsAsFactors = FALSE
)
print(view_summary)
```

## Configure and prepare MOFA2

```{r}
#| label: prepare-mofa

dir.create(dirname(model_output), recursive = TRUE, showWarnings = FALSE)
mofa <- MOFA2::create_mofa(views)
view_names <- MOFA2::views_names(mofa)

default_model_options <- MOFA2::get_default_model_options(mofa)
if (length(likelihoods) == 0L) {
  print(default_model_options$likelihoods)
  stop("Declare likelihoods explicitly for every prepared view before fitting.")
}
if (is.null(names(likelihoods)) || anyDuplicated(names(likelihoods)) ||
    !setequal(names(likelihoods), view_names)) {
  stop("likelihoods must be named exactly for the prepared views.")
}
if (any(!likelihoods %in% c("gaussian", "bernoulli", "poisson", "exponential"))) {
  stop("Use a likelihood supported by the current MOFA2 model options.")
}

data_options <- MOFA2::get_default_data_options(mofa)
data_options$scale_views <- scale_views
data_options$use_float32 <- TRUE

model_options <- default_model_options
model_options$num_factors <- num_factors
model_options$likelihoods <- likelihoods[view_names]
model_options$spikeslab_weights <- TRUE
model_options$ard_weights <- TRUE
model_options$ard_factors <- FALSE

training_options <- MOFA2::get_default_training_options(mofa)
training_options$seed <- seed
training_options$convergence_mode <- convergence_mode
training_options$maxiter <- max_iterations
training_options$drop_factor_threshold <- drop_factor_threshold
training_options$verbose <- verbose

model_option_table <- data.frame(
  option = c("num_factors", "scale_views", "use_float32",
             "spikeslab_weights", "ard_weights", "ard_factors", "likelihoods"),
  value = c(
    as.character(model_options$num_factors),
    as.character(data_options$scale_views),
    as.character(data_options$use_float32),
    as.character(model_options$spikeslab_weights),
    as.character(model_options$ard_weights),
    as.character(model_options$ard_factors),
    paste(paste(names(likelihoods), likelihoods, sep = "="), collapse = ";")
  ),
  stringsAsFactors = FALSE
)
training_option_table <- data.frame(
  option = c("seed", "convergence_mode", "maxiter",
             "drop_factor_threshold", "save_data", "use_basilisk"),
  value = as.character(c(seed, convergence_mode, max_iterations,
                         drop_factor_threshold, save_data, use_basilisk)),
  stringsAsFactors = FALSE
)
readr::write_tsv(model_option_table, model_options_output)
readr::write_tsv(training_option_table, training_options_output)

mofa <- MOFA2::prepare_mofa(
  object = mofa,
  data_options = data_options,
  model_options = model_options,
  training_options = training_options
)
```

## Train and save the native model

```{r}
#| label: run-mofa

mofa <- MOFA2::run_mofa(
  object = mofa,
  outfile = model_output,
  save_data = save_data,
  use_basilisk = use_basilisk
)

if (!file.exists(model_output)) {
  stop("MOFA2 did not create the expected native HDF5 model.")
}
cat("Saved trained MOFA2 model to", model_output, "\n")
```

## Output contract

The output is a trained MOFA2 model in native HDF5 format. The prepared views
remain the `.qs2` input, while concise model/training option manifests are TSV.
The requested number of factors is an initial unsupervised choice; inactive or
redundant factors and explained variance must be reviewed in
`diagnostics.qmd` before interpretation.

::: {.callout-note}
## R/Python environment

MOFA2 calls the Python `mofapy2` backend. `use_basilisk` is exposed here rather
than hiding an environment choice or a local Python path. Use the simplest
working environment for the installation and consult current MOFA2 guidance if
the backend, reticulate, or basilisk setup changes.
:::
