---
title: "MOFA2: interpret factors and feature weights"
format: html
editor: source
---

> **Status:** `draft`<br>
> **Template class:** `SOURCE-BACKED WORKFLOW`

## Purpose

Interpret a fitted MOFA2 model through sample-level factor scores, feature
weights and optional metadata associations. This page loads the native model;
it never retrains it. Factor selection is explicit and post hoc.

## Inputs and parameters

```{r}
#| label: parameters

model_path <- "output/mofa/mofa_model.hdf5"
factor_scores_output <- "output/mofa/factor_scores.tsv"
factor_weights_output <- "output/mofa/factor_weights.tsv"
top_weights_output <- "output/mofa/top_feature_weights.tsv"
continuous_association_output <- "output/mofa/factor_continuous_associations.tsv"
categorical_association_output <- "output/mofa/factor_categorical_associations.tsv"
group_summary_output <- "output/mofa/factor_group_summaries.tsv"

# Empty means export all factors/views. Otherwise use current factor/view names,
# after reviewing diagnostics; do not assume Factor1 is privileged.
factors_to_interpret <- character()
views_to_report <- character()
top_features_per_factor <- 20L

metadata_path <- NULL
metadata_sample_column <- "sample_id"
continuous_metadata_columns <- character()
categorical_metadata_column <- NULL
```

::: {.callout-warning}
## Post hoc factor selection

MOFA fitting is unsupervised, but downstream factor selection can still become
outcome-driven. If you screen many factors against a phenotype and focus only
on the strongest association, treat that as exploratory post hoc analysis and
report the multiple-testing context. Do not describe the selected factor as if
the phenotype had been an independent validation target.
:::

::: {.callout-note}
## Interpretation language

Factor weights identify covariance structure and high-weight features help
characterize a latent factor. Factor/phenotype associations are associations;
MOFA does not establish causal regulatory relationships. Prefer
“top-weight features” or “features contributing strongly to the factor” over
causal language.
:::

## Load the fitted model and extract factors

```{r}
#| label: extract-factor-scores

model <- MOFA2::load_model(model_path, verbose = FALSE)
factor_scores <- MOFA2::get_factors(
  model,
  groups = "all",
  factors = "all",
  as.data.frame = TRUE
)
names(factor_scores)[names(factor_scores) == "sample"] <- "sample_id"
readr::write_tsv(factor_scores, factor_scores_output)

factor_list <- MOFA2::get_factors(model, groups = "all", factors = "all")
if (length(factor_list) != 1L) {
  stop("This compact interpretation template expects one sample group.")
}
factor_matrix <- factor_list[[1L]]
available_factors <- colnames(factor_matrix)
selected_factors <- if (length(factors_to_interpret) == 0L) {
  available_factors
} else {
  if (any(!factors_to_interpret %in% available_factors)) {
    stop("factors_to_interpret contains an unavailable factor name.")
  }
  factors_to_interpret
}
```

The factor table is tidy (`sample_id`, `factor`, `value`) and the matrix used
for plots/associations has one row per sample. Factor sign and order are
model-specific and should not be compared across independent fits without an
explicit alignment step.

## Extract signed feature weights

```{r}
#| label: extract-feature-weights

factor_weights <- MOFA2::get_weights(
  model,
  views = "all",
  factors = "all",
  scale = FALSE,
  as.data.frame = TRUE
)
names(factor_weights)[names(factor_weights) == "value"] <- "weight"
factor_weights$abs_weight <- abs(factor_weights$weight)
readr::write_tsv(factor_weights, factor_weights_output)

if (length(views_to_report) > 0L) {
  if (any(!views_to_report %in% unique(factor_weights$view))) {
    stop("views_to_report contains an unavailable view name.")
  }
  factor_weights <- factor_weights[factor_weights$view %in% views_to_report,
                                   , drop = FALSE]
}
factor_weights <- factor_weights[factor_weights$factor %in% selected_factors,
                                 , drop = FALSE]

weight_groups <- split(
  seq_len(nrow(factor_weights)),
  list(factor_weights$view, factor_weights$factor),
  drop = TRUE
)
top_weight_indices <- unlist(lapply(weight_groups, function(indices) {
  ordered <- indices[
    order(factor_weights$abs_weight[indices], decreasing = TRUE)
  ]
  head(ordered, top_features_per_factor)
}))
top_weights <- factor_weights[top_weight_indices, , drop = FALSE]
readr::write_tsv(top_weights, top_weights_output)
```

Top features are ranked by absolute weight while retaining the signed weight.
This is a characterization of the fitted factor, not a causal feature list.

## Optional metadata associations

```{r}
#| label: metadata-associations

if (!is.null(metadata_path) &&
    (length(continuous_metadata_columns) > 0L ||
     !is.null(categorical_metadata_column))) {
  metadata <- readr::read_tsv(metadata_path, show_col_types = FALSE)
  if (!metadata_sample_column %in% names(metadata)) {
    stop("metadata_sample_column is absent from the metadata table.")
  }
  if (anyDuplicated(metadata[[metadata_sample_column]])) {
    stop("Metadata sample IDs must be unique.")
  }
  if (any(!rownames(factor_matrix) %in% metadata[[metadata_sample_column]])) {
    stop("All model samples must be present in the metadata table.")
  }
  metadata <- metadata[
    match(rownames(factor_matrix), metadata[[metadata_sample_column]]),
    ,
    drop = FALSE
  ]

  continuous_rows <- list()
  continuous_index <- 0L
  for (metadata_column in continuous_metadata_columns) {
    if (!metadata_column %in% names(metadata) ||
        !is.numeric(metadata[[metadata_column]])) {
      stop("continuous_metadata_columns must name numeric columns.")
    }
    for (factor_name in selected_factors) {
      keep <- stats::complete.cases(
        metadata[[metadata_column]],
        factor_matrix[, factor_name]
      )
      if (sum(keep) >= 3L &&
          length(unique(metadata[[metadata_column]][keep])) > 1L &&
          length(unique(factor_matrix[keep, factor_name])) > 1L) {
        test <- suppressWarnings(stats::cor.test(
          metadata[[metadata_column]][keep],
          factor_matrix[keep, factor_name],
          method = "spearman",
          exact = FALSE
        ))
        continuous_index <- continuous_index + 1L
        continuous_rows[[continuous_index]] <- data.frame(
          metadata_column = metadata_column,
          factor = factor_name,
          n_samples = sum(keep),
          rho = unname(test$estimate),
          p_value = test$p.value,
          stringsAsFactors = FALSE
        )
      }
    }
  }
  if (length(continuous_rows) > 0L) {
    continuous_associations <- do.call(rbind, continuous_rows)
    continuous_associations$adjusted_p_value <- stats::p.adjust(
      continuous_associations$p_value,
      method = "BH"
    )
    readr::write_tsv(
      continuous_associations,
      continuous_association_output
    )
  }

  if (!is.null(categorical_metadata_column)) {
    if (!categorical_metadata_column %in% names(metadata)) {
      stop("categorical_metadata_column is absent from the metadata table.")
    }
    category_rows <- list()
    group_rows <- list()
    category_index <- 0L
    group_index <- 0L
    for (factor_name in selected_factors) {
      grouping <- metadata[[categorical_metadata_column]]
      keep <- !is.na(grouping) & !is.na(factor_matrix[, factor_name])
      if (sum(keep) >= 3L && length(unique(grouping[keep])) >= 2L) {
        test <- stats::kruskal.test(
          factor_matrix[keep, factor_name] ~ as.factor(grouping[keep])
        )
        category_index <- category_index + 1L
        category_rows[[category_index]] <- data.frame(
          metadata_column = categorical_metadata_column,
          factor = factor_name,
          n_samples = sum(keep),
          groups = length(unique(grouping[keep])),
          p_value = test$p.value,
          stringsAsFactors = FALSE
        )
        for (group_name in unique(grouping[keep])) {
          group_index <- group_index + 1L
          group_rows[[group_index]] <- data.frame(
            metadata_column = categorical_metadata_column,
            group = as.character(group_name),
            factor = factor_name,
            n_samples = sum(keep & grouping == group_name),
            median_score = stats::median(
              factor_matrix[keep & grouping == group_name, factor_name]
            ),
            stringsAsFactors = FALSE
          )
        }
      }
    }
    if (length(category_rows) > 0L) {
      categorical_associations <- do.call(rbind, category_rows)
      categorical_associations$adjusted_p_value <- stats::p.adjust(
        categorical_associations$p_value,
        method = "BH"
      )
      readr::write_tsv(
        categorical_associations,
        categorical_association_output
      )
      readr::write_tsv(do.call(rbind, group_rows), group_summary_output)
    }
  }
}
```

Continuous associations use Spearman correlation and adjust the tested
factor-variable combinations with Benjamini-Hochberg correction. The optional
categorical section uses a modest Kruskal-Wallis comparison without assuming two
groups and reports group medians. In both cases, one sample is one observation;
multiple views do not create replicated samples.

## Factor-score heatmap

```{r}
#| label: factor-score-heatmap

heatmap_matrix <- factor_matrix[, selected_factors, drop = FALSE]
if (nrow(heatmap_matrix) > 1L && ncol(heatmap_matrix) > 0L) {
  heatmap(
    heatmap_matrix,
    scale = "none",
    Colv = NA,
    main = "Selected MOFA factor scores"
  )
}
```

The heatmap needs no biological group annotation and therefore does not require
a fixed phenotype label. Ordering by metadata, if added for a particular study,
should remain an explicit post hoc choice.

## Output contract

This page consumes `mofa_model.hdf5` and exports tidy factor scores, signed
feature weights, absolute-weight top-feature tables, and optional metadata
associations/group summaries. It does not refit the model and does not recreate
signature scoring, pathway enrichment, motif analysis or TF-activity workflows.
