---
title: "Untitled"
format: html
---

# Overview

This notebook analyzes the public dataset `GSE171145` as an external LUAD comparison cohort. The workflow merges the raw samples, adds CopyKAT and clinical metadata, performs standard Seurat processing, annotates cell states, and then quantifies the same C1 plasticity program used in the CSS project.

Biologically, this notebook asks whether the plasticity axis observed in the CSS samples is recapitulated in an independent set of EGFR-mutant LUAD tumors.

# Setup

The opening chunks load the analysis packages and define a helper to reconstruct one Seurat object per raw sample before merging them into a common atlas.


```{r}
library(tidyverse)
library(Seurat)
library(patchwork)
library(scCustomize)
library(qs2)
library(harmony)
library(ggsci)

```

# Load and merge raw samples

This block reads each raw count table plus its barcode map, harmonizes cell names, and merges all samples into a single Seurat object that becomes the starting point for the downstream cohort-level analysis.

```{r}

library(Seurat)
library(data.table)

umap_theme <- ggpubr::theme_classic2() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 11, color = "grey30"),

    axis.title = element_text(size = 13),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),

    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10)
  )

load_bd_sample <- function(count_file, cell_file, sample_id) {
  
  message("Loading: ", sample_id)
  
  # Read counts
  counts <- fread(count_file, data.table = FALSE)
  rownames(counts) <- counts$gene
  counts <- counts[,-1]
  
  # Read cell map
  cell_map <- fread(cell_file, sep = "\t", header = TRUE)
  
  # Check alignment
  if(!all(colnames(counts) %in% cell_map$CellIndex)) {
    stop(paste("Mismatch in:", sample_id))
  }
  
  # Reorder mapping
  cell_map <- cell_map[match(colnames(counts), cell_map$CellIndex), ]
  
  # Replace with real barcodes
  colnames(counts) <- cell_map$CellName
  
  # Create Seurat object
  obj <- CreateSeuratObject(counts = counts, project = "GSE171145")
  
  # Add metadata
  obj$sample <- sample_id
  
  return(obj)
}

base_path <- "input/tables/GSE171145_RAW"

count_files <- list.files(base_path, pattern = "counts.tsv$", full.names = TRUE)

seurat_list <- lapply(count_files, function(count_file) {
  
  sample_id <- gsub(".counts.tsv", "", basename(count_file))
  cell_file <- sub("counts.tsv", "cellname.list.txt", count_file)
  
  load_bd_sample(count_file, cell_file, sample_id)
})

names(seurat_list) <- gsub(".counts.tsv", "", basename(count_files))

combined <- merge(
  seurat_list[[1]],
  y = seurat_list[-1],
  add.cell.ids = names(seurat_list)
)

combined

table(combined$sample)

qs_save(object = combined,file = "output/objects/GSE171145_merged_preQC.qs2",compress_level = 5,nthreads = 10)

```

# copykat CNV annotataion

CopyKAT is run here as an orthogonal malignancy layer so the plasticity analysis can later be interpreted alongside inferred CNV status.
```{r}

library(copykat)

combined <- qs_read(file = "output/objects/GSE171145_merged_preQC.qs2", nthreads = 10)

# P1
exp.rawdata <- as.matrix(combined@assays$RNA$counts.1)
gc()
copykat.P1 <- copykat(rawmat=exp.rawdata, id.type="S", ngene.chr=5, win.size=25, KS.cut=0.1, sam.name="P1", distance="euclidean", norm.cell.names="", output.seg=FALSE, plot.genes=FALSE, genome="hg20", n.cores=8)

# P2
exp.rawdata <- as.matrix(combined@assays$RNA$counts.2)
gc()
copykat.P2 <- copykat(rawmat=exp.rawdata, id.type="S", ngene.chr=5, win.size=25, KS.cut=0.1, sam.name="P2", distance="euclidean", norm.cell.names="", output.seg=FALSE, plot.genes=FALSE, genome="hg20", n.cores=8)

# P3
exp.rawdata <- as.matrix(combined@assays$RNA$counts.3)
gc()
copykat.P3 <- copykat(rawmat=exp.rawdata, id.type="S", ngene.chr=5, win.size=25, KS.cut=0.1, sam.name="P3", distance="euclidean", norm.cell.names="", output.seg=FALSE, plot.genes=FALSE, genome="hg20", n.cores=8)

# P4
exp.rawdata <- as.matrix(combined@assays$RNA$counts.4)
gc()
copykat.P4 <- copykat(rawmat=exp.rawdata, id.type="S", ngene.chr=5, win.size=25, KS.cut=0.1, sam.name="P4", distance="euclidean", norm.cell.names="", output.seg=FALSE, plot.genes=FALSE, genome="hg20", n.cores=8)

# P5
exp.rawdata <- as.matrix(combined@assays$RNA$counts.5)
gc()
copykat.P5 <- copykat(rawmat=exp.rawdata, id.type="S", ngene.chr=5, win.size=25, KS.cut=0.1, sam.name="P5", distance="euclidean", norm.cell.names="", output.seg=FALSE, plot.genes=FALSE, genome="hg20", n.cores=8)

# P6
exp.rawdata <- as.matrix(combined@assays$RNA$counts.6)
gc()
copykat.P6 <- copykat(rawmat=exp.rawdata, id.type="S", ngene.chr=5, win.size=25, KS.cut=0.1, sam.name="P6", distance="euclidean", norm.cell.names="", output.seg=FALSE, plot.genes=FALSE, genome="hg20", n.cores=8)

# P7
exp.rawdata <- as.matrix(combined@assays$RNA$counts.7)
gc()
copykat.P7 <- copykat(rawmat=exp.rawdata, id.type="S", ngene.chr=5, win.size=25, KS.cut=0.1, sam.name="P7", distance="euclidean", norm.cell.names="", output.seg=FALSE, plot.genes=FALSE, genome="hg20", n.cores=8)

# P8
exp.rawdata <- as.matrix(combined@assays$RNA$counts.8)
gc()
copykat.P8 <- copykat(rawmat=exp.rawdata, id.type="S", ngene.chr=5, win.size=25, KS.cut=0.1, sam.name="P8", distance="euclidean", norm.cell.names="", output.seg=FALSE, plot.genes=FALSE, genome="hg20", n.cores=5)

# P9
exp.rawdata <- as.matrix(combined@assays$RNA$counts.9)
gc()
copykat.P9 <- copykat(rawmat=exp.rawdata, id.type="S", ngene.chr=5, win.size=25, KS.cut=0.1, sam.name="P9", distance="euclidean", norm.cell.names="", output.seg=FALSE, plot.genes=FALSE, genome="hg20", n.cores=8)



results <- list()

for (i in 1:9) {
  layer_name <- paste0("counts.", i)
  message("Running CopyKAT on ", layer_name)
  
  exp.rawdata <- as.matrix(combined@assays$RNA[[layer_name]])
  gc()
  
  results[[layer_name]] <- copykat(
    rawmat = exp.rawdata,
    id.type = "S",
    ngene.chr = 5,
    win.size = 25,
    KS.cut = 0.1,
    sam.name = paste0("P", i),
    distance = "euclidean",
    norm.cell.names = "",
    output.seg = FALSE,
    plot.genes = FALSE,
    genome = "hg20",
    n.cores = 8
  )
}

```

# Basic QC and object cleanup

After merging, the object is joined across layers, annotated with mitochondrial content, and filtered to remove poor-quality cells before deeper interpretation.
```{r}

combined <- qs_read(file = "output/objects/GSE171145_merged_preQC.qs2",nthreads = 10)

combined <- JoinLayers(combined)

combined[["percent.mt"]] <- PercentageFeatureSet(combined, pattern = "^MT-")

VlnPlot(combined, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"))

combined <- subset(
  combined,
  subset = nFeature_RNA > 500 &
           nFeature_RNA < 6000 &
           percent.mt < 20
)

combined

```

# Add EGFR metadata 

This metadata step links the merged object back to patient-level variables, which is essential for relating C1-Core to genotype and response-like clinical features.
```{r}

combined <- qs_read(file = "output/objects/GSE171145_merged_preQC.qs2", nthreads = 10)

patient_meta <- vroom::vroom("output/tables/patient_meta.csv") |>
  select(-...1)

meta <- combined@meta.data

sample_to_patient <- c(
  "GSM5219674_LJQ-T"  = "P1",
  "GSM5219675_GBG-T"  = "P2",
  "GSM5219676_LYB-T1" = "P3",
  "GSM5219677_LYB-T2" = "P4",
  "GSM5219678_CYD-T"  = "P5",
  "GSM5219679_CYZ-T"  = "P6",
  "GSM5219680_XMS"    = "P7",
  "GSM5219681_ZYQ"    = "P8",
  "GSM5219682_TGS"    = "P9"
)

meta$patient_id <- sample_to_patient[meta$sample]

meta

meta2 <- merge(
  meta,
  patient_meta,
  by = "patient_id",
  all.x = TRUE
)

rownames(meta2) <- rownames(meta)

combined@meta.data <- meta2

p2 <- DimPlot_scCustom(combined, group.by = "egfr") +
  theme_bw(base_size = 12) +
  ggtitle("EGFR annotation") +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12)
  )
p2

ggsave(
  filename = "output/figures/EGFR_DimPlot.png",
  plot = p2,
  dpi = 600,
  width = 10,
  height = 8
)

```

# Doublets
```{r}

# 4. Doublet detection
library(scDblFinder)
sce <- as.SingleCellExperiment(combined)
library(BiocParallel)
sce <- scDblFinder(sce, samples="orig.ident", BPPARAM=MulticoreParam(5))
combined$doublet_score <- sce$scDblFinder.score
combined$doublet_class <- sce$scDblFinder.class

table(combined$doublet_class)

prop.table(table(combined$doublet_class))

combined <- subset(combined, subset = doublet_class == "singlet")

```

## minimal processing

```{r}

plan(multicore, workers = 5)
options(future.globals.maxSize = 10 * 1024^3)  # 2 GiB
combined <- NormalizeData(combined)
combined <- FindVariableFeatures(combined, nfeatures = 5000)
combined <- ScaleData(combined)
combined <- RunPCA(combined)

PC_Plotting(combined, dim_number=1)
PC_Plotting(combined, dim_number=2)

# Check elbow plot to justify PC selection
ElbowPlot_scCustom(combined, ndims=50)

combined <- RunHarmony(object = combined,group.by.vars = "sample")

# 5. Neighbors + clustering
combined <- FindNeighbors(
  combined,
  dims = 1:42,
  reduction = "harmony"
)

plan(multicore, workers = 1)
combined <- FindClusters(combined, resolution = 0.8,random.seed = 1234)

plan(multicore, workers = 4)
combined <- RunUMAP(object = combined,umap.method = "uwot2",  seed.use = 1234,dims = 1:42,reduction = "harmony")


library(lisi)

emb <- Embeddings(combined, "harmony")

data <- data.frame(
  harmony_1 = emb[,1],
  harmony_2 = emb[,2]
)

# meta
meta <- data.frame(batch = combined$sample)

# LISI on Harmony space
lisi <- compute_lisi(data, meta, "batch") |>
  as.data.frame() |>
  rownames_to_column("cell")

skimr::skim(lisi)

summary(lisi)

```

```{r}

p4 <- combined@meta.data %>%
  dplyr::count(sample, name = "n_cells") %>%
  dplyr::arrange(desc(n_cells)) %>%
  ggplot(aes(x = reorder(sample, n_cells), y = n_cells)) +
  geom_col() +
  geom_text(aes(label = n_cells), hjust = -0.2, size = 3) +
  labs(x = "Patient", y = "Number of cells", fill = "Stage") +
  coord_flip() +
  theme_bw(base_size = 12)

p4

DimPlot_scCustom(combined, reduction = "umap", group.by = "sample") + 
  theme_bw(base_size = 10) +
  ggtitle("Sample annotation") +
  theme(legend.position = "right")

```

# decoupleR
```{r}

library(decoupleR)

# TF–target network
net <- get_collectri(
  organism = "human",
  split_complexes = FALSE
)

# Use SCT normalized data
mat <- LayerData(combined[["RNA"]], layer = "scale.data")

# Run ULM
acts <- run_ulm(
  mat = mat,
  net = net,
  .source = "source",
  .target = "target",
  .mor = "mor",
  minsize = 5
)

acts

# Extract ulm and store it in tfsulm in pbmc
combined[['tfsulm']] <- acts %>%
                    tidyr::pivot_wider(id_cols = 'source', 
                                       names_from = 'condition',
                                       values_from = 'score') %>%
                    tibble::column_to_rownames('source') %>%
                    Seurat::CreateAssayObject(.)

# Change assay
DefaultAssay(object = combined) <- "tfsulm"

# Scale the data
combined <- Seurat::ScaleData(combined)
combined@assays$tfsulm@data <- combined@assays$tfsulm@scale.data

DefaultAssay(object = combined) <- "RNA"

```

## Plots
```{r}

DefaultAssay(object = combined) <- "tfsulm"

FeaturePlot_scCustom(combined, features = c("TEAD4")) + 
  labs(x = "UMAP1", y = "UMAP2", title = "TEAD4 activity") +
  umap_theme

DefaultAssay(object = combined) <- "RNA"

FeaturePlot_scCustom(combined, features = c("TEAD4")) + 
  labs(x = "UMAP1", y = "UMAP2", title = "TEAD4 expression") +
  umap_theme

```

# CytoTRACE2

```{r}

set.seed(1234)
library(BiocParallel)
library(CytoTRACE2)

# counts
cts <- GetAssayData(combined, assay="RNA", layer="counts")

# run CytoTRACE2
ct <- cytotrace2(cts, is_seurat=FALSE, species="human", seed=1234,ncores = 5)

# add to metadata
combined$CytoTRACE2 <- ct$CytoTRACE2_Score
combined$CytoTRACE2_Potency <- factor(ct$CytoTRACE2_Potency)

```

# Cell-Cycle Scoring
```{r}

# A list of cell cycle markers, from Tirosh et al, 2015, is loaded with Seurat.  We can
# segregate this list into markers of G2/M phase and markers of S phase
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

combined <- CellCycleScoring(combined, s.features = s.genes, g2m.features = g2m.genes, set.ident = F)

```

# Adding CopyKat

```{r}

sample_to_patient <- c(
  "GSM5219674_LJQ-T"  = "P1",
  "GSM5219675_GBG-T"  = "P2",
  "GSM5219676_LYB-T1" = "P3",
  "GSM5219677_LYB-T2" = "P4",
  "GSM5219678_CYD-T"  = "P5",
  "GSM5219679_CYZ-T"  = "P6",
  "GSM5219680_XMS"    = "P7",
  "GSM5219681_ZYQ"    = "P8",
  "GSM5219682_TGS"    = "P9"
)

copykat_P1 <- vroom::vroom("output/tables/P1_copykat_prediction.txt")
copykat_P2 <- vroom::vroom("output/tables/P2_copykat_prediction.txt")
copykat_P3 <- vroom::vroom("output/tables/P3_copykat_prediction.txt")
copykat_P4 <- vroom::vroom("output/tables/P4_copykat_prediction.txt")
copykat_P5 <- vroom::vroom("output/tables/P5_copykat_prediction.txt")
copykat_P6 <- vroom::vroom("output/tables/P6_copykat_prediction.txt")
copykat_P7 <- vroom::vroom("output/tables/P7_copykat_prediction.txt")
copykat_P8 <- vroom::vroom("output/tables/P8_copykat_prediction.txt")
copykat_P9 <- vroom::vroom("output/tables/P9_copykat_prediction.txt")

copykat_list <- list(
  P1 = copykat_P1,
  P2 = copykat_P2,
  P3 = copykat_P3,
  P4 = copykat_P4,
  P5 = copykat_P5,
  P6 = copykat_P6,
  P7 = copykat_P7,
  P8 = copykat_P8,
  P9 = copykat_P9
)

copykat_all <- dplyr::bind_rows(copykat_list)

meta <- combined@meta.data
meta$cell.names <- rownames(meta)

merged <- dplyr::inner_join(meta, copykat_all, by = "cell.names")

rownames(merged) <- merged$cell.names

combined@meta.data <- merged

meta_ep <- Epithelial_obj@meta.data
meta_ep$cell.names <- rownames(meta_ep)

merged <- merge(meta_ep, copykat_all, by = "cell.names")

rownames(merged) <- merged$cell.names

Epithelial_obj@meta.data <- merged

```

# Annotation with reference HCLA Azimuth:lungref
```{r}
# Use terminal, with proper env 

library(Seurat)
library(Azimuth)
library(future)
plan("sequential")              

ref_data <- tibble(SeuratData::AvailableData())
options(timeout = 10000)
SeuratData::InstallData("lungref")

combined <- qs_read(file = "GSE171145_processed.qs2", nthreads = 10)

options(future.globals.maxSize = 1000 * 1024^4)   # start with 2 GiB
combined <- RunAzimuth(combined, reference = "lungref")

meta_to_remove <- grep("\\.score$|^mapping\\.score$", colnames(combined@meta.data), value = TRUE)
combined@meta.data <- combined@meta.data[, !colnames(combined@meta.data) %in% meta_to_remove]

combined <- DietSeurat(
  combined,
  assays = "RNA",
  dimreducs = c("pca","harmony", "umap"),   # or c("harmony", "umap")
  graphs = NULL
)

qs_save(object = combined,compress_level = 8,file = "GSE171145_processed.qs2", nthreads = 10)


table(combined$predicted.ann_level_1)

table(combined$predicted.ann_finest_level)

table(combined$predicted.ann_level_1, combined$predicted.ann_finest_level)

```

# Save the standard analysis
```{r}

qs_save(object = combined,compress_level = 8,file = "output/objects/GSE171145_CopyKat.qs2", nthreads = 10)

combined <- qs_read(file = "output/objects/GSE171145_CopyKat.qs2", nthreads = 10)

```


# Score C1

This is the core transfer step from the CSS project into the Maynard cohort: the same UP and DOWN signature logic is used to build a continuous C1 plasticity score in tumor epithelial cells.
```{r}

set.seed(1234)
library(UCell)
library(BiocParallel)

# parallel
BPPARAM <- MulticoreParam(workers = 2)
register(BPPARAM)

C1_Core_signature <- vroom::vroom("input/tables/C1_Core_signature.csv")

C1_Core_DOWN <- C1_Core_signature |>
  filter(C1_Core_direction == "DOWN")

C1_Core_UP <-  C1_Core_signature |>
  filter(C1_Core_direction == "UP")

# Score the refined UP and DOWN signatures.
# The final exported score is C1_Core; raw UCell helper columns are removed after smoothing.

# C1 EpiCore scoring with UCell

# Set a reproducible seed for scoring.
set.seed(1234)

# Load the scoring packages used in this block.
library(UCell)
library(BiocParallel)

# Register the BiocParallel backend for UCell scoring.

BPPARAM <- MulticoreParam(workers = 9)
register(BPPARAM)

# Define small helpers for z-scoring and signature checks.

zscore <- function(x) {
  as.numeric(scale(x))
}

check_signature <- function(gene_vector, object_genes, signature_name) {
  genes_input <- unique(na.omit(gene_vector))
  genes_found <- intersect(genes_input, object_genes)
  genes_lost  <- setdiff(genes_input, object_genes)

  message(
    signature_name, ": ",
    length(genes_found), "/", length(genes_input),
    " genes found in object."
  )

  if (length(genes_found) < 5) {
    stop(
      signature_name,
      " has fewer than 5 genes present in the object. ",
      "The score would be unstable."
    )
  }

  list(
    genes_found = genes_found,
    genes_lost  = genes_lost
  )
}

# Build C1 EpiCore signatures

object_genes <- rownames(combined)

down_checked <- check_signature(
  gene_vector    = C1_Core_DOWN$gene,
  object_genes   = object_genes,
  signature_name = "DOWN_Signature"
)

up_checked <- check_signature(
  gene_vector    = C1_Core_UP$gene,
  object_genes   = object_genes,
  signature_name = "UP_Signature"
)

gene_list <- list(
  DOWN_Signature = down_checked$genes_found,
  UP_Signature   = up_checked$genes_found
)

# Optional report of genes missing from the Seurat object
missing_signature_genes <- list(
  DOWN_Signature_missing = down_checked$genes_lost,
  UP_Signature_missing   = up_checked$genes_lost
)

missing_signature_genes

# Add UCell scores

combined <- AddModuleScore_UCell(
  obj      = combined,
  features = gene_list,
  BPPARAM  = BPPARAM
)

# Smooth UCell scores in Harmony space
# Smoothing is performed in Harmony space, not UMAP space.
# This gives a neighborhood-level estimate of UP and DOWN activity.

combined <- SmoothKNN(
  obj             = combined,
  signature.names = c("UP_Signature_UCell", "DOWN_Signature_UCell"),
  reduction       = "harmony"
)

# Compute smoothed C1 EpiCore score
# Formula:
#   C1_Core = scale(scale(UP UCell kNN) - scale(DOWN UCell kNN))
#
# Interpretation:
#   - Higher values = stronger C1-like epithelial state
#   - Lower values  = stronger retained differentiated / lineage-like state
#
# Note:
#   This is a smoothed neighborhood-level score.
#   It is ideal for visualization, gradients, and trajectory/state analysis.

md <- combined@meta.data

up_z <- zscore(md$UP_Signature_UCell_kNN)
down_z <- zscore(md$DOWN_Signature_UCell_kNN)

combined$C1_Core_raw <- up_z - down_z
combined$C1_Core <- zscore(up_z - down_z)

# Clean temporary UCell columns
# Keep only the final C1_Core score.
# Remove both raw UCell and smoothed kNN UCell intermediate columns.

cols_remove <- c(
  "UP_Signature_UCell",
  "DOWN_Signature_UCell",
  "UP_Signature_UCell_kNN",
  "DOWN_Signature_UCell_kNN"
)

cols_remove <- intersect(cols_remove, colnames(combined@meta.data))

combined@meta.data <- combined@meta.data[
  ,
  setdiff(colnames(combined@meta.data), cols_remove),
  drop = FALSE
]

# Stop backend

bpstop(BPPARAM)

```


# Plot
```{r}

# Plot main continuous scores on UMAP
score_umap_plot <- FeaturePlot_scCustom(
  combined,
  features = c("C1_Core", "CytoTRACE2","G2M.Score","S.Score"),
  reduction = "umap",
  na_cutoff = NA
) &
  scale_color_viridis_c(option = "magma") &
  labs(x = "UMAP1", y = "UMAP2") &
  umap_theme

score_umap_plot

ggsave(
  filename = "output/figures/score_umap_plot.png",
  plot = score_umap_plot,
  dpi = 600,
  width = 12,
  height = 8
)

# Plot main continuous scores on UMAP
C1_core_umap_plot <- FeaturePlot_scCustom(
  combined,
  features = c("C1_Core"),
  reduction = "umap",
  na_cutoff = NA
) &
  scale_color_viridis_c(option = "magma") &
  labs(x = "UMAP1", y = "UMAP2", title = "C1-Core") &
  umap_theme

C1_core_umap_plot

ggsave(
  filename = "output/figures/C1_core_UMAP.png",
  plot = C1_core_umap_plot,
  dpi = 600,
  width = 10,
  height = 6
)


DimPlot_scCustom( combined, reduction = "umap", group.by = "seurat_clusters") +
  labs(x = "UMAP1", y = "UMAP2", title = "Leiden Clusters") +
  umap_theme

ggsave(
  filename = "output/figures/seurat_clusters.png",
  dpi = 600,
  width = 10,
  height = 6
)
 
DimPlot_scCustom( combined, reduction = "umap",group.by = "predicted.ann_level_1",label = T,label.box = T)  +
  labs(x = "UMAP1", y = "UMAP2", title = "Annotation (HLCA)") +
  umap_theme +
  NoLegend()

ggsave(
  filename = "output/figures/AnnotationLV1.png",
  dpi = 600,
  width = 12,
  height = 8
)


DimPlot_scCustom( combined, reduction = "umap",group.by = "predicted.ann_finest_level",label = T,label.box = T,label.size = 3) +   
  labs(x = "UMAP1", y = "UMAP2", title = "Annotation finest (HLCA)") +
  umap_theme +
  NoLegend()

ggsave(
  filename = "output/figures/AnnotationFINEST.png",
  dpi = 600,
  width = 24,
  height = 18
)


DimPlot_scCustom( combined, reduction = "umap",group.by = "copykat.pred",label = T,label.box = T,label.size = 3) +   
  labs(x = "UMAP1", y = "UMAP2", title = "CopyKat tumor annotation") +
  umap_theme +
  NoLegend()

ggsave(
  filename = "output/figures/Annotation_CopyKat.png",
  dpi = 600,
  width = 12,
  height = 8
)


DimPlot_scCustom(seurat_object = combined,group.by = "tmb_group") +   
  labs(x = "UMAP1", y = "UMAP2", title = "TMB annotation") +
  umap_theme +
  scale_color_npg()

ggsave(
  filename = "output/figures/Annotation_tmb_group.png",
  dpi = 600,
  width = 12,
  height = 8
)

DimPlot_scCustom(seurat_object = combined,group.by = "smoking") +   
  labs(x = "UMAP1", y = "UMAP2", title = "Smoking annotation") +
  umap_theme +
  scale_color_aaas()

ggsave(
  filename = "output/figures/Annotation_smoking.png",
  dpi = 600,
  width = 12,
  height = 8
)

p_A <- FeaturePlot_scCustom(
  combined,
  features = "C1_Core",
  reduction = "umap",
  na_cutoff = NA
) &
  scale_color_viridis_c(option = "magma") &
  labs(x = "UMAP1", y = "UMAP2", title = "C1-Core") &
  umap_theme

p_B <- DimPlot_scCustom(
  combined,
  reduction = "umap",
  group.by = "copykat.pred"
) +
  labs(x = "UMAP1", y = "UMAP2", title = "CopyKat tumor annotation") +
  umap_theme +
  scale_color_bmj()

p_C <- DimPlot_scCustom(
  seurat_object = combined,
  reduction = "umap",
  group.by = "tmb_group"
) +
  labs(x = "UMAP1", y = "UMAP2", title = "TMB annotation") +
  umap_theme +
  scale_color_npg()

p_D <- DimPlot_scCustom(
  seurat_object = combined,
  reduction = "umap",
  group.by = "smoking"
) +
  labs(x = "UMAP1", y = "UMAP2", title = "Smoking annotation") +
  umap_theme +
  scale_color_aaas()

p_umap_annotations <- (p_A | p_B) / (p_C | p_D) +
  plot_annotation(tag_levels = "A") &
  theme(
    plot.tag = element_text(size = 18, face = "bold")
  )

p_umap_annotations

ggsave(
  filename = "output/figures/Full_umap_annotations_patchwork.png",
  plot = p_umap_annotations,
  width = 14,
  height = 10,
  dpi = 600
)

```

# Epithelial only

```{r}

epithelial_labels <- c(
  "AT1",
  "AT2",
  "AT2 proliferating",
  "Basal resting",
  "Club (nasal)",
  "Club (non-nasal)",
  "Goblet (nasal)",
  "Ionocyte",
  "Multiciliated (non-nasal)",
  "SMG duct",
  "SMG mucous",
  "SMG serous (bronchial)",
  "Suprabasal",
  "Transitional Club-AT2",
  "Tuft"
)

Epithelial_obj <- subset(
  combined,
  subset = predicted.ann_finest_level %in% epithelial_labels
)

Epithelial_obj$predicted.ann_finest_level <- droplevels(
  factor(Epithelial_obj$predicted.ann_finest_level)
)

table(Epithelial_obj$predicted.ann_finest_level)
ncol(Epithelial_obj)

Epithelial_obj <- NormalizeData(Epithelial_obj)
Epithelial_obj <- FindVariableFeatures(Epithelial_obj, nfeatures = 5000)
Epithelial_obj <- ScaleData(Epithelial_obj)
Epithelial_obj <- RunPCA(Epithelial_obj)

PC_Plotting(Epithelial_obj, dim_number = 1)
PC_Plotting(Epithelial_obj, dim_number = 2)
ElbowPlot_scCustom(Epithelial_obj, ndims = 50)

Epithelial_obj <- RunHarmony(
  object = Epithelial_obj,
  group.by.vars = "sample",
  reduction.use = "pca",
  dims.use = 1:42
)

Epithelial_obj <- FindNeighbors(
  Epithelial_obj,
  reduction = "harmony",
  dims = 1:42
)

plan(multicore, workers = 4)
Epithelial_obj <- RunUMAP(
  object = Epithelial_obj,
  reduction = "harmony",
  dims = 1:42,
  umap.method = "uwot2",
  seed.use = 1234
)

```


# Plot
```{r}

# Plot main continuous scores on UMAP
score_umap_plot <- FeaturePlot_scCustom(
  Epithelial_obj,
  features = c("C1_Core", "CytoTRACE2","G2M.Score","S.Score"),
  reduction = "umap",
  na_cutoff = NA
) &
  scale_color_viridis_c(option = "magma") &
  labs(x = "UMAP1", y = "UMAP2") &
  umap_theme

score_umap_plot

ggsave(
  filename = "output/figures/Epithelial_obj_score_umap_plot.png",
  plot = score_umap_plot,
  dpi = 600,
  width = 12,
  height = 8
)

# Plot main continuous scores on UMAP
C1_core_umap_plot <- FeaturePlot_scCustom(
  Epithelial_obj,
  features = c("C1_Core"),
  reduction = "umap",
  na_cutoff = NA
) &
  scale_color_viridis_c(option = "magma") &
  labs(x = "UMAP1", y = "UMAP2", title = "C1-Core") &
  umap_theme

C1_core_umap_plot

ggsave(
  filename = "output/figures/Epithelial_obj_C1_core_UMAP.png",
  plot = C1_core_umap_plot,
  dpi = 600,
  width = 10,
  height = 6
)


DimPlot_scCustom( Epithelial_obj, reduction = "umap", group.by = "seurat_clusters") +
  labs(x = "UMAP1", y = "UMAP2", title = "Leiden Clusters") +
  umap_theme

ggsave(
  filename = "output/figures/Epithelial_obj_seurat_clusters.png",
  dpi = 600,
  width = 10,
  height = 6
)



DimPlot_scCustom( Epithelial_obj, reduction = "umap",group.by = "predicted.ann_finest_level",label = T,label.box = T,label.size = 3) +   
  labs(x = "UMAP1", y = "UMAP2", title = "Annotation finest (HLCA)") +
  umap_theme +
  NoLegend()

ggsave(
  filename = "output/figures/Epithelial_obj_AnnotationFINEST.png",
  dpi = 600,
  width = 24,
  height = 18
)

DimPlot_scCustom( Epithelial_obj, reduction = "umap",group.by = "copykat.pred") +   
  labs(x = "UMAP1", y = "UMAP2", title = "CopyKat tumor annotation") +
  umap_theme

ggsave(
  filename = "output/figures/Epithelial_obj_Annotation_CopyKat.png",
  dpi = 600,
  width = 12,
  height = 8
)


DimPlot_scCustom(seurat_object = Epithelial_obj,group.by = "tmb_group") +   
  labs(x = "UMAP1", y = "UMAP2", title = "TMB annotation") +
  umap_theme +
  scale_color_npg()

ggsave(
  filename = "output/figures/Epithelial_obj_Annotation_tmb_group.png",
  dpi = 600,
  width = 12,
  height = 8
)

DimPlot_scCustom(seurat_object = Epithelial_obj,group.by = "smoking") +   
  labs(x = "UMAP1", y = "UMAP2", title = "Smoking annotation") +
  umap_theme +
  scale_color_aaas()

ggsave(
  filename = "output/figures/Epithelial_obj_Annotation_smoking.png",
  dpi = 600,
  width = 12,
  height = 8
)

library(patchwork)
library(ggsci)

p_A <- FeaturePlot_scCustom(
  Epithelial_obj,
  features = "C1_Core",
  reduction = "umap",
  na_cutoff = NA
) &
  scale_color_viridis_c(option = "magma") &
  labs(x = "UMAP1", y = "UMAP2", title = "C1-Core") &
  umap_theme

p_B <- DimPlot_scCustom(
  Epithelial_obj,
  reduction = "umap",
  group.by = "copykat.pred"
) +
  labs(x = "UMAP1", y = "UMAP2", title = "CopyKat tumor annotation") +
  umap_theme +
  scale_color_bmj()

p_C <- DimPlot_scCustom(
  seurat_object = Epithelial_obj,
  reduction = "umap",
  group.by = "tmb_group"
) +
  labs(x = "UMAP1", y = "UMAP2", title = "TMB annotation") +
  umap_theme +
  scale_color_npg()

p_D <- DimPlot_scCustom(
  seurat_object = Epithelial_obj,
  reduction = "umap",
  group.by = "smoking"
) +
  labs(x = "UMAP1", y = "UMAP2", title = "Smoking annotation") +
  umap_theme +
  scale_color_aaas()

p_umap_annotations <- (p_A | p_B) / (p_C | p_D) +
  plot_annotation(tag_levels = "A") &
  theme(
    plot.tag = element_text(size = 18, face = "bold")
  )

p_umap_annotations

ggsave(
  filename = "output/figures/epithelial_umap_annotations_patchwork.png",
  plot = p_umap_annotations,
  width = 14,
  height = 10,
  dpi = 600
)

DefaultAssay(Epithelial_obj) <- "RNA"

p_A <- FeaturePlot_scCustom(
  Epithelial_obj,
  features = "C1_Core",
  reduction = "umap",
  na_cutoff = NA
) &
  scale_color_viridis_c(option = "magma") &
  labs(x = "UMAP1", y = "UMAP2", title = "C1-Core") &
  umap_theme

p_B <- FeaturePlot_scCustom(
  Epithelial_obj,
  features = "TEAD4",
  reduction = "umap",
  na_cutoff = NA
) &
  labs(x = "UMAP1", y = "UMAP2", title = "TEAD4 expression") &
  umap_theme

DefaultAssay(Epithelial_obj) <- "tfsulm"

p_C <- FeaturePlot_scCustom(
  Epithelial_obj,
  features = "TEAD4",
  reduction = "umap",
  na_cutoff = NA
) &
  labs(x = "UMAP1", y = "UMAP2", title = "TEAD4 activity") &
  umap_theme

DefaultAssay(Epithelial_obj) <- "RNA"

p_umap_annotations <- (p_A | p_B) / (p_C) +
  plot_annotation(tag_levels = "A") &
  theme(
    plot.tag = element_text(size = 18, face = "bold")
  )

p_umap_annotations

ggsave(
  filename = "output/figures/epithelial_TEAD4_umap_annotations_patchwork.png",
  plot = p_umap_annotations,
  width = 12,
  height = 10,
  dpi = 600
)

```

# Show violin of diffent metadata
```{r}

library(FSA)
library(ggpubr)

df <- Epithelial_obj@meta.data |>
  as_tibble() |>
  mutate(
    mutation = case_when(
      egfr == "L858R" ~ "EGFR-L858R",
      egfr == "19Del" ~ "EGFR-19Del",
      egfr == "Wild" ~ "Other / EGFR-wild",
      TRUE ~ paste0("EGFR-", egfr)
    ),
    patient_id = factor(patient_id),
    mutation = factor(
      mutation
    ),
    stage = factor(stage),
    predicted.ann_finest_level = factor(predicted.ann_finest_level)
  )

pairwise.wilcox.test(
  df$C1_Core,
  df$egfr,
  p.adjust.method = "BH"
)

# 1. STATISTICS
pairwise.wilcox.test(df$C1_Core, df$predicted.ann_finest_level, p.adjust.method = "BH")

library(rstatix)

pwt <- df %>%
  wilcox_test(C1_Core ~ predicted.ann_finest_level) %>%
  adjust_pvalue(method = "BH") %>%
  add_significance()

pwt

write_csv(pwt, "output/tables/Epithelial_obj_C1_wilcox_test_cells.csv")

dunn_res <- dunnTest(C1_Core ~ predicted.ann_finest_level, data = df, method = "bh")

sig_dunn <- dunn_res$res %>%
  as_tibble() %>%
  separate(Comparison, c("group1","group2"), sep = " - ") %>%
  filter(P.adj < 0.05) %>%
  arrange(desc(Z)) %>%
  mutate(direction = ifelse(Z > 0, "higher", "lower"))

sig_dunn

write_csv(sig_dunn, "output/tables/Epithelial_obj_C1_dunn_cells.csv")


# 2. CELL TYPE PLOTS

library(dplyr)
library(ggplot2)
library(ggpubr)
library(RColorBrewer)
library(patchwork)

dir.create("output/figures", recursive = TRUE, showWarnings = FALSE)

cell_levels <- df %>%
  group_by(predicted.ann_finest_level) %>%
  summarise(med_C1 = median(C1_Core, na.rm = TRUE), .groups = "drop") %>%
  arrange(med_C1) %>%
  pull(predicted.ann_finest_level)

pastel_cols <- setNames(
  colorRampPalette(brewer.pal(9, "Pastel1"))(length(cell_levels)),
  cell_levels
)

base_theme <- theme_pubclean(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(size = 11, color = "grey30"),
    legend.position = "none"
  )

p1 <- ggplot(df, aes(predicted.ann_finest_level, C1_Core,
                     fill = predicted.ann_finest_level,
                     color = predicted.ann_finest_level)) +
  geom_violin(trim = TRUE, scale = "width", alpha = 0.75, linewidth = 0.25) +
  geom_boxplot(width = 0.13, outlier.shape = NA, fill = "white",
               color = "black", linewidth = 0.35) +
  scale_fill_manual(values = pastel_cols, drop = FALSE) +
  scale_color_manual(values = pastel_cols, drop = FALSE) +
  coord_flip() +
  labs(
    x = "Cell type",
    y = "C1-Core",
    title = "All cell types",
    subtitle = "Azimuth::Lung v2 (HLCA) | Kruskal–Wallis p < 2e-16"
  ) +
  base_theme

focus_cells <- c("AT2", "AT2 proliferating", "Basal resting", "Suprabasal")

df_focus <- df %>%
  filter(predicted.ann_finest_level %in% focus_cells) %>%
  mutate(predicted.ann_finest_level = factor(predicted.ann_finest_level, levels = focus_cells))

p2 <- ggplot(df_focus, aes(predicted.ann_finest_level, C1_Core,
                           fill = predicted.ann_finest_level,
                           color = predicted.ann_finest_level)) +
  geom_violin(trim = TRUE, scale = "width", alpha = 0.75, linewidth = 0.25) +
  geom_boxplot(width = 0.15, outlier.shape = NA, fill = "white",
               color = "black", linewidth = 0.35) +
  stat_compare_means(
    comparisons = list(
      c("AT2", "AT2 proliferating"),
      c("Basal resting", "Suprabasal")
    ),
    method = "wilcox.test",
    label = "p.format",
    size = 4
  ) +
  scale_fill_manual(values = pastel_cols, drop = FALSE) +
  scale_color_manual(values = pastel_cols, drop = FALSE) +
  labs(
    x = "Cell type",
    y = "C1-Core",
    title = "Focused epithelial populations",
    subtitle = "Wilcoxon tests between selected groups"
  ) +
  base_theme

comb <- (p1 | p2) +
  plot_layout(widths = c(1.35, 1)) +
  plot_annotation(
    title = "C1-Core distribution across Azimuth lung cell annotations — GSE171145",
    theme = theme(
      plot.title = element_text(face = "bold", size = 17),
      plot.subtitle = element_text(size = 12, color = "grey30")
    )
  )

p1
p2
comb

ggsave("output/figures/C1_Core_by_celltype.png", p1, width = 12, height = 8, dpi = 600)
ggsave("output/figures/C1_Core_by_celltype_focus.png", p2, width = 10, height = 6, dpi = 600)
ggsave("output/figures/C1_Core_by_celltype_Combined.png", comb, width = 15, height = 8, dpi = 600)


# 3. ONCOGENIC MUTATION
pairwise.wilcox.test(df$C1_Core, df$mutation, p.adjust.method = "BH")
kruskal.test(C1_Core ~ mutation, data = df)

dunn_res <- dunnTest(C1_Core ~ mutation, data = df, method = "bh")

sig_dunn <- dunn_res$res %>%
  as_tibble() %>%
  separate(Comparison, c("group1","group2"), sep = " - ") %>%
  filter(P.adj < 0.05) %>%
  arrange(desc(Z)) %>%
  mutate(direction = ifelse(Z > 0, "higher", "lower"))

sig_dunn

write_csv(sig_dunn,"output/tables/C1_dunn_mutation.csv")

library(gt)
table_mutation <- sig_dunn %>%
  arrange(Z) %>%
  gt() %>%
  tab_header(
    title = "Dunn’s Post-hoc Test Results",
    subtitle = "Pairwise comparisons between Oncogenic Driver Mutation"
  ) %>%
  fmt_number(
    columns = c(Z),
    decimals = 2
  ) %>%
  fmt_scientific(
    columns = c(P.unadj, P.adj),
    decimals = 2
  ) %>%
  cols_label(
    group1 = "Cell type 1",
    group2 = "Cell type 2",
    Z = "Z score",
    P.unadj = "P (unadjusted)",
    P.adj = "P (adjusted)",
    direction = "Effect direction"
  ) %>%
  tab_style(
    style = list(
      cell_text(weight = "bold")
    ),
    locations = cells_column_labels()
  ) %>%
  opt_table_lines() %>%
  tab_options(
    table.font.size = px(12)
  )

table_mutation

table_mutation |> gtsave(filename = "output/figures/tab_mutation.png")

mutation_comparisons <- list(
  c("EGFR-19del", "EGFR-L858R"),
  c("EGFR-19del", "Other / EGFR-wild"),
  c("EGFR-L858R", "Other / EGFR-wild")
)

p3 <- ggplot(df, aes(mutation, C1_Core, fill = mutation)) +
  geom_violin(trim = TRUE, scale = "width", alpha = 0.75, linewidth = 0.25) +
  geom_boxplot(
    width = 0.13,
    outlier.shape = NA,
    fill = "white",
    color = "black",
    linewidth = 0.35
  ) +
  stat_compare_means(
    comparisons = mutation_comparisons,
    method = "wilcox.test",
    p.adjust.method = "BH",
    label = "p.format",
    size = 4
  ) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    x = "Mutation",
    y = "C1-Core",
    title = "C1-Core by mutation mutation status",
    subtitle = "Pairwise Wilcoxon tests with BH adjustment"
  ) +
  base_theme

p3

ggsave(
  "output/figures/C1_Core_by_mutation.png",
  p3,
  width = 10,
  height = 6,
  dpi = 600
)

```

```{r}

plot_wilcox_violin <- function(
  data,
  x_col,
  y_col,
  title = NULL,
  x_lab = NULL,
  y_lab = NULL,
  palette = "Set2",
  output_file = NULL,
  width = 10,
  height = 8
) {
  x_col <- rlang::ensym(x_col)
  y_col <- rlang::ensym(y_col)

  plot_df <- data %>%
    dplyr::filter(!is.na(!!x_col), !is.na(!!y_col)) %>%
    dplyr::mutate(!!x_col := factor(!!x_col))

  r_val <- wilcoxonRG(
    x = dplyr::pull(plot_df, !!y_col),
    g = dplyr::pull(plot_df, !!x_col),
    verbose = FALSE
  )

  w_test <- wilcox.test(
    stats::as.formula(paste(rlang::as_string(y_col), "~", rlang::as_string(x_col))),
    data = plot_df
  )

  subtitle_text <- paste0(
    "Wilcoxon p = ", format.pval(w_test$p.value, digits = 3),
    " | RG = ", round(r_val, 2)
  )

  p <- ggplot(plot_df, aes(x = !!x_col, y = !!y_col, fill = !!x_col)) +
    geom_violin(trim = TRUE, alpha = 0.25, color = NA) +
    geom_boxplot(
      width = 0.15,
      fill = "white",
      color = "black",
      outlier.shape = NA,
      linewidth = 0.4
    ) +
    scale_fill_brewer(palette = palette) +
    theme_bw(base_size = 14) +
    labs(
      x = x_lab %||% rlang::as_string(x_col),
      y = y_lab %||% rlang::as_string(y_col),
      title = title %||% paste(rlang::as_string(y_col), "by", rlang::as_string(x_col)),
      subtitle = subtitle_text
    ) +
    theme(
      legend.position = "none",
      axis.title = element_text(face = "bold"),
      axis.text = element_text(color = "black")
    )

  if (!is.null(output_file)) {
    ggsave(output_file, p, width = width, height = height, dpi = 600)
  }

  return(p)
}

```

```{r}

p3 <- plot_wilcox_violin(
  data = df,
  x_col = smoking,
  y_col = C1_Core,
  title = "C1-Core by smoking status",
  x_lab = "Smoking status",
  y_lab = "C1-Core",
  output_file = "output/figures/C1_Core_by_smoking.png"
)

p3

p4 <- plot_wilcox_violin(
  data = df,
  x_col = tmb_group,
  y_col = C1_Core,
  title = "C1-Core by TBM status",
  x_lab = "TMB status",
  y_lab = "C1-Core",
  output_file = "output/figures/C1_Core_by_smoking.png"
)

p4


```

```{r}

dir.create("output/figures", recursive = TRUE, showWarnings = FALSE)

library(dplyr)
library(tibble)
library(ggplot2)
library(ggpubr)
library(broom)
library(performance)
library(emmeans)
library(ggrepel)
library(patchwork)
library(RColorBrewer)
library(forcats)

meta <- Epithelial_obj@meta.data |>
  as_tibble() |>
  mutate(
    mutation = case_when(
      egfr == "L858R" ~ "EGFR-L858R",
      egfr == "19Del" ~ "EGFR-19Del",
      egfr == "Wild" ~ "Other / EGFR-wild",
      TRUE ~ paste0("EGFR-", egfr)
    ),
    patient_id = factor(patient_id),
    mutation = factor(
      mutation
    ),
    stage = factor(stage),
    predicted.ann_finest_level = factor(predicted.ann_finest_level)
  )

pastel_cols <- RColorBrewer::brewer.pal(max(3, nlevels(meta$mutation)), "Pastel2")
names(pastel_cols) <- levels(meta$mutation)

patient_summary <- meta |>
  group_by(patient_id, mutation, stage, predicted.ann_finest_level) |>
  summarise(
    median_C1_Core = median(C1_Core, na.rm = TRUE),
    n_cells = n(),
    .groups = "drop"
  ) |>
  filter(n_cells >= 20) |>
  droplevels()

m_patient <- lm(
  median_C1_Core ~ mutation + predicted.ann_finest_level,
  data = patient_summary,
  weights = n_cells
)

broom::tidy(m_patient, conf.int = TRUE)
performance::check_collinearity(m_patient)
emmeans::emmeans(m_patient, ~ mutation)

patient_mutation_summary <- patient_summary |>
  group_by(patient_id, mutation, stage) |>
  summarise(
    median_C1_Core = weighted.mean(median_C1_Core, w = n_cells, na.rm = TRUE),
    n_cells = sum(n_cells),
    .groups = "drop"
  )

p_patient <- ggplot(
  patient_mutation_summary,
  aes(x = mutation, y = median_C1_Core, fill = mutation)
) +
  geom_violin(trim = TRUE, alpha = 0.6, color = NA) +
  geom_boxplot(
    width = 0.13,
    fill = "white",
    color = "grey25",
    outlier.shape = NA,
    linewidth = 0.35
  ) +
  geom_point(
    aes(size = n_cells, color = stage),
    position = position_jitter(width = 0.08, height = 0),
    alpha = 0.95
  ) +
  ggrepel::geom_text_repel(
    aes(label = patient_id),
    size = 3,
    show.legend = FALSE,
    max.overlaps = Inf
  ) +
  scale_fill_manual(values = pastel_cols) +
  scale_size_continuous(range = c(2.5, 6)) +
  ggpubr::theme_pubclean(base_size = 13) +
  labs(
    title = "Patient-level C1_Core by mutation group",
    subtitle = "Each point is one patient",
    x = "Mutation group",
    y = "Weighted median C1_Core",
    color = "Stage",
    size = "Cells"
  ) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey35"),
    axis.text.x = element_text(angle = 25, hjust = 1)
  )

coef_df <- broom::tidy(m_patient, conf.int = TRUE) |>
  filter(term != "(Intercept)", !is.na(estimate)) |>
  mutate(
    term_clean = term |>
      gsub("mutation", "Mutation: ", x = _) |>
      gsub("predicted.ann_finest_level", "Cell type: ", x = _),
    term_clean = forcats::fct_reorder(term_clean, estimate)
  )

p_coef <- ggplot(coef_df, aes(x = estimate, y = term_clean)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey55") +
  geom_errorbarh(
    aes(xmin = conf.low, xmax = conf.high),
    height = 0.18,
    linewidth = 0.45,
    color = "grey35"
  ) +
  geom_point(size = 2.4, color = "grey20") +
  ggpubr::theme_pubclean(base_size = 12) +
  labs(
    title = "Patient-level model effects",
    subtitle = "Weighted model: median C1_Core ~ mutation group + epithelial subtype",
    x = "Regression coefficient ± 95% CI",
    y = NULL
  ) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey35"),
    axis.text.y = element_text(size = 8)
  )

p_slide <- p_patient | p_coef +
  plot_annotation(
    title = "C1_Core is associated with mutation group and epithelial cell state",
    subtitle = "Patient/cell-state summaries reduce cell-level pseudoreplication and collinearity",
    tag_levels = "A",
    theme = theme(
      plot.title = element_text(face = "bold", size = 16),
      plot.subtitle = element_text(size = 11, color = "grey35"),
      plot.tag = element_text(face = "bold", size = 15)
    )
  )

p_slide

ggsave(
  "output/figures/C1_Core_patient_level_mutation_model_summary.png",
  p_slide,
  width = 15,
  height = 8,
  dpi = 600
)

```


```{r}

qs_save(object = Epithelial_obj,compress_level = 8,file = "output/objects/Epithelial_obj_CopyKat.qs2", nthreads = 10)

Epithelial_obj <- qs_read(file = "output/objects/Epithelial_obj.qs2", nthreads = 10)

```


